---
title: "10_train_wrangle"
author: "Jake Bergren"
date: "November 16, 2018"
output: html_document
---

Wrangling, scaling data for downstream machine learning.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#install.packages("RSQLite")
#install.packages("dplyr")
#install.packages("BiocManager")

library(dplyr)
library(DBI)
#library(mygene)
library(RSQLite)

```

First things first, load the data.

```{r}
con <- dbConnect(RSQLite::SQLite(),dbname="D:/GitHub/BMIN503_Final_Project/protein_training.db")
# This takes a little while
train_df <- dbGetQuery(con, "SELECT * FROM train")
# gram_num_df <- dbGetQuery(con, "SELECT protein, COUNT(gram_num) as gram_total FROM train GROUP BY protein")
# Get list of amyloids to map with
# amyloid_df <- dbGetQuery(con, "SELECT * FROM amyloid")
dbDisconnect(con)
```

Wrangle the data for machine learning purposes! Factor, scale, drop columns of low utility, add the labels for training and eval.

```{r}
# Drop protein, gram_num, U_num, U_perc and the amyloid_num cols
#train_df$protein <- NULL
train_df$U_num <- NULL
train_df$U_perc <- NULL
train_df$amyloid_1 <- NULL
train_df$amyloid_2 <- NULL
train_df$amyloid_3 <- NULL
train_df$amyloid_4 <- NULL
train_df$amyloid_5 <- NULL

factor_cols <- c("gram_1","gram_2","gram_3","gram_4","gram_5")
# Make a consistent amino acid map, factor gram columns
amino_acid_factors <- c("R"=0,"H"=1,"K"=2,"D"=3,"E"=4,"S"=5,"T"=6,"N"=7,"Q"=8,"C"=9,"U"=10,"G"=11,"P"=12,"A"=13,
                        "V"=14,"I"=15,"L"=16,"M"=17,"F"=18,"Y"=19,"W"=20)

train_df[factor_cols] <- lapply(train_df[factor_cols], function(i){i = amino_acid_factors[i]})

# Scale certain columns
non_scale_cols <- c("protein","gram_num","amyloid","gram_1","gram_2","gram_3","gram_4","gram_5")
scale_cols <- names(train_df)[!names(train_df) %in% non_scale_cols]
train_df[,scale_cols] <- lapply(train_df[,scale_cols], scale)
```

Now we record the scaled table in our sqlite database.

```{r}
con <- dbConnect(RSQLite::SQLite(),dbname="D:/GitHub/BMIN503_Final_Project/protein_training.db")
dbWriteTable(con, "train_scaled", train_df)
dbDisconnect(con)
```

Finally we mark each n-gram with it's amyloid marker. This will take awhile.

```{r}
con <- dbConnect(RSQLite::SQLite(),dbname="D:/GitHub/BMIN503_Final_Project/protein_training.db")
dbExecute(con,
  "ALTER TABLE train_scaled ADD amy INTEGER;"
)
# THIS TAKES A LONG TIME!
dbExecute(con,
  "UPDATE train_scaled 
  SET amy = (
    SELECT amyloid as amy
    FROM amyloid
    WHERE protein = train_scaled.protein
    );"
)
dbDisconnect(con)

```

And now the data are ready for training!
