---
title: "16_final_train_feature_selection"
author: "Jake Bergren"
date: "November 28, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Feature Selection

The sheer size of my training set begs for feature slimming prior to training. This notebook will be used to determine which features to utilize.

```{r}
library(dplyr)
library(tibble)
library(tidyr)
library(DBI)
library(RSQLite)
library(ggplot2)
library(reshape2)
library(pwr)
```

Lets find which columns are numeric, then generate some sample indices for sampling.

```{r eval=TRUE, fig.width=7, fig.height=10}
# First lets get the numeric Columns.
con <- dbConnect(RSQLite::SQLite(),dbname="D:/GitHub/BMIN503_Final_Project/training_final.db")
cols_df <- dbGetQuery(
  con, 
  "SELECT * FROM train LIMIT 100"
  )

n_row <- first(dbGetQuery(
  con,
  "SELECT COUNT(*) FROM train"
))
dbDisconnect(con)

# Get rep sample size
samp_size <- ceiling(pwr.r.test(r=0.01, sig.level = 0.01, power = 0.99)$n)
train_ind <- seq(n_row)

sample_inds <- list()
for (i in seq(1,10)){
  sample_inds[i] <- list(sample(train_ind, samp_size, replace=FALSE))
}

num_cols_index <- unlist(lapply(cols_df, is.numeric))
num_cols <- colnames(cols_df[,num_cols_index])
```

Now we can sample the dataset and get the columns that should be dropped.

```{r}
get_drop_cols <- function(dframe){
  drop_cols_final<- c()
  # Get ranked list of correlates (num of connections in correlation "graph")
  rank_list <- dframe %>%
    melt(measure.vars=c("V1","V2")) %>%
    select(value) %>%
    group_by(value) %>%
    tally() %>%
    arrange(desc(n))
  # Get list of unique values to check for dropping
  unique_list <- rank_list %>%
    select(value) %>%
    distinct()

  while (length(unique_list$value)>0){
    # Get top val
    top_val <- first(rank_list$value)
    
    # Mark values connected to top value for dropping
    drop_cols <- dframe %>%
     filter(V1==top_val | V2==top_val) %>%
     melt(measure.vars=c("V1","V2")) %>%
     filter(value!=top_val) %>%
     select(value)

    drop_cols_final <- c(drop_cols_final, drop_cols$value)
    # Build new rank list
    rank_list <- filter(rank_list, value %in% drop_cols_final == FALSE & value!=top_val)
    unique_list <- filter(unique_list, value %in% drop_cols_final == FALSE & value!=top_val)
}
  return(drop_cols_final)
}

get_sample <- function(cols, row_ids){
  cols_str <- paste(cols, collapse = ", ")
  row_ids_str <- paste(unlist(row_ids), collapse = ", ")
  query <- paste0(
    "SELECT ",cols_str," FROM train WHERE rowid IN (",row_ids_str,")"
    )

  con <- dbConnect(RSQLite::SQLite(),dbname="D:/GitHub/BMIN503_Final_Project/training_final.db")
  # This takes a little while
  sample <- dbGetQuery(
    con, 
    query
    )
  dbDisconnect(con)
  sample$prot_U_num <- NULL
  sample$prot_U_perc <- NULL
  return(sample)
}

col_results <- c()
pb = txtProgressBar(0, 10, initial=0)

for (i in seq(1,10)){
  # Get sample
  sample <- get_sample(num_cols, sample_inds[i])

  # Calculate correlation, with replacement of NA with zero
  corr_mat <- sample %>%
    mutate_if(is.numeric, funs(replace(., is.na(.), 0))) %>%
    cor %>%
    melt
  
  # Mark correlations as significant if they are 0.95 or greater
  corr_mat$sig <- ifelse(corr_mat$value >= 0.95, 1, 0)

  # Build the "graph" of correlates
  corr_mat_unique <- corr_mat[, c("Var1","Var2","sig")] %>%
    filter(sig == 1 & Var1!=Var2 & !(grepl("amyloid",Var1)==TRUE & grepl("amyloid",Var2)==TRUE)) %>%
    select("Var1","Var2") %>%
    apply(1,sort) %>%
    t() %>%
    apply(2, as.character) %>%
    as.data.frame() %>%
    distinct() %>%
    group_by(V1, V2) %>%
    tally()
  
  # For some reason melt turns everything into factors, lets reverse that
  corr_mat_unique[,c("V1","V2")] <- lapply(corr_mat_unique[,c("V1","V2")], as.character)
  
  # Get the columns we can remove from the dataset for training
  col_results <- c(col_results, get_drop_cols(corr_mat_unique))
  setTxtProgressBar(pb, i)
}

col_res_df <- as.data.frame(col_results, stringsAsFactors = FALSE) %>%
  group_by(col_results) %>%
  tally()

colnames(col_res_df) <- c("column_to_drop", "n_times_marked_for_deletion")

ggplot(col_res_df) +
  geom_bar(aes(x=column_to_drop, y=n_times_marked_for_deletion), stat="identity") +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) +
  ylim(0, 10)
```

That's good that the vast majority of numerical values aren't correlative, that means there's a lot of good information in there!

Lets remove the few variables that are correlative.

```{r}
drop_corr_cols_final <- unlist(filter(col_res_df, n_times_marked_for_deletion >= 5)$column_to_drop)
drop_corr_cols_final
```

Now for the categoricals.

```{r eval=TRUE, fig.width=7, fig.height=10}
chisqmatrix <- function(x) {
  names = colnames(x);  num = length(names)
  m = matrix(nrow=num, ncol=num, dimnames=list(names,names))
  for (i in 1:(num-1)) {
    for (j in (i+1):num) {
      p_val <- chisq.test(x[,i],x[,j], simulate.p.value = TRUE, B = 100)$p.value
      m[i,j] = p_val
      m[j,i] = p_val
    }
  }
  return (m)
}

ignore_non_num_cols <- c("protein")
non_num_col_names <- as.data.frame(colnames(cols_df[, !num_cols_index]), stringsAsFactors = FALSE)
colnames(non_num_col_names) <- c("colnames")
non_num_col_names <- filter(non_num_col_names, colnames %in% ignore_non_num_cols == FALSE)$colnames

csqr_col_results <- c()
pb = txtProgressBar(0, 10, initial=0)

for (i in seq(1,10)){
#  # Get sample
  sample <- as.data.frame(lapply(get_sample(non_num_col_names, sample_inds[i]), factor))

  chsq_mat = sample %>%
    chisqmatrix %>%
    melt

  chsq_mat$sig <- ifelse(chsq_mat$value > 0.05, 1, 0)
  if (nrow(filter(chsq_mat, sig==1))==0){
    setTxtProgressBar(pb, i)
    next
  }
  # Build the "graph" of correlates
  chsqr_mat_unique <- chsq_mat[, c("Var1","Var2","sig")] %>%
    filter(sig == 1 & Var1!=Var2) %>%
    select("Var1","Var2") %>%
    apply(1,sort) %>%
    t() %>%
    apply(2, as.character) %>%
    as.data.frame() %>%
    distinct()# %>%
    #group_by(V1, V2) %>%
    #tally()
  
  # For some reason melt turns everything into factors, lets reverse that
  chsqr_mat_unique[,c("V1","V2")] <- lapply(chsqr_mat_unique[,c("V1","V2")], as.character)
  # Get the columns we can remove from the dataset for training
  csqr_col_results <- c(csqr_col_results, get_drop_cols(corr_mat_unique))
  setTxtProgressBar(pb, i)
  
}

length(csqr_col_results)
```

Looks like none of the categorical values are dependent on any other value.

```{r}
# Write the cols to a file for later
write(drop_corr_cols_final, file="drop_cols.txt", sep="\n")
```