---
title: "2017-09-21 - Identify Cohort"
author: "Adam Dziorny"
date: "September 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r nzConn, echo=FALSE}
source('F_rodbc_cdw.R')

nzConn <- makeConn(user = 'dziornya', pwd = getPassword('CHOPDW'), 
                   server = 'binbiaclv2.chop.edu', db = 'CDWPRD')

# Source the paging function to notify me when results are back
source('F_pageXDS.r')
```

## Identify the Cohort

We need to identify a cohort of PICU / CICU patients who are admitted between 7/1/2012 and 7/1/2017.  

Each section below contains a chunk of SQL code, which will be combined below to a single query and run.

### Cohort

The starting cohort is all VISIT_KEY values with any PICU or CICU VISIT_EVENTs of type ADMISSION, TRANSFER_IN, TRANSFER_OUT, DISCHARGE or CENSUS within the time frame.  Having a VISIT_KEY in this cohort means that at some point during the patient's stay they were in the PICU or CICU within the date range specified.

From the `CDW_DICTIONARY` table (where `DICT_CAT_KEY = 10050`), mapping to `VISIT_EVENT`.`DICT_ADT_EVENT_KEY`:

DICT_KEY | SRC_ID | Name
-------- | ------ | ----
20666 | 1 | Admission
20667 | 2 | Discharge
20668 | 3 | Transfer In
20669 | 4 | Transfer Out
20670 | 5 | Patient Update
20671 | 6 | Census
20672 | 7 | Hospital Outpatient
150739 | 8 | Leave of Absence Out
150740 | 9 | Leave of Absence Return
515921 | 10 | Leave of Absence Census

From the `CDW_DICTIONARY` table (where `DICT_CAT_KEY = 100118`), mapping to `VISIT_EVENT`.`DICT_EVENT_SUBTYPE_KEY`:

DICT_KEY | SRC_ID | NAME
-------- | ------ | ----
29720 | 1 | Original
29721 | 2 | Cancelled
29722 | 3 | Update

From the `DEPARTMENT` table:

DEPT_KEY | DEPT_ID | Name | Abbrev
-------- | ------- | ---- | ------
5149 | 34 | 7 EAST PICU | 7EP
5307 | 36 | 6 SOUTH TOWER | 6ST
4920 | 43 | 7 SOUTH TOWER | 7ST
7512 | 123 | 7 WEST PICU | 7WPICU
4778 | 66 | 6 NORTHEAST | 6NE
5166 | 58 | 6 NORTHWEST | 6NWS
5223 | 101001069 | PERIOP COMPLEX | PERIOP

From the CDW Data Dictionary, the variable `EFF_EVENT_DT` is the instant when the event was supposed to have happened.

```{r cohort}
q.cohort <- strwrap("
SELECT 
  DISTINCT vis_evnt.VISIT_KEY
FROM 
  CDW.VISIT_EVENT vis_evnt  
    INNER JOIN CDW.CDW_DICTIONARY event_type ON vis_evnt.DICT_ADT_EVENT_KEY = event_type.dict_key
        AND event_type.src_id IN (1,2,3,4,6) 
    INNER JOIN CDW.CDW_DICTIONARY event_sub_type ON vis_evnt.DICT_EVENT_SUBTYPE_KEY = event_sub_type.dict_key
        AND event_sub_type.src_id <> 2 
    INNER JOIN CDW.DEPARTMENT dept ON vis_evnt.DEPT_KEY = dept.DEPT_KEY
WHERE 
  dept.DEPT_ID IN (34, 43, 123, 36) 
  AND vis_evnt.VISIT_KEY NOT IN (-1, 0)
  AND vis_evnt.PAT_KEY NOT IN (0)
  AND vis_evnt.EFF_EVENT_DT BETWEEN DATE('07/01/2012') AND DATE('07/01/2017')
  ", width=10000)
```

### All ADT Events

Now for this group of DISTINCT VISIT_KEYs we obtain *ALL* of the ADT events for each VISIT_KEY, from *ALL* departments.  This is important because there are many patient movement workflows that require knowing other departments. 

Example patient movement:
- ED -> PICU -> Floor -> PICU
- ED -> PICU -> PICU (new room) -> OR -> PICU
- PICU -> OR -> PICU -> Floor -> CICU -> Floor
- ED -> Floor -> OR -> PICU

Our goal is to capture each individual entry / exit to the PICU and CICU, excluding OR transfers (of DEPT_ID 101001069 [PERIOP], 58 and 66 [Cardiac ORs]). OR transfers should be considered part of the same PICU / CICU stay.

Set the EVENT_DIR (or the direction of the event) as either move-in or move-out, grouping ADMIT/Transfer In as well as DISCHARGE/Transfer Out.

Set the DEPT_GRP as either PICU (34,43,123), CICU, PERIOP or OTHER.

The Bed was just included here for debugging purposes. It adds nothing to the query and could be removed.

event_type SRC_ID: 
- 1, 3 are move-in
- 2, 4 are move-out

Departments:
- 34, 43, 123 are PICU
- 36 is CICU
- 101001069, 58, 66 are PERIOP, Cardiac OR, Cardiac Cath

Event_sub_type: 2 is CANCELLED

```{r all_adt}
q.all_adt <- strwrap("
SELECT
  vis_evnt.VISIT_KEY,
  vis_evnt.VISIT_EVENT_KEY,
  pat.DOB AS PAT_DOB,
  pat.SEX AS PAT_SEX,
  vis_evnt.EFF_EVENT_DT as IN_DT,
  event_type.DICT_NM AS EVENT_TYPE,
  CASE WHEN event_type.src_id IN (1,3) THEN 1 
       WHEN event_type.src_id IN (2,4) THEN 0 
	   END AS EVENT_DIR,
  dept.DEPT_NM as d_name,
  dept.DEPT_KEY AS d_key,
  CASE WHEN dept.DEPT_ID IN (34,43,123) THEN 1 
       WHEN dept.DEPT_ID IN (36) THEN 2 
	   WHEN dept.DEPT_ID IN (101001069,58,66) THEN 3 
	   ELSE 0 
	   END AS dept_grp,
  bed.BED_NM AS bed_name,
  vis_evnt.PAT_KEY,
  dict_svc.DICT_NM AS PAT_SVC
FROM 
  cohort
    INNER JOIN CDW.VISIT_EVENT vis_evnt ON cohort.VISIT_KEY = vis_evnt.VISIT_KEY
    INNER JOIN CDW.PATIENT pat ON pat.PAT_KEY = vis_evnt.PAT_KEY
    INNER JOIN CDW.CDW_DICTIONARY event_type ON vis_evnt.DICT_ADT_EVENT_KEY = event_type.dict_key
        AND event_type.src_id IN (1,2,3,4) 
    INNER JOIN CDW.CDW_DICTIONARY event_sub_type ON vis_evnt.DICT_EVENT_SUBTYPE_KEY = event_sub_type.dict_key
        AND event_sub_type.src_id <> 2 
    INNER JOIN CDW.CDW_DICTIONARY dict_svc ON dict_svc.DICT_KEY = vis_evnt.DICT_PAT_SVC_KEY
    INNER JOIN CDW.DEPARTMENT dept ON vis_evnt.DEPT_KEY = dept.DEPT_KEY
    INNER JOIN CDW.MASTER_BED bed ON vis_evnt.BED_KEY = bed.BED_KEY
  ", width=10000)
```

```{r, eval=TRUE, echo=FALSE}
# Temporary query from where we have built the cohort so far
q.temp <- strwrap(paste("
WITH 
cohort AS (", q.cohort, "), 
all_adt AS (", q.all_adt, ")
SELECT * FROM all_adt", sep="\n"), width=100000)

cohort.1 <- sqlQuery(nzConn, q.temp, 
                  believeNRows=TRUE, 
                  stringsAsFactors=FALSE)
rm(q.temp)

cat(sprintf('Unique PAT_KEY values: %d\nUnique Visit_KEY values: %d\n',
            length(unique(cohort.1$PAT_KEY)),
            length(unique(cohort.1$VISIT_KEY))))
```

### Remove PERIOP Encounters

All queries are now done partitioned by VISIT_KEY sorted by in_dt and event_dir. The values of event_dir were chosen to ensure that the "OUT" will always come before the "IN" when sorting in this manner (recall they have the exact same DT timestamp).

While we have all of the ADT events for each VISIT_KEY we grab the HOSP_ADMIT_DT and HOSP_DC_DT, so we can calculate an overall LOS.

Next we flag all peri-op jumps:
- Either: CURRENT and LAG dept are PERIOP, CURRENT is "Transfer Out" and LAG is "Transfer In"
- OR     CURRENT and LEAD dept are PERIOP, CURRENT is "Transfer In" and LEAD is "Transfer Out"

If either of these conditions is satisfied, flag the case as OUT_PERIOP meaning the patient is out at a periop procedure. These will be removed in the next step.

```{r remv_periop}
q.remv_periop <- strwrap("
SELECT 
  *,
  FIRST_VALUE(in_dt) OVER (PARTITION BY visit_key ORDER BY in_dt) AS HOSP_ADMIT_DT,
  FIRST_VALUE(in_dt) OVER (PARTITION BY visit_key ORDER BY in_dt DESC) AS HOSP_DC_DT,
  CASE WHEN (LAG(dept_grp) OVER (PARTITION BY visit_key ORDER BY in_dt,event_dir) IN (3) AND 
             dept_grp IN (3) AND
			 LAG(event_type) OVER (PARTITION BY visit_key ORDER BY in_dt,event_dir) IN ('Transfer In') AND
			 event_type IN ('Transfer Out')) OR 
			(LEAD(dept_grp) OVER (PARTITION BY visit_key ORDER BY in_dt,event_dir) IN (3) AND 
             dept_grp IN (3) AND
			 LEAD(event_type) OVER (PARTITION BY visit_key ORDER BY in_dt,event_dir) IN ('Transfer Out') AND
			 event_type IN ('Transfer In')) THEN 'Y' END AS out_periop
FROM all_adt
  ", width=10000)
```

```{r, eval=TRUE, echo=FALSE}
# Temporary query from where we have built the cohort so far
q.temp <- strwrap(paste("
WITH 
cohort AS (", q.cohort, "), 
all_adt AS (", q.all_adt, "),
remv_periop AS (", q.remv_periop, ")
SELECT * FROM remv_periop ORDER BY VISIT_KEY, in_dt, event_dir", sep="\n"), width=100000)

cohort.2 <- sqlQuery(nzConn, q.temp, 
                  believeNRows=TRUE, 
                  stringsAsFactors=FALSE)
rm(q.temp)

cat(sprintf('Unique PAT_KEY values: %d\nUnique Visit_KEY values: %d\n',
            length(unique(cohort.2$PAT_KEY)),
            length(unique(cohort.2$VISIT_KEY))))
```

### Flag Same Departments

Now that we've excluded all PERIOP movement, we set a flag if the LAG and CURRENT are of the same DEPT_GRP (meaning the patient was just moving around the DEPT_GRP).

```{r same_dept}
q.same_dept <- strwrap("
SELECT 
  *,
  CASE WHEN LAG(dept_grp) OVER (PARTITION BY visit_key ORDER BY in_dt,event_dir) IN (dept_grp) THEN 'Y' END AS same_dept
FROM remv_periop
WHERE out_periop IS NULL
  ", width=10000)
```

```{r, eval=TRUE, echo=FALSE}
# Temporary query from where we have built the cohort so far
q.temp <- strwrap(paste("
WITH 
cohort AS (", q.cohort, "), 
all_adt AS (", q.all_adt, "),
remv_periop AS (", q.remv_periop, "),
same_dept AS (", q.same_dept, ")
SELECT * FROM same_dept ORDER BY VISIT_KEY, in_dt, event_dir", sep="\n"), width=100000)

cohort.3 <- sqlQuery(nzConn, q.temp, 
                  believeNRows=TRUE, 
                  stringsAsFactors=FALSE)
rm(q.temp)

cat(sprintf('Unique PAT_KEY values: %d\nUnique Visit_KEY values: %d\n',
            length(unique(cohort.3$PAT_KEY)),
            length(unique(cohort.3$VISIT_KEY))))
```

### Remove the Duplicate Departments

At this point we have removed PERIOP, and we exclude all DEPT_GRP except 1 or 2 (PICU or CICU). Each "transit" through that DEPT_GRP runs from the VISIT_IN_DT, or when SAME_DEPT is NULL, down through the VISIT_OUT_DT, or the last "Y" before a NULL under SAME_DEPT.  These are set in separate rows here, and combined in the next step.

```{r visit_dt_set}
q.visit_dt_set <- strwrap("
SELECT 
  *,
  CASE WHEN SAME_DEPT IS NULL THEN IN_DT END AS VISIT_IN_DT,
  CASE WHEN LEAD(SAME_DEPT) OVER (PARTITION BY visit_key ORDER BY in_dt,event_dir) IS NULL THEN IN_DT END AS VISIT_OUT_DT
FROM same_dept
WHERE DEPT_GRP IN (1,2)
  ", width=10000)
```

```{r, eval=TRUE, echo=FALSE}
# Temporary query from where we have built the cohort so far
q.temp <- strwrap(paste("
WITH 
cohort AS (", q.cohort, "), 
all_adt AS (", q.all_adt, "),
remv_periop AS (", q.remv_periop, "),
same_dept AS (", q.same_dept, "),
visit_dt_set AS (", q.visit_dt_set, ")
SELECT * FROM visit_dt_set ORDER BY VISIT_KEY, in_dt, event_dir", sep="\n"), width=100000)

cohort.4 <- sqlQuery(nzConn, q.temp, 
                  believeNRows=TRUE, 
                  stringsAsFactors=FALSE)
rm(q.temp)

cat(sprintf('Unique PAT_KEY values: %d\nUnique Visit_KEY values: %d\n',
            length(unique(cohort.4$PAT_KEY)),
            length(unique(cohort.4$VISIT_KEY))))
```

### Combine Lines 

At this stage we have rows with alternating NULL / VISIT_IN_DT and opposing alternating VISIT_OUT_DT / NULL values.  We must combine VISIT_IN and VISIT_OUT DT's into a single row for each entry-exit to an ICU.  We chose to keep the VISIT_IN_DT row, and when VISIT_OUT_DT is NULL, get the LEAD(VISIT_OUT_DT) - in other words, just move it up a row.

```{r comb_lines}
q.comb_lines <- strwrap("
SELECT
  *,
  CASE WHEN VISIT_OUT_DT IS NULL THEN LEAD(VISIT_OUT_DT) OVER (PARTITION BY visit_key ORDER BY in_dt,event_dir) END AS OUT_DT
FROM visit_dt_set
WHERE VISIT_IN_DT IS NOT NULL OR VISIT_OUT_DT IS NOT NULL
  ", width=10000)
```

```{r, eval=TRUE, echo=FALSE}
# Temporary query from where we have built the cohort so far
q.temp <- strwrap(paste("
WITH 
cohort AS (", q.cohort, "), 
all_adt AS (", q.all_adt, "),
remv_periop AS (", q.remv_periop, "),
same_dept AS (", q.same_dept, "),
visit_dt_set AS (", q.visit_dt_set, "),
comb_lines AS (", q.comb_lines, ")
SELECT * FROM comb_lines ORDER BY VISIT_KEY, in_dt, event_dir", sep="\n"), width=100000)

cohort.5 <- sqlQuery(nzConn, q.temp, 
                  believeNRows=TRUE, 
                  stringsAsFactors=FALSE)
rm(q.temp)

cat(sprintf('Unique PAT_KEY values: %d\nUnique Visit_KEY values: %d\n',
            length(unique(cohort.5$PAT_KEY)),
            length(unique(cohort.5$VISIT_KEY))))
```

### Final Cohort

Restrict our final output to those rows with non-NULL IN_DT and OUT_DT and limit our SELECT criteria to get our final set of rows.  Note that VISIT_KEY is *NOT* DISTINCT because patients might move in and out of the PICU / CICU (or even both) in a single hospital visit.  This captures each movement and records IN_DT and OUT_DT. Results are sorted by VISIT_KEY and IN_DT.

```{r final_cohort}
q.final_cohort <- strwrap("
SELECT 
  VISIT_KEY,
  PAT_KEY,
  PAT_SVC,
  PAT_DOB,
  PAT_SEX,
  CASE WHEN DEPT_GRP = 1 THEN 'PICU'
       WHEN DEPT_GRP = 2 THEN 'CICU' END AS DEPT,
  VISIT_IN_DT AS IN_DT,
  OUT_DT,
  HOSP_ADMIT_DT,
  HOSP_DC_DT
FROM comb_lines
WHERE VISIT_IN_DT IS NOT NULL AND OUT_DT IS NOT NULL
  ", width=10000)
```

Now we evaluate the above expressions to build the final.cohort data frame.

```{r}
q.final_query <- strwrap(paste("
WITH 
cohort AS (", q.cohort, "), 
all_adt AS (", q.all_adt, "),
remv_periop AS (", q.remv_periop, "),
same_dept AS (", q.same_dept, "),
visit_dt_set AS (", q.visit_dt_set, "),
comb_lines AS (", q.comb_lines, "),
final_cohort AS (", q.final_cohort, ")
SELECT * FROM final_cohort ORDER BY VISIT_KEY, IN_DT", sep="\n"), width=100000)

final.cohort <- sqlQuery(nzConn, q.final_query, 
                  believeNRows=TRUE, 
                  stringsAsFactors=FALSE)
```

This is saved out, along with all the build strings (as they will be needed to join the lab tables), to an output data file:

```{r}
save(cohort.1,cohort.2,cohort.3,cohort.4,cohort.5, 
     file = 'D_interim_cohorts')

save(final.cohort,
     q.cohort,q.all_adt,q.remv_periop,
     q.same_dept,q.visit_dt_set,q.comb_lines,
     q.final_cohort,
     file = 'D_final_cohort_w_queries')

# And page to notify that we are done
pageMe(profileID = 'E7582503', msg='Finished gathering cohort!', displayResult = F)
```



