---
title: "Split VS Data"
author: "Adam Dziorny"
date: "November 25, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)

source('F_pageXDS.r')
```

## Load File

First we load the pre-processed VS file, created from `2017-11-25 - Pre-Process Vital Signs.Rmd`.  This file is ~ 150 MB and loads into about 12 GB of RAM.

```{r loader}
tic <- proc.time()
load('D_vs_preprocessed')
toc <- proc.time() - tic

cat(sprintf('Elapsed time to load (min): %0.2f\n',
            toc[3]/60))
```

```{r, echo=F}
rm(tic,toc)
```

### Most Commonly Filled

Let's look at the most commonly filled VS rows - we sort by `DISP_NM` and summarize by N and %.  We then display the head of the sorted list.

```{r}
count.by.nm <-
  vs.df %>%
  dplyr::group_by(DISP_NM) %>%
  dplyr::summarize(n = n(),
                   perc = n() / nrow(vs.df) * 100) %>%
  dplyr::arrange(desc(n))

kable(head(count.by.nm, n=20), col.names = c('Row Name', 'N', '%'), digits = 2)
```

```{r, echo=FALSE}
rm(count.by.nm)
```

### Pulse

Let's export the Pulse after we convert the `MEAS_VAL` into a numeric field.  

```{r}
vs.hr.df <-
  vs.df %>%
  dplyr::filter(DISP_NM == 'Pulse') %>%
  dplyr::mutate(VAL = as.numeric(MEAS_VAL)) %>%
  dplyr::select(VISIT_KEY, IN_DT, OUT_DT, VAL, FS_REC_DT, REC_DT, ENTRY_DT)

save(vs.hr.df, file='D_vs_hr')
```

What's the histogram of the values of this field look like?

```{r}
vs.hr.df %>%
  ggplot() +
  geom_histogram(aes(x = VAL), bins = 30) +
  xlab('HR (beats/min)') + 
  ylab('Count') +
  theme_bw()
```

There appear to be some outliers here, so we will drill down into the summary of values:

```{r}
summary(vs.hr.df$VAL)
```

Let's pull off the 1st and 99th percentile values.  Below are those values, and the histogram resulting from plotting every value between the tails:

```{r}
tails <- quantile(vs.hr.df$VAL, c(0.01, 0.99), na.rm = T)

cat(sprintf('1st Percentile: %0.f\n99th Percentile: %0.f\n',
            tails[1], tails[2]))
```

```{r}
vs.hr.df %>%
  dplyr::filter(VAL >= tails[1] & VAL <= tails[2]) %>%
  ggplot() +
  geom_histogram(aes(x = VAL), bins = 20) +
  xlab('HR (beats/min)') + 
  ylab('Count') +
  theme_bw()
```

```{r, echo=FALSE}
rm(vs.hr.df, tails)
```

### SpO2

Now let's export the SpO2 after we convert the `MEAS_VAL` into a numeric field.  

```{r}
vs.spo2.df <-
  vs.df %>%
  dplyr::filter(DISP_NM == 'SpO2') %>%
  dplyr::mutate(VAL = as.numeric(MEAS_VAL)) %>%
  dplyr::select(VISIT_KEY, IN_DT, OUT_DT, VAL, FS_REC_DT, REC_DT, ENTRY_DT)

save(vs.spo2.df, file='D_vs_spo2')
```

What's the histogram of the values of this field look like?

```{r}
vs.spo2.df %>%
  ggplot() +
  geom_histogram(aes(x = VAL), bins = 30) +
  xlab('SpO2 (%)') + 
  ylab('Count') +
  theme_bw()
```

```{r echo=FALSE}
rm(vs.spo2.df)
```
