---
title: "Pre-Process Vital Signs"
author: "Adam Dziorny"
date: "November 25, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(scales)

source('F_pageXDS.r')
```

## Load File

First we load the raw VS file, created from `2017-11-24 - Gather Vital Signs.Rmd`.  This raw file is ~ 100 MB and loads into about 10 GB of RAM.

```{r loader}
tic <- proc.time()
load('D_vs_df')
toc <- proc.time() - tic

cat(sprintf('Elapsed time to load (min): %0.2f\n',
            toc[3]/60))
```

How many rows are in our dataset:

```{r}
cat(sprintf('Millions of rows: %0.2f\n',
            nrow(vs.df)/1e6))
```

## Convert Date-Times to POSIX

We used the `as.is` clause in our query to prevent inappropriate POSIX conversion of the column, so now we have to post-hoc convert the five date-time columns to POSIX format.  This will shrink the dataset (eventually), but the conversion is slow.

```{r}
tic <- proc.time()

# Full DT
vs.df[,2] <- as.POSIXct(vs.df[,2], format='%Y-%m-%d %H:%M:%S') # IN_DT
vs.df[,3] <- as.POSIXct(vs.df[,3], format='%Y-%m-%d %H:%M:%S') # OUT_DT

vs.df[,8] <- as.POSIXct(vs.df[,8], format='%Y-%m-%d %H:%M:%S') # REC_DT
vs.df[,9] <- as.POSIXct(vs.df[,9], format='%Y-%m-%d %H:%M:%S') # ENTRY_DT

# Just Date
vs.df[,7] <- as.POSIXct(vs.df[,7], format='%Y-%m-%d') # FS_REC_DT

toc <- proc.time() - tic

cat(sprintf('Elapsed time to convert DTs (min): %0.2f\n',
            toc[3]/60))
```

## Remove Out-of-Range INs

We next will remove the rows with IN date-times out of range of our IRB-designated date range: `2012-07-01` through `2017-07-01`:

```{r}
outOfRange <- vs.df$IN_DT < as.POSIXct('2012-07-01') | 
  vs.df$IN_DT > as.POSIXct('2017-07-01')

vs.df <- vs.df[!outOfRange,]

cat(sprintf('The number of records removed: %d\n',sum(outOfRange)))
```

## Add Factors

We add a factor on the Display Name of the Flowsheet Row:

```{r}
vs.df$DISP_NM <- factor(vs.df$DISP_NM)
```

## Save Preprocessed Data

And now we save out the pre-processed vital sign data:

```{r}
save(vs.df, file = 'D_vs_preprocessed')

# And page to notify that we are done
pageMe(profileID = 'E7582503', msg='Finished preprocessing VS!', displayResult = F)
```