---
title: "Gather Lab Values"
author: "Adam Dziorny"
date: "September 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r nzConn, echo=FALSE}
source('F_rodbc_cdw.R')
source('F_lab_ids.R')

nzConn <- makeConn(user = 'dziornya', pwd = getPassword('CHOPDW'), 
                   server = 'binbiaclv2.chop.edu', db = 'CDWPRD')

# Source the paging function to notify me when results are back
source('F_pageXDS.r')
```

# Lab Values

In this notebook we collect lab value data.  First we must load the cohort information we have previously saved:

```{r}
load(file = 'D_final_cohort_w_queries')
```

## Combine Queries

Included with the above load are all of the sub-queries (the small parts of the large `WITH` query) onto which we will join the individual lab result queries.  The labs of interest are stored in a separate `F_lab_ids.R` file, which contains a data frame:

```{r}
head(lab.ids)
```

```{r}
summary(lab.ids)
```

Each `base` represents a result component, and each `proc.id` / `proc.name` represents a procedure.  Some procedures use different `rslt.id` values for the same `base` - for example, PO2 is `rslt.id` = 29 in an arterial blood gas sample, but `rslt.id` = 4112 in a venous blood gas sample.  We desire to represent both as PO2 (but maintain the procedure for subsequent analysis).

First we use the function `makeQuery()`, which composes a data frame of all queries for each element in the `lab.ids` data frame.  We then use loops to combine these queries (and a `join` query for each) into to cohort `WITH` statement.  Finally we union all of the lab queries together.  

```{r combine_queries}
query.df <- makeQuery()

query <- strwrap(paste("
WITH 
cohort AS (",       q.cohort, "), 
all_adt AS (",      q.all_adt, "), 
remv_periop AS (",  q.remv_periop, "), 
same_dept AS (",    q.same_dept, "),
visit_dt_set AS (", q.visit_dt_set, "), 
comb_lines AS (",   q.comb_lines, "), 
final_cohort AS (", q.final_cohort, "), ", 
  sep="\n"), width=1000000)

# Loop over each individual query and write out the lab and '_jt' query
for (i in 1 : nrow(query.df)) {
  endLine <- ifelse(i < nrow(query.df),',','')

  query <- paste0(query, 
                  ' ', 'tbl_', query.df[i,'md5'], ' AS (',
                  query.df[i,'query'],'), ',
                  'tbl_', query.df[i,'md5'], '_jt AS (',
                  query.df[i,'joinTable'], ')',
                  endLine, ' ')
}
rm(i, endLine)

# Now loop over all joined tables '_jt' and UNION
for (i in 1 : nrow(query.df)) {
  endLine <- ifelse(i < nrow(query.df),'\n UNION \n','')
  
  query <- paste0(query,
                  'SELECT * FROM ',
                  'tbl_', query.df[i,'md5'], '_jt',
                  endLine)
}
rm(i, endLine)

# Wrap the whole query
query <- strwrap(query, width=1000000)
```

We need to specify an "as-is" vector, a boolean vector which tells the RODBC connection which columns to leave alone (and not attempt to convert to a time class).  The columns of the result table will be:

| Col # | Name | Type |
| ----- | ---- | ---- |
| 1 | `VISIT_KEY` | Integer |
| 2 | `PAT_KEY` | Integer |
| 3 | `PROC_ORD_KEY` | Integer |
| 4 | `LAB_DEPT` | Character (Factor) |
| 5 | `RSLT_DT` | Date-Time |
| 6 | `SPECIMEN_TAKEN_DT` | Date-Time |
| 7 | `VALUE` | Numeric |
| 8 | `PROC_NM` | Character (Factor) |
| 9 | `BASE_NM` | Chracter (Factor) | 
| 10 | `DEPT_GRP` | Character (Factor) |
| 11 | `IN_DT` | Date-Time |
| 12 | `RSLT_DATE` | Date |

Based on this, we would like to leave the Date-Time and Date columns `as-is` and perform offline conversion of characters to dates. All others are `F` in the `as.is` vector. 

```{r}
as.is <- rep(F,12)

as.is[c(5,6,11,12)] <- T
```

## Run Query & Save

Now we run the full query, timing it with `proc.time()` on each end. After it is completed, we save out the results immediately.

```{r}
# Run the query, timing it with proc.time() 
tic <- proc.time()
labs.df <- sqlQuery(nzConn, query, 
                  believeNRows=TRUE, 
                  stringsAsFactors=FALSE,
                  as.is = as.is)
toc <- proc.time() - tic

save(labs.df, file='D_labs_df')

# And page to notify that we are done
pageMe(profileID = 'E7582503', msg='Finished gathering lab values!', displayResult = F)
```

The total query duration is as follows:

```{r}
cat(sprintf('Elapsed time: %0.2f min\n',toc[3]/60))
```