---
title: "BMIN503_Migraine_Pain_Craig"
author: "Sansanee Craig"
date: "November 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```
## Overview

#overview and introduction for your final project. The overview consists of 2-3 sentences summarizing the project and goals. For the introduction, the first paragraph describes the problem addressed, its significance, and some background to motivate the problem. In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff

## Introduction



## Descriptive Statistics

```{r, eval=TRUE}
data_all_rows <- read.csv("data_all_rows.csv")
str(data_all_rows) # 3642 observations of 33 variables
summary(data_all_rows)

sex <-data_all_rows$Sex #2561 F, 1080 M, 1: blank?

race <-data_all_rows$Race.Ethnicity  #H 255, NHB 1487, NHW 1720, Other 179, blank 1
sum(is.na(data_all_rows$Race.Ethnicity)) #0

payer_type <-data_all_rows$Payer.Type #Commercial 2047, MA 1480, missing 114, blank 1
lang <- data_all_rows$Primary.Language #Eng 3573, Non-Eng 68, blank 1
pathway <- data_all_rows$Pathway #No 637, Yes 3004, blank 1


#Ethnicity (not on file on snapshot, not hispanic in demographics), race B, coded as nonhispanic black. Options in Epic: Not Hispanic, Hispanic, Refused / American Indian or Alaska Native, Asian, Black or African American, Indian, Native Hawaiian or Other Pacific Islander, Other, Refused, White

#Language (preferred lang, not on file), coded as english. Options in Epic: Preferred language, spoken language, written language. 

#Find and table all  missing fields for all data
#For Race: Does other also include missing fields? Mandatory epic field?
#Figure out this blank entry
#Continue summarizing descriptive data stats for table 1


```

