---
title: "BMIN503_Migraine_Pain_Craig"
author: "Sansanee Craig"
date: "November 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```
## Overview
Racial disparities in pain management have been documented in the adult and pediatric population. Most studies have focused on pathologies with definitive diagnostic testing, such as appendicitis, long-bone fractures, and sickle cell disease. Few studies have assessed variability of pain management in pediatric migraine. The purpose of this project is to utilize secondary analysis of EHR data to determine if race/ethnicity-based differences exist in the management of pediatric migraine pain in the emergency room at the Children’s Hospital of Philadelphia (CHOP). 

## Introduction
Pain management cross-sects most fields of medical practice. In the case of migraine pain, a patient’s care team may comprise a primary care physician and a neurologist, with ER and hospitalist providers added during refractory migraine episodes. Alleviating pain efficiently, safely, and equally for all patients is the ultimate goal during an emergency room visit. In the few pediatric studies to date, however, some sociodemographic disparities have been shown to exist in pediatric headache care across the United States.  
In beginning to study this issue, thoughtful data queries in conjunction with data analysts is vital as the choices behind codifying race and ethnicity alone could significantly alter study results. Workflow analysis during direct observations in the ER provides insight into missing field factors, such as non-mandatory data entry in assigning a pain scale/score or completing demographic fields. Nurses’ feedback sheds light on why pain reassessment alerts were being ignored or pain scales not being assigned during registration. Discussions with ER physicians and the ER director provides background information on previous disparities study results and QI projects, identifying most relevant outcome to study, and factors in choosing which statistical result to report. Meeting with a pediatric neurologist, the director of the headache clinic, results in an additional view of focusing QI efforts on giving primary care providers better tools to care for migraine pain in order to decrease unnecessary neurology referrals or ER visits. 
In short, data scientists and quality improvement (QI) teams can contribute to understanding the best methods to study where these disparities lie and how to rapidly conduct improvement projects to decrease known disparities. Clinical informaticists help to facilitate change management by navigating the socio-technical factors in hospital and clinic organizations. Thus, this is a multidisciplinary project with stakeholders spanning the field biomedical and health informatics.



## Descriptive Statistics

```{r, eval=TRUE}
data_all_rows <- read.csv("https://raw.githubusercontent.com/scraig10/BMIN503_Final_Project/master/data_all_rows.csv", header=TRUE)

str(data_all_rows) # 3642 observations of 33 variables
summary(data_all_rows)

sex <-data_all_rows$Sex #2561 F, 1080 M, 1: blank?

race <-data_all_rows$Race.Ethnicity  #H 255, NHB 1487, NHW 1720, Other 179, blank 1
sum(is.na(data_all_rows$Race.Ethnicity)) #0

payer_type <-data_all_rows$Payer.Type #Commercial 2047, MA 1480, missing 114, blank 1
lang <- data_all_rows$Primary.Language #Eng 3573, Non-Eng 68, blank 1
pathway <- data_all_rows$Pathway #No 637, Yes 3004, blank 1


#Ethnicity (not on file on snapshot, not hispanic in demographics), race B, coded as nonhispanic black. Options in Epic: Not Hispanic, Hispanic, Refused / American Indian or Alaska Native, Asian, Black or African American, Indian, Native Hawaiian or Other Pacific Islander, Other, Refused, White

#Language (preferred lang, not on file), coded as english. Options in Epic: Preferred language, spoken language, written language. 

#Find and table all  missing fields for all data
#For Race: Does other also include missing fields? Mandatory epic field?
#Figure out this blank entry
#Continue summarizing descriptive data stats for table 1


```

