---
title: "BMIN503/EPID600 Final Project"
author: "Silvia Canelon"
output: 
  html_document:
    toc: true
    toc_float: true
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE, collapse=TRUE, message=FALSE}
options(width = 400)
```  

### Overview

The project makes use of the freely-available MIMIC-III Clinical Database to investigate the characteristics of pregnant patients admitted to the Beth Israel Deaconess Medical Center, and their inpatient care. This large dataset of over forty thousand patients admitted to critical care units will be used to identify predictors significantly associated with mortality and/or specific cause of death.

Three faculty/staff being consulted for the project:
- Mary Regina Boland; Assistant Professor of Informatics; Department of Biostatistics, Epidemiology, and Informatics
- Ray Bai; Postdoctoral Researcher; Department of Biostatistics, Epidemiology, and Informatics
- Lena Davidson; Data Analyst; Department of Biostatistics, Epidemiology, and Informatics

### Introduction 

The secondary use of Electronic Health Records (EHR) facilitates retrospective research on data collected during the course of patients' clinical care. The data available through the MIMIC-III Clinical Database provides detailed health-related information collected from a diverse population of patients admitted to intensive care units of a large me dical center, some of whom may have been pregnant at the time of admission. Given the nature of intensive care units, any pregnant patients admitted were likely considered high-risk pregnancies before or at the time of admission. In addition, a subset of these pregnant patients will have deceased or suffered other adverse outcomes during their care, information which can be gleaned through analysis of in-hospital mortality and billing-related information recorded in the EHR. Understanding the characteristics of the pregnant patient population, as well as their care, will shed light on patterns related to adverse outcomes experienced during critical care and hopefully provide preliminary data pertinent to the clinical decison-making process.

The analysis and subsequent interpretation of EHR data is interdisciplinary because it combines domains of the biomedical informatics, epidemiology, biostatistics, and clinical fields. Uncovering potential correlations between patient and patient care information and clinical outcomes involves the use of informatics and statistical analysis skills, whereas epidemiology and clinical domains provide the necessary context for defining/refining the analysis approach and interpreting the results. Consultation with all three faculty/staff has confirmed that adverse outcomes among the patient population admitted to critical care are important to understand from a public health perspective, and likely to not be "rare" events given the intensive care unit context, which allows for standard statistical approaches.


### Methods

The Mimic-III Clinic Database contains de-identified health data records for over 40,000 patients admitted to the Beth Israel Deaconess Medical Center of Boston, Massachusetts from 2001 and 2012. The database includes information on demographics, vital sign measurements, laboratory test results, procedures, medications, caregiver notes, imaging reports, billing-related information, and mortality.
Demographic, procedure, mortality, and billing-related information such as International Classification of Disease (ICD) codes, will be used to explore patient characteristics and care while being treated at Beth Isreal Deaconess Medical Center, and mortality studied as an outcome.

#### Connection to server

Packages are loaded and the MIMIC-III dataset is imported from the MySQL database it is stored in.
```{r Connection, include=FALSE, collapse=TRUE, message=FALSE}
rm(list=ls()) # clears the environment

# load packages
pkgs = c("RMySQL", "DBI", "tidyverse", "stringr", "dplyr", "formatR", "lubridate", "tictoc", "ggThemeAssist", "viridis", "GGally") # define packages
lapply(pkgs, library, character.only = TRUE) # load packages

drv = dbDriver("MySQL")

# connecting to the secure server "Curie"
mimic_iii = dbConnect(drv, dbname="clinical_mimic_III", 
                           host="127.0.0.1",
                           port=3307,
                           user=keyring::key_list("Curie")[1,2], #username for mysql
                           password=keyring::key_get("Curie"))  #password for mysql

#on.exit(dbDisconnect(mimic_iii)) # had to comment out in order to knit document
```

#### Importing data

This section imports MIMIC-III data from separate tables, most of them including health records and a couple of them including dictionaries for ICD codes.
```{r Data import, cache=TRUE, collapse=TRUE, message=FALSE}
tic("data import") # takes about 20 secs

rs = dbSendQuery(mimic_iii, "select distinct * from ADMISSIONS;")
data_admissions = fetch(rs, n=-1) #46520
length(unique(data_admissions$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from PATIENTS;")
data_patients = fetch(rs, n=-1) #46520
length(unique(data_patients$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from SERVICES;")
data_services = fetch(rs, n=-1) #46496
length(unique(data_services$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from TRANSFERS;")
data_transfers = fetch(rs, n=-1) #46520
length(unique(data_transfers$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from ICUSTAYS;")
data_icustays = fetch(rs, n=-1) #46476
length(unique(data_icustays$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from DIAGNOSES_ICD;")
data_icd_diag = fetch(rs, n=-1) #46520
length(unique(data_icd_diag$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from PROCEDURES_ICD;")
data_icd_proc = fetch(rs, n=-1) #42214
length(unique(data_icd_proc$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from D_ICD_DIAGNOSES;")
dd_icd_diag = fetch(rs, n=-1) 
dim(unique(dd_icd_diag)) #14567

rs = dbSendQuery(mimic_iii, "select distinct * from D_ICD_PROCEDURES;")
dd_icd_proc = fetch(rs, n=-1) 
dim(unique(dd_icd_proc)) #3882

toc()
```

#### Data tyding

##### Trimming/formatting data
This section selects only the relevant variables from admissionss, expired patient, services, and ICU stay data, and adjusts variable class as needed (namely to `date` or `numeric` classes).
```{r Trimming/formatting, collapse=TRUE, message=FALSE}
# trimming and formatting admissions data
admissions <- data_admissions %>% 
  select(-(ADMISSION_LOCATION:DISCHARGE_LOCATION),-(EDREGTIME:DIAGNOSIS),
         -(LANGUAGE:MARITAL_STATUS),-HAS_CHARTEVENTS_DATA) %>%
  mutate(ADMITTIME = as.Date(ADMITTIME),
         DISCHTIME = as.Date(DISCHTIME))

# formatting dates of death in expired patient data
patients_dod <- data_patients %>% select(-DOD,-DOD_SSN) %>% filter(EXPIRE_FLAG == 1) %>%
  mutate(DOD_HOSP = as.Date(DOD_HOSP))
# joining expired patient data with larger patients data frame
patients <- left_join(data_patients %>% select(-(DOD:DOD_SSN)), patients_dod) %>%
  mutate(DOB = as.Date(DOB))

# trimmming and formatting services data
services <- data_services %>% select(-(TRANSFERTIME:PREV_SERVICE))

# trimmming and formatting ICU stay data
icustays <- data_icustays %>% select(-(FIRST_WARDID:LAST_WARDID)) %>%
  mutate(INTIME = as.Date(INTIME),
         OUTTIME = as.Date(OUTTIME),
         LOS = as.numeric(LOS))
```

##### Joining data
This section joins diagnosis and procedure records to the ICD diagnosis and procedure data dictionaries by `ICD9_CODE`.
```{r Joining diagnosis/procedures, collapse=TRUE, message=FALSE}
# joining diagnosis and procedure data to ICD codes
icd_diag <- left_join(data_icd_diag, dd_icd_diag %>% select(-ROW_ID), by = "ICD9_CODE")
icd_proc <- left_join(data_icd_proc, dd_icd_proc %>% select(-ROW_ID), by = "ICD9_CODE")

# counting diagnosis records with and without proper ICD9_CODEs from the data dictionary
dim(icd_diag) #651047
length(unique(icd_diag$SUBJECT_ID)) #46520
diag_w_code <- icd_diag %>% filter(!(is.na(SHORT_TITLE)))
dim(diag_w_code) #634709
diag_wo_code <- icd_diag %>% filter(is.na(SHORT_TITLE))
dim(diag_wo_code) #16338

# counting procedure records with and without proper ICD9_CODEs from the data dictionary
dim(icd_proc) #240095
length(unique(icd_proc$SUBJECT_ID)) #42214
proc_w_code <- icd_proc %>% filter(!(is.na(SHORT_TITLE)))
dim(proc_w_code) #237948
proc_wo_code <- icd_proc %>% filter(is.na(SHORT_TITLE))
dim(proc_wo_code) #2147
```

This section joins admission records to a table containing patient data, and then to Intensive Care Unit (ICU) stay records. It also record dates and dates of birth to calculate the age at the time of admission (`AGE_ADM`), hospital discharge (`AGE_DIS`), and hospital death (`AGE_DOD`).
```{r Joining admission/patient, collapse=TRUE, message=FALSE}
# joining admission data to adult patient data
patients_adm <- left_join(admissions, patients %>% select(-ROW_ID), by = "SUBJECT_ID") %>%
  mutate(AGE_ADM = round(as.duration(DOB %--% ADMITTIME)/dyears(1),2), # age at admission
         AGE_DIS = round(as.duration(DOB %--% DISCHTIME)/dyears(1),2), # age at discharge
         AGE_DOD = round(as.duration(DOB %--% DOD_HOSP)/dyears(1),2)) %>% # age at death
  filter(AGE_ADM >= 18) # age at admission was used to select records from only adults

# joining ICU data to admission+patient data
patients_adm_icu <- left_join(patients_adm, icustays %>% select(SUBJECT_ID, HADM_ID, LOS),
                              by = c("SUBJECT_ID","HADM_ID")) %>% rename(LOS_ICU = LOS)

# joining ICD data to admission+patient+ICU data
patients_adm_icu_diag <- left_join(patients_adm_icu, icd_diag %>% select(-ROW_ID), 
                               by = c("SUBJECT_ID","HADM_ID"))
patients_adm_icu_proc <- left_join(patients_adm_icu, icd_proc %>% select(-ROW_ID), 
                               by = c("SUBJECT_ID","HADM_ID"))
```

##### Recoding data

This section identifies the different categories present in the dataset `ETHNICITY` variable and then recodes those that could reasonably be binned together in order to increase the sample size in each major ethnicity category. For example, ethnicity subcategory `ASIAN - JAPANESE` was recoded as `ASIAN`. It should be noted that with a large enough dataset with larger representation of each ethnicity subcategory, important nuance could be captured between these subpopulations that experience different experiences with the health care system. This is not possible with datasets that, by default, allow for only a handful of ethnicity categories. Produces primary data frame `patients_recoded`.
```{r Recoding ethnicity, collapse=TRUE, message=FALSE}
# identifying the frequency different ethnicity categories present in the dataset
cat_ethnicity <- table(patients_adm_icu$ETHNICITY) %>% data.frame()

# binning subethnicities into larger ethnicity categories
patients_recoded <- patients_adm_icu %>% 
  mutate(ETHNICITY_BROAD = as.factor(ETHNICITY)) %>%
  mutate(ETHNICITY_BROAD = recode(ETHNICITY_BROAD,
                                  "AMERICAN INDIAN/ALASKA NATIVE" = "AIAN",
                                  "AMERICAN INDIAN/ALASKA NATIVE FEDERALLY RECOGNIZED TRIBE" = "AIAN",
                                  "ASIAN - ASIAN INDIAN" = "ASIAN",
                                  "ASIAN - CAMBODIAN" = "ASIAN",
                                  "ASIAN - CHINESE" = "ASIAN",
                                  "ASIAN - FILIPINO" = "ASIAN",
                                  "ASIAN - JAPANESE" = "ASIAN",
                                  "ASIAN - KOREAN" = "ASIAN",
                                  "ASIAN - OTHER" = "ASIAN",
                                  "ASIAN - THAI" = "ASIAN",
                                  "ASIAN - VIETNAMESE" = "ASIAN",
                                  "BLACK/AFRICAN" = "BLACK",
                                  "BLACK/AFRICAN AMERICAN" = "BLACK",
                                  "BLACK/CAPE VERDEAN" = "BLACK",
                                  "BLACK/HAITIAN" = "BLACK",
                                  "HISPANIC OR LATINO" = "HISPANIC",
                                  "HISPANIC/LATINO - CENTRAL AMERICAN (OTHER)" = "HISPANIC",
                                  "HISPANIC/LATINO - COLOMBIAN" = "HISPANIC",
                                  "HISPANIC/LATINO - CUBAN" = "HISPANIC",
                                  "HISPANIC/LATINO - DOMINICAN" = "HISPANIC",
                                  "HISPANIC/LATINO - GUATEMALAN" = "HISPANIC",
                                  "HISPANIC/LATINO - HONDURAN" = "HISPANIC",
                                  "HISPANIC/LATINO - MEXICAN" = "HISPANIC",
                                  "HISPANIC/LATINO - PUERTO RICAN" = "HISPANIC",
                                  "HISPANIC/LATINO - SALVADORAN" = "HISPANIC",
                                  "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER" = "NHPI",
                                  "WHITE - BRAZILIAN" = "WHITE",
                                  "WHITE - EASTERN EUROPEAN" = "WHITE",
                                  "WHITE - OTHER EUROPEAN" = "WHITE",
                                  "WHITE - RUSSIAN" = "WHITE"))

# identifying the frequency of different ethnicity categories after the recoding efforts
cat_ethnicity_broad <- table(patients_recoded$ETHNICITY_BROAD) %>% data.frame()
```

This section adjusts the class of certain variables in the new `patients_recoded` data frame and recodes mortality and age variables. The MIMIC-III dataset shifts all dates to protect patient confidentiality and all patients 89 years or older are marked as being 300 years or older. These 300 year old ages were adjusted to 89 for easier interpretation.
```{r Formatting and recoding other vars, collapse=TRUE, message=FALSE}
# formatting/recoding insurance, gender, hospital death, and age variables
patients_recoded <- patients_recoded %>% 
  mutate(ADMISSION_TYPE = as.factor(ADMISSION_TYPE),
         GENDER = factor(GENDER),
         MORTALITY = factor(HOSPITAL_EXPIRE_FLAG, levels = c(0,1), labels = c("LIVED","DIED")),
         AGE_ADM = ifelse(AGE_ADM > 300, 89, AGE_ADM),
         AGE_DIS = ifelse(AGE_DIS > 300, 89, AGE_DIS),
         AGE_DOD = ifelse(AGE_DOD > 300, 89, AGE_DOD),
         INSURANCE = as.factor(INSURANCE)) %>%
  mutate(INSURANCE = recode(INSURANCE, 
                            "Government" = "GOV", "Medicaid" = "MEDICAID","Medicare" = "MEDICARE",
                            "Private" = "PRIVATE", "Self Pay" = "SELF PAY"))

# trimming data to remove variables no longer needed
patients_recoded <- patients_recoded %>% select(-ETHNICITY, -HOSPITAL_EXPIRE_FLAG, -EXPIRE_FLAG)
```

#### Delivery codes
This section imports tables of ICD codes for delivery diagnoses and procedures.
```{r Importing delivery codes, collapse=TRUE, message=FALSE}
# Importing all delivery DIAGNOSIS ICD9 codes
codes_delivery_diag = read.csv("delivery_diag_codes_SPC.csv", header = F) %>% 
  mutate(V1 = as.character(V1))
head(codes_delivery_diag)

# adjusting the CODE variable to remove the decimal point
codes_delivery_diag_icd9 <- codes_delivery_diag %>% filter(V2 == "ICD9") %>% 
  mutate(CODE = str_remove(V1, "\\.")) %>% select(CODE)
head(codes_delivery_diag_icd9)

# Importing all delivery PROCEDURE ICD9 codes
codes_delivery_proc = read.csv("delivery_proc_codes_SPC.csv", header = F)
head(codes_delivery_proc)

# adjusting the CODE variable to remove the decimal point
codes_delivery_proc_icd9 <- codes_delivery_proc %>% filter(V2 == "ICD9") %>% 
  mutate(CODE = str_remove(V1, "\\.")) %>% select(CODE)
head(codes_delivery_proc_icd9)
```

This section identifies patient records assigned a delivery ICD diagnosis or procedure code.
```{r Delivery records, collapse=TRUE, message=FALSE}
# Records with delivery diagnosis codes
records_delivery_diag = icd_diag %>% filter(ICD9_CODE %in% codes_delivery_diag_icd9$CODE) %>% 
  mutate(CODE_TYPE = "DIAGNOSIS")
head(records_delivery_diag)

# Records with delivery procedure codes
records_delivery_proc = icd_proc %>% filter(ICD9_CODE %in% codes_delivery_proc_icd9$CODE) %>%
  mutate(CODE_TYPE = "PROCEDURE")
head(records_delivery_proc)

# Joining diagnosis delivery and procedure codes
records_delivery <- rbind(records_delivery_diag, records_delivery_proc) %>% unique()
length(unique(records_delivery$ROW_ID)) #203
length(unique(records_delivery$SUBJECT_ID)) #48
length(unique(records_delivery$HADM_ID)) #48
```
- There are 203 delivery diagnosis or procedure records from 48 patients

#### C-section codes
This section imports tables of ICD codes for cesarean section (C-section) procedures.
```{r Importing C-section codes, collapse=TRUE, message=FALSE}
# Importing all C-section PROCEDURE codes
codes_csection_proc = read.csv("csection_proc_codes_SPC.csv", header = F)
head(codes_csection_proc)
codes_csection_proc_icd9 <- codes_csection_proc %>% filter(V2 == "ICD9") %>%
  mutate(CODE = str_remove(V1, "\\.")) %>% select(CODE)
head(codes_csection_proc_icd9)
```

This section identifies patient records assigned a C-section ICD procedure code.
```{r C-section records, collapse=TRUE, message=FALSE}
# Records with C-section procedure codes
records_csection_proc = icd_proc %>% filter(ICD9_CODE %in% codes_csection_proc_icd9$CODE) %>%
  mutate(CODE_TYPE = "PROCEDURE")
head(records_csection_proc)

records_csection <- records_csection_proc

length(unique(records_csection$ROW_ID)) #38
length(unique(records_csection$SUBJECT_ID)) #37
length(unique(records_csection$HADM_ID)) #37
```
- There are 38 csection procedure codes from 37 patients

#### Joining delivery and C-section
This section adds binary variables to the primary `patients_recoded` dataframe to account for the presence/absence of a delivery or C-section ICD code.
```{r Delivery + C-section as variables, collapse=TRUE, message=FALSE}
# adding in variables for delivery and csection
patients_recoded <- patients_recoded %>%
  mutate(DELIVERY = ifelse(HADM_ID %in% records_delivery$HADM_ID, "DELIVERY", "NO DELIVERY"),
         CSECTION = ifelse(HADM_ID %in% records_csection$HADM_ID, "CSECTION", "NO CSECTION")) %>%
  mutate(DELIVERY = as.factor(DELIVERY),
         CSECTION = as.factor(CSECTION))
```

### Results

#### Descriptive information
```{r GGPairs, warning=FALSE, collapse=TRUE, message=FALSE}
# trimmming data to only the predictor variables of interest
patients_recoded_trim <- patients_recoded %>% 
  select(ADMISSION_TYPE,INSURANCE,GENDER,AGE_ADM,AGE_DOD,LOS_ICU,ETHNICITY_BROAD,DELIVERY,CSECTION, MORTALITY) %>% unique()
dim(unique(patients_recoded_trim)) #54362 unique records and 10 variables of interest

# ggpairs(patients_recoded_trim, mapping = aes(col = MORTALITY), columns = c("GENDER","ETHNICITY_BROAD", "AGE_ADM","AGE_DOD"))
```

#### Visualization
```{r Color palette, collapse=TRUE, message=FALSE}
# Penn and DBEI/CCEB colors
PennBlue = "#1e376c" #DBEI/CCEB
PennRed = "#a90533" #DBEI/CCEB
DarkBlue = "#2e5d97" #DBEI/CCEB
LightBlue = "#2e9ad5" #DBEI/CCEB
Purple = "#7e6b94" #DBEI
Orange = "#d47d3a" #Excel orange
Yellow = "#efbf2c" #Excel yellow 
DeptYellow = "ffd400" #DBEI/CCEB yellow
Beige = "#e8eade" #DBEI/CCEB
OrangeRed = "#fef3742" #DBEI
LightRed = "#d99792" #DBEI
Green = "#83944e" #CCEB
Grey = "#999999"
```

```{r Age, collapse=TRUE, message=FALSE, warning=FALSE}
# gathering different age variables into a single variable "AGE" and corresponding age type "TIMEPOINT"
pats_age <- patients_recoded_trim %>% 
  gather(AGE_ADM:AGE_DOD, key = "TIMEPOINT", value = "AGE") %>%
  mutate(TIMEPOINT = recode(TIMEPOINT, "AGE_ADM" = "ADMISSION", "AGE_DOD" = "DEATH"))
head(pats_age)

# creating a histogram for AGE at admission and at death
ggplot(pats_age, aes(x = AGE), bins = 50) +
  geom_histogram(aes(fill = TIMEPOINT), color = "white") + 
  scale_fill_manual(values = c(Grey, "black")) +
  theme_bw() +
  labs(y = "Record Count", x = "Patient Age")

# AGE_ADM descriptive information
table(patients_recoded_trim$AGE_ADM > 89) # only 36 records with patient age > 89

summary(patients_recoded_trim$AGE_ADM) # mean 64.02
sd(patients_recoded_trim$AGE_ADM) # std dev 16.9921

# AGE_DOD descriptive information
age_dod <- patients_recoded_trim %>% filter(!(is.na(AGE_DOD)))
summary(age_dod$AGE_DOD) # mean 70.13
sd(age_dod$AGE_DOD) # std dev 14.86695

# AGE_ADM as predictor
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, y = AGE_ADM)) +
    geom_boxplot() +
  labs(y = "Age at Admission (years)", x = "Mortality")

# AGE_ADM as a predictor
summary((glm(MORTALITY ~ AGE_ADM, data = patients_recoded_trim, family = binomial()))) # p-value <2e-16
```
- AGE_ADM was positively associated with MORTALITY with p-value <2e-16

```{r Gender, collapse=TRUE, message=FALSE}
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, fill = GENDER)) +
    geom_bar(position = "fill", color = "white") + 
  scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# GENDER as predictor
summary((glm(MORTALITY ~ GENDER, data = patients_recoded_trim, family = binomial()))) # 0.00226
```
- There is a relationship between GENDER and MORTALITY with p-value 0.00226

```{r Ethnicity, collapse=TRUE, collapse=TRUE, message=FALSE}
summary(patients_recoded_trim$ETHNICITY_BROAD)

ggplot(patients_recoded_trim, aes(x = MORTALITY, fill = ETHNICITY_BROAD)) + 
  geom_bar(aes(fill = ETHNICITY_BROAD), position = "dodge")
  #scale_fill_manual(values = c(Grey, Yellow, PennRed))

ggplot(data = patients_recoded_trim, aes(x = MORTALITY, fill = ETHNICITY_BROAD)) +
    geom_bar(position = "fill", color = "white") + 
  #scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# ETHNICITY_BROAD as predictor
summary((glm(MORTALITY ~ ETHNICITY_BROAD, data = patients_recoded_trim, family = binomial()))) # 0.00226
```
- There is no significant association between any particular `ETHNICITY_BROAD` and `MORTALITY`

```{r Insurance, collapse=TRUE, message=FALSE}
summary(patients_recoded_trim$INSURANCE)

# standard bar plot
ggplot(patients_recoded_trim, aes(x = MORTALITY, fill = INSURANCE)) + 
  geom_bar(aes(fill = INSURANCE), position = "dodge") +
  labs(y = "Count", x = "Mortality")
  #scale_fill_manual(values = c(Grey, Yellow, PennRed)) +

# stacked bar plot
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, fill = INSURANCE)) +
    geom_bar(position = "fill", color = "white") + 
  #scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# INSURANCE_TYPE as predictor
summary((glm(MORTALITY ~ INSURANCE, data = patients_recoded_trim, family = binomial())))
```
- There is a signficant association between all `INSURANCE` and `MORTALITY` (i.e. having any insurance)

```{r LOS_ICU, collapse=TRUE, message=FALSE, warning=FALSE}
# Length of stay (LOS) in the ICU as predictor
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, y = LOS_ICU)) +
    geom_boxplot() +
  labs(y = "Length of Stay in ICU (days)", x = "Mortality")

# LOS_ICU as predictor
summary((glm(MORTALITY ~ LOS_ICU, data = patients_recoded_trim, family = binomial()))) # p-value <2e-16
```
- `LOS_ICU` was positively associated with `MORTALITY` with p-value <2e-16

```{r ADMISSION_TYPE, collapse=TRUE, message=FALSE}
# standard bar plot
ggplot(patients_recoded_trim, aes(x = MORTALITY, fill = ADMISSION_TYPE)) + 
  geom_bar(aes(fill = ADMISSION_TYPE), position = "dodge") +
  labs(y = "Count", x = "Mortality")
  #scale_fill_manual(values = c(Grey, Yellow, PennRed)) +

# Type of ICU admission as predictor
summary((glm(MORTALITY ~ ADMISSION_TYPE, data = patients_recoded_trim, family = binomial()))) # p-value <2e-16

```
- `ADMISSION_TYPE`s EMERGENCY and URGENT were both positively associated with `MORTALITY` with p-value <2e-16

This section summarizes the record count for each variable of interest and calculates the proportion (%).
```{r Summary, message=FALSE}
summary(patients_recoded_trim)

patients_recoded_trim %>% group_by(MORTALITY, GENDER) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(MORTALITY, INSURANCE) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(MORTALITY, ADMISSION_TYPE) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(MORTALITY, ETHNICITY_BROAD) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(MORTALITY, DELIVERY, GENDER) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(MORTALITY, CSECTION, GENDER) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(MORTALITY) %>% summarize(COUNT = n(), PROP = round(n()/54362*100,2))
```
- From the summarized data it is clear that there are no records of patients that died within the hospital and also delivered a child during their care. Subsequently there are also no records of patients that died within the hospital and delivered a child via a C-section procedure. 
- Therefore, any logistic model assessing `DELIVERY` or `CSECTION` as a predictor of `MORTALITY` will be inadequate, as evidenced below.

```{r Delivery, collapse=TRUE, message=FALSE}
# narrowing down to female patients
patients_recoded_trim_f <- patients_recoded_trim %>% filter(GENDER == "F")

ggplot(data = patients_recoded_trim_f, aes(x = MORTALITY, fill = DELIVERY)) +
    geom_bar(position = "dodge", color = "white") + 
  scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# Delivery as predictor
summary((glm(MORTALITY ~ DELIVERY, data = patients_recoded_trim_f, family = binomial()))) # 0.00226
```
- There is a **non-significant** positive association between `NO DELIVERY` and `MORTALITY`

```{r C-section, collapse=TRUE, message=FALSE}
ggplot(data = patients_recoded_trim_f, aes(x = MORTALITY, fill = CSECTION)) +
    geom_bar(position = "dodge", color = "white") + 
  scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# C-section as predictor
summary((glm(MORTALITY ~ CSECTION, data = patients_recoded_trim_f, family = binomial()))) # 0.00226
```
- There is a **non-significant** positive association between `NO CSECTION` and `MORTALITY`

#### Logistic regression

This section uses a logistic regression model that includes all variables of interest as predictors of `MORTALITY`.
```{r GLM Model, collapse=TRUE, message=FALSE}
# narrowing to only binary variables of interest
patients_recoded_trim_glm <- patients_recoded_trim %>% select(-(AGE_ADM:LOS_ICU),-(DELIVERY:CSECTION))

# logistic regression model
patients_glm <- glm(MORTALITY ~ ., data = patients_recoded_trim_glm, family = binomial(logit))
summary(patients_glm)

# There were signficant positive associations between all insurance types and MORTALITY

glm_pred <- predict(patients_glm, patients_recoded_trim_glm, type = "response")
head(glm_pred)

table(glm_pred > 0.5) # 0 admissions had probability > 0.5
```
- There were no admissions with a probability greater than 0.5 using this model, meaning the prediction was no better than chance.

#### Model Evaluation

This section performs a 10-fold cross-validation of the model
```{r Cross Validation, message=FALSE}
#K-Fold Cross Validation
N = nrow(patients_recoded_trim_glm)
K = 10
set.seed(1234)
s = sample(1:K, size = N, replace = T)
pred.outputs.glm <- vector(mode = "numeric", length = N)
obs.outputs <- vector(mode = "numeric", length = N)
offset <- 0
for(i in 1:K){
    train <- filter(patients_recoded_trim_glm, s != i)
    test <- filter(patients_recoded_trim_glm, s == i)
  obs.outputs[1:length(s[s == i]) + offset] <- test$MORTALITY
    
  #GLM train/test
    glm <- glm(MORTALITY ~ ., data = train, family = binomial(logit))
  glm.pred.curr <- predict(glm, test, type = "response")
  pred.outputs.glm[1:length(s[s == i]) + offset] <- glm.pred.curr

    offset <- offset + length(s[s == i])
}
head(pred.outputs.glm)

table(pred.outputs.glm > .5) # No admissions had probability > 0.5 
```

```{r ROC Curve, message=FALSE, message=FALSE}
library(pROC)
roc(patients_recoded_trim_glm$MORTALITY, glm_pred, ci = TRUE) # AOC: 0.6399

plot.roc(patients_recoded_trim_glm$MORTALITY, glm_pred, ci = TRUE)
plot.roc(obs.outputs, pred.outputs.glm, col = "red", add = TRUE)
```
- An area under the curve (AUC) of 0.6399 showed the logistic model was slightly better at fitting the admissions data than chance (~0.5).

#### Conclusions and Limitations
- The logistic model as utilized in this project was not an ideal for the admissions data in the MIMIC-III database. Approaching the analysis of mortality as a binary outcome might benefit from exploration using a random forest classifier, potentially in addition to regularization of the data to identify the most important variables in the dataset.
- The patient population is mostly of an age beyond the reproductive lifespan, which means fewer pregnancies, particularly those unassisted by Artificial Reproductive Technology, so the MIMIC-III databse is not adequate for studying pregnancy-related outcomes.
- Because all dates in the database have been shifted to protect patient confidentiality, C-section rates could not be compared to national rates reported by the Centers for Disease Control and Prevention (CDC) because years could not be matched. Reference: https://mimic.physionet.org/mimicdata/time/
  - Date shifting removed:
      - Year: The year was randomly distributed between 2100 and 2200
      - Day of the month: The absolute day of the month is not preserved (though the day of the week is)
      - Inter-patient information: Meaning two patients in the ICU on the same date were not actually in the ICU at the same time
- Not all `ICD9_CODES` in the MIMIC-III database matched to an exisiting ICD code, which reduced the number of records available for analysis.


### References
- Johnson, A., Pollard, T., Mark, R. (2016) MIMIC-III, a freely accessible critical care database. Scientific Data (2016). DOI: 10.1038/sdata.2016.35. Available from: http://www.nature.com/articles/sdata201635
