---
title: "The Childhood Interstitial Lung Disease (chILD) Registry: A Key Instrument in the Study of Rare Childhood Lung Disease"
author: "Katie Tsukahara"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
bibliography: BMIN503_Final_bib.bib
csl: american-medical-association.csl
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

### Overview
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.


>OVERVIEW:<br>Childhood interstitial lung disease (chILD) is a heterogeneous group of rare childhood lung diseases affecting the interstitial region of the lung, leading to abnormal gas exchange and a range of morbidity and mortality. Research and development of therapies is challenging due to the low prevalence and heterogeneity of disease, thus the CHILD Research Network and Registry was established to collect data from multiple centers across the United States. Our goal is to compare patient demographics across subsets of chILD subjects by diagnosis, and to examine the diagnostic and clinical factors that contribute to diagnosis.

Link to my Final Project Repo is [here](https://github.com/ktsukah/BMIN503_Final_Project).

### Introduction 
Childhood interstitial lung disease (chILD) encompasses a variety of lung diseases that manifest via changes to the pulmonary interstitium. The pulmonary interstitium is, simply put, the lung tissue between the airspace and the pulmonary blood vessels, and excludes other lung structures such as the airways. These diseases are grouped together owing to similarities in the area of the lung affected and in clinical presentation. While both children and adults can develop interstitial lung disease (ILD), many causes of chILD are seen only in the pediatric age group. These diseases can be genetic in origin, or associated with abnormal lung development, immunologic dysfunction, rheumatologic disease, or infectious history. Some cases do not have a clear etiology. Certain diagnoses confer a poor prognosis of early lung transplant or mortality. Other diagnoses confer a more positive prognosis, including clinical improvement and resolution of disease over early childhood.
<br><br>To study this group of diseases, the Children’s Interstitial and Diffuse Lung Disease Research Network (ChILDRN) established a national registry in 2016, with participation by 13 sites across the United States. The original registry contains 436 subjects, enrolled from 2016 to 2020, with data on demographics, medical history, diagnostics, management and outcomes. With this project, we aim to examine differences in patient demographics, including age of onset and severity of presentation, among the more common chILD diagnoses. We also aim to examine the diagnostic and clinical factors that contribute to diagnosis, if possible examining the role that different diagnostic tests play in diagnosis and whether invasive testing can be avoided in these patients that are at higher risk for procedures due to their underlying pulmonary disease. The database and this subsequent project use the skills and expertise of clinicians across many fields (Pulmonology, Radiology, Pathology, Rheumatology, and others), clinical researchers, basic science researchers, and biomedical informaticians to collect, manage, and interpret data on this complex group of patients.



### Methods
The Children’s Interstitial and Diffuse Lung Disease Research Network (ChILDRN) established a national registry for chILD under an IRB for a longitudinal observational study with the participation of 13 centers across the United States [@YoungOA3786]. This has since expanded to 20 centers, with over 400 enrolled subjects. The IRB allows for both retrospective enrollment as well as prospective data collect, including clinical characteristics as well as diagnostic testing, longitudinal results, and outcomes such as supplemental oxygen use, mechanical ventilation or death. The data is housed by the REDCap (Research Electronic Data Capture) data platform. [@HARRIS2019103208; @HARRIS2009377] For the purposes of this project, data of interest was filtered using the REDCap Report feature, then de-identified and date-shifted data was exported for analysis, in accordance with the IRB. A CSV file contained the raw data, and the R file contained R code to read in the CSV, and appropriately label columns, data type, and factor levels using the Hmisc package.

```{r, include=TRUE, warning = FALSE, message = FALSE}
# loading RedCAP data
setwd("C:/Users/tsukaharak/Documents/R files/BMIN503_Final_Project")
source("CHOPChildhoodILDNati-110620BMIN_R_2020-11-27_1739.R", local = knitr::knit_global())

# a cursory check for duplicate entries - all entries have a unique chILD Registry ID
length(data$child_registry_id)
length(unique(data$child_registry_id))

# copying the data frame to create a working version, so that if going back to the original data is required, 
#it will not require reloading in all of the data
child <- data

```

After loading in the data, basic enrollment demographics were examined. This included center where the child was enrolled, age at enrollment, age at diagnosis, race and ethnicity.


```{r, echo = FALSE, warning = FALSE, message = FALSE}
# loading libraries for data cleaning and graphics
library(tidyr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(lubridate)

```

In order to create a visualization of enrollment by center, a new data frame was created with with city and chILD Registry ID prefix corresponding to the enrollment center. Latitude and longitude were retrieved from the us.cities data frame in the maps package. Due to two centers being located in New York city boroughs, these coordinates were added manually as they were not included in the us.cities data frame. These were then plotted on a map of the continental United States. These geographic locations are representative only of the location of enrollment. It is not uncommon for children with rare disease to travel to a specialized center for, thus, it is difficult to infer any geographic distrubtion of the patient's location of residence from these numbers.

```{r, warning = FALSE, message = FALSE, eval = TRUE}
library(maps)
library(mapproj)

DAG.df <- data.frame(
  name = c("Boston MA", "Atlanta GA", "Philadelphia PA", "Cleveland OH", "Queens NY", "Aurora CO", "Indianapolis IN", "Kansas City MO", "Chicago IL", "Ann Arbor MI", "Columbia MO", "Bronx NY", "Portland OR", "Pittsburgh PA", "San Diego CA", "Seattle WA", "Palo Alto CA", "Chapel Hill NC", "Nashville TN", "Saint Louis MO"),
  prefix = c(704,720,712,709,702,703,718,711,717,716,715,708,713,705,706,710,707,719,701,714),
  stringsAsFactors = FALSE
)

DAG.df <- DAG.df %>% 
  arrange(prefix) %>%
  left_join(us.cities, by = "name") %>%
  mutate(prefix = as.character(prefix))

# latitude and longitude for NY boroughs
# Queens 40.728, -73.795
# Bronx 40.845, -73.865
DAG.df[which(grepl("Queens",DAG.df$name)), ]$lat <- 40.7282
DAG.df[which(grepl("Queens",DAG.df$name)), ]$long <- -73.7949
DAG.df[which(grepl("Bronx",DAG.df$name)), ]$lat <- 40.8448
DAG.df[which(grepl("Bronx",DAG.df$name)), ]$long <- -73.8648

DAG.child <- child %>%
  select(child_registry_id) %>%
  mutate(prefix = substr(child_registry_id, 0, 3)) %>%
  count(prefix) %>%
  mutate(prefix = as.character(prefix)) %>%
  left_join(DAG.df, by = "prefix")

USA <- map_data("world") %>% filter(region=="USA")

# Create breaks for the point sizing scale
mybreaks <- c(20,40,60,80,100)

# Build the map
ggplot() +
    geom_polygon(data = USA, aes(x=long, y = lat, group = group), fill="grey", alpha=0.4) +
    geom_point(data = DAG.child, aes(x=long, y=lat, size=n), color = "#2b8cbe", shape=20, stroke=FALSE) +
    scale_size_continuous(name="Number of Enrolled \nSubjects", trans="identity", range=c(0,12), breaks = mybreaks) +
    theme_void() +
    coord_map() + ylim(25,50) + xlim(-130, -60) +
    guides(colour = guide_legend()) +
    ggtitle("Subject Enrollment by Center") +
    theme(
      legend.position = c(0.85, 0.15),
      text = element_text(color = "#22211d"),
      plot.title = element_text(size= 16, hjust=0.1, color = "#4e4d47", margin = margin(b = -0.1, t = 0.4, l = 2, unit = "cm")),
    )

```

This study allowed for retrospective enrollment as well as prospective data collection. Using summary statistics and histograms, the age at enrollment in the Registry and age at diagnosis were examined. When the Registry was first established, it was expected that many subjects would be enrolled retrospectively. 

```{r, warning = FALSE, message = FALSE}
# Age at enrollment and diagnosis
summary(child$age_studyentry)
ggplot(data = child, aes(x = age_studyentry)) + 
  geom_histogram(color = "black", fill = "white", binwidth=3) +
  labs(x = "Age (in months)", title = "Age at Study Entry in chILD Registry")
summary(child$age_diagnosis)
ggplot(data = child, aes(x = age_diagnosis)) + 
  geom_histogram(color = "black", fill = "white", binwidth=3) +
  labs(x = "Age (in months)", title = "Age at Diagnosis in chILD Registry")
```

We were also interested to see if subjects are being enrolled closer to their time of diagnosis now that the Registry is more established. This was plotted as a scatter plot below, and showed that there continues to be a significant amount of retrospective enrollment, with many subjects being enrolled up to several years after their diagnosis. The reasons for this are unclear, and may reflect a growing awareness of the chILD Registry, a move towards providing care for chILD at specialized centers, or the need for further resources devoted to recruitment.

```{r, warning = FALSE, message = FALSE}
time.to.enroll <- child %>% 
  select(age_studyentry, age_diagnosis, cons_date) %>%
  mutate(age.diff = age_studyentry - age_diagnosis)
time.to.enroll <- time.to.enroll %>%
  mutate(cons_date = as.Date(time.to.enroll$cons_date))
ggplot(time.to.enroll, aes(x=cons_date, y=age.diff)) +
  geom_point() +
  ggtitle("Time Intervals between Diagnosis and Enrollment in chILD Registry") +
  ylab("Time from Diagnosis to Enrollment (in Months)") +
  xlab("Approximate Time of Enrollment (Date-Shifted)") +
  theme(axis.text.x=element_text(angle=60, hjust=1)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")

```


Overall demographic and clinical characteristics of the registry are shown below. Further analysis was then performed to compare clinical characteristics across diagnoses. 

```{r, warning = FALSE, message = FALSE}
# attempt at a Table 1
# not able to download ethnicity data
library(gtsummary)
library(gt)

table.data <- child %>%
  dplyr::select(age_studyentry, age_diagnosis, patient_gender.factor,  relative_in_registry.factor, fibrosis.factor, airway.factor, home_o2.factor, mech_vent___1.factor, mech_vent___2.factor, transplant_yn.factor) 
levels(table.data$mech_vent___1.factor)=c("No","Yes")
levels(table.data$mech_vent___2.factor)=c("No","Yes")

table.data <- table.data %>%
  rename("Patient Gender" = patient_gender.factor) %>%
  rename("Relative in Registry" = relative_in_registry.factor) %>%
  rename("Pulmonary Fibrosis" = fibrosis.factor) %>%
  rename("Airway Disease" = airway.factor) %>%
  rename("Past or Current Home Supplemental Oxygen" = home_o2.factor) %>%
  rename("Past or Current Noninvasive Ventilation (>3 weeks)" = mech_vent___1.factor) %>%
  rename("Past or Current Invasive Ventilation (>3 weeks)" = mech_vent___2.factor) %>%
  rename("Lung Transplant" = transplant_yn.factor) %>%
  mutate(Death = ifelse(!is.na(child$death_age), "Yes", "No"))

table.one <- tbl_summary(table.data,  missing = "no") %>%
as_gt() %>%
gt::tab_header("Table 1. Summary of Clinical Characteristics of the Study Cohort") %>%
gt::opt_align_table_header(align = "left") %>%
gt::tab_options(table.font.size = 'small')

table.one

```



### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

### References
<div id="refs"></div>
