---
title: "BMIN503/EPID600 Final Project"
author: "Silvia Canelon"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  

### Overview

The project makes use of the freely-available MIMIC-III Clinical Database to investigate the characteristics of pregnant patients admitted to the Beth Israel Deaconess Medical Center, and their inpatient care. This large dataset of over forty thousand patients admitted to critical care units will be used to identify predictors significantly associated with mortality and/or specific cause of death.

Three faculty/staff being consulted for the project:
- Mary Regina Boland; Assistant Professor of Informatics; Department of Biostatistics, Epidemiology, and Informatics
- Ray Bai; Postdoctoral Researcher; Department of Biostatistics, Epidemiology, and Informatics
- Lena Davidson; Data Analyst; Department of Biostatistics, Epidemiology, and Informatics

### Introduction 

The secondary use of Electronic Health Records (EHR) facilitates retrospective research on data collected during the course of patients' clinical care. The data available through the MIMIC-III Clinical Database provides detailed health-related information collected from a diverse population of patients admitted to intensive care units of a large me dical center, some of whom may have been pregnant at the time of admission. Given the nature of intensive care units, any pregnant patients admitted were likely considered high-risk pregnancies before or at the time of admission. In addition, a subset of these pregnant patients will have deceased or suffered other adverse outcomes during their care, information which can be gleaned through analysis of in-hospital mortality and billing-related information recorded in the EHR. Understanding the characteristics of the pregnant patient population, as well as their care, will shed light on patterns related to adverse outcomes experienced during critical care and hopefully provide preliminary data pertinent to the clinical decison-making process.

The analysis and subsequent interpretation of EHR data is interdisciplinary because it combines domains of the biomedical informatics, epidemiology, biostatistics, and clinical fields. Uncovering potential correlations between patient and patient care information and clinical outcomes involves the use of informatics and statistical analysis skills, whereas epidemiology and clinical domains provide the necessary context for defining/refining the analysis approach and interpreting the results. Consultation with all three faculty/staff has confirmed that adverse outcomes among the patient population admitted to critical care are important to understand from a public health perspective, and likely to not be "rare" events given the intensive care unit context, which allows for standard statistical approaches.


### Methods

The Mimic-III Clinic Database contains de-identified health data records for over 40,000 patients admitted to the Beth Israel Deaconess Medical Center of Boston, Massachusetts from 2001 and 2012. The database includes information on demographics, vital sign measurements, laboratory test results, procedures, medications, caregiver notes, imaging reports, billing-related information, and mortality.
Demographic, procedure, mortality, and billing-related information such as International Classification of Disease (ICD) codes, will be used to explore patient characteristics and care while being treated at Beth Isreal Deaconess Medical Center, and mortality studied as an outcome.

#### Connection to server
```{r}
rm(list=ls()) # clears the environment

# load packages
pkgs = c("RMySQL", "DBI", "tidyverse", "stringr", "dplyr", "formatR", "lubridate", "tictoc", "ggThemeAssist", "viridis", "GGally") # define packages
lapply(pkgs, library, character.only = TRUE) # load packages

drv = dbDriver("MySQL")

#change these below
username= #username for mysql
mysql_password=  #password for mysql

###
mimic_iii = dbConnect(drv, dbname="clinical_mimic_III", 
                           host="127.0.0.1",
                           port=3307,
                           user=keyring::key_list("Curie")[1,2], #username for mysql
                           password=keyring::key_get("Curie"))  #password for mysql

#on.exit(dbDisconnect(mimic_iii)) # had to comment out in order to knit document
```

#### Importing data
```{r}
tic("data import") # takes about 20 secs

rs = dbSendQuery(mimic_iii, "select distinct * from ADMISSIONS;")
data_admissions = fetch(rs, n=-1) #46520
length(unique(data_admissions$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from PATIENTS;")
data_patients = fetch(rs, n=-1) #46520
length(unique(data_patients$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from SERVICES;")
data_services = fetch(rs, n=-1) #46496
length(unique(data_services$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from TRANSFERS;")
data_transfers = fetch(rs, n=-1) #46520
length(unique(data_transfers$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from ICUSTAYS;")
data_icustays = fetch(rs, n=-1) #46476
length(unique(data_icustays$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from DIAGNOSES_ICD;")
data_icd_diag = fetch(rs, n=-1) #46520
length(unique(data_icd_diag$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from PROCEDURES_ICD;")
data_icd_proc = fetch(rs, n=-1) #42214
length(unique(data_icd_proc$SUBJECT_ID))

rs = dbSendQuery(mimic_iii, "select distinct * from D_ICD_DIAGNOSES;")
dd_icd_diag = fetch(rs, n=-1) 
dim(unique(dd_icd_diag)) #14567

rs = dbSendQuery(mimic_iii, "select distinct * from D_ICD_PROCEDURES;")
dd_icd_proc = fetch(rs, n=-1) 
dim(unique(dd_icd_proc)) #3882

toc()
```

#### Data tyding

##### Trimming/formatting data
```{r}
# trimming and formatting admissions data
admissions <- data_admissions %>% 
  select(-(ADMISSION_LOCATION:DISCHARGE_LOCATION),-(EDREGTIME:DIAGNOSIS),
         -(LANGUAGE:MARITAL_STATUS),-HAS_CHARTEVENTS_DATA) %>%
  mutate(ADMITTIME = as.Date(ADMITTIME),
         DISCHTIME = as.Date(DISCHTIME))

# formatting dates of death in expired patient data
patients_dod <- data_patients %>% select(-DOD,-DOD_SSN) %>% filter(EXPIRE_FLAG == 1) %>%
  mutate(DOD_HOSP = as.Date(DOD_HOSP))
# joining expired patient data with larger patients data frame
patients <- left_join(data_patients %>% select(-(DOD:DOD_SSN)), patients_dod) %>%
  mutate(DOB = as.Date(DOB))

# trimmming and formatting services data
services <- data_services %>% select(-(TRANSFERTIME:PREV_SERVICE))

# trimmming and formatting ICU stay data
icustays <- data_icustays %>% select(-(FIRST_WARDID:LAST_WARDID)) %>%
  mutate(INTIME = as.Date(INTIME),
         OUTTIME = as.Date(OUTTIME),
         LOS = as.numeric(LOS))
```

##### Joining data
```{r}
# joining diagnosis and procedure data to ICD codes
icd_diag <- left_join(data_icd_diag, dd_icd_diag %>% select(-ROW_ID), by = "ICD9_CODE")
icd_proc <- left_join(data_icd_proc, dd_icd_proc %>% select(-ROW_ID), by = "ICD9_CODE")

# counting diagnosis records with and without proper ICD9_CODEs from the data dictionary
dim(icd_diag) #651047
length(unique(icd_diag$SUBJECT_ID)) #46520
diag_w_code <- icd_diag %>% filter(!(is.na(SHORT_TITLE)))
dim(diag_w_code) #634709
diag_wo_code <- icd_diag %>% filter(is.na(SHORT_TITLE))
dim(diag_wo_code) #16338

# counting procedure records with and without proper ICD9_CODEs from the data dictionary
dim(icd_proc) #240095
length(unique(icd_proc$SUBJECT_ID)) #42214
proc_w_code <- icd_proc %>% filter(!(is.na(SHORT_TITLE)))
dim(proc_w_code) #237948
proc_wo_code <- icd_proc %>% filter(is.na(SHORT_TITLE))
dim(proc_wo_code) #2147
```

```{r}
# joining admission data to adult patient data
patients_adm <- left_join(admissions, patients %>% select(-ROW_ID), by = "SUBJECT_ID") %>%
  mutate(AGE_ADM = round(as.duration(DOB %--% ADMITTIME)/dyears(1),2),
         AGE_DIS = round(as.duration(DOB %--% DISCHTIME)/dyears(1),2),
         AGE_DOD = round(as.duration(DOB %--% DOD_HOSP)/dyears(1),2)) %>%
  filter(AGE_ADM >= 18)

# joining ICU data to admission+patient data
patients_adm_icu <- left_join(patients_adm, icustays %>% select(SUBJECT_ID, HADM_ID, LOS),
                              by = c("SUBJECT_ID","HADM_ID")) %>% rename(LOS_ICU = LOS)

# joining ICD data to admission+patient+ICU data
patients_adm_icu_diag <- left_join(patients_adm_icu, icd_diag %>% select(-ROW_ID), 
                               by = c("SUBJECT_ID","HADM_ID"))
patients_adm_icu_proc <- left_join(patients_adm_icu, icd_proc %>% select(-ROW_ID), 
                               by = c("SUBJECT_ID","HADM_ID"))
```


##### Recoding data
```{r}
# simplifying ethnicity category
cat_ethnicity <- table(patients_adm_icu$ETHNICITY) %>% data.frame()

patients_recoded <- patients_adm_icu %>% 
  mutate(ETHNICITY_BROAD = as.factor(ETHNICITY)) %>%
  mutate(ETHNICITY_BROAD = recode(ETHNICITY_BROAD,
                                  "AMERICAN INDIAN/ALASKA NATIVE" = "AIAN",
                                  "AMERICAN INDIAN/ALASKA NATIVE FEDERALLY RECOGNIZED TRIBE" = "AIAN",
                                  "ASIAN - ASIAN INDIAN" = "ASIAN",
                                  "ASIAN - CAMBODIAN" = "ASIAN",
                                  "ASIAN - CHINESE" = "ASIAN",
                                  "ASIAN - FILIPINO" = "ASIAN",
                                  "ASIAN - JAPANESE" = "ASIAN",
                                  "ASIAN - KOREAN" = "ASIAN",
                                  "ASIAN - OTHER" = "ASIAN",
                                  "ASIAN - THAI" = "ASIAN",
                                  "ASIAN - VIETNAMESE" = "ASIAN",
                                  "BLACK/AFRICAN" = "BLACK",
                                  "BLACK/AFRICAN AMERICAN" = "BLACK",
                                  "BLACK/CAPE VERDEAN" = "BLACK",
                                  "BLACK/HAITIAN" = "BLACK",
                                  "HISPANIC OR LATINO" = "HISPANIC",
                                  "HISPANIC/LATINO - CENTRAL AMERICAN (OTHER)" = "HISPANIC",
                                  "HISPANIC/LATINO - COLOMBIAN" = "HISPANIC",
                                  "HISPANIC/LATINO - CUBAN" = "HISPANIC",
                                  "HISPANIC/LATINO - DOMINICAN" = "HISPANIC",
                                  "HISPANIC/LATINO - GUATEMALAN" = "HISPANIC",
                                  "HISPANIC/LATINO - HONDURAN" = "HISPANIC",
                                  "HISPANIC/LATINO - MEXICAN" = "HISPANIC",
                                  "HISPANIC/LATINO - PUERTO RICAN" = "HISPANIC",
                                  "HISPANIC/LATINO - SALVADORAN" = "HISPANIC",
                                  "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER" = "NHPI",
                                  "WHITE - BRAZILIAN" = "WHITE",
                                  "WHITE - EASTERN EUROPEAN" = "WHITE",
                                  "WHITE - OTHER EUROPEAN" = "WHITE",
                                  "WHITE - RUSSIAN" = "WHITE"))

cat_ethnicity_broad <- table(patients_recoded$ETHNICITY_BROAD) %>% data.frame()

# formatting/recoding insurance, gender, hospital death, and age variables
patients_recoded <- patients_recoded %>% 
  mutate(ADMISSION_TYPE = as.factor(ADMISSION_TYPE),
         GENDER = factor(GENDER),
         MORTALITY = factor(HOSPITAL_EXPIRE_FLAG, levels = c(0,1), labels = c("LIVED","DIED")),
         AGE_ADM = ifelse(AGE_ADM > 300, 89, AGE_ADM),
         AGE_DIS = ifelse(AGE_DIS > 300, 89, AGE_DIS),
         AGE_DOD = ifelse(AGE_DOD > 300, 89, AGE_DOD),
         INSURANCE = as.factor(INSURANCE)) %>%
  mutate(INSURANCE = recode(INSURANCE, 
                            "Government" = "GOV", "Medicaid" = "MEDICAID","Medicare" = "MEDICARE",
                            "Private" = "PRIVATE", "Self Pay" = "SELF PAY"))

# trimming data
patients_recoded <- patients_recoded %>% select(-ETHNICITY, -HOSPITAL_EXPIRE_FLAG, -EXPIRE_FLAG)
```

#### Delivery codes

```{r}
# Importing all delivery DIAGNOSIS ICD9 codes
codes_delivery_diag = read.csv("delivery_diag_codes_SPC.csv", header = F) %>% 
  mutate(V1 = as.character(V1))
head(codes_delivery_diag)
codes_delivery_diag_icd9 <- codes_delivery_diag %>% filter(V2 == "ICD9") %>% 
  mutate(CODE = str_remove(V1, "\\.")) %>% select(CODE)
head(codes_delivery_diag_icd9)

# Importing all delivery PROCEDURE ICD9 codes
codes_delivery_proc = read.csv("delivery_proc_codes_SPC.csv", header = F)
head(codes_delivery_proc)
codes_delivery_proc_icd9 <- codes_delivery_proc %>% filter(V2 == "ICD9") %>% 
  mutate(CODE = str_remove(V1, "\\.")) %>% select(CODE)
head(codes_delivery_proc_icd9)
```

```{r}
# Records with delivery diagnosis codes
records_delivery_diag = icd_diag %>% filter(ICD9_CODE %in% codes_delivery_diag_icd9$CODE) %>% 
  mutate(CODE_TYPE = "DIAGNOSIS")
head(records_delivery_diag)

# Records with delivery procedure codes
records_delivery_proc = icd_proc %>% filter(ICD9_CODE %in% codes_delivery_proc_icd9$CODE) %>%
  mutate(CODE_TYPE = "PROCEDURE")
head(records_delivery_proc)

# Joining diagnosis delivery and procedure codes
records_delivery <- rbind(records_delivery_diag, records_delivery_proc) %>% unique()
length(unique(records_delivery$ROW_ID)) #203
length(unique(records_delivery$SUBJECT_ID)) #48
length(unique(records_delivery$HADM_ID)) #48
```
- There are 203 delivery diagnosis or procedure records from 48 patients

#### C-section codes
```{r}
# Importing all c-section PROCEDURE codes
codes_csection_proc = read.csv("csection_proc_codes_SPC.csv", header = F)
head(codes_csection_proc)
codes_csection_proc_icd9 <- codes_csection_proc %>% filter(V2 == "ICD9") %>%
  mutate(CODE = str_remove(V1, "\\.")) %>% select(CODE)
head(codes_csection_proc_icd9)
```

```{r}
# Records with csection procedure codes
records_csection_proc = icd_proc %>% filter(ICD9_CODE %in% codes_csection_proc_icd9$CODE) %>%
  mutate(CODE_TYPE = "PROCEDURE")
head(records_csection_proc)

records_csection <- records_csection_proc

length(unique(records_csection$ROW_ID)) #38
length(unique(records_csection$SUBJECT_ID)) #37
length(unique(records_csection$HADM_ID)) #37
```
- There are 38 csection procedure codes from 37 patients

#### Joining delivery and csection
```{r}
# adding in variables for delivery and csection
patients_recoded <- patients_recoded %>%
  mutate(DELIVERY = ifelse(HADM_ID %in% records_delivery$HADM_ID, "DELIVERY", "NO DELIVERY"),
         CSECTION = ifelse(HADM_ID %in% records_csection$HADM_ID, "CSECTION", "NO CSECTION")) %>%
  mutate(DELIVERY = as.factor(DELIVERY),
         CSECTION = as.factor(CSECTION))
```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

#### Descriptive information
```{r}
patients_recoded_trim <- patients_recoded %>% 
  select(ADMISSION_TYPE,INSURANCE,GENDER,AGE_ADM,AGE_DOD,LOS_ICU,ETHNICITY_BROAD,DELIVERY,CSECTION, MORTALITY) %>% unique()
length(unique(patients_recoded_trim)) #38888

ggpairs(patients_recoded_trim, mapping = aes(col = MORTALITY), columns = c("GENDER","ETHNICITY_BROAD", "AGE_ADM","AGE_DOD"))
```

#### Visualization
```{r}
# Penn and DBEI/CCEB colors
PennBlue = "#1e376c" #DBEI/CCEB
PennRed = "#a90533" #DBEI/CCEB
DarkBlue = "#2e5d97" #DBEI/CCEB
LightBlue = "#2e9ad5" #DBEI/CCEB
Purple = "#7e6b94" #DBEI
Orange = "#d47d3a" #Excel orange
Yellow = "#efbf2c" #Excel yellow 
DeptYellow = "ffd400" #DBEI/CCEB yellow
Beige = "#e8eade" #DBEI/CCEB
OrangeRed = "#fef3742" #DBEI
LightRed = "#d99792" #DBEI
Green = "#83944e" #CCEB
Grey = "#999999"
```

```{r Plot:Age}
pats_age <- patients_recoded_trim %>% 
  gather(AGE_ADM:AGE_DOD, key = "TIMEPOINT", value = "AGE") %>%
  mutate(TIMEPOINT = recode(TIMEPOINT, "AGE_ADM" = "ADMISSION", "AGE_DOD" = "DEATH"))

ggplot(pats_age, aes(x = AGE), bins = 50) + #bins = 25 looks good
  geom_histogram(aes(fill = TIMEPOINT), color = "white") + 
  scale_fill_manual(values = c(Grey, "black")) +
  theme_bw() +
  labs(y = "Record Count", x = "Patient Age")

# Age summaries
table(patients_recoded_trim$AGE_ADM > 89) # only 36 records with patient age > 89

summary(patients_recoded_trim$AGE_ADM) # mean 64.02
sd(patients_recoded_trim$AGE_ADM) # std dev 16.9921

# AGE_DOD
age_dod <- patients_recoded_trim %>% filter(!(is.na(AGE_DOD)))
summary(age_dod$AGE_DOD) # mean 70.13
sd(age_dod$AGE_DOD) # std dev 14.86695

# AGE at admission as predictor
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, y = AGE_ADM)) +
    geom_boxplot() +
  labs(y = "Age at Admission (years)", x = "Mortality")

summary((glm(MORTALITY ~ AGE_ADM, data = patients_recoded_trim, family = binomial()))) # p-value <2e-16

# AGE_ADM was positively associated with MORTALITY with p-value <2e-16
```

```{r Gender/Sex}
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, fill = GENDER)) +
    geom_bar(position = "fill", color = "white") + 
  scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# Gender as predictor
summary((glm(MORTALITY ~ GENDER, data = patients_recoded_trim, family = binomial()))) # 0.00226

# There is a relationship between GENDER and MORTALITY with p-value 0.00226

#exp(coef(glm(MORTALITY ~ GENDER, data = patients_recoded_trim, family = binomial())))
```
```{r Ethnicity}
summary(patients_recoded_trim$ETHNICITY_BROAD)

ggplot(patients_recoded_trim, aes(x = MORTALITY, fill = ETHNICITY_BROAD)) + 
  geom_bar(aes(fill = ETHNICITY_BROAD), position = "dodge")
  #scale_fill_manual(values = c(Grey, Yellow, PennRed))

ggplot(data = patients_recoded_trim, aes(x = MORTALITY, fill = ETHNICITY_BROAD)) +
    geom_bar(position = "fill", color = "white") + 
  #scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# Ethnicity as predictor
summary((glm(MORTALITY ~ ETHNICITY_BROAD, data = patients_recoded_trim, family = binomial()))) # 0.00226

# There is no signficant association between any particular ETHNICITY and MORTALITY
```

```{r Insurance}
summary(patients_recoded_trim$INSURANCE)

ggplot(patients_recoded_trim, aes(x = MORTALITY, fill = INSURANCE)) + 
  geom_bar(aes(fill = INSURANCE), position = "dodge") +
  labs(y = "Count", x = "Mortality")
  #scale_fill_manual(values = c(Grey, Yellow, PennRed)) +

ggplot(data = patients_recoded_trim, aes(x = MORTALITY, fill = INSURANCE)) +
    geom_bar(position = "fill", color = "white") + 
  #scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# Insurance type as predictor
summary((glm(MORTALITY ~ INSURANCE, data = patients_recoded_trim, family = binomial())))

# There is a signficant association between all INSURANCE and MORTALITY (i.e. having any insurance)
```

```{r LOS_ICU}
# Length of stay (LOS) in the ICU as predictor
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, y = LOS_ICU)) +
    geom_boxplot() +
  labs(y = "Length of Stay in ICU (days)", x = "Mortality")

summary((glm(MORTALITY ~ LOS_ICU, data = patients_recoded_trim, family = binomial()))) # p-value <2e-16

# LOS_ICU was positively associated with MORTALITY with p-value <2e-16
```

```{r Delivery}
ggplot(data = patients_recoded_trim %>% filter(GENDER == "F"), aes(x = MORTALITY, fill = DELIVERY)) +
    geom_bar(position = "dodge", color = "white") + 
  scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# Delivery as predictor
summary((glm(MORTALITY ~ DELIVERY, data = patients_recoded_trim, family = binomial()))) # 0.00226

# There is a non-significant positive association between NO DELIVERY and MORTALITY
```

```{r C-section}
ggplot(data = patients_recoded_trim, aes(x = MORTALITY, fill = CSECTION)) +
    geom_bar(position = "dodge", color = "white") + 
  scale_fill_manual(values = c(Green, DarkBlue)) +
  theme_bw() +
  labs(y = "Proportion", x = "Mortality")

# C-section as predictor
summary((glm(MORTALITY ~ CSECTION, data = patients_recoded_trim, family = binomial()))) # 0.00226

# There is a non-significant positive association between NO CSECTION and MORTALITY
```

```{r Summary}
summary(patients_recoded_trim)

patients_recoded_trim %>% group_by(GENDER) %>% summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(INSURANCE) %>% summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(ADMISSION_TYPE) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(ETHNICITY_BROAD) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(DELIVERY, GENDER) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(CSECTION, GENDER) %>% 
  summarize(COUNT = n(), PROP = round(n()/54362*100,2))
patients_recoded_trim %>% group_by(MORTALITY) %>% summarize(COUNT = n(), PROP = round(n()/54362*100,2))
```

#### Logistic regression
```{r}
# narrowing to only binary variables of interest
patients_recoded_trim_glm <- patients_recoded_trim %>% select(-(AGE_ADM:LOS_ICU))

# logistic regression model
patients_glm <- glm(MORTALITY ~ ., data = patients_recoded_trim_glm, family = binomial(logit))
summary(patients_glm)

# There were signficant positive associations between all insurance types and MORTALITY
```

#### Limitations
- Because all dates in the database have been shifted to protect patient confidentiality, I could not compare c-section rates to national rates reported by the CDC per year. Reference: https://mimic.physionet.org/mimicdata/time/
  - Date shifting removed:
      - Year - The year is randomly distributed between 2100 - 2200
      - Day of the month - The absolute day of the month is not preserved
      - Inter-patient information - Two patients in the ICU on 2150-01-01 were not in the ICU at the same time
- Not all `ICD9_CODES` in the database match to a ICD code
- Population is mostly older which means fewer pregnancies, so pregnancy-related outcomes are not easily studied

### References
- Johnson, A., Pollard, T., Mark, R. (2016) MIMIC-III, a freely accessible critical care database. Scientific Data (2016). DOI: 10.1038/sdata.2016.35. Available from: http://www.nature.com/articles/sdata201635
