```{r}
# BMIN503 Final Project
# Author: David Hammill
# Date: 10 Dec 2021
# Description: Final Project For BMIN503
```



### Overview
  - This project is hypothesizing that hemoconcentration on cardiopulminary bypass (CPB) with del Nido cardioplegia used as an arresting agent leads to increased Acute Kidney Injury postoperatively. 
  - The data we are using is threefold
    - Data from the Society of Thorastic Surgeouns conisiting of open heart bypass cases from the UPHS Health System
    - Data from PennChart EMR
    - Data from Quantum (bypass data)
  - Factulty 
    - Nemesh Desai, M.D.
    - Dr. Assad, 
  - [Link](https://github.com/DavidMHammill/BMIN503_Final_Project) to repository
 
### Introduction
  - Cardiac surgery patients experience AKI at a rate higher than any other group of patients which impacts length of stay and outcomes dramatically.  Thirty Percent of patients who receive CPB will experience AKI.  Cardioplegia which is used to arrest the heart contributes additional volume on CPB.  THis is especially true in patients who receive del Nido cardioplegia.  Typically this volume is then removed on CPB with a hemoconcentrator.  Previous studies have proven an increase in AKI associated with hemoconcentration.  It's clear then that there is a correlation. This study will go one step further and look at this group of patients and determine if using  a diuretic to improve the removal of this volume with the use of the kidneys improved post op creatinine levels.
  
  
### Methods
  - The data will be cleaned, removing any Personal Patient Data
  - This study will categorize the cases assigning two levels, cases which use del Nido cardioplegia along with hemoconcentration.
  - This study will adjust for patients with pre-existing conditions that will place them at greater risk for AKI
  - Perform bivariate analysis on the two groups, the outcome being whether or not the patient experienced AKI
  - Data
    - STS Data:
    
```{r}
# Read the data into dataframe
sts = read.csv("BMIN503finalproject_data/sts_clean.csv")

# Summarize the dataframe
str(sts)
```
    - Dataset 2
    - Dataset 3
  
```{r}
# I have received the quantum data and just received the Pennchart data with the correct creatinine column. In the meantime I will test the correlation of bypass time with AKI
# Compare creatinine levels with bypass time (bivariate analysis) 

library(tidyr)

# Create a function to normalize a vector based on min / max values
min_max_norm <- function(x) {
    (x - min(x)) / (max(x) - min(x))
}

df = sts[,c("Postoperative.Creatinine.Level", "Cardiopulmonary.Bypass.Time")]
df = drop_na(df)
df_norm = as.data.frame(lapply(df, min_max_norm))
results = cor.test(df_norm$Cardiopulmonary.Bypass.Time, df_norm$Postoperative.Creatinine.Level)

```
  
  
### Results
  - p-value of 2.514e-12  => Strong indicator that bypass time is strongly correlated with postoperative creatinine level.
  - Bullet 2
  
### References
  - Bullet 1
  - Bullet 2
  - Bullet 3

