---
title: "Your_Title"
subtitle: "BMIN503/EPID600 Final Project"
author: "First_Name Last_Name"
format: html
editor: visual
embed-resources: true
---

------------------------------------------------------------------------

```{r}
library("nhanesA")
library("haven")
library("dplyr")
```

### Overview

Volatile Organic Compounds (VOC's) are a class of air pollutants present via environmental air and water systems with several health implications, including eye/nose/throat irritation, headache and loss of coordination, nausea, and liver, kidney, and central nervous system damage.^1^ This project will assess prominent social factors on VOC exposure, including socioeconomic status, locality, and occupation. To examine this relationship, I will assess the NHANES 2019 -- 2020 urine VOC metabolite data for three prominent classes: xylenes, acrylamides, and butanes. To develop this plan, I have met with Dr. Kelly Getz and Dr. John Holmes, who provided insights on classes of toxicants, limitations regarding the dose-response of toxicants, and potential analytic approaches for NHANES data, including k-means clustering. Materials for the final project can be [found in this Github repository.](https://github.com/jennko1230/BMIN503_Final_Project)

### Introduction

VOCs are a classification of pollutants present in the atmosphere that are linked with poor health outcomes in humans. Many VOCs travel through air and water sources to travel through the human body and are associated with symptoms/diseases, including asthma, atopic dermatitis, neurological symptoms, and cancer.^1^ Up to 25% of all VOC's in our atmosphere are linked to anthropogenic sources, including petroleum and gas extraction, fossil fuel burning, chemical and industrial processes, mining, and agricultural practices.^1^ A method of quantifying VOC exposure in humans is to assess urine metabolites, a biomarker that has marked elevated concentrations for longer than their blood/exhaled air specimens.

VOC exposure is a pressing that needs interdisciplinary collaboration to evaluate and combat alarming VOC emissions thoroughly. VOC exposure cutoffs and their determination of safety levels implement principles common in toxicology. Toxicologists and public health organizations must collaborate with prominent policy members to thoroughly address/minimize community exposures. Many social determinants of health can influence VOC exposure and substantially negatively influence individuals' health. Factors, including place of residence, occupation, and socioeconomic status, can impact one's exposure to pollutants. This data analysis will seek to evaluate a series of community factors and how these factors can influence one's exposure to VOCs. This analysis aims to build a predictive model to determine individuals' VOC exposure, which could support the development of public health programs tailored to factors that influence VOC exposure.

### Methods

```{r}
#Load NHANES dataset
varlist <- c("DEMO", "DUQ") 
varlist_years <- paste0(rep(varlist, each = 2), c("_I", "_J"))
list_all <- sapply(varlist_years, function(x) {data.frame(nhanes(x))}) 
for(i in 1:length(list_all)) {
  assign(names(list_all)[i], list_all[[i]])
} 
for (i in 1:length(varlist)) {
  assign(varlist[i], plyr::rbind.fill(mget(grep(varlist[i], ls(), value = T))))
} 
rm(list = grep("_[IJ]", ls(), value = T))
nhanes.data <- full_join(get(varlist[1]), get(varlist[2]), by = "SEQN")
rm(list = ls()[-which(ls() == "nhanes.data")])

#Pull variables of interest - age, gender, race, education, income, 
nhanes.practice <- nhanes.data |>
  select(id = SEQN, gender = RIAGENDR, age = RIDAGEYR, 
           race = RIDRETH1, education = DMDEDUC2, income = INDHHIN2,
           marijuana = DUQ200, cocaine = DUQ250, heroin = DUQ290,
           meth = DUQ330) |>
  mutate(gender = factor(gender, levels = c(1, 2), 
                           labels = c("men", "women"))) |>
  mutate(race = factor(race, levels = c(3, 1, 2, 4, 5),
                         labels = c("White", "MexicanAmerican", 
                                    "Hispanic", "Black", "Other"))) |>
 
  filter(income %in% c(1:10, 14, 15)) |>
  mutate(income.short = cut(income, breaks=c(0, 5, 10, 15),
                              labels=c("<$25K", "$25-75K", ">$75K"))) |>
  filter(age >= 18)
```

### Results

Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
