---
title: "Examining Pregnancy Prevention Methods among Adolescents AFAB"
subtitle: "BMIN503/EPID600 Final Project"
author: "Gabrielle Zuckerman"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

## Overview {#sec-overview}

Preventing unintended teenage pregnancy is a public health priority that can be mitigated by individual uptake of effective contraception in sexually active teens. The Centers for Disease Control and Prevention publishes a publicly available dataset called the National High School Youth Risk Behavior Survey (YRBS) that contains information on important health behaviors and experiences in American high school students. Thus, the goal of this project is to use data from the YRBS to see what proportion of sexually active adolescents that were assigned female at birth (AFAB) utilize pregnancy prevention methods, which pregnancy methods are most popular among them, and whether these outcomes vary by sexual minority status.

## Introduction {#sec-introduction}

Despite decreases over the past several decades, unintended teenage pregnancy continues to persist as a public health problem that, compared to other high income countries, is especially prevalent in the United States.^1^ One objective of Healthy People 2030, an initiative from the US Department of Health and Human Services that identifies public health priorities to improve health and well-being across the United States,^2^ is to \"increase the proportion of adolescent females who used effective birth control the last time they had sex.\"^3^ In fact, several Healthy People 2030 objectives are related to adolescent sexual education, behaviors, and outcomes.^4^ There has also been extensive evidence demonstrating that teen pregnancy has negative lifelong consequences for both the birthing parent and the child.^5^ Teen pregnancy is also a health disparities issue. Disparities have been observed by age, race, geography, and socioeconomic status.^6^ Moreover, prior work has found that sexual minority adolescents are at a greater risk of teenage pregnancy than their heterosexual peers.^7--10^ This justifies an examination of contraceptive use by sociodemographic factors.

Addressing and acquiring knowledge about adolescent contraceptive use requires an interdisciplinary approach. In addition to thinking about unintended pregnancy as a public health issue, behavioral experts (e.g., psychologists, health behaviorists) can contribute to our understanding of which adolescents choose to use contraception methods and what characteristics they express. Additionally, contraceptive access is of concern as adolescents may not have the financial or transportational means to ascertain pregnancy prevention and/or can lack health literacy around this sexual health. Lastly, because sex is particularly stigmatized among adolescents, teens cannot necessarily rely on their guardians or medical providers for sex-related advice or assistance in utilizing contraception.

## Methods {#sec-methods}

### Load Necessary Packages

We will first load several packages that will be used to analyze the YRBS data.

```{r}
library(tidyverse)
library(sf)
library(RColorBrewer)
library(tigris)
library(tmap)
library(tmaptools)
library(grid)
library(leaflet)
library(forecast)
library(ggthemes)
library(png)
```

### Data Import

YRBS does not have data in CSV form, so the dataset had to be downloaded and opened in SAS. Additionally, YRBS provides SAS code for loading and formatting the data. Thus, the data is already properly labelled and formatted. I additionally filtered out data from years outside of 2015-2021. No other data cleaning was conducted.

```{r}
yrbs <- read.csv("/Users/gmzuck/Library/Mobile Documents/com~apple~CloudDocs/Desktop/UPenn/Coursework/Fall 2023/EPID 6000/yrbs2015-2021.csv")
```

### Data Cleaning

Missing values are currently coded as the string "Missing" and must be changed to NA.

```{r}
yrbs[yrbs=="Missing"] <- NA
```

We only want to include students who report that their sex is female.

```{r}
yrbs.female <- yrbs[yrbs$sex == "Female" & !is.na(yrbs$sex),]
```

Now, we will limit the dataset to only include variables of interest as there are currently \>200 variables to look through in the dataset.

```{r}
#yrbs.female <- yrbs.female %>% select(c("year","weight","stratum","PSU","record", #Survey Design "age","sex","grade","race4","race7","sexid","sexid2", #Demographics "q59","q60","q61","q62","q63","q64","q65","sexpart","sexpart2", #Sex-related "q30","q34","q37","q38","q40","q45","q48","q49","q51","q52","q53","q54","q55","q56","q57" #Risky Behaviors))
```

### Data Checks

It is good statistical practice to check the distributions of the variables that will be used to ensure that they make sense.

```{r}
```

## Results {#sec-results}

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

## Conclusion

This the conclusion. The @sec-results can be invoked if you'd like.

## References

1.  Trends in Teen Pregnancy and Childbearing \| HHS Office of Population Affairs. Accessed November 9, 2023. https://opa.hhs.gov/adolescent-health/reproductive-health-and-teen-pregnancy/trends-teen-pregnancy-and-childbearing
2.  About Healthy People 2030 - Healthy People 2030 \| health.gov. Accessed November 9, 2023. https://health.gov/healthypeople/about
3.  Increase the proportion of adolescent females who used effective birth control the last time they had sex --- FP‑05 - Healthy People 2030 \| health.gov. Accessed November 9, 2023. https://health.gov/healthypeople/objectives-and-data/browse-objectives/family-planning/increase-proportion-adolescent-females-who-used-effective-birth-control-last-time-they-had-sex-fp-05
4.  Adolescents - Healthy People 2030 \| health.gov. Accessed November 9, 2023. https://health.gov/healthypeople/objectives-and-data/browse-objectives/adolescents
5.  Hofferth SL, Hayes CD, eds. *Risking the Future: Adolescent Sexuality, Pregnancy, and Childbearing, Volume II: Working Papers and Statistical Appendices*.; 1987. doi:10.17226/946
6.  Martin JA, Hamilton BE, Osterman MJK, Driscoll AK, Drake P. Births: Final Data for 2016. *Natl Vital Stat Rep Cent Dis Control Prev Natl Cent Health Stat Natl Vital Stat Syst*. 2018;67(1):1-55.
7.  Saewyc EM, Poon CS, Homma Y, Skay CL. Stigma management? The links between enacted stigma and teen pregnancy trends among gay, lesbian, and bisexual students in British Columbia. *Can J Hum Sex*. 2008;17(3):123-139.
8.  Lindley LL, Walsemann KM. Sexual Orientation and Risk of Pregnancy Among New York City High-School Students. *Am J Public Health*. 2015;105(7):1379-1386. doi:10.2105/AJPH.2015.302553
9.  Goldberg SK, Reese BM, Halpern CT. Teen Pregnancy Among Sexual Minority Women: Results From the National Longitudinal Study of Adolescent to Adult Health. *J Adolesc Health Off Publ Soc Adolesc Med*. 2016;59(4):429-437. doi:10.1016/j.jadohealth.2016.05.009
10. Charlton BM, Corliss HL, Missmer SA, Rosario M, Spiegelman D, Austin SB. Sexual orientation differences in teen pregnancy and hormonal contraceptive use: an examination across 2 generations. *Am J Obstet Gynecol*. 2013;209(3):204.e1-8. doi:10.1016/j.ajog.2013.06.036
