---
title: "Influenza"
subtitle: "BMIN503/EPID600 Final Project"
author: "Grace Li"
format: html
editor: visual
embed-resources: true
---

------------------------------------------------------------------------

Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers. Feel free to change the theme and other display settings, although this is not required.

### Overview

Give a brief a description of your project and its goal(s), what data you are using to complete it, and what two faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

### Introduction

Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

### Methods

Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

```{r}
#| label: load-packages
#| include: false

library(tidyverse)
library(dplyr)
```

```{r}
#| label: ili_case

state_ili_case <- read_csv("StateDatabySeason.csv")

state_ili_case <- state_ili_case |>
        select(-c(URL, WEBSITE)) |>
        rename(state = STATENAME, 
               level = `ACTIVITY LEVEL`, 
               label = `ACTIVITY LEVEL LABEL`,
               weekend = WEEKEND,
               week = WEEK,
               season = SEASON) |> 
        mutate(level = as.numeric(gsub("Level ", "", level))) |>
        mutate(ordered_week = (week - 40) %% 52 + 1) 

states <- data.frame(state = datasets::state.name, division = datasets::state.division) 

state_ili_case <- inner_join(state_ili_case, state_division, by = "state")

pa_ili_case <- state_ili_case |>
        filter(state == "Pennsylvania")

pa_ili_case |> filter(season == "2018-19") |>
        ggplot(aes(x = ordered_week, y = level)) +
        geom_point() +
        geom_line() +
        scale_x_continuous(breaks = 1:52, labels = c(40:52, 1:39)) +
        labs(x = "Week", y = "ILI Activity Level", title = "2018-19 ILI Activity") +
        theme_classic()
        
```

### Results

Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
