---
title: "BMIN503/EPID600 Final Project"
author: "Elizabeth Mahanna Gabrielli, MD"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  


### Overview

This project will be investigating 6 month outcomes in stroke patients who are discharged from the hospital to a nursing home. We will formulate 3 models. One model will address the question: What factors predict if these patients survive to 6 months? The second model will address the question: What factors predict if these patients improve their living situation (either living at home, a relatives or a residential community) or remain in the nursing home? The third model will address the question: What factors predict their level of function? Are they dependent or not-dependent? 


https://github.com/emahannagabrielli/BMIN503_Final_Project

Faculty advisors:

Dr. Monisha Kumar, Associate Professor Department of Neurology. At first I was going to have 2 research questions about final functional outcome and final living place, which each included survival as an outcome.  After, discussing with Dr.Kumar, predicting survival on its own is an important question. Then we decided because we will be looking at survival separately, we will only investigate final functional outcome and final living place for patients who survived 6 months after discharge. In addition, Dr. Kumar helped me decide how to group the final outcomes into clinical relevant groups. 

Dr. Rebecca Hubbard, Associate Professor of Biostatistics and Todd Miano helped me in determing how to analyze the data.  Dr. Hubbard at first suggested I use multinominal regression when my original research questions had 3 categorial variables with the mlogit().  After we changed questions, she suggested multivariate linear regression with the glm() function in R. She helped me decide how to group patients and which data to remove because of incomplete outcomes or not a clear outcome.  

### Introduction 

Acute stroke is a common and highly morbid disease. Worldwide it is the second most common cause of mortality and third most common cause of disability. More than 7 million people in the United States survive a stroke each year. Improvements in acute stroke treatment such as intravenous tPA and thombectomy have greatly decreased stroke mortality and disabilty. However, there are still a large number of patients who are not candidates for these therapies due to timing of stroke onset, medical comorbidities, medications and more. When a patient is not doing well in the hospital and we predict they will end up being discharged to a nursing home, families have many questions. Key questions are often: "Will they be in a nursing home forever or only temporarily?" "Will they be dependent on others for their care?" If we knew factors associated with these poor outcomes, it would allow physicians to better council families. This would allow the families to either plan for the future expected outcomes, or in some cases the families may decide against full aggressive care and concentrate on hospice measures only.

This problem requires a bioinformatician, statistical analysis and clinical knowledge of the problem from both neurology and intensive care point of view. As a bioinformatician in training, I have the knowledge on how to access this database in order to perform a novel secondary analysis.  With statistical expertise, our team is able to appropriately model and analyse our data. Along with this we require the clinical knowledge for the scope of the problem, the clinical significance and to help trouble shoot our analysis. For example, it was discovered some final cohorts on functional status at discharge had very small cohorts.  Using expert consultation from a neurologist we determined it is appropriate to group both the recovered patients and the not-recovered but not-dependent patients together. 

In patients who are discharged to a nursing home after an acute ischemic stroke, we aimed to 1) Build clinical prediction models that can identify key in-hospital variables which determine 6 month outcome, final living place and final functional outcome.
2) Describe the preditive performance of such models using both traditional regression and random forest approaches.
3) Use an easily reproducible, open-source workflow. 

We hypothesized we would be able to develop accurate models as described above. 

### Methods
The data used comes from the International Stroke Trial (IST) database.  This trial occured in the late 1990s and had over 19,000 subjects. It took place in over 35 countries and had a disproportionally older aged patients compared to other stroke trials.  This trial occured prior to thrombolysis or thrombectomy.  Therefore, it is a good database to analyze outcomes in patients for whom these therapies either were contraindicated or for whom they did not work. There is available data on variables assessed at randomization, which occured within 48 hours of admission, early outcome at 14 days and long term outcome at 6 months. 

First, we determined how many subject were in each outcome of interest. Based on this data and some low number of patients per outcome we made decisions to group outcomes together. 

Second, we created a clean database. This database included new dependent variables.  It also changed some independent variables into factors or into new groupings or renamed them into more understandable names.  Finally we kept only independent variables that pertained to the originial acute hospital stay as we wanted to develop a prediciton model based upon intial hospital characteristics. We then made 3 databases from this database with only the specific outcome of interest (survival at 6 months, final living place, or final functional outcome).  We did this as we expected final living place and final functional outcome to have an association and we did not want that association to be taken into account by the Random Forest. 

We then began exploratory analyses and descriptive statistics.  We made histograms of age and length of stay.  We compared age versus each outcome variable.  We compared stroke type and each outcome variable.  We compared gender and each outcome variable. We compared initial blood pressure readings and each outcome variable. 

In our next step, we aimed to determine which variables are most important and should be included in our prediction models.  We did this by performing Random Forest on all included independent variables for each of the outcomes.  We then chose the top 7 independent variables accross all 3 outcomes.  We chose to use 7 independent variables as our outcome with the smallest cohort was 69 subjects. After choosing the 7 most important independent variables, we used these variables in a linear regresssion prediction model and a random forest prediction model.  We did cross validation using a K-fold method, meaning data was split into K randomly selected sub-samples. K-1 sub-samples were used for model training. One subsample was used as an independent test set. We evaulated the models with Receiver-Operating Characgteristics (ROC) and Area Under the ROC curve (AUC).

```{r, eval=TRUE}
library(dplyr)
library(tidyr)
library(tidyverse)
IST <- read.csv("IST_corrected.csv", header=TRUE) #Link to websiste with CSV file https://datashare.is.ed.ac.uk/handle/10283/128
#class(IST)
#str(IST)
#dim(IST)
sum(is.na(IST$DPLACE))
IST_nh <- filter(IST, DPLACE=="D" & DNOSTRK=="N") #Filtering patients who were discharged to place "D," which is a nursing home and filtering out patients who did not have a stroke. 
dim(IST_nh) #cohort is now 527 subjects

#Link to datadictionary: https://datashare.is.ed.ac.uk/bitstream/handle/10283/128/IST_variables.pdf?sequence=10&isAllowed=y

#First, we determined how many subject were in each outcome of interest. Based on this data and some low number of patients per outcome we made decisions to group outcomes together. 
#Alive 6 month outcome
IST_nh_dead <- filter(IST_nh, OCCODE=="1") #subsetting subjects who were dead at 6 months
dim(IST_nh_dead) #132 dead

#Living Place outcomes
IST_nh_home <- filter(IST_nh, FPLACE=="A" & DIED=="0") #subsetting to subjects who were living at home at 6 months
dim(IST_nh_home) #118 subjects at home in 6 months
IST_nh_rel <- filter(IST_nh, FPLACE=="B" & DIED=="0") #subsetting to subjects who were living at a relatives home at 6 months
dim(IST_nh_rel) #12 subjects at relatives home
IST_nh_rescare <- filter(IST_nh, FPLACE=="C" & DIED=="0") #subsetting to subjects who were living in residential care at 6 months
dim(IST_nh_rescare) #55 in residential care
IST_nh_improvedliving <- filter(IST_nh, FPLACE=="A"|FPLACE=="B"|FPLACE=="C" & DIED=="0") #subsetting to subjects who improved to a lower level of care living situation at 6 months. 
dim(IST_nh_improvedliving) #185 improved living condition
IST_nh_nh <- filter(IST_nh, FPLACE=="D" & DIED=="0") #subsetting to subjects who remained in a nursing home at 6 months
dim(IST_nh_nh) #187 remained in NH
IST_nh_hosptial <- filter(IST_nh, FPLACE=="E" & DIED=="0") #subsetting to subjects who were in a hospital setting at 6 months.  I discussed with Dr. Hubbard and Dr. Kumar and we will remove these from the dataset as it is unknown their living place when not in the hospital. 
dim(IST_nh_hosptial) #10 in hospital
IST_nh_unknown <- filter(IST_nh, FPLACE=="U" & DIED=="0")
dim(IST_nh_unknown) #0 unknown

#Functional Outcomes
IST_nh_dependent <- filter(IST_nh, OCCODE=="2" & DIED=="0") #subsetting subjects who were dependent at 6 months
dim(IST_nh_dependent) #313 dependent
IST_nh_notrecovered <- filter(IST_nh, OCCODE=="3" & DIED=="0") #subsetting subjects who were not recovered at 6 months
dim(IST_nh_notrecovered) #43 not recovered
IST_nh_recovered <- filter(IST_nh, OCCODE=="4" & DIED=="0") #subsetting subjects who were recovered at 6 months
dim(IST_nh_recovered) #26 recovered
IST_nh_notdependent <- filter(IST_nh, OCCODE=="3"|OCCODE=="4" & DIED=="0") #combining subset of subjects who were alive and not dependent at 6 months. 
dim(IST_nh_notdependent) #69 nondependent
IST_nh_missingstatus <- filter(IST_nh, OCCODE=="8"|OCCODE=="9" & DIED=="0")
dim(IST_nh_missingstatus) #1 missing status --> will remove

#Next step is to clean the dataset.
sum(is.na(IST_nh)) #There are 5466 missing values. 

NH_clean <-IST_nh %>% #Create a new data set with outcome variables and only pertinent data
  mutate(alive6m = factor(DIED, levels=c(0,1), labels=c("survived", "died"))) %>% #create alive at 6 months variable
  mutate(FLP = ifelse(FPLACE=="A"|FPLACE=="B"|FPLACE=="C", "improved",ifelse(FPLACE=="D", "remainsNH", NA))) %>% #create outcome of Final Living Place (FLP) with 2 outcomes: remains in NH or improved living situation.
  mutate(FLP = factor(FLP, levels=c("improved","remainsNH"), labels=c("improved","remainsNH"))) %>%
  mutate(FFO = ifelse(OCCODE=="3"|OCCODE=="4", "notdependent",ifelse(OCCODE=="2", "dependent", NA))) %>% #create outcome variable Final Functional Outcome (FFO) as either dependent or notdependent
  mutate(FFO = factor(FFO, levels=c("notdependent", "dependent"), labels=c("notdependent", "dependent"))) %>%
  #Mutate each deficit outcome into Yes, No/Unkown instead of Yes, No or Can't assess. This is because Can't Assess may be associated with other worse outcomes and not have anything to do with the deficit in question. 
 mutate(DefFace = ifelse(RDEF1=="Y", "yes", "no.unk")) %>%
  mutate(DefFace = factor(DefFace, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(DefArm = ifelse(RDEF2=="Y", "yes", "no.unk")) %>%
  mutate(DefArm = factor(DefArm, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(DefLeg = ifelse(RDEF3=="Y", "yes", "no.unk")) %>%
  mutate(DefLeg = factor(DefLeg, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(DefDysphagia = ifelse(RDEF4=="Y", "yes", "no.unk")) %>%
  mutate(DefDysphagia = factor(DefDysphagia, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(DefHemianopia = ifelse(RDEF5=="Y", "yes", "no.unk")) %>%
  mutate(DefHemianopia = factor(DefHemianopia, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(DefVisual = ifelse(RDEF6=="Y", "yes", "no.unk")) %>%
  mutate(DefVisual = factor(DefVisual, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(DefPostfossa = ifelse(RDEF7=="Y", "yes", "no.unk")) %>%
  mutate(DefPostfossa = factor(DefPostfossa, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(DefOther = ifelse(RDEF8=="Y", "yes", "no.unk")) %>%
  mutate(DefOther = factor(DefOther, levels=c("yes", "no.unk"), labels=c("yes", "no.unk"))) %>%
  mutate(Cerebralbleed14 = factor(H14, levels=c(1,0), labels=c("yes", "no"))) %>%
  mutate(Ischemic14 = factor(ISC14, levels=c(1,0), labels=c("yes","no"))) %>%
  mutate(Stroke14 = factor(STRK14, levels=c(1,0), labels=c("yes", "no"))) %>%
  mutate(StrokeType = factor(STYPE, levels=c("LACS","PACS","POCS","TACS"), labels=c("Lacunar", "Partial Anterior Circ", "Posterior Circ", "Total Anterior Circ"))) %>%
  mutate(gender = factor(SEX, levels=c("M", "F"), labels=c("male", "female"))) %>%
  dplyr::select(-(HOSPNUM:RDELAY), -(RATRIAL:RASP3), -(RDEF1:RDEF8), -(RDATE:DALIVE), -(DPLACE:NCB14), -STYPE) %>% #removing columns not pertinent
  rename(HospLOS=DALIVED) %>%
  drop_na(alive6m) #remove subjects that don't have survival outcome

head(NH_clean)

#Make dataset for Alive at 6 months outcome
NH_alive6m <- NH_clean %>%
  select(-FLP, -FFO)  #remove other outcome variables
sum(is.na(NH_alive6m)) #0 NA values in deficits
Count_alive6m <- NH_alive6m %>% count("survived",alive6m)


#Make dataset for Final Living Place at 6 months (remains in NH vs improved level of living)
NH_FLP <- NH_clean %>%
  select(-alive6m, -FFO) %>% #remove other outcome variables
    drop_na(FLP) #remove subjects that don't have a FLP outcome variable
sum(is.na(NH_FLP)) #No NA values
Count_FLP <- NH_FLP %>% count("remainsNH",FLP)


#Make dataset for Final Functional Outcome at 6 months (dependent vs notdependent)
NH_FFO <- NH_clean %>% 
  select(-alive6m, -FLP) %>% #remove other outcome variables
  drop_na(FFO) #remove subjects without FFO outcome
sum(is.na(NH_FFO)) #No NA values
Count_FFO <- NH_FFO %>% count("dependent",FFO)

```

``` {r eval=FALSE}
install.packages("mlogit")
install.packages("pROC")
install.packages("randomForest")
```
```{r eval = TRUE}
#Exploratory analysis and Descriptive Statistics
#Main results shown in RESULTS section. This is other exploratory analysis not to be reported in manuscript.  
library(ggplot2)
mean(NH_clean$AGE, na.rm=TRUE) #Mean age is 78.6
quantile(NH_clean$AGE, na.rm=TRUE) #0%  25%  50%  75% 100%: 40   74   80   84   98
#Histogram of Age
ggplot(data=NH_clean, aes(x=AGE)) +
  geom_histogram(binwidth = 3)+
  labs(title = "Histogram of Age in entire cohort")

mean(NH_clean$HospLOS, na.rm=TRUE) #Mean Hospital LOS is 27.6 days
quantile(NH_clean$HospLOS, na.rm=TRUE) #0%  25%  50%  75% 100%: 0   13   20   33  369

#Histogram of Hospital LOS
ggplot(data=NH_clean, aes(x=HospLOS)) +
  geom_histogram(binwidth = 3)+
  labs(title = "Histogram of Hospital Length of Stay in entire cohort")

#Box plot of Hosp LOS vs 6 m survival 
ggplot(data=NH_clean, aes(x=factor(alive6m), y=HospLOS)) +
  geom_boxplot() +
  labs(title="Hospital Length of Stay (LOS) vs 6 Month survival") +
  labs(x=("6 month survival"), y=("Hospital LOS (days)")) +
  ylim(c(0,100))

#Box plot of Systolic Blood Pressure at randomization vs 6 m survival
ggplot(data=NH_clean, aes(x=factor(alive6m), y=RSBP)) +
  geom_boxplot() +
  labs(title="Systolic blood pressure vs 6 Month survival") +
  labs(y="Systolic BP (mmHg)", x=("6 month survival"))

#Box plot of SBP vs Final Living Place
ggplot(data=NH_FLP, aes(x=factor(FLP), y=RSBP)) +
  geom_boxplot() +
  labs(title="Systolic blood pressure and Final Living Place") +
  labs(y="Systolic BP (mmHg)", x=("Final Living Place"))

#Box plot of SBP vs Final Functional Outcome
ggplot(data=NH_FFO, aes(x=factor(FFO), y=RSBP)) +
  geom_boxplot() +
  labs(title="Systolic blood pressure and Final Funcitonal Outcome") +
  labs(y="Systolic BP (mmHg)", x=("Final Funcitonal Outcome"))
```

``` {r eval=TRUE}
library(randomForest)
library(mlogit)
library(ggplot2)
library(tidyverse)
library(dplyr)


#Random forests for outcome of 6 months survival of patients discharged from the hospital to a nursing home.
Alive6m.rf <- randomForest(alive6m ~ ., data=NH_alive6m, ntree=100, importance=TRUE)
Alive6m.rf
imp.alive6m <- Alive6m.rf$importance
head(sort(imp.alive6m[,4], decreasing=TRUE), n=10)
alive6m.rf.pred <- predict(Alive6m.rf, NH_alive6m, type="prob")
rf.pred.alive6m <- alive6m.rf.pred[, 2]
head(alive6m.rf.pred)


#Model evaluation it K-fold Cross validation
N = nrow(NH_alive6m)
K = 10
set.seed(1234)
s = sample(1:K, size=N, replace=T)
alive6m.pred_outputs.rf <- vector(mode="numeric", length=N)
alive6m.obs_outputs <- vector(mode="numeric", length=N)
offset <- 0
    
for(i in 1:K){
    train <- filter(NH_alive6m, s != i)
    test <- filter(NH_alive6m, s == i)
    alive6m.obs_outputs[1:length(s[s==i]) + offset] <- test$alive6m

    #RF train/test
    rf <- randomForest(alive6m ~., data=train, ntree=100)
    rf.pred.curr <- predict(rf, newdata=test, type="prob") 
    alive6m.pred_outputs.rf[1:length(s[s==i]) + offset] <- rf.pred.curr[,2]

    offset <- offset + length(s[s==i])
}


#Random forests for outcome of final Living Place of patients discharged from the hospital to the nursing home and alive at 6 months.
FLP.rf <- randomForest(FLP ~ ., data=NH_FLP, ntree=100, importance=TRUE)
FLP.rf
imp.FLP <- FLP.rf$importance
head(sort(imp.FLP[,4], decreasing=TRUE), n=10)
FLP.rf.pred <- predict(FLP.rf, NH_FLP, type="prob")
rf.pred.FLP <- FLP.rf.pred[, 2]

#Model evaluation it K-fold Cross validation
N = nrow(NH_FLP)
K = 10
set.seed(1234)
s = sample(1:K, size=N, replace=T)
#pred_outputs.glm <- vector(mode="numeric", length=N)
FLP.pred_outputs.rf <- vector(mode="numeric", length=N)
FLP.obs_outputs <- vector(mode="numeric", length=N)
offset <- 0
for(i in 1:K){
    train <- filter(NH_FLP, s != i)
    test <- filter(NH_FLP, s == i)
    FLP.obs_outputs[1:length(s[s==i]) + offset] <- test$FLP
   
    #RF train/test
    rf <- randomForest(FLP ~., data=train, ntree=100)
    rf.pred.curr <- predict(rf, newdata=test, type="prob") 
    FLP.pred_outputs.rf[1:length(s[s==i]) + offset] <- rf.pred.curr[,2]
    offset <- offset + length(s[s==i])
}

#Random forests for outcome of final functional outcome (FFO) of patients discharged from the hospital to the nursing home and alive at 6 months
FFO.rf <- randomForest(FFO ~ ., data=NH_FFO, ntree=100, importance=TRUE)
FFO.rf
imp.FFO <- FFO.rf$importance
head(sort(imp.FFO[,4], decreasing=TRUE), n=10)
FFO.rf.pred <- predict(FFO.rf, NH_FFO, type="prob")

#Model evaluation it K-fold Cross validation
N = nrow(NH_FFO)
K = 10
set.seed(1234)
s = sample(1:K, size=N, replace=T)
#pred_outputs.glm <- vector(mode="numeric", length=N)
FFO.pred_outputs.rf <- vector(mode="numeric", length=N)
FFO.obs_outputs <- vector(mode="numeric", length=N)
offset <- 0
for(i in 1:K){
    train <- filter(NH_FFO, s != i)
    test <- filter(NH_FFO, s == i)
    FFO.obs_outputs[1:length(s[s==i]) + offset] <- test$FFO
   
    #RF train/test
    rf <- randomForest(FFO ~., data=train, ntree=100)
    rf.pred.curr <- predict(rf, newdata=test, type="prob") 
    FFO.pred_outputs.rf[1:length(s[s==i]) + offset] <- rf.pred.curr[,2]
    offset <- offset + length(s[s==i])
}


#Assessing all 3 models the top 4 variables are consistent: AGE, HospLOS (hospital length of stay), RSBP (first SBP at time of randomization), STYPE (stroke type). Then the next most common was def.dysphagia (dysphagia deficits)

#GLM of alive at 6 months with top variables
alive6m.top.glm <-glm(alive6m ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data= NH_alive6m, family = binomial(logit))
summary(alive6m.top.glm) #Age, Stroke type:Total Anterior circulation, hospital LOS, No Dysphagia
alive6m.top.glm.pred <- predict(alive6m.top.glm, NH_alive6m, type="response")


#Random forests for outcome of 6 month survival with top variables.
alive6m.top.rf <- randomForest(alive6m ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=NH_alive6m, ntree=100, importance=TRUE)
alive6m.top.rf
imp.alive6m.top <- alive6m.top.rf$importance
head(sort(imp.alive6m.top[,4], decreasing=TRUE), n=10)
alive6m.top.rf.pred <- predict(alive6m.top.rf, NH_alive6m, type="prob")

#Model evaluation it K-fold Cross validation
N = nrow(NH_alive6m)
K = 10
set.seed(1234)
s = sample(1:K, size=N, replace=T)
alive6m.top.pred_outputs.glm <- vector(mode="numeric", length=N)
alive6m.top.pred_outputs.rf <- vector(mode="numeric", length=N)
alive6m.top.obs_outputs <- vector(mode="numeric", length=N)
offset <- 0
for(i in 1:K){
    train <- filter(NH_alive6m, s != i)
    test <- filter(NH_alive6m, s == i)
    alive6m.top.obs_outputs[1:length(s[s==i]) + offset] <- test$alive6m
   #GLM train/test
    glm <- glm(alive6m ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=train, family=binomial(logit))
    glm.pred.curr <- predict(glm, test, type="response")
    alive6m.top.pred_outputs.glm[1:length(s[s==i]) + offset] <- glm.pred.curr
    
    #RF train/test
    rf <- randomForest(alive6m ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=train, ntree=100)
    rf.pred.curr <- predict(rf, newdata=test, type="prob") 
    alive6m.top.pred_outputs.rf[1:length(s[s==i]) + offset] <- rf.pred.curr[,2]
    
    offset <- offset + length(s[s==i])
}

#GLM of Final Living Place with top variables
FLP.top.glm <-glm(FLP ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data= NH_FLP, family = binomial(logit))
summary(FLP.top.glm) #Age, stroke type(total anterior circulation), hospital LOS, No dysphagia
FLP.top.glm.pred <- predict(FLP.top.glm, NH_FLP, type="response")


#Random forests for outcome of final Living Place with top variables.
FLP.top.rf <- randomForest(FLP ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=NH_FLP, ntree=100, importance=TRUE)
FLP.top.rf
imp.FLP <- FLP.top.rf$importance
head(sort(imp.FLP[,4], decreasing=TRUE), n=10)
FLP.top.rf.pred <- predict(FLP.rf, NH_FLP, type="prob")

#Model evaluation it K-fold Cross validation
N = nrow(NH_FLP)
K = 10
set.seed(1234)
s = sample(1:K, size=N, replace=T)
FLP.top.pred_outputs.glm <- vector(mode="numeric", length=N)
FLP.top.pred_outputs.rf <- vector(mode="numeric", length=N)
FLP.top.obs_outputs <- vector(mode="numeric", length=N)
offset <- 0
for(i in 1:K){
    train <- filter(NH_FLP, s != i)
    test <- filter(NH_FLP, s == i)
    FLP.top.obs_outputs[1:length(s[s==i]) + offset] <- test$FLP
   
    #GLM train/test
    glm <- glm(FLP ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=train, family=binomial(logit))
    glm.pred.curr <- predict(glm, test, type="response")
    FLP.top.pred_outputs.glm[1:length(s[s==i]) + offset] <- glm.pred.curr
    
    #RF train/test
    rf <- randomForest(FLP ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=train, ntree=100)
    rf.pred.curr <- predict(rf, newdata=test, type="prob") 
    FLP.top.pred_outputs.rf[1:length(s[s==i]) + offset] <- rf.pred.curr[,2]

    offset <- offset + length(s[s==i])
}


#GLM of Final Functional Outcome with top variables
FFO.top.glm <-glm(FFO ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data= NH_FFO, family = binomial(logit))
summary(FFO.top.glm) #Hosp length of stay + Stroke type (Total anterior circulation)
FFO.top.glm.pred <- predict(FFO.top.glm, NH_FFO, type="response")

#Random forests for outcome of Final functional outcome with top variables.
FFO.top.rf <- randomForest(FFO ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=NH_FFO, ntree=100, importance=TRUE)
FFO.top.rf
imp.FFO <- FFO.top.rf$importance
head(sort(imp.FFO[,4], decreasing=TRUE), n=10)
FFO.top.rf.pred <- predict(FFO.rf, NH_FFO, type="prob")

#Model evaluation it K-fold Cross validation
N = nrow(NH_FFO)
K = 10
set.seed(1234)
s = sample(1:K, size=N, replace=T)
FFO.top.pred_outputs.glm <- vector(mode="numeric", length=N)
FFO.top.pred_outputs.rf <- vector(mode="numeric", length=N)
FFO.top.obs_outputs <- vector(mode="numeric", length=N)
offset <- 0
for(i in 1:K){
    train <- filter(NH_FFO, s != i)
    test <- filter(NH_FFO, s == i)
    FFO.top.obs_outputs[1:length(s[s==i]) + offset] <- test$FFO
   
    #GLM train/test
    glm <- glm(FFO ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=train, family=binomial(logit))
    glm.pred.curr <- predict(glm, test, type="response")
    FFO.top.pred_outputs.glm[1:length(s[s==i]) + offset] <- glm.pred.curr
    
    #RF train/test
    rf <- randomForest(FFO ~ AGE + HospLOS + RSBP + StrokeType + RCONSC + DefDysphagia + gender, data=train, ntree=100)
    rf.pred.curr <- predict(rf, newdata=test, type="prob") 
    FFO.top.pred_outputs.rf[1:length(s[s==i]) + offset] <- rf.pred.curr[,2]

    offset <- offset + length(s[s==i])
}

```
### Results
Of patients from the IST trial who were discharged to a nursing home, we found 386 survived 6 months, 69 were not dependent on others for care and 186 improved their level of living and no longer lived in a nursing home (Table 1). We found age differed between between groups in all 3 outcomes in bivariate analyses (Figure 1a, 1b, 1c).  We found stroke type differed between all outcomes with Total Anterior Stroke having worse outcomes (Figure 2a, 2b, 2c).  We found males were more likely to have improved outcomes in bivariate analyses (Figure 3a, 3b, 3c).

Our Random Forest model using all independent variables gave 4 variables, which consistently had the highest importance for each outcome: Age, length of hospital stay, systolic blood pressure at randomization and stroke type. Other variables that were of high importance were consciousness at time of randomization, dysphagia symptoms and sex. These 7 variables were used in the final models. 

####Model Evaulation:

#####6 month Survival
All models predicting 6 month survival had a moderate to fair performance (Figure 4a,4b)(Table 2). The Logistic Regression model using top variables performed slightly better than other models with an AUC of 0.67. There is evidence for overfitting in both models of random forests as shown by the large difference between the AUC of the training versus testing data. Age, Total Anterior Stroke, No/Unknown Dysphagia were all significantly associated with increase chance of dying, while increased hospital length of stay and full consciousness at time of randomization where significantly associated with surviving at 6 months. Table 3 shows the p values for these variables. 

#####Final Living Place
All models predicting 6 month survival had moderate performance (Figure 5a,5b)(Table 4). The Logistic Regression model using top variables again performed slightly better than other models with an AUC of 0.7135. There is evidence for overfitting in both models of random forests as shown by the large difference between the AUC of the training versus testing data. Age, hospital length of stay, total anterior stroke, no/unknown dysphagia were all significantly associated with remaining in a nursing home in the logistic regression model (Table 5).

#####Final Functional OUtcome
All models predicting 6 month survival had a moderate to fair performance (Figure 6a,6b)(Table 6). The Logistic Regression model using top variables again performed slightly better than other models with an AUC of 0.71. There is evidence for overfitting in both models of random forests as shown by the large difference between the AUC of the training versus testing data. Increased hospital length of stay and total anterior stroke were both significantly associated with being in a dependent state in the logistic regression model (Table 7).

###Conclusions
For ischemic stroke patients who are discharged to a nursing home after a stroke were able to develop prediction models for 6 month outcomes of survival, final living place and functional outcome.  For all outcomes the logistic regression models using top variables slightly outperformed the other models.  Considering the multitude of factors the subjects are exposed to between hospital discharge and 6 month outcomes, it would be very difficult to have a highly predictive model using only in-hospital variables.  We therefor are satisfied with the moderate performance of our models. Future directions would be to apply and test this model on future stroke populations that were not able to receive thrombectomy or thrombolysis. 

```{r eval=TRUE}
library(pROC)
library(dplyr)

#Descriptive Statistics

#First make table of all outcomes and how many subjects are in each
Count_alive6m <- Count_alive6m %>%
  rename(Outcome=alive6m) %>%
  select(Outcome, n)
Count_FLP <- Count_FLP %>%
  rename(Outcome=FLP) %>%
  select(Outcome, n)
Count_FFO <- Count_FFO %>%
  rename(Outcome=FFO) %>%
  select(Outcome, n)

Table1 <- data.frame(rbind(Count_alive6m, Count_FFO, Count_FLP, deparse.level = 1, make.row.names = TRUE, stringsAsFactors = default.stringsAsFactors()))
Table1

#Figure 1: BoxPlots of Age vs outcomes
#Figure 1a.Box plot of Age vs 6 m survival
ggplot(data=NH_clean, aes(x=factor(alive6m), y=AGE)) +
  geom_boxplot() +
  labs(title="Figure 1a. Age vs 6 Month survival") +
  labs(y="Age (years)", x=("6 month survival"))
#Figure 1b. Box plot of Age vs Final Living Place
ggplot(data=NH_FLP, aes(x=factor(FLP), y=AGE)) +
  geom_boxplot() +
  labs(title="Figure 1b. Age and Final Living Place") +
  labs(y="Age (years)", x=("Final Living Place"))
#Figure 1c. Box plot of Age vs Final Functional Outcome
ggplot(data=NH_FFO, aes(x=factor(FFO), y=AGE)) +
  geom_boxplot() +
  labs(title="Figure 1c. Age vs 6 Final Funcitonal Outcome") +
  labs(y="Age (years)", x=("Final Functional Outcome"))

#Figure 2: Stroke Type vs Outcomes
#Figure 2a. Box plot of Stroke type and survival
ggplot(data=NH_alive6m, aes(x=StrokeType, fill=alive6m)) +
  geom_bar(position="fill") +
  labs(title="Figure 2a. Stroke Type vs 6 Month survival")
#Figure 2b. Box plot of Stroke type and Final Living Place
ggplot(data=NH_FLP, aes(x=StrokeType, fill=FLP)) +
  geom_bar(position="fill") +
  labs(title="Figure 2b. Stroke Type vs Final Living Place")
#Figure 2c. Box plot of Stroke Type and Final Functional Outcome
ggplot(data=NH_FFO, aes(x=StrokeType, fill=FFO)) +
  geom_bar(position="fill") +
  labs(title="Figure 2c. Stroke Type vs Final Functional Outcome")

#Figure 3. Gender vs Outcomes
#Figure 3a. Box plot of gender and survival
ggplot(data=NH_alive6m, aes(x=alive6m, fill=gender)) +
  geom_bar(position="fill") +
  labs(title="Figure 3a. Gender and 6 Month survival") +
  labs(x="6 month survival")
#Figure 3b. Box plot of gender and Final Living Place
ggplot(data=NH_FLP, aes(x=FLP, fill=gender)) +
  geom_bar(position="fill") +
   labs(title="Figure 3b. Gender and Final Living Place") +
  labs(x="Final Living Place")
#Figure 3c. Box plot of gender and Final Functional Outcome
ggplot(data=NH_FFO, aes(x=FFO, fill=gender)) +
  geom_bar(position="fill") + 
   labs(title="Figure 3c. Gender and Final Functional Outcome") +
  labs(x="Final Functional Outcome")

#Figure 4a. ROC curves of RF with all Variables for 6 Month Survival
pROC::plot.roc(NH_alive6m$alive6m, rf.pred.alive6m, ci=TRUE, main="Figure 4a. ROC of RF with all Variables \n for 6 Month Survival", col="darkgreen", add=FALSE)
pROC::plot.roc(alive6m.obs_outputs, alive6m.pred_outputs.rf, ci=TRUE, col="red", add=TRUE) 
legend("right", legend=c("RF Training", "RF Cross-Validation"), col=c("darkgreen", "red"), lwd=2)

#Figure 4b. ROC curves of alive 6 m with top variables
plot.roc(NH_alive6m$alive6m, alive6m.top.glm.pred, ci+TRUE, main="Figure 4b. ROC curves of Top Variable Prediction Models \n for 6 Month Survival")
plot.roc(alive6m.top.obs_outputs, alive6m.top.pred_outputs.glm, ci+TRUE, col="darkblue", add = TRUE)
plot.roc(NH_alive6m$alive6m, alive6m.top.rf.pred[,2], ci=TRUE, col="darkgreen", add=TRUE)
plot.roc(alive6m.top.obs_outputs, alive6m.top.pred_outputs.rf, ci=TRUE, col="red", add=TRUE)
legend("right", legend=c("GLM Training", "GLM Cross-Validation", "RF Training", "RF Cross-Validation"), col=c("black", "darkblue", "darkgreen", "red"), lwd=2)

#Table 2: AUC for Models predicting 6 month survival
Model_6month_survival <- list("RF with all variables", "RF with Top Predictors", "GLM with Top Predictors")
AUC <- list((auc(alive6m.obs_outputs, alive6m.pred_outputs.rf)), (auc(alive6m.top.obs_outputs, alive6m.top.pred_outputs.rf)), (auc(alive6m.top.obs_outputs, alive6m.top.pred_outputs.glm)))
Table2 <- cbind(Model_6month_survival, AUC)
Table2 #Table2: Comparison of AUC for each Prediction Model for 6 Month Survival

#Table 3: Variables with significant p values for 6 month survival
Variable <- list("Age", "Hospital Length of Stay", "Total Anterior Stroke", "No/Unknown Dysphagia", "Full Consciousness at randomization")
PValue <- list(3e-05, 0.022, 0.007, 0.024, 0.047)
Table3 <- cbind(Variable, PValue)
Table3 #Table 3: Variables with significant p values for 6 month survival

#Figure 5a. ROC curves of RF using all variables for Final Living Place
plot.roc(NH_FLP$FLP, FLP.rf.pred[,2], ci=TRUE, main="Figure 5a. ROC curves of RF using all variables \n for Final Living Place", col="darkgreen", add=FALSE)
plot.roc(FLP.obs_outputs, FLP.pred_outputs.rf, ci=TRUE, col="red", add=TRUE) 
legend("right", legend=c("RF Training", "RF Cross-Validation"), col=c("darkgreen", "red"), lwd=2)

#Figure 5b. ROC curves of RF and GLM predictors of Final Living Place (FLP) with top variables
plot.roc(NH_FLP$FLP, FLP.top.glm.pred, ci=TRUE, main="Figure 5b. ROC curves of Top Variable Prediction Models \n for Final Living Place")
plot.roc(FLP.top.obs_outputs, FLP.top.pred_outputs.glm, ci=TRUE, col="darkblue", add=TRUE)
plot.roc(NH_FLP$FLP, FLP.top.rf.pred[,2], ci=TRUE, col="darkgreen", add=TRUE)
plot.roc(FLP.top.obs_outputs, FLP.top.pred_outputs.rf, ci=TRUE, col="red", add=TRUE) 
legend("right", legend=c("GLM Training", "GLM Cross-Validation", "RF Training", "RF Cross-Validation"), col=c("black", "darkblue", "darkgreen", "red"), lwd=2)

#Table 4: AUC for Final Living Place Models
Model_FLP <- list("RF with all variables", "RF with Top Predictors", "GLM with Top Predictors")
AUC <- list((auc(FLP.obs_outputs, FLP.pred_outputs.rf)), (auc(FLP.top.obs_outputs, FLP.top.pred_outputs.rf)), (auc(FLP.top.obs_outputs, FLP.top.pred_outputs.glm)))
Table4 <- cbind(Model_FLP, AUC)
Table4 #Table4: Comparison of AUC for each Prediction Model for Final Living Place

#Table 5: Variables with significant p values for Final Living Place
FLP_Sig_Variable <- list("Age", "Hospital Length of Stay", "Total Anterior Stroke", "No/Unknown Dysphagia")
PValue <- list(3e-08, 0.046, 0.0004, 0.03)
Table5 <- cbind(FLP_Sig_Variable, PValue)
Table5 #Table 5: Variables with significant p values for Final Living Place

#Figure 6a. ROC curves of RF using all variables for Final Functional Outcome
plot.roc(NH_FFO$FFO, FFO.rf.pred[,2], ci=TRUE, main="Figure 6a. ROC curves of RF using all variables \n for Final Functional Outcome", col="darkgreen", add=FALSE)
plot.roc(FFO.obs_outputs, FFO.pred_outputs.rf, ci=TRUE, col="red", add=TRUE) 
legend("right", legend=c("RF Training", "RF Cross-Validation"), col=c("darkgreen", "red"), lwd=2)

#Figure 6b. RF and GLM Prediction Model ROC curves for Final Functional Outcome
plot.roc(NH_FFO$FFO, FFO.top.glm.pred, ci=TRUE, main="Figure 6b. ROC curves of Top Variable Prediction Models \n for Final Functional Outcome")
plot.roc(FFO.top.obs_outputs, FFO.top.pred_outputs.glm, ci=TRUE, col="darkblue", add=TRUE)
plot.roc(NH_FFO$FFO, FFO.top.rf.pred[,2], ci=TRUE, col="darkgreen", add=TRUE)
plot.roc(FFO.top.obs_outputs, FFO.top.pred_outputs.rf, ci=TRUE, col="red", add=TRUE) 
legend("right", legend=c("GLM Training", "GLM Cross-Validation", "RF Training", "RF Cross-Validation"), col=c("black", "darkblue", "darkgreen", "red"), lwd=1)


#Table 6: AUC for Final Functional Outcome (FFO) Models
Model_FFO <- list("RF with all variables", "RF with Top Predictors", "GLM with Top Predictors")
AUC <- list((auc(FFO.obs_outputs, FFO.pred_outputs.rf)), (auc(FFO.top.obs_outputs, FFO.top.pred_outputs.rf)), (auc(FFO.top.obs_outputs, FFO.top.pred_outputs.glm)))
Table6 <- cbind(Model_FFO, AUC)
Table6 #Table6: Comparison of AUC for each Prediction Model for Final Functional Outcome

#Table 7: Variables with significant p values for Final Functional Outcome
FFO_Sig_Variable <- list("Hospital Length of Stay", "Total Anterior Stroke")
PValue <- list(0.011, 0.004)
Table7 <- cbind(FFO_Sig_Variable, PValue)
Table7 #Table 7: Variables with significant p values for Final Functional Outcome
```
