---
title: "Due to the change of topic, my project progress has fallen behind. I will update my templete by Dec. 7"
author: "To peer reviewers: please review my final project on Dec. 8 or Dec. 9"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but keep the headers.

### Overview
###Due to the change of topic, my project progress has fallen behind. I will update my templete by Dec. 7
###To peer reviewers: please review my final project on Dec. 8 or Dec. 9

In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

### Introduction 
###Due to the change of topic, my project progress has fallen behind. I will update my templete by Dec. 7
###To peer reviewers: please review my final project on Dec. 8 or Dec. 9
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.
In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

Personalized genetic testing and anlysis services have become a hot topic over the past five years. Major companies such as 23andMe and Wegene generate reports though collecting and analyzing DNA samples from customers. These reports usually contain ancestry information and predictions of genetically related diseases. While many genetically related diseases, such as type 2 diabetes, are phenotypically developed under the impact from both genes and environment, disease predictions generated from these companies only take genetic information into account. In another word, these predictions cannot give a person's real time risk prediction of getting type 2 diabetes under certain physical conditions.

Different from current prediction models, my model will take into account data other than DNA, such as age, sex, race, BMI, and waistline (with the help from Dr. John Holmes). By doing this, my model can give a "real time" prediction for a person's probability of having/getting the type 2 diabetes. Theoretically, this model can help people to have a specific caution of how they should control their BMI and/or waistline given their genotype and other predictors. 
  
  
### Methods
###Due to the change of topic, my project progress has fallen behind. I will update my templete by Dec. 7
###To peer reviewers: please review my final project on Dec. 8 or Dec. 9
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

I have just got my access to PennOmics today. Therefore, I'm unable to make any change to the methods section. As my presentation is relatively late (Dec. 12), I still have some time to finish my methods and results section before presentation. But basically, I will write a function to calculate the risk of getting DM2 with/with out an SNP and run some machine learning functions to build up the prediction model.
```{r, eval = FALSE}

```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
