---
title: "Predicting Acute Kidney Injury in Hospitalized Patients from EHR Data"
author: "Haedi Thelen"
output: 
  html_document:
    theme: paper 
    highlight: tango
---

***
#### Overview

Acute kidney injury (AKI) is one of the most common complications seen in hospitalized patients. The ability to predict which patients are at highest risk for AKI would allow clinicians to intervene and prevent poor patient outcomes. Despite many attempts, however, AKI predicting models fail to perform in way that motivates their use in clinical decision support. Understanding which risk factors contribute most to AKI prediction, and how those risk factors vary based on ICU status and admission type, can inform future studies seeking to predict AKI. This study aims to quantify the predictive power of various risk factors for AKI based on admission type and patient location in the hospital.


### Introduction 
A common condition in hospitalized patients, rates of AKI are estimated to be as high as 7% in hospitalized patients and 30% in ICU patients. (Goyal et al.). AKI is a sudden loss of renal function due to non-renal causes, such as dehydration, reduced blood flow to the kidney, nephrotoxic medications, or sepsis. AKI is  usually reversible, but if not addressed can progress to permanent kidney damage and multi-organ failure. Diagnosing AKI early can prevent poor patient outcomes (Goyal et al.). Many groups have attempted to build decision making tools to assist clinicians in predicting AKIs early in the disease stage allowing rapid intervention and preventing sequelae. However, the majority of AKI prediction tools, including many using state of the art machine learning methods on large, diverse datasets, struggle to improve on a trained clinicians ability to identify AKIs (De Vlieger et al). In particular, the black box issue of machine learning algorithms limits the clinicians ability to trust the models, understand what is happening to the patient, and base decisions on model outputs. One way to enhance model transparency, while improving performance is to understand which risk factors are important to model decision making. Understanding which risk factors contribute most to AKI prediction, and how those risk factors vary based on ICU status and admission type can inform future work seeking to predict AKI. 

Prior research has built a data set of patients on combination anti-hypertensive with either NSAID or oxycodone to assess AKI risk (Miano et al). This dataset can be re-analyzed to identify which of the studied risk factors contributes most to prediction of AKI. Patients on combination antihypertensives and either NAISDs or oxycodone represent a common subset of hospitalized patients, thus results on this population can be expected to generalize well to all patients. This study seeks to answer the question: in patients receiving treatment for hypertension and pain, do the most predictive risk factors for AKI vary by ICU vs non-ICU and admission type (medical vs surgery patients)?

AKI prediction is a prominent interdisciplinary space. Clinicians, informaticians, computer scientists, and epidemiologists frequently work together and publish in this space. The work typically combines new machine learning approaches with classic clinical definitions of kidney injury and risk factors. No discipline alone has enough expertise to address the problem, making interdisciplinary collaboration key to success. Conversations with Todd Miano (PharmD, PhD research at University of Pennsylvania's Department of Biostatistics, Epidemiology, and Bioinformatics (DBEI) and  Critical Care Pharmacist) highlighted the difficulty in defining AKI though KDIGO's multipart definition based on the non-specific marker, serum creatinine. Informatician, John Holmes (Professor of Medical Bioinformatics in Epidemiology also at the University of Pennsylvania DBEI) emphasized the fact that the various prediction models incorporate similar, large sets of risk factors. He suggested that different risk factors my play varying roles for patients in different clinical settings, such as patients in the ICU or post-operative patients. These discussions motivated the proposed study of quantifying predictive power of various risk factors, but splitting the analysis based on admission type and patient location. 


A git repo for this project is available [here](https://github.com/HaediThelen/BMIN503_Final_Project). 


### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

#### Data
Todd Miano provided the data used in this project, and was previously described in Miano et al. Briefly, data was collected on 27,741 adult patients with greater than 24-hours exposure to one both an anti-hypertensive agent and an analgesic. Ant-hypertensives of interest included either a renin-angiotensin inhibitor (RAS-I) or the calcium channel blocker amlodipine. Analgesics of interest included wither an NSAID or oxycodone. Data was collected between 2004 and 2017 from patients admitted to University of Pennsylvania Health System hospitals using the Penn Data Store warehouse for electronic health records in the health system. Patients were excluded if they had contraindications to NSAIDs, including unresolved AKI within 2 weeks before entry, baseline serum creatinine >2mg/dl (an indicator of renal injury), end stage renal disease, renal replacement therapy, platelet count $< 100*10^{11}$ (risk of bleed), pregnancy (a contraindication to RAS-I treatment), lack of baseline or follow-up serum creatinine, and history of solid organ transplant. Information was also collected variables associated with AKI, including cardiovascular conditions, diabetes, and cancer. Laboratory values and presence of nephrotoxic medications at entry are also included. In this data set, AKI's and AKI severity were labeled according to Kidney Disease Improving Global Outcomes (KDIGO) criteria. 

In this data set, comorbid conditions and laboratory values were assessed at the point in time where the patient qualified for the study (entry). That is when the patient had greater than 24-hours exposure to the drug-combinations of interest. Thus, when using this data to predict AKI, we necessarily are predicting AKI at the same point in time. Therefore, the prediction methods evaluated in this study attempt to predict AKIs using information available to clinicians up to 24 hours after initiating combination therapy of anti-hypertensives and opioids. The goal of this research is to identify which predictors are the most helpful in predicting AKIs in this setting, and how they differ, if at all, between the medical floor and the ICU. 

To do this, we first will import and examine the dataset. Then separate it into 1) the complete dataset, 2) ICU cohort, and 3) the medical floor cohort. (WHAT ABOUT TRAIN AND TEST STES and K FOLDS)

#### Linear Regression Analysis
Linear regression models will be created for each data sub-set. K-fold cross validation will be used to estimate prediction error. Key predictors will be identified based on statistical significance. 

#### Decision Tree Analysis
Next, decision tree models will be trained for each model. Key predictors will be identified based on GINA importance scores. 


#### Import and Load Packages
```{r, eval = F}
# Import packages, if needed
install.packages("dplyr")
install.packages("gtsummary")
install.packages("modelsummary")
```

```{r, eval = T}
# Load packages
library(dplyr)
library(gtsummary)
library(modelsummary)
library(ggplot2)
library(randomForest)
library(tibble)
library(tidyr)
library(stringr)
library(pROC)
```

#### Read in the Data
Data are found at [find data storage place](). 
```{r, eval = T}
#Load the data
data <- read.csv("/Users/haedi/Repos/BMIN503_Final_Project/BMIN503_Final_Project/aki_ml_dataset.csv", header = TRUE)

# Drop columns not needed:
# ddiGroup was needed to classify exposure in the prior study, but not needed
# here, nsaidType is missing data for 81% and so was left out. 
# aki_stage, rrt, and mortHosp30 are outcome variables needed in the prior study
# but not here; we are only interested in binary AKI as the outcome variable. 

# Table 1, sorted by ICU stay, variables renamed for readability
data %>%
  select(!c(pid, ddiGroup, nsaidType, aki_stage, rrt, mortHosp30)) %>%
  rename(Analgesic = pain, Anithypertensive = bp, AKI = aki, Age = age, 
         Sex = sex, Race = race, BMI = bmi,  "Admission Type" = admType, 
         "Prior Length of Stay" = priorLos, "Perioperative Day" = periOp, 
         Ventilator = vent, "Congestive Heart Failure" = chf, "Chronic Pulmonary Disease" = cpd, 
         HIV = hiv, "Liver Disease" = liver, "periveral Vascular Disease" = pvd,
         "Ceribrovascular Disease" = cva, "Myocardial Infarction" = mif, 
         "Valvular Disease" = valve, Hypertension = htn, "Cardiac Arrhythmias" = arry, 
         "Pulmonary Circulation Disorder" = pCirc, Obesity = obese, "Weight Loss" = wtLoss,
         "Fluid and Electrolyte Disorder"= fluid, "Chronic Kidney Disease" = ckd,
         "Atrial Fibrillation" = afib, "Obstructive Sleep Apnea" = osa, 
         "Diabetes Mellitus" = dm, Cancer = cancer, "Prior AKI" = prior_aki, Vancomycin = vanco,
         "Bactrim" = bactrim, "Vasopressors" = pressor, "Other Nephrotoxins" = ntxOther, 
         "Nephrotoxic Antibiotics" = abxNTX, "Diruetics" = diuretic, 
         "Broad Spectrum Antibiotics" = gramNegBroad, "Narrow Spectrum Antibiotics" = gramNegNarrow,
         "WBC x10^8cells/L" = wbc, "Hmoglobin g/dl" = hgb, "Platelets x10^11/L" = platelets, 
         "Sodium, mEq/L" = sodium, "Potassium, mEq/L" = potassium, "Chloride mEq/L" = chloride,
         "Serum Creatinine at entry" = creatinine) %>% #  THIS IS AT ENTRY, RIGHT? NOT WHAT WAS USED TO CALCULATE AKI??
   mutate(Cancer = factor(Cancer, level = c(0,1,2), labels = c("None", "Non-metastatic", "Metastatic"))) %>%
  mutate(icu = factor(icu, levels = c(0,1), labels = c("non-ICU", "ICU"))) %>%
  tbl_summary(by = icu)

# Mutate binary variables to factors, keep only variables we will use
data.c.f <- data %>%
  select(!c(pid, ddiGroup, nsaidType, aki_stage, rrt, mortHosp30,)) %>%
  mutate(across('race', str_replace, 'Other / Unk', 'Other')) %>%
  mutate(aki = factor(aki, level = c(0,1), labels = c("No AKI", "AKI"))) %>%
  mutate(cancer = factor(cancer, level = c(0,1,2), 
                         labels = c("no_cancer", "non-metastatic", "metastatic"))) %>%
   mutate(icu = factor(icu, levels = c(0,1), labels = c("non-ICU", "ICU"))) %>%
  mutate_at(c("pain", "bp", "sex", "race", "admType", "periOp", "vent", "chf", 
              "cpd", "hiv","liver", "pvd", "cva", "mif", "valve", "htn", "arry",
              "pCirc", "obese", "wtLoss", "fluid", "ckd", "afib", "osa", "dm", 
              "prior_aki", "vanco", "bactrim", "pressor","ntxOther", "abxNTX",
              "diuretic","gramNegBroad", "gramNegNarrow"), factor) %>%
  relocate(where(is.factor))%>%
  relocate(aki)

#####  I think this is unecessary: 
data.f <- data.c.f %>% # create dataframe of just factors, for changing to design matrix
  select_if(is.factor)

x.factors <-model.matrix( aki ~ ., data = data.f) [ , -1]
head(x.factors)

#reassemble dataframe, this time with dummy variables and numeric variables
data.c <- cbind(data.c.f[, c("aki", "age", "bmi", "priorLos", "wbc", "hgb", "platelets", "sodium", "potassium", "chloride", "creatinine", "indexGFR")], x.factors)
data.c <- data.c %>%
  rename(periOpPODone = `periOpPOD one`, periOpPODtwo = `periOpPOD two`, periOpPODthree = `periOpPOD three`, periOpPODzero = `periOpPOD zero`, dmDMnonComp = `dmDM-Non-comp`, cancerNonMetastatic = `cancernon-metastatic`, prior_akiYes_resolved = `prior_akiYes-resolved`) 
###


# Break up the datasets to ICU and Non-ICU cohorts 

#### REDO with matrix stuff??
icu <- data.c.f %>%
  filter(icu=="ICU") %>%
  select(!c(icu))

nonicu <- data.c.f%>%
  filter(icu=="non-ICU")%>%
  select(!c(icu))
```
### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

#### Visualizing Data
From Table 1, we can see that AKI rates are higher in the ICU. Next, we visualize the relationship between ICU stay and AKI. 
```{r, eval = T}
ggplot(data = data.complete.factors, aes(x=icu, fill = aki)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of study patients with AKI by location", x = "ICU Status", y ="Proporiton" )
```

Using a Chi-square test, we can check if there is a statistically significant difference in AKI rates.
```{r, eva; = T}
# Chi-Square test
chisq.test(table(data.complete$aki, data.complete$icu))
```
At the $\alpha = 0.5$ level, we can conclude that in our study population, the AKI rates are different between non-ICU and ICU settings. This is helpful contextual information for our data, though, it is not part of the main analysis. 

#### Logistic Regression and Random Forest Models
Next we create a logistic regression model for all three data subsets (complete, ICU, and non-ICU), then compare the predictors. 

##### Analysis Regression for Complete Dataset

```{r, eval=T}
N = nrow(data.c.f)
K = 10
set.seed(1234)
s = sample(1:K, size = N, replace = T)
pred.outputs.glm <- vector(mode = "numeric", length = N)
pred.outputs.rf <- vector(mode = "numeric", length = N)
obs.outputs <- vector(mode = "numeric", length = N)
offset <- 0
for(i in 1:K){
	train <-filter(data.c.f, s != i)
	test <- filter(data.c.f, s == i)
	obs.outputs[1:length(s[s == i]) + offset] <- test$aki
	
	#GLM train/test
	glm.aki <- glm(aki ~ ., data = train, family = binomial(logit))
  glm.predictions <- predict(glm.aki, test, type = "response")
  pred.outputs.glm[1:length(s[s == i]) + offset] <- glm.predictions
  
	#RF train/test
	rf <- randomForest(aki ~ ., data = train, ntree = 100)
	rf.pred.curr <- predict(rf, newdata = test, type = "prob") 
	pred.outputs.rf[1:length(s[s == i]) + offset] <- rf.pred.curr[ , 2]
	
	offset <- offset + length(s[s == i])
}
```


```{r, eval = T}

#AUC
roc(obs.outputs, pred.outputs.glm) # glm with cross validation
roc(obs.outputs, pred.outputs.rf, ci = TRUE) # rf with cross validation

#ROC Curves
plot.roc(obs.outputs, pred.outputs.glm, col = "darkred") # Glm, k fold cross validated
plot.roc(obs.outputs, pred.outputs.rf, ci = TRUE, col = "navy", add = TRUE) # RF, K-fold cross validated
legend("bottomright", legend = c("glm cross-validation", "rf cross-validation"), col = c("darkred", "navy"), lwd = 1)

```
Next, we need to identify the top predictors

##### Mobel Building Regression for ICU Dataset

```{r, eval=T}
N = nrow(icu)
K = 10
set.seed(1234)
s = sample(1:K, size = N, replace = T)
pred.outputs.glm.icu <- vector(mode = "numeric", length = N)
pred.outputs.rf.icu <- vector(mode = "numeric", length = N)
obs.outputs.icu <- vector(mode = "numeric", length = N)
offset <- 0
for(i in 1:K){
	train <-filter(icu, s != i)
	test <- filter(icu, s == i)
	obs.outputs.icu[1:length(s[s == i]) + offset] <- test$aki
	
	#GLM train/test
	glm.aki.icu <- glm(aki ~ ., data = train, family = binomial(logit))
  glm.predictions.icu <- predict(glm.aki.icu, test, type = "response")
  pred.outputs.glm.icu[1:length(s[s == i]) + offset] <- glm.predictions.icu
  
	#RF train/test
	rf.icu <- randomForest(aki ~ ., data = train, ntree = 100)
	rf.pred.curr.icu <- predict(rf.icu, newdata = test, type = "prob") 
	pred.outputs.rf.icu[1:length(s[s == i]) + offset] <- rf.pred.curr.icu[ , 2]
	
	offset <- offset + length(s[s == i])
}
```


```{r, eval = T}

#AUC
roc(obs.outputs.icu, pred.outputs.glm.icu) # glm with cross validation
roc(obs.outputs.icu, pred.outputs.rf.icu, ci = TRUE) # rf with cross validation

#ROC Curves
plot.roc(obs.outputs.icu, pred.outputs.glm.icu, col = "darkred") # Glm, k fold cross validated
plot.roc(obs.outputs.icu, pred.outputs.rf.icu, ci = TRUE, col = "navy", add = TRUE) # RF, K-fold cross validated
legend("bottomright", legend = c("glm cross-validation", "rf cross-validation"), col = c("darkred", "navy"), lwd = 1)

```
##### Mobel Building Regression for ICU Dataset

```{r, eval=T}
N = nrow(nonicu)
K = 10
set.seed(1234)
s = sample(1:K, size = N, replace = T)
pred.outputs.glm.nonicu <- vector(mode = "numeric", length = N)
pred.outputs.rf.nonicu <- vector(mode = "numeric", length = N)
obs.outputs.nonicu <- vector(mode = "numeric", length = N)
offset <- 0
for(i in 1:K){
	train <-filter(nonicu, s != i)
	test <- filter(nonicu, s == i)
	obs.outputs.nonicu[1:length(s[s == i]) + offset] <- test$aki
	
	#GLM train/test
	glm.aki.nonicu <- glm(aki ~ ., data = train, family = binomial(logit))
  glm.predictions.nonicu <- predict(glm.aki.nonicu, test, type = "response")
  pred.outputs.glm.nonicu[1:length(s[s == i]) + offset] <- glm.predictions.nonicu
  
	#RF train/test
	rf.nonicu <- randomForest(aki ~ ., data = train, ntree = 100)
	rf.pred.curr.nonicu <- predict(rf.nonicu, newdata = test, type = "prob") 
	pred.outputs.rf.nonicu[1:length(s[s == i]) + offset] <- rf.pred.curr.nonicu[ , 2]
	
	offset <- offset + length(s[s == i])
}
```


```{r, eval = T}

#AUC
roc(obs.outputs.nonicu, pred.outputs.glm.nonicu) # glm with cross validation
roc(obs.outputs.nonicu, pred.outputs.rf.nonicu, ci = TRUE) # rf with cross validation

#ROC Curves
plot.roc(obs.outputs.nonicu, pred.outputs.glm.nonicu, col = "darkred") # Glm, k fold cross validated
plot.roc(obs.outputs.nonicu, pred.outputs.rf.nonicu, ci = TRUE, col = "navy", add = TRUE) # RF, K-fold cross validated
legend("bottomright", legend = c("glm cross-validation", "rf cross-validation"), col = c("darkred", "navy"), lwd = 1)

```
Performance decreased when the data was split between ICU and Non ICU, vs when data was combined. 
But were the top predictors different? 


# References
1) Goyal A, Daneshpajouhnejad P, Hashmi MF, et al. Acute Kidney Injury. [Updated 2022 Aug 18]. In: StatPearls [Internet]. Treasure Island (FL): StatPearls Publishing; 2022 Jan-. Available from: https://www.ncbi.nlm.nih.gov/books/NBK441896/.

2) De Vlieger, Greeta; Kashani, Kianoushb; Meyfroidt, Geerta. Artificial intelligence to guide management of acute kidney injury in the ICU: a narrative review. Current Opinion in Critical Care: December 2020 - Volume 26 - Issue 6 - p 563-573 doi: 10.1097/MCC.0000000000000775.

3) Miano, Todd A., et al. "Effect of renin-angiotensin system inhibitors on the comparative nephrotoxicity of NSAIDs and opioids during hospitalization." Kidney360 1.7 (2020): 604.

