---
title: "BMIN503/EPID600 Project Template"
author: "Your Name"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
library(dplyr)
library(ggplot2)
library(tibble)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but keep the headers.

### Overview
In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.


### Introduction 
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.

In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.


### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 
```{r}
#functions to help get specific columns from different sets

#selects columns for demographics
demographics <- function(info) {
  info <- select(info, 1, 4:5, 9:12, 17:19, 21, 41:47, 227:508)
  return(info)
}

#selects columns about tech
pttech <- function(info) {
  info <- select(info, 1, 48:124, 509:609)
  return(info)
}

#selects columns for pedsql
pedsql <- function(info) {
  info <- select(info, 1, 125:149, 610:634)
  return(info)
}

#selects columns for physical activity and nutrition
physnut <- function(info) {
  info <- select(info, 1, 150:228, 635:717)
  return(info)
}

#selects columns for weight questionnaire
weight <- function(info) {
  info <- select(info, 1, 229:247)
  return(info)
}

#selects columns for asthma questionnaire
asthma <- function(info) {
  info <- select(info, 1, 248:426, 718:846)
  return(info)
}

#selects columns for usherwood questionnaire
asthma <- function(info) {
  info <- select(info, 1, 847:885)
  return(info)
}
```


````{r}
#Get the data - with some formatting
#Clear existing data and graphics
#rm(list=ls())
graphics.off()
#Load Hmisc library
library(Hmisc)
#Read Data
data=read.csv('ChronicConditionSpec_DATA_2019-10-02_1538.csv')

#Set factors
source('factors.r')

#Set Levels
source('levels.r')

#remove '.factor' from names
data <- data[c(1, 473, 3, 474:478, 9:10, 479:483, 16, 484, 18, 485:487, 22, 488, 24, 489:491, 28, 492, 30, 493:504, 43, 505:510, 50, 511:536, 77, 537:560, 102, 561:588, 131, 589:606, 150, 607:684, 229, 685:715, 261, 716:765, 312, 766:789, 337, 790:821, 370, 822:899, 449, 900:917, 468, 918:919, 471, 920)]
names(data) <- gsub(".factor", "", names(data), fixed = TRUE)

#Name all columns
source('labels.r')

#collect the 4 columns on language preference into 2
data <- add_column(data, "Patient's Language" = NA, .before = "eng")
data$`Patient's Language` <- case_when(
  data$eng == 1 ~ "English",
  data$esp == 1 ~ "Spanish",
  TRUE ~ NA_character_
)
data <- add_column(data, "Parent's Language" = NA, .before = "eng")
data$`Parent's Language` <- case_when(
  data$eng2 == 1 ~ "English",
  data$esp2 == 1 ~ "Spanish",
  TRUE ~ NA_character_
)
data[5:8] <- NULL
```

```{r}
#cleaning data

```

```{r}
#make dataframe for anyone completing at least 1 fu, and then their initial visit (to collect demographics)
participants <- subset(data, data$redcap_id %in% data$redcap_id[which(data$redcap_event_name=="fu_visit_1_arm_1")])
participantsFirst <- subset(participants, participants$redcap_event_name=="initial_visit_arm_1")

complete <- subset(data, data$redcap_id %in% data$redcap_id[which(data$redcap_event_name=="final_visit_arm_1")])
completeFirst <- subset(complete, complete$redcap_event_name=="initial_visit_arm_1")

#get demographic data
completeFirst <- demographics(completeFirst)
```


### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
