---
title: "Measuring the Participation in the ACCELERATE Natural History Registry"
subtitle: "BMIN503/EPID600 Final Project"
author: "Bridget Austin"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

## Overview {#sec-overview}

Castleman disease is a rare disease that is very difficult to diagnose and treat. Right now, only one FDA-approved drug exists to treat the most deadly form of CD. The best way to gain more knowledge and find other treatments for this disease is through the implementation of a rare disease registry. This project aims to characterize the first several hundred patients in the ACCELERATE natural history registry to find opportunities for further patient enrollment so that this important research can continue.

## Introduction {#sec-introduction}

Castleman disease is a rare disease characterized by lymph node enlargement, specific microscopic changes to the lymph nodes, and a broad range of symptoms and laboratory findings. Castleman disease patients can belong in one of two groups: unicentric Castleman disease (UCD) or multicentric Castleman disease (MCD). UCD is typically localized and easily treated via lymph node excision. However, MCD is a systemic condition associated with heterogeneous symptoms. MCD, being the more severe of the subtypes, is currently very poorly understood. It is thought that MCD patients have a 5-year overall survival rate between 50% and 65%. There is currently only one FDA-approved drug for the treatment of idiopathic MCD, the most aggressive type of MCD . Unfortunately, there has not been much of an effort historically to collaborate on Castleman disease research. However, more recently, the Castleman Disease Collaborative Network (CDCN) has made strides in the research due to the implementation of a biobank and a medical record registry. In 2013 the CDCN initiated a medical record registry to promote the study of Castleman disease. This registry, called ACCELERATE (Advancing Castleman Care with an Electronic Longitudinal Registry, E-Repository, and Treatment/Effectiveness research) collects real-world demographic, clinical, laboratory, and outcomes data on patients with Castleman disease. Currently, the registry has around 700 patients who have partially or fully completed the enrollment process.

Due to the nature of rare disease research, it is important to be able to collect and store patient information for as many patients as possible. That way, researchers can compare disease manifestation across age, sex, race, and more. When collecting both medical records and biospecimens in a rare disease, there is a need to collect, track, and monitor information in an systematic fashion. Especially in a research laboratory, where specimen and patient data not only be manipulated to make discoveries regarding patient treatment but may also change over time. The need to join information using biobank databases, the ACCELERATE data set, and other resources makes this an interdisciplinary problem. My main goal of this project is to further understand the population enrolled in the ACCELERATE registry and find ways to increase engagement. From meeting with my mentor, I found that one interest of our research team is to explore gaps in enrollment and errors in the enrollment process. First, I will describe enrollees in the registry to discover gaps and opportunities in patient engagement to further enrollment. To characterize the Castleman disease population in ACCELERATE, clinical and demographic characteristics will be summarized for all patients enrolled in the registry. I will use an interactive map to illustrate the locations of the patients enrolled in the US and other countries. Then I will use enrollment data to take a deep dive into engagement, by finding information like the average length of symptom onset to time of diagnosis (to tell us more about how our diagnostic criteria have affected the population), date of diagnosis on different forms (to check for patient understanding/gaps in patient engagement), date of diagnosis to date of enrollment in registry (to tell us more about patient engagement). This information can be crucial in showing us how patients interact with the registry and can help to find barriers in enrollment.

## Methods {#sec-methods}

First I am downloading relevant libraries and forms from the ACCELERATE Natural History Registry. Any information about names of patients has been removed. From here on out, patients are identified using a Global Unique Identifier (GUID).

```{r}
#libraries

library(readr)
library(gtsummary)
library(tidyverse)
library(lubridate)
library(dplyr)
library(lubridate)
library(ggplot2)

```

```{r}
#Read in registry documents. All downloaded on 11.7.23.

demographic <- read.csv("Y:/BMIN 503/Data Final Project/DemographicForm.csv")
diagnosis <- read.csv("Y:/BMIN 503/Data Final Project/DiagnosisForm.csv")
eligibility<- read.csv("Y:/BMIN 503/Data Final Project/EligibilityForm.csv")

```

```{r}
#Cleaning the demographic form for what we are interested in.

demographic <- rename(demographic,  
                 sex = sex_of_participant, 
                 dob = participant_date_of_birth_guid, 
                 race = race_of_participant, 
                 country=country_of_birth_guid,
                 ) |>
  mutate(sex = factor(sex, levels = c(1, 2, 3, 4),
                      labels = c("male", "female", "Other", "Transsexual"))) |>
  mutate(race = factor(race, levels = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17  ,18), 
                         labels = c("AmericanIndianOrAlaskanNative", "Asian Indian", "Chinese", "Filipino", "Japanese", "Korean","Vietnamese", "Other Asian", "Black or Afrian American", "Native Hawaiian", "Guamanian", "Chamorro", "Samoan", "Other Pacific Islander", "White", "Other", "Unknown", "Refuse to answer"))) 
  
```

Now that we have loaded and edited our data to be a little more user friendly, I will begin describing the population of patients in our registry.

```{r}
#Some descriptive statistics

demographic|>
  tbl_summary()

#We have 674 patients who filled out this form. 

demographic|>
  select(c(sex, country))|>
  tbl_summary()

#59% of enrollees are female and 73% of enrollees are from the USA.

#calculate current age of these patients.

now=Sys.Date()
now

demographic$dob<-as.Date(demographic$dob, format="%m/%d/%Y")

demographic<-
  mutate(demographic, today=now)

demographic<-
  mutate(demographic, age=as.numeric(today-demographic$dob)/365)

#median current age of patients is 47 years old.
```

Here, I want to look at age at diagnosis. In a rare disease with no known etiology, it is important to find if there is a trend in age at diagnosis. Here we see that the median age at diagnosis is 39 years old, which is similar to previously reported results (Zhang et al., 2018). However, as you can see in the graph below, the range is quite wide.

```{r}
#Combine diagnosis and demographic. The goal is to calculate age at diagnosis using diagnosis date from diagnosis form and age from demographic form.

#converting dates into actual date format.
diagnosis$date_at_diagnosis<-as.Date(diagnosis$date_at_diagnosis, format="%m/%d/%Y")

diagnosis$date_symptoms_began<-as.Date(diagnosis$date_symptoms_began, format="%m/%d/%Y")
#joining the two dataframes.
demographic.with.diagnosis <- inner_join(demographic, diagnosis, by="patient_id")

#new column- age at diagnosis
demographic.with.diagnosis<-
  mutate(demographic.with.diagnosis, age.at.diagnosis=as.numeric(demographic.with.diagnosis$date_at_diagnosis-demographic.with.diagnosis$dob)/365)

#graph age at diagnosis
ggplot(demographic.with.diagnosis, aes(x = age.at.diagnosis)) +
    geom_histogram(binwidth = 4)

#median age at diagnosis. 

demographic.with.diagnosis$age.at.diagnosis<- as.numeric(unlist(demographic.with.diagnosis$age.at.diagnosis))

demographic.with.diagnosis|>
  summarize(age.at.diagnosis=mean(age.at.diagnosis, na.rm=TRUE))
#Mean age at diagnosis is 36.64 years old.


median(demographic.with.diagnosis$age.at.diagnosis)

#median age at diagnosis is 39 years old.
```

Below you can see that we have more female participants in our registry than male participants.

```{r}
#chart age of participants
ggplot(data=demographic, aes(x=sex))+
  geom_bar()
```

Below you can see that most of our registry participants identify as white.

```{r}
ggplot(data=demographic, aes(x=race))+
  geom_bar()+
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

When reviewing diagnosis information, we can see that most of our participants self identify as having HHV8-unknown multicentric Castleman disease (MCD) or unicentric Castleman disease (UCD).

```{r}

#Here I am going to breakdown diagnosis by sex.

#changing the column name to something more useful.
colnames(demographic.with.diagnosis)[22]="CD_Diagnosis"

class(demographic.with.diagnosis$CD_Diagnosis)

#Labelling the diagnoses
demographic.with.diagnosis<- mutate(demographic.with.diagnosis, CD_Diagnosis
                                    =case_match(CD_Diagnosis, 
                                                1~"CD subtype unknown",
                                                2~"UCD",
                                                3~"HHV8 Pos MCD",
                                                4~"HHV8 Unknown MCD",
                                                5~"iMCD",
                                                6~"Not CD but has CD path report"
                                                                                        ))

#First, just a plot of diagnosis count.
ggplot(data=demographic.with.diagnosis, aes(x=CD_Diagnosis))+
  geom_bar()+
  theme(axis.text.x = element_text(angle=45, hjust=1))


#Plot of diagnosis by gender.
ggplot(data=demographic.with.diagnosis, aes(x = CD_Diagnosis, fill = sex)) +
    geom_bar(position = "fill")+
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

## Results {#sec-results}

In the results, I will begin to review the interactions and impacts between the registry and our population. I will review 4 main ideas. First, the average length of time from symptom onset to time of diagnosis to indicate how fast our population is receiving a diagnosis. You can see below that the average time from symptom onset to diagnosis is 147 days. This can be concerning for our patient population as some experience severe symptoms that potentially lead to death. We would hope patients would be diagnosed at a faster rate.

```{r}
#Average length of time from symptom onset to time of diagnosis. This tells us whether our recently published diagnostic criteria have affected the the rate in which patients are diagnosed.

#adding a new column called years from symptom onset to diagnosis.
demographic.with.diagnosis<-
  mutate(demographic.with.diagnosis, Years_from_symptom_onset_to_diagnosis=as.numeric(demographic.with.diagnosis$date_at_diagnosis-demographic.with.diagnosis$date_symptoms_began)/365)

demographic.with.diagnosis|>
  summarize(Years_from_symptom_onset_to_diagnosis=median(Years_from_symptom_onset_to_diagnosis, na.rm=TRUE))

#Average time from symptom onset to diagnosis is 0.4 years.

#Will do the same in days.
demographic.with.diagnosis<-
  mutate(demographic.with.diagnosis, Time_from_symptom_onset_to_diagnosis=as.numeric(demographic.with.diagnosis$date_at_diagnosis-demographic.with.diagnosis$date_symptoms_began))

demographic.with.diagnosis|>
  summarize(Time_from_symptom_onset_to_diagnosis=median(Time_from_symptom_onset_to_diagnosis, na.rm=TRUE))

#Average time from symptom onset to diagnosis is 147 days.

ggplot(demographic.with.diagnosis, aes(x = Time_from_symptom_onset_to_diagnosis)) +
    geom_histogram(binwidth = 4)

ggplot(demographic.with.diagnosis, aes(x = Years_from_symptom_onset_to_diagnosis)) +geom_histogram(binwidth = 4)
```

Second, I will review the difference between the date of diagnosis on different forms our patients fill out, to see how accurately they are filling out their forms. This tells us whether patients are understanding our questions and/or paying attention to what they enter. As seen below, most patients are reporting the same date on two different forms, which is good. However there is a large portion of patients who are reporting different dates, indicating a potential lack of understanding.

```{r}
#Now lets look at how the date of diagnosis differs on forms to see if patients are filling out forms correctly.to check for patient understanding/gaps in patient engagement.

#joining the two forms: diagnostic and eligibility.
eligibility.with.diagnosis <- inner_join(eligibility, diagnosis, by="patient_id")


#converting both objects to a date format
eligibility.with.diagnosis$date_at_diagnosis<-as.Date(eligibility.with.diagnosis$date_at_diagnosis, format="%m/%d/%Y")

eligibility.with.diagnosis$date_on_diagnostic_pathology_report<-as.Date(eligibility.with.diagnosis$date_on_diagnostic_pathology_report, format="%m/%d/%Y")

#adding new column= diagnosis.date.diff (difference between the two dates on the two forms)
eligibility.with.diagnosis<-
  mutate(eligibility.with.diagnosis, diagnosis.date.diff=as.numeric(eligibility.with.diagnosis$date_on_diagnostic_pathology_report-eligibility.with.diagnosis$date_at_diagnosis))

#Median difference between the two dates is 0, indicating that most patients are accurately/consistently recording their date of a diagnostic path report in one form as the date of diagnosis in another.
eligibility.with.diagnosis|>
  summarize(diagnosis.date.diff=median(diagnosis.date.diff, na.rm=TRUE))

#Plot
ggplot(eligibility.with.diagnosis, aes(x = diagnosis.date.diff)) +
    geom_histogram(binwidth = 4)+
  xlim(-100, 100)

```

Third, I am analyzing the date of diagnosis to date of enrollment in our registry. This will shed some light on our recruitment strategies. Median date of consent is 346 days after date of diagnosis. This means, it takes patients about 1 year to become involved in our research. This is a gap we would like to shorten.

```{r}
#Date of diagnosis to date of enrollment in registry.

#converting consent date to a date format.
eligibility.with.diagnosis$consent_date<-as.Date(eligibility.with.diagnosis$consent_date, format="%m/%d/%Y")

#adding new column= consent.date.diff (difference between the two dates on the two forms)
eligibility.with.diagnosis<-
  mutate(eligibility.with.diagnosis, consent.date.diff=as.numeric(eligibility.with.diagnosis$consent_date-eligibility.with.diagnosis$date_at_diagnosis))


#Median difference between the two dates. Median date of consent is 346 days after date of diagnosis.
eligibility.with.diagnosis|>
  summarize(consent.date.diff=median(consent.date.diff, na.rm=TRUE))

#plot
ggplot(eligibility.with.diagnosis, aes(x = consent.date.diff)) +
    geom_histogram(binwidth = 4)+
  ylim(0,15)

count(eligibility.with.diagnosis, consent.date.diff)
#This count tells us that 3 patients consented to the registry before even having a diagnosis and that 8 patients consented the same day. This seems unlikely and most likely means that patients are automatically writing their date of diagnosis as "today's date" when they are completing the forms. However, this is only a small portion of patients.

```

\
Lastly, I would like to visually demonstrate our registry enrollment in the form of an interactive map. This helps us to visualize where we need to improve our recruitment. You can see below, that most of our patient population lives in Pennsylvania, New York, and California. This makes sense given the location of our center, populations of those states, and our network of doctors in those states.

```{r}

library(tidyverse)
library(sf)
library(RColorBrewer)
library(tigris)
library(grid)
library(leaflet)
library(tigris)
library(tmap)
library(RColorBrewer)

```

```{r}
#Downloading state data.
st<-states(cb=T, resolution="20m")
plot(st$geometry)

us.states<-st|>
  filter(!STATEFP %in% c("02", "11", 
                            "15", "60", 
                            "66", "69", 
                            "72", "78")) 

colnames(demographic.with.diagnosis)[10]="STUSPS"


#Created a dataframe with patient counts for each state.
df1<-demographic.with.diagnosis|>
  group_by(STUSPS)|>
  summarise(patient_id=n())

#joining with demographic data from registry with the geographic data.

map2<- left_join(us.states, df1, by="STUSPS")



```

```{r}
#making interactive leaflet map. 

pal_fun <- colorBin(palette = brewer.pal(9, "Blues"), 
                    bins = c(1, 5, 10, 20, 30, 40, 50))

#add pop up message
pu_message <- paste0(map2$STUSPS,
                     "<br># patients: ",       
                     map2$patient_id)

leaflet(map2) |>
  addPolygons(stroke = FALSE,                        
              fillColor = ~ pal_fun(patient_id),
              fillOpacity = 0.7, smoothFactor = 0.5, #
              popup = pu_message) |>
  addProviderTiles(providers$Stadia.StamenTonerLite) |>
  addLegend("bottomright",               
            pal = pal_fun,               
            values = ~ patient_id,  
            title = 'Patients Participating in ACCELERATE',      
            opacity = 1) |>              
  addScaleBar()


```

```{r}
#Static map in case leaflet map does not load.

myPalette <- colorRampPalette(brewer.pal(9, "BuPu"))

my_theme <- function() {
  theme_minimal() +  
  theme(axis.line = element_blank(),  # further customization of theme components
        axis.text = element_blank(),  # remove x and y axis text and labels
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),  # make grid lines invisible
        legend.key.size = unit(0.8, "cm"),           # increase size of legend
        legend.text = element_text(size = 16),       # increase legend text size
        legend.title = element_text(size = 16))      
}

ggplot() +
  geom_sf(data = us.states, aes(fill = map2$patient_id)) +
  ggtitle("Participation in ACCELERATE Registry") + 
  my_theme()+
  scale_fill_gradientn(name = "patient participation",      
                    colours = myPalette(100))    
```

## Conclusion {#conclusion}

In this project, I began to summarize the population of Castleman disease patients participating in the ACCELERATE Natural History Registry. From this we now know that 59% of enrollees are female and 73% of enrollees are from the USA. The mean age at diagnosis of our cohort is 36.64 years old. The average time from symptom onset to diagnosis is 147 days.

Regarding how patients interact with our registry, there are a few interesting findings. From the data on patient accuracy we can see that most patients are entering information accurately and consistently into the registry, indicating that our information is likely to be accurate and useful for our research. We also learned that it takes patients about 1 year to become involved in our research. This is a gap we would like to shorten and something we can focus on in our recruitment strategies.

Lastly, the patient participation interactive map is a great addition to our team's resources as we can continuously track patient participation throughout the US and clearly identify regions that need more contact, support, and recruitment into our research studies.

## References {#refs}

Zhang X, Rao H, Xu X, et al. Clinical characteristics and outcomes of Castleman disease: A multicenter study of 185 Chinese patients. *Cancer Sci*. 2018;109(1):199-206. doi:10.1111/cas.13439
