---
title: "BMIN503/EPID600 Project Template"
author: "Your Name"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

### **Overview**
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.
    \
    \
    **This project aims to geospatially compare social determinants of health. voting patterns, and COVID epidemiology to answer the questions:**\
    **1) Are county-level demographics (ex. urbanicity), social determinants of health (ex. poverty rates, education level, area deprivation index), and previous voting patterns (outcomes and turnout data for 2016, 2018, 2020 elections) associated with COVID-19 disease epidemiology (cases and deaths) and**\
    **2) Does the socioeconomic gradient in voting correlates with SES gradient in COVID epidemiology?**
    \
    \
    **I have spoken with the following faculty:**\
    **Julia Lynch, PhD** | Professor, Political Science, School of Arts and Sciences\
    **John H. Holmes** | Professor, Biostatistics and Epidemiology, Perelman School of Medicine\
    \
    **Dr. Holmes advised that my question initially may have had a very obvious answer (conservative voting history may be associated with greater COVID cases and deaths). Dr. Lynch advised me to narrow my question and define it much more clearly, such as which sociodemographic factors I would use. Dr. Lynch also suggested that voter turnout might be an interesting metric that has not yet been looked at in this context. Furthermore, in order to go beyond a purely descriptive project, I could also assess if the socioeconomic status gradient in voting correlates with SES gradient in COVID epidemiology.**\
    
    [Link to final project GitHub repository](https://github.com/tuhinas/EPID600_FinalProject)

### **Introduction**
Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.
   \
    \
    COVID-19 is an urgent and deadly public health crisis that has infected over 9.67 million Americans and claimed over 235,000 lives. There is evidence that political partisanship has influenced individuals’ voluntary decisions to engage prevention practices such as distancing and mask-wearing** (Grossman G, et.al., Proceedings of the National Academy of Sciences Sep 2020, 117 (39) 24144-24153; DOI: 10.1073/pnas.2007835117). Social determinants of health such as poverty and homelessness also have been linked to increased COVID-19 transmission (Abrams E, Szefler S. COVID-19 and the impact of social determinants of health. The Lancet Respiratory Medicine. 2020;8(7):659-661. doi:10.1016/s2213-2600(20)30234-4). However, voter turnout compared alongside social determinants of health has not yet been evaluated in the context of COVID-19 epidemiology. Given the great partisan divide currently plaguing American politics and the catastrophic toll COVID-19 has had on Americans, such analyses are timely and important in understanding what populations are vulnerable to infection. Therefore, the objective of this analyses is to geospatially compare social determinants of health. voting patterns, and COVID epidemiology.
    \
    \
    The problem described above is interdisciplinary because it looks at COVID-19 health outcomes in both a socioeconomic and political context. It is generally accepted that lower socioeconomic status is associated with lower civic engagement, and data has shown that higher levels of income inequality is associated with lower voter turnout (Galbraith, J., & Hale, J. (2008). State Income Inequality and Presidential Election Turnout and Outcomes. Social Science Quarterly, 89(4), 887-901. http://www.jstor.org/stable/42956351). Since COVID-19 transmission is both linked to socioeconomic status and influenced by personal behavior, interdisciplinary analyses using recent election results might help health officials better understand vulnerable populations in their communities. The partisan spin irrationally applied to mask-wearing practices and the lack of federal leadership has only amplified the partisan divide and amplified COVID-19 infections.

### **Methods**
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

#Installing/loading R packages
```{r}
library(dplyr)
library(tidyverse)
library(sf)
library(tidycensus)
library(leaflet)
```

## Installing MIT Election Data Package
```{r}
if (!require('devtools', quietly = TRUE)) install.packages('devtools')
library(devtools)
devtools::install_github('MEDSL/elections')
install.packages("elections")
library(elections)

## I can't get this next set of code to work even though this is how the MIT Election Data Science Lab says to import the data with the "elections" package

#Importing Presidential Precincts 2008 dataset
data(presidential_precincts_2008)
presidential_precincts_2008 %>%
  select(state, county_fips, precinct, candidate, office, votes) %>%
  head()
#Importing Presidential Precincts 2012 dataset
data(presidential_precincts_2012)
presidential_precincts_2012 %>%
  select(state, county_fips, precinct, candidate, office, votes) %>%
  head()
#Importing Presidential Precincts 2016 dataset
data(presidential_precincts_2016)
presidential_precincts_2016 %>%
  select(state, county_fips, precinct, candidate, office, votes) %>%
  head()

```

#Because the above installation is not working, I'm importing data from my Github repository
```{r}
presidential_precincts_2000_2016 <- read.csv("https://raw.githubusercontent.com/tuhinas/EPID600_FinalProject/master/countypres_2000-2016.csv", header = TRUE, sep = ",")
head(presidential_precincts_2000_2016)

pacounts_presidential_2020 <- read.csv("https://raw.githubusercontent.com/tuhinas/EPID600_FinalProject/master/20201110_pa_results.csv", header = TRUE, sep = ",")
head(pacounts_presidential_2020)
```

#Importing SES data
```{r}
#CDC Social Vulnerability Index Data
pa_svi <- read.csv("https://raw.githubusercontent.com/tuhinas/EPID600_FinalProject/master/CDC_SVI_Pennsylvania_downloaded20201116.csv", header = TRUE, sep = ",")
head(pa_svi)
```
#Importing COVID-19 case and death data from NYTimes GitHub repo
```{r}
#COVID-19 case data
covid19_county <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv", header = TRUE, sep = ",")
head(covid19_county)
```


### **Results**
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

Questions to answer:
1) Are county-level demographics (ex. urbanicity), social determinants of health (ex. poverty rates, education level, area deprivation index), and previous voting patterns (outcomes and turnout data for 2016, 2018, 2020 elections) associated with COVID-19 disease epidemiology (cases and deaths)

2) Does the socioeconomic gradient in voting correlates with SES gradient in COVID epidemiology?**


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

