---
title: "Accessing Differences in Breast Cancer Screening Utilization by Race"
subtitle: "BMIN503/EPID600 Final Project"
author: "Mattia A. Mahmoud"
format: 
  html:
    toc: true
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

## Overview {#sec-overview}

Breast cancer is the second leading cause of cancer death among women in the U.S., with 43,000 deaths projected during the year 2023. Worse, Black women have 40% higher risk of breast cancer mortality compared to White women and are more likely to be diagnosed with poor prognostic, later stage cancers such as stage IV triple negative breast cancer (ER/PR/HER2-). 

Breast density is a strong risk factor for breast cancer, with women with dense breasts having a five fold higher risk of breast cancer compared to women with non-dense breasts and thus they are considered "high risk". Because of this, sixteen states, including Pennsylvania (PA), have recently required coverage with co-pays and deductibles for supplemental screenings for breast cancer with MRI or ultrasound for women with dense breasts. This is a major step forward for high risk women. The study purpose is to estimate the potential impact of the new PA supplemental screening law on the mammography screening population in the Penn Medicine health system (specifically Hospital of the University of Pennsylvania), and examine differences in screening utilization and false negative mammogram rates by race. Of note, false negative mammograms are considered a failure of the screening system, given that women are more likely to have their cancers (if detected) detected at later stages with worse outcomes.

## Introduction {#sec-introduction}

Early detection and screening is vital for reducing racial breast cancer disparities. Breast density is the relative amount of fibroglandular versus fatty tissue in the breast.  Women with dense breasts have increased risk both for developing breast cancer as well as for false-negative mammograms, due to some failures of screening intervals to detect dense breasts on mammograms. Supplemental screening with breast MRI has been shown to reduce the interval cancer rate (false negative mammogram rate) among patients with dense breasts. In May 2023, Pennsylvania (PA) mandated insurance coverage of one supplemental screening study (ultrasound or breast MRI) annually for women with dense breasts. Prior research suggests that Black women have lower breast density than white women based on subjective grading from the American College of Radiology Breast Imaging and Reporting Data System (BI-RADS), partially due to larger average breast size and higher body mass index (BMI) than white women. Given that Black women have lower prevalence of dense breasts and higher prevalence of triple negative breast cancer, the new supplemental screening law may meaningfully reduce racial disparities in breast cancer prognosis and mortality by catching more cancers thereby, reducing false negative mammogram rates. The purpose of this study is to compare supplemental screening utilization by race and see if there are differences in false negative mammogram rates between eligible and non eligible women for the supplemental screening under the new PA law. 

My research problem is interdisciplinary and requires appropriate input from a **multidisciplinary team** including an epidemiologist (Dr. Anne Marie McCarthy), surgeon (Dr. Lola Fayanju), radiologist (Dr. Emily Conant), biostatistician (Dr. Jinbo Chen) to utilize the appropriate methods and have an appropriate clinical context for understanding the results of my research project. Dr. McCarthy's mentorship will focus on epidemiological methods to examine factors that underlie differences in screening utilization by race; with Dr. Fayanju, I will elucidate the importance of carefully considering race and ethnicity in the conduct of the epidemiologic research; with Dr. Conant, I will understand the differences in the interpretation of density measures and with Dr. Chen, I will use methods to evaluate the additive value of the screening law to improve breast cancer incidence and mortality by race.

## Methods {#sec-methods}

The dataset I utilized in this project was obtained from a screening cohort within the Penn Medicine database. Penn Medicine has a large network of breast screening practices that perform thousands of mammograms per year. Women are given a breast cancer risk factor questionnaire at the time of mammogram screening, later stored in the electronic health record (EHR). Information from mammograms, questionnaires, subsequent MRI screenings, and local (Pennsylvania, New Jersey, Delaware) Cancer Registries are linked together for patients as part of Dr. McCarthy's parent grant. 

We included 76,888 mammograms among women aged 40-74 years who identified as non Hispanic Black or White without prior breast cancer or known BRCA1/2 mutations who underwent screening mammography from 2015 to 2021 at the Hospital for the University of Pennsylvania. For each mammogram, breast density was characterized according to BI-RADS (Breast Imaging-Reporting and Data System) density categories: almost entirely fatty, scattered areas of fibroglandular density, heterogeneously dense, and extremely dense.

Women eligible for insurance coverage for supplemental screening in PA include those with:

-   extremely dense breasts

-   heterogeneously dense breasts and \>20% lifetime risk of breast cancer, estimated using the Gail model, which is a breast cancer risk prediction tool that provides a percentage of breast cancer risk for women at 5 years and for a lifetime.

We defined a false negative mammogram as a cancer diagnosis within one year of a non-actionable mammogram (initial BI-RADS Assessment Negative or Benign) as per American College of Radiology guidelines. We tested for associations of eligibility according to the PA law with risk of a false negative mammogram, adjusting for age and race among Black and White women with negative mammograms.

First of all, the raw clinical and outcome data will be imported. The raw data are in excel, and they were imported as two dataframes.

```{r}
library(tidyverse) 
library(comorbidity) # For Elixhauser comorbidity scores
library(ggplot2) # For graphs
library(lubridate) # For manipulating dates
library(gtsummary) #Summary statistics
library(RColorBrewer) # For coloring the plots
library(cowplot) # For combining multiple plots into one
library(ggpubr) # Combining multiple graphs into one figure 
library(pROC) # For cross-validation
library(dplyr) # For data cleaning
```

```{r}
#setwd("/Users/mattiamahmoud/Downloads")
data <- read_csv("/Users/mattiamahmoud/Downloads/HUP Screening Mammograms 2015-2021.csv")
```

Summary views of the raw patients clinical data were reviewed.

```{r}
str(data)
```

However, lots of information in raw data are not related to this project. The dataset was cleaned for better utilization based on the information needed.

```{r}
data <- data %>% rename("accession" = "Accession") %>% 
  select(Race,Age,GailRisk5Yr,GailRiskLife,OverallBreastComposition, StudyLevelOutcome, ComponentLevelOutcome, reg_dxdate, reg_subtype4, accession, bc_prior)
```

To investigate the relationships between race, eligibility status and false negative mammography rates, I will select another dataset that has definitive race/ethnicity category and merge it with the dataset with other patient characteristics to facilitate further use.

```{r}
race_variable <- read_csv("/Users/mattiamahmoud/Downloads/Race variable.csv")
```

```{r}
merger <- merge(x = data, y = race_variable, by = c("accession"))
```

I then defined a new variable as eligibility (binary: eligible or not) based on the criteria outlined in the methods section as "heterogeneously dense and Gail lifetime risk of \>20% OR extremely dense" according to the BI-RADS system among women who are between the ages of 40- 74 (when mammography screening first begins)

```{r}
age40_74 <- merger[merger$Age > 40 & merger$Age < 75,]

#make age into a numeric
age40_74$Age <- sub(" .*", "", age40_74$Age)
age40_74$Age <- as.numeric(age40_74$Age)

#####make Gail Lifetime risk into a numeric 
#remove % after numbers
age40_74$GailRiskLife <- sub("%", "", age40_74$GailRiskLife)
age40_74$GailRiskLife <- as.numeric(age40_74$GailRiskLife)

#make new variable with ONLY values of Gail life time risk
age40_74 <- age40_74[which(! age40_74$GailRiskLife=="No" | is.na(age40_74$GailRiskLife)==TRUE),]
age40_74 $GailRiskLife <- as.numeric(age40_74$GailRiskLife)
age40_74 <- age40_74[which(! is.na(age40_74$GailRiskLife)==TRUE),]

#extremely dense breasts
age40_74$ed <- ifelse((age40_74$OverallBreastComposition== 'Extremely dense [4]'), 1, 0)

#heterogenously dense breasts and Gail lifetime risk greater than 20 
age40_74$hd_gail <- ifelse((age40_74$OverallBreastComposition == 'Heterogeneously dense [3]' & age40_74$GailRiskLife >= 20), 1,0)

#eligibility variable
age40_74$eligible <- ifelse((age40_74$ed ==1 | age40_74$hd_gail== 1), 1,0)

```

Next I subsetted the cohort to just women with true negative or false negative outcomes only since we are interested in the rates of screening failure of "false negative" among negative mammograms and since "high risk" women benefit from supplemental screening due to catching "false negative" women who would have otherwise had their cancer undetected or detected at later stage disease. Additionally, I also created divided the dataset by race in order to look at associations between eligibility and rate of false negative mammograms among Black or among White women alone. The final datasets for everyone is entitled "new negative."

```{r}
negative <- subset(age40_74, StudyLevelOutcome == "TN" | age40_74$StudyLevelOutcome == "FN")

negative$outcome <- ifelse((negative$StudyLevelOutcome == 'FN'), 1,0)
```

```{r}
white_neg <- subset(negative, race_eth == "White")

black_neg <- subset(negative, race_eth == "Black or African American")

new_negative <- subset(negative, race_eth == "Black or African American" | race_eth == "White")

new_negative$race_eth <- ifelse((new_negative$race_eth == "Black or African American"), 0,1)

# new_negative$OverallBreastComposition <- ifelse((new_negative$OverallBreastComposition == 'Almost entirely fatty [1]'), 1,ifelse(new_negative$OverallBreastComposition=='Scattered fibroglandular density [2]'), 2, ifelse(new_negative$OverallBreastComposition== 'Heterogenously dense'), 3, 4)

new_negative$BreastDensity = ifelse((new_negative$OverallBreastComposition == 'Almost entirely fatty [1]'), 1, ifelse((new_negative$OverallBreastComposition=='Scattered fibroglandular density [2]'), 2,
                       ifelse((new_negative$OverallBreastComposition== 'Heterogeneously dense [3]'), 3,
                              ifelse((new_negative$OverallBreastComposition== 'Extremely dense [4]'), 4,5))))

new_negative %>% select(OverallBreastComposition, BreastDensity)
```

To get an overview of the characteristics of patients enrolled in this study, summary of patients characteristics based on race, age, Gail Score, BI-RADS density and different receptors status are necessary. I first changed variables as needed from character values to numerical values in order to obtain summary statistics in the results section.

```{r}
new_negative <- new_negative %>%
  select(accession, Age, GailRiskLife, StudyLevelOutcome, race_eth, BreastDensity, outcome, reg_subtype4, bc_prior, eligible) %>%
   mutate(race_eth = factor(race_eth, levels=c(0,1), labels=c("Black or African American", "White"))) %>%
 mutate(BreastDensity = factor(BreastDensity, levels=c(1,2,3,4), labels=c("Almost entirely fatty", "Scattered fibrograndular density", "Heterogeneously dense", "Extremely dense"))) %>%
  mutate(outcome = factor(outcome, levels=c(1,0), labels=c("False Negative","True Negative"))) %>%
  mutate(reg_subtype4 = factor(reg_subtype4,levels=c(0,1,2,3,4,9), labels=c("DCIS","ER/PR+/HER-","ER/PR+/HER+", "ER/PR-/HER2+", "ER/PR-/HER2-", "Invasive with unknown subtype"))) %>% 
  mutate(bc_prior = factor(bc_prior,levels=c(0,1), labels=c("No Prior History of Breast Cancer","Prior History of Breast Cancer"))) %>%
 mutate(eligible = factor(eligible,levels=c(0,1), labels=c("Not Eligible for Supplemental Screening","Eligible for Supplemental Screening")))
```

```{r}
table <- new_negative %>%
 select(race_eth,Age,GailRiskLife,BreastDensity, outcome, reg_subtype4, eligible) 
library(knitr)
library(kableExtra)
options(knitr.kable.NA='')
kable(summary(table))
```

## Results {#sec-results}

As with any epidemiologic study, it is important to have a Table 1 showing the characteristics of the study population. The final population had information from 45,873 unique screening mammograms. 44% of the screening population that met the inclusion/exclusion criteria were White and 56% were Black. The median age in the cohort was 57 and the median Gail Lifetime Risk score was 8.5%, indicating that greater than 50% of women are high risk in this cohort. For context, a woman with a 1.66% or higher Gail Lifetime risk score is considered high risk for breast cancer. Most women fell in the BI-RADS Category B or C (scattered fibroglandular tissue or heterogeneously dense). There was a total of 42 false negative mammograms among the whole cohort, indicating that it is a rare event. Finally, only 4.4% of Black or White women in the cohort were considered eligible for supplemental screening.

```{r}
data_summary <- new_negative %>%
   select(race_eth,Age,GailRiskLife,BreastDensity, outcome, reg_subtype4, eligible) 

table1<- tbl_summary(
  data_summary, 
  missing = "no",
) %>%
  modify_header(label = "**Variable**") %>%
  bold_labels()
table1
```

Then, I explored the descriptive statistics of age, BI-RADS density, race/ethnicity and eligibility status visually.

```{r}
library(ggplot2)
ggplot(data = age40_74, aes(x=Age, colour = "darkgoldenrod")) +
  geom_histogram(binwidth = 3) +
  ggtitle(paste("Patients age distribution")) +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

The age distribution appeared to be relatively normal around the median age of 57 years.

```{r}
ggplot(new_negative, aes(x=race_eth, fill = BreastDensity)) +
  geom_bar(position="fill")+
  scale_fill_manual(values=c("darkseagreen", "darkseagreen2", "chartreuse3", "aquamarine4"))+
  theme_bw()+
  labs(title = "Race/Ethnicity vs. BI-RADS Density",
       x = "Race/Ethnicity",
       y = "BI-RADS Density") +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

Looking at the bar plot of race by BI-RADS above we see that White women have higher percentages of people in the extremely dense and heterogeneously dense categories than Black women.

```{r}
ggplot(data=new_negative, aes(x=Age, fill=eligible)) +
  geom_histogram(binwidth = 3) +
  scale_fill_manual(values=c("burlywood", "cornsilk4", "gray90"))+
  theme_bw()+
  ggtitle(paste("Eligibility status by Age")) +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

Then, I want to look at the eligibility status of all patients as a whole to see if there is any relationship between eligibility according to the PA law and patient Ages. The result above shows that percentages of eligibility actually decrease with increasing age.

```{r}
ggplot(data=new_negative, aes(x=race_eth)) +
  geom_bar() +
  labs(title = "Patients race distribution",
       x = "Race",
       y = "Count") +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

```{r}
ggplot(new_negative, aes(x=eligible, fill = race_eth)) +
  geom_bar(position="fill")+
  scale_fill_manual(values=c("cadetblue1", "darkturquoise"))+
  theme_bw()+
  labs(title = "Race/Ethnicity vs. Eligibility Status",
       x = "Eligibility Status",
       y = "Race/Ethnicity") +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

There is a higher percentage of White women than Black women among the 4.4% of the total cohort who are eligible for supplemental screening.

Now, I will investigate the associations being eligible for supplemental screening and outcome of false negative rates BY race.

```{r}
white_new_neg <- subset(new_negative, race_eth == "White")
black_new_neg <- subset(new_negative, race_eth == "Black or African American")
```

```{r}

# FN - outcome, race, age, eligible are the exposures for White women
mylogit10 <- glm(outcome ~ Age + eligible, data = white_new_neg, family = "binomial")
summary(mylogit10)
exp(coef(mylogit10))
exp(cbind(OR = coef(mylogit10), confint(mylogit10)))

# FN - outcome, race, age, eligible are the exposures for Black women
mylogit11 <- glm(outcome ~ Age + eligible, data = black_new_neg, family = "binomial")
summary(mylogit11)
exp(coef(mylogit11))
exp(cbind(OR = coef(mylogit11), confint(mylogit11)))

```

Among White women, those who are eligible are 84% less likely to have a false negative mammogram, after adjusting for age. We are not able to accurately compute the same association for Black women, given that there are no false negative mammograms among Black women.

```{r}
# race = exposure, outcome = eligibility
mylogit6 <- glm(eligible ~ race_eth + Age, data = new_negative, family = "binomial")
summary(mylogit6)
exp(coef(mylogit6))
exp(cbind(OR = coef(mylogit6), confint(mylogit6)))

t5 <- tbl_regression(mylogit6, exponentiate = TRUE) %>%
  bold_p(t = 0.10) %>%
  bold_labels() %>%
  italicize_levels()
t5

# FN - outcome, race, age, eligible are the exposures
mylogit9 <- glm(outcome ~ race_eth + Age + eligible, data = new_negative, family = "binomial")
summary(mylogit9)
exp(coef(mylogit9))
exp(cbind(OR = coef(mylogit9), confint(mylogit9)))

t4 <- tbl_regression(mylogit9, exponentiate = TRUE) %>%
  bold_p(t = 0.10) %>%
  bold_labels() %>%
  italicize_levels()
t4
```

After adjusting for age, White women are 5.96 times more likely than Black women to be eligible for supplemental screening with statistical significance. After adjusting for age and eligibility status, White women are 50% less likely to have false negative mammograms compared to Black women, with statistical significance. Additionally, after adjusting for age and race, women who are eligible for supplemental screening are 72% less likely to have a false negative mammogram.

## Conclusion

Our findings show that Black women have less dense breasts, making them less likely to be eligible for supplemental screening under recently passed PA law. This is in part due to the fact that the BI-RADS breast density tool less captured the full range of densities for Black women, with most classified as having Category B or C density. PA eligibility criteria did not identify any of the Black women at HUP who had a false-negative mammogram yet identified 36% of White women as having a false negative mammogram. White women were more likely to be eligible for supplemental screening and less likely to have a false negative mammogrma compared to Black women Given that Black women have lower prevalence of dense breasts and higher prevalence of triple negative breast cancer, the new supplemental screening law is unlikely to meaningfully reduce racial disparities in breast cancer risk and screening. Methods to better assess breast cancer risk among high risk Black women are urgently needed to identify Black women that may benefit from supplemental screening. 

Limitations of this study included not being able to examine changes in the false negative mammography rate and eligibility rates over time since this was a cross sectional analysis of women only at one site at Penn Medicine (HUP). Additionally, given that there was no systemic criteria codified in the law to characterize "high risk women" , prior to the recently passed PA law, I couldn't compare longitudinally differences in eligibility rates nor report some of the characteristics according to new and old criteria to see how many more women would be eligible. Finally, more risk factor data could have been included in this study such as BMI, or history of genetic mutations that could influence eligibility and false negative rates. We were only able to adjust for age in the models given that Gail risk factor data was unavailable at the time of the analysis. Future studies can investigate whether there is a disparity in breast cancer detection rates by race which may influence disparities in mortality.

## References

1.  Kohler BA, Sherman RL, Howlader N, et al. Annual Report to the Nation on the Status of Cancer, 1975-2011, Featuring Incidence of Breast Cancer Subtypes by Race/Ethnicity, Poverty, and State. Journal of the National Cancer Institute. 2015;107(6):djv048.

2.  Bodewes, F. T. H., van Asselt, A. A., Dorrius, M. D., Greuter, M. J. W., & de Bock, G. H. (2022). Mammographic breast density and the risk of breast cancer: A systematic review and meta-analysis. Breast (Edinburgh, Scotland), 66, 62--68.

3.  Bakker, M. F., de Lange, S. V., Pijnappel, R. M., Mann, R. M., Peeters, P. H. M., Monninkhof, E. M., Emaus, M. J., Loo, C. E., Bisschops, R. H. C., Lobbes, M. B. I., de Jong, M. D. F., Duvivier, K. M., Veltman, J., Karssemeijer, N., de Koning, H. J., van Diest, P. J., Mali, W. P. T. M., van den Bosch, M. A. A. J., Veldhuis, W. B., van Gils, C. H., ... DENSE Trial Study Group (2019). Supplemental MRI Screening for Women with Extremely Dense Breast Tissue. The New England journal of medicine, 381(22), 2091--2102.

4.  Hussein, H., Abbas, E., Keshavarzi, S., Fazelzad, R., Bukhanov, K., Kulkarni, S., Au, F., Ghai, S., Alabousi, A., & Freitas, V. (2023). Supplemental Breast Cancer Screening in Women with Dense Breasts and Negative Mammography: A Systematic Review and Meta-Analysis. Radiology, 306(3), e221785.

5.  Are You Dense Advocacy. D.E.N.S.E. State Efforts. Available from: http://areyoudenseadvocacy.org/ (Accessed June 1, 2020).

6.  McCarthy AM, Keller BM, Pantalone LM, et al. Racial Differences in Quantitative Measures of Area and Volumetric Breast Density. Journal of the National Cancer Institute. 2016;108(10)

7.  Bissell, M. C. S., Kerlikowske, K., Sprague, B. L., Tice, J. A., Gard, C. C., Tossas, K. Y., Rauscher, G. H., Trentham-Dietz, A., Henderson, L. M., Onega, T., Keegan, T. H. M., Miglioretti, D. L., & Breast Cancer Surveillance Consortium (2020). Breast Cancer Population Attributable Risk Proportions Associated with Body Mass Index and Breast Density by Race/Ethnicity and Menopausal Status. Cancer epidemiology, biomarkers & prevention : a publication of the American Association for Cancer Research, cosponsored by the American Society of Preventive Oncology, 29(10), 2048--2056.

8.  Arasu, V. A., Miglioretti, D. L., Sprague, B. L., Alsheik, N. H., Buist, D. S. M., Henderson, L. M., Herschorn, S. D., Lee, J. M., Onega, T., Rauscher, G. H., Wernli, K. J., Lehman, C. D., & Kerlikowske, K. (2019). Population-Based Assessment of the Association Between Magnetic Resonance Imaging Background Parenchymal Enhancement and Future Primary Breast Cancer Risk. Journal of clinical oncology : official journal of the American Society of Clinical Oncology, 37(12), 954--963.

