---
title: "Ultradistal Radius Bone Mineral Density in Children: Results from the Bone Mineral Density in Childhood Study"
author: "Joseph M. Kindler"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

#Forward
##Frequently Used Abbreviations
UD: ultradistal (see figure at https://github.com/kindlerj/BMIN503_Final_Project/blob/master/Radius%20DXA%20and%20pQCT%20Scan.pdf)
aBMD (g/cm2): areal bone mineral density 
DXA: dual-energy X-ray absorptiometry
pQCT: peripheral quantitative computed tomography
Ct.vBMD (g/cm3): cortical volumetric bone mineral density (volumetric density of radius cortical bone)
Tt.Ar (cm2): total bone cross-sectional area (size of the cortical bone shell)
BMDCS: Bone Mineral Density in Childhood Study
AA: African American
non-AA: non-African American


# 1. Overview

Dual energy X-ray absorptiometry (DXA) is a cornerstone of clinical bone health assessment. The preferred skeletal sites for bone density measurement in pediatric patients include the total body and lumbar spine. However, due to a variety of reasons including the presence of implants, prior fracture, or excess body weight, obtaining reliable measurements at these skeletal regions is not practical. Therefore, efforts have been made to develop valid and informative measures of bone density at alternative skeletal regions of interest. The ultradistal radius (i.e., the wrist) is one skeletal site with strong potential, but this specific measure has yet to be thoroughly characterized in children with respect race and sex differences, changes during growth, tracking throughout childhood and adolescence, the confounding of stature, and it's ability to predict fracture. These gaps in knowledge should be addressed before this measurement can be used with confidence in subsequent studies or in clinical practice. Therefore, the goal of this project is to characterize ultradistal (UD) radius areal bone mineral density (aBMD) during childhood with respect to race, sex, age, stature, bone geometry, and fracture history. We will accomplish this study by performing a secondary analysis of the Bone Mineral Density in Childhood Study (BMDCS) which includes >2,000 black and non-black males and females ages 5-19 at baseline who completed over 10,000 measurements collected over a six-year period. Dr. Zemel and Dr. Kalkwarf's background on childhood growth and Dr. McCormack's background in R programming were excellent sources of knowledge for the development of this project, which can be located at the following link: https://github.com/kindlerj/BMIN503_Final_Project. 

# 2. Introduction 

The (BMDCS) has been instrumental in characterizing bone growth during childhood and identifying the factors that contribute to these complex processes. Several traditional and non-traditional bone phenotypes have been developed using BMDCS data, with the exception of bone mineral density (aBMD) at the UD radius. Understanding the factors associated with UD radius aBMD during childhood, such as age, sex, race, and bone geometry, will be criticial in developing this specific phenotype for clinical application, as well as understanding fracture etiology during growth since the forearm is the most common fracture site during the growing years.  

Both modifiable and non-modifiable factors are involved in achieving optimal bone health during childhood. The attainment of "peak bone mass," or maximum biological bone growth potential, during growth has significant ramifications on the risk for fracture and osteoporosis in adulthood. Bone-augmenting modifiable factors include physical activity, diet, and medications, while non-modifiable factors include genetics, population ancestry, sex, age, and sexual maturation. I am a nutritionist and bone biologist by training and my post-doc mentor, Dr. Babette Zemel (UPenn and Children's Hostpial of Philadelphia), is a biological anthropologist with vast knowledge of human growth. Dr. Zemel was the principal investigator on the BMDCS, which was a prospective multi-site study of greater than 2,000 children funded by the NIH, and has used this rich dataset to identify maturational, dietary, and genetic determinants of bone health. Dr. Heidi Kalkwarf (Cincinnati Children's Hospital Medical Center) is also a nutritionist and bone biologist, has extensive expertise in childhood growth, and is an active member of the BMDCS investigative team. Finally, Dr. Shana McCormack (UPenn and Children's Hostpial of Philadelphia) is an endocrinologist and is an expert in childhood growth and R programming. Collectively, this collaborative team has aided in developing the study aims, conducting longitudinal statistical analyses in R, and developing a plan of action to further develop this new phenotype (i.e., UD radius aBMD) in future studies with respect to understanding the genetic and behavioral factors involved in forearm bone density accrual. Current pediatric recommendations for clincal bone health assessment suggest using either the total body or lumbar spine as the two most appropriate skeletal regions for osteoporosis diagnosis. However, developing the UD radius aBMD phenotype will help efforts toward establishing this outcome as an ancillary measure of bone health in instances that a total body or lumbar spine bone density scan is not feasible or appropriate at these skeletal regions of interest. 

# 3. Methods

We used data from the Bone Mineral Density in Childhood Study (BMDCS) to complete this study. The BMDCS was a large, multi-center, multi-ethnic study including 2014 healthy black and non-black children and adolescents that completed approximately 10,000 laboratory visits over a six year period. Areal bone mineral density (aBMD) was assessed at the ultradistal (UD) radius using dual-energy X-ray absorptiometry (DXA), fracture history was assessed via self-report, bone geometry was assessed at the distal radius using peripheral quantitative computed tomography (pQCT), and height was measured using a wall-mounted stadiometer. Sample DXA and pQCT images can be found at the following link: https://github.com/kindlerj/BMIN503_Final_Project/blob/master/Radius%20DXA%20and%20pQCT%20Scan.pdf. Longitudinal mixed regression was used to assess changes in UD radius aBMD during childhood, as well as sex- and race-related differences in these changes. Additionally, tracking of UD radius bone density, as well as relationships between UD radius aBMD and fracture history and bone geometry were assessed using longitudinal mixed regression.

## 3.1. R Code: 

### 3.1.1. Install packages needed for this study.
```{r eval=TRUE}
install.packages("foreign")
install.packages("haven")
install.packages("gamlss")
install.packages("xlsx")
install.packages("tidyverse")
install.packages("powerlmm")
install.packages("nlme")
install.packages("lme4")
install.packages("lm.beta")
install.packages("plotly")
install.packages("mass")
install.packages("ppcor")
install.packages("dotwhisker")
library(dotwhisker)
library(broom)
library(dplyr)
library(dplyr)
library(ggplot2)
install.packages("gamlss")
library(broom)
library(lm.beta)
library(reshape2)
library(plyr)
library(lme4)
install.packages("lmerTest")
install.packages("bbmle")
install.packages("psych")
```

### 3.1.2. Import data
```{r eval=TRUE}
#Data set containing forearm cortical bone measurements
library(haven)
old_pqct.dta <- read_dta("~/Desktop/BMIN_503/Final_Project/BMDCS Data/old_pqct.dta")
pqct <- subset(old_pqct.dta, select = c(1:54))
head(pqct)

#Data set containing filter variable to exclude necessary measurements due to a variety of reasons (e.g., developed chronic condition, bedrest, unexplained weight loss/gain, initiation of exclusionary medication, etc.)
library(haven)
excl <- read_dta("~/Desktop/BMIN_503/Final_Project/BMDCS Data/old_excluded.dta")
head(excl)

#Data set containing pertinent deomgraphic, anthropometric, and forearm DXA data
library(haven)
old_ultradistal_radius_workfile_2018.09.10.dta <- read_dta("~/Desktop/BMIN_503/Final_Project/BMDCS Data/old_ultradistal_radius_workfile_2018.09.10.dta")
dxa <- subset(old_ultradistal_radius_workfile_2018.09.10.dta, select = c(randid, visit, sex, gender, height, weight, age, black, race, fore_r_ud_bmd_pc, fore_r_ud_bmc_pc, wb_tot_fat, wb_tot_lean, brknbone))
head(dxa)

#Data set containing UD BMD Z scores
ultra_distal_radius_bmd <- read_dta("~/Desktop/BMIN_503/Final_Project/BMDCS Data/ultra_distal_radius_bmd.dta")

#Removing duplicate variables from the Z score data set
ultra_distal_radius_bmd$bid <- NULL
ultra_distal_radius_bmd$age <- NULL
ultra_distal_radius_bmd$gender <- NULL
ultra_distal_radius_bmd$site <-NULL
ultra_distal_radius_bmd$visit_num <-NULL
ultra_distal_radius_bmd$black <-NULL
ultra_distal_radius_bmd$weight <- NULL
ultra_distal_radius_bmd$height <- NULL
#ultra_distal_radius_bmd$waz <- NULL
#ultra_distal_radius_bmd$bmiz <- NULL
#ultra_distal_radius_bmd$haz <- NULL
ultra_distal_radius_bmd$fore_r_ud_bmd_pc <- NULL
ultra_distal_radius_bmd$fore_r_ud_bmd_fulc <- NULL

```

### 3.1.3. Merge the data files (merge on 'randid' and 'visit')
```{r eval=TRUE}
fake <- dplyr::full_join(pqct, dxa, by = c("randid", "visit"))
bmdcs <- dplyr::full_join(fake, excl, by = c("randid", "visit"))
bmdcs <- dplyr::full_join(bmdcs, ultra_distal_radius_bmd, by = c("randid", "visit"))
```

### 3.1.4. Removing necessary measurements/subjects from the dataset
```{r eval=TRUE}
#Create a new variable for the participants that should be deleted. If a person has a number (1-7) (meaning, they need to be deleted), they'll receive a "FALSE", but if they have a "NA", then they'll receive a "TRUE".
bmdcs$exclude.filter <- as.factor(is.na(bmdcs$extype)) 
#Copy the 'extype' variable so it appears next to the newly generated variable from above... scroll through to confirm that this worked properly (all numbers (1 to 7) should have a corresponding "FALSE", and NA's should have a corresponding "TRUE").
bmdcs$dummy <- bmdcs$extype 
#This command uses the "exclude.filter" variable that we created in command 1 to get rid of the "FALSE" rows, thereby deleting the the necessary measures/subjects and retaining our final and valid measurements.
bmdcs <- bmdcs[bmdcs$exclude.filter==TRUE,] 
```

### 3.1.5. Creating new gender variable
```{r eval=TRUE}
bmdcs$gender.2 = as.factor(bmdcs$gender) #gender was numeric in 'bmdcs', so we need it to be a factor
```

### 3.1.6. Creating polynomial terms for age (for regression analyses)
```{r eval=TRUE}
bmdcs$age_squared <- bmdcs$age * bmdcs$age
bmdcs$age_cubed <- bmdcs$age * bmdcs$age * bmdcs$age
bmdcs$age_four <- bmdcs$age * bmdcs$age * bmdcs$age * bmdcs$age
```

### 3.1.7. Creating data frames for males and females, as well as for AA males, AA females, non-AA males, and non-AA females
```{r eval=TRUE}
#Males (includes both AA and non-AA)
bmdcs.male <- bmdcs[bmdcs$gender==1, ] 
#Females (includes both AA and non-AA)
bmdcs.female <- bmdcs[bmdcs$gender==2, ] 
#AA Male
bmdcs.AA.male <- bmdcs[bmdcs$gender==1 & bmdcs$black==1, ] 
#AA Female
bmdcs.AA.female <- bmdcs[bmdcs$gender==2 & bmdcs$black==1, ] 
#non-AA Male
bmdcs.nonAA.male <- bmdcs[bmdcs$gender==1 & bmdcs$black==0, ] 
#non-AA Female
bmdcs.nonAA.female <- bmdcs[bmdcs$gender==2 & bmdcs$black==1, ] 
```

### 3.1.8. Creating interaction terms for regression analyses
```{r eval=TRUE}
#Main BMDCS data set
#Age squared
bmdcs$age_squared_by_black <- bmdcs$age_squared * bmdcs$black
bmdcs$age_squared_by_gender <- bmdcs$age_squared * bmdcs$gender
bmdcs$age_squared_by_gender_by_black <- bmdcs$age_squared * bmdcs$gender * bmdcs$black

#Age cubed
bmdcs$age_cubed_by_black <- bmdcs$age_cubed * bmdcs$black
bmdcs$age_cubed_by_gender <- bmdcs$age_cubed * bmdcs$gender
bmdcs$age_cubed_by_gender_by_black <- bmdcs$age_cubed * bmdcs$gender * bmdcs$black

#Age fourth
bmdcs$age_four_by_black <- bmdcs$age_four * bmdcs$black
bmdcs$age_four_by_gender <- bmdcs$age_four * bmdcs$gender
bmdcs$age_four_by_gender_by_black <- bmdcs$age_four * bmdcs$gender * bmdcs$black

#BMDCS male and female data sets
#Age cubed
bmdcs.male$black_by_age_cubed <- bmdcs.male$age_cubed * bmdcs.male$black
bmdcs.female$black_by_age_cubed <- bmdcs.female$age_cubed * bmdcs.female$black

```

### 3.1.9. New data frames with labels for figures.
```{r eval=TRUE}
#BMDCS
#bmdcs.figs <- mutate(bmdcs, black = factor(black, levels=c(0, 1), labels=c("non-African American", "African American")))
#bmdcs.figs <- mutate(bmdcs.figs, gender.2 = factor(gender.2, levels=c(1, 2), labels=c("Male", "Female")))

#Female
#bmdcs.figs.female <- mutate(bmdcs.female, black = factor(black, levels=c(0, 1), labels=c("non-African American", "African American")))
#bmdcs.figs.female <- mutate(bmdcs.figs.female, gender.2 = factor(gender.2, levels=c(1, 2), labels=c("Male", "Female")))

#Male
#bmdcs.figs.male <- mutate(bmdcs.male, black = factor(black, levels=c(0, 1), labels=c("non-African American", "African American")))
#bmdcs.figs.male <- mutate(bmdcs.figs.male, gender.2 = factor(gender.2, levels=c(1, 2), labels=c("Male", "Female")))

#AA Female
#bmdcs.figs.aa.female <- mutate(bmdcs.AA.female, black = factor(black, levels=c(0, 1), labels=c("non-African American", "African American")))
#bmdcs.figs.aa.female <- mutate(bmdcs.figs.aa.female, gender.2 = factor(gender.2, levels=c(1, 2), labels=c("Male", "Female")))

##AA Male
#bmdcs.figs.aa.male <- mutate(bmdcs.AA.male, black = factor(black, levels=c(0, 1), labels=c("non-African American", "African American")))
#bmdcs.figs.aa.male <- mutate(bmdcs.figs.aa.male, gender.2 = factor(gender.2, levels=c(1, 2), labels=c("Male", "Female")))

##non-AA Female
#bmdcs.figs.non.aa.female <- mutate(bmdcs.nonAA.female, black = factor(black, levels=c(0, 1), labels=c("non-African American", "African American")))
#bmdcs.figs.non.aa.female <- mutate(bmdcs.figs.non.aa.female, gender.2 = factor(gender.2, levels=c(1, 2), labels=c("Male", "Female")))

#non-AA Male
#bmdcs.figs.non.aa.male <- mutate(bmdcs.nonAA.male, black = factor(black, levels=c(0, 1), labels=c("non-African American", "African American")))
#bmdcs.figs.non.aa.male <- mutate(bmdcs.figs.non.aa.male, gender.2 = factor(gender.2, levels=c(1, 2), labels=c("Male", "Female")))
```

### 3.1.10. Baseline data frame
```{r eval=TRUE}
#save a new df
#baseline <- bmdcs[bmdcs$visit==0, ] 
```

### 3.1.11. Handling biologically implausible outliers in Tt.Ar and Tb.vBMD
```{r eval=TRUE}
#Tt.Ar
#There is an outlier in Tt.Ar that needs to be removed, so we're going to save new DFs for the Tt.Ar analyses to bypass this biologically implausable value 
#new.ttar.df <- bmdcs.figs
#new.ttar.df.males <- bmdcs.figs.male
#new.ttar.df.females <- bmdcs.figs.female
#new.ttar.df.nonAA.male <- bmdcs.figs.non.aa.male

#This df replaces the 'bmdcs' data set
#new.ttar.df <- new.ttar.df[new.ttar.df$tot_a_r30>=10, ]

#This df replaces the 'bmdcs.males' data set
#new.ttar.df.males <- new.ttar.df.males[new.ttar.df.males$tot_a_r30>=10, ]

#This df replaces the 'bmdcs.females' data set
#new.ttar.df.females <- new.ttar.df.females[new.ttar.df.females$tot_a_r30>=10, ]

#This df replaces the 'bmdcs.nonAA.male' data set
#new.ttar.df.nonAA.male <- new.ttar.df.nonAA.male[new.ttar.df.nonAA.male$tot_a_r30>=10, ]


#Tb.vBMD
#Need to remove one outlier in Tb.vBMD, so we're going to save new DFs for the Tb.vBMD analyses to bypass this biologically implausable value 
#new.tbvbmd.bmdcs <- bmdcs
#new.tbvbmd.bmdcs.males <- bmdcs.male
#new.tbvbmd.bmdcs.figs.male <- bmdcs.figs.male

#This df replaces the 'bmdcs' data set 
#new.tbvbmd.bmdcs <- new.tbvbmd.bmdcs[new.tbvbmd.bmdcs$trab_den_r3>=50, ]

#This df replaces the 'bmdcs.males' data set
#new.tbvbmd.bmdcs.males <- new.tbvbmd.bmdcs.males[new.tbvbmd.bmdcs.males$trab_den_r3>=50, ]

#This df replaces the 'bmdcs.figs.male' data set
#new.tbvbmd.bmdcs.figs.male <- new.tbvbmd.bmdcs.figs.male[new.tbvbmd.bmdcs.figs.male$trab_den_r3>=50, ]
```

### 3.1.12. Tracking data frames
```{r eval=TRUE}
#save a new df
bmdcs.tracking <- bmdcs[bmdcs$visit==0 | bmdcs$visit==6, ] 
#save a df with only baseline data
bmdcs.tracking.0 <- bmdcs.tracking[bmdcs.tracking$visit==0, ]
#save a new df with only 6-year data
bmdcs.tracking.6 <- bmdcs.tracking[bmdcs.tracking$visit==6, ]
#merge the baseline and 6-year dfs
final.bmdcs.tracking <- dplyr::full_join(bmdcs.tracking.0, bmdcs.tracking.6, all.x = TRUE, by = c("randid"))
#check that they merged properly ---- good!
head(final.bmdcs.tracking)

final.bmdcs.tracking.male <- final.bmdcs.tracking[final.bmdcs.tracking$gender.x==1, ] #tracking df for males
final.bmdcs.tracking.female <- final.bmdcs.tracking[final.bmdcs.tracking$gender.x==2, ]  #tracking df for females
```

### 3.1.13. Age groups at baseline
```{r eval=TRUE}
#(Young group, < 9 years old) Exclude every row with an age greater than 9.999
track.age.young <- final.bmdcs.tracking[final.bmdcs.tracking$age.x <= 9.999, ]
#summary(track.age.young$age.x)

#(Medium age group, 10-14.9 years old) First exclude everyone with an age less than 10, next exclude eveyone with an age greater than 14.999
track.age.medium <- final.bmdcs.tracking[final.bmdcs.tracking$age.x >= 10, ]
track.age.medium <- track.age.medium[track.age.medium$age.x <= 14.999, ]
#summary(track.age.medium$age.x)

#(Old age group, >10 years old) Exclude every row with an age less than 15
track.age.old <- final.bmdcs.tracking[final.bmdcs.tracking$age.x >= 15, ]
#summary(track.age.old$age.x)
```

### 3.1.14. Fracture data frame
```{r eval=TRUE}
bmdcs.fracture <- bmdcs
bmdcs.fracture$brknbone[bmdcs.fracture$brknbone == 2] <- 0 #for logistic regression, values need to be 0 and 1, so we need to change all 2s to 0s
head(bmdcs.fracture$brknbone, 20) #confirm this transformation worked... successful!
```

# 4. Results

## 4.1. Baseline characteristics

```{r eval=TRUE}
#Baseline age
#describe(baseline$age)
#Baseline height Z-score
#describe(baseline$haz)
#Baseline height Z-score
#describe(baseline$waz)
#Baseline BMI Z-score
#describe(baseline$bmiz)

#Race and sex breakdown
#table(baseline$gender, baseline$black) #1 = male, 2 = female... #1 = AA, 0 = non-AA

#Number (and percent) of children with UD radius BMD less than -2.0 (i.e., have "low bone mass for age"). For reference, based on a standard normal distribution, 2.3% of the population should fall below -2.0.
#bmdcs$low.bone.mass.group <- cut(bmdcs$ud_radius_bmd_z, c(-8,-2, 8))
#summary(bmdcs$low.bone.mass.group)
#low.bone.mass.percent = (224/9741)*100
#low.bone.mass.percent

#Number of visits at each annual time point (0 = baseline)
#hist(bmdcs$visit)
#table(bmdcs$visit)

#Number of visits at each annual time point based on sex
#table(bmdcs$visit, bmdcs$black, bmdcs$sex)

#Number of visits at each annual time point based on sex and AA/non-AA
#table(bmdcs$visit, bmdcs$black, bmdcs$sex, bmdcs$black) #1 = AA, 0 = non-AA

```


## 4.2. Characterization of changes in UD radius aBMD during childhood

```{r eval=TRUE}
library(ggbiplot)
library(dplyr)
library(ggplot2)

#Raw Values
ggplot(data=bmdcs.figs, aes(x=factor(gender.2), fore_r_ud_bmd_pc)) + 
    geom_boxplot(color="black", fill="lightblue") +
    facet_grid(. ~black) +
    labs(title="A) UD Radius BMD in African American and non-African 
    American Males and Females") +
    labs(x="Gender", y="UD Radius BMD (g/cm2)")

#Z Scores
ggplot(data=bmdcs.figs, aes(x=factor(gender.2), ud_radius_bmd_z)) + 
    geom_boxplot(color="black", fill="lightblue") +
    facet_grid(. ~black) +
    labs(title="B) UD Radius BMD in African American and non-African 
    American Males and Females") +
    labs(x="Gender", y="UD Radius BMD (Z-score)")

```
### Figure 1. UD radius raw values (A) and Z-scores (B) by race and sex 

```{r eval=TRUE}

ggplot(data=bmdcs.figs.aa.female, aes(age, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="A) AA Female: Age vs. UD Radius BMD") + 
    geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) +
    geom_smooth(color = "lightblue", method = loess) +
    labs(x="Age (years)", y="UD Radius BMD (g/cm2)") +
    ylim(c(.1, .7)) 

ggplot(data=bmdcs.figs.aa.male, aes(age, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="B) AA Male: Age vs. UD Radius BMD") + 
    geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) +
    geom_smooth(color = "lightblue", method = loess) +
    labs(x="Age (years)", y="UD Radius BMD (g/cm2)")+
    ylim(c(.1, .7)) 
    
ggplot(data=bmdcs.figs.non.aa.female, aes(age, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="C) non-AA Female: Age vs. UD Radius BMD") + 
    geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) +
    geom_smooth(color = "lightblue", method = loess) +
    labs(x="Age (years)", y="UD Radius BMD (g/cm2)")+
    ylim(c(.1, .7)) 
    
ggplot(data=bmdcs.figs.non.aa.male, aes(age, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="D) non-AA male: Age vs. UD Radius BMD") + 
    geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) +
    geom_smooth(color = "lightblue", method = loess) +
    labs(x="Age (years)", y="UD Radius BMD (g/cm2)")+
    ylim(c(.1, .7)) 
```
### Figure 2. UD radius by age in AA females (A), AA males (B), non-AA females (C), and non-AA males (D)

```{r eval=TRUE}
ud.radius.model.1.2 = lmer(scale(fore_r_ud_bmd_pc) ~ (1 | randid) + black + gender + scale(age) + scale(age_squared), data = bmdcs)
summary(ud.radius.model.1.2)
#confint(ud.radius.model.1.2)
#nobs(ud.radius.model.1.2)
#extractAIC(ud.radius.model.1.2)
```
Race, gender, age, and age squared were significantly associated with UD radius aBMD. The positive relationship with race indicates greater aBMD in blacks versus non-blacks, and a negative relationship with gender indicates greater aBMD in males versus females. These results pertaining to race and sex differences in UD radius aBMD are supported by Figure 1 above. That the polynomial age term was signicicantly assiated with UD radius aBMD suggest that bone density increases during childhood in a non-linear fashion. These results pertaining to non-linear changes in UD radius aBMD are supported by Figure 2 above. It should be noted that additional polynomial terms (age cubed and age to the fourth power) were also investigated with respect to UD radius aBMD. Age squared was determined to be the most appropriate age term considering the strength of the relationship with the bone density outcome, as well as the best model fit which was assessed by comparing the AIC. For simplicity, only the age squared model is presented in this final report.

```{r eval=TRUE}
ud.radius.model.2.c = lmer(scale(fore_r_ud_bmd_pc) ~ (1 | randid) + black + gender + scale(age) + scale(age_squared) + scale(age_squared_by_black) + scale(age_squared_by_gender), data = bmdcs)
summary(ud.radius.model.2.c)
#confint(ud.radius.model.2.c)
#nobs(ud.radius.model.2.c)
#extractAIC(ud.radius.model.2.c)
```
Two-way interactions between age squared with race and sex, with respect to UD radius BMD Z-scores, were statistically significant. These results are consistent with the data presented in Figure 2. It should be noted that the three way interaction between age squared, race, and sex was also investigated, but was not statistically significant. For simplicity, the most parsimonious model is presented in this final report.


### 4.3. Characterizing the relationship between radius cortical and trabecular bone and UD radius aBMD during childhood

### r.3.1. Relationship between radius Tb.vBMD and UD radius aBMD
```{r eval=TRUE}
ggplot(data=new.tbvbmd.bmdcs.figs.male, aes(trab_den_r3, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="Males: Radius Tb.vBMD vs. UD Radius BMD") + 
    #geom_smooth(color = "red") #+ 
    geom_smooth(color = "green", method = lm) +
  #  geom_smooth(color = "lightblue", method = loess) + 
    labs(x="Radius Tb.vBMD (g/cm3)", y="UD Radius (g/cm2)") +
    ylim(c(.2, .6)) +
    xlim(c(100, 300))

ggplot(data=bmdcs.figs.female, aes(trab_den_r3, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="Females: Radius Tb.vBMD vs. UD Radius BMD") + 
    #geom_smooth(color = "red") #+
    geom_smooth(color = "green", method = lm) +
  #  geom_smooth(color = "lightblue", method = loess) +   
    labs(x="Radius Tb.vBMD (g/cm3)", y="UD Radius (g/cm2)") +
    ylim(c(.2, .6)) +
    xlim(c(100, 300))

```
### Figure 3. Relationship between UD radius aBMD and Tb.vBMD in males (A) and females (B). 

```{r eval=TRUE}
#Male
full.tbvbmd.udBMD.males = lmer(scale(fore_r_ud_bmd_pc) ~ (1 | randid) + black + scale(age) + scale(age_squared) + scale(trab_den_r3), data = new.tbvbmd.bmdcs.males)
summary(full.tbvbmd.udBMD.males)
#confint(full.tbvbmd.udBMD.males)
nobs(full.tbvbmd.udBMD.males)
extractAIC(full.tbvbmd.udBMD.males)

#Female
full.tbvbmd.udBMD.females = lmer(scale(fore_r_ud_bmd_pc) ~ (1 | randid) + black + scale(age) + scale(age_squared) + scale(trab_den_r3), data = bmdcs.female)
summary(full.tbvbmd.udBMD.females)
#confint(full.tbvbmd.udBMD.females)
nobs(full.tbvbmd.udBMD.females)
extractAIC(full.tbvbmd.udBMD.females)

#Sex interaction
new.tbvbmd.bmdcs$gender_by_trab_den_r3 <- new.tbvbmd.bmdcs$gender * new.tbvbmd.bmdcs$trab_den_r3

full.tbvbmd.udBMD.interaction = lmer(scale(fore_r_ud_bmd_pc) ~ (1 | randid) + gender + black + scale(age) + scale(age_squared) + scale(trab_den_r3) + scale(gender_by_trab_den_r3), data = new.tbvbmd.bmdcs)
summary(full.tbvbmd.udBMD.interaction)
#confint(full.tbvbmd.udBMD.interaction)
nobs(full.tbvbmd.udBMD.interaction)
extractAIC(full.tbvbmd.udBMD.interaction)

```
Radius Tb.vBMD is positively associated with UD radius aBMD in both females and males. The significant negative interaction indicates that the relationship between Tb.vBMD and UD radius aBMD is stronger in males (coded as 1) versus females (coded as 2). 

### 4.3.1. Relationship between radius Ct.vBMD and UD radius aBMD 
```{r eval=TRUE}
#All
ggplot(data=bmdcs, aes(crt_den_r30, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="Radius Ct.vBMD vs. UD Radius BMD") + 
    #geom_smooth(color = "red") #+ 
    geom_smooth(color = "green", method = lm) +
    geom_smooth(color = "lightblue", method = loess) + 
    labs(x="Radius Ct.vBMD (g/cm3)", y="UD Radius (g/cm2)") +
    ylim(c(.2, .625)) +
    xlim(c(950, 1250))
```
### Figure 4. Relationship between UD radius aBMD and Ct.vBMD in children. 

```{r eval=TRUE}
#All (males and females together since there was no significant sex interaction)
full.ctvbmd.udBMD.all = lmer(scale(fore_r_ud_bmd_pc) ~ (1 | randid) + black + gender + scale(age) + scale(age_squared) + scale(crt_den_r30), data = bmdcs)
summary(full.ctvbmd.udBMD.all)
confint(full.ctvbmd.udBMD.all)
nobs(full.ctvbmd.udBMD.all)
extractAIC(full.ctvbmd.udBMD.all)
```
Radius Ct.vBMD is positively associated with UD radius aBMD. The interaction between Ct.vBMD and sex was investigated, but was not statistically significant. Therefore, we present our findings in the total cohort only.

### 4.3.2. Relationship between Tt.Ar and UD Radius aBMD
```{r eval=TRUE}
#All
ggplot(data=new.ttar.df, aes(tot_a_r30, fore_r_ud_bmd_pc)) + 
    geom_point(color = "gray") +
    labs(title="Radius Tt.Ar vs. UD Radius BMD") + 
    #geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) +
    #geom_smooth(color = "lightblue", method = loess)
    labs(x="Radius Tt.Ar (cm2)", y="UD Radius BMD (g/cm2)") +
    ylim(c(.15, .6)) +
    xlim(c(40, 125))
```
### Figure 5. Relationship between UD radius aBMD and Tt.Ar in children. 

```{r eval=TRUE}
#All (males and females together since there was no significant sex interaction)
full.ttar.udBMD.all = lmer(scale(fore_r_ud_bmd_pc) ~ (1 | randid) + black + gender + scale(age) + scale(age_squared) + scale(tot_a_r30), data = new.ttar.df)
summary(full.ttar.udBMD.all)
#confint(full.ttar.udBMD.all)
nobs(full.ttar.udBMD.all)
extractAIC(full.ttar.udBMD.all)
```
Tt.Ar is positively associated with UD radius aBMD. The interaction between Tt.Ar and sex was investigated, but was not statistically significant. Therefore, we present our findings in the total cohort only.

## 4.4. Tracking, or consistency, of UD radius aBMD during childhood

### 4.4.1. Tracking of UD radius aBMD in total cohort 
```{r eval=TRUE}
ggplot(data=final.bmdcs.tracking, aes(ud_radius_bmd_z.x, ud_radius_bmd_z.y)) + 
    geom_point(color = "gray") +
    labs(title="Baseline UD Radius BMD vs. 6-Year UD Radius BMD") + 
    geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) +
    geom_smooth(color = "lightblue", method = loess)
```
### Figure 5. Relationship between baseline and year 6 UD radius aBMD.  

```{r eval=TRUE}
summary.lm(lm(ud_radius_bmd_z.x ~ ud_radius_bmd_z.y + black.x + gender.x, data=final.bmdcs.tracking))

```
UD radius aBMD Z-scores track strongly and linearly during childhood. The race and sex interactions were explored, but were not significant. Thus, the most parsimonious model was included in this final report. 

### 4.4.2. Tracking of UD radius aBMD based on baseline age
```{r eval=TRUE}
ggplot(data=track.age.young, aes(ud_radius_bmd_z.x, ud_radius_bmd_z.y)) + 
    geom_point(color = "gray") +
    labs(title="< 10 Years Old: Baseline UD Radius BMD vs. 6-Year UD Radius BMD") + 
    #geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) #+
    #geom_smooth(color = "lightblue", method = loess) 

ggplot(data=track.age.medium, aes(ud_radius_bmd_z.x, ud_radius_bmd_z.y)) + 
    geom_point(color = "gray") +
    labs(title="10 to 14.9 Years Old: Baseline UD Radius BMD vs. 6-Year UD Radius BMD") + 
    #geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) #+
    #geom_smooth(color = "lightblue", method = loess) 

ggplot(data=track.age.old, aes(ud_radius_bmd_z.x, ud_radius_bmd_z.y)) + 
    geom_point(color = "gray") +
    labs(title=">15 Years old: Baseline UD Radius BMD vs. 6-Year UD Radius BMD") + 
    #geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) #+
    #geom_smooth(color = "lightblue", method = loess)
```
### Figure 5. Relationship between baseline and year 6 UD radius aBMD based on age at baseline (< 10 years vs. 10 to 14.9 years vs. >= 15 years).  

```{r eval=TRUE}
library(dotwhisker)
library(broom)
library(dplyr)

m1_df <- tidy(younger.tracking.model) %>% filter(term != "(Intercept)") %>% mutate(model = "< 10 Years")
m2_df <- tidy(medium.tracking.model) %>% filter(term != "(Intercept)") %>% mutate(model = "10 to 14.9 Years")
m3_df <- tidy(older.tracking.model) %>% filter(term != "(Intercept)") %>% mutate(model = "> 15 Years")

three_model_whisker <- rbind(m3_df, m2_df, m1_df)

dwplot(three_model_whisker) +
  theme_bw() + 
  xlab("Coefficient Estimate") +
  ylab("") +
  ggtitle("Differences in Tracking of UD Radius aBMD Across 
  Age Groups During Childhood") 
```
### Figure 6. Tracking coefficients for UD radius aBMD in children < 10 years (blue) vs. 10 to 14.9 years (green) vs. >= 15 years (red).  

```{r eval=TRUE}
#Younger age (< 10 years old)
younger.tracking.model <- lm(ud_radius_bmd_z.y ~ ud_radius_bmd_z.x , data=track.age.young)
summary(younger.tracking.model)
confint(younger.tracking.model)
#Medium age (10 - 14.9 years old)
medium.tracking.model <- lm(ud_radius_bmd_z.y ~ ud_radius_bmd_z.x , data=track.age.medium)
summary(medium.tracking.model)
confint(medium.tracking.model)
#Older age (> 10 years old)
older.tracking.model <- lm(ud_radius_bmd_z.y ~ ud_radius_bmd_z.x , data=track.age.old)
summary(older.tracking.model)
confint(older.tracking.model)
```
There was a strong linear relationship between UD radius aBMD Z-scores at the baseline and 6-year timepoint in each of the three age groups. These relationships are visually presented via scatterplot in Figure 5, above. To visually demonstrate whether the tracking relationship differed across age groups, dot and whisker plots are presented in Figure 6. That the 95% confidence intervals for the tracking coefficient for all three age groups overlap, this suggests that UD radius aBMD does not track stronger or weaker at a given age. 

## 4.5. Characterizing the relationship between UD radius aBMD and stature during childhood

### 4.5.1. Linear relationship between UD Radius aBMD and baseline HAZ
```{r eval=TRUE}
ggplot(data=bmdcs, aes(haz, ud_radius_bmd_z)) + 
    geom_point(color = "gray") +
    labs(title="HAZ versus UD Radius aBMD Z-score") + 
    #geom_smooth(color = "red") +
    geom_smooth(color = "green", method = lm) #+
    #geom_smooth(color = "lightblue", method = loess)
```
### Figure 7. Relationship between height Z-score and UD radius aBMD Z-score.  

```{r eval=TRUE}
summary.lm(lm(ud_radius_bmd_z ~ haz, data=bmdcs))
```
The positive relationship between HAZ and bone density indicate that individuals with taller stature have greater UD radius aBMD Z-scores. However, the relatively weak estimate (estiamte = 0.098, P<0.0001) suggest that this relationship is not overwhelmingly strong, such that one standard deviation increase in HAZ corresponds to an approximately 0.1 standard deviation increase in UD radius aBMD.

### 4.5.2. Differences in UD radius aBMD across HAZ groups (HAZ <-1, -1 to 1, >1)
```{r eval=TRUE}
bmdcs$haz.groups <- cut(bmdcs$haz, c(-5, -1, 1, 5))

summary(bmdcs$haz.groups)
ggplot(data=bmdcs, aes(x=factor(haz.groups), y=ud_radius_bmd_z)) +
    geom_boxplot(color="black", fill="lightblue", na.action=na.exclude)

```
### Figure 8. Comparison of UD radius aBMD across groups based on stature (HAZ < -1 vs. HAZ = -1 to 1 vs. HAZ > 1).    

```{r eval=TRUE}
haz.group.anova <- aov(haz ~ haz.groups, data = bmdcs)
summary(haz.group.anova)
```
In agreement with the results presented above from the linear regression analysis, there was a significant difference in UD radius aBMD in groups based on HAZ.


## 4.6. Understanding whether UD radius aBMD Z-scores are associated with fracture in childhood

```{r eval=TRUE}
ggplot(data=bmdcs.fracture, aes(x=factor(brknbone), y=ud_radius_bmd_z)) +
    geom_boxplot(color="black", fill="lightblue", na.action=na.exclude)
```
### Figure 9. Comparison of UD radius aBMD between children with and without a history of fracture.   

```{r eval=TRUE}
summary((glm(brknbone~ ud_radius_bmd_z, data=bmdcs.fracture, family=binomial())))
```
UD radius aBMD Z-scores did not differ between children with (coded as 0) vs. without (coded as 1) a history of fracture. 

# Summary of Results 
In summary, UD radius aBMD increases non-linearly during childhood, and these changes are greater in blacks versus non-blacks and males versus females. In both sexes, distal radius Tb.vBMD was positively associated with changes in UD radius aBMD but this relationship appeared to be stronger in males versus females. Both radius Ct.vBMD and Tt.Ar were positively associated with UD radius aBMD and these relationships did not appreciably differ between males and females. Tracking correlation coefficients were strong, but differed only slightly across childhood. Specifically, UD radius aBMD appeared to track strongest in those who were younger (< 10 years old at baseline) and older (> 15 years old at baseline), but the strength of this relationship differed only marginally compared to those ages 10 to 15 years at baseline. There was a modest but significant positive correlation between UD radius aBMD and HAZ, indicating that the taller children had greater bone density at the forearm. However, given the relatively low strength of this correlation coefficient, the meaning of this relationship should not be overstates, as this might be an artifactual finding due to our large sample size (>10,000 measurements) and our lack of subjects with higher (>2.0) and lower (<-2.0) height Z-scores. Thus, suggesting that UD radius aBMD might provide a relatively unbiased measure of bone density, albeit compared to other more biased measures such as lumbar spine and total body skeletal sites, for individuals with short stature (HAZ > 2.0). Finally, logistic regression was performed to determine whether UD radius aBMD was associated with self-reported fracture history. Forearm bone density was slightly lower in those with a prior fracture, but this difference was not statistically significant. Since a substantial fracture history was part of the exclusion criteria for the original study, the link between UD radius aBMD and fracture still warrants more rigorous testing in appropriate pediatric cohorts.  
