---
title: "Adherence to AAP Guidelines for Treatment Duration for Acute Otitis Media and Implications for Clinical Decision Support"
author: "Jonathan Beus"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but keep the headers.

#### Abbreviations Used
Abbreviation | 
---|---
AAP | American Academy of Pediatrics
AOM | Acute Otitis Media
ASP | Antimicrobial Stewardship Program
CDS | Clinical Decision Support
ED | Emergency Department
QI | Quality Improvement


### Overview

[Jonathan Beus BMIN503 Final Project](https://github.com/beusj/BMIN503_Final_Project)

This project examines adherence to treatment guidelines for antibiotic duration for [acute otitis media](https://www.healthychildren.org/English/health-issues/conditions/ear-nose-throat/Pages/Middle-Ear-Infections.aspx) (common ear infection) for patients seen and discharged from our emergency department. An American Academy of Pediatrics (AAP) guideline was published in 2013 with soft recommendations regarding treatment duration [AAP Guidleline](https://pediatrics.aappublications.org/content/131/3/e964) which were reflected in a local "Clinical Pathway." Essentially, while 10 days of therapy has classicaly been used, for patients 2 years and older, 7 days of therapy is likely adequate. Our antimicrobial stewardship program (ASP) has asked that we consider implementing clinical decision support to encourage clinicians to follow these guidelines. In preparation for this, we would like to evaluate past rates of adherence to these guidelines and to identify patient and provider characteristics associated with adherence or non-adherence to these guidelines. This information will serve to define the magnitude of the problem and potentially highlight patient or provider populations for additional intervention.

To explore this question, we are using an internal data warehouse, extracted from the Electronic Health Record. We have written SQL queries to extract the pertinent information, identifying patients seen in our ED for Otitis Media from July 2011 through June 2019.

Name | Title | Division/Department | Primary Discussion Points
---|---|---|---
Joe Zorc, MD, MSCE | Professor of Pediatrics, Director of Emergency Information Systems | Department of Emergency Medicine | We discussed which factors were most likely to be associated with adherence to the guidelines, including provider type (trainee, advanced practice provider, attending), calendar year, and use of an Order Set. We also discussed potentially un-intended consequences of our CDS intervention including driving providers away from using the Order Set or encouraging providers to use broader-spectrum antibiotics.
Brandon Ku, MD | Associate Director for ED QI | Department of Emergency Medicine | We primarily discussed post-intervention assessment of the impact of a CDS-based QI intervention. Due to the rapid-cycle nature of improvement, we discussed using control charts for improvement monitoring. With this in mind, we set a goal of an absolute 10% increase in adherence to the guidelines for patients >= 2 years.
Talene Metjian, PharmD | Manager, Antimicrobial Stewardship Program | Antimicrobial Stewardship Program | The majority of non-adherence occurs in patients >= 2 years as the shorter duration was a change from prior practice. The use of 10 days in patients < 2 years was and is the standard duration. As a result, we should focus our analysis on the cohort of patients 2 years and older.



### Introduction 
#### Background and Significance
Antimicrobial resistance is a significant threat to modern healthcare. Antimicrobial resistance is developing more rapidly than new antimicrobials can be discovered. It is generally felt that antibiotic overuse is a major factor driving worsening resistance patterns. Antimicrobial stewardship is a field developed to help providers limit inappropriate antibiotic use. As discussed above, in 2013 an AAP policy guideline noted that, for patients >= 2 years, 7 days of antibiotic therapy was likely as effective as longer durations. The policy was well received, but probably more notable in that it provided guidance to providers about when *not* treating with antibiotics could be appropriate. The discussion about duration seems to have been overshadowed by the more surprising endorsement to not treat in specific cases. For many providers, when the decision was made to treat a patient with antibiotics, they continued to use 10 days of therapy.

When a patient diagnosed with AOM is discharged from our ED, there are several ways that an order for antibiotics can be placed. The provider could go to their "order tab" and enter the name of the antibiotic they would like to prescribe and, subsequently, the dosing and duration. This occurs some in the ED, but most ED providers have become accustomed to using a discharge Order Set. There is such an Order Set that currently exists for ear complaints. Historically, it has provided a list of antibiotics for providers to choose from (in addition to the ability to add a discharge diagnosis, instructions, etc). The antibiotic group is shared with Order Sets for other kinds of problems that might present to the ED and has historically been defaulted to 10 days. This has made ordering 10 days of antibiotics, the easiest thing to do. While outside the scope of the current report, we are implementing some features in the Order Set to help providers choose an appropriate antibiotic and also to guide them to the recommended duration based on age.

The goal of the current analysis and report is preparatory to CDS interventions and assessment of their effectiveness. As discussed above, we are interested in patient and provider factors that influence the selection of antibiotic duration.


#### Interdisciplinary Approach
For this project, we have primarily gathered resources with expertise in Clinical Informatics, Antimicrobial Stewardship (pharmacy), and Quality Improvement. Clinical informatics generally includes clinicians and draws from disciplines including human-computer interaction, decision science, analytics, and computer science. Clinical informatics is a burgeoning field and includes broad definitions, although here, we will generally consider its primary intent to be to help clinicians and patients by supporting quality, timely, efficient, and safe care through the use of information technology. While a software engineer might easily be able to add antibiotics and durations to an Order Set, a clinical informaticist gathers additional information from providers and/or their own clinical experience to think about the workflow by which a process is performed to improve the likelihood that the anticipated goal will be achieved in a way that supports, rather than disrupts, the provider's clinical activities. The clinical informaticists involved (JZ and JB) have extensively discussed which dependent-variables are reasonably extracted from our data warehouse and might be informative in targeting CDS.

Our ASP pharmacist (TM) has been crucial to help define which clinical actions are desired and to incoporate the appropriate antibiotics as well as durations for appropriate patients. Our ASP pharmacist and QI expert (BK) have been instrumental in confirming the decision to focus our analysis on patients >= 2 years and to exclude patients prescribed azithromycin. Azithromycin is not a preferred antibiotic for the treatment of AOM, but is sometimes used in patients with sever-allergies to the preferred agents. Unlike other antibiotics used to treat AOM, the duration of azithromycin is dependent on the dose used, not associated with age, and typically just 3 to 5 days. As a result, for patients treated with azithromycin, the prescription would almost always be adherent to the guideline to use 7 days of therapy. Its inclusion would thus artificially inflate the estimate of adherence to the prescribing recommendations.


### Methods
**In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.**

#### Describe data and general methodologic approach

```{r}



```




### Results
**Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.**
