---
title: "BMIN503/EPID600 Project"
author: "Margalit Haber"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
###Overview 
An abnormality on a head computed tomography scan (CT) is commonly used as an indication that a traumatic brain injury (TBI) event occurred for clinical studies, however, the diagnostic utility of a head CT abnormality for a TBI is not well understood. This project aims to identify whether a head CT abnormality is useful as a prognostic biomarker or whether other factors are more relevant for prognosis.

###Introduction
There are over 2 million occurrences of traumatic brain injury (TBI) in the united states every year. Of those 2 million cases, there is currently no means of predicting who will recover and who will develop chronic long-lasting symptoms. The inability to determine who will recover without intervention makes designing clinical research studies sensitive to patient selection. The inclusion and exclusion criteria selected for a TBI study is crucial to the success or failure of that study. A common inclusion criteria for TBI studies is the presence of a tissue abnormality on a head computed tomography scan (CT). It is utilized since it ensures that the underlying brain tissue is damaged after the impact, however, the diagnostic utility of head CT abnormalities has not been well characterized. This project aims to determine the diagnostic and prognostic utility of a head CT abnormality for TBI.
	
The analysis from this project will not only focus on head CT abnormalities as a biomarker for TBI, but will also investigate socioeconomic status and age as factors contributing to TBI outcome. The knowledge learned from this project will inform the medical field as to factors that make an individual susceptible to poor outcome after a TBI. This project is also aimed at taking a critical look at clinical research study design, and whether more informed decisions about subject selection is needed for future studies. Initial study design and subject selection criteria is important for the success of clinical research studies as well as epidemiological studies of secondary datasets.



### Methods
The proposed project utilizes a cohort of 100 subjects recruited at Penn Presbyterian Medical Center. The subjects were recruited within 72 hours of their injury, and then followed-up at 2 weeks and 6 months. The dataset includes basic demographic information as well as neuropsychiatric tests for memory and executive function. Linear regression will be used to determine whether positive head CT is associated with long lasting symptoms or outcomes. The symptoms and outocmes being tested will be drawn from the standardized tests RAVLT, BSI, GOSE, Insomnia severity scale, PHQ-9.


`#In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 
`

```{r eval=TRUE}
Data <- read.csv("DataSet_FinalProject.csv", header=TRUE)
library(dplyr)
Data.clean <- Data %>%
  mutate(gender=factor(sex_M, levels=c(0, 1, 2), labels=c(NA, "male", "female"))) %>%
  mutate(hct=factor(hct, levels=c(1, 2), labels=c("positive", "negative"))) %>%
  mutate(race=factor(race, levels=c(0, 1, 2, 3, 4, 5, 6), labels=c(NA, "Indian", "AlaskanNative.Inuit", "Asian", "NativeHawaiian.PacificIslander", "Black", "White"))) %>%
  mutate(ethnicity=factor(ethnicity, levels=c(1, 2), labels=c("Hispanic.Latin", "NonHispanic"))) %>%
  mutate(severity=recode(gcs, "c('1', '2', '3', '4', '5', '6', '7', '8')=3 ; c('9', '10', '11', '12')=2 ; c('13', '14', '15')=3")) %>%
  mutate(severity=factor(severity, levels=c(3, 2, 1), labels=c("Severe", "Moderate", "Mild"))) %>% 
  mutate(severity=recode(gcs, '1'=3, '2'=3, '3'=3, '4'=3, '5'=3, '6'=3, '7'=3, '8'=3 , '9'=2, '10'=2, '11'=2, '12'=2 , '13'=1, '14'=1, '15'=1)) %>%
  mutate(severity=factor(severity, levels=c(1, 2, 3), labels=c("Mild", "Moderate","Severe"))) %>%
  mutate(alcohol=factor(alcohol, levels=c(1, 2), labels=c("Sober", "Drunk"))) %>%
  filter(group == 1) %>%
Data.clean$education_years <- as.integer(as.character(Data.clean$education_years))
```

### Results
The study population consists of males and females between the ages of 18-93 with a median age of 56 years. The years of education for the study population ranges from 9-20 years with an average education of 12 years.
```{r eval=TRUE}
library(ggplot2)
library(plyr)

max(Data.clean$age, na.rm = TRUE)
min(Data.clean$age, na.rm = TRUE)
median(Data.clean$age, na.rm = TRUE) 

max(Data.clean$education_years, na.rm = TRUE)
min(Data.clean$education_years, na.rm = TRUE)
median(Data.clean$education_years, na.rm = TRUE)
```
There are proportionally more men enrolled in the study than females, which is consistant with previous literature regarding incidence of TBIs. The population consists of more white individuals. Incidence of inebriation at emergency department presentation for a TBI was higher in men than in women. 
```{r eval=TRUE}
ggplot(data=Data.clean, aes(gender)) + 
      geom_bar() +
      labs(title="Study Population: Demographics by gender")+
      labs(x="Gender")+
      theme_bw()
      
ggplot(data=Data.clean, aes(x=race, fill=factor(gender))) +
      geom_bar(position="stack") +
      labs(title="Study Population: Demographics by race") +
      theme_bw()

ggplot(data=Data.clean, aes(x=alcohol, fill=factor(gender))) +
      geom_bar(position="stack") +
      labs(title="Study Population: Sobriety at ED presentation") +
      labs(x="Alcohol") +
      theme_bw()

ggplot(data=Data.clean, aes(age, gcs, color=gender)) + 
    geom_point() + 
    labs(title="Study Population: Injury presentation across age") +
    labs(x="Age", y="GCS") + #Adds a layer with labels
    theme_bw()
```
Severity for TBI is currently classified by glascow coma scale scores. TBI severity is generally broken down in 3 categories: Mild(gcs=15-13), Moderate(gcs=9-12), and Severe(gcs=1-8). The population in this study showed a high proportion of Mildly injured subjects (primarily gcs=15). The age of subjects was well distributed over the Mild and Severe injury categories. There are few subjects falling within the Moderate injury category.
```{r eval=TRUE}
median(Data.clean$gcs, na.rm=TRUE)

ggplot(data=Data.clean, aes(severity, age)) +
    geom_violin(fill="lavender", draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(height=0, width=0.1) +
    labs(title="Study Population: Severity of injury at ED presentation") +
    xlab("Severity (GCS)") +
    ylab("Age") +
    theme_bw() #violin plot indicating age distribution across injury severity 

ggplot(data=Data.clean, aes(x=severity, fill=factor(hct))) +
      geom_bar(position="stack") +
      labs(title="Study Population: HCT+/- by severity") +
      theme_bw()

count(Data.clean$hct) #n for each group
```
`#Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.`