---
title: "An Examination of Socio-demographic Differences in Sedentary Behavior"
author: "Liming Huang"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but leave the headers.

### Overview
Sedentary behavior is increasingly being recognized as a determinant of poor cardiovascular health, independent of physical activity levels. However, the socio-demographic correlates of both screen-based sedentary behavior (i.e., television and computer use) and transport based (i.e., driving time) have yet to be defined. To address this gap, this population level, cross-sectional study will examine the relationship between several key socio-demographic factors (age, sex, race, education, employment, shift-work urban/rural location, depression status, season of assessment) with screen-based and transport-based sedentary time. Data will be obtained from the baseline sample (N~500,000) of the United Kingdom Biobank Project.  The specific goals of this study are to determine:

1.	Are there socio-demographic differences in sedentary behavior among the UK Biobank sample?

2.	What is the socio-demographic profile of high sedentary behavior?





### Introduction 
Sedentary behavior is defined as time spent engaged in sitting or lying down activities that require an energy expenditure of 1.0 to 1.5 basal metabolic rates. Scientific evidence is accumulating that adult sedentary behavior might be associated with cardiovascular disease, cancer, Type 2 diabetes, metabolic syndrome and individual cardiovascular risk factors, obesity, overweight and adiposity, mental health, musculoskeletal disorders, cardiovascular mortality, and all-cause mortality. Research into the prevalence and determinants of sedentary behavior is essential to identify populations with high levels of sedentary behavior, and plan targeted interventions for those populations.

Correlates of sedentary behavior span multiple disciplines, including social and psychological domains. Examining the association of variables within these domains with sedentary behavior can generate plausible hypotheses to be tested and can provide initial insights relevant to the development of appropriate and targeted interventions. The collaborators on the current research team span these disciplines. Specifically, Dr. Freda Patterson is a public health researcher with expertise in cardiovascular health behavior change, including physical activity and sedentary behavior while Dr. Susan Malone has a nursing science background with expertise in cardiometabolic health. Together, this interdisciplinary team has the expertise to guide and disseminate the findings from the current investigation. 


### Methods
The analytic sample for the current study is extracted from the UK Biobank cohort study baseline data. It has 100423 subjects and contains self-reported sedentary behavior metrics of daily hours of television viewing, recreational computer use and driving. Eight socio-demographic characteristics, i.e., age, gender, race, education,  empoyment, shift work, urban/rural location, current depression status, and season of assessment, are also included. Socio-demographic differences in sedentary behavior are examined using ANOVA (for discrete independent variables) or Poisson regression (for continuous independent variables). Socio-demographic profile of high sedentary behavior is explored using regressionn tree analysis.

### Data cleaning

```{r, eval=TRUE}
## Load dplyr for data cleaning
library(dplyr)
## Load ggplot2 for graphs
library(ggplot2)
## Load pastecs for descriptive statistics
library(pastecs)
## Load pscl for zero-inflated Poisson regression
library(pscl)
## Load rpart for decision tree analysis
library(rpart)
## Load partykit for decision tree plot
library(partykit)


## Import the data
rm(list=ls())
getwd()
setwd ("F:/projects/002_biobank/derived_data")
temp <- read.csv(file = "F:/epid600_penn/EPID600_Final_Project/sedentary_data_for_R.csv", header = TRUE)
## Define categorical variables as factors 
temp1 <- dplyr::mutate(temp, GENDER = factor(SEX, labels=c("Female", "Male")),
           RACE = factor(RACE_NEW, labels=c("Mixed/Other", "Asian","Black","White")),
           COLLEGE = factor(COLLEGE,labels = c("No", "Yes")),
           EMPLOYMENT = factor(EMPLOYMENT, labels = c("Employed",
                                                        "Unemployed",
                                                        "Retired")),
           SHIFT_WORK = factor(SHIFT_WORK, labels = c("No", "Yes")),
           RESIDENCE = factor(RESIDENCE, labels = c("Urban", "Rural")),
           DEPRESSION = factor(MH_STATUS, labels = c("No", "Yes")),
           SEASON_ASSESS = factor(SEASON_ASSESS, labels = c("Fall", 
                                                    "Winter",
                                                    "Spring",
                                                  "Summer")),
           TV_HRS2 = TV_HRS*2,
           COMP_HRS2 = COMP_HRS*2,
           DRIVE_HRS2 = DRIVE_HRS*2,
           offset = 2)

## Keep only the variables that are needed for the analysis  
temp2 <- dplyr::select(temp1, TV_HRS, COMP_HRS, DRIVE_HRS, AGE, RACE, GENDER, COLLEGE, EMPLOYMENT, SHIFT_WORK, RESIDENCE, DEPRESSION, SEASON_ASSESS, offset, TV_HRS2, COMP_HRS2, DRIVE_HRS2)

## Keep only the observations with nonomissing values for all variables of interest
data <- temp2[complete.cases(temp2),]
```

### Descriptive statistics for sedentary behavior (TV_HRS, COMP_HRS, and DRIVE_HRS)

```{r, eval=TRUE}
## Histogram for TV_HRS
 qplot(TV_HRS, data=data, geom="histogram", binwidth=1, col=I("blue"), fill=I("blue"), alpha=I(0.5))
## Descriptives for TV_HRS
stat.desc(data$TV_HRS) 
## Histogram for COMP_HRS
qplot(COMP_HRS, data=data, geom="histogram", binwidth=1, col=I("blue"), fill=I("blue"), alpha=I(0.5))
## Descriptives for COMP_HRS
stat.desc(data$COMP_HRS) 
## Histogram for DRIVE_HRS
qplot(DRIVE_HRS, data=data, geom="histogram", binwidth=1, col=I("blue"), fill=I("blue"), alpha=I(0.5))
## Descriptives for DRIVE_HRS
 stat.desc(data$DRIVE_HRS) 
```

### Descriptive statistics for socio-demographic variables

```{r, eval=TRUE}
## Bar plot for RACE
ggplot(data,aes(x=RACE)) +   geom_bar(fill="blue")
## Proportion table for RACE
prop.table(table(data$RACE))
## Bar plot for GENDER
ggplot(data,aes(x=GENDER)) +   geom_bar(fill="blue")
## Proportion table for GENDER
prop.table(table(data$GENDER))
## Bar plot for COLLEGE
ggplot(data,aes(x=COLLEGE)) +   geom_bar(fill="blue")
## Proportion table for COLLEGE
prop.table(table(data$COLLEGE))
## Bar plot for EMPLOYMENT
ggplot(data,aes(x=EMPLOYMENT)) +   geom_bar(fill="blue")
## Proportion table for EMPLOYMENT
prop.table(table(data$EMPLOYMENT))
## Histogram for AGE
qplot(AGE, data=data, geom="histogram", binwidth=1, col=I("blue"), fill=I("blue"), alpha=I(0.5))
## Descriptives for AGE
stat.desc(data$AGE)
## Bar plot for SHIFT_WORK
ggplot(data,aes(x=SHIFT_WORK)) +   geom_bar(fill="blue")
## Proportion table for SHIFT_WORK
prop.table(table(data$SHIFT_WORK))
 ## Bar plot for RESIDENCE
ggplot(data,aes(x=RESIDENCE)) +   geom_bar(fill="blue")
## Proportion table for RESIDENCE
prop.table(table(data$RESIDENCE))
 ## Bar plot for DEPRESSION
ggplot(data,aes(x=DEPRESSION)) +   geom_bar(fill="blue")
## Proportion table for DEPRESSION
prop.table(table(data$DEPRESSION))
## Bar plot for SEASON_ASSESS
ggplot(data,aes(x=SEASON_ASSESS)) +   geom_bar(fill="blue")
## Proportion table for SEASON_ASSESS
prop.table(table(data$SEASON_ASSESS))
```

### Socio-demographic differences in sedentary behavior

1. Socio-demographic differences in daily hours of television viewing

```{r, eval=TRUE}
## Zero inflation Poisson model for association between TV_HRS and AGE
summary(zeroinfl(TV_HRS2 ~ AGE + offset(offset), data=data ,dist="poisson"))$coefficients
## Boxplots of TV_HRS by GENDER
ggplot(data=data, aes(GENDER, TV_HRS)) +  geom_boxplot() 
## ANOVA model for TV_HRS with GENDER as a factor
summary.lm(aov(TV_HRS ~ GENDER, data=data))
## Boxplots of TV_HRS by COLLEGE
ggplot(data=data, aes(COLLEGE, TV_HRS)) + geom_boxplot() 
## ANOVA model for TV_HRS with COLLEGE as a factor
summary.lm(aov(TV_HRS ~ COLLEGE, data=data))
## Boxplots of TV_HRS by RACE
ggplot(data=data, aes(RACE, TV_HRS)) +   geom_boxplot() 
## ANOVA model for TV_HRS with RACE as a factor
summary.lm(aov(TV_HRS ~ RACE, data=data))
## Multiple comparison 
TukeyHSD(aov(TV_HRS ~RACE, data=data))
## Boxplots of TV_HRS by EMPLOYMENT
ggplot(data=data, aes(EMPLOYMENT, TV_HRS)) +   geom_boxplot() 
## ANOVA model for TV_HRS with EMPLOYMENT as a factor
summary.lm(aov(TV_HRS ~ EMPLOYMENT, data=data))
## Multiple comparison 
TukeyHSD(aov(TV_HRS ~EMPLOYMENT, data=data))
## Boxplots of TV_HRS by SHIFT_WORK
ggplot(data=data, aes(SHIFT_WORK, TV_HRS)) +  geom_boxplot() 
## ANOVA model for TV_HRS with SHIFT_WORK as a factor
summary.lm(aov(TV_HRS ~ SHIFT_WORK, data=data))
## Boxplots of TV_HRS by RESIDENCE
ggplot(data=data, aes(RESIDENCE, TV_HRS)) + geom_boxplot() 
## ANOVA model for TV_HRS with RESIDENCE as a factor
summary.lm(aov(TV_HRS ~ RESIDENCE, data=data))
## Boxplots of TV_HRS by DEPRESSION
ggplot(data=data, aes(DEPRESSION, TV_HRS)) +  geom_boxplot() 
## ANOVA model for TV_HRS with DEPRESSION as a factor
summary.lm(aov(TV_HRS ~ DEPRESSION, data=data))
## Boxplots of TV_HRS by SEASON_ASSESS
ggplot(data=data, aes(SEASON_ASSESS, TV_HRS)) + geom_boxplot() 
## ANOVA model for TV_HRS with SEASON_ASSESS as a factor
summary.lm(aov(TV_HRS ~ SEASON_ASSESS, data=data))
## Multiple comparison 
TukeyHSD(aov(TV_HRS ~SEASON_ASSESS, data=data))
```

2. Socio-demographic differences in daily hours of recreational computer use

```{r, eval=TRUE}
## Zero inflation Poisson model for association between COMP_HRS and AGE
summary(zeroinfl(COMP_HRS2 ~ AGE + offset(offset), data=data ,dist="poisson"))$coefficients
## Boxplots of COMP_HRS by GENDER
ggplot(data=data, aes(GENDER, COMP_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with GENDER as a factor
summary.lm(aov(COMP_HRS ~ GENDER, data=data))
## Boxplots of COMP_HRS by COLLEGE
ggplot(data=data, aes(COLLEGE, COMP_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with COLLEGE as a factor
summary.lm(aov(COMP_HRS ~ COLLEGE, data=data))
## Boxplots of COMP_HRS by RACE
ggplot(data=data, aes(RACE, COMP_HRS)) +  geom_boxplot() 
## ANOVA model for COMP_HRS with RACE as a factor
summary.lm(aov(COMP_HRS ~ RACE, data=data))
## Multiple comparison 
TukeyHSD(aov(COMP_HRS ~RACE, data=data))
## Boxplots of COMP_HRS by EMPLOYMENT
ggplot(data=data, aes(EMPLOYMENT, COMP_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with EMPLOYMENT as a factor
summary.lm(aov(COMP_HRS ~ EMPLOYMENT, data=data))
## Multiple comparison 
TukeyHSD(aov(COMP_HRS ~EMPLOYMENT, data=data))
## Boxplots of COMP_HRS by SHIFT_WORK
ggplot(data=data, aes(SHIFT_WORK, COMP_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with SHIFT_WORK as a factor
summary.lm(aov(COMP_HRS ~ SHIFT_WORK, data=data))
## Boxplots of COMP_HRS by RESIDENCE
ggplot(data=data, aes(RESIDENCE, COMP_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with RESIDENCE as a factor
summary.lm(aov(COMP_HRS ~ RESIDENCE, data=data))
## Boxplots of COMP_HRS by DEPRESSION
ggplot(data=data, aes(DEPRESSION, COMP_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with DEPRESSION as a factor
summary.lm(aov(COMP_HRS ~ DEPRESSION, data=data))
## Boxplots of COMP_HRS by SEASON_ASSESS
ggplot(data=data, aes(SEASON_ASSESS, COMP_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with SEASON_ASSESS as a factor
summary.lm(aov(COMP_HRS ~ SEASON_ASSESS, data=data))
## Multiple comparison 
TukeyHSD(aov(COMP_HRS ~SEASON_ASSESS, data=data))
```

3. Socio-demographic differences in dalily hours of driving

```{r, eval=TRUE}
## Zero inflation Poisson model for association between DRIVE_HRS and AGE
summary(zeroinfl(DRIVE_HRS2 ~ AGE + offset(offset), data=data ,dist="poisson"))$coefficients
## Boxplots of DRIVE_HRS by GENDER
ggplot(data=data, aes(GENDER, DRIVE_HRS)) + geom_boxplot() 
## ANOVA model for COMP_HRS with GENDER as a factor
summary.lm(aov(DRIVE_HRS ~ GENDER, data=data))
## Boxplots of DRIVE_HRS by COLLEGE
ggplot(data=data, aes(COLLEGE, DRIVE_HRS)) +  geom_boxplot()
## ANOVA model for DRIVE_HRS with COLLEGE as a factor
summary.lm(aov(DRIVE_HRS ~COLLEGE, data=data))
## Boxplots of DRIVE_HRS by RACE
ggplot(data=data, aes(RACE, DRIVE_HRS)) + geom_boxplot() 
## ANOVA model for DRIVE_HRS with RACE as a factor
summary.lm(aov(DRIVE_HRS ~RACE, data=data))
## Multiple comparison 
TukeyHSD(aov(DRIVE_HRS ~RACE, data=data))
## Boxplots of DRIVE_HRS by COLLEGE
ggplot(data=data, aes(COLLEGE, DRIVE_HRS)) +  geom_boxplot() 
## ANOVA model for DRIVE_HRS with COLLEGE as a factor
summary.lm(aov(DRIVE_HRS ~COLLEGE, data=data))
## Boxplots of DRIVE_HRS by EMPLOYMENT
ggplot(data=data, aes(EMPLOYMENT, DRIVE_HRS)) + geom_boxplot() 
## ANOVA model for DRIVE_HRS with EMPLOYMENT as a factor
summary.lm(aov(DRIVE_HRS ~EMPLOYMENT, data=data))
## Multiple comparison 
TukeyHSD(aov(DRIVE_HRS ~EMPLOYMENT, data=data))
## Boxplots of DRIVE_HRS by SHIFT_WORK
ggplot(data=data, aes(SHIFT_WORK, DRIVE_HRS)) + geom_boxplot() 
## ANOVA model for DRIVE_HRS with SHIFT_WORK as a factor
summary.lm(aov(DRIVE_HRS ~SHIFT_WORK, data=data))
## Boxplots of DRIVE_HRS by RESIDENCE
ggplot(data=data, aes(RESIDENCE, DRIVE_HRS)) + geom_boxplot()
## ANOVA model for DRIVE_HRS with RESIDENCE as a factor
summary.lm(aov(DRIVE_HRS ~RESIDENCE, data=data))
## Boxplots of DRIVE_HRS by DEPRESSION
ggplot(data=data, aes(DEPRESSION, DRIVE_HRS)) + geom_boxplot() 
## ANOVA model for DRIVE_HRS with DEPRESSION as a factor
summary.lm(aov(DRIVE_HRS ~DEPRESSION, data=data))
## Boxplots of DRIVE_HRS by SEASON_ASSESS
ggplot(data=data, aes(SEASON_ASSESS, DRIVE_HRS)) + geom_boxplot()
## ANOVA model for DRIVE_HRS with SEASON_ASSESS as a factor
summary.lm(aov(DRIVE_HRS ~SEASON_ASSESS, data=data))
## Multiple comparison 
TukeyHSD(aov(DRIVE_HRS ~SEASON_ASSESS, data=data))
```

### Socio-demographic profile of high sedentary behavior

1. Regression tree for daily hours of television viewing

```{r, eval=TRUE}
## Decision tree for TV_HRS with 10 fold cross validation and cp = 0.0001
tree_tv <- rpart(TV_HRS~AGE+SEASON_ASSESS+GENDER+RACE+COLLEGE+EMPLOYMENT+
             SHIFT_WORK+RESIDENCE+DEPRESSION,  method="anova", data=data, cp = 0.0001)
## The best tree has  3 splits at cp in [0.002, 0.01]
printcp(tree_tv)
## Save the best tree in best_tree_tv
best_tree_tv <- prune(tree_tv, cp = 0.01)
## Plot the tree for TV_HRS
plot(as.party(best_tree_tv), type="simple")
## Get R-squared 
1-best_tree_tv$cptable[nrow(best_tree_tv$cptable), "xerror"]
```

2. Regression tree for daily hours of recreational computer use

```{r, eval=TRUE}
## Decision tree for COMP_HRS with 10 fold cross validation and cp = 0.0001
tree_comp <- rpart(COMP_HRS~AGE+SEASON_ASSESS+GENDER+RACE+COLLEGE+EMPLOYMENT+
             SHIFT_WORK+RESIDENCE+DEPRESSION,  method="anova", data=data, cp = 0.0001)
## The best tree has  3 splits at cp in [0.003, 0.007]
printcp(tree_comp)
## Save the best tree in best_tree_comp
best_tree_comp <- prune(tree_comp, cp = 0.003)
## Plot the tree for COMP_HRS
plot(as.party(best_tree_comp), type="simple")
## Get R-squared
1-best_tree_comp$cptable[nrow(best_tree_comp$cptable), "xerror"]
```


3. Regression tree for daily hours of driving

```{r, eval=TRUE}
## Decision tree for DRIVE_HRS with 10 fold cross validation and cp = 0.0001
tree_drive <- rpart(DRIVE_HRS~AGE+SEASON_ASSESS+GENDER+RACE+COLLEGE+EMPLOYMENT+
             SHIFT_WORK+RESIDENCE+DEPRESSION,  method="anova", data=data,cp = 0.0001)
## The best tree has  3 splits at cp in [0.004, 0.02]
printcp(tree_drive)
## Save the best tree in best_tree_drive
best_tree_drive <- prune(tree_drive, cp = 0.005)
## Plot the tree for DRIVE_HRS
plot(as.party(best_tree_drive), type="simple")
## Get R-squared 
1-best_tree_drive$cptable[nrow(best_tree_drive$cptable), "xerror"]
```


### Results

Most participants were White (93.1%), female (55.5%), did not attend college (65.6%), had a regular work shift (89.8%), lived in urban areas (86.6%), did not have depression (73.0%), were employed (59.7%), and the mean age was 57.08 (SD=8.134). 

```{r, eval=TRUE}
## Proportion table for RACE
prop.table(table(data$RACE))
## Proportion table for GENDER
prop.table(table(data$GENDER))
## Proportion table for COLLEGE
prop.table(table(data$COLLEGE))
## Proportion table for EMPLOYMENT
prop.table(table(data$EMPLOYMENT))
## Descriptives for AGE
stat.desc(data$AGE)
## Proportion table for SHIFT_WORK
prop.table(table(data$SHIFT_WORK))
## Proportion table for RESIDENCE
prop.table(table(data$RESIDENCE))
 ## Bar plot for DEPRESSION
prop.table(table(data$DEPRESSION))
## Bar plot for SEASON_ASSESS
prop.table(table(data$SEASON_ASSESS))
```


On average, participants reported 2.73 hours of daily television viewing (SD=1.567), 1.18 hours of recreational computer use (SD=1.417), and, 0.80 hours of driving (SD=1.270). 


```{r, eval=TRUE}
## Descriptives for TV_HRS
stat.desc(data$TV_HRS) 
## Descriptives for COMP_HRS
stat.desc(data$COMP_HRS) 
## Descriptives for DRIVE_HRS
 stat.desc(data$DRIVE_HRS) 
```

There are statistically significant socio-demographic differences in sedentary behavior. In daily TV watching, males spent 0.02 hours more than females (p=0.08), blacks 0.18 hours more than whites and 0.54 hours more than Asians (all p<0.0001),  retired participants 0.54 hours more than the unemployed and 0.83 hours more than the employed (all p<0.0001),  those who did not attend college 0.88 hours more than college attendees (p<0.0001), those with regular work 0.15 hours more than shift workers (p<0.0001), and urban residents 0.15 hours more than rural residents (p<0.0001). In winter time spent watching TV was 0.06 hours more than in spring, 0.10 hours more than in fall, and 0.15 hours more than in summer (all p<0.0003). The daily hours of TV viewing increased by 1.53% with one year increase in age (p<0.0001). 

```{r, eval=TRUE}

## Zero inflation Poisson model for association between TV_HRS and AGE
summary(zeroinfl(TV_HRS2 ~ AGE + offset(offset), data=data ,dist="poisson"))$coefficients
## ANOVA model for TV_HRS with GENDER as a factor
summary.lm(aov(TV_HRS ~ GENDER, data=data))
## ANOVA model for TV_HRS with COLLEGE as a factor
summary.lm(aov(TV_HRS ~ COLLEGE, data=data))
## ANOVA model for TV_HRS with RACE as a factor
summary.lm(aov(TV_HRS ~ RACE, data=data))
## Multiple comparison of TV_HRS by RACE
TukeyHSD(aov(TV_HRS ~RACE, data=data))
## ANOVA model for TV_HRS with EMPLOYMENT as a factor
summary.lm(aov(TV_HRS ~ EMPLOYMENT, data=data))
## Multiple comparison of TV_HRS by employment status
TukeyHSD(aov(TV_HRS ~EMPLOYMENT, data=data))
## ANOVA model for TV_HRS with SHIFT_WORK as a factor
summary.lm(aov(TV_HRS ~ SHIFT_WORK, data=data))
## ANOVA model for TV_HRS with RESIDENCE as a factor
summary.lm(aov(TV_HRS ~ RESIDENCE, data=data))
## ANOVA model for TV_HRS with DEPRESSION as a factor
summary.lm(aov(TV_HRS ~ DEPRESSION, data=data))
## ANOVA model for TV_HRS with SEASON_ASSESS as a factor
summary.lm(aov(TV_HRS ~ SEASON_ASSESS, data=data))
## Multiple comparison of TV_HRS by season of assessment
TukeyHSD(aov(TV_HRS ~SEASON_ASSESS, data=data))
```


In daily recreational computer use, males spent 0.36 hours more than females (p<0.0001), blacks 0.18 hours more than asians and 0.48 hours more than whites (all p<0.0001), the employed 0.14 hours more than the unemployed and 0.19 hours more than the retired (all p<0.0001),  college attendees  0.44 hours more than those who did not attend college (p<0.0001), shift workers  0.04 hours more than those with regular work (p=0.02), urban residents 0.11 hours more than rural residents (p<0.0001), people with depression 0.09 hours more than those without depression (p<0.0001). In spring time spent using computer was 0.08 hours more than in fall, 0.09 hours more than in summer (all p<0.0001), but not significantly different from that in winter. The daily hours of recreational computer use decreased by 0.30% with one year increase in age (p<0.0001). 




```{r, eval=TRUE}
## Zero inflation Poisson model for association between COMP_HRS and AGE
summary(zeroinfl(COMP_HRS2 ~ AGE + offset(offset), data=data ,dist="poisson"))$coefficients
## ANOVA model for COMP_HRS with GENDER as a factor
summary.lm(aov(COMP_HRS ~ GENDER, data=data))
## ANOVA model for COMP_HRS with COLLEGE as a factor
summary.lm(aov(COMP_HRS ~ COLLEGE, data=data))
## ANOVA model for COMP_HRS with RACE as a factor
summary.lm(aov(COMP_HRS ~ RACE, data=data))
## Multiple comparison of COMP_HRS by RACE
TukeyHSD(aov(COMP_HRS ~RACE, data=data))
## ANOVA model for COMP_HRS with COLLEGE as a factor
summary.lm(aov(COMP_HRS ~ COLLEGE, data=data))
## ANOVA model for COMP_HRS with COLLEGE as a factor
summary.lm(aov(COMP_HRS ~ COLLEGE, data=data))
## ANOVA model for COMP_HRS with EMPLOYMENT as a factor
summary.lm(aov(COMP_HRS ~ EMPLOYMENT, data=data))
## Multiple comparison of COMP_HRS by employment status
TukeyHSD(aov(COMP_HRS ~EMPLOYMENT, data=data))
## ANOVA model for COMP_HRS with SHIFT_WORK as a factor
summary.lm(aov(COMP_HRS ~ SHIFT_WORK, data=data))
## ANOVA model for COMP_HRS with RESIDENCE as a factor
summary.lm(aov(COMP_HRS ~ RESIDENCE, data=data))
## ANOVA model for COMP_HRS with DEPRESSION as a factor
summary.lm(aov(COMP_HRS ~ DEPRESSION, data=data))
## ANOVA model for COMP_HRS with SEASON_ASSESS as a factor
summary.lm(aov(COMP_HRS ~ SEASON_ASSESS, data=data))
## Multiple comparison of COMP_HRS by season of assessment
TukeyHSD(aov(COMP_HRS ~SEASON_ASSESS, data=data))
```

In daily driving, males spent 0.50 hours more than females (p<0.0001), blacks 0.26 hours more than asians and 0.42 hours more than whites (all p<0.0001), the employed 0.52 hours more than the unemployed and 0.60 hours more than the retired (all p<0.003),  those who did not attend college  0.22 hours more than college attendees (p<0.0001), shift workers 0.70 hours more than those with regular work (p<0.0001), rural residents 0.14 hours more than urban residents (p<0.0001), those without depression 0.02 hours more than people with depression (p=0.013). Driving hours did not differ significantly across seasons. The daily hours of driving decreased by 2.2% with one year increase in age (p<0.0001). 



```{r, eval=TRUE}
## Zero inflation Poisson model for association between DRIVE_HRS and AGE
summary(zeroinfl(DRIVE_HRS2 ~ AGE + offset(offset), data=data ,dist="poisson"))$coefficients
## ANOVA model for COMP_HRS with GENDER as a factor
summary.lm(aov(DRIVE_HRS ~ GENDER, data=data))
## ANOVA model for DRIVE_HRS with COLLEGE as a factor
summary.lm(aov(DRIVE_HRS ~COLLEGE, data=data))
## ANOVA model for DRIVE_HRS with RACE as a factor
summary.lm(aov(DRIVE_HRS ~RACE, data=data))
## Multiple comparison of DRIVING_HRS by RACE
TukeyHSD(aov(DRIVE_HRS ~RACE, data=data))
## ANOVA model for DRIVE_HRS with COLLEGE as a factor
summary.lm(aov(DRIVE_HRS ~COLLEGE, data=data))
## ANOVA model for DRIVE_HRS with EMPLOYMENT as a factor
summary.lm(aov(DRIVE_HRS ~EMPLOYMENT, data=data))
## Multiple comparison of DRIVE_HRS by employment status
TukeyHSD(aov(DRIVE_HRS ~EMPLOYMENT, data=data))
## ANOVA model for DRIVE_HRS with SHIFT_WORK as a factor
summary.lm(aov(DRIVE_HRS ~SHIFT_WORK, data=data))
## ANOVA model for DRIVE_HRS with RESIDENCE as a factor
summary.lm(aov(DRIVE_HRS ~RESIDENCE, data=data))
## ANOVA model for DRIVE_HRS with DEPRESSION as a factor
summary.lm(aov(DRIVE_HRS ~DEPRESSION, data=data))
## ANOVA model for DRIVE_HRS with SEASON_ASSESS as a factor
summary.lm(aov(DRIVE_HRS ~SEASON_ASSESS, data=data))
## Multiple comparison of DRIVE_HRS by season of assessment
TukeyHSD(aov(DRIVE_HRS ~SEASON_ASSESS, data=data))
```

In the decision tree model of daily television viewing, education and employment status are the most predictive factors. Not attending college (tier 1) and being retired (tier 2) accumulated the highest mean daily hours of television viewing (M=3.50) versus the 1.98 mean hours reported by college attendees who were employed or unemployed. 

```{r, eval=TRUE}
## Plot the tree for TV_HRS
plot(as.party(best_tree_tv), type="simple")
## Get R-squared on the test set
1-best_tree_tv$cptable[nrow(best_tree_tv$cptable), "xerror"]
```


In the decision tree model of daily recreational computer use, education and gender became two most important factors. Male college graduates averaged 1.68 hours per day versus the 0.896 hours recorded by female non-college graduates. 

```{r, eval=TRUE}
## Plot the tree for TV_HRS
plot(as.party(best_tree_comp), type="simple")
## Get R-squared on the test set
1-best_tree_comp$cptable[nrow(best_tree_comp$cptable), "xerror"]
```


In the regression tree of transport based sedentariness, being employed (tier 1), male (tier 2) and not attending college (tier 3) distinguished the highest mean daily hours of driving (1.69) versus the 0.446 mean hours reported by unemployed or retired participants.

```{r, eval=TRUE}
## Plot the tree for DRIVE_HRS
plot(as.party(best_tree_drive), type="simple")
## Get R-squared on the test set
1-best_tree_drive$cptable[nrow(best_tree_drive$cptable), "xerror"]
```

These data showed a different socio-demographic profile to be most predictive for each sedentary behavior metric considered. Thus, depending on the form of sedentary behavior being intervened upon, a different demographic group will be at higher risk. These data support the multi-dimensional nature of sedentary behavior and underscore the need to measure and consider its multiple forms. 










