---
title: "Combining aggregate food and housing costs with individual factors for BRFSS Health Outcomes"
author: "Alexander Chan"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)

```  


### Overview
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

This project aims at identifying predictors of health outcomes using socioeconomic factors. Specifically, income, housing costs, food security, and food costs will be tested for association with surveyed outocmes from Behavioral Risk Factor Surveillance System (BRFSS) epidemiologic data. Measures of mental and physical health will be analyzed using secondary data from the American Community Survey (ACS), Housing Urban Development (HUD) fair market rent (fmr) data, Feeding America dataset. The main faculty for this project is Blanca Himes and Sherry Xie, who assisted with geospatial plotting and formatting.  

### Introduction 
Describe the problem addressed, its significance, and some background to motivate the problem.

Do aggregate factors compare well to individual factors? This would be particularly useful since census data is extensive. 

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. Use 5-year ACS data from 2014-2018 for poverty rate and median income. Collect HUD fmr rate for the past from 2017-2020. Finally I use the  "Map the Gap" calculated county-level meal cost to estimate food expense. The cost-of-food index is calculated by applying county-level sales tax rates to Nielsen basket-of-goods based on actual pounds purchased. This value was last calculated in 2010 by Feeding America. Necessary corrections to nominal housing costs for inflation were corrected using CPI values from 2010-2020. BRFSS data is ta


factors: poverty rate, median income,  median income spent on food, median income spent on rent, race, sex, education level. Look into high versus low cost-of-living metropolitan areas. 




### Exploratory Analyses

###
Download and install Rtools40 to use the BLS api: https://cran.r-project.org/bin/windows/Rtools/

### Reading and Cleaning Datasets
```{r, eval=TRUE}
# commonly-used packages
library(ggplot2)
library(dplyr)
library(magrittr)
library(tidyr)
library(gtsummary)

# tools for accessing BLS data
# install_github("mikeasilva/blsAPI", force = TRUE) # Install the blsAPI by Mike Silva
library(devtools)
library(blsAPI)
library(rjson)
library(blsAPI)
```


# Query CPI data from BLS
```{r, eval=TRUE}
# query data from BLS.gov

# load api key (must have the txt file)
api_key <-  read.csv("bls_api_key.txt", header=FALSE)[[1]]

# signatures to be queried from BLS.gov, 10-year CPI data from major metropolitan areas
signatures <- read.csv("metro_area_cpi_signatures.csv") 

# payload signatures to query BLS 
month_payload <- list('seriesid'= signatures$signature_monthly) #2018-2020 monthly CPI 
annual_payload <- list('seriesid'= signatures$signature_annual) #2018-2020 annual CPI

# query CPI 2018-2020 data from major metropolitan areas from BLS 
months_cpi <- blsAPI(month_payload, return_data_frame = TRUE) # monthly CPI dataframe
annual_cpi <- blsAPI(annual_payload, return_data_frame = TRUE) # annual CPI dataframe
```


```{r, eval=TRUE}

# add columns to join region labels by
months_cpi$signature_monthly <- months_cpi$seriesID
annual_cpi$signature_annual <- annual_cpi$seriesID

# join region and state labels to the cpi data
# complete monthly dataframe for 2018-2020 regional metro cpi data
months_cpi %<>%
  left_join(signatures[, c("signature_monthly","state","area")],by="signature_monthly") %>%
  dplyr::select(-seriesID)

# complete annual dataframe for 2018-2020 regional metro cpi data
annual_cpi %<>%
  left_join(signatures[, c("signature_annual","state","area")],by="signature_annual") %>%
  dplyr::select(-seriesID)

head(months_cpi)
head(annual_cpi)

```
```{r, eval=TRUE}
# Further format dataframes
# pivot data by metro area and years, add column for mean annual CPI

# annual cpi 
a_cpi <- annual_cpi %>% 
  pivot_wider(id_cols = c("area","year"),names_from = "periodName",
              values_from = "value")%>% # pivot on area and year using values from period
  mutate_at(vars(3:4), as.numeric) %>% # change CPI values to numeric
  mutate(annual = rowMeans(select(., c("1st Half", "2nd Half")), na.rm= TRUE))# mean annual cpi

# monthly cpi
m_cpi <- months_cpi %>%
  dplyr::select(-period) %>%
  pivot_wider(id_cols = c("area","year"), names_from = "periodName", 
              values_from = "value") %>% # pivot on area and year using values from period
  mutate_at(vars(2:14), as.numeric) %>% # change CPI values to numeric
  mutate(annual = rowMeans(select(.,-c("area", "year")), na.rm = TRUE)) # mean annual cpi

head(a_cpi)
head(m_cpi)
```

# Now let's visualize economic growth in various metropolitan areas around the US
```{r, eval=TRUE}
a_cpi_growth_rate <- as.data.frame(a_cpi) %>%
  dplyr::select(area, annual) %>%
  dplyr::group_by(area) %>%
  dplyr::summarise(growth = (max(annual)-min(annual))/min(annual)*100) #growth from 2018-2020

ggplot(a_cpi_growth_rate, aes(x = reorder(area, -growth), y = growth))+
  geom_col() + 
  theme(axis.text.x = element_text(angle = 90)) +
  theme(axis.title.x = element_blank())+
  geom_hline(yintercept=2.2, linetype="dashed", color = "red")+
  ylab("Growth (% CPI Change 2018-2020)")

```



### Query Fair market rent (FMR) data from HUD
```{r, eval=TRUE}
library(sf)
library(ggsn)
library(stringr)
library(grid)

# function to format the geo id column of FMR dataframes
format_id <- function(df){
  
  df$ID <- str_replace(as.character(df$fips2010),
                       as.character(df$CouSub),"")
  df$ID <- as.numeric(df$ID) + 50000000000
  df$ID <- paste0("0",as.character(df$ID))
  
  return(df)
}

# read in the housing costs: fair market rates (2017-2020)
# format relevent columns into a standard form
fmr_2017 <- read.csv("FY2017_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  dplyr::mutate_at(c("fmr0", "fmr1", "fmr2", "fmr3", "fmr4"), as.numeric)
fmr_2018 <- read.csv("FY2018_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  dplyr::mutate_at(c("fmr_0", "fmr_1", "fmr_2", "fmr_3", "fmr_4"), as.numeric) %>%
  dplyr::rename("CouSub" = "cousub", "fmr0" = "fmr_0","fmr1" = "fmr_1","fmr2" = "fmr_2", 
         "fmr3" = "fmr_3", "fmr4" = "fmr_4", "Metro_code" = "metro_code")
fmr_2019 <- read.csv("FY2019_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  dplyr::mutate_at(c("fmr_0", "fmr_1", "fmr_2", "fmr_3", "fmr_4"), as.numeric) %>%
  dplyr::rename("CouSub" = "cousub","fmr0" = "fmr_0","fmr1" = "fmr_1","fmr2" = "fmr_2", 
         "fmr3" = "fmr_3", "fmr4" = "fmr_4", "Metro_code" = "metro_code") 
fmr_2020 <- read.csv("FY2020_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  dplyr::mutate_at(c("fmr_0", "fmr_1", "fmr_2", "fmr_3", "fmr_4"), as.numeric) %>%
  dplyr::rename("CouSub" = "cousub", "fmr0" = "fmr_0","fmr1" = "fmr_1","fmr2" = "fmr_2", 
         "fmr3" = "fmr_3", "fmr4" = "fmr_4", "Metro_code" = "metro_code") 
fmr_2021 <- read.csv("FY2021_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  dplyr::mutate_at(c("rent50_0", "rent50_1", "rent50_2", "rent50_3", "rent50_4"),
            as.numeric) %>%
  dplyr::rename("CouSub" = "cousub", "fmr0" = "rent50_0",
         "fmr1" = "rent50_1","fmr2" = "rent50_2", 
         "fmr3" = "rent50_3", "fmr4" = "rent50_4",
         "Metro_code" = "cbsasub21", "areaname" = "areaname21",
         "countyname" = "cntyname") 

# read in shape file of US counties
counties <- readRDS(gzcon(url('https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/uscounties_2010.rds'))) # decompress  rds file and read in shape file
counties$GEO_ID <- as.character(counties$GEO_ID)#change geo id to characters in counties df
counties$ID <- gsub("US","", counties$GEO_ID)  #remove US from ID

# read in metro area shape file and convert into spatial components
setwd("tl_2017_us_cbsa")
cbsa.sf <- st_read("tl_2017_us_cbsa.shp")
cbsa.sf %<>%
  dplyr::select(GEOID, geometry,NAME, NAMELSAD) %>%
  dplyr::rename(ID2 = GEOID, Metro = NAMELSAD)


# function to merge FMR dataframes with county geometries 
geo_format <- function(geo_df){
  fmr_geo <- counties %>%
    dplyr::select(c("NAME", "ID", "geometry")) %>%
    left_join(geo_df[,c("ID","fmr0", "fmr1", "fmr2", "fmr3", "fmr4",
                        "Metro_code", "areaname", "countyname")], by="ID") %>%
    distinct(Metro_code, .keep_all = TRUE)
}

# function to merge FMR dataframes with metro area geometries 
geo_format2 <- function(geo_df){
  
  geo_df %<>%
    dplyr::select(ID,fmr0,fmr1,fmr2, fmr3, fmr4, Metro_code, areaname, countyname)%>%
    distinct(Metro_code, .keep_all = TRUE) %>% # removed repeated counties
    dplyr::mutate(ID2 = substr(Metro_code, 12,16)) # metro IDs
  
  fmr_geo <- cbsa.sf %>%
    dplyr::select(c("NAME", "ID2", "geometry", "Metro")) %>%
    left_join(geo_df, by="ID2") # join by metro area

}

# format geographical IDs for all FMRs 2017-20201
fmr_2017 <- format_id(fmr_2017)
fmr_2018 <- format_id(fmr_2018)
fmr_2019 <- format_id(fmr_2019)
fmr_2020 <- format_id(fmr_2020)
fmr_2021 <- format_id(fmr_2021)

# add county geographical data to all FMRs 2017-20201
fmr_2017_geo <- geo_format(fmr_2017)
fmr_2018_geo <- geo_format(fmr_2018)
fmr_2019_geo <- geo_format(fmr_2019)
fmr_2020_geo <- geo_format(fmr_2020)
fmr_2021_geo <- geo_format(fmr_2021)

# add metro area geographical data to all FMRs 2017-20201
fmr_2017_geo2 <- geo_format2(fmr_2017)
fmr_2018_geo2 <- geo_format2(fmr_2018)
fmr_2019_geo2 <- geo_format2(fmr_2019)
fmr_2020_geo2 <- geo_format2(fmr_2020)
fmr_2021_geo2 <- geo_format2(fmr_2021)


```

```{r, eval=TRUE}
library(RColorBrewer)
library(leaflet)

my_theme <- function() {
  theme_minimal() +                                  # shorthand for white background color
  theme(axis.line = element_blank(),                 # further customization of theme components
        axis.text = element_blank(),                 # remove x and y axis text and labels
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),  # make grid lines invisible
        legend.key.size = unit(0.8, "cm"),           # increase size of legend
        legend.text = element_text(size = 16),       # increase legend text size
        legend.title = element_text(size = 16),
        plot.title = element_text(hjust = 0.5))      # increase legend title size
}

myPalette <- colorRampPalette(brewer.pal(9, "BuPu"))    # Blue-Purple
myPalette2 <- colorRampPalette(brewer.pal(9, "YlOrRd")) # Yellow-Orange-Red
myPalette3 <- colorRampPalette(brewer.pal(9, "YlGnBu")) # Yellow-Green-Blue

# geographic visualizations of housing costs
cpi_by_year <- read.csv("annual_cpi_us.csv", fileEncoding = "UTF-8-BOM")
cpi_by_year$adjust <- cpi_by_year$cpi/258.2937 # adjustment factor into 2020 dollars

# use the cpi_by_year adjustment to adjust rents to 2010 dollars
# rent in y year = (rent in current year) X (CPI y year/CPI current year)

# Select a color palette with which to run the palette function
pal_fun  <- colorNumeric("BuPu", NULL)      # Blue-Purple from RColorBrewer
pal_fun2 <- colorNumeric("YlOrRd", NULL)    # Yellow-Orange-Red from RColorBrewer

# function to get inflation correction factor for a given year into 2020 dollars
i_factor <- function(year){
  c_factor <- cpi_by_year$adjust[which(cpi_by_year$year == year)]
  return(c_factor)
}

# function to generate static map of rent costs by year across all coutnies in US
map_rent <- function(df, year){
    correction = i_factor(year)
 
  fmr_map = ggplot() +
    geom_sf(data=df,lwd=0, 
            aes(fill = fmr1/correction)) + # adjust factor included
    ggtitle(paste(year, "Median Rent for 1-bed")) + 
    scale_fill_gradientn(name = "Rent (2020 $)", 
                         colours = myPalette2(100),
                         limits = c(0,2000),
                         breaks = c(500,1000,1500,2000)) + # add limits and breaks to normalize scaling between years  
    north(data = df, symbol = 12, location = "bottomright") +
    scalebar(data = df, dist = 500, dist_unit = 'mi', transform = TRUE,
             location = "bottomleft",
             st.dist = 0.1) +
    my_theme()
}

# call static map function and display
fmr1_2017_map <- map_rent(fmr_2017_geo, 2017) # generate static rent map for counties
fmr1_2017_map # display rent map

# call static map function and display
fmr1_2017_map2 <- map_rent(fmr_2017_geo2, 2017) # generate static rent map for metro areas
fmr1_2017_map2 # display rent map

```

```{r, eval=TRUE}
# an interactive map of rent prices
library(htmlwidgets)
library(htmltools)

pal_fun <- colorNumeric("YlOrRd", NULL)  # Yellow-Orange-Red from RColorBrewer


# ------------------------------------------------- #
# HTML code to fix misaligned "na" in map legend
# ------------------------------------------------- #
library(htmlwidgets)
css_fix <- "div.info.legend.leaflet-control br {clear: both;}" # CSS to correct spacing
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML


# Funciton allows to plot interactive leaflet maps of median US rents by year
map_rent_i <- function(df, year){

#---------------------------------------------------#
# HTML code to add title to interactive maps
#---------------------------------------------------#
tag.map.title <- tags$style(HTML("
.leaflet-control.map-title { 
    transform: translate(-50%,20%);
    position: fixed !important;
    left: 50%;
    text-align: center;
    padding-left: 10px; 
    padding-right: 10px; 
    background: rgba(255,255,255,0.75);
    font-weight: bold;
    color: #000000;
    font-size: 20px;
  }
"))

title <- tags$div(
  tag.map.title, HTML(paste(year, "Median Rent:", 1, "Bed"))
)  
    
  correction = i_factor(year)# find correction factor
  
  # Pop-up message
  pu_message <- paste0("County: ", df$countyname,  
                     "<br>Rent: ", "$", floor(df$fmr1/correction)) # in 2010 dollars
  
  # Adding more customization 
  interactive_map <- leaflet(df) %>%
    addPolygons(stroke = FALSE,                        # remove polygon borders
                fillColor = ~pal_fun(fmr1/correction), # in 2010 dollars
                fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
                popup = ~pu_message) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
    #addProviderTiles(providers$Stamen.Toner) %>%
    #addProviderTiles(providers$Esri.NatGeoWorldMap)
    addLegend("bottomright",                           # location of legend
              pal=pal_fun,                             # palette function
              values=~fmr1/correction,                  # variable to be passed to palette function
            title = paste("Rent",1,"bed (2020 $)"),  # legend title
            opacity = 1) %>% # legend opacity (1 = completely opaque)
    addScaleBar() %>%
    addTiles() %>%
    addControl(title, position = "topleft", className="map-title") #add title panel
  
  return(interactive_map %<>% htmlwidgets::prependContent(html_fix))
}

# Call interactive map function and display 
fmr_2020_mapi <- map_rent_i(fmr_2020_geo, 2020) 
fmr_2020_mapi


```


### ACS poverty and income data

```{r, eval=TRUE}
library(tidycensus)
library(tidyr)

census_key <- read.delim("census_key.txt", header=FALSE)[1,1] # read in api key from local


# census_api_key(census_key,install=TRUE)

```

```{r, eval=TRUE}
acs.data <- get_acs(geography = "county",       # query data at the county level 
                   year = 2018,                 # end year (these will give us ACS 5-year estimates for 2014-2018)
                   variables = c("B17010_002",  # number of families falling below the poverty threshold
                                 "B17010_001",  #total number of families for which poverty was determined
                                 "B19013_001"), # median household income
                   key = census_key) 
acs.data %<>%
  dplyr::select(-moe) #remove the margin of error

# pivot on both estimate and total variables
acs.data <- pivot_wider(data=acs.data, names_from="variable", values_from="estimate")

acs.data %<>%
  dplyr::rename(poverty_total = B17010_001, 
         poverty_estimate = B17010_002,
         med_house_income = B19013_001)  %>% # give vars meaningful names
  dplyr::mutate(ID = paste0("0", as.character(050000000000 + as.numeric(GEOID)))) #create ID column to merge on


```

```{r, eval=TRUE}
# geospatial exploration of ACS median income from 2014-2018 by county 

# join acs data
acs.data.geo <- left_join(counties, acs.data[,c("ID", "poverty_total",
                                                "poverty_estimate", "med_house_income")],
                          by = "ID") # merge to counties

acs.data.geo %<>% # clean up the dataframe
  dplyr::select(GEO_ID, ID, COUNTY, poverty_total, poverty_estimate,
         med_house_income,geometry)

# plot income map
income_map <- ggplot() +
  geom_sf(data=acs.data.geo,lwd=0, 
          aes(fill = med_house_income)) + # adjust factor included
  ggtitle("Median Household Income (2014-2018)") + 
  scale_fill_gradientn(name = "Income", 
                       colours = myPalette2(100)) + # add limits and breaks to normalize scaling between years  
  north(data = acs.data.geo, location = "bottomright", symbol = 12) +
  scalebar(data = acs.data.geo, dist = 500, dist_unit = 'mi', transform = TRUE,
           location = "bottomleft",
           st.dist = 0.1) +
  my_theme()

income_map # show the median income map
```


```{r, eval=TRUE}
library(ggpubr)
options(scipen=10000)

# joins acs data and fmr data for any given fmr year
acs_join <- function(acs.geo.df, fmr.geo.df){
  acs.df <- dplyr::select(as.data.frame(acs.data.geo), -geometry)
  acs_fmr <- left_join(fmr.geo.df, acs.df, by = "ID")
  return(acs_fmr)
}

# combine acs data with fmr data
acs.fmr.2017 <- acs_join(acs.data.geo, fmr_2017_geo) 
acs.fmr.2018 <- acs_join(acs.data.geo, fmr_2018_geo) 
acs.fmr.2019 <- acs_join(acs.data.geo, fmr_2019_geo) 
acs.fmr.2020 <- acs_join(acs.data.geo, fmr_2020_geo) 
acs.fmr.2021 <- acs_join(acs.data.geo, fmr_2021_geo) 

```


```{r, eval=TRUE}
# calculate poverty rate and percent spent on housing by county for one year

thresh <- 0 # threshold for filtering poverty estimates

acs.fmr.2017 %<>%
  dplyr::filter(poverty_estimate > thresh) %>%
  dplyr::mutate(poverty_percent = poverty_estimate/poverty_total) %>%
  dplyr::mutate(housing_percent = (fmr1*12)/med_house_income)
acs.fmr.2018 %<>%
  dplyr::filter(poverty_estimate > thresh) %>%
  dplyr::mutate(poverty_percent = poverty_estimate/poverty_total) %>%
  dplyr::mutate(housing_percent = (fmr1*12)/med_house_income)
acs.fmr.2019 %<>%
  dplyr::filter(poverty_estimate > thresh) %>%
  dplyr::mutate(poverty_percent = poverty_estimate/poverty_total) %>%
  dplyr::mutate(housing_percent = (fmr1*12)/med_house_income)
acs.fmr.2020 %<>%
  dplyr::filter(poverty_estimate > thresh) %>%
  dplyr::mutate(poverty_percent = poverty_estimate/poverty_total) %>%
  dplyr::mutate(housing_percent = (fmr1*12)/med_house_income)


# plot map of percentage spent on median 1-bed apartment by median income
rent_income_map <- ggplot() +
  geom_sf(data=acs.fmr.2017,lwd=0, 
          aes(fill = housing_percent)) + # adjust factor included
  ggtitle("Percent of income spent on rent (2017 1-Bed)") + 
  scale_fill_gradientn(name = "% spent on rent", 
                       colours = myPalette2(100)) + # add limits and breaks to normalize scaling between years  
  north(data = acs.fmr.2017, location = "bottomright", symbol = 12) +
  scalebar(data = acs.fmr.2017, dist = 500, dist_unit = 'mi', transform = TRUE,
           location = "bottomleft",
           st.dist = 0.1) +
  my_theme()

rent_income_map # show the median income map

# relationship between amount spent on housing and poverty rate 
# FMR 1-bed data from 2017-2020
# Median income estimated from 2014-2018 ACS data
# generate linear regression fits for the relationship between housing burdern and poverty 

# looking at income alone across US counties

income_2017 <- ggplot(data = acs.fmr.2017, aes(x=med_house_income, y=poverty_percent))+
  geom_point() +
  xlab("Median Income ($)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2017 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))
  

income_2018 <- ggplot(data = acs.fmr.2018, aes(x=med_house_income, y=poverty_percent))+
  geom_point() +
  xlab("Median Income ($)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2018 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))

income_2019 <- ggplot(data = acs.fmr.2019, aes(x=med_house_income, y=poverty_percent))+
  geom_point() +
  xlab("Median Income ($)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2019 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))

income_2020 <- ggplot(data = acs.fmr.2020, aes(x=med_house_income, y=poverty_percent))+
  geom_point() +
  xlab("Median Income ($)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2020 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))

combined_income <- cowplot::plot_grid(income_2017,income_2018,
                                      income_2019,income_2020)

combined_income

# looking at housing cost alone across US counties

housing_2017 <- ggplot(data = acs.fmr.2017, aes(x=fmr1, y=poverty_percent))+
  geom_point() +
  xlab("Median Rent (1-Bed)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2017 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))

housing_2018 <- ggplot(data = acs.fmr.2018, aes(x=fmr1, y=poverty_percent))+
  geom_point() +
  xlab("Median Rent (1-Bed)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2018 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))

housing_2019 <- ggplot(data = acs.fmr.2019, aes(x=fmr1, y=poverty_percent))+
  geom_point() +
  xlab("Median Rent (1-Bed)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2019 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))

housing_2020 <- ggplot(data = acs.fmr.2020, aes(x=fmr1, y=poverty_percent))+
  geom_point() +
  xlab("Median Rent (1-Bed)") + 
  ylab("Poverty Rate %") + 
  ggtitle("2020 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5))

combined_housing <- cowplot::plot_grid(housing_2017, housing_2018,
                                      housing_2019, housing_2020)

combined_housing

# looking at income and housing cost combined across US counties

h_burden_2017 <- ggplot(data = acs.fmr.2017, aes(x=housing_percent, y=poverty_percent))+
  geom_point() +
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Percent Spent on Rent") + 
  ylab("Poverty Rate %") + 
  ggtitle("2017 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

h_burden_2018 <- ggplot(data = acs.fmr.2018, aes(x=housing_percent, y=poverty_percent))+
  geom_point() +
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Percent Spent on Rent") + 
  ylab("Poverty Rate %") + 
  ggtitle("2018 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

h_burden_2019 <-ggplot(data = acs.fmr.2019, aes(x=housing_percent, y=poverty_percent))+
  geom_point() +
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Percent Spent on Rent") + 
  ylab("Poverty Rate %") + 
  ggtitle("2019 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

h_burden_2020 <-ggplot(data = acs.fmr.2020, aes(x=housing_percent, y=poverty_percent))+
  geom_point() +
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Percent Spent on Rent") + 
  ylab("Poverty Rate %") + 
  ggtitle("2020 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

combined_h_burden <- cowplot::plot_grid(h_burden_2017, h_burden_2018,
                                        h_burden_2019, h_burden_2020) 

combined_h_burden # display the combined graph

```

# mapthegap data
```{r, eval=TRUE}
setwd("Map the Meal Gap data")

# map the gap county data 2010
food <- read.csv("MMG2012_2010Data_ToShare.csv", fileEncoding = "UTF-8-BOM") 

food %<>%
  dplyr::select(FIPS, State,County..State,X2010.Food.Insecurity.Rate,X2010.Cost.Per.Meal) %>%
  dplyr::rename(County = County..State, food_insecure = X2010.Food.Insecurity.Rate, 
         meal_cost = X2010.Cost.Per.Meal) %>%
  dplyr::mutate(ID = paste0("0", as.character(050000000000 + as.numeric(FIPS)))) 
  
food$meal_cost = as.numeric(gsub("[$]","",food$meal_cost))


```

```{r, eval=TRUE}
# join map the gap data to acs/fmr data
acs.fmr.food.2017 <- left_join(acs.fmr.2017, food[,c("ID","food_insecure", 
                                                     "meal_cost")], by = "ID") %>%
  dplyr::mutate(meal_percent = meal_cost*3*365/med_house_income)
acs.fmr.food.2018 <- left_join(acs.fmr.2018, food[,c("ID","food_insecure", 
                                                     "meal_cost")], by = "ID") %>%
  dplyr::mutate(meal_percent = meal_cost*3*365/med_house_income)
acs.fmr.food.2019 <- left_join(acs.fmr.2019, food[,c("ID","food_insecure", 
                                                     "meal_cost")], by = "ID") %>%
  dplyr::mutate(meal_percent = meal_cost*3*365/med_house_income)
acs.fmr.food.2020 <- left_join(acs.fmr.2020, food[,c("ID","food_insecure", 
                                                     "meal_cost")], by = "ID") %>%
  dplyr::mutate(meal_percent = meal_cost*3*365/med_house_income)


# look at food costs in isolation 
food_cost <- ggplot(acs.fmr.food.2017, aes(x=meal_cost, y=poverty_percent))+
  geom_point()+
  xlab("Meal Cost") +
  ylab("Poverty Rate %") +
  ggtitle("Meal Cost") +
  theme(plot.title = element_text(hjust = 0.5)) 


# look at food cost as percent of median income
food_burden <- ggplot(acs.fmr.food.2017, aes(x=meal_percent, y=poverty_percent))+
  geom_point()+  
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Percent Spent on Food") + 
  ylab("Poverty Rate %") + 
  ggtitle("Income spent on Food") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

combined_food_burden <- cowplot::plot_grid(food_cost, food_burden) 

combined_food_burden # display the combined graph


```

```{r, eval=TRUE}
# generate a two-factor cost of living index by adding food and housing costs

acs.fmr.food.2017 %<>%
  mutate(COL = meal_percent + housing_percent)

acs.fmr.food.2018 %<>%
  mutate(COL = meal_percent + housing_percent)

acs.fmr.food.2019 %<>%
  mutate(COL = meal_percent + housing_percent)

acs.fmr.food.2020 %<>%
  mutate(COL = meal_percent + housing_percent)

COL_2017 <- ggplot(acs.fmr.food.2017, aes(x=COL, y=poverty_percent))+
  geom_point()+  
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Combined food and housing costs") + 
  ylab("Poverty Rate %") + 
  ggtitle("2017 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

COL_2018 <- ggplot(acs.fmr.food.2018, aes(x=COL, y=poverty_percent))+
  geom_point()+  
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Combined food and housing costs") + 
  ylab("Poverty Rate %") + 
  ggtitle("2018 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

COL_2019 <- ggplot(acs.fmr.food.2019, aes(x=COL, y=poverty_percent))+
  geom_point()+  
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Combined food and housing costs") + 
  ylab("Poverty Rate %") + 
  ggtitle("2019 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

COL_2020 <- ggplot(acs.fmr.food.2020, aes(x=COL, y=poverty_percent))+
  geom_point()+  
  geom_smooth(method = "lm", formula = y~x) +
  xlab("Combined food and housing costs") + 
  ylab("Poverty Rate %") + 
  ggtitle("2020 US Counties ") +
  theme(plot.title = element_text(hjust = 0.5)) +
  stat_regline_equation() 

combined_COL <- cowplot::plot_grid(COL_2017,COL_2018,COL_2019,COL_2020) 

combined_COL # display the combined graph

```


Based on these analyses, poverty rate and income are directly correlated as expected. However, median housing cost and food expenses alone do not seem to be directly associated with the poverty rate. Furthermore, calculated variables such as percent-spent-on-housing and percent-spent-on-food do not seem to give better correlations to poverty rate compared to median income alone, From these results I will use housing costs, food costs, and either poverty rate or median income from aggreate county data to predict health outcomes from BRFSS data. 

```{r, eval=TRUE}
library(gtools)
library(plyr)

pov_quant_2017 <- acs.fmr.food.2017 %>%
  dplyr::mutate(pov_quant = quantcut(poverty_percent, q=4,by=0.25, na.rm=TRUE))

pov_meal_quart <- ggplot(data=pov_quant_2017, aes(x=med_house_income, 
                                                  y=meal_cost,
                                                  color = pov_quant))+
  geom_point() +
  xlab("Median Household Income ($)") +
  ylab("Meal Cost ($)") +
  ggtitle("US Counties 2017") +
  theme(plot.title = element_text(hjust = 0.5)) 

pov_meal_quart

pov_house_quart <- ggplot(data=pov_quant_2017, aes(x=med_house_income, 
                                                  y=fmr1,
                                                  color = pov_quant))+
  geom_point() +
  xlab("Median Household Income ($)") +
  ylab("Median Rent ($)")  +
  ggtitle("US Counties 2017") +
  theme(plot.title = element_text(hjust = 0.5)) 

pov_house_quart

```

# Get BRFSS data
```{r, eval=TRUE}
library(foreign)
setwd("MMSA2017_XPT")
brfss2017_raw <- read.xport("MMSA2017.xpt")
```

```{r, eval=TRUE}
# select variables of interest
# clean BRFSS 2017 dataframe

brfss2017 <- brfss2017_raw %>%
  dplyr::select('MMSANAME', 'X_MMSA', 'X_RACE', 'SEX', 'X_AGEG5YR','EDUCA','INCOME2',
                'ADDEPEV2', 'CHCCOPD1', 'X_RFHLTH') %>%
  dplyr::rename('Metro Area'='MMSANAME','Metro_ID'='X_MMSA','age'='X_AGEG5YR','race'='X_RACE', 
         'sex'='SEX', 'education'='EDUCA','income'='INCOME2',
         'depression'='ADDEPEV2','copd'='CHCCOPD1', 'health'='X_RFHLTH') %>%
  dplyr::filter(race %in%  c(1,2,3,5,8)) %>%
  dplyr::filter(sex %in% c(1,2)) %>%
  dplyr::filter(age %in% seq(13)) %>%
  dplyr::filter(education %in%  c(1,2,3,4,5,6)) %>%
  dplyr::filter(income %in% c(1,2,3,4,5,6,7,8)) %>%
  dplyr::filter(depression %in%  c(1,2)) %>%
  dplyr::filter(copd %in% c(1,2)) %>%
  dplyr::filter(health %in% c(1,2))

brfss2017 %<>%
  dplyr::mutate('Metro_ID' = as.character(brfss2017$'Metro_ID')) %>%
  dplyr::mutate(race = factor(race, levels = c(1,2,3,5,8), 
                        labels = c( "White", "Black", "American Indian", 
                                    "Asian", "Hispanic"))) %>%
  dplyr::mutate(sex = factor(sex, levels = c(1,2), 
                        labels = c("Male", "Female"))) %>%
  
  dplyr::mutate(age = factor(age, levels = seq(13), 
                        labels = c("18-24","25-29","30-34","35-39","40-44","45-49","50-54",
                                   "55-59","60-64","65-69","70-74","75-29",">79"))) %>%
  dplyr::mutate(education = factor(education, levels = c(1,2,3,4,5,6), 
                        labels = c("None","Elementary", "Some HS", "HS",
                                   "Some College", "College"))) %>%
  dplyr::mutate(income = factor(income, levels = c(1,2,3,4,5,6,7,8), 
                        labels = c("10k","15k", "20k", "25k", "35k", "50k", "75k","> 75k"))) %>%
  dplyr::mutate(depression = factor(depression, levels = c(1,2), 
                        labels = c("yes", "no"))) %>%
  dplyr::mutate(copd = factor(copd, levels = c(1,2), 
                        labels = c("yes", "no"))) %>%
  dplyr::mutate(health = factor(health, levels = c(1,2), 
                        labels = c("good", "poor"))) 

```


Let's look at the summary of the clean 2017 data frame. I am interested in sex, age, income, education, and race as demographic factors. I am interested in previous depression, COPD, and general health as outcomes. From the summary, there is a bias in sex and age. This might be explained by the fact that BRFSS is a phone based survey. It seems that income and education is skewed twoards highly educated, high earning individuals as well. The BRFSS might be biased towards home-owners and landline users over cell-phone users.
```{r, eval=TRUE}
summary(brfss2017, maxsum = 15)

```

# Data Wrangling to apply an appropriate metro code to every county

```{r, eval=TRUE}
cbsa2fips <- read.csv("cbsa2fipsxw.csv") 

cbsa2fips %<>%
  dplyr::mutate(fipsstatecode = as.character(cbsa2fips$fipsstatecode)) %>%
  dplyr::mutate(fipscountycode = as.character(cbsa2fips$fipscountycode)) %>%
  dplyr::mutate(fips_state = ifelse(nchar(fipsstatecode) == 1, paste0("0",fipsstatecode),
         fipsstatecode)) %>%
  dplyr::mutate(fips_county = ifelse(nchar(fipscountycode) == 2, paste0("0",fipscountycode),
         ifelse(nchar(fipscountycode) == 1, paste0("00",fipscountycode), fipscountycode))) %>%
  dplyr::mutate(ID = paste0("0", as.character(050000000000 + as.numeric(paste0(fips_state,fips_county))))) %>%
  dplyr::rename('Metro_ID' = cbsacode) %>%
  dplyr::select(ID, 'Metro_ID')

cbsa2fips$`Metro_ID` = as.character(cbsa2fips$`Metro_ID`)

# associate all counties to given metro area
acs.fmr.food.2017 %<>% 
  left_join(cbsa2fips, by="ID")

```

```{r, eval=TRUE}
# function to aggregate all acs.fmr.food cost of living data by metro area
# returns mean for all counties in a given metro area
metro_costs <- function(df){
  df %<>%
    dplyr::select(fmr0, fmr1, fmr2, fmr3, fmr4, Metro_ID, poverty_percent,
                  housing_percent, meal_cost, meal_percent, med_house_income)
  
  df = as.data.frame(df) 
  
  df %<>%
    dplyr::select(-geometry) %>%
    group_by(Metro_ID) %>%
    summarise_at(vars('fmr0', 'fmr1', 'fmr2', 'fmr3', 'fmr4', 'poverty_percent',
                  'housing_percent', 'meal_cost', 'meal_percent',
                  'med_house_income'), mean)
  return(df)
}

col_2017 <- metro_costs(acs.fmr.food.2017)

# join 2017 metro cost of living to BRFSS2017 dataframe
brfss2017 %<>%
  left_join(col_2017, by = "Metro_ID")

# remove NAs
brfss2017 %<>%
  na.omit(brfss2017)
```

# Look at how aggregate costs/income data and individual income data affect health outcomes

```{r, eval=TRUE}
metro_income_ghealth <- ggplot(data=brfss2017, aes(x=health, y = med_house_income))+
  geom_boxplot() +
  xlab("General Health") +
  ylab("Median Household Income ($)")  +
  ggtitle("BRFSS2017 General Health") +
  theme(plot.title = element_text(hjust = 0.5)) 


metro_income_copd <- ggplot(data=brfss2017, aes(x=copd, y = med_house_income))+
  geom_boxplot() +
  xlab("COPD History") +
  ylab("Median Household Income ($)")  +
  ggtitle("BRFSS2017 COPD History") +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_income_depression <- ggplot(data=brfss2017, aes(x=depression, y = med_house_income))+
  geom_boxplot() +
  xlab("Depression History") +
  ylab("Median Household Income ($)")  +
  ggtitle("BRFSS2017 Depression") +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_income_health_outcomes <- cowplot::plot_grid(metro_income_ghealth,metro_income_copd,
                                                   metro_income_depression) 

metro_income_health_outcomes
```

# Rent effects on health outcomes
```{r, eval=TRUE}
metro_fmr1_ghealth <- ggplot(data=brfss2017, aes(x=health, y = fmr1))+
  geom_boxplot() +
  xlab("General Health") +
  ylab("Median 1-Bed Rent ($)")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_fmr1_copd <- ggplot(data=brfss2017, aes(x=copd, y = fmr1))+
  geom_boxplot() +
  xlab("COPD History") +
  ylab("Median 1-Bed Rent ($)")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_fmr1_depression <- ggplot(data=brfss2017, aes(x=depression, y = fmr1))+
  geom_boxplot() +
  xlab("Depression History") +
  ylab("Median 1-Bed Rent ($)")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_rent_health_outcomes <- cowplot::plot_grid(metro_fmr1_ghealth,metro_fmr1_copd,
                                                   metro_fmr1_depression) 

metro_rent_health_outcomes
```

# aggregate food expenses and health outcomes
```{r, eval=TRUE}
metro_food_ghealth <- ggplot(data=brfss2017, aes(x=health, y = meal_cost))+
  geom_boxplot() +
  xlab("General Health") +
  ylab("Cost of Meal ($)")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_food_copd <- ggplot(data=brfss2017, aes(x=copd, y = meal_cost))+
  geom_boxplot() +
  xlab("COPD History") +
  ylab("Cost of Meal ($)")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_food_depression <- ggplot(data=brfss2017, aes(x=depression, y = meal_cost))+
  geom_boxplot() +
  xlab("Depression History") +
  ylab("Cost of Meal ($)")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_food_health_outcomes <- cowplot::plot_grid(metro_food_ghealth,metro_food_copd,
                                                   metro_food_depression) 

metro_food_health_outcomes

```

# Aggregate poverty effect on health outcomes
```{r, eval=TRUE}
metro_pov_ghealth <- ggplot(data=brfss2017, aes(x=health, y = poverty_percent))+
  geom_boxplot() +
  xlab("General Health") +
  ylab("Poverty Rate")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_pov_copd <- ggplot(data=brfss2017, aes(x=copd, y = poverty_percent))+
  geom_boxplot() +
  xlab("COPD History") +
  ylab("Poverty Rate")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_pov_depression <- ggplot(data=brfss2017, aes(x=depression, 
                                                    y= poverty_percent))+
  geom_boxplot() +
  xlab("Depression History") +
  ylab("Poverty Rate")  +
  theme(plot.title = element_text(hjust = 0.5)) 

metro_pov_health_outcomes <- cowplot::plot_grid(metro_pov_ghealth,metro_pov_copd,
                                                   metro_pov_depression) 

metro_pov_health_outcomes

```


# Individual income associations with health visualizations
```{r, eval=TRUE}

# income and general health 
ggplot(data=brfss2017, aes(x=income,  fill=health)) + 
  geom_bar(position="dodge") +
  ggtitle('Individual Income and General Health')+
  xlab('Income')+
  ylab('Count')  +
  theme(plot.title = element_text(hjust = 0.5)) 

# income and depression
ggplot(data=brfss2017, aes(x=income,  fill=depression)) + 
  geom_bar(position="dodge") +
  ggtitle('Individual Income and Depression History')+
  xlab('Income')+
  ylab('Count')  +
  theme(plot.title = element_text(hjust = 0.5)) 

# income and COPD
ggplot(data=brfss2017, aes(x=income,  fill=copd)) + 
  geom_bar(position="dodge") +
  ggtitle('Individual Income and COPD History')+
  xlab('Income')+
  ylab('Count')  +
  theme(plot.title = element_text(hjust = 0.5)) 
```


# Now we will run chi-square tests to see if income is indepdent from chosen 
health outcomes.

```{r, eval=TRUE}

# chi-square test for individual income and general health
brfss2017_health_table <- table(brfss2017$health, brfss2017$income)
brfss2017_health_table
chisq.test(brfss2017_health_table)

# chi-square test for individual income and depression history
brfss2017_depression_table <- table(brfss2017$depression, brfss2017$income)
brfss2017_depression_table
chisq.test(brfss2017_depression_table)

# chi-square test for individual income and COPD history
brfss2017_COPD_table <- table(brfss2017$copd, brfss2017$income)
brfss2017_COPD_table
chisq.test(brfss2017_COPD_table)

```

# Based on visualizaitons and chi-squared tests, it seems that there is an association between income and general health and history of both COPD and depression 




# Exploring the health outcomes (depression, COPD, general health) using all individual demographic factors only by logistic regression.

```{r, eval=TRUE}

# Train a logistic regression model using BRFSS2017 data

depression_glm <- glm(formula=depression ~ race + sex + age + education + income,
                      data=brfss2017, family = binomial)

health_glm <- glm(formula=health ~ race + sex + age + education + income, 
                  data=brfss2017, family = binomial(logit))

copd_glm <- glm(formula=copd ~ race + sex + age + education + income, 
                  data=brfss2017, family = binomial(logit))

```


# Check model accuracy using individual demographics alone

```{r, eval=TRUE}
library(PRROC)

# ROC for depression logistic regression
depression_pred <- predict(depression_glm, data=brfss2017, type="response")

depression_glm_roc <- roc.curve(depression_pred[brfss2017$depression == "no"],
                                depression_pred[brfss2017$depression == "yes"], curve=TRUE)

# ROC for general health logistic regression
health_pred <- predict(health_glm, data=brfss2017, type="response")
health_glm_roc <- roc.curve(health_pred[brfss2017$health == "poor"],
                                health_pred[brfss2017$health == "good"], curve=TRUE)

# ROC for COPD logistic regression
copd_pred <- predict(copd_glm, data=brfss2017, type="response")
copd_glm_roc <- roc.curve(copd_pred[brfss2017$copd == "no"],
                                copd_pred[brfss2017$copd == "yes"], curve=TRUE)

plot(depression_glm_roc)
plot(health_glm_roc)
plot(copd_glm_roc)

```

# Exploring the health outcomes (depression, COPD, general health) using both individual demographic factors and aggregate metropolitan factors by logistic regression.

```{r, eval=TRUE}

# Train a logistic regression model using BRFSS2017 data

depression_glm2 <- glm(formula=depression ~ race + sex + age + education + income +
                         med_house_income + fmr1 + meal_cost,
                      data=brfss2017, family = binomial(logit))


health_glm2 <- glm(formula=health ~ race + sex + age + education + income + med_house_income
                      + fmr1 + meal_cost,
                  data=brfss2017, family = binomial(logit))

copd_glm2 <- glm(formula=copd ~ race + sex + age + education + income + med_house_income
                      + fmr1 + meal_cost, 
                  data=brfss2017, family = binomial(logit))


```

# check model accuracy using individual demographics and aggregate metro data

```{r, eval=TRUE}
# ROC for depression logistic regression
depression_pred2 <- predict(depression_glm2, data=brfss2017, type="response")

depression_glm_roc2 <- roc.curve(depression_pred2[brfss2017$depression == "no"],
                                depression_pred2[brfss2017$depression == "yes"], curve=TRUE)
head(depression_pred2)
# ROC for general health logistic regression
health_pred2 <- predict(health_glm2, data=brfss2017, type="response")
health_glm_roc2 <- roc.curve(health_pred2[brfss2017$health == "poor"],
                                health_pred2[brfss2017$health == "good"], curve=TRUE)

# ROC for COPD logistic regression
copd_pred2 <- predict(copd_glm2, data=brfss2017, type="response")
copd_glm_roc2 <- roc.curve(copd_pred2[brfss2017$copd == "no"],
                                copd_pred2[brfss2017$copd == "yes"], curve=TRUE)

plot(depression_glm_roc2)
plot(health_glm_roc2)
plot(copd_glm_roc2)
```

Adding in the additional variables from ACS median income an housing costs only slightly increased AUC, but not enough to merit including them into GLM models. Overall, the value of individual survey data is much more valuable in predicting history of depression, COPD, and general well-being/health.


