---
title: "Another comparison of two area-level socioeconomic deprivation indices: SVI  (Social Vulnerability Index) and ADI (Area Deprivation Index) and their associations with asthma and smoking outcomes from survey data"
subtitle: "BMIN503/EPID600 Final Project"
author: "Amy Goodwin Davies"
format: 
  html:
    toc: true
    toc-depth: 6
    theme: minty
editor: visual
number-sections: true
embed-resources: true
execute: 
  warning: false
---

------------------------------------------------------------------------

```{r packages}
library(haven)
library(tidyverse)
library(readxl)
library(tigris)
library(sf)
library(leaflet)
library(gridExtra)
library(RColorBrewer)
options(tigris_use_cache = TRUE)
options(progress_enabled = FALSE)
```

```{r functions}

```

<!-- Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers. Feel free to change the theme and other display settings, although this is not required. -->

## Overview {#sec-overview}

The goal of this project is to compare the Agency for Toxic Substances and Disease Registry Social Vulnerability Index (SVI) and the University of Wisconsin-developed Area Deprivation Index (ADI) in their association with asthma and smoking outcomes from the Behavioral Risk Factor Surveillance System (BRFSS) CDC dataset.

For SVI, the most recent available dataset is from 2020. For this project, national ranking at the county level will be used. The most recent ADI dataset is from 2021. National ranking for ADI is provided at the census block group level. For the asthma and smoking outcomes in the BRFSS CDC dataset, the 2021 Selected Metropolitan/Micropolitan Area Risk Trends (SMART) subset will used. The SMART BRFSS dataset provides the CBSA (Core-based statistical area), which "consist of the county or counties (or equivalent entities) associated with at least one core (urban area) of at least 10,000 population, plus adjacent counties having a high degree of social and economic integration with the core as measured through commuting ties." <https://www.census.gov/programs-surveys/metro-micro/about/glossary.html>

<!-- Give a brief a description of your project and its goal(s), what data you are using to complete it, and what two faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository. -->

## Introduction {#sec-introduction}

As Rollings et al 2023 discuss, there exist several area-level socioeconomic deprivation indices without consensus within healthcare research and policy fields about which indices should be used for a variety of analyses. In comparing two frequently used indices, SVI and ADI, in their associations with SMART BRFSS dataset, this project has the potential to inform selection of the appropriate index for future work. This project is interdisciplinary in that it spans data science, social science, and biomedical informatics.

A challenge for this project, as with many projects which incorporate geospatial variables, is differing geographic units. For this project, SVI is at the county level, ADI is at the census block group level, and the BRFSS outcomes are at the CBSA level. Fortunately, all of these geographic units are census-defined and are supersets or subsets of each other. The current plan is to link these datasets at the county level. Based on guidance from Dr. Sherrie Xie, ADI will be aggregated to the county-level using a weighted average that incorporates census counts for census block groups. The BRFSS smoking and asthma outcomes will be aggregated at the CBSA level. CBSAs and CBSA level smoking and asthma outcomes will be mapped to counties. An alternative approach uses weighted averages to aggregate both SVI and ADI to the CBSA level, which can also be explored.

Once the linked dataset is prepared, logistic regression models will be used to compare SVI and ADI as predictors of the aggregated smoking and asthma outcomes. Spatial autoregressive regression, as used by Tipirneni et al 2022 will also be considered. For smoking, the smoking status variable `_SMOKER3` will be aggregated at the CBSA level. For asthma, the variable for lifetime asthma prevalence `_LTASTH1` will be aggregated at the CBSA level. More information about the smoking and asthma variables available in the SMART BRFSS dataset is provided in @sec-app.

There are several limitations of the project. One clear limitation is that the analyses will be restricted to areas of the US which are included in a CBSA grouping included in the SMART BRFSS dataset.

<!-- Describe the problem addressed, its significance, and some background to motivate the problem. This should extend what is in the @sec-overview. -->

<!-- Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff. -->

## Methods {#sec-methods}

<!-- Describe the data used and general methodological approach used to address the problem described in the @sec-introduction. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. -->

### SMART BRFSS CDC dataset

```{r loadmmsa}
# MMSA2021.xpt retrieved from https://www.cdc.gov/brfss/smart/smart_2021.html
# Accompanying documentation available at https://www.cdc.gov/brfss/annual_data/2021/pdf/2021_smart_brfss_mmsa_methodology-508.pdf
MMSA2021 <- read_xpt("input_data/mmsa/MMSA2021_XPT/MMSA2021.xpt")

MMSA2021 |> glimpse()
```

```{r mmsalabels}
# construct table of variable names and labels
mmsa_colname_labels <- MMSA2021 %>%
  map_dfc(attr, "label") |>
  pivot_longer(everything(),
               names_to = "colname",
               values_to = "label")
# include variable names without labels
mmsa_colnames_all <- MMSA2021 |>
  colnames() |>
  as_tibble() |>
  rename(colname = value) |>
  left_join(mmsa_colname_labels, by = join_by(colname))
# add missing labels documented in https://www.cdc.gov/brfss/annual_data/2021/pdf/2021_smart_brfss_mmsa_methodology-508.pdf
attr(MMSA2021[["_MMSA"]], "label") <- toupper("the code of metropolitan or micropolitan statistical area where the respondent lives")
attr(MMSA2021[["MMSANAME"]], "label") <- toupper("the MMSA name")
attr(MMSA2021[["_MMSAWT"]], "label") <- toupper(" the MMSA-level weight that is used when generating MMSA-level estimates for variables in the data set")
mmsa_colnames_all <- MMSA2021 %>%
  map_dfc(attr, "label") |>
  pivot_longer(everything(),
               names_to = "colname",
               values_to = "label")
```
```{r}
# rename columns to make them easier to work with using R (not requiring ``)
# underscore prefix indicates derived or computed, replace with d_
smart_2021 <- MMSA2021 |> 
  rename_with(~ tolower(gsub("_", "d_", .x, fixed = TRUE)))
smart_2021 |> View()
```

```{r mmsatocounty}
# https://www.cdc.gov/brfss/annual_data/2021/pdf/2021_smart_brfss_mmsa_methodology-508.pdf
# Typically, BRFSS data are used to produce state-level estimates; however, for the SMART
# project, BRFSS data were used to produce small area-level estimates for MMSAs as defined by
# the US Census Bureau. On June 6, 2003, OMB issued new definitions for MMSA and
# metropolitan divisions. OMB periodically updates the list of MMSAs. The list of areas used for this
# analysis can be found here: https://www.census.gov/geographies/reference-files/timeseries/demo/metro-micro/delineation-files.html. The March 2020 release was used for defining
# the MMSAs in the 2021 SMART data set.
# retrieved from https://www.census.gov/geographies/reference-files/time-series/demo/metro-micro/historical-delineation-files.html
delin_2020 <- read_xls("input_data/mmsa/list1_2020.xls",
                       skip = 2) |>
  rename_with(~ tolower(gsub(" ", "_", .x, fixed = TRUE))) |> 
  rename_with(~ tolower(gsub("/", "_", .x, fixed = TRUE)))


delin_2020_cbsa_join <- delin_2020 |> filter(is.na(metropolitan_division_code)) # need to explore why this is necessary
delin_2020_mdc_join <- delin_2020 |> filter(!is.na(metropolitan_division_code)) # need to explore why this is necessary

# TODO check this
mmsa_county_mapping_cbsa <- smart_2021 |> 
  mutate(d_mmsa_char = as.character(d_mmsa)) |>
  distinct(d_mmsa, d_mmsa_char, d_mmsa_char, mmsaname) |> 
  left_join(delin_2020_cbsa_join, by = c("d_mmsa_char" = "cbsa_code"))  |> 
  distinct(d_mmsa, d_mmsa_char, csa_title, county_county_equivalent, fips_state_code, fips_county_code)
mmsa_county_mapping_mdc <- smart_2021 |> 
  mutate(d_mmsa_char = as.character(d_mmsa)) |>
  distinct(d_mmsa, d_mmsa_char, d_mmsa_char, mmsaname) |> 
  left_join(delin_2020_mdc_join, by = c("d_mmsa_char" = "metropolitan_division_code"))  |> 
  distinct(d_mmsa, d_mmsa_char, csa_title, county_county_equivalent, fips_state_code, fips_county_code)
mmsa_county_mapping <- mmsa_county_mapping_cbsa |> 
  dplyr::union(mmsa_county_mapping_mdc) |> 
  distinct()


# smart_2021 |> summarize(n_distinct(d_mmsa)) |> pull() == mmsa_county_mapping |> summarize(n_distinct(d_mmsa)) |> pull()

# check for issues with leading zeros
# smart_2021 |> 
#   filter(nchar(d_mmsa) != 5)
# smart_2021 |> 
#   mutate(d_mmsa_char = as.character(d_mmsa)) |> 
#   filter(nchar(d_mmsa_char) != 5)
# mmsa_county_mapping |> 
#   filter(nchar(d_mmsa_char) != 5)
# delin_2020 |> 
#   filter(nchar(cbsa_code) != 5)
```

```{r plotmmsacounties}
tigris_counties <- counties() |>
  filter(!STATEFP %in% c("02",
                         "15", "60",
                         "66", "69",
                         "72", "78"))

mmsa_county_mapping_geoid <- mmsa_county_mapping |>
  mutate(smart = TRUE) |> 
  right_join(tigris_counties, by = c("fips_state_code"= "STATEFP", "fips_county_code" = "COUNTYFP")) |>
  mutate(smart = replace_na(smart, FALSE))

mmsa_county_mapping_geoid |> filter(smart == TRUE) |> nrow() == mmsa_county_mapping |> nrow()

mmsa_county_mapping_geoid_sf <- mmsa_county_mapping_geoid |> st_as_sf()

# simple map
# mmsa_county_mapping_geoid_sf |>
#   st_simplify(dTolerance = 1e3) |>
#   ggplot(aes(fill = smart)) +
#   geom_sf()
```

```{r convertsmart}
# Before we create the leaflet maps, we will change the CRS (Coordinate Reference System) to 4326 because leaflet expects the data provided to it to be in WGS84, which is EPSG:4326. We can do the transformation with st_transform.
# This is slow so I did it once and saved the RDS

# mmsa_county_mapping_geoid_sf_4326 <- st_transform(mmsa_county_mapping_geoid_sf, crs = 4326)
# mmsa_county_mapping_geoid_sf_4326 |> saveRDS("mmsa_county_mapping_geoid_sf_4326.rds")
mmsa_county_mapping_geoid_sf_4326 <- readRDS("mmsa_county_mapping_geoid_sf_4326.rds")
mmsa_county_mapping_geoid_sf_4326
```


```{r smartleaflet}
pal_fun <- colorFactor(topo.colors(2), mmsa_county_mapping_geoid_sf_4326$smart)

mmsa_county_mapping_geoid_sf_4326  |>
  filter(!is.na(FUNCSTAT)) |> # need to explore why this is necessary 
  st_simplify(dTolerance = 1e3)  |>
  leaflet() |>
  addPolygons(
    stroke = FALSE,
    # remove polygon borders
    fillColor = ~ pal_fun(smart),
    fillOpacity = 0.5,
    smoothFactor = 0.5,
    # increase opacity and resolution
    popup = mmsa_county_mapping_geoid_sf_4326$county_county_equivalent
  ) |>
  addProviderTiles(providers$CartoDB.Voyager) |> # add third party provider tile
  addLegend(
    "bottomright",
    # location of legend
    pal = pal_fun,
    # palette function
    values = ~ smart,
    # variable to pass to palette function
    title = 'Included in SMART BRFSS dataset',
    # legend title
    opacity = 1 # legend opacity (1 = completely opaque)
  ) |>
  addScaleBar()
```

### SVI dataset

```{r loadsvi}
# retrieved from https://www.atsdr.cdc.gov/placeandhealth/svi/data_documentation_download.html
svi2020 <- read_csv("input_data/svi/SVI_2020_US_county.csv")
svi2020 |> glimpse()
```

```{r mapsvi}
svi2020_sf <- svi2020 |>
  select(RPL_THEME1,
         RPL_THEME2,
         RPL_THEME3,
         RPL_THEME4,
         RPL_THEMES,
         FIPS) |>
  left_join(tigris_counties, by = c("FIPS" = "GEOID")) |>
  st_as_sf()

# svi2020_sf_4326 <- st_transform(svi2020_sf, crs = 4326)
# svi2020_sf_4326 |> saveRDS("svi2020_sf_4326.rds")
svi2020_sf_4326 <- readRDS("svi2020_sf_4326.rds")
svi2020_sf_4326
```

```{r leafletsvi}
pal_svi <- colorNumeric("inferno", NULL, reverse = TRUE)

svi2020_sf_4326  |>
  filter(!is.na(FUNCSTAT)) |> # need to explore why this is necessary
  st_simplify(dTolerance = 1e3)  |>
  leaflet() |>
  addPolygons(
    stroke = FALSE,
    # remove polygon borders
    fillColor = ~ pal_svi(RPL_THEME1),
    fillOpacity = 0.5,
    smoothFactor = 0.5
    # increase opacity and resolution
  ) |>
  addProviderTiles(providers$CartoDB.Voyager) |> # add third party provider tile
  addLegend(
    "bottomright",
    # location of legend
    pal = pal_svi,
    # palette function
    values = ~ RPL_THEMES,
    # variable to pass to palette function
    title = 'SVI',
    # legend title
    opacity = 1 # legend opacity (1 = completely opaque)
  ) |>
  addScaleBar()
```



### ADI dataset

```{r loadadi}
# retrieved from https://www.neighborhoodatlas.medicine.wisc.edu/
adi2021 <- read_csv("input_data/adi/adi-download/US_2021_ADI_Census_Block_Group_v4_0_1.csv")
adi2021 |> glimpse()
```

```{r mapadi}
tigris_cbg <- block_groups(cb = TRUE)

adi2021_sf <- adi2021 |>
  left_join(tigris_cbg, by = c("FIPS" = "GEOID")) |>
  st_as_sf()

# adi2021_sf_4326 <- st_transform(adi2021_sf, crs = 4326)
# adi2021_sf_4326 |> saveRDS("adi2021_sf_4326.rds")
adi2021_sf_4326 <- readRDS("adi2021_sf_4326.rds")
adi2021_sf_4326


# simple map
# adi2021_sf_4326 |>
#   mutate(ADI_NATRANK = as.numeric(ADI_NATRANK)) |>
#   filter(!is.na(ADI_NATRANK)) |>
#   st_simplify(dTolerance = 1e3) |>
#   ggplot(aes(fill = ADI_NATRANK)) +
#   geom_sf()
```

```{r leafletadi}
# pal_adi <- colorNumeric("plasma", NULL)
# 
# adi2021_sf_4326  |>
#   head(100) |> 
#   mutate(ADI_NATRANK = as.numeric(ADI_NATRANK)) |>
#   filter(!is.na(ADI_NATRANK)) |> 
#   st_simplify(dTolerance = 1e3)  |>
#   leaflet() |>
#   addPolygons(
#     stroke = FALSE,
#     # remove polygon borders
#     fillColor = ~ pal_adi(ADI_NATRANK),
#     fillOpacity = 0.5,
#     smoothFactor = 0.5
#     # increase opacity and resolution
#   ) |>
#   addProviderTiles(providers$CartoDB.Voyager) |> # add third party provider tile
#   addLegend(
#     "bottomright",
#     # location of legend
#     pal = pal_adi,
#     # palette function
#     values = ~ ADI_NATRANK,
#     # variable to pass to palette function
#     title = 'adi',
#     # legend title
#     opacity = 1 # legend opacity (1 = completely opaque)
#   ) |>
#   addScaleBar()
# ```

### Combine datasets



## Results {#sec-results}

<!-- Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required. -->

## Conclusion

<!-- This the conclusion. The @sec-results can be invoked if you'd like. -->

## References

<!-- Need to figure out how to properly manage references -->

*Rollings KA, Noppert GA, Griggs JJ, Melendez RA, Clarke PJ. Comparison of two area-level socioeconomic deprivation indices: Implications for public health research, practice, and policy. PLoS One. 2023 Oct 5;18(10):e0292281. doi: 10.1371/journal.pone.0292281. PMID: 37797080; PMCID: PMC10553799.*

*Tipirneni R, Schmidt H, Lantz PM, Karmakar M. Associations of 4 Geographic Social Vulnerability Indices With US COVID-19 Incidence and Mortality. Am J Public Health. 2022 Nov;112(11):1584-1588. doi: 10.2105/AJPH.2022.307018. Epub 2022 Sep 15. PMID: 36108250; PMCID: PMC9558191.*

*Xie S, Himes BE. Approaches to Link Geospatially Varying Social, Economic, and Environmental Factors with Electronic Health Record Data to Better Understand Asthma Exacerbations. AMIA Annu Symp Proc. 2018 Dec 5;2018:1561-1570. PMID: 30815202; PMCID: PMC6371292.*

*Xie S, Hubbard RA, Himes BE. Neighborhood-level measures of socioeconomic status are more correlated with individual-level measures in urban areas compared with less urban areas. Ann Epidemiol. 2020 Mar;43:37-43.e4. doi: 10.1016/j.annepidem.2020.01.012. Epub 2020 Feb 11. PMID: 32151518; PMCID: PMC7160852.*

## Appendices {#sec-app}

### Asthma and smoking BRFSS variables

The table below summarizes the variables of interest, combining information from <https://www.cdc.gov/brfss/annual_data/2021/summary_matrix_21.html> and <https://www.cdc.gov/brfss/annual_data/2021/pdf/2021-calculated-variables-version4-508.pdf>

+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Output Variables\   | **Description or Result\                                                                                                                                                          | **Values**                                                                                                                                                                                                                                                                                                                                                                                                                                          |
| (In Final Data Set)** | of Calculation**                                                                                                                                                                  |                                                                                                                                                                                                                                                                                                                                                                                                                                                     |
+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| \_LTASTH1             | CV for lifetime asthma prevalence / Calculated variable for adults who have ever been told they have asthma.                                                                      | +--------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                                                                                                       |
|                       |                                                                                                                                                                                   | | **1**  | No                                     | *Respondents who have not been told by a doctor, nurse, or health professional that they had asthma.*                                                                                                                                                                                   |                                                                                                       |
|                       | \_LTASTH1 is derived from ASTHMA3.                                                                                                                                                | +--------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                                                                                                       |
|                       |                                                                                                                                                                                   | | **2**  | Yes                                    | *Respondents who have been told by a doctor, nurse, or health professional that they had asthma.*                                                                                                                                                                                       |                                                                                                       |
|                       |                                                                                                                                                                                   | +--------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                                                                                                       |
|                       |                                                                                                                                                                                   | | **9**  | Don't know/Not Sure or Refused/Missing | *Respondents who reported they did not know if they had been told by a doctor, nurse, or health professional that they had asthma, those who refused to answer if they had been told by a doctor, nurse. or health professional that they had asthma, or those with missing responses.* |                                                                                                       |
|                       |                                                                                                                                                                                   | +--------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                                                                                                       |
+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| \_CASTHM1             | CV for current asthma prevalence / Calculated variable for adults who have been told they currently have asthma.                                                                  | +-------+----------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+ |
|                       |                                                                                                                                                                                   | | **1** | No                                     | *Respondents who have not been told by a doctor, nurse, or health professional that they had asthma or do not still have asthma.*                                                                                                                                                                                                                                                              | |
|                       | \_CASTHM1 is derived from ASTHMA3 and ASTHNOW.                                                                                                                                    | +-------+----------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+ |
|                       |                                                                                                                                                                                   | | **2** | Yes                                    | *Respondents who have been told by a doctor, nurse, or health professional that they had asthma and that they still have asthma.*                                                                                                                                                                                                                                                              | |
|                       |                                                                                                                                                                                   | +-------+----------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+ |
|                       |                                                                                                                                                                                   | | **9** | Don't know/Not Sure or Refused/Missing | *Respondents who reported they did not know if they had been told by a doctor, nurse, or health professional that they had asthma, those who refused to answer if they had been told by a doctor, nurse, or health professional that they had asthma, those who did not know if they still had asthma, those who refused to answer if they still had asthma, or those with missing responses.* | |
|                       |                                                                                                                                                                                   | +-------+----------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+ |
+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| \_ASTHMS1             | Computed asthma status / Calculated variable for computed asthma status.                                                                                                          | +-------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+  |
|                       |                                                                                                                                                                                   | | **1** | Current                                | *Respondents who have been told by a doctor, nurse, or health professional that they had asthma and that they still have asthma.*                                                                                                                                                                                                                                                             |  |
|                       | \_ASTHMS1 is derived from ASTHMA3 and ASTHNOW                                                                                                                                     | +-------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+  |
|                       |                                                                                                                                                                                   | | **2** | Former                                 | *Respondents who have been told by a doctor, nurse, or health professional that they had asthma but do not still have asthma.*                                                                                                                                                                                                                                                                |  |
|                       |                                                                                                                                                                                   | +-------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+  |
|                       |                                                                                                                                                                                   | | **3** | Never                                  | *Respondents who have not been told by a doctor, nurse, or health professional that they had asthma.*                                                                                                                                                                                                                                                                                         |  |
|                       |                                                                                                                                                                                   | +-------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+  |
|                       |                                                                                                                                                                                   | | **9** | Don't know/Not Sure or Refused/Missing | *Respondents who reported they didn't know if they had been told by a doctor, nurse, or health professional that they had asthma, those who refused to answer if they had been told by a doctor, nurse, or health professional that they had asthma, those who didn't know if they still had asthma, those that refused to answer if they still had asthma, or those with missing responses.* |  |
|                       |                                                                                                                                                                                   | +-------+----------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+  |
+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| \_SMOKER3             | Smoking Status: Everyday smoker, someday smoker, former smoker and non smoker / Calculated variable for four-level smoker status: everyday smoker, someday smoker, former smoker, | +-------+--------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+               |
|                       |                                                                                                                                                                                   | | **1** | Current smoker--now smokes every day | *Respondents who reported having smoked at least 100 cigarettes in their lifetime and now smoke every day.*                                                                                                                                                                                                                                                                        |               |
|                       | non-smoker. \_SMOKER3 is derived from SMOKE100 and SMOKDAY2.                                                                                                                      | +-------+--------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+               |
|                       |                                                                                                                                                                                   | | **2** | Current smoker--now smokes some days | *Respondents who reported having smoked at least 100 cigarettes in their lifetime and now smoke some days.*                                                                                                                                                                                                                                                                        |               |
|                       |                                                                                                                                                                                   | +-------+--------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+               |
|                       |                                                                                                                                                                                   | | **3** | Former smoker                        | *Respondents who reported having smoked at least 100 cigarettes in their lifetime and currently do not smoke.*                                                                                                                                                                                                                                                                     |               |
|                       |                                                                                                                                                                                   | +-------+--------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+               |
|                       |                                                                                                                                                                                   | | **4** | Never smoked                         | *Respondents who reported they had not smoked at least 100 cigarettes in their lifetime.*                                                                                                                                                                                                                                                                                          |               |
|                       |                                                                                                                                                                                   | +-------+--------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+               |
|                       |                                                                                                                                                                                   | | **9** | Don't know/Refused/Missing           | *Respondents who reported they didn't know if they had smoked 100 cigarettes in their lifetime, those who refused to answer if they had smoked 100 cigarettes in their lifetime, those who didn't know if they now smoked every day, some days or not at all, those who refused to answer if they now smoked every day, some days or not at all, or those with missing responses.* |               |
|                       |                                                                                                                                                                                   | +-------+--------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+               |
+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| \_RFSMOK3             | CV for current smoking status / \_RFSMOK3 Calculated variable for adults who are current smokers. \_RFSMOK3 is derived from \_SMOKER3.                                            | +-------+----------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                        |
|                       |                                                                                                                                                                                   | | **1** | No                         | *Respondents who reported they had not smoked at least 100 cigarettes in their lifetime, those who reported having smoked 100 cigarettes in their lifetime but do not currently smoke.*                                                                                                                                                                                             |                        |
|                       |                                                                                                                                                                                   | +-------+----------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                        |
|                       |                                                                                                                                                                                   | | **2** | Yes                        | *Respondents who reported having smoked at least 100 cigarettes in their lifetime and currently smoke.*                                                                                                                                                                                                                                                                             |                        |
|                       |                                                                                                                                                                                   | +-------+----------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                        |
|                       |                                                                                                                                                                                   | | **9** | Don't know/Refused/Missing | *Respondents who reported they did not know if they had smoked 100 cigarettes in their lifetime, those who refused to answer if they had smoked 100 cigarettes in their lifetime, those who didn't know if they now smoked every day, some days or not at all, those who refused to answer if they now smoked every day, some days or not at all, or those with missing responses.* |                        |
|                       |                                                                                                                                                                                   | +-------+----------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                        |
|                       |                                                                                                                                                                                   | |       |                            |                                                                                                                                                                                                                                                                                                                                                                                     |                        |
|                       |                                                                                                                                                                                   | +-------+----------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+                        |
+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| \_CURECI1             | CV for current e-cigarette smoker status / Calculated variable for adults who are current e-cigarette users.                                                                      | +-------+----------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+    |
|                       |                                                                                                                                                                                   | | **1** | Not currently using E-cigarettes | *Respondents who reported they had not used E-cigarettes in their lifetime, those who reported having used E-cigarettes in their lifetime but do not currently use E-cigarettes.*                                                                                                                                                                                                                 |    |
|                       | \_CURECI1 is derived from ECIGNOW1.                                                                                                                                               | +-------+----------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+    |
|                       |                                                                                                                                                                                   | | **2** | Current E-cigarette user         | *Respondents who reported having used E-cigarettes in their lifetime and currently use E-cigarettes.*                                                                                                                                                                                                                                                                                             |    |
|                       |                                                                                                                                                                                   | +-------+----------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+    |
|                       |                                                                                                                                                                                   | | **9** | Don't know/Refused/Missing       | *Respondents who reported they did not know if they had used E-cigarettes in their lifetime, those who refused to answer if they had used E-cigarettes in their lifetime, those who didn't know if they now used E-cigarettes every day, some days or not at all, those who refused to answer if they now used E-cigarettes every day, some days or not at all, or those with missing responses.* |    |
|                       |                                                                                                                                                                                   | +-------+----------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+    |
+-----------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
