---
title: "BMIN503/EPID600 Final Project"
author: "Srinath Adusumalli"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  

```{r include=FALSE}
library(icd)
library(magrittr)
library(dplyr)
library(lubridate)
library(stringr)
library(tidyverse)
library(ggplot2)
library(readr)
library(knitr)
library(MASS)
library(readr)
library(naniar)
library(table1)
library(htmlTable)
library(readxl)
library(eeptools)
```

***
### Overview
My project aims to examine patterns of consultation of the cardiovascular medicine consult service since the implementation of the PennChart EHR in March 2017.  The purpose of doing this is to inform staffing for this service, particularly with an advanced practice provider, going forward as we move into HUP East/New Patient Pavilion (new patient hospital).

The data being used to complete the project are exclusively extracted from the Penn Medicine Epic/PennChart Clarity enterprise data warehouse.  We queried a number of tables throughout Clarity including those holding information on patient demographics, hospitalizations, medical history, problem list, visit/encounter diagnoses, inpatient consultation/procedure orders, and primary/secondary teams following patients during thier hospitalization.  We have data on patients admitted to the Hospital of the University of Pennsylvania from March 4th, 2017 (date of inpatient go-live) through November 2019.  

Additionally, these data are supplemented by queries from the Epic/PennChart tool SlicerDicer which draws information from the Caboodle data warehouse.

I spoke to the following individuals about my project:
- Srinivas Denduluri, PhD (senior clinical informaticist from the Division of Cardiovascular Medicine): Working with him and the Penn Medicine Data Analytic Center (analyst there is Erik Hossein), we were able to determine the location of patient care team information within Clarity tables.  This is surprisingly hard to to find and took some work to identify the appropriate tables.  I worked with Dr. Denduluri to ultimately write SQL queries to extract relevant data from Clarity and Caboodle.

- Monika Sanghavi, MD (consultative cardiology attending from the Division of Cardiovascular Medicine): Dr. Sanghavi is the director of the inpatient cardiovascular medicine consult service at the Hospital of the University Pennsylvania and is the physician who initially requested data on cardiology consultation.  Speaking with her helped to inform the queries we were conducting and the information we were searching for.

- Frank Silvestry, MD/Daniel Kolansky, MD (echocardiography and interventional cardiology attendings from the Divsion of Cardiovascular Medicine):  Drs. Silvestry and Kolansky are the associate chiefs of the Division and are also stakeholders in the project we are conducting.  Dr. Silvestry is also the director of the Cardiovascular Disease fellowship training program which staffs the consult service with fellows.  From them, I have learned we are hiring an advanced practice provider to supplement the fellows on service, so data on consult service volume will be of use to help inform how we staff that additional provider.

Final GitHub repository: https://github.com/sri-adu/BMIN503_Final_Project

### Introduction 
Introduction:  Cardiovascular disease is the leading cause of mortality and morbidity in the United States and beyond.  As such, it is one of the most common presenting complaints at the Hospital of the University of Pennsylvania, a busy quarternary care academic medical center.  The cardiovascular medicine consult service is one of the most busy (if not the busiest) consult services in the hospital.  Despite this, we have little data on which to base the operations of our service.  This is critical both for current operational needs as well as the future as we think about how to staff the new HUP East/New Patient Pavilion.  My aim is to examine patterns of consultation of the cardiovascular disease consult service since the implementation of an inpatient EHR in March 2017.  As a clinical informatician, I helped to build the informatics infrastucture for using the EHR to answer these questions and now would like to use it 

The problem of how to staff the cardiovascular medicine consult service is multidisciplinary as our service interacts with all types of providers from all specialities (ranging from MDs, to RNs, to physical therapists, and beyond).  Additionally, I am working with our division senior clinical informatician (different discipline than my primary one) as well as a cardiovascular medicine colleague on this project.  In talking with my background sources, I have discovered data needed to address my question is stored in a complex format in PennChart and Clarity.  Even a seemingly simple task such as determining what sequence of teams cared for a particular patient is not so simple through PennChart, Epic, or any EHR.  Despite this, we have now developed a dataset of patients seen by the cardiovascular medicine service as well as their background characteristics and teams from which consults came from which we will use to answer my question.  

### Methods
Given considerable effort was required to compile this dataset from EHR data, the analysis itself will be primarily descriptive and aimed at characterizing patterns of consultation.  As mentioned above, these data are derived from the Clarity data warehouse which represents an extract of data from the PennChart/Epic production environment done every morning at 2am. The SQL code to obtain these data is contained within this repo and also linked here (https://github.com/sri-adu/BMIN503_Final_Project/blob/master/Epic_Consult_Query_SriA.sql).  These data were supplemented by summary data obtained via a self-service tool called SlicerDicer (for total number of HUP admissions).  

If you look closely at the SQL code, part of the work we did was define logic for obtaining patient comorbidities.  As it turns out, Epic/PennChart leaves much to be desired regarding clinically relevant groupers of diagnoses (meaning they are very "dirty" and, at times, nonspecific"). An example of this is "Unspecified cardiovascular disease" being equal to "atherosclerotic cardiovascular disease". As such, we created our own groupings/groupers of diagnoses based on ICD-9-CM and ICD-10-CM diagnosis codes which are modified from Elixhauser comorbitity ICD coding algorithms.  Additionally, we defined search criteria of how to find those data in the EHR by looking in medical history, problem list, and visit/encounter diagnosis tables.

With regards to determining who had a cardiology consult, this is based on tools I created within the EHR.  When we went live with PennChart in 2017, I created consult orders and subsequent patient lists for each of our consult services.  In the majority of instances, when a consult is ordered, and order is placed into PennChart - this order (procedure in Epic terms) is what we are using to define consult.  Furthermore, these orders are grouped into types of cardiology consult which we are using to determine consult buckets (i.e. general cardiology consults, consultative attending rounder consults, etc.).  We are also using provider team information, which corresponds to whether a consult team decided to follow a patient in the hospital and how long they followed a patient while in the hospital.

```{r eval = FALSE}
## Import of data

setwd("~/Box")

# Overall consult patient list
consult_patient_list <- read_excel("consult patient list.xlsx", col_types = c("numeric", "text", "numeric", "date", "text", "text", "text", "text", "text", "text", "text", "text", "text", "text"))

# Hospital visit overview information
consults_hospital_visit_IDs <- read_excel("consults hospital visit IDs.xlsx", col_types = c("numeric", "text", "numeric", "numeric", "date"))

# Hospital visit diagnoses
consults_hospital_visits_dx <- read_excel("consults hospital vistis.xlsx", col_types = c("text", "text", "text", "numeric", "date", "numeric", "date", "date", "text", "numeric", "text", "text", "date", "text", "text", "numeric"))

# Consult team data
> consults_teams <- read_excel("consults teams.xlsx", col_types = c("numeric", "numeric", "text", "numeric", "date", "date", "text", "text", "text", "date", "text", "numeric", "text", "text", "text", "date", "date", "numeric", "numeric", "numeric", "numeric"))

# Patient comorbidity data
consults_comorbids <- read_excel("consults comorbids.xlsx", col_types = c("numeric", "text", "numeric", "text", "text", "text", "text", "text", "text", "text", "text", "text", "text", "text", "text", "text", "text", "text"))

```

```{r eval = TRUE}
# Data cleaning of comorbidities 
consults_comorbids <- consults_comorbids %>%
  replace_with_na_all(condition = ~.x %in% common_na_strings)

# Renaming MRN field in consult patient list to join with comorbidities
consult_patient_list <- consult_patient_list %>% 
  rename(HUP_MRN = MRN) 

# Linking comorbidity data with patient demographic data table in order to obtain full demographic/baseline data
consults_table1 <- inner_join(consult_patient_list,consults_comorbids,by="HUP_MRN")
View(consults_table1)

# Adding in consult type and renaming MRN field for inner join
consults_teams <- consults_teams %>% 
  rename(HUP_MRN = MRN) 

consults_teams <- distinct(consults_teams, consults_teams$HUP_MRN, .keep_all = TRUE)

consults_table_complete <- inner_join(consults_table1,consults_teams,by="HUP_MRN")
View(consults_table_complete)

# Creation of patient age
consults_table_complete$Birth_Date <- as.Date(consults_table_complete$Birth_Date)
consults_table_complete$age <- floor(age_calc(consults_table_complete$Birth_Date, units = "years"))

```

### Results
From March 2017 - November 2019, there were 3,220 unique cardiology consults at the Hospital of the University of Pennsylvania.  By combining these data with a quick search in PennChart/Epic Hyperspace using the SlicerDicer tool (from the Caboodle data warehouse), we find that during that same period of time, there were a total of 75,990 admissions to HUP.  Therefore, our services consulted on 4.23% of all admitted patients through the hospital and in the emergency department during this period of time.  

```{r eval = FALSE}
# Find number of distinct consults based on consult teams
n_distinct(consults_teams$Pat_Enc_CSN_ID)

# Percentage of hospitalized patients who recieved a cardiology consult
3220/75990*100
```

```{r eval = TRUE}
# Creation of table 1 for consults
table1(~ factor(EAP_Procedure) + age + factor(Gender) + factor(Race) + factor(AFib) + factor(CAD) + factor(Dyslipidemia) + factor(HF) + factor(HTN) + factor(Ischemic_Stroke) + factor(Aortic_Aneurysm), data=consults_table_complete, topclass = "Rtable1-zebra")

# Creation of day of the week table for consults
table1(~ factor(Day_of_Week), data=consults_teams, topclass = "Rtable1-zebra")

```

Overall, we see that our consult population is fairly representative of a population of patients with cardiovascular disease.  With regards to our initial question of how to staff the consult service, we note that the least number of consults are seen on Sundays with the most being seen during the middle of the week, which is reflective of the experience most of us have had while on service.  Therefore, we would like to staff the service with additional providers Tuesdays - Fridays.  Presumably, teams have not yet aggregated their thoughts for consults on Mondays, so likely we can give additional providers days off on Saturday, Sunday, and Monday.