---
title: "Accessing Differences in Breast Cancer Screening Utilization by Race"
subtitle: "BMIN503/EPID600 Final Project"
author: "Mattia A. Mahmoud"
format: 
  html:
    toc: true
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

## Overview {#sec-overview}

Breast cancer is the second leading cause of cancer death among women in the U.S., with 43,000 deaths projected in 2023. Worse, Black women have 40% higher risk of breast cancer mortality compared to White women and are more likely to be diagnosed with poor prognostic, later stage cancers. 

Breast density is a strong risk factor for breast cancer. Sixteen states, including Pennsylvania (PA), have recently required coverage with co-pays and deductibles for supplemental screenings for breast cancer with MRI or ultrasound for women with dense breasts. The study purpose is to estimate the potential impact of the PA supplemental screening law on the mammography screening population in the Penn Medicine health system, particularly differences in screening utilization and false negative mammogram rates by race.

## Introduction {#sec-introduction}

Early detection and screening is vital for reducing racial breast cancer disparities. Breast density is the relative amount of fibroglandular versus fatty tissue in the breast.  Women with dense breasts have increased risk both for developing breast cancer as well as for false-negative mammograms. Supplemental screening with breast MRI has been shown to reduce the interval cancer rate (false negative mammogram rate) among patients with dense breasts. In May 2023, Pennsylvania (PA) mandated insurance coverage of one supplemental screening study (ultrasound or breast MRI) annually for women with dense breasts. Prior research suggests that Black women have lower breast density than white women based on subjective grading from the American College of Radiology Breast Imaging and Reporting Data System (BI-RADS), partially due to larger average breast size and higher body mass index (BMI) than white women. Given that Black women have lower prevalence of dense breasts and higher prevalence of triple negative breast cancer, the new supplemental screening law may meaningfully reduce racial disparities in breast cancer prognosis and mortality by reducing false negative mammogram rates. The purpose of this study is to compare supplemental screening utilization by race and see if there are differences in false negative mammogram rates between eligible and non eligible women for the supplemental screening. 

My research problem is interdisciplinary and requires appropriate input from a **multidisciplinary team** including an epidemiologist (Dr. Anne Marie McCarthy), surgeon (Dr. Lola Fayanju), radiologist (Dr. Emily Conant), biostatistician (Dr. Jinbo Chen) to utilize the appropriate methods and have an appropriate clinical context for understanding the results of my research project. Dr. McCarthy's mentorship will focus on epidemiological methods to examine factors that underlie differences in screening utilization by race; with Dr. Fayanju, I will elucidate the importance of carefully considering race and ethnicity in the conduct of the epidemiologic research; with Dr. Conant, I will understand the differences in the interpretation of density measures and with Dr. Chen, I will use methods to evaluate the additive value of the screening law to improve breast cancer incidence and mortality by race.

## Methods {#sec-methods}

The dataset I utilized in this project was obtained from a screening cohort within the Penn Medicine database. Penn Medicine has a large network of breast screening practices that perform thousands of mammograms per year. Women are given a breast cancer risk factor questionnaire at the time of mammogram screening, later stored in the electronic health record (EHR). Information from mammograms, questionnaires, subsequent MRI screenings, and local (Pennsylvania, New Jersey, Delaware) Cancer Registries are linked together for patients as part of Dr. McCarthy's parent grant. 

We included 76,888 mammograms among women aged 40-74 years who identified as Black or White without prior breast cancer or known BRCA1/2 mutations who underwent screening mammography from 2015 to 2021 at the Hospital for the University of Pennsylvania. For each mammogram, breast density was characterized according to BI-RADS (Breast Imaging-Reporting and Data System) density categories: almost entirely fatty, scattered areas of fibroglandular density, heterogeneously dense, and extremely dense.

Women eligible for insurance coverage for supplemental screening in PA include those with:

-   extremely dense breasts

-   heterogeneously dense breasts and \>20% lifetime risk of breast cancer, estimated using the Gail model.

We defined a false negative mammogram as a cancer diagnosis within one year of a non-actionable mammogram (initial BI-RADS Assessment Negative or Benign). We tested for associations of eligibility according to the PA law with risk of a false negative mammogram, adjusting for age and race among Black and White women with negative mammograms.

First of all, the raw clinical and outcome data will be imported. The raw data are in excel, and they were imported as two dataframes.

```{r}
library(tidyverse) 
library(comorbidity) # For Elixhauser comorbidity scores
library(ggplot2) # For graphs
library(lubridate) # For manipulating dates
library(gtsummary) #Summary statistics
library(RColorBrewer) # For coloring the plots
library(cowplot) # For combining multiple plots into one
library(ggpubr) # Combining multiple graphs into one figure 
library(pROC) # For cross-validation
library(dplyr) # For data cleaning
```

```{r}
#setwd("/Users/mattiamahmoud/Downloads")
data <- read_csv("/Users/mattiamahmoud/Downloads/HUP Screening Mammograms 2015-2021.csv")
```

Summary views of the raw patients clinical data were reviewed.

```{r}
str(data)
```

However, lots of information in raw data are not related to this project. Two datasets are therefore cleaned for better utilization based on the information needed.

```{r}
data <- data %>% rename("accession" = "Accession") %>% 
  select(Race,Age,GailRisk5Yr,GailRiskLife,OverallBreastComposition, StudyLevelOutcome, ComponentLevelOutcome, reg_dxdate, reg_subtype4, accession, bc_prior)
```

To investigate the relationships between race, eligibility status and false negative mammography rates, I will select another dataset that has definitive race/ethnicity category and merge it with the dataset with other patient characteristics to facilitate further use.

```{r}
race_variable <- read_csv("/Users/mattiamahmoud/Downloads/Race variable.csv")
```

```{r}
merger <- merge(x = data, y = race_variable, by = c("accession"))
```

Define a new variable as eligibility based on the criteria outlined in the methods section

```{r}
age40_74 <- merger[merger$Age > 40 & merger$Age < 75,]

#make age into a numeric
age40_74$Age <- sub(" .*", "", age40_74$Age)
age40_74$Age <- as.numeric(age40_74$Age)

#####make Gail Lifetime risk into a numeric 
#remove % after numbers
age40_74$GailRiskLife <- sub("%", "", age40_74$GailRiskLife)
age40_74$GailRiskLife <- as.numeric(age40_74$GailRiskLife)

#make new variable with ONLY values of Gail life time risk
age40_74 <- age40_74[which(! age40_74$GailRiskLife=="No" | is.na(age40_74$GailRiskLife)==TRUE),]
age40_74 $GailRiskLife <- as.numeric(age40_74$GailRiskLife)
age40_74 <- age40_74[which(! is.na(age40_74$GailRiskLife)==TRUE),]

#extremely dense breasts
age40_74$ed <- ifelse((age40_74$OverallBreastComposition== 'Extremely dense [4]'), 1, 0)

#heterogenously dense breasts and Gail lifetime risk greater than 20 
age40_74$hd_gail <- ifelse((age40_74$OverallBreastComposition == 'Heterogeneously dense [3]' & age40_74$GailRiskLife >= 20), 1,0)

#eligibility variable
age40_74$eligible <- ifelse((age40_74$ed ==1 | age40_74$hd_gail== 1), 1,0)

```

```{r}
negative <- subset(age40_74, StudyLevelOutcome == "TN" | age40_74$StudyLevelOutcome == "FN")

negative$outcome <- ifelse((negative$StudyLevelOutcome == 'FN'), 1,0)
```

```{r}
white_neg <- subset(negative, race_eth == "White")

black_neg <- subset(negative, race_eth == "Black or African American")

new_negative <- subset(negative, race_eth == "Black or African American" | race_eth == "White")

new_negative$race_eth <- ifelse((new_negative$race_eth == "Black or African American"), 0,1)

# new_negative$OverallBreastComposition <- ifelse((new_negative$OverallBreastComposition == 'Almost entirely fatty [1]'), 1,ifelse(new_negative$OverallBreastComposition=='Scattered fibroglandular density [2]'), 2, ifelse(new_negative$OverallBreastComposition== 'Heterogenously dense'), 3, 4)

new_negative$BreastDensity = ifelse((new_negative$OverallBreastComposition == 'Almost entirely fatty [1]'), 1, ifelse((new_negative$OverallBreastComposition=='Scattered fibroglandular density [2]'), 2,
                       ifelse((new_negative$OverallBreastComposition== 'Heterogeneously dense [3]'), 3,
                              ifelse((new_negative$OverallBreastComposition== 'Extremely dense [4]'), 4,5))))

new_negative %>% select(OverallBreastComposition, BreastDensity)
```

To get an overall information of patients enrolled in this clinical trial, summary of patients characteristics based on race, age, and different receptors status.

```{r}
new_negative <- new_negative %>%
  select(accession, Age, GailRiskLife, StudyLevelOutcome, race_eth, BreastDensity, outcome, reg_subtype4, bc_prior, eligible) %>%
   mutate(race_eth = factor(race_eth, levels=c(0,1), labels=c("Black or African American", "White"))) %>%
 mutate(BreastDensity = factor(BreastDensity, levels=c(1,2,3,4), labels=c("Almost entirely fatty", "Scattered fibrograndular density", "Heterogeneously dense", "Extremely dense"))) %>%
  mutate(outcome = factor(outcome, levels=c(1,0), labels=c("False Negative","True Negative"))) %>%
  mutate(reg_subtype4 = factor(reg_subtype4,levels=c(0,1,2,3,4,9), labels=c("DCIS","ER/PR+/HER-","ER/PR+/HER+", "ER/PR-/HER2+", "ER/PR-/HER2-", "Invasive with unknown subtype"))) %>% 
  mutate(bc_prior = factor(bc_prior,levels=c(0,1), labels=c("No Prior History of Breast Cancer","Prior History of Breast Cancer"))) %>%
 mutate(eligible = factor(eligible,levels=c(0,1), labels=c("Not Eligible for Supplemental Screening","Eligible for Supplemental Screening")))
```

```{r}
table <- new_negative %>%
 select(race_eth,Age,GailRiskLife,BreastDensity, outcome, reg_subtype4, eligible) 
library(knitr)
library(kableExtra)
options(knitr.kable.NA='')
kable(summary(table))
```

## Results {#sec-results}

I will then explore the descriptive statistics of age, race/ethnicity and eligibility status visually

```{r}
library(ggplot2)
ggplot(data = age40_74, aes(x=Age, colour = "darkgoldenrod")) +
  geom_histogram(binwidth = 3) +
  ggtitle(paste("Patients age distribution")) +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

```{r}
ggplot(new_negative, aes(x=race_eth, fill = BreastDensity)) +
  geom_bar(position="fill")+
  scale_fill_manual(values=c("darkseagreen", "darkseagreen2", "chartreuse3", "aquamarine4"))+
  theme_bw()+
  labs(title = "Race/Ethnicity vs. BI-RADS Density",
       x = "Race/Ethnicity",
       y = "BI-RADS Density") +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

First, I want to look at the eligibility status of all patients as a whole to see if there is any relationship between eligibility according to the PA law and patient Ages. The result below, however, doesn't indicate any association of survival status and patients democracy.

```{r}
ggplot(data=new_negative, aes(x=Age, fill=eligible)) +
  geom_histogram(binwidth = 3) +
  scale_fill_manual(values=c("burlywood", "cornsilk4", "gray90"))+
  theme_bw()+
  ggtitle(paste("Eligibility status by Age")) +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

```{r}
ggplot(data=new_negative, aes(x=race_eth)) +
  geom_bar() +
  labs(title = "Patients race distribution",
       x = "Race",
       y = "Count") +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

```{r}
ggplot(new_negative, aes(x=eligible, fill = race_eth)) +
  geom_bar(position="fill")+
  scale_fill_manual(values=c("cadetblue1", "darkturquoise"))+
  theme_bw()+
  labs(title = "Race/Ethnicity vs. Eligibility Status",
       x = "Eligibility Status",
       y = "Race/Ethnicity") +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

```{r}
ggplot(new_negative, aes(x=outcome, fill = eligible)) +
  geom_bar(position="fill")+
  scale_fill_manual(values=c("coral4", "darksalmon"))+
  theme_bw()+
  labs(title = "Eligibility Status and False Negative Mammograms",
       x = "Mammograms",
       y = "Eligibility Status") +
  theme(
    text = element_text(family = "Palatino"),
    plot.title = element_text(hjust = 0, face="bold", size=20))
```

Sensitivity/Specificity

```{r}
mylogit1 <- glm(eligible ~ race_eth + Age, data = new_negative, family = "binomial")
summary(mylogit1)
exp(coef(mylogit1))
exp(cbind(OR = coef(mylogit1), confint(mylogit1)))

#Confusion Matrix
threshold=0.01
predicted_values<-factor(ifelse(predict(mylogit1,type="response")>threshold,1,0))
actual_values<-factor(mylogit1$y)
conf_matrix<-table(predicted_values,actual_values)
conf_matrix

#install.packages("caret")
library(caret)

sensitivity(predicted_values, actual_values)
specificity(conf_matrix)

t3 <- tbl_regression(mylogit1, exponentiate = TRUE) %>%
  bold_p(t = 0.10) %>%
  bold_labels() %>%
  italicize_levels()
t3
```

```{r}
# race = exposure, outcome = eligibility
mylogit6 <- glm(eligible ~ race_eth + Age, data = new_negative, family = "binomial")
summary(mylogit6)
exp(coef(mylogit6))
exp(cbind(OR = coef(mylogit6), confint(mylogit6)))

t3 <- tbl_regression(mylogit6, exponentiate = TRUE) %>%
  bold_p(t = 0.10) %>%
  bold_labels() %>%
  italicize_levels()
t3

# FN - outcome, race, age, eligible are the exposures
mylogit9 <- glm(outcome ~ race_eth + Age + eligible, data = new_negative, family = "binomial")
summary(mylogit9)
exp(coef(mylogit9))
exp(cbind(OR = coef(mylogit9), confint(mylogit9)))

t4 <- tbl_regression(mylogit9, exponentiate = TRUE) %>%
  bold_p(t = 0.10) %>%
  bold_labels() %>%
  italicize_levels()
t4
```

## Conclusion

This the conclusion. The @sec-results can be invoked if you'd like.
