---
title: "Multiple micronutrient supplementation pilot intervention for pregnant women in Botswana"
subtitle: "BMIN503/EPID600 Final Project"
author: "Jingni Pan"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

# Overview

This project was a hybrid implementation-effectiveness pilot intervention of Multiple Micronutrient Supplementation (MMS) for pregnant women in Botswana. The primary goal was to assess the program's reach (penetration) and its effects on maternal hemoglobin levels and adverse birth outcomes using a Difference-in-Difference (DiD) analysis of data from the Tsepamo Study.Dr. Caniglia shared this idea and access to the dataset with me, and help me set up a rigorous analysis plan. From her, I learned the rigorous process of defining the intervention periods and using the non-intervention clinics as a comparison group to ensure the validity of the causal estimate when studying the effects of MMS strategies. Dr. Ruggieri helped narrow down to my current topic focusing on Botswana.

# Introduction

Micronutrient deficiencies during pregnancy are a major public health issue, particularly in low- and middle-income countries (LMICs), where they contribute significantly to maternal anemia, poor fetal growth, and increased risk of adverse birth outcomes.1,2 According to the World Health Organization (WHO), an estimated 32 million pregnant women globally are affected by anemia, most of whom live in LMICs.3 Iron and folic acid supplementation (IFA) has long been the standard of care, yet it often fails to address the multiple concurrent micronutrient deficiencies that women face during pregnancy.

Multiple micronutrient supplementation (MMS), particularly the United Nations International Multiple Micronutrient Antenatal Preparation (UNIMMAP), has been recommended as a more effective alternative to IFA. MMS provides 15 essential vitamins and minerals, including iron, folic acid, zinc, vitamin A, and iodine.4 A systematic review of randomized controlled trials found that compared to IFA, MMS reduced the risk of low birthweight by 12%, preterm birth by 8%, and small for gestational age (SGA) births by 9%.2 MMS has also been shown to improve maternal hemoglobin levels, reduce maternal mortality in some contexts, and promote better long-term health outcomes in children.5 Moreover, recent cost-effectiveness analyses suggest that MMS offers greater health and economic benefits than IFA, with studies reporting positive benefit–cost ratios ranging from 2:1 to 23:1 depending on country context.6,7

Despite these promising findings, MMS is not yet widely adopted in sub-Saharan Africa. In Botswana, maternal and child health continue to face significant challenges. The country has a maternal mortality ratio estimated at 155 deaths per 100,000 live births, and nearly 50% of pregnant women register for antenatal care (ANC) after the first trimester, missing the opportunity for early intervention.8,9 An HIV prevalence of 24% among pregnant women further compounds the risks of anemia and poor pregnancy outcomes.10 A recent large-scale cohort study in Botswana involving over 96,000 women found that MMS users had lower risks of preterm birth, very preterm birth, and low birth weight compared to those on IFA, with the strongest link seen among women with HIV.11 Further, this study found that fewer than 60% of pregnant women in Botswana received guideline-concordant supplementation. Qualitative interviews conducted with pregnant women identified important barriers to IFA, such as stockouts, insufficient education, side effects, poverty, pill burden, and poor adherence, which may similarly affect MMS uptake.12 Taken together, this evidence suggests that MMS may outperform both IFA and iron-only supplementation in Botswana’s clinic-based ANC system, while also underscoring the need to address structural barriers like supplement stockouts, education, and delayed ANC registration.

In light of the efficacy of MMS, well documented in clinical trials, the WHO has updated its guidelines to now recommend MMS “in the context of rigorous research.” WHO and UNICEF specifically call for countries considering MMS programs to generate rigorous implementation research to understand the acceptability, feasibility, equity, and sustainability of switching from IFA to MMS.13 Despite studies in Bangladesh, Indonesia, and Mali, no implementation study has yet been conducted in Botswana. 14–16 This highlights the importance of generating evidence specific to Botswana’s health system and maternal population.

To address these gaps, we conducted a hybrid implementation-effectiveness pilot intervention of providing MMS to ANCs in Botswana. Specifically, we assessed the reach, feasibility, acceptability, and appropriateness of providing MMS to pregnant women through routine ANC services at clinics across three urban, rural, and semi-urban regions in Botswana and measured its effect on maternal hemoglobin levels and adverse birth outcomes through linkage with a nationwide birth outcomes surveillance study (The Tsepamo Study). In doing this, we hope to prove the necessity of transitioning the policy recommendation from IFA to MMS.

# Methods

##Linkage to Tsepamo Study (Data Cleaning)

To assess the primary implementation outcome and all effectiveness outcomes of interest, we leveraged individual-level clinical data collected in the Tsepamo study, which has been conducting birth outcomes surveillance in Botswana since 2014. The Tsepamo study captures delivery data from government hospitals throughout Botswana, including hospitals in Gaborone, Molepolole, and Palapye, and covers up to 72% of births nationwide. Data are abstracted from maternity obstetric records at the time of delivery, including ANC clinic and visits, diagnoses, medications (including MMS), HIV status, and birth outcomes. We used the ANC clinic and date of the first ANC clinic visit (the booking date) recorded in Tsepamo to link our MMS intervention data with the Tsepamo database. To minimize the potential for oversampling shorter pregnancies and to ensure all individuals included in the analysis had the potential to be pregnant for 40 weeks, we restricted our analysis to individuals with an estimated start of pregnancy at least 40 weeks before the day we abstracted the data from Tsepamo (Sep 3, 2025).

```{r}
# Load all neceaary packages
library(haven)
library(dplyr)
library(lubridate)
library(tidyverse)
```

```{r}
#Read a SAS dataset (.sas7bdat)
tsepamo <- read_sas("~/Desktop/tsepamo_updated_5sep25.sas7bdat")
# View the data
print(tsepamo)

# Only keep data from the 13 intervention clinics
mms <- tsepamo %>%         #=intervclinics
  filter(
    cliniccode %in% c(5302, 5303, 5323, 15312, 3319, 15325, 5306, 
                      15331, 15341, 15329, 3201, 5318, 15335)
  )

# Separately, only keep data from the non-intervention clinics (all other clinics in gabs, moleps, palapye
# Creating non-intervention clinics dataset
nonintervclinics <- tsepamo %>%
  mutate(flagtokeep = if_else((cliniccode >= 5000 & cliniccode < 6000) |
                                (cliniccode >= 3000 & cliniccode < 4000) |
                                (cliniccode >= 15000 & cliniccode < 16000), 1, 0))
# Filter the non-intervention clinics based on the flag
nonintervclinics <- nonintervclinics %>%
  filter(flagtokeep == 1) %>%
  filter(!(cliniccode %in% c(5302, 5303, 5323, 15312, 3319, 15325, 5306, 
                             15331, 15341, 15329, 3201, 5318, 15335)))

# Creating the MMS start date and end date based on clinic code T1
mms_timeline <- mms %>%                                 # Has 158+2 variables here
  mutate(
    mmsstartdate = case_when(
      cliniccode == 5302 ~ ymd("2024-08-23"),
      cliniccode == 5303 ~ ymd("2024-08-23"),
      cliniccode == 5323 ~ ymd("2024-08-23"),
      cliniccode == 15312 ~ ymd("2024-08-13"),
      cliniccode == 3319 ~ ymd("2024-09-06"),
      cliniccode == 15325 ~ ymd("2024-08-16"),
      cliniccode == 5306 ~ ymd("2024-08-23"),
      cliniccode == 15331 ~ ymd("2024-08-21"),
      cliniccode == 15341 ~ ymd("2024-08-16"),
      cliniccode == 15329 ~ ymd("2024-08-19"),
      cliniccode == 3201 ~ ymd("2024-09-06"),
      cliniccode == 5318 ~ ymd("2024-08-23"),
      cliniccode == 15335 ~ ymd("2024-08-16")
    ),
    mmsenddate = case_when(
      cliniccode == 5302 ~ ymd("2025-02-13"),
      cliniccode == 5303 ~ ymd("2025-03-17"),
      cliniccode == 5323 ~ ymd("2025-02-13"),
      cliniccode == 15312 ~ ymd("2025-02-20"),
      cliniccode == 3319 ~ ymd("2025-04-15"),
      cliniccode == 15325 ~ ymd("2025-02-20"),
      cliniccode == 5306 ~ ymd("2025-02-13"),
      cliniccode == 15331 ~ ymd("2025-02-20"),
      cliniccode == 15341 ~ ymd("2025-02-20"),
      cliniccode == 15329 ~ ymd("2025-02-20"),
      cliniccode == 3201 ~ ymd("2025-04-15"),
      cliniccode == 5318 ~ ymd("2025-02-13"),
      cliniccode == 15335 ~ ymd("2025-02-20")
    )
  ) %>%
  # Restriction step, 40 weeks before cut-off date Sep 03, 2025
  filter(lmp <= ymd("2024-11-27"))

# Filter mms data for women whose booking date is on or after the start of the pilot
Interv_T1 <- mms_timeline %>%    #Rename from mms_T1 to Interv_T1, for intervention clincis during intervention period  
  filter(bookdate >= mmsstartdate)       

# Creating T0 for intervention clinics 
Interv_timeline_T0 <- mms_timeline %>%             #List past three timelines 
  mutate(
    t0_start_1 = mmsstartdate - years(1),
    t0_end_1   = mmsenddate - years(1),
    t0_start_2 = mmsstartdate - years(2),
    t0_end_2   = mmsenddate - years(2),
    t0_start_3 = mmsstartdate - years(3),
    t0_end_3   = mmsenddate - years(3)
  ) %>%
  distinct(cliniccode, .keep_all = TRUE) %>%
  select(cliniccode, starts_with("t0_"))

Interv_T0 <- mms %>%        #For intervention clincis during non-intervention period  
  left_join(Interv_timeline_T0, by = "cliniccode") %>%
  filter(
    (bookdate >= t0_start_1 & bookdate <= t0_end_1) |
      (bookdate >= t0_start_2 & bookdate <= t0_end_2) |
      (bookdate >= t0_start_3 & bookdate <= t0_end_3)
  )

# Create start and end dates for non-intervention clinics
nonintervclinics_timeline <- nonintervclinics %>%                          # Has 159+2 variables
  mutate(
    mmsstartdate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2024-08-23"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2024-08-16"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2024-09-06")
    ),
    mmsenddate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2025-02-13"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2025-02-20"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2025-04-15")
    )
  ) %>%

  # Restriction step, 40 weeks before cut-off date Sep 03, 2025
  filter(lmp <= ymd("2024-11-27"))

# Filter non-intervention clinics for women whose booking date is on or after the start of the pilot
noninterv_T1 <- nonintervclinics_timeline %>%
  filter(bookdate >= mmsstartdate)

# Creating T0 for nonintervention clinics 
Noninterv_timeline_T0_ <- nonintervclinics_timeline %>%             #List past three timelines 
  mutate(
    t0_start_1 = mmsstartdate - years(1),
    t0_end_1   = mmsenddate - years(1),
    t0_start_2 = mmsstartdate - years(2),
    t0_end_2   = mmsenddate - years(2),
    t0_start_3 = mmsstartdate - years(3),
    t0_end_3   = mmsenddate - years(3)
  ) %>%
  distinct(cliniccode, .keep_all = TRUE) %>%
  select(cliniccode, starts_with("t0_"))

noninterv_T0 <- nonintervclinics %>%   #use original nonintervention clinics dataset without T1 timeline
  left_join(Noninterv_timeline_T0_, by = "cliniccode") %>%
  filter(
    (bookdate >= t0_start_1 & bookdate <= t0_end_1) |
      (bookdate >= t0_start_2 & bookdate <= t0_end_2) |
      (bookdate >= t0_start_3 & bookdate <= t0_end_3)
  )# Create start and end dates for non-intervention clinics
nonintervclinics_timeline <- nonintervclinics %>%                          # Has 159+2 variables
  mutate(
    mmsstartdate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2024-08-23"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2024-08-16"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2024-09-06")
    ),
    mmsenddate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2025-02-13"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2025-02-20"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2025-04-15")
    )
  ) %>%
  # Restriction step, 40 weeks before cut-off date Sep 03, 2025
  filter(lmp <= ymd("2024-11-27"))

# Filter non-intervention clinics for women whose booking date is on or after the start of the pilot
noninterv_T1 <- nonintervclinics_timeline %>%
  filter(bookdate >= mmsstartdate)

# Creating T0 for nonintervention clinics 
Noninterv_timeline_T0_ <- nonintervclinics_timeline %>%             #List past three timelines 
  mutate(
    t0_start_1 = mmsstartdate - years(1),
    t0_end_1   = mmsenddate - years(1),
    t0_start_2 = mmsstartdate - years(2),
    t0_end_2   = mmsenddate - years(2),
    t0_start_3 = mmsstartdate - years(3),
    t0_end_3   = mmsenddate - years(3)
  ) %>%
  distinct(cliniccode, .keep_all = TRUE) %>%
  select(cliniccode, starts_with("t0_"))

noninterv_T0 <- nonintervclinics %>%   #use original nonintervention clinics dataset without T1 timeline
  left_join(Noninterv_timeline_T0_, by = "cliniccode") %>%
  filter(
    (bookdate >= t0_start_1 & bookdate <= t0_end_1) |
      (bookdate >= t0_start_2 & bookdate <= t0_end_2) |
      (bookdate >= t0_start_3 & bookdate <= t0_end_3)
  )
```

## Implementation Outcomes

The implementation outcome was MMS penetration (often called “reach”),18 which is defined as the proportion of eligible individuals who participate in the intervention. We estimated penetration by calculating the proportion of deliveries for which there was documentation in the obstetric card that MMS was the first supplement prescribed during pregnancy. We used a difference-in-difference (DiD) approach to compare this proportion across time and clinic. We fit a linear probability model with indicators for the intervention group, the post-intervention period, and an interaction between the intervention group and post-intervention period. The coefficient for the interaction term quantifies the change in the proportion who received MMS from the pre-intervention to the post-intervention period, comparing the intervention clinics with the non-intervention clinics (a difference-in-difference). Each clinic had a unique intervention period, defined as the time between the first MMS distribution and 30 days after the sixth month of distribution. The pre-intervention period refers to the same calendar period in the three consecutive prior years. We included all clinics in Gaborone, Molepolole, and Palapye as non-intervention clinics in the primary analysis. Individuals were included in the analysis if they initiated ANC at one of the 13 intervention clinics or at a non-intervention clinic during either the intervention period or the corresponding pre-intervention period and delivered at a site included in Tsepamo. For non-intervention clinics, the intervention and pre-intervention periods were defined using the median intervention dates from clinics within the same geographic region.

```{r}
#Coding variable for MMNS, IFAS only, Iron only, No iron
tsepamo <- tsepamo %>%
  mutate(
    mmns = case_when(
      multivit == 1 ~ 3,                                      # MMNS
      (iron == 1 & folate == 1) | (fefol == 1) ~ 2,            # Iron + folatic acid but no other micronutrients
      (iron == 1 & folate == 0 & fefol == 0 & multivit == 0) ~ 1, # Iron only
      (iron == 0 & fefol == 0 & multivit == 0) ~ 0             # Not on iron
    )
  )

# Number MMS filled in Tsepamo among Women Presenting to ANC
# During T1
##Intervention
Interv_MMS_T1 <- Interv_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))
Interv_MMS_T1
##nonintervention
noninterv_MMS_T1 <- noninterv_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))
noninterv_MMS_T1

# During T0
##Intervention
Interv_MMS_T0 <- Interv_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))
Interv_MMS_T0
##nonintervention
noninterv_MMS_T0 <- noninterv_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))
noninterv_MMS_T0
```

## Effectiveness Outcomes

The primary effectiveness outcome was the mean change in hemoglobin during pregnancy, calculated as the difference between the second and first hemoglobin measurements recorded in the Tsepamo database. Specifically, the first hemoglobin must be measured before 24 weeks of gestation, and the second hemoglobin must be measured on or after 24 weeks of gestation. Secondary outcomes included the second hemoglobin value measured in pregnancy and a composite indicator of adverse birth outcomes, including stillbirth, preterm delivery (\<37 weeks of gestation), in-hospital neonatal death (\<28 days), and small-for-gestational-age (SGA). Preterm delivery, SGA, and low birth weight (\<2500 grams) were also examined individually. As above, analyses of the effectiveness outcomes were restricted to individuals who delivered at a site included in Tsepamo.

### Adverse Birth Outcome

```{r}
# Number any adverse birth outcome in Tsepamo among Women Presenting to ANC

## SGA and VSGA calculation functions

calculate_sga <- function(nga, bwt, gender, bwtgm) {
  # Vectorized version using case_when
  dplyr::case_when(
    is.na(nga) | is.na(bwt) | is.na(gender) | is.na(bwtgm) ~ NA_real_,
    bwt < bwtgm ~ 1L,  # Meets SGA criteria
    TRUE ~ 0L          # Does not meet criteria
  )
}

calculate_vsga <- function(nga, bwt, gender, bwtgm) {
  # Identical structure since logic is the same, just different thresholds
  dplyr::case_when(
    is.na(nga) | is.na(bwt) | is.na(gender) | is.na(bwtgm) ~ NA_real_,
    bwt < bwtgm ~ 1L,  # Meets VSGA criteria
    TRUE ~ 0L          # Does not meet criteria
  )
}

# Correct SGA and VSGA weight thresholds
sgawham <- setNames(c(530,600,690,780,880,1000,1130,1270,1430,1550,1820,2050,2270,2470,2640,2800,2940,3060,3170),
                    24:42)
sgawhaf <- setNames(c(500,570,650,740,830,940,1070,1200,1350,1530,1790,2020,2230,2410,2570,2710,2830,2930,3010),
                    24:42)

vsgawham <- setNames(c(470,540,610,690,790,890,1000,1130,1270,1300,1560,1800,2020,2220,2390,2550,2690,2820,2920),
                     24:42)
vsgawhaf <- setNames(c(440,510,580,660,740,840,950,1070,1200,1320,1570,1800,2000,2180,2340,2480,2600,2690,2770),
                     24:42)

calculate_growth_status <- function(bwt, bwtgm) {
  dplyr::case_when(
    is.na(bwt) | is.na(bwtgm) ~ NA_real_,  # Handle missing data
    bwt < bwtgm ~ 1L,                     # Meets criteria
    TRUE ~ 0L                              # Does not meet criteria
  )
}


## For Intervention period

# Intervention clinics 
Interv_BD_T1 <- Interv_T1 %>%
  # First create prerequisite variables
  mutate(
    
    # Set implausible values to NA
    nga_clean = if_else(nga %in% 24:42, nga, NA_real_),
    
    # Flag outliers for review
    nga_flag = case_when(
      nga < 24 ~ "Below viability threshold",
      nga > 42 ~ "Post-term error",
      is.na(nga) ~ "Missing GA",
      TRUE ~ NA_character_
    ),
    
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga_clean %in% 24:36 ~ 1L,  # WHO preterm definition
      nga_clean >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga_clean %in% 24:31 ~ 1L,  # WHO very preterm
      nga_clean >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birth weight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(round(nga_clean)),  # Handle decimal weeks if needed
    bwtgm_sga = case_when(
      gender == 0 ~ sgawham[nga_char],
      gender == 1 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 ~ vsgawham[nga_char],
      gender == 1 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt <= bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt <= bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

Interv_BD_clinics_T1 <- Interv_BD_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )
Interv_BD_clinics_T1

# Non-intervention clincis
noninterv_BD_T1 <- noninterv_T1 %>%
  # First create prerequisite variables
  mutate(
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga >= 24 & nga < 37 ~ 1L,
      nga >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga >= 24 & nga < 32 ~ 1L,
      nga >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birthweight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(nga),
    bwtgm_sga = case_when(
      gender == 0 & nga %in% 24:42 ~ sgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 & nga %in% 24:42 ~ vsgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt < bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt < bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

noninterv_BD_clinics_T1 <- noninterv_BD_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )
noninterv_BD_clinics_T1

## For Non-interventnion period
#Intervenetion clincis
Interv_BD_T0 <- Interv_T0 %>%
  # First create prerequisite variables
  mutate(
    
    # Set implausible values to NA
    nga_clean = if_else(nga %in% 24:42, nga, NA_real_),
    
    # Flag outliers for review
    nga_flag = case_when(
      nga < 24 ~ "Below viability threshold",
      nga > 42 ~ "Post-term error",
      is.na(nga) ~ "Missing GA",
      TRUE ~ NA_character_
    ),
    
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga_clean %in% 24:36 ~ 1L,  # WHO preterm definition
      nga_clean >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga_clean %in% 24:31 ~ 1L,  # WHO very preterm
      nga_clean >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birthweight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(round(nga_clean)),  # Handle decimal weeks if needed
    bwtgm_sga = case_when(
      gender == 0 ~ sgawham[nga_char],
      gender == 1 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 ~ vsgawham[nga_char],
      gender == 1 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt <= bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt <= bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

Interv_BD_clinics_T0 <- Interv_BD_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )
Interv_BD_clinics_T0

# Non-interverntion clinics
noninterv_BD_T0 <- noninterv_T0 %>%
  # First create prerequisite variables
  mutate(
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga >= 24 & nga < 37 ~ 1L,
      nga >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga >= 24 & nga < 32 ~ 1L,
      nga >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birthweight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(nga),
    bwtgm_sga = case_when(
      gender == 0 & nga %in% 24:42 ~ sgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 & nga %in% 24:42 ~ vsgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt < bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt < bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

noninterv_BD_clinics_T0 <- noninterv_BD_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )
noninterv_BD_clinics_T0
```

### Hemoglobins

```{r}
# Number deliveries in Tsepamo among Women Presenting to ANC with 2 hemoglobins
#During T1

##Interventnion
Interv_2hb_T1 <- Interv_T1 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())
Interv_2hb_T1
##Non~
noninterv_2hb_T1 <- noninterv_T1 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())
noninterv_2hb_T1

# During T0

##Interventnion
Interv_2hb_T0 <- Interv_T0 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())
Interv_2hb_T0
##Non~
noninterv_2hb_T0 <- noninterv_T0 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())
noninterv_2hb_T0
```

# Results

## Descriptive Statistics

The MMS intervention was conducted between August 2024 and April 2025. Overall, the study team dispensed 19,373 30-count bags of MMS (10,409 during the intervention period and 8,964 at the conclusion of the intervention period). Table below presents the sociodemographic and clinical characteristics of women who initiated ANC at the intervention and non-intervention clinics during the 3-year pre-intervention period and during the intervention period. During the intervention period, there were 1,262 deliveries among women presenting to ANC at the 13 intervention clinics, compared with 5,116 deliveries during the corresponding pre-intervention period. In non-intervention clinics, there were 2,460 deliveries during the intervention period and 10,279 deliveries during the corresponding pre-intervention period. Across both groups and time periods, over half of the women-initiated ANC in the second trimester. Approximately 14 to 16% of women were living with HIV across groups and time periods, and the distribution of hemoglobin groups was generally similar. Most women weighed between 50 and 80 kg, and the majority were aged 20 to 35 years. Over half of the women had one or more children, and the vast majority had attained at least secondary education.

```{r}
# Recode variables to match the demo table's requirement
clean_variables <- function(df){
  df %>%
    mutate(
      hiv = case_when(
        hivpos == 1 ~ "Women with HIV",
        hivpos == 0 ~ "Women without HIV",
        TRUE ~ NA_character_
      ),
      hb_group = case_when(
        firsthb < 11 ~ "<11",
        firsthb >= 11 ~ "≥11",
        TRUE ~ NA_character_
      ),
      weight_group = case_when(
        firstwt < 50 ~ "<50kg",
        firstwt >= 50 & firstwt < 80 ~ "50-80kg",
        firstwt >= 80 ~ "≥80kg",
        TRUE ~ NA_character_
      ),
      age_group = case_when(
        matage < 20 ~ "<20 years",
        matage >= 20 & matage <= 35 ~ "20-35 years",
        matage > 35 ~ "≥35 years",
        TRUE ~ NA_character_
      ),
      parity_group = case_when(
        para >= 1 ~ "1 or more children",
        para == 0 ~ "0 children",
        TRUE ~ "Missing"
      ),
      occupation_group = case_when(
        occup == 3 ~ "Salaried",
        occup %in% c(1, 2, 99) ~ "Other, unknown, missing",
        TRUE ~ NA_character_
      ),
      education_group = case_when(
        educ %in% c(2, 3) ~ "Secondary or higher",
        educ %in% c(1, 4, 99) ~ "None or primary or missing",
        TRUE ~ NA_character_
      )
    )
}

Interv_T1 <- clean_variables(Interv_T1)
Interv_T0 <- clean_variables(Interv_T0)
noninterv_T1 <- clean_variables(noninterv_T1)
noninterv_T0 <- clean_variables(noninterv_T0)
```

```{r}
library(dplyr)
library(gtsummary)
library(flextable)
library(labelled)
library(rifttable)
library(forcats)

# 1. Data preperation, combination, and label

## Combine data and create groups labelling
Sum_table_2 <- bind_rows(
  Interv_T1 %>% mutate(group = "Intervention T1"),
  Interv_T0 %>% mutate(group = "Intervention T0"),
  noninterv_T1 %>% mutate(group = "Non-Intervention T1"),
  noninterv_T0 %>% mutate(group = "Non-Intervention T0")
)

## Define and order all the variables (Factor Levels)
Sum_table_2 <- Sum_table_2 %>%
  mutate(
    lmp_calc = delivdate - nga * 7,
    daystobooking = as.numeric(bookdate - lmp_calc),   # First ANC weeks
    weekstobooking_demo = daystobooking / 7,
    trimester_group = case_when(
      !is.na(weekstobooking_demo) & weekstobooking_demo < 12 ~ "First (<12 weeks)",
      !is.na(weekstobooking_demo) & weekstobooking_demo < 24 ~ "Second (12-24 weeks)",
      !is.na(weekstobooking_demo) ~ "≥24 weeks",
      TRUE ~ "Missing") |> factor(levels = c("First (<12 weeks)",
                           "Second (12-24 weeks)",
                           "≥24 weeks", 
                           "Missing")),
    clinic_type = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ "Molepolole",
      cliniccode >= 3000 & cliniccode < 4000 ~ "Palapye",
      cliniccode >= 15000 & cliniccode < 16000 ~ "Gaborone",
      TRUE ~ "Other/Unknown" 
    ),
    group = factor(group, levels = c(
      "Intervention T1",
      "Intervention T0",
      "Non-Intervention T1",
      "Non-Intervention T0"
    )),
    strata = sub("^(Intervention|Non-Intervention).*$", "\\1", group),
    strata = factor(strata, levels = c("Intervention", "Non-Intervention")),
    time_point = sub(".*(T[01])$", "\\1", group),
    time_point = factor(time_point, levels = c("T1", "T0")),
    
    hb_group = fct_na_value_to_level(hb_group, level = "Missing"),
    hb_group = factor(hb_group,
                      levels = c("<11", "≥11", "Missing")),
    hiv = fct_na_value_to_level(hiv, level = "Missing"),
    hiv = factor(hiv,
                 levels = c("Women with HIV","Women without HIV", "Missing")),
    weight_group = fct_na_value_to_level(weight_group, level = "Missing"),
    weight_group = factor(weight_group,
                          levels = c("<50kg","50-80kg","≥80kg", "Missing")),
    age_group = fct_na_value_to_level(age_group, level = "Missing"),
    age_group = factor(age_group,
                       levels = c("<20 years","20-35 years","≥35 years", "Missing")),
    parity_group = factor(parity_group,
                          levels = c("1 or more children","0 children or missing")),
    occupation_group = factor(occupation_group,
                              levels = c("Salaried","Other, unknown, missing")),
    education_group = factor(education_group,
                             levels = c("Secondary or higher","None or primary or missing")),
    clinic_type = factor(clinic_type, 
                         levels = c("Molepolole", "Palapye", "Gaborone", "Other/Unknown"))
  )

## Set label for variables
Sum_table_2 <- Sum_table_2 %>% 
  set_variable_labels(
    clinic_type = "Clinic Type",
    trimester_group = "Trimester of first ANC",
    hiv = "HIV Status",
    firsthb = "Hemoglobin (g/dL), Median (IQR)", # Cotinuous 
    hb_group = "Hemoglobin Group, N (%)",       # binary 
    weight_group = "Weight Group",
    age_group = "Age Group",
    parity_group = "Parity Group",
    occupation_group = "Occupation",
    education_group = "Education",
  )

# 2. Generate Table 1 (gtsummary)

Table1_gtsummary <- Sum_table_2 %>%
  
  # include：firsthb (con) and hb_group (binary)
  select(group, clinic_type, trimester_group, hiv, firsthb, hb_group, 
         weight_group, age_group, parity_group, occupation_group, education_group) %>%
  
  tbl_summary(
    by = group, 
    #number format in the table
    statistic = list(
      clinic_type ~ "{n} ({p}%)",
      all_continuous() ~ c("{median} ({p25}, {p75})"),
      all_categorical() ~ "{n} ({p}%)"
    ),
     
    # KEY ADDITION: Set the number of decimal places for percentages
    digits = list(
      all_categorical() ~ c(0, 2), # Set all categorical percentages to 2 decimal places
      all_continuous() ~ 2   # Set continuous variables (Median, IQR) to 2 decimal places as well
  )) %>%
    
  bold_labels() %>%
  
  gtsummary::modify_spanning_header(
    c(stat_1, stat_2) ~ "**Intervention Clinics**",
    c(stat_3, stat_4) ~ "**Non-intervention Clinics**"
  ) %>%
  
  gtsummary::modify_header(
    stat_1 = "**T1**\nN={n} ({style_percent(p)}%)",
    stat_2 = "**T0**\nN={n} ({style_percent(p)}%)",
    stat_3 = "**T1**\nN={n} ({style_percent(p)}%)",
    stat_4 = "**T0**\nN={n} ({style_percent(p)}%)"
  )

# Transfer gtsummary to flextable 
Table1_1 <- Table1_gtsummary %>% as_flex_table()
Table1_1
```

Table 1. Baseline characteristics measured at the first antenatal visit by supplementation strategy, the Tsepamo Study

## Implementation Outcomes

During the pre-intervention period, receipt of MMS was similar between the intervention (40.9%) and non-intervention clinics (40.9%). During the intervention period, receipt of MMS increased in intervention clinics (84.6%) but remained similar in non-intervention clinics (37.2%). The resulting DiD estimate was 47.4% (95% CI: 0.438, 0.511).

```{r}
#Add treated and post to each dataset for future use
Interv_T0 <- Interv_T0 %>%
  mutate(treated = 1, post = 0)

Interv_T1 <- Interv_T1 %>%
  mutate(treated = 1, post = 1)

noninterv_T0 <- noninterv_T0 %>%
  mutate(treated = 0, post = 0)

noninterv_T1 <- noninterv_T1 %>%
  mutate(treated = 0, post = 1)

combined_table <- bind_rows(Interv_T0, Interv_T1, noninterv_T0, noninterv_T1)

#Add outcome mms variable
did_table_2 <- combined_table %>%
  mutate(mms_received = ifelse(multivit == 1, 1, 0))

did_model_2 <- lm(mms_received ~ treated + post + treated:post, data = did_table_2)
summary(did_model_2)   #new model, looks better

#CI
confint(did_model_2)
#get 4 variables listed
did_table_2 %>%
  group_by(treated, post) %>%
  summarise(
    pMMS = mean(mms_received, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()
```

Results were similar in sensitivity and subgroup analyses.

```{r}
#Subgroup DID for pMMS
did_table_2_sub <- did_table_2 %>%
  mutate(region = case_when(
    cliniccode %in% c(5302, 5303, 5306, 5318, 5323) ~ "Molepolole",     #compare to specific noninterventions 
    cliniccode %in% c(15312, 15325, 15329, 15331, 15335, 15341) ~ "Gaborone",
    cliniccode %in% c(3201, 3319) ~ "Palapye",
    TRUE ~ "Other"
  ))
## Define the Gaborone
intervention_Gab_subset <- c(15312, 15325, 15329, 15331, 15335, 15341)
nonintervention_Gab_subset <- c(15305, 15314, 15315, 15330, 15336)
## Define the Molepolole
intervention_Mol_subset <- c(5302, 5303, 5323, 5306, 5318)
nonintervention_Mol_subset <- c(5301, 5305, 5307, 5313, 5419)
## Define the Palapye
intervention_Pal_subset <- c(3201, 3319)
nonintervention_Pal_subset <- c(3313, 3320, 3322)

##Gaborone: Gaborone vs Gaborone non-intervention clinics
gabs_did_pMMS <- did_table_2 %>%
  filter(cliniccode %in% c(intervention_Gab_subset, nonintervention_Gab_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Gab_subset, 1, 0))

gabs_model <- lm(mms_received ~ treated + post + treated:post, data = gabs_did_pMMS)
summary(gabs_model)
#CI
confint(gabs_model)
#get 4 variables listed
gabs_did_pMMS %>%
  group_by(treated, post) %>%
  summarise(
    pMMS = mean(mms_received, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

##Molepolole
Mol_did_pMMS <- did_table_2 %>%
  filter(cliniccode %in% c(intervention_Mol_subset, nonintervention_Mol_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Mol_subset, 1, 0))

mole_model <- lm(mms_received ~ treated + post + treated:post, data = Mol_did_pMMS)
summary(mole_model)
#CI
confint(mole_model)
#get 4 variables listed
Mol_did_pMMS %>%
  group_by(treated, post) %>%
  summarise(
    pMMS = mean(mms_received, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

##Palapye
pal_did_pMMS <- did_table_2 %>%
  filter(cliniccode %in% c(intervention_Pal_subset, nonintervention_Pal_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Pal_subset, 1, 0))

pal_model <- lm(mms_received ~ treated + post + treated:post, data = pal_did_pMMS)
summary(pal_model)
#CI
confint(pal_model)
#get 4 variables listed
pal_did_pMMS %>%
  group_by(treated, post) %>%
  summarise(
    pMMS = mean(mms_received, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()

##HIV status
# HIV-positive subset
pMMS_hivpos <- did_table_2 %>% filter(hivpos == 1)     #use dataset from analysis
# HIV-negative subset
pMMS_hivneg <- did_table_2 %>% filter(hivpos == 0)
# DID 
# For HIV-positive
did_hivpos_pMMS <- lm(mms_received ~ treated + post + treated:post, data = pMMS_hivpos)
summary(did_hivpos_pMMS)
confint(did_hivpos_pMMS) #CI
did_table_2 %>% # 4 proportions
  filter(hivpos == 1) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    prop_mms = mean(mms_received == 1, na.rm = TRUE)
  )

# For HIV-negative
did_hivneg_pMMS <- lm(mms_received ~ treated + post + treated:post, data = pMMS_hivneg)
summary(did_hivneg_pMMS)
confint(did_hivneg_pMMS)
did_table_2 %>% # 4 proportions
  filter(hivpos == 0) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    prop_mms = mean(mms_received == 1, na.rm = TRUE)
  )

## Sensiivity Analysis 
# List 13 intervention and 13 matched control clinics
intervention_clinics_did <- c(5302, 5303, 5323, 15312, 3319, 15325, 5306, 
                              15331, 15341, 15329, 3201, 5318, 15335)  
nonintervention_clinics_did <- c(3313, 3320, 3322, 5301, 5305, 5307, 5313, 5419, 15305, 15314, 15315, 15330, 15336)  

pMMS_sensitivity <- combined_table %>%
  filter(cliniccode %in% c(intervention_clinics_did, nonintervention_clinics_did))

pMMS_sensitivity <- pMMS_sensitivity %>%
  mutate(mms_received = ifelse(multivit == 1, 1, 0))

sensitivity_model_MMS <- lm(mms_received ~ treated + post + treated:post, data = pMMS_sensitivity)
summary(sensitivity_model_MMS)
#CI
confint(sensitivity_model_MMS)
#get 4 variables listed
pMMS_sensitivity %>%        #use table instead of model!
  group_by(treated, post) %>%
  summarise(
    pMMS = mean(mms_received, na.rm = TRUE),
    n = n()
  ) %>%
  ungroup()
```
```{r}
# Create dataset
did_data_new <- tibble(
  # Column representing the Y-axis outcomes/models
  Model = c(
    "13 Intervention: All Non-intervention",
    "13 Intervention: 13 Non-intervention",
    "Subgroup: Gabrone",
    "Subgroup: Molepolole",
    "Subgroup: Palapye",
    "Subgroup: HIV +",
    "Subgroup: HIV -"
  ),
  # DiD Estimate (pMMS column)
  Estimate = c(0.474, 0.551, 0.666, 0.552, 0.200, 0.447, 0.483)*100,
  # Lower CI (2.5 %)
  Lower_CI = c(0.438, 0.506, 0.606, 0.468, 0.088, 0.355, 0.443)*100,
  # Upper CI (97.5 %)
  Upper_CI = c(0.511, 0.596, 0.726, 0.636, 0.311, 0.538, 0.522)*100
) %>%
  mutate(
    # Check if the confidence interval crosses zero (Sig if both bounds are > 0 or < 0)
    Significance = if_else(
      condition = (Lower_CI * Upper_CI) > 0,
      true = "Sig",
      false = "n.s."
    ),
    # Ensure the outcomes are plotted in the correct order (reversing the list order for presentation)
    Model = fct_inorder(Model)
  )

print(did_data_new)

forest_plot_new <- ggplot(
  data = did_data_new, 
  aes(
    x = Estimate, 
    y = Model, 
    color = Model, 
    shape = Model
  )
) +
  scale_color_viridis_d() +
  # A: 95% CI
  geom_errorbarh(aes(xmin = Lower_CI, xmax = Upper_CI), height = 0.2) +
  
  # B: point estimate
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  
  # C: Set X range and label 
  scale_x_continuous(
    name = "DiD Estimate (pMMS, 95% CI)",
    limits = c(0, 80), 
    breaks = seq(0, 80, by = 10)
  ) +
  
  # D: tidy pattern
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(), 
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  ) +
  
  # E: Y axis label 
  labs(y = NULL, title = "Difference-in-Differences Results")

print(forest_plot_new)
```
This graph includes the absolute difference-in-difference results for the subgroup, 13 intervention versus 13 non-intervention clinics, and 13 intervention versus all non-intervention clinics. It shows that results are pretty consistent. 

## Effectiveness Outcomes
### Mean change in difference between first and second hemoglobin
During the intervention period, there were 421 pregnant women in the intervention clinics and 747 in the non-intervention clinics who had one hemoglobin measurement before 24 weeks of gestation and a second hemoglobin measurement on or after 24 weeks of gestation. During the pre-intervention period, there were 1,944 pregnant women in the intervention clinics and 3,727 pregnant women in the non-intervention clinics, who had one hemoglobin measurement before 24 weeks of gestation and a second hemoglobin measurement on or after 24 weeks of gestation. The results indicate that the change in hemoglobin was reduced by 0.048 g/dL more in the intervention clinics relative to the non-intervention clinics post-intervention. 
```{r}
# Filter for those with both hemoglobins and compute change
filtered_table <- combined_table %>%
  mutate(
    lmp_calc = delivdate - nga * 7,
    gage_firsthb  = as.numeric((hb_date  - lmp_calc) / 7),
    gage_secondhb = as.numeric((hb2_date - lmp_calc) / 7)
  ) %>%
  filter(gage_firsthb < 24, gage_secondhb >= 24) %>%
  filter(!is.na(firsthb), !is.na(secondhb))

hb_table <- filtered_table %>%
  mutate(hb_change = secondhb - firsthb)

# Run DID model
hb_did_model <- lm(hb_change ~ treated + post + treated:post, data = hb_table)
summary(hb_did_model)
confint(hb_did_model)
# Get group means
hb_table %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(hb_change, na.rm = TRUE),
    sd_hb_change = sd(hb_change, na.rm = TRUE)
  )
```
Results were consistent when restricting the non-intervention clinics to the 13 clinics originally eligible and in the subgroup analyses by region and HIV status. 
```{r}
## Subgroup Analysis
hb_table_sub <- hb_table %>%
  mutate(region = case_when(
    cliniccode %in% c(5302, 5303, 5306, 5318, 5323) ~ "Molepolole",     #compare tp all other noninterventions 
    cliniccode %in% c(15312, 15325, 15329, 15331, 15335, 15341) ~ "Gaborone",
    cliniccode %in% c(3201, 3319) ~ "Palapye",
    TRUE ~ "Other"
  ))
##Gaborone
gabs_did_hb <- hb_table %>%
  filter(cliniccode %in% c(intervention_Gab_subset, nonintervention_Gab_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Gab_subset, 1, 0))

gabs_model_hb <- lm(hb_change ~ treated + post + treated:post, data = gabs_did_hb)
summary(gabs_model_hb)
#CI
confint(gabs_model_hb)
#get 4 variables listed
gabs_did_hb %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(hb_change, na.rm = TRUE),
    sd_hb_change = sd(hb_change, na.rm = TRUE)
  )
##Molepolole
mole_did_hb <- hb_table %>%
  filter(cliniccode %in% c(intervention_Mol_subset, nonintervention_Mol_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Mol_subset, 1, 0))

mole_model_hb <- lm(hb_change ~ treated + post + treated:post, data = mole_did_hb)
summary(mole_model_hb)
#CI
confint(mole_model_hb)
#get 4 variables listed
mole_did_hb %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(hb_change, na.rm = TRUE),
    sd_hb_change = sd(hb_change, na.rm = TRUE)
  )

##Palapye
pal_did_hb <- hb_table %>%
  filter(cliniccode %in% c(intervention_Pal_subset, nonintervention_Pal_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Pal_subset, 1, 0))

pal_model_hb <- lm(hb_change ~ treated + post + treated:post, data = pal_did_hb)
summary(pal_model_hb)
#CI
confint(pal_model_hb)
#get 4 variables listed
pal_did_hb %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(hb_change, na.rm = TRUE),
    sd_hb_change = sd(hb_change, na.rm = TRUE)
  )

##HIV status
# HIV-positive subset
hb_hivpos <- hb_table %>% filter(hivpos == 1)     #use dataset from total analysis
# HIV-negative subset
hb_hivneg <- hb_table %>% filter(hivpos == 0)
# DID 
# For HIV-positive
did_hivpos_hb <- lm(hb_change ~ treated + post + treated:post, data = hb_hivpos)
summary(did_hivpos_hb)
confint(did_hivpos_hb) #CI
hb_table %>% # 4 proportions, need to change the dataset to be consistent
  filter(hivpos == 1) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(hb_change, na.rm = TRUE),
    sd_hb_change = sd(hb_change, na.rm = TRUE)
  )

# For HIV-negative
did_hivneg_hb <- lm(hb_change ~ treated + post + treated:post, data = hb_hivneg)
summary(did_hivneg_hb)
confint(did_hivneg_hb)
hb_table %>% # 4 proportions
  filter(hivpos == 0) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(hb_change, na.rm = TRUE),
    sd_hb_change = sd(hb_change, na.rm = TRUE)
  )
## Sensitivity Analysis
hb_change_sensitivity <- hb_table %>%
  filter(cliniccode %in% c(intervention_clinics_did, nonintervention_clinics_did))
sensitivity_hb_change <- lm(hb_change ~ treated + post + treated:post, data = hb_change_sensitivity)
summary(sensitivity_hb_change)
confint(sensitivity_hb_change) #CI

hb_change_sensitivity %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(hb_change, na.rm = TRUE),
    sd_hb_change = sd(hb_change, na.rm = TRUE)
  )
```
```{r}
# Create dataset
did_data_new_2 <- tibble(
  # Column representing the Y-axis outcomes/models
  Model = c(
    "13 Intervention: All Non-intervention",
    "13 Intervention: 13 Non-intervention",
    "Subgroup: Gabrone",
    "Subgroup: Molepolole",
    "Subgroup: Palapye",
    "Subgroup: HIV +",
    "Subgroup: HIV -"
  ),
  # DiD Estimate (change in hemoglobin column)
  Estimate = c(-0.048, -0.061, -0.126, 0.085, -0.054, -0.005, -0.053),
  # Lower CI (2.5 %)
  Lower_CI = c(-0.237, -0.279, -0.431, -0.310, -0.588, -0.548, -0.254),
  # Upper CI (97.5 %)
  Upper_CI = c(0.142, 0.157, 0.180, 0.481, 0.481, 0.539, 0.148)
) %>%
  mutate(
    # Check if the confidence interval crosses zero (Sig if both bounds are > 0 or < 0)
    Significance = if_else(
      condition = (Lower_CI * Upper_CI) > 0,
      true = "Sig",
      false = "n.s."
    ),
    # Ensure the outcomes are plotted in the correct order (reversing the list order for presentation)
    Model = fct_inorder(Model)
  )

print(did_data_new_2)

forest_plot_new_2 <- ggplot(
  data = did_data_new_2, 
  aes(
    x = Estimate, 
    y = Model, 
    color = Model, 
    shape = Model
  )
) +
  scale_color_viridis_d() +
  # A: 95% CI
  geom_errorbarh(aes(xmin = Lower_CI, xmax = Upper_CI), height = 0.2) +
  
  # B: point estimate
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  
  # C: Set X range and label 
  scale_x_continuous(
    name = "DiD Estimate (g/dL, 95% CI)",
    limits = c(-0.6, 0.6), 
    breaks = seq(-0.6, 0.6, by = 0.2),
    labels = function(x) sprintf("%.1f", x) 
  ) +
  
  # D: tidy pattern
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(), 
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  ) +
  
  # E: Y axis label 
  labs(y = NULL, title = "Difference-in-Differences Results for Change in Hemoglobin")

print(forest_plot_new_2)
```

### Mean change in second hemoglobin (H2)

During the intervention period, there were 710 pregnant women in intervention clinics and 1,213 in non-intervention clinics who had a hemoglobin measured on or after 24 weeks gestation, During the pre-intervention period, there were 3,178 pregnant women and 6,437 pregnant women who had a hemoglobin measured on or after 24 weeks gestation. The overall DiD estimate was 0.016 g/dL (95% CI: –0.129 to 0.160), with similar findings in the sensitivity and subgroup analyses. 
```{r}
# Filter for those with a second hb value
secondhb_table <- combined_table %>%
  filter(!is.na(secondhb))
# Run DID model
secondhb_did_model <- lm(secondhb ~ treated + post + treated:post, data = secondhb_table)
summary(secondhb_did_model)
# Group summaries
secondhb_table %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_secondhb = mean(secondhb, na.rm = TRUE),
    sd_secondhb = sd(secondhb, na.rm = TRUE)
  )
#CI
confint(secondhb_did_model)
```

```{r}
## Subgroup Analysis
hb2_table_sub <- secondhb_table %>%
  mutate(region = case_when(
    cliniccode %in% c(5302, 5303, 5306, 5318, 5323) ~ "Molepolole",     #compare tp all other noninterventions 
    cliniccode %in% c(15312, 15325, 15329, 15331, 15335, 15341) ~ "Gaborone",
    cliniccode %in% c(3201, 3319) ~ "Palapye",
    TRUE ~ "Other"
  ))
##Gaborone
gabs_did_hb2 <- secondhb_table %>%
  filter(cliniccode %in% c(intervention_Gab_subset, nonintervention_Gab_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Gab_subset, 1, 0))

gabs_model_hb2 <- lm(secondhb ~ treated + post + treated:post, data = gabs_did_hb2)
summary(gabs_model_hb2)
#CI
confint(gabs_model_hb2)
#get 4 variables listed
gabs_did_hb2 %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb = mean(secondhb, na.rm = TRUE),
    sd_hb = sd(secondhb, na.rm = TRUE)
  )
##Molepolole
mole_did_hb2 <- secondhb_table %>%
  filter(cliniccode %in% c(intervention_Mol_subset, nonintervention_Mol_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Mol_subset, 1, 0))

mole_model_hb2 <- lm(secondhb ~ treated + post + treated:post, data = mole_did_hb2)
summary(mole_model_hb2)
#CI
confint(mole_model_hb2)
#get 4 variables listed
mole_did_hb2 %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean = mean(secondhb, na.rm = TRUE),
    sd = sd(secondhb, na.rm = TRUE)
  )

##Palapye
pal_did_hb2 <- secondhb_table %>%
  filter(cliniccode %in% c(intervention_Pal_subset, nonintervention_Pal_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Pal_subset, 1, 0))

pal_model_hb2 <- lm(secondhb ~ treated + post + treated:post, data = pal_did_hb2)
summary(pal_model_hb2)
#CI
confint(pal_model_hb)
#get 4 variables listed
pal_did_hb2 %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb = mean(secondhb, na.rm = TRUE),
    sd_hb = sd(secondhb, na.rm = TRUE)
  )

##HIV status
# HIV-positive subset
hb2_hivpos <- secondhb_table %>% filter(hivpos == 1)     #use dataset from total analysis
# HIV-negative subset
hb2_hivneg <- secondhb_table %>% filter(hivpos == 0)
# DID 
# For HIV-positive
did_hivpos_hb2 <- lm(secondhb ~ treated + post + treated:post, data = hb2_hivpos)
summary(did_hivpos_hb2)
confint(did_hivpos_hb2) #CI
secondhb_table %>% # 4 proportions, need to change the dataset to be consistent
  filter(hivpos == 1) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb = mean(secondhb, na.rm = TRUE),
    sd_hb = sd(secondhb, na.rm = TRUE)
  )

# For HIV-negative
did_hivneg_hb2 <- lm(secondhb ~ treated + post + treated:post, data = hb2_hivneg)
summary(did_hivneg_hb2)
confint(did_hivneg_hb2)
secondhb_table %>% # 4 proportions
  filter(hivpos == 0) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb = mean(secondhb, na.rm = TRUE),
    sd_hb = sd(secondhb, na.rm = TRUE)
  )

## Sensitivity Analysis
secondhb_sensitivity <- secondhb_table %>%
  filter(cliniccode %in% c(intervention_clinics_did, nonintervention_clinics_did))
sensitivity_secondhb <- lm(secondhb ~ treated + post + treated:post, data = secondhb_sensitivity)
summary(sensitivity_secondhb)
confint(sensitivity_secondhb) #CI

secondhb_sensitivity %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    mean_hb_change = mean(secondhb, na.rm = TRUE),
    sd_hb_change = sd(secondhb, na.rm = TRUE)
  )
```
```{r}
# Create dataset
did_data_new_3 <- tibble(
  # Column representing the Y-axis outcomes/models
  Model = c(
    "13 Intervention: All Non-intervention",
    "13 Intervention: 13 Non-intervention",
    "Subgroup: Gabrone",
    "Subgroup: Molepolole",
    "Subgroup: Palapye",
    "Subgroup: HIV +",
    "Subgroup: HIV -"
  ),
  # DiD Estimate (change in hemoglobin column)
  Estimate = c(0.016, -0.007, -0.076, -0.090, 0.336, 0.081, -0.011),
  # Lower CI (2.5 %)
  Lower_CI = c(-0.129, -0.176, -0.294, -0.428, -0.588, -0.331, -0.161),
  # Upper CI (97.5 %)
  Upper_CI = c(0.160, 0.161, 0.142, 0.248, 0.481, 0.493, 0.139)
) %>%
  mutate(
    # Check if the confidence interval crosses zero (Sig if both bounds are > 0 or < 0)
    Significance = if_else(
      condition = (Lower_CI * Upper_CI) > 0,
      true = "Sig",
      false = "n.s."
    ),
    # Ensure the outcomes are plotted in the correct order (reversing the list order for presentation)
    Model = fct_inorder(Model)
  )

print(did_data_new_3)

forest_plot_new_3 <- ggplot(
  data = did_data_new_3, 
  aes(
    x = Estimate, 
    y = Model, 
    color = Model, 
    shape = Model
  )
) +
  scale_color_viridis_d() +
  # A: 95% CI
  geom_errorbarh(aes(xmin = Lower_CI, xmax = Upper_CI), height = 0.2) +
  
  # B: point estimate
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  
  # C: Set X range and label 
  scale_x_continuous(
    name = "DiD Estimate (g/dL, 95% CI)",
    limits = c(-0.6, 0.6), 
    breaks = seq(-0.6, 0.6, by = 0.2),
    labels = function(x) sprintf("%.1f", x) 
  ) +
  
  # D: tidy pattern
  theme_bw() +
  theme(
    panel.grid.major.y = element_blank(), 
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  ) +
  
  # E: Y axis label 
  labs(y = NULL, title = "Difference-in-Differences Results for Change in Second Hemoglobin")

print(forest_plot_new_3)
```

### Adverse Birth Outcomes 

In the pre-intervention period, the risk of the composite adverse birth outcome was 30.8% for the intervention clinics and 30.2% for the non-intervention clinics. During the intervention period, the risk of the composite outcome was 30.4% for the intervention clinics and 31.4% for the non-intervention clinics. The resulting DiD estimate was -1.7% (95% CI: -5.2%, 1.9%). Results were consistent when restricting the non-intervention clinics to the 13 clinics originally eligible and in the subgroup analyses by region and HIV status. 
```{r}
Interv_BD_T0 <- Interv_BD_T0 %>%
  mutate(treated = 1, post = 0)

Interv_BD_T1 <- Interv_BD_T1 %>%
  mutate(treated = 1, post = 1)

noninterv_BD_T0 <- noninterv_BD_T0 %>%
  mutate(treated = 0, post = 0)

noninterv_BD_T1 <- noninterv_BD_T1 %>%
  mutate(treated = 0, post = 1)

BD_table <- bind_rows(Interv_BD_T0, Interv_BD_T1, noninterv_BD_T0, noninterv_BD_T1)

BD_model <- lm(composite ~ treated + post + treated:post, data = BD_table)
summary(BD_model)

# Check proportion with birth composite outcome by treated x post group
BD_table %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    n_composite = sum(composite == 1, na.rm = TRUE),
    proportion_composite = mean(composite == 1, na.rm = TRUE)
  )
#CI
confint(BD_model)
```

```{r}
## Subgroup Analysis
BD_table_sub <- BD_table %>%
  mutate(region = case_when(
    cliniccode %in% c(5302, 5303, 5306, 5318, 5323) ~ "Molepolole",     #compare tp all other noninterventions 
    cliniccode %in% c(15312, 15325, 15329, 15331, 15335, 15341) ~ "Gaborone",
    cliniccode %in% c(3201, 3319) ~ "Palapye",
    TRUE ~ "Other"
  ))

##Gaborone
gabs_did_BD <- BD_table %>%
  filter(cliniccode %in% c(intervention_Gab_subset, nonintervention_Gab_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Gab_subset, 1, 0))

gabs_model_BD <- lm(composite ~ treated + post + treated:post, data = gabs_did_BD)
summary(gabs_model_BD)
#CI
confint(gabs_model_BD)
#get 4 variables listed
gabs_did_BD %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    n_composite = sum(composite == 1, na.rm = TRUE),
    proportion_composite = mean(composite == 1, na.rm = TRUE)
  )

##Molepolole
mole_did_BD <- BD_table %>%
  filter(cliniccode %in% c(intervention_Mol_subset, nonintervention_Mol_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Mol_subset, 1, 0))

mole_model_BD <- lm(composite ~ treated + post + treated:post, data = mole_did_BD)
summary(mole_model_BD)
#CI
confint(mole_model_BD)
#get 4 variables listed
mole_did_BD %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    n_composite = sum(composite == 1, na.rm = TRUE),
    proportion_composite = mean(composite == 1, na.rm = TRUE)
  )

##Palapye
pal_did_BD <- BD_table %>%
  filter(cliniccode %in% c(intervention_Pal_subset, nonintervention_Pal_subset)) %>%
  mutate(treated = ifelse(cliniccode %in% intervention_Pal_subset, 1, 0))

pal_model_BD <- lm(composite ~ treated + post + treated:post, data = pal_did_BD)
summary(pal_model_BD)
#CI
confint(pal_model_BD)
#get 4 variables listed
pal_did_BD %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    n_composite = sum(composite == 1, na.rm = TRUE),
    proportion_composite = mean(composite == 1, na.rm = TRUE)
  )

##HIV status
# HIV-positive subset
BD_hivpos <- BD_table %>% filter(hivpos == 1)     #use dataset from total analysis
# HIV-negative subset
BD_hivneg <- BD_table %>% filter(hivpos == 0)
# DID 
# For HIV-positive
did_hivpos_BD <- lm(composite ~ treated + post + treated:post, data = BD_hivpos)
summary(did_hivpos_BD)
confint(did_hivpos_BD) #CI
BD_table %>% # 4 proportions, need to change the dataset to be consistent
  filter(hivpos == 1) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    prop_composite = mean(composite == 1, na.rm = TRUE)
  )

# For HIV-negative
did_hivneg_BD <- lm(composite ~ treated + post + treated:post, data = BD_hivneg)
summary(did_hivneg_BD)
confint(did_hivneg_BD)
BD_table %>% # 4 proportions
  filter(hivpos == 0) %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    prop_composite = mean(composite == 1, na.rm = TRUE)
  )

## Sensitivity Analysis
BD_sensitivity <- BD_table %>%
  filter(cliniccode %in% c(intervention_clinics_did, nonintervention_clinics_did))

sensitivity_model_BD <- lm(composite ~ treated + post + treated:post, data = BD_sensitivity)
summary(sensitivity_model_BD)
#CI
confint(sensitivity_model_BD)
#get 4 variables listed
BD_sensitivity %>%
  group_by(treated, post) %>%
  summarise(
    n = n(),
    n_composite = sum(composite == 1, na.rm = TRUE),
    proportion_composite = mean(composite == 1, na.rm = TRUE)
  )
```

# Conclusion
In this pilot evaluation of transitioning from IFA to MMS within antenatal care services in Botswana, findings indicate that implementation of MMS is both feasible and operationally acceptable. The program achieved a 47.4% relative increase in effective coverage compared with the pre-intervention period, highlighting strong potential for scale-up within routine maternal health systems.

Preliminary effectiveness results suggest a favorable trend toward reduced adverse birth outcomes following MMS introduction. Although the observed reduction (–1.7%) did not reach statistical significance in this early phase, ongoing data collection and larger sample sizes

Safety assessments showed non-inferiority in maternal hemoglobin levels among women receiving MMS relative to those receiving IFA (DiD: –0.048 g/dL), addressing concerns related to MMS’s lower iron content. These findings support the continued use of MMS without increased risk of iron deficiency.

Overall, the pilot provides encouraging early evidence across feasibility, effectiveness, and safety domains. Continued monitoring and expanded analyses are warranted to confirm clinical impact and to inform national decision-making regarding full-scale adoption of MMS in Botswana’s antenatal care program.

# References
1.	Darnton-Hill I, Mkparu UC. Micronutrients in pregnancy in low- and middle-income countries. Nutrients.MDPI AG. 2015;7(3):1744-1768. doi:10.3390/nu7031744
2.	Bourassa MW, Osendarp SJM, Adu-Afarwuah S, et al. Review of the evidence regarding the use of antenatal multiple micronutrient supplementation in low- and middle-income countries. Ann N Y Acad Sci.Blackwell Publishing Inc. 2019;1444(1):6-21. doi:10.1111/nyas.14121
3.	World Health Organization. Anemia. February 10, 2025. Accessed July 13, 2025. https://www.who.int/news-room/fact-sheets/detail/anaemia.
4.	Expert consensus on an open-access United Nations International Multiple Micronutrient Antenatal Preparation–multiple micronutrient supplement product specification. Ann N Y Acad Sci. 2020;1470(1):3-13. doi:10.1111/nyas.14322
5.	West KP, Shamim AA, Mehra S, et al. Effect of Maternal multiple micronutrient vs iron-folic acid supplementation on infant mortality and adverse birth outcomes in rural Bangladesh: The JiVitA-3 randomized trial. JAMA - Journal of the American Medical Association. 2014;312(24):2649-2658. doi:10.1001/jama.2014.16819
6.	Verney AMJ, Busch-Hallen JF, Walters DD, Rowe SN, Kurzawa ZA, Arabi M. Multiple micronutrient supplementation cost–benefit tool for informing maternal nutrition policy and investment decisions. Matern Child Nutr. 2023;19(4). doi:10.1111/mcn.13523
7.	Alfiani F, Utami AM, Zakiyah N, Daud NAA, Suwantika AA, Puspitasari IM. Cost-Effectiveness Analysis of Multiple Micronutrient Supplementation (MMS) Compared to Iron Folic Acid (IFA) in Pregnancy: A Systematic Review. Int J Womens Health. 2025;17:639-649. doi:10.2147/IJWH.S489159
8.	Ramotsababa M, Setlhare V. Late registration for antenatal care by pregnant women with previous history of caesarean section. Published online 2021. doi:10.4102/phcfm
9.	World Bank Group. Botswana. https://ilostat.ilo.org/data/.
10.	Kapoor A, Mussa A, Diseko M, et al. Cross-sectional trends in HIV prevalence among pregnant women in Botswana: an opportunity for PrEP? J Int AIDS Soc. 2022;2022:25892. doi:10.1002/jia2.25892/full
11.	Caniglia EC, Zash R, Swanson SA, et al. Iron, folic acid, and multiple micronutrient supplementation strategies during pregnancy and adverse birth outcomes in Botswana. Lancet Glob Health. 2022;10(6):e850-e861. doi:10.1016/S2214-109X(22)00126-7
12.	Kebaabetswe P, Diseko M, Zash R, et al. A qualitative assessment of barriers to iron and folic acid supplementation among pregnant women in Botswana. BMC Public Health. 2024;24(1). doi:10.1186/s12889-024-20685-5
13.	WHO Antenatal Care Recommendations for a Positive Pregnancy Experience Nutritional Interventions Update: Multiple Micronutrient Supplements during Pregnancy. Accessed July 13, 2025. https://www.who.int/news-room/fact-sheets/detail/anaemia.
14.	Ba A, Fox MJ, Keita AM, et al. Qualitative evaluation of a package of implementation strategies codesigned to support the introduction of multiple micronutrient supplementation (MMS) for pregnant women in Bamako, Mali. Matern Child Nutr. Published online January 1, 2024. doi:10.1111/mcn.13712
15.	Battung SM, Groen H, van der Beek EM. Prenatal multiple micronutrient supplementation in the Parepare district, Indonesia; population characteristics and intake adherence. BMC Public Health. 2025;25(1):983. doi:10.1186/s12889-025-22129-0
16.	Hossain M, Sanin KI, Haque MA, et al. Evaluating the impact of a countrywide, market-based roll-out of multiple micronutrient supplementation on low birth weight in Bangladesh: Protocol for a two-arm, quasi-experimental and mixed-methods evaluation study. BMJ Open. 2022;12(5). doi:10.1136/bmjopen-2021-060230
17.	Weiner BJ, Lewis CC, Stanick C, et al. Psychometric assessment of three newly developed implementation outcome measures. Implementation Science. 2017;12(1). doi:10.1186/s13012-017-0635-3
18.	Proctor E, Silmere H, Raghavan R, et al. Outcomes for implementation research: Conceptual distinctions, measurement challenges, and research agenda. Administration and Policy in Mental Health and Mental Health Services Research. 2011;38(2):65-76. doi:10.1007/s10488-010-0319-7
 