---
title: "BMIN503/EPID600 Final Project"
author: "Dominique Medaglio"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***


### Overview

> Medication nonadherence amongst diabetics is a widespread issue in the United States, and differences in insurance coverage (which differs by residential location) may be a contributing factor. The purpose of this project is to determine if there are differences in cost-related nonadherence (CRN) prevalence based on urban/rural residence. I will acheive this aim by creating static maps and logistic regression models based on adult respondents of the 2019 National Health Interview Survey (NHIS).

>I have spoken to Mary Regina Boland, Robert Gross, Sean Hennessy, and Karen Glanz. With Mary Regina, I learned about special considerations for using regional data, which is what I'll be using for my maps. Robert and Sean provided guidance on how to interpret CRN measures appropriately. Karen provided overall advice on adherence behaviors as a part of self-care behaviors, and variables that I should consider for my models.

>The link to the final GitHub repository can be found here: https://github.com/dmedaglio/BMIN503_Final_Project


### Introduction 

>Medication adherence is a mainstay to the management of chronic conditions. Medication error, including the non-initiation of life-saving treatment, is one of the top reasons why patients are readmitted to the hospital. Unfortunately, medication nonadherence is very common; it is estimated that 30-50% of medications are not taken as prescribed.

>Cost is a common reason why patients do not their medications as they should. This can especially be compounded if patients take multiple medications to manage a condition, such as common with diabetes. Insurance coverage can be important to help many afford their medications, yet sufficient coverage may vary by where patients live. In urban areas, one is more likely to benefit from more provider competition in the Marketplace and have lower premiums. In rural areas, such competition may not exist, resulting in higher premiums on average. It is unknown if urban/rural residence also has an impact on the affordability of prescription medications.

>This research question requires an interdisciplinary approach, given the need for geospacial approaches, pharmacoepi expertise, and an understanding of health policy. It will be important to accurately interpreret NHIS survey responses, as CRN is only one form of medication nonadherence. Additionally, the NHIS is administered using geographically clustered sampling techniques, so static maps will likely zoom in on these clusters. Finally, an understanding of policies around the health insurance marketplace will be needed to interpret access to pharmacy benefits. 


### Methods

>I will use the 2019 responses of the National Health Interview Survey, focusing on adult responses only. Adult responses will include the Sample Adult (individual selected to respond to all household questions) or their proxy, if that proxy lives in the same household. 

>The available dataset had to be manipulated for the needs of this project. A number of relevant predictors were selected, including participant demographics, urban/rural residence, health insurance variables, medication variables, and the adherence outcomes.I pulled all 4 CRN related survey questions to serve as my outcomes of interest: not filled a medication due to cost, delaying a medication due to cost, skipping a dose due to cost, and taking less of a medication due to cost. 

>Variables with multiple levels were collasped to fewer levels, to allow for easier interpretation.


```{r}
library(readxl)
adult19 <- read_excel("C:/Users/dmcom/Documents/NHISadult2019.xlsx")
str(adult19)


library(dplyr)
library(forcats)
NHIS19 <- adult19%>%
  mutate(urbanrur = factor(URBRRL, levels = c(1,2,3,4), labels = c("Central Metro", "Fringe Metro", "MedSmall Metro", "Nonmetro")),
         age = AGEP_A, #all values above 85 are NOT ages but refusal/don't know
         sex = factor(SEX_A, levels = c(1,2), labels = c("Male", "Female")),
         edu = factor(EDUC_A, levels = c(00, 01, 02, 03, 04, 05, 06, 07, 08, 09, 10, 11), labels = c("Never attend/K only", "Grade 1-11", "Grade 12, no diploma", "GED or equiv", "High school grad", "Some college, no degree", "Assoc: occup/tech/voc", "Assoc: academ", "Bachelors", "Masters", "Professional", "Doctoral")),
         raceeth = factor(HISPALLP_A, levels = c(01, 02, 03, 04, 05, 06, 07), labels = c("Hispanic", "NH White", "NH Black", "NH Asian", "NH AIAN only", "NH AIAN and other", "Other/Multiple")),
         htnmed = factor(HYPMED_A, levels = c(1,2), labels = c("Yes", "No")),
         cholmed = factor(CHLMED_A, levels = c(1,2), labels = c("Yes", "No")),
         diabpill = factor(DIBPILL_A, levels = c(1,2), labels = c("Yes", "No")),
         insulin = factor(DIBINS_A, levels = c(1,2), labels = c("Yes", "No")), 
         insurance = factor(COVER_A, levels = c(1,2,3,4), labels = c("Private", "Public", "Other", "Uninsured")),
         ins_65 = factor(COVER65_A, levels = c(1,2,3,4,5,6), labels = c("Private", "Dual eligible", "Medicare Advan", "Medicare", "Other", "Uninsured")),
         mktplace = factor(MAXCHNG_A, levels = c(1,2), labels = c("Yes", "No")),
         skipmed = factor(RXSK12M_A, levels = c(1,2), labels = c("Yes", "No")),
         lessmed = factor(RXLS12M_A, levels = c(1,2), labels = c("Yes", "No")),
         delaymed = factor(RXDL12M_A, levels = c(1,2), labels = c("Yes", "No")),
         nofill = factor(RXDG12M_A, levels = c(1,2), labels = c("Yes", "No")),
         depressmed = factor(DEPMED_A, levels = c(1,2), labels = c("Yes", "No")),
         resregion = factor(REGION, levels = c(1,2,3,4), labels = c("Northeast", "Midwest", "South", "West"))) %>%
  select (urbanrur, HHRESPSA_FLG, age, sex, edu, raceeth, htnmed, cholmed, diabpill, insulin, insurance, ins_65, mktplace, RX12M_A, skipmed, lessmed, delaymed, nofill, depressmed, resregion) %>%
  filter(HHRESPSA_FLG == 1, RX12M_A == 1) #this filter ensures that data only reflects responses from the randomly selected Adult or their proxy who lives in the household. this also filters for only adults who took at least one prescription in the past 12 months

NHIS19$edu <- fct_collapse(NHIS19$edu,
                       "No HS Diploma" = c("Never attend/K only", "Grade 1-11", "Grade 12, no diploma"),
                       "Diploma or GED" = c("GED or equiv", "High school grad"), 
                       "Associates" = c("Assoc: occup/tech/voc", "Assoc: academ"), 
                       "Grad Degree" = c("Masters", "Professional", "Doctoral")) 

NHIS19$raceeth <- fct_collapse(NHIS19$raceeth,
                               "Other" = c("NH Asian", "NH AIAN only", "NH AIAN and other", "Other/Multiple"))
```
### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

>As there are a number of cost-related nonadherence outcomes, descriptive tables were created for each outcome.

```{r}
library(table1)
label(NHIS19$age) <- "Age"
label(NHIS19$sex) <- "Sex"
label(NHIS19$raceeth) <- "Race/Ethnicity"
label(NHIS19$edu) <- "Education"
label(NHIS19$urbanrur) <- "Urban/Rural Classification"
label(NHIS19$resregion) <- "US Region"
label(NHIS19$htnmed) <- "Hypertention medication"
label(NHIS19$cholmed) <- "Cholesterol medication"
label(NHIS19$diabpill) <-"Diabetes medication, oral"
label(NHIS19$insulin) <- "Insulin"
label(NHIS19$depressmed) <- "Antidepressant"
label(NHIS19$insurance) <- "Primary insurance type"
label(NHIS19$mktplace) <- "Used the Marketplace for insurance"
label(NHIS19$nofill) <- "Ever Not Fill Med Due to Cost?"
label(NHIS19$skipmed) <- "Ever Skip Med Due to Cost?"
label(NHIS19$lessmed) <- "Ever Take Less Med Due to Cost?"
label(NHIS19$delaymed) <- "Ever Delay a Med Due to Cost?"

tbl_nofill <- NHIS19 %>%
  select(age, sex, edu, raceeth, htnmed, cholmed, diabpill, insulin, depressmed, insurance, mktplace, nofill, resregion, urbanrur)

tbl_delay <- NHIS19 %>%
  select(age, sex, edu, raceeth, htnmed, cholmed, diabpill, insulin, depressmed, insurance, mktplace, delaymed, resregion, urbanrur)

tbl_skip <- NHIS19 %>%
  select(age, sex, edu, raceeth, htnmed, cholmed, diabpill, insulin, depressmed, insurance, mktplace, skipmed, resregion, urbanrur)

tbl_less <- NHIS19 %>%
  select(age, sex, edu, raceeth, htnmed, cholmed, diabpill, insulin, depressmed, insurance, mktplace, lessmed, resregion, urbanrur)


#Descriptive statistics for the four CRN outcomes

library(gtsummary)
tbl_nofill %>% 
  tbl_summary(by = nofill) %>%
  add_p() %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "Med Not Filled")%>%
  bold_labels()
 
tbl_delay %>% 
  tbl_summary(by = delaymed) %>%
  add_p() %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "Delayed a Med")%>%
  bold_labels()

tbl_less %>% 
  tbl_summary(by = lessmed) %>%
  add_p() %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "Took Less of a Med")%>%
  bold_labels()

tbl_skip %>% 
  tbl_summary(by = skipmed) %>%
  add_p() %>%
  modify_spanning_header(c("stat_1", "stat_2") ~ "Skipped a Med")%>%
  bold_labels()

```

>Most of the variables included in the tables were statistically significant in the univariate analyses. Younger age, female sex, and lower education levels were all associated with more CRN. Respondents with a hypertension or a cholesterol medication were more likely to not report CRN, yet those on insulin or an antidepressant were more likely to report CRN. Higher proportions of those with public insurance, other types of insurance, and those who were uninsured reported CRN more often.

>Visualizations for urban/rural residence (the exposure of interest) were then created for each CRN outcome. As the exposure and outcomes are both categorical data, bar graphs were selected. Graphs were normalized to 100% to more readily show proportional differences between the urban/rural categories.

```{r}
library(ggplot2)

p_nofill <-ggplot(data = NHIS19, aes(x = urbanrur, fill = factor(nofill)))+
  geom_bar(position = "fill")+
  labs(title = "CRN by Urban/Rural Category")+
  labs(x = "Urban/Rural Classification", y = "Percentage")
  
p_nofill + labs(fill = "Did Not Fill?")


p_nofill <-ggplot(data = NHIS19, aes(x = urbanrur, fill = factor(delaymed)))+
  geom_bar(position = "fill")+
  labs(title = "CRN by Urban/Rural Category")+
  labs(x = "Urban/Rural Classification", y = "Percentage")
  
p_nofill + labs(fill = "Delayed a Med?")


p_nofill <-ggplot(data = NHIS19, aes(x = urbanrur, fill = factor(lessmed)))+
  geom_bar(position = "fill")+
  labs(title = "CRN by Urban/Rural Category")+
  labs(x = "Urban/Rural Classification", y = "Percentage")
  
p_nofill + labs(fill = "Took Less of a Med?")


p_nofill <-ggplot(data = NHIS19, aes(x = urbanrur, fill = factor(skipmed)))+
  geom_bar(position = "fill")+
  labs(title = "CRN by Urban/Rural Category")+
  labs(x = "Urban/Rural Classification", y = "Percentage")
  
p_nofill + labs(fill = "Skipped a Med?")

```

>In all four bar graphs, it is clear that the majority of respondants do not report CRN. However, as you move from the most urban classification (Central Metro) to the most rural classification (Nonmetro), the proportion of respondants who do report CRN increases.

>Next, logistic regression models for each outcome were built. Variables that were significant in the univariate analyses were included in each respective model. the 'nofill', 'delaymed' and 'lessmed' models had the same inputs, while 'skipmed' had 2 differences (the exclusion of the antihypertensives variable and the inclusion of the Marketplace variable).

```{r}
m_nofill <- glm(nofill ~ age + sex + raceeth + edu + urbanrur + resregion + htnmed + cholmed + insulin + depressmed + insurance, data = tbl_nofill, family = binomial())

m_delay <- glm(delaymed ~ age + sex + raceeth + edu + urbanrur + resregion + htnmed + cholmed + insulin + depressmed + insurance, data = tbl_delay, family = binomial())

m_skip <- glm(skipmed ~ age + sex + raceeth + edu +  urbanrur + resregion + cholmed + diabpill + insulin + depressmed + insurance + mktplace, data = tbl_skip, family = binomial())

m_less <- glm(lessmed ~ age + sex + raceeth + edu + urbanrur + resregion + htnmed + cholmed + insulin + depressmed + insurance, data = tbl_less, family = binomial())

#tbl_regression takes all 4 models and formats results into tables

m_nofill %>%
  tbl_regression(
    exponentiate = TRUE,
    pvalue_fun = ~style_pvalue(.x, digits = 2),
    ) %>%
  bold_labels()
  

m_delay %>%
  tbl_regression(
    exponentiate = TRUE,
    pvalue_fun = ~style_pvalue(.x, digits = 2),
    ) %>%
  bold_labels()

m_less %>%
  tbl_regression(
    exponentiate = TRUE,
    pvalue_fun = ~style_pvalue(.x, digits = 2),
    ) %>%
  bold_labels()

m_skip %>%
  tbl_regression(
    exponentiate = TRUE,
    pvalue_fun = ~style_pvalue(.x, digits = 2),
    ) %>%
  bold_labels()

```

>In all four regression models, urban/rural residence was not statistically significant, yet other variables remained significant in the models. Respondents who were uninsured (relative to having private insurance) were less likely to not fill a med, delay a med, or take less of a med due to cost. Also, respondents who were on insulin were significantly more likely to report the same three forms of CRN, compared to those not on insulin.

>There are a number of limitations that should be considered in the interpretation of these results. First, CRN is only one aspect of medication nonadherence: people can not adhere to a medication regimen for a number of reasons (forgetfulness, health literacy, lack of trust, etc.). The NHIS is a cross-sectional survey, which only captures a limited perspective of the respondant's experiences. There is very likely unmeasured confounding that the survey does not capture, yet is important to understanding CRN. Finally, there may be misclassification bias, as urban/rural residence is measured based on where the respondant currently lives, and not necessarily where they spent most of their residence across the past year.