---
title: "Food Insecurity in the United States"
author: "Nawar Naseer"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

### Overview
Food insecurity, or a "lack of consistent access to enough food for an active, healthy life", is a major public health issue in the United States, where approximately 11% of households experienced food insecurity in 2018 (1). The United States Department of Agriculture (USDA) publishes a Food Environment Atlas, which contains data on over 275 variables with respect to food insecurity, including access to food retail, use of government food assistance programs, health and socioeconomic indicators, etc. These data are collected from various sources, span several years and geographic levels, and have been made publicly available. The goal of this project is to assess how the food retail environment varies across time and space, and to test for associations between access to food and the development of adverse health outcomes. In addition, the effect of other variables such as socioeconomic indicators and the use of government assistance programs such as SNAP will be incorporated into the analyses. Geospatial differences will be analyzed, with a specific focus on Philadelphia in comparision to 1) other counties in Pennslyvania, and 2) other large cities in the United States. 

[Github Repository](insert url for final github repo)

### Introduction
In the United States, approximately 11% of households were food insecure in 2018 (1). In Philadelphia, this rate is nearly double the national average at 18.3%, and more than one out of every five Philadelphian struggles with basic food access (2). Food insecurity contributes to the development of chronic diseases such as heart disease and diabetes, both of which are two of the United State’s top ten leading causes of death (3). Additionally, food insecurity affects children’s cognitive development, and their ability to excel in school (4). Thus, food insecurity has long-lasting impacts on the productivity of a population, and understanding the regional factors that contribute to food insecurity and adverse health outcomes is vital for creating effective local and national interventions. 

A food secure environment provides both physical and financial access to nutritious food. By analyzing the food retail environment in context of socioeconmic factors and government assistance programs, this project spans the disciplines of sociology, economics, geography, public policy, and public health. Based on consulting with Dr. Alyson Karpyn, an analysis on a vulnerable but previously under-researched population, senior citizens, has also been incorporated. This project thus aims to use a multidisciplinary approach to understanding factors influencing food insecurity, and in doing so, may provide critical insight on how to tackle food insecurity in the United States.

### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

The main data set used for this project was obtained from the USDA's publicly available data on the [Food Environment Atlas](https://www.ers.usda.gov/data-products/food-environment-atlas/data-access-and-documentation-downloads/). The downloaded Excel file contained several sheets pertinent to this project and have been loaded into R as a list of dataframes called "mysheets". These dataframes contain county-specific information on access and proximity to food retail providers (grocery stores, convenience stores, farmers' markets, etc), food assistance usage, food insecurity rates, diabetes and obesity rates, and socioeconomic indicators including race, age, income, etc. 

All necessary packages will be installed and loaded first.
```{r, eval=FALSE}
#Load all necessary packages/libraries
install.packages("readxl")
library(readxl)
library(dplyr)
library(tidyverse)
```

The data downloaded has been saved into the working directory and is loaded into R. Note that names of some of the sheets were slightly modified in Excel to exclude spaces and make them R-compatible
```{r, eval=FALSE}
#Load in data
#The following function allows one to read in multiple sheets from one Excel file into a list, with each sheet representing a separate dataframe within the list.
read_excel_allsheets <- function(filename, tibble = TRUE) {
    # I prefer straight data.frames
    # but if you like tidyverse tibbles (the default with read_excel)
    # then just pass tibble = TRUE
    sheets <- readxl::excel_sheets(filename)
    x <- lapply(sheets, function(X) readxl::read_excel(filename, sheet = X))
    if(!tibble) x <- lapply(x, as.data.frame)
    names(x) <- sheets
    x
}
mysheets <- read_excel_allsheets("USDAFoodEnvironmentAtlasMarch2018.xls")
head(mysheets$ACCESS) #to check that data has been loaded properly
```

The next step is to clean the data using the dplyr and tidyverse packages. Each dataframe will be updated to only include variables that will be analyzed in this project. 
```{r, eval=TRUE}

```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

### References

1. Coleman-Jensen, A., Rabbitt, M. P., Gregory, C. A., & Singh, A. (2019, September). Household Food Security in the United States in 2018. Retrieved November 6, 2019, from https://www.ers.usda.gov/publications/pub-details/?pubid=94848.
2. Hunger Free America. Philadelphia Falling Behind in Freedom to Eat: Greater Philadelphia Hunger Report, 2018.
3. Centers for Disease Control and Prevention, Leading Causes of Death, <https://www.cdc.gov/nchs/fastats/leading-causes-of-death.htm>
4. Feeding America. Help Kids Facing Hunger This School Year, <https://www.feedingamerica.org/hunger-blog/help-kids-facing-hunger-this>
