---
title: 'Risk Analyses and Prediction of US Adult Chronic Diseases'
author: 'Final Project of Data Science for Biomedical Informatics (BMIN503/EPID600) <br>Instructor: Blanca Himes <br><br> Nianfu Song '
output: 
  html_document:
  toc: FALSE
depth: 3
theme: paper 
highlight: tango
---

```{r set-options, echo=FALSE, cache=FALSE}

knitr::opts_chunk$set(warning=FALSE, message=FALSE, error=FALSE, width = 80)
options(width = 80)
```  

```{r libs, echo=FALSE, cache=FALSE, message = FALSE, warning=FALSE}
# install.packages("MatchIt")
# install.packages('Zelig')
library("MatchIt")
library("Zelig")
library(tidyverse)
library(tidycensus)
library(datasets)
library(dplyr)

library(sf)
library(ggmap)
library(ggplot2)
library(gplots)
library(GGally)
library(cowplot)
library(RColorBrewer)
library(leaflet)
#library(xtable)
#library("expss")

library(randomForest)
library(pROC)
library(gplots)

```    
# 1. Introduction

## 1.1. Overview
The goal of my final project is to use survey data to analyze health related factors and predict potential risk of diseases. The National Health Interview Data is used and logistic, rare event logistic and machine learning methods are used to train models and to predict risk of cancers. Cross validation will be applied to evaluate the power of prediction of estimated models.

## 1.2 Aims of the project
Chronic diseases such as cancer, lung disease, and stroke have been major health concers and reasons of death in the U.S. Risk evaluation of these can lead to prventional treatments or life style changes that help patients survive and improve quality of life. The National Health Interview database enables us to use data analytical tools to predict the risk of many types of diseases with only simple interview answers to social, economic, and life style questions. The risk diseases could be estimated before any labs and clinical visits. When integrated into web tools, anybody can do the prediction by answering simple questions about themselves. This model can also help doctors to focus on high risk patient groups and help people to change their diet, and life style to reduce their risk of specific health problems.

This project is interdisciplinary because it needs not only the knowledge and skill of data analysis, it also needs physicians to reason and explain the predictions and causality relationships between risk and prediction variables, the predictors of the models. A predictor in a model could be a result or the cause of individual’s awareness of risk. For example, a breast cancer women might follow suggestions of her friends to become a vegetarian, if there are many such cases, vegetarians may be a good predictor for risk of this group of women. However, this predictor may not be the cause of the disease. It may be statistically correlated to existing cancer cases, but may not be a good predictor for the risk of potentially unknown breast cancers. In such a case, if physicians could help to confirm the irrelevance of vegetable lifestyle, the model can predict the risk of diseases before the first attack of these diseases.

# 2. Method and data
The data from National Health Interview Survey [NHIS 2018](https://www.cdc.gov/nchs/nhis/data-questionnaires-documentation.htm) are analyzed first by plots, tables, and maps, then by logistic and random forest models before rare event logistic models are estimated. The estimated results are represented in ROCs and heatmaps specifically designed for this project using truncated and signed log-transformed p-values that are less than 0.05. 

## 2.1. Data Downloading, Dimension, IDs, and Relations between Data Tables 
Please read comments to help understand the code. Some comments, however, are inactivated code lines for program testing and data examination.

```{r, ch2.1 }

hshld <- read.csv("/home/Nianfu/BMIN503/data/househldcsv/househld.csv")
#names(hshld)
#dim(hshld) ## 46500x17
fmly <- read.csv("/home/Nianfu/BMIN503/data/familyxxcsv/familyxx.csv")
#names(fmly) 
#dim(fmly) ## 30307x127 
prsn <- read.csv("/home/Nianfu/BMIN503/data/personsxcsv/personsx.csv")
#names(prsn)  
#dim(prsn) ## 72831x602 
adlt <- read.csv("/home/Nianfu/BMIN503/data/samadultcsv/samadult.csv")
#names(adlt)
#dim(adlt) ## 25417x742
chldrn <- read.csv("/home/Nianfu/BMIN503/data/samchildcsv/samchild.csv")
#names(chldrn)
#dim(chldrn) ## 8269x158

# data table number of rows and columns

NHIStables <- data.frame(c("Household", "Family", "Persons","Adults","Children"),
                        rbind(dim(hshld),dim(fmly),dim(prsn),dim(adlt),dim(chldrn))) 
names(NHIStables) <- c("dataName","rows", "columns")
NHIStables

#common columns compared with family table
com.hs_fm <- na.omit(match(names(hshld), names(fmly))) #4:6
names(fmly[,com.hs_fm])
com.ps_fm <- na.omit(match(names(prsn), names(fmly)))  #3:6
names(fmly[,com.ps_fm])
com.ad_fm <- na.omit(match(names(adlt), names(fmly)) ) #3:6
names(fmly[,com.ad_fm])
com.ch_fm <- na.omit(match(names(chldrn), names(fmly)))  #3:6
names(chldrn[,com.ch_fm])

#children and adults
com.ch_ad <- na.omit( match(names(chldrn), names(adlt)))
length(com.ch_ad) # 158 columns are all in adult
com.ad_ch <- na.omit( match(names(adlt), names(chldrn)))
length(com.ad_ch) # 158 columns are all in adult

#person and adults
com.pr_ad <- na.omit( match(names(prsn), names(adlt)))
length(com.pr_ad) # 158 columns are all in adult
sort(names(adlt[,com.pr_ad]))
com.ad_pr <- na.omit( match(names(adlt), names(prsn)))
length(com.ad_pr) # 158 columns are all in adult
sort(names(prsn[,com.ad_pr]))

## common names
head(hshld[, names(fmly[,c(4:6)])])
head(prsn[, names(fmly[,c(3:6)])])
head(adlt[, names(fmly[,c(3:6)])])
#head(chldrn[, names(fmly[,c(3:6)])])


##uniqueness of ids

dim(distinct(adlt[,c( "FMX", "HHX" )])) # unique ID, FPX is a number of head of Household
dim(adlt)

dim(distinct(prsn[,c("FMX","HHX", "FPX")])) # unique ID
dim(prsn)

dim(distinct(fmly[,c("FMX","HHX")])) # unique ID
dim(fmly)

##join to construction data of unique variables. persons<adults<household<family
Allxx <- left_join(prsn, adlt, by=names(adlt[,com.pr_ad])[-11]) %>% 
  left_join(hshld, by ="HHX") %>% 
   left_join(fmly, by =c("HHX","FMX") ) 
names(Allxx)
#View(xx)

```

* Four of the 5 data tables are potentially the data sources for the study project. The child table for children information is excluded from this project because their information is unlikely to be predictors for adults' health. These 4 tables have a total 72831 rows and 1466 columns
* After reading all the several hundreds of pages of documents, only 1 column of data for family size in the family data table will useful for this study. None are from household table.
* Personsx table includes data of all interviewed persons of families. Diseases are included for relatives information of adults in the adults table. This columns diseases in this personsx table are systematically named differently from the simillar information in the adults table.

> The 4 data tables have 3 shared ids\ 
--FMX:   Family number\
--HHX:   Household number\
--FPX:   Person number\

# 2.2. Variable Selection and Transformation
Potential response variables are selected diseases such as cancers and other 15 types of chronic diseases. All democgraphic, activities, smoking, drinking, and other life style related factors are included as potential predictors. The unknown, not answered, not applicable are change into NA values. All categorical and some continuous variables such as first smoking age, hours of sleep are transformed into categorical variables to simplify estimation for logistic model by constructing only linear links function in the logistic model. Such a categoricalization of continuous variables is an approximation to nonlinear link function. All data categoies are labeled to reduce errors in coding. Consistency of transformation is checked with their corresponding original data. thus you may expect to see some examination code lines for consistency. The three columns of data for vigorous, strength, and moderate physical activities are combined into a categorical variable to measure different types of activity. Age and BMI are the only continuous variables included in the estimation. They are kept to be continuous variables just to keep the diversity of the model and see if the code may work better or worse with them. For details of the data transformation please read the comments in the code chunks. 

```{r, ch2.2,warning=FALSE, message=FALSE, error=FALSE}
#family table
fm <- fmly %>% 
  select(HHX, FMX, FM_SIZE)

#person table
ps <- ps0 <- prsn %>% 
  select(HHX, FMX, FPX, REGION, REGIONBR,
         HISCODI3, FSPOUS2,FMOTHER1,FFATHER1,MOM_ED,DAD_ED, 
         
         LADURA1,  LADURA2,  LADURA3, LADURA4, LADURA7,LADURA8, LADURA9,LADURA10,
         LADURA11,  LADURA12,  LADURA18, LADURA21, LADURA28,  LADURA29,LDURA14A, 
         
         LADURB1,LADURB2,LADURB3, LADURB4, LADURB7, LADURB8,  LADURB9,LADURB10,
         LADURB11, LADURB12, LADURB18, LADURB21, LADURB28, LADURB29, LDURB14A,

         PHSTAT, WRKLYR1, WRKFTALL)
# Check names
# names(ps0)[12:26]
# names(ps0)[27:41]

#replace missings in with 0 where no diseases are blank
ps[,c(12:41)] <- replace(ps0[,c(12:41)],is.na(ps0[,c(12:41)]),-1)   ##missing cells are 0

# 97,98,99, 7,8,9 cases for some variables
ps[,c(12:26)] <- replace(ps[,c(12:26)],(ps[,c(12:26)] %in% c(96,97,98,99)),NA)   ##group into 99
ps[,c(27:41)] <- replace(ps[,c(27:41)],(ps[,c(27:41)] %in% c(7,8,9)),NA)   ##group into 99
#table(ps[,27], exclude = NULL)

#adult table
ad <- ad0 <- adlt %>% 
  select(HHX, FMX, FPX, SEX, AGE_P, MRACRPI2, R_MARITL,PAR_STAT, INDSTRN1, WRKCATA,
         
         ALDURA1,  ALDURA2,  ALDURA3, ALDURA4, ALDURA7,ALDURA8, ALDURA9,ALDURA10,
         ALDURA11,  ALDURA12,  ALDURA18,  ALDURA21, ALDURA28,  ALDURA29,ADURA14A, 
         
         ALDURB1,ALDURB2,ALDURB3, ALDURB4, ALDURB7,   ALDURB8,  ALDURB9,  ALDURB10,
         ALDURB11,ALDURB12,ALDURB18,ALDURB21, ALDURB28, ALDURB29, ADURB14A, LIVEV,
         
         SMKEV, SMKREG, SMKNOW, SMKSTAT2, VIGFREQW,MODFREQW, STRFREQW, ALCSTAT,AHEIGHT, AWEIGHTP, BMI,
         ASISLEEP,AASSTILL )

# Check names
#names(ad0)[11:25]
#names(ad0)[26:41]

#replace 7,8,9,97, 98, 99 for refuse, uncertain, and unknown, 
ad[,c(11:41)] <- replace(ad0[,c(11:41)],is.na(ad0[,c(11:41)]),-1)   ##missing cells are 0
ad[,c("SMKEV")] <- replace(ad[,c("SMKEV")],(ad[,c("SMKEV")] %in% c(7,8,9)), NA)   #### above 99.5
ad[,c("SMKREG")] <- replace(ad[,c("SMKREG")],(ad[,c("SMKREG")] %in% c(96,97,98,99)), NA)   #### above 99.5
ad[,c("SMKNOW")] <- replace(ad[,c("SMKSTAT2")],(ad[,c("SMKSTAT2")] %in% c(7,8,9)), NA)   #### above 99.5
ad[,c("SMKSTAT2")] <- replace(ad[,c("SMKSTAT2")],(ad[,c("SMKSTAT2")] %in% c(9)), NA)   #### above 99.5
ad[,c("ALCSTAT")] <- replace(ad[,c("ALCSTAT")],(ad[,c("ALCSTAT")] %in% c(9, 10)),NA)   ####Alcohol remove unknown frequency (very few)
ad[,c("ASISLEEP")] <- replace(ad[,c("ASISLEEP")],(ad[,c("ASISLEEP")] %in% c(97,98,99)),NA)   ####sleep hours

ad[,c("VIGFREQW")] <- replace(ad[,c("VIGFREQW")],(ad[,c("VIGFREQW")] %in% c(0)),1)   #### vigorous excercise times 0=<1, 
ad[,c("VIGFREQW")] <- replace(ad[,c("VIGFREQW")],(ad[,c("VIGFREQW")] %in% c(95,96)),0)   #### vigorous excercise 0 times
ad[,c("VIGFREQW")] <- replace(ad[,c("VIGFREQW")],(ad[,c("VIGFREQW")] %in% c(97,98,99)),NA)   #### 

ad[,c("MODFREQW")] <- replace(ad[,c("MODFREQW")],(ad[,c("MODFREQW")] %in% c(0)),1)   #### moderate excercise times 0=<1, 
ad[,c("MODFREQW")] <- replace(ad[,c("MODFREQW")],(ad[,c("MODFREQW")] %in% c(95,96)),0)   #### moderate excercise 0 times
ad[,c("MODFREQW")] <- replace(ad[,c("MODFREQW")],(ad[,c("MODFREQW")] %in% c(97,98,99)),NA)   #### no moderate exc

ad[,c("STRFREQW")] <- replace(ad[,c("STRFREQW")],(ad[,c("STRFREQW")] %in% c(0)),1)   #### strength excercise times 0=<1, 
ad[,c("STRFREQW")] <- replace(ad[,c("STRFREQW")],(ad[,c("STRFREQW")] %in% c(95,96)),0)   #### strength excercise 0 times
ad[,c("STRFREQW")] <- replace(ad[,c("STRFREQW")],(ad[,c("STRFREQW")] %in% c(97,98,99)),NA)   #### no strength exc

ad[,c("AWEIGHTP")] <- replace(ad[,c("AWEIGHTP")],(ad[,c("AWEIGHTP")] %in% c(996,997,998,999)),NA)   ####weight lbs.
ad[,c("AHEIGHT")] <- replace(ad[,c("AHEIGHT")],(ad[,c("AHEIGHT")] %in% c(96,97,98,99)),NA)   ####height inch

ad[,c("BMI")] <- replace(ad[,c("BMI")],(ad[,c("BMI")] %in% c(9999)),NA)   ####sleep hours
ad[,c("BMI")] <- replace(ad[,c("BMI")],(ad[,c("BMI")] %in% c(9995)),99.5)   #### above 99.5

ad[,c("PAR_STAT")] <- replace(ad[,c("PAR_STAT")],(ad[,c("PAR_STAT")] %in% c(9)),NA)   ####parents status
ad[,c("R_MARITL")] <- replace(ad[,c("R_MARITL")],(ad[,c("R_MARITL")] %in% c(9)),NA)   ####Marriage status
ad[,c("LIVEV")] <- replace(ad[,c("LIVEV")],(ad[,c("LIVEV")] %in% c(7,8,9)),NA)   ####liver condition

#add a flag for adult data rows to make it different from other data sources
ad$adult <- 1

# this columns is not included in the models because of BMI 
ps[,c("FSPOUS2")] <- replace(ps[,c("FSPOUS2")],(ps[,c("FSPOUS2")] %in% c(98)),NA)   ####weight lbs.

#head(ad$adult)
#table(ad$R_MARITL)

# data are mostly integers now

# list of variables and number of na
mis.fm <- sapply(fm, function(x) sum(is.na(x)))
mis.ps <-sapply(ps, function(x) sum(is.na(x)))
mis.ad <- sapply(ad, function(x) sum(is.na(x)))

# MERGE and convert data into analyzable data types; Disease (nmaed dis?) are response variables for health status .
# 4 variables (smk, activ, sleep, and dis99) are tranformated with fct_rev() function to facilitate the default baseline as wanted. 
hlth <- left_join(ps, ad, by=c('HHX', 'FMX', 'FPX')) %>% 
  left_join(fm, by =c("HHX","FMX") ) %>% 
  mutate(sex=factor(SEX, levels=c(1,2), labels=c("male", "female")), 
         alc=factor(cut(ALCSTAT,breaks = c(0,1,5, 7,8), labels=c("Abstainer", "Former", "Light", "Heavier"))),
         smk=fct_rev(factor(SMKSTAT2, levels=c(1,2,3,4), labels=c("Daily", "Some", "Former", "Never") )), 
         age_smk=ifelse(smk=="Never",-5, SMKREG),
         age_smk=factor(cut(age_smk, breaks = c(-6,0,18,100), labels=c("Never","6-18", "above19"), useNA='ifany' )),

         # three columns are merged into 1. 
         activ=fct_rev(factor(ifelse(VIGFREQW >2, 1, 
                              ifelse(STRFREQW>2, 2, 
                              ifelse(MODFREQW>2, 3, 
                              ifelse(VIGFREQW==0 & STRFREQW==0 & MODFREQW==0, 5, 4)))), 
                              levels=c(1,2,3,4,5), labels = c("Vigorous","Strength","Moderate", "Some", "None"))),
         sleep=fct_rev(factor(cut(ASISLEEP, breaks = c(-1,0,4,5,6,7,24), useNA='ifany' ), 
                              labels = c("<=4 hrs","5 hrs","6 hrs","7 hrs", ">7 hrs"))),
         fm_size=factor(cut(FM_SIZE, breaks = c(0,1,2,4, 20), useNA='ifany' ), 
                              labels = c("1 person","2 persons","3-4 persons", ">5 persons")),
         
         age=as.numeric(AGE_P),
         BMI=as.numeric(BMI)/100,
         
         region=factor(REGION, levels=c(1,2,3,4), labels = c("Northeast", "Midwest", "South", "West")),
         race0= factor(HISCODI3,levels=c(1:5), labels = c("Hispanic", "White","Black", "Asian","Other")),
         race= factor(race0, levels = c("White", "Hispanic", "Black", "Asian","Other")),
         marriage = factor(cut(R_MARITL, breaks = c(0,2,6, 7,8), useNA='ifany' ), labels = c("married","D/S/W", "Never", "Partner")),
         par_stat = factor(cut(PAR_STAT, breaks = c(0,2,3), useNA='ifany'), labels= c("Children","No children")),
         
         dis1=factor(cut(ALDURB1, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No vision disease", "Vision disease")),
         dis2=factor(cut(ALDURB2, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No hearing disease", "Hearing disease")),
         dis3=factor(cut(ALDURB3, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No arthritis disease", "Arthritis disease")),
         dis4=factor(cut(ALDURB4, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No back or neck disease", "Back or neck ")),
         dis7=factor(cut(ALDURB7, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No heart disease", "Heart disease")),
         
         dis8=factor(cut(ALDURB8, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No stroke disease", "Stroke")),
         dis9=factor(cut(ALDURB9, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No hypertension disease", "Hypertension disease")),
         dis10=factor(cut(ALDURB10, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No diabetes disease", "Diabetes disease")),
         dis11=factor(cut(ALDURB11, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No lung or breathing disease", "Lung or breathing disease")),
         dis12=factor(cut(ALDURB12, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No cancer disease", "Cancer disease")),
         
         dis14=factor(cut(ADURB14A, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No mental retardation", "Mental retardation")),
         dis18=factor(cut(ALDURB18, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No weight problem", "Weight problem")),
         dis21=factor(cut(ALDURB21, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No circulation disease", "Circulation disease")),
         dis28=factor(cut(ALDURB28, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No benign tumor disease", "Benign tumor disease")),
         dis29=factor(cut(ALDURB29, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No alcohol/drug problem", "Alcohol/drug problem")),
         dis99=fct_rev(factor(cut(LIVEV, breaks =c(0,1,2), useNA='ifany' ), labels=c("Chronic liver disease", "No chronic liver disease")))
              # Chronic liver, original 1=yes 2=no, reversed 1=no, 2=yes to be consistent with others
          ) %>% 
  filter(adult==1)

# Combine/NA some confusing categories, several others were correct in the above code.
hlth$smk[hlth$smk=="smk/unknown"]= NA

## check data
names(hlth)
nrow(hlth)
nrow(ad)

# liver, smoking age special check
#table(hlth$dis99, exclude = NULL)
#table(ad$LIVEV, exclude = NULL)
#table(hlth$LIVEV, exclude = NULL)
#table(hlth$age_smk, exclude = NULL)

#check number of no activity
#table(hlth$activ, exclude = NULL)
#hlth %>% filter(VIGFREQW==0 & STRFREQW==0 & MODFREQW==0) %>% summarise(n=n())

#check transormed data with original data
#sum(ad0$ALDURB1 %in% c(6,7,8,9))
#table(hlth$dis1, exclude = NULL)

```

# 3. DATA Descriptions and Exploration of Potential Correlations. 
This section uses mainly plots ot represents distribution of variables.

## 3.1. Demographics and Activities

```{r, ch3.1, warning=FALSE, message=FALSE, error=FALSE, fig.width=12}
#personal demographics and behaviors

XX0 <- hlth %>% 
  select(HHX, FMX, FPX, age, sex, race,  smk, age_smk, alc, sleep, activ,
          fm_size, marriage, par_stat, BMI)
head(XX0)
#summary(XX0)
#ggpairs(XX0)

##plots
p1 <- ggplot(data = XX0, aes(x=age )) + 
  geom_histogram(bins = 8, color="black", fill = "purple") +
  labs(y="People", x="Age")  +
  theme_bw()
p2 <- ggplot(data = XX0, aes(x=sex )) + 
  geom_bar(color="black", fill = "pink", width=0.7) +
  labs(y="People", x="Sex")  +
  theme_bw()
p3 <- ggplot(data = XX0, aes(x=smk )) + 
  geom_bar(color="black", fill = "green", width=0.7) +
  labs(y="People", x="Smoking")  +
  theme_bw()
p4 <- ggplot(data = XX0, aes(x=age_smk )) + 
  geom_bar(color="black", fill = "grey", width=0.7) +
  labs(y="People", x="Age first smoked regularly") + 
  theme_bw()
p5 <- ggplot(data = XX0, aes(x=alc )) + 
  geom_bar(color="black", fill = "Red", width=0.7) +
  labs(y="People", x="Alcohol drinking")  +
  theme_bw()
p6 <- ggplot(data = XX0, aes(x=sleep )) + 
  geom_bar(color="black", fill = "brown", width=0.7) +
  labs(y="People", x="Hours of spleep")  +
  theme_bw()
p7 <- ggplot(data = XX0, aes(x=activ)) + 
  geom_bar(color="black", fill = "purple", width=0.7) +
  labs(y="People", x="Physical activity")  +
  theme_bw()
p8 <- ggplot(data = XX0, aes(x=BMI )) + 
  geom_histogram(bins = 8, color="blue", fill = "red") +
  labs(y="People", x="BMI")  +
  theme_bw()
p9 <- ggplot(data = XX0, aes(x=fm_size )) + 
  geom_bar(color="black", fill = "Brown", width=0.7) +
  labs(y="People", x="Family size")  +
  theme_bw()
p10 <- ggplot(data = XX0, aes(x=race)) + 
  geom_bar(color="black", fill = "lightblue", width=0.7) +
  labs(y="People", x="Race") + 
  theme_bw()
p11 <- ggplot(data = XX0, aes(x=marriage)) + 
  geom_bar(color="black", fill = "pink", width=0.7) +
  labs(y="People", x="Marriage")  +
  theme_bw()
p12 <- ggplot(data = XX0, aes(x=par_stat)) + 
  geom_bar(color="black", fill = "blue", width=0.7) +
  labs(y="People", x="Children")  +
  theme_bw()
  
plot_grid(p1,p2,p3,p4,p5,p6,labels = "AUTO")
plot_grid(p7,p8,p9,p10,p11,p12,labels = c("G","H","I","J","K","L","M"))

# clearn memory
rm("Allxx", "chldrn", "XX0", "fm","ps0","ad","ad0", "p1","p2","p3","p4","p5","p6","p7","p8","p9","p10","p11","p12",
   "mis.ad","mis.fm","mis.ps",'com.ad_ch', 'com.ad_fm', 'com.ad_pr', 'com.ch_ad', 'com.ch_fm', 'com.hs_fm', 'com.pr_ad', 'com.ps_fm', "prsn","adlt","fmly", "hshld", "NHIStables")

```

## 3.2. Exploration of Chronic Diseases vs. Alcohol and Smoking
Ratios of cases of selected disease are ploted against drinking and smoking, which are the 2 most intereting factor I would like to analyze.

```{r, ch3.2 first, fig.width=10,warning=FALSE, message=FALSE, error=FALSE}

YY <- hlth %>% 
  select(dis1, dis2, dis3, dis4, dis7, 
         dis8,dis9, dis10, dis11, dis12,
         dis18, dis28,dis21, dis29, dis14, dis99)
#total number diseases by type
dis <- data.frame(sapply(YY, function(x) sum(as.numeric(x)==2, na.rm = TRUE)))
table(YY$dis99)
dis$name <-  c("vision", "hearing", "arthritis","back or neck", "heart",
                "strock","hypertension","diabetes", "lung", "cancer",
                "weight", "circulation", "benign tumor", "alcohol/drug", "mental", "liver")
names(dis)[1] <- c("number.disease")

## try some contract
ggplot(data = hlth[!is.na(hlth$smk), ], aes(x = smk, fill = factor(dis11))) + 
  geom_bar(position = "dodge", color="blue") +
  theme_bw()+
  labs(x="Smoking", y="People")+
  theme(legend.title=element_blank())
```

>The above plot dispays that bar plots of absolute numbers of each disease categories are not suitable for comparison against the categories of other variables, thus bar plots of ratios will be applied. 

```{r, ch3.2_second, fig.width=10, warning=FALSE, message=FALSE, error=FALSE}
# check ratio: smoke
par(mfrow=c(2,3))
tmt <- with(hlth[!is.na(hlth$smk), ], table(smk,dis11)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])
barplot(r, xlab = "Smoke", ylab="lung or breathing", col=c("green", "red","purple","brown"))

tmt <- with(hlth[!is.na(hlth$smk), ], table(smk,dis12)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Smoke", ylab="Cancer %", col=c("green", "red","purple","brown"))

tmt <- with(hlth[!is.na(hlth$smk), ], table(smk,dis99)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Smoke", ylab="Liver %", col=c("green", "red","purple","brown"))

# alcohol
tmt <- with(hlth[!is.na(hlth$alc), ], table(alc,dis11)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])
barplot(r, xlab = "Alcohol", ylab="lung or breathing", col=c("green", "red","purple","brown"))

tmt <- with(hlth[!is.na(hlth$alc), ], table(alc,dis12)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Alcohol", ylab="Cancer %", col=c("green", "red","purple","brown"))

tmt <- with(hlth[!is.na(hlth$alc), ], table(alc,dis99)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Alcohol", ylab="Liver %", col=c("green", "red","purple","brown"))

```

>Smoking is bad to health, but drinking alcohol has mixed relationships with health.

## 3.3. Diseases and Parents
This section checked correlation between parent diseases and adult diseases.  The diseases are linked using a parents flag in the adult data.
```{r, ch3.3, warning=FALSE, message=FALSE, error=FALSE}
## liver condition is not in the persons data
names(ps)
mom <- ps %>% 
  select(HHX, FMX, FPX,
         LADURB1,LADURB2,LADURB3, LADURB4, LADURB7, LADURB8,  LADURB9,LADURB10,
         LADURB11, LADURB12, LDURB14A, LADURB18, LADURB21, LADURB28, LADURB29) %>% 
  mutate(FMOTHER1c=FPX,
         mdis1=factor(cut(LADURB1, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No vision disease", "Vision disease")),
         mdis2=factor(cut(LADURB2, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No hearing disease", "Hearing disease")),
         mdis3=factor(cut(LADURB3, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No arthritis disease", "Arthritis disease")),
         mdis4=factor(cut(LADURB4, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No back or neck disease", "Back or neck ")),
         mdis7=factor(cut(LADURB7, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No heart disease", "Heart disease")),
         
         mdis8=factor(cut(LADURB8, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No stroke disease", "Stroke")),
         mdis9=factor(cut(LADURB9, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No hypertension disease", "Hypertension disease")),
         mdis10=factor(cut(LADURB10, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No diabetes disease", "Diabetes disease")),
         mdis11=factor(cut(LADURB11, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No lung or breathing disease", "Lung or breathing disease")),
         mdis12=factor(cut(LADURB12, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No cancer disease", "Cancer disease")),
         
         mdis14=factor(cut(LDURB14A, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No mental retardation", "Mental retardation")),
         mdis18=factor(cut(LADURB18, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No weight problem", "Weight problem")),
         mdis21=factor(cut(LADURB21, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No circulation disease", "Circulation disease")),
         mdis28=factor(cut(LADURB28, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No benign tumor disease", "Benign tumor disease")),
         mdis29=factor(cut(LADURB29, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No alcohol/drug problem", "Alcohol/drug problem")),
         
   #      mdis99=fct_rev(factor(cut(LIVEV, breaks =c(0,1,2), useNA='ifany' ), labels=c("Chronic liver disease", "No chronic liver disease")))
            ) 

# Dad data, the same variable name pattern as mom's
dad <- ps %>% 
  select(HHX, FMX, FPX,
         LADURB1,LADURB2,LADURB3, LADURB4, LADURB7, LADURB8,  LADURB9,LADURB10,
         LADURB11, LADURB12, LDURB14A, LADURB18, LADURB21, LADURB28, LADURB29) %>% 
  mutate(FFATHER1c=FPX,
         ddis1=factor(cut(LADURB1, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No vision disease", "Vision disease")),
         ddis2=factor(cut(LADURB2, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No hearing disease", "Hearing disease")),
         ddis3=factor(cut(LADURB3, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No arthritis disease", "Arthritis disease")),
         ddis4=factor(cut(LADURB4, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No back or neck disease", "Back or neck ")),
         ddis7=factor(cut(LADURB7, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No heart disease", "Heart disease")),
         
         ddis8=factor(cut(LADURB8, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No stroke disease", "Stroke")),
         ddis9=factor(cut(LADURB9, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No hypertension disease", "Hypertension disease")),
         ddis10=factor(cut(LADURB10, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No diabetes disease", "Diabetes disease")),
         ddis11=factor(cut(LADURB11, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No lung or breathing disease", "Lung or breathing disease")),
         ddis12=factor(cut(LADURB12, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No cancer disease", "Cancer disease")),
         
         ddis14=factor(cut(LDURB14A, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No mental retardation", "Mental retardation")),
         ddis18=factor(cut(LADURB18, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No weight problem", "Weight problem")),
         ddis21=factor(cut(LADURB21, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No circulation disease", "Circulation disease")),
         ddis28=factor(cut(LADURB28, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No benign tumor disease", "Benign tumor disease")),
         ddis29=factor(cut(LADURB29, breaks =c(-1.1,-0.1,4), useNA='ifany' ), labels = c("No alcohol/drug problem", "Alcohol/drug problem")),
        
   #      mdis99=fct_rev(factor(cut(LIVEV, breaks =c(0,1,2), useNA='ifany' ), labels=c("Chronic liver disease", "No chronic liver disease")))
          )
# merge to hlth
hlth_par <- hlth %>% 
  mutate(FFATHER1c=FFATHER1, FMOTHER1c=FMOTHER1) %>% 
  left_join(mom, by = c("HHX","FMX", "FMOTHER1c")) %>% 
  left_join(dad, by = c("HHX","FMX", "FFATHER1c"))
  
#table(hlth_par$dis11)
#table(hlth_par$dis11, hlth_par$mdis11)
#table(hlth_par$dis1, hlth_par$mdis1)
## try some relation between adult and mom
ggplot(data = hlth_par[!is.na(hlth_par$mdis11),], aes(x = dis11, fill = mdis11)) + 
  geom_bar(position = "dodge", color="blue", na.rm = TRUE) +
  theme_bw()+
  labs(x="Mom lung problem", y="People")+
  theme(legend.title=element_blank())
```

>The above plot suggests again the difficulty in displaying the diseases with absolution number count. So in this project most of the result will be shown in relative values or percents.

```{r, ch4.2_second, warning=FALSE, message=FALSE, error=FALSE}
# rate aldults and their parents
fig.dim=c(10,7)
fig.width=9
par(mfrow=c(3,2))
tmt <- with(hlth_par[!is.na(hlth_par$mdis11), ], table(mdis11,dis11)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Mom lung or breath problem", ylab="lung or breath problem %",  col=c("green","red"))
#dady
tmt <- with(hlth_par[!is.na(hlth_par$ddis11), ], table(ddis11,dis11)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Dad lung or breath problem", ylab="lung or breath problem %",  col=c("green","blue"))

# in this case the sample size is too small, 1 case would be too big
tmt <- with(hlth_par[!is.na(hlth_par$mdis7), ], table(mdis7,dis7)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Mam heart disease", ylab="Heart disease %", col=c("green","red"))
# in this case the sample size is too small, 1 case would be too big
tmt <- with(hlth_par[!is.na(hlth_par$ddis7), ], table(ddis7,dis7)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Dad heart disease", ylab="Heart disease %", col=c("green","blue"))

# in this case the sample size is too small, 1 case would be too big
tmt <- with(hlth_par[!is.na(hlth_par$mdis3), ], table(mdis3,dis3)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Mom arthritis", ylab="Arthritis %",  col=c("green","red"))
# in this case the sample size is too small, 1 case would be too big
tmt <- with(hlth_par[!is.na(hlth_par$ddis3), ], table(ddis3,dis3)) 
r=tmt[,2]/(tmt[,1]+tmt[,2])*100
barplot(r, space =0.3,xlab = "Dad arthritis", ylab="Arthritis %",  col=c("green","blue"))

#names(hlth_par)
```

>Parents' health are relavent to adults' health.

# 4. Regional Variation
Maps of selected predictors and diseases are produced to represent regional variations

## 4.1. Regional Variation of Drinking and Smoking
Drinking and smoking ratios are display on the US regional map. A map function is created to map variables.

```{r, ch4.1, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}

# US state level geometry data from ACS
state<- get_acs(geography = "state",  geometry=TRUE,       # query data at the state level 
                   year = 2016,                 # end year (these will give us ACS 5-year estimates for 2011-2016)
                    variables = c("B01003_001")) #population
```


```{r, ch4.1_Second, warning=FALSE, message=FALSE, error=FALSE}
#map
# assign a region for each state. Region definition is based on maps of the US Census Beaurau
state$STATE=as.integer(state$GEOID)
NorthEast <- c(9,10,11,23,24,25,33,34,36,42,44,50)
MidWest <- c(17,18,19,20,26,27,29,31,38,39,46,55)
South <- c(1,5,12,13,21,22,28,37,40,45,47,48, 51, 54)
West <-  c(4,6,8,16,30,32,35,41,49,53,56)

#geos is only a object name for the geometric data
geos <- state[!(state$STATE %in% c(2,15,72)),]
geos$REGION[geos$STATE %in% NorthEast] <- 1
geos$REGION[geos$STATE %in% MidWest] <- 2
geos$REGION[geos$STATE %in% South] <- 3
geos$REGION[geos$STATE %in% West] <- 4

#check data
#table(geos$REGION)
geos[geos$REGION==2,]

#Percent of Abstainers by region
regionx <- hlth %>% filter(!is.na(alc)) %>% group_by(REGION, alc) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))

# Abst for alcohol abstainer 
Abst <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(alc=="Abstainer") %>%  
  mutate( per=100-n_1/n_t*100) %>% 
  select(REGION, per) 

#theme
my_theme <- function() {
  theme_minimal() +                                  
  theme(axis.line = element_blank(),                 
        axis.text = element_blank(),                 
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),  
        legend.key.size = unit(0.8, "cm"),          
        legend.text = element_text(size = 16),       
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 22))      
}
#coordinates of labels for Regions
coordinates = st_coordinates(st_point_on_surface(geos))

##Map function that will be used for all region maps
Rmap <- function(vrb, title,colors){
   # merge data
  map1 <- left_join( geos, vrb,by = "REGION")
  prev_min <-min(as.numeric(map1$per), na.rm = TRUE)
  prev_max <- max(as.numeric(map1$per), na.rm = TRUE)
  
  myPalette <- colorRampPalette(brewer.pal(9, colors))
  
   # Center of geometry
  map2 <- data.frame(map1,coordinates)
  map2$perc=as.character(paste0(round(map2$per,1),"%"))
  map2$lab=ifelse(map2$STATE %in% c(42, 19, 28,32), map2$perc,'')
 
  ggplot() +
    geom_sf(data=map1, lwd = 0, aes(fill = map1$per)) +
    my_theme() +
    ggtitle(title)+
    scale_fill_gradientn(name = "Rate (%)", colours = myPalette(50),
                         limit = range(0, prev_max))  +
    geom_text(data = map2, aes(x=X, y=Y, label=lab), size = 5, fontface = "bold",color="white")
}

# map for Alcohol
Rmap(Abst, "US Map of Adults Ever been Alcohol Consumers", "YlGnBu")

#percent of Never smoke by region
regionx <- hlth %>% filter(!is.na(smk)) %>% group_by(REGION, smk) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))
# Abst is for "never smoking" here
nonsmk <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(smk=="Never") %>%  
  mutate( per=100-n_1/n_t*100) %>% 
  select(REGION, per) 
# map for Alcohol
Rmap(nonsmk, "US Map of Adults Ever been Smokers", "YlOrRd")

```

## 4.2. Regional Variabtion of Diseases
The database does not include state information and regions have to be used to display location variation. For those rarely occured diseases the regional level percentages of specific categorie are still quite meaningful. This section of analysis display some examples of veriation among the 4 regions.Maps are based on continental state multipolygon vector geometry data from US Census Beaureu website. Each state in the same region are shown in the same color that indicate the percentage values of the region. 
```{r, ch4.2, warning=FALSE, message=FALSE, error=FALSE}
par(mfrow=c(3,2))
#lung diseases %
regionx <- hlth %>% filter(!is.na(dis11)) %>% group_by(REGION, dis11) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))
Lung <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(dis11=="Lung or breathing disease") %>%  
  mutate( per=n_1/n_t*100) %>% 
  select(REGION, per) 
  class(Abst$REGION ) 
  class (geos$REGION)
  colors="Purples"
  title="Title"

Rmap(Lung,"US Map of Adult Lung Diseases", "YlGn")

#Heart diseases %
regionx <- hlth %>% filter(!is.na(dis7)) %>% group_by(REGION, dis7) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))
Heart <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(dis7=="Heart disease") %>%  
  mutate( per=n_1/n_t*100) %>% 
  select(REGION, per)
Rmap(Heart,"US Map of Heart Diseases", "Purples")

#Liver %
regionx <- hlth %>% filter(!is.na(dis99)) %>% group_by(REGION, dis99) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))
Liver <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(dis99=="Chronic liver disease") %>%  
  mutate( per=n_1/n_t*100) %>% 
  select(REGION, per)
Rmap(Liver, "US Map of Liver Diseases", "OrRd")

#Arthritis diseases %
regionx <- hlth %>% filter(!is.na(dis3)) %>% group_by(REGION, dis3) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))
Arthritis <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(dis3=="Arthritis disease") %>%  
  mutate( per=n_1/n_t*100) %>% 
  select(REGION, per)
Rmap(Arthritis, "US Adult Arthritis", "PuRd")

#stroke %
regionx <- hlth %>% filter(!is.na(dis8)) %>% group_by(REGION, dis8) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))
stroke <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(dis8=="Stroke") %>%  
  mutate( per=n_1/n_t*100) %>% 
  select(REGION, per)
Rmap(stroke,"US Adult Stoke", "GnBu")

#Mental retardation
regionx <- hlth %>% filter(!is.na(dis14)) %>% group_by(REGION, dis14) %>% summarise(n= n())
tot <- regionx %>% group_by(REGION) %>% summarise(n= sum(n))
Mental <- left_join(regionx, tot, by = "REGION", suffix = c("_1", "_t")) %>% 
  filter(dis14=="Mental retardation") %>%  
  mutate( per=n_1/n_t*100) %>% 
  select(REGION, per)
Rmap(Mental,"US Adult Mental Retardation", "RdPu")


```

# 5. Modeling Relationships between Diseases and Predictors. 
All the diseases have binary values and logistic model is an intuitive method. Only age and BMI of all the feathures of interviewed adults are kept to be numeric variables. Factors are first transformed into binary variables before estimation of models. Baseline categories are the ones that omitted in the estimation results. 

## 5.1. Variables Significantly Correlated to Health
Variables/categories of variables that are significantly related to diseases are identified using logistic model.

```{r, ch5.1, warning=FALSE, message=FALSE, error=FALSE}

dis_Names <- c("Vision","Hearing", "Arthritis","Back_neck", "Heart",
              "Stroke","Hypertension", "Diabetes", "Lung_breath", "Cancer",
              "Mental", "Weight", "Circulation", "Benign_tumor","Alcohol_drug", 
              "Chronic_liver")
# reset the XX incase order change
# age, BMI are numeric, others are factors 

XX <- hlth %>% 
  select(age, sex, race,  smk, alc, sleep, activ,  
          fm_size, marriage, par_stat, BMI, region) ##HHX, FMX, FPX,age_smk, 
# Reset the YY in case order change
YY <- hlth %>% 
  select(dis1, dis2, dis3, dis4, dis7, 
         dis8,dis9, dis10, dis11, dis12,
         dis14, dis18, dis21, dis28, dis29,  
         dis99)
i=1
cef <- NA
sigName =NA 
dstName=NA
sigcef <- NA
pval <- NA

for (i in 1:16) {

  Y=YY[,i]
  YXX <- data.frame(Y, XX)
  YXX <- YXX[complete.cases(YXX),]
  #dim(data.frame(Y, XX))
  dim(YXX)
  Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))
  dt <-  data.frame(y=YXX[,"Y"],Xmat)
  dt.glm <- glm( y ~ . , data =dt, family = binomial(logit))
  s <- summary(dt.glm)$coefficients[,4]
  c <- summary(dt.glm)$coefficients[s<0.05, "Estimate"]
  nam <- names(c)[2:length(c)]
  sigName[i] <- list(nam)   #significant var names
  sigcef[i] <- list(c)      #significant coefficients
  cef[i] <-  list(summary(dt.glm)$coefficients[, "Estimate"]) ## all estimate
  pval[i] <- list(s)
  }
#sigName
cefTab <- as.data.frame(cef)
pvalTab <- as.data.frame(pval)

sig <- as.data.frame(pval)<0.05
sig=ifelse(sig==0, NA, sig)
sigTab <- cefTab*sig
names(sigTab)=names(YY)
sigTabf <- round(sigTab,2)
print(sigTabf)

```

>The above table contends all the significant coefficients whose p-values are less than 0.05. Those "NA" values in the table represent insignificant coefficients. The next section will display them in a heat map.

## 5.2. Heatmap of P-values
```{r, ch5.2, fig.height=8, warning=FALSE, message=FALSE, error=FALSE}

# Obtain the signs of estimated coefficients, 1= positive, -1=negtative
HMsign <- cefTab
HMsign[cefTab>0]=1
HMsign[cefTab<0]=-1

#log transformation
lps <- log(pvalTab)

# significance ones <0.05 are kept
lps[pvalTab>=0.05]=NaN

#add sign
pmat <- lps*HMsign*(-1)
X <- as.matrix(pmat[-1,])

#Truncation for managing extremely small p-values for reasonable color scheme
Z <- X
Z[X>-log(0.0001)]=-log(0.0001)
Z[X<=log(0.0001)]=log(0.0001)

# add names of diseases for each model
colnames(Z) <- dis_Names

# mgp and mar are used to adjust position of key, 
# color panel is defined for transision of colors from green to yellow to red
# margin to make room for labels on the margin
heatmap.2(Z, col=colorpanel(10,"green","yellow", "red"),  Rowv = FALSE, Colv = FALSE,  na.color = "white",
          margin=c(8,8), trace="none", 
          main="                    Significance of Estimated Coefficients",
          dendrogram = "none", key.xlab="Truncated and signed log p-value", 
          key.par = list(mgp=c(1.2,0.3,0), mar=c(2.5,2.5,4,0)), key.title = "- coeff               + coeff")
```

>The heatmap was designed to display the significant correlations (by the standard of p-values <=0.05) between diseases in columns and predictors in rows. Green in the the above heatmap is used to indicate negatively correlated to diseases, and thus positively correlated to health. Red is for positively correlated to diseases and thus negatively related to health. The color scheme as shown in the color key is realized using a truncated log-transformed scores based on p-values and the sign of its corresponding coefficients. There are 4 steps to transform the p-values: (1). Log-transform the p-values from the estimation; (2). Replace all cells with NaN when corresponding p-values>=0.05; (3). Add the opposite sign of the corresponding estimated coefficients; (4). Truncate large values by + and - log(0.0001) so that colors display in a right range.

## 5.3. Implications of Correlations between Predictors and Health 
A continuous variable or a category of a factor is defined to be "much" healthier if the difference between the number of models in which it is negatively significant is 10 or above more than the number of models in which it is positively significant.  A continuous variable or a category of a factor is defined to be healthier if the numbers of models in which it is negatively significant is 5 or above more than the number of models in which it is positively significant.  A continuous variable or a category of a factor is defined to be moderately healthier if the number of models in which it is negatively significant is 4 or below 4 more than the number of models in which it is ositively significant. If opposite of the three situations are true, then they are defined as much worse, worse or moderately worse correspondingly. 

```{r, ch5.3, warning=FALSE, message=FALSE, error=FALSE}

nPsig <- apply(sigTabf, 1, function(x) sum(x>=0, na.rm = TRUE))
nNsig <- apply(sigTabf, 1, function(x) sum(x<0, na.rm = TRUE))
nsig <- as.data.frame(cbind(Var=names(nPsig), nPsig, nNsig))
nsig$Contrast <- nNsig-nPsig
nsig$Baseline <- c("","","male", 
                   "white","white","white","white",
                   "Never smoke","Never smoke","Never smoke",
                   "abstainer", "abstainer", "abstainer", 
                   ">8 hrs of sleep",">8 hrs of sleep",">8 hrs of sleep",">8 hrs of sleep",
                   "No activities", "No activities", "No activities", "No activities", 
                   "1-person family", "1-person family", "1-person family",
                   "married couple","married couple","married couple",
                   "parents with kids",
                   "",
                   "Northeast", "Northeast", "Northeast")
nsig$Implication <- c("", "Young people are much healthier",
          "Female are moderately healthier", 
          "Blacks are moderately worse","Hispanics are moderately worse", "Asians, moderately healthier", "Other races are worse",
          "Former smokers, much worse","Some smokers, worse","Daily smokers, much worse",
          "Former alcohol users, moderately worse", "Current light drinker, much healthier", "Current heavy drinker, healthier",
          "7 hrs sleep, moderately healthier","6 hrs sleep, healthier",
              "5 hrs sleep, worse", "<4 hrs sleep, much worse",
          "Some activity is healthier",  "Moderate activity is much healthier",  "Strength activity is moderately healthier",  
              "Vigorous activity is much healthier",  
          "2-or-more-member family, moderately worse", ">=2 member family, moderately worse", ">=2 member family, moderately worse",
          "Divorce, separate, widow, much worse", "Never married, moderately worse", "live with partner, moderately worse", 
              "parents, healthier", "",
          "Midwest people are moderately healthier", "Southern people are as health as Northeast", "west people are as moderately healthier")
names(nsig) <- c("Variables", "# +coeff","# -coeff", "difference", "Baseline", "Implication")
print(nsig[2:nrow(nsig),c(2:5)])
print(nsig[2:nrow(nsig),c(4,6)])
rm("Abst", "coordinates", "geos", "dis", "regionx", "state", "tot","dad","mom","dt", "lps")

```

# 6. Trainining Models for Prediction

## 6.1. Full Prediction Model Training
All predictors are included in the estimation of the prediction models for the 16 diseases. Records with missing values are excluded in the model training process. There are only a small number of incomplete records. 

The current directory is specified for my computer, please change the folder name you want to use, or delete the directory to use your default workspace for saving and reading the temporary results. 

Please be patient, the training may take several minutes or long depending on the environment it runs.

### 6.1.1. Estimation and Cross-Validation 
```{r, ch6.1.1, eval=FALSE, message=FALSE,warning=FALSE} 
j=1
pred.y.logit <- NA
obs.y.logit <- NA 
pred.y.rf <- NA
obs.y.rf <- NA 

for (j in 1:16) {
  # j is the jth diseases such as dis1, dis2,...dis29, dis99
  Y=YY[,j]
  YXX <- data.frame(Y, XX)
  YXX <- YXX[complete.cases(YXX),]
  #dim(data.frame(Y, XX))
  dim(YXX)
  Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))
  dt <-  data.frame(y=YXX[,"Y"],Xmat)

  ###logistic CV with all variables
  dtxx <- dt
  N = nrow(dtxx)
  K = 10
  set.seed(1234)
  # randomly select one of the K from N, the number of rows of data
  s = sample(1:K, size = N, replace = T) # used for label rows of data titanic for each group       
  pred.outputs.logit <- vector(mode = "numeric", length = N)
  obs.outputs.logit <- vector(mode = "numeric", length = N)
  offset <- 0 # offset is a pointer used for determine where to pending results for the next loop	
  
  i=1
  for(i in 1:K){
    	train <- filter(dtxx, s != i)
    	test <- filter(dtxx, s == i)
    	obs.outputs.logit[1:length(s[s == i]) + offset] <- test$y
    
      dt.glm <- glm(y ~ . , data =train, family = binomial(logit))
      logit.pred.curr <- predict(dt.glm, test, type = "response")
      pred.outputs.logit[1:length(s[s == i]) + offset] <- logit.pred.curr 
     	offset <- offset + length(s[s == i])
     # 	return(list(obs.outputs.logit, pred.outputs.logit))
    }
  pred.y.logit[j] <- list(pred.outputs.logit)
  obs.y.logit[j] <- list(obs.outputs.logit) 

###random forest with top variables
dtxx <- dt
  N = nrow(dtxx)
  K = 10
  set.seed(1234)
  # randomly select one of the K for N times, the number of rows of data
  s = sample(1:K, size = N, replace = T) # used for label rows of data titanic for each group       
  pred.outputs.rf <- vector(mode = "numeric", length = N) # to store prediction
  obs.outputs.rf <- vector(mode = "numeric", length = N) # to store response yi
  offset <- 0
  i=1

  for(i in 1:K){
  	train <- filter(dtxx, s != i)
  	test <- filter(dtxx, s == i)
  	obs.outputs.rf[1:length(s[s == i]) + offset] <- test$y 
    
  	#RF train/test
  	rf <- randomForest(y ~ ., data = train, ntree = 100)
  	
  	rf.pred.curr <- predict(rf, newdata = test, type = "prob") 
  	pred.outputs.rf[1:length(s[s == i]) + offset] <- rf.pred.curr[ , 2]
  
      # offset is used for determine where to pending results for the next loop	
  	offset <- offset + length(s[s == i])

#  	return (obs.outputs.rf)
  }
  pred.y.rf[j] <- list(pred.outputs.rf)
  obs.y.rf[j] <- list(obs.outputs.rf)
}
Xmat <- X <- NULL
Y <- YXX <-  NULL
saveRDS(pred.y.logit,file="pred_y_logit.rds") 
saveRDS(obs.y.logit, file="obs_y_logit.rds") 
saveRDS(pred.y.rf, file="pred_y_rf.rds") 
saveRDS(obs.y.rf, file="obs_y_rf.rds") 

```

### 6.1.2. Full Model AUC and ROC
```{r, ch6.1.2, warning=FALSE, message=FALSE, error=FALSE}
#getwd()

#please change the directory or delete the directory for reading the files from the above code
pred.y.logit <- readRDS("/home/Nianfu/pred_y_logit.rds")
obs.y.logit <- readRDS("/home/Nianfu/obs_y_logit.rds") 
pred.y.rf <- readRDS("/home/Nianfu/pred_y_rf.rds") 
obs.y.rf <- readRDS("/home/Nianfu/obs_y_rf.rds") 
# AUC in one table
auc <- matrix(data=NA, nrow = 16, ncol=2)
i=1
for (i in c(1:16)){
  r1 <- roc(obs.y.logit[[i]], pred.y.logit[[i]], ci = TRUE)
  auc[i,1]=r1$auc[1]
  r2 <- roc(obs.y.rf[[i]], pred.y.rf[[i]], ci = TRUE)
  auc[i,2]=r2$auc[1]
}
aucdt <- as.data.frame(auc)
names(aucdt)=c("Logistic", "Random.Forest")
aucdt
print(paste("logistic 10-fold CV model AUC= ", round(mean(auc[,1]), 3)))
print(paste("Random Forest 10-fold CV model AUC= ", round(mean(auc[,2]), 3)))


#Plot ROCs
par(mfrow=c(1,1))

i=1
plot.roc(obs.y.logit[[1]], pred.y.logit[[1]],  lwd = 0.7, lty=1, ci = TRUE, col = "blue") #CV of svm
plot.roc(obs.y.rf[[i]], pred.y.rf[[i]],  lwd = 0.7, lty=2, ci = TRUE, col = "brown",add = TRUE) #CV of rf
legend("bottomright", legend = c("Logistic Cross-Validation", "RF Cross-Validation"), col = c("blue", "brown"), lwd = 2, lty = c(1,1,2,1))
for (i in c(2:16)){
plot.roc(obs.y.logit[[i]], pred.y.logit[[i]], lwd = 0.7,ci = TRUE, lty=1, col = "blue", add=TRUE) #CV of svm
plot.roc(obs.y.rf[[i]], pred.y.rf[[i]], lwd = 0.7,ci = TRUE, lty=2, col = "brown", add = TRUE) #CV of rf
}

```

> The logit models predict better than random forest models

## 6.2. Reduced Models for Prediction
In this section, the top 10 variables/categories from the full models are used for prediction model training with 10-fold cross-validation. AUC and ROC are produced to compared the power of predictions.

Again, as for the previous full model training, please be change the directories and be patient to wait for the running of the code. 

### 6.2.1. Reduced Model Cross-Validation 
```{r,ch6.2.1,  eval=FALSE, message=FALSE,warning=FALSE}

j=1
pred.y.logit10 <- NA
obs.y.logit10 <- NA 
pred.y.rf10 <- NA
obs.y.rf10 <- NA 

for (j in 1:16) {
  # j is the jth diseases such as dis1, dis2,...dis29, dis99
  Y=YY[,j]
  YXX <- data.frame(Y, XX)
  YXX <- YXX[complete.cases(YXX),]
  Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))

  # reduced 10 variables
  z <-  rownames(cefTab)[-1] # logistic coefficient from above code
  zp <- pvalTab[-1,j] #pvalue matrix from above code
  tp10nam <- z[order(zp)][1:10] # top 10 x
  tp10X <- Xmat[, tp10nam]

  Y10X <- data.frame(YXX$Y, tp10X)
  Xmat10 <- data.frame(model.matrix(~., data = Y10X[, c(2:ncol(Y10X))] ))
  dt <-  data.frame(y=YXX[,"Y"],Xmat10)

  ###logistic CV with all variables
  dtxx <- dt
  N = nrow(dtxx)
  K = 10
  set.seed(1234)
  # randomly select one of the K from N, the number of rows of data
  s = sample(1:K, size = N, replace = T) # used for label rows of data titanic for each group       
  pred.outputs.logit <- vector(mode = "numeric", length = N)
  obs.outputs.logit <- vector(mode = "numeric", length = N)
  offset <- 0 # offset is a pointer used for determine where to pending results for the next loop	
  
  i=1
  for(i in 1:K){
    	train <- filter(dtxx, s != i)
    	test <- filter(dtxx, s == i)
    	obs.outputs.logit[1:length(s[s == i]) + offset] <- test$y
    
      dt.glm <- glm(y ~ . , data =train, family = binomial(logit))
      logit.pred.curr <- predict(dt.glm, test, type = "response")
      pred.outputs.logit[1:length(s[s == i]) + offset] <- logit.pred.curr 
     	offset <- offset + length(s[s == i])
     # 	return(list(obs.outputs.logit, pred.outputs.logit))
    }
  pred.y.logit10[j] <- list(pred.outputs.logit)
  obs.y.logit10[j] <- list(obs.outputs.logit) 

###random forest with top variables
dtxx <- dt
  N = nrow(dtxx)
  K = 10
  set.seed(1234)
  # randomly select one of the K for N times, the number of rows of data
  s = sample(1:K, size = N, replace = T) # used for label rows of data titanic for each group       
  pred.outputs.rf <- vector(mode = "numeric", length = N) # to store prediction
  obs.outputs.rf <- vector(mode = "numeric", length = N) # to store response yi
  offset <- 0
  i=1

  for(i in 1:K){
  	train <- filter(dtxx, s != i)
  	test <- filter(dtxx, s == i)
  	obs.outputs.rf[1:length(s[s == i]) + offset] <- test$y
  
  	#RF train/test
  	rf <- randomForest(y ~ ., data = train, ntree = 100)
  	
  	rf.pred.curr <- predict(rf, newdata = test, type = "prob") 
  	pred.outputs.rf[1:length(s[s == i]) + offset] <- rf.pred.curr[ , 2]
  
      # offset is used for determine where to pending results for the next loop	
  	offset <- offset + length(s[s == i])

#  	return (obs.outputs.rf)
  }
  pred.y.rf10[j] <- list(pred.outputs.rf)
  obs.y.rf10[j] <- list(obs.outputs.rf)
}
Xmat <- X <- NULL
Y <- YXX <-  NULL
saveRDS(pred.y.logit10,file="pred_y_logit10.rds") 
saveRDS(obs.y.logit10, file="obs_y_logit10.rds") 
saveRDS(pred.y.rf10, file="pred_y_rf10.rds") 
saveRDS(obs.y.rf10, file="obs_y_rf10.rds") 
```


### 6.2.2.Reduced Model AUC and ROC
```{r, chunk6.2.2, warning=FALSE, message=FALSE, error=FALSE}
pred.y.logit10 <- readRDS("/home/Nianfu/pred_y_logit10.rds")
obs.y.logit10 <- readRDS("/home/Nianfu/obs_y_logit10.rds") 
pred.y.rf10 <- readRDS("/home/Nianfu/pred_y_rf10.rds") 
obs.y.rf10 <- readRDS("/home/Nianfu/obs_y_rf10.rds")  
# AUC in one table
auc10 <- matrix(data=NA, nrow = 16, ncol=2)
i=1
for (i in c(1:16)){
  r1 <- roc(obs.y.logit10[[i]], pred.y.logit10[[i]], ci = TRUE)
  auc10[i,1]=r1$auc[1]
  r2 <- roc(obs.y.rf10[[i]], pred.y.rf10[[i]], ci = TRUE)
  auc10[i,2]=r2$auc[1]
}
aucdt10 <- as.data.frame(auc10)
names(aucdt10)=c("Logistic", "Random.Forest")
aucdt10
print(paste("Reduced logistic 10-fold CV model AUC= ", round(mean(auc10[,1]), 3)))
print(paste("Reduced random Forest 10-fold CV model AUC= ", round(mean(auc10[,2]), 3)))
   
  
#Plot ROCs
par(mfrow=c(1,1))

i=1
plot.roc(obs.y.logit10[[i]], pred.y.logit10[[i]],  lwd = 0.7, lty=1, ci = TRUE, col = "red") #CV of svm
plot.roc(obs.y.rf10[[i]], pred.y.rf10[[i]],  lwd = 0.7, lty=2, ci = TRUE, col = "green",add = TRUE) #CV of rf
legend("bottomright", legend = c("Logistic Cross-Validation", "RF Cross-Validation"), col = c("red", "green"), lwd = 2, lty = c(1,1,2,1))
for (i in c(2:16)){
plot.roc(obs.y.logit10[[i]], pred.y.logit10[[i]], lwd = 0.7,ci = TRUE, lty=1, col = "red", add=TRUE) #CV of svm
plot.roc(obs.y.rf10[[i]], pred.y.rf10[[i]], lwd = 0.7,ci = TRUE, lty=2, col = "green", add = TRUE) #CV of rf
}

```

>Area Under the Curve (AUC) of the logistic regression for all the 16 full models are larger than those of the Random Ferest models. 

## 6.3., Reduced vs. Full Model Prediction
If only top 10 variables/catigories of variables are used, the prediction model can be applied for cheaper surveys. This section compare the similarity of full and top 10 variable logistic models.
```{r, ch6.3, warning=FALSE, message=FALSE, error=FALSE}
i=1
plot.roc(obs.y.logit[[i]], pred.y.logit[[i]],  lwd = 0.7, lty=1, ci = TRUE, col = "blue") #CV of svm
plot.roc(obs.y.logit10[[i]], pred.y.logit10[[i]],  lwd = 0.7, lty=1, ci = TRUE, col = "red", add=TRUE) #CV of svm
legend("bottomright", legend = c("Full Logistic Model", "Reduced Logistic Model"), col = c("blue", "red"), lwd = 2, lty = c(1,1,2,1))
for (i in c(2:16)){
plot.roc(obs.y.logit[[i]], pred.y.logit[[i]], lwd = 0.7,ci = TRUE, lty=1, col = "blue", add=TRUE) #CV of svm
plot.roc(obs.y.logit10[[i]], pred.y.logit10[[i]], lwd = 0.7,ci = TRUE, lty=1, col = "red", add = TRUE) #CV of rf
}

allauc <- cbind(aucdt, aucdt10)
names(allauc)[] <- c("Logistic_AUC", "RF_AUC", "Top10_Logistic_AUC", "Top10_RF_AUC")
allauc[, c(1,3)]

```

>The prediction qualities are similar measured by AUC values of the 2 groups of models. Only for the NOT-converged estimtation of alcohol_drug abuse models there are obvious difference. But the difference does not matter much because both full and reduced top 10 variable models are not converged. Thus the reduced models are as good as full model in prediction.

# 7. Discussion about Adding Parent Health as a Predictor
In the previous section the parents diseases have shown strong correlation with adults diseases. But the estimation performed with the code below cannot find many converged coefficients. The reason is the numbers of adults who have parents information are too small. Thus this part of inconverged results was not reported. 

```{r, ch7, eval=FALSE, warning=FALSE, message=FALSE, error=FALSE}
  cef <- NA
  sigName =NA 
  dstName=NA
  sigcef <- NA
  pval <- NA
  ncomplete <- rep(NA , 16)
tryfunc <- function(){
  dim(hlth_par)
  dis_Names <- c("Vision","Hearing", "Arthritis","Back_neck", "Heart",
                "Stroke","Hypertension", "Diabetes", "Lung_breath", "Cancer",
                "Mental", "Weight", "Circulation", "Benign_tumor","Alcohol_drug", 
                "Chronic_liver")
  # reset the XX incase order change
  # age, BMI are numeric, others are factors 
  
  XX <- hlth_par %>% 
    select(age, sex, race,  smk, alc, sleep, activ,  
            fm_size, marriage, par_stat, BMI, REGION ) ##HHX, FMX, FPX,age_smk, 
  # reset the YY incase order change
  YY <- hlth_par %>% 
    select(dis1, dis2, dis3, dis4, dis7, 
           dis8,dis9, dis10, dis11, dis12,
           dis14, dis18, dis21, dis28, dis29,  
           dis99)
  mY <- hlth_par %>% 
    select(mdis1, mdis2, mdis3, mdis4, mdis7, 
           mdis8, mdis9, mdis10, mdis11, mdis12,
           mdis14, mdis18, mdis21, mdis28, mdis29)
  dY <- hlth_par %>% 
    select(ddis1, ddis2, ddis3, ddis4, ddis7, 
           ddis8, ddis9, ddis10, ddis11, ddis12,
           ddis14, ddis18, ddis21, ddis28, ddis29)
  i=2

  for (i in 1:15) {
  
    Y=YY[,i]
    mom=mY[,i]
    dad=dY[,i]
    YXX <- data.frame(Y, XX, mom, dad)
    YXX <- YXX[complete.cases(YXX),]
    #dim(data.frame(Y, XX))
    dim(YXX)
    Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))
    dt <-  data.frame(y=YXX[,"Y"],Xmat)
    dt.glm <- glm( y ~ . , data =dt, family = binomial(logit))
    s <- summary(dt.glm)$coefficients[,4]
    names(s)[substr(names(s),1,3)=="mom"] <- "mom"
    names(s)[substr(names(s),1,3)=="dad"] <- "dad"
    c <- summary(dt.glm)$coefficients[s<0.05, "Estimate"]
    names(c)[substr(names(c),1,3)=="mom"] <- "mom"
    names(c)[substr(names(c),1,3)=="dad"] <- "dad"
    nam <- names(c)[2:length(c)]
    sigName[i] <- list(nam)   #significant var names
    sigcef[i] <- list(c)      #significant coefficients
    cef[i] <-  list(summary(dt.glm)$coefficients[, "Estimate"]) ## all estimate
    pval[i] <- list(s)
    names(cef[[i]])[substr(names(cef[[i]]),1,3)=="mom"] <- "mom"
    names(cef[[i]])[substr(names(cef[[i]]),1,3)=="dad"] <- "dad"
    ncomplete[i] <- nrow(YXX)
    }

# special case, there are no mom dad for liver diseases
for (i in 16) {

  Y=YY[,i]
  YXX <- data.frame(Y, XX)
  YXX <- YXX[complete.cases(YXX),]
  #dim(data.frame(Y, XX))
  dim(YXX)
  Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))
  dt <-  data.frame(y=YXX[,"Y"],Xmat)
  dt.glm <- glm( y ~ . , data =dt, family = binomial(logit))
  s <- summary(dt.glm)$coefficients[,4]
  c <- summary(dt.glm)$coefficients[s<0.05, "Estimate"]
  nam <- names(c)[2:length(c)]
  sigName[i] <- list(nam)   #significant var names
  sigcef[i] <- list(c)      #significant coefficients
  cef[i] <-  list(summary(dt.glm)$coefficients[, "Estimate"]) ## all estimate
  pval[i] <- list(s)
  ncomplete[i] <- nrow(YXX)
 
 }
 return(pval)
}
########################################################################################
# function created based on https://stackoverflow.com/questions/17999920/merge-vectors-of-a-list-using-row-names-in-r
comlstVect <- function (x){
  Rows <- unique(unlist(lapply(x, names)))
  
  ## Create a matrix of NA values 
  ##   with appropriate dimensions and dimnames
  myMat <- matrix(NA, nrow = length(Rows), ncol = length(x), 
                  dimnames = list(Rows, sapply(x, colnames)))
  ## Use your `for` loop to fill it in
  ##   with the appropriate values from your list
  for (i in seq_along(x)) {
    myMat[names(x[[i]]), i] <- x[[i]]
  }
  return(myMat)
}
# x <- tryfunc()
# pval <- comlstVect(x)

```

>The last chunk of R code is inactivated because the estimations are most not converged. if readers want to check the results, please decomment the last severa commented lines of commands such as "x =  tryfunc()", and "pval <- comlstVect(pval)". pval is a matrix for all the p-values of the estimations for the 16 models with parents health information. 

# 8. Rare Events and Power of Predictions
Unfortunitly, we have unbalanced data, and most diseases occure rarely. Thus the true false in the confusion table is large. This is a situation when sensitivity is large, precision is small. Such a situatioin implies that if the prediction rule of the logistic models capture most of the diseases, there would be a larger number of false positive prediction even when the rate of error is small and specificity is large. IIn this section, F-score is constructed as the harmonic mean of sensitivity and precision for a sequence of prediciton rules using different probability. The best prediction rule corresponding to the maximum F-score for each model is found and reported in this section.

```{r, ch8, warning=FALSE, message=FALSE, error=FALSE}

# reduced model predicted probability for each aldults
pred.y.logit10 <- readRDS("/home/Nianfu/pred_y_logit10.rds")
obs.y.logit10 <- readRDS("/home/Nianfu/obs_y_logit10.rds") 

# Diseases data
YY <- hlth_par %>% 
    select(dis1, dis2, dis3, dis4, dis7, 
           dis8,dis9, dis10, dis11, dis12,
           dis14, dis18, dis21, dis28, dis29,  
           dis99)
#Binary matrix
Ymat <- data.frame(model.matrix(~., data = YY ))
#rate of diseases/cases
case <- sapply(Ymat, function(x) sum(x, na.rm = TRUE))
all <- sapply(Ymat, function(x) sum(!is.na(x)))
Case.ratio <-case/all 
r <- data.frame(case[-1], round(Case.ratio[-1],3))
rownames(r) <- dis_Names
colnames(r) <- c("cases","rate_of-occurance")
r
taul<- r
```

> Fourteen out of 16 diseases have rates of occurance less than 3%. 

```{r, Ch8_second, warning=FALSE, message=FALSE, error=FALSE}
#confusion matrix
conf <- pred <- pred.y.logit10

maxl <- matrix(0, nrow = 16, ncol = 7)
for (f in seq(from =0, to=1, by=0.01)){
  # TRUE = casse or yes disease 
  pred <- lapply(pred.y.logit10, function(x) {x>f})
  
  for (i in 1:16){
    t <- table(obs.y.logit10[[i]], pred[[i]])
    if (nrow(t)==2 & ncol(t)==2) {
      sensitivity <- t[2,2]/(t[1,2]+t[2,2])
      specificity <- t[1,1]/(t[1,1]+t[2,1])
      precision <- t[2,2]/(t[2,1]+t[2,2])
      Fscore <- 2*sensitivity*precision/(sensitivity+precision)
    
      if(Fscore>maxl[i,7] & !is.na(Fscore)) {
        maxF=Fscore
        maxl[i,] <- c(i, f, round(sensitivity,2), round(specificity,2), round(precision,2), round(Fscore,2), round(maxF,4))
      }
    }
  }
}

Fscor <- data.frame(dis_Names, maxl)
names(Fscor) <- c("","","pred.pr", "sensitivity","specificity","precision","Max F score", "max F")
Fscor

#clearn memory
rm( "dt.glm", "hlth_par", "obs.y.logit", "obs.outputs.rf",  "obs.y.logit10", "obs.y.rf", "obs.y.rf10", "pred.y.logit", "pred.outputs.rf", "pred.y.logit", "pred.y.logit10", "pred.y.rf", "pred.y.rf10","r1","r2")
#"dt.lg", "dt0", "fX","m.data", "m.out","ps","dtxx", "obs.y.logit",,"rf.pred.curr" 

```

> The best F-score for these models are less than 0.5. About half of them have only F-scores equal or less than 0.1.

# 9. Discussion of Rare Event Models

The low rates of diseases in the general US population imply the dificulty of prediction with such unbalanced data. An alternative is to estimate Rare Event Models which use only all the cases and selected proportion of controls from the original survey data. The estimates are adjusted for such data selection. The method used for such models is called Rare Event Logistic Model.

## 9.1. Rare Event Full Models
Randomly selected 5 controls are selected for each of all the case. R package "Zelig" can estimate rare event logistic models. Variables/categories of variables that are significantly correlated to diseases are identified with rare event logistic model estimation. 
```{r, ch9.1, warning=FALSE, message=FALSE, error=FALSE}

# install.packages("MatchIt")
#library("MatchIt")
# install.packages('Zelig')
library("Zelig")
library(dplyr)

## region is excluded 
XX <- hlth %>% 
  select(age, sex, race,  smk, alc, sleep, activ,  
          fm_size, marriage, par_stat, BMI) ##HHX, FMX, FPX,age_smk, region
# reset the YY incase order change
YY <- hlth %>% 
  select(dis1, dis2, dis3, dis4, dis7, 
         dis8,dis9, dis10, dis11, dis12,
         dis14, dis18, dis21, dis28, dis29,  
         dis99)

j=1
m=5
cef <- NA
sigName =NA 
dstName=NA
sigcef <- NA
pval <- NA

for (j in c(1:10,13, 16)) {
  # j is the jth diseases such as dis1, dis2,...dis29, dis99
  Yj=YY[,j]
  Y=ifelse(Yj==levels(Yj)[1], 0, 1)
  YXX <- data.frame(Y, XX)
  YXX <- YXX[complete.cases(YXX),]
  Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))

  # reduced 10 variables
  nam <- colnames(Xmat)[-1]
  fX <- Xmat[,-1]

  YX <- data.frame(YXX$Y, fX)
  Xmat <- data.frame(model.matrix(~., data = YX[, c(2:ncol(YX))] ))  #exclude y
  dt0 <-  data.frame(y=YXX[,"Y"],Xmat[,-1])
  
  # model formula
  formulaX <- as.formula(paste("y ~", nam[1],"+",nam[2],"+",nam[3],"+",nam[4],"+",nam[5],"+",
                               nam[6], "+",nam[7],"+",nam[8],"+",nam[9],"+",nam[10],"+",
                               nam[11],"+",nam[12],"+",nam[13],"+",nam[14],"+",nam[15],"+",
                               nam[16], "+",nam[17],"+",nam[18],"+",nam[19],"+",nam[20],"+",
                               nam[21],"+",nam[22],"+",nam[23],"+",nam[24],"+",nam[25],"+",
                               nam[26], "+",nam[27],"+",nam[28]))
  # number controls needed, m is the number controls for each case
  nmatch <- sum(dt0$y)*m
  #Id the controls
  mr <- sample(rownames(dt0[dt0$y==0,]), size=nmatch, replace=FALSE)
  # select controls
  dtm <- dt0[mr,]
  # form a new data, m controls for each case
  dt <- as.data.frame(rbind(dt0[dt0$y==1,],dtm))
  # Rare event logistic estimation
  dt.zlig  <- zelig(y ~ .,  data =dt, model = "relogit", tau = taul[j,2], cite = FALSE)
  # coefficients
  b <- coefficients(dt.zlig)
  cef[j]=list(b)
  #p-values
  pval[j] <- list(get_pvalue(dt.zlig))
}   

cefTab <- as.data.frame(cef[c(1:10,13,16)])
pvalTab <- as.data.frame(pval[c(1:10,13,16)])
sig <- as.data.frame(pval[c(1:10,13,16)])<0.05
sig=ifelse(sig==0, NA, sig)
sigTab <- cefTab*sig
names(sigTab)=names(YY[c(1:10,13,16)])
sigTabf <- round(sigTab,2)
#significant coefficients only
print(sigTabf)  

```

>Included in the table above are only significant coefficents from the estimation of full rare event model, using all variables. These significant coefficients can also be respresented on a heeat map.

## 9.2 Rare Event Heatmap of P-Values of Full Models
Similar to the heat map produced for the regular logistic model, a heat map for a rare event model is constructed to show the correlation of predictors and diseases. Four diseases are excluded from the heat map because of smaller number of records used in the estimation. The regressor for US regions is excluded because the Zelig package does not produce coefficient for it in some of the models. 

```{r, ch9.2, fig.height=8, warning=FALSE, message=FALSE, error=FALSE}
library(gplots)

HMsign <- cefTab
HMsign[cefTab>0]=1
HMsign[cefTab<0]=-1
# manage small p values
ps <- log(pvalTab)
ps[pvalTab>0.05]=NaN
pmat <- ps*HMsign*(-1)
X <- as.matrix(pmat[-1,])

Z <- X
Z[X>-log(0.0001)]=-log(0.0001)
Z[X<=log(0.0001)]=log(0.0001)
colnames(Z) <- dis_Names[c(1:10,13,16)]
#heatmap(X, Rowv = NA, Colv = NA, scale = "column",
#        main = "heatmap(*, NA, NA) ~= image(t(X))")
heatmap.2(Z, col=colorpanel(10,"green","yellow", "red"),  Rowv = FALSE, Colv = FALSE,  na.color = "white",
          margin=c(8,8), trace="none", 
          main="                    Significance of Rare Event Model",
          
          dendrogram = "none", key.xlab="Truncated and signed log p-value", 
          key.par = list(mgp=c(1.2,0.3,0), mar=c(2.5,2.5,4,0)), key.title = "- coeff               + coeff")

```

## 9.3 Prediction with Rare Event Logistic Model

### 9.3.1 Prediction of Case-Control Samples, Top 10 Variable Rare Event Logistic Model

Following the previous regular logistic model using top 10 variables, the top 10 most significant variables from section 9.2 are selected for the prediction with rare even model. ROC and AUC are produced to assess the quality of prediction

```{r Ch9.3.1, warning=FALSE, message=FALSE, error=FALSE}
# install.packages("MatchIt")
library("MatchIt")
#install.packages('Zelig')
library("Zelig")

# taul is from above chunk "Ch8" r

j=1
m=5
pred.zelig <- NA
obs.zelig <- NA 

## only 1:10, 13, 16 diseases converge
for (j in c(1:12) ){
  # j is the jth diseases such as dis1, dis2,...dis29, dis99
  YY0 <- YY[,c(1:10,13,16)]
  Yj=YY0[,j]
  Y=ifelse(Yj==levels(Yj)[1], 0, 1)
  YXX <- data.frame(Y, XX)
  YXX <- YXX[complete.cases(YXX),]
  Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))

  # reduced 10 variables
  z <-  colnames(Xmat)[-1] # logistic coefficient from above code
  zp <- pvalTab[-1,j] #pvalue matrix from above full code
  tp10nam <- z[order(zp)][1:10] # top 10 x
  #print(paste("j=",j))
  tp10X <- Xmat[, tp10nam]

  Y10X <- data.frame(YXX$Y, tp10X)
  Xmat10 <- data.frame(model.matrix(~., data = Y10X[, c(2:ncol(Y10X))] ))  #exclude y
  dt0 <-  data.frame(y=YXX[,"Y"],Xmat10[,-1])
  
  #propensity score matching formula
  formulaX <- as.formula(paste("y ~", tp10nam[1],"+",tp10nam[2],"+",tp10nam[3],"+",tp10nam[4],"+",tp10nam[5],"+",
                               tp10nam[6], "+",tp10nam[7],"+",tp10nam[8],"+",tp10nam[9],"+",tp10nam[10]))
  
  nmatch <- sum(dt0$y)*m
  mr <- sample(rownames(dt0[dt0$y==0,]), size=nmatch, replace=FALSE)
  dtm <- dt0[mr,]
  dtxx <- as.data.frame(rbind(dt0[dt0$y==1,],dtm))

  #dtxx <- m.data
  n = nrow(dtxx)
  #N = nrow(dt0)
  K = 10
  set.seed(1234)
  # randomly select one of the K from N, the number of rows of data
  s = sample(1:K, size = n, replace = T) # used for label rows of data titanic for each group       
  pred.outputs.logit <- vector(mode = "numeric", length = n)
  obs.outputs.logit <- vector(mode = "numeric", length = n)
  offset <- 0 # offset is a pointer used for determine where to pending results for the next loop	
  i=1
  for(i in 1:K){
    	train <- filter(dtxx, s != i)
    	
    	test1 <- filter(dtxx, s == i)
    	#test2 <- ex.data
    	test <- test1
    	obs.outputs.logit[1:(n-length(s[s != i])) + offset] <- test$y
    
      dt.zlig  <- zelig(y ~ .,
                data =train, model = "relogit", tau = taul[i,2], cite = FALSE)

      b <- coefficients(dt.zlig)
      test.x <- as.matrix(test %>% mutate(y=1) )
      prb <- 1/(1+exp(test.x%*%b*(-1)))

      pred.outputs.logit[1:(n-length(s[s != i])) + offset] <- prb
     	offset <- offset + (n-length(s[s != i]))
     # 	return(list(obs.outputs.logit, pred.outputs.logit))
    }
  pred.zelig[j] <- list(pred.outputs.logit)
  obs.zelig[j] <- list(obs.outputs.logit) 
}


#confusion matrix for 1:11, 13, 16
## A combination of propensity score match and rare event logistic model
conf <- pred <- pred.zelig
f=0.2
maxl <- matrix(0, nrow = 12, ncol = 7)
for (f in seq(from =0, to=0.3, by=0.001)){
  # TRUE = casse or yes disease 
  pred <- lapply(pred.zelig, function(x) {x>f})
  
  for (i in 1:12){
    t <- table(obs.zelig[[i]], pred[[i]])
    if (nrow(t)==2 & ncol(t)==2) {
      sensitivity <- t[2,2]/(t[1,2]+t[2,2])
      specificity <- t[1,1]/(t[1,1]+t[2,1])
      precision <- t[2,2]/(t[2,1]+t[2,2])
      Fscore <- 2*sensitivity*precision/(sensitivity+precision)
    
      if(Fscore>maxl[i,7] & !is.na(Fscore)) {
        maxF=Fscore
        maxl[i,] <- c(i, f, round(sensitivity,2), round(specificity,2), round(precision,2), round(Fscore,2), round(maxF,4))
      }
    }
  }
}

Fscor <- data.frame(dis_Names[c(1:10,13, 16)], maxl)
names(Fscor) <- c("","","pred.pr", "sensitivity", "specificity", "precision","MaxFscore", "maxF")
Fscor
```

>Even for the predictions of the selected sample data, the maxmum F-score (MaxFscore) are mostly around 0.5, and the sensitivity is still about 0.3, and specificity about 0.9 for most of the models. but these data do not reflect the survey population. Predictions for the population should predict all the data as the next section will do.

### 9.3.2. Prediction of All Data, Top-10 Variable Rare Event Logistic Model 
The test data used in the cross validation of the top-10 variable rare event logistic model are all the data except the training data. Thus the predictions represent all the population. The models take a parameter tau from the test data. Please see variables, "top10nam", "ex.data","e" and command lines with "zelig()" for the method used. "MatchIt" could be used but not in this section, because controls in the matched data are randomly selected. When controls selected with nearest match, most models do not converge. Random controls perform better than nearest match. 

```{r Ch9.3.2, warning=FALSE, message=FALSE, error=FALSE}
# install.packages("MatchIt")
#library("MatchIt")
#install.packages('Zelig')
library("Zelig")

j=1
pred.zeligAll <- NA
obs.zeligAll <- NA 

## only 1:11, 13, 16 diseases converge
for (j in c(1:12) ){
  # j is the jth diseases such as dis1, dis2,...dis29, dis99
  YY0 <- YY[,c(1:10,13,16)]
  Yj=YY0[,j]
  Y=ifelse(Yj==levels(Yj)[1], 0, 1)
  YXX <- data.frame(Y, XX)
  #complete records only
  YXX <- YXX[complete.cases(YXX),]
  # reconstruct egressor matrix used in section 9.1 
  Xmat <- data.frame(model.matrix(~., data = YXX[, c(2:ncol(YXX))] ))

  # Top 10 variables from section 9.1 for full rare event model
  z <-  rownames(cefTab)[-1] # logistic coefficient from above section 9.1 code
  zp <- pvalTab[-1,j] # pvalue matrix from above section 9.1 code
  tp10nam <- z[order(zp)][1:10] # top 10 x
  #print(paste("j=",j))
  
  #select the top 10 from Xmat
  tp10X <- Xmat[, tp10nam]
  
  # matched data for rare event regression
  Y10X <- data.frame(YXX$Y, tp10X)
  Xmat10 <- data.frame(model.matrix(~., data = Y10X[, c(2:ncol(Y10X))] ))  #exclude y
  
  # Top 10 variables and the jth disease data for the rare event model
  dt0 <-  data.frame(y=YXX[,"Y"],Xmat10[,-1])
  
  # Model formula
  formulaX <- as.formula(paste("y ~", tp10nam[1],"+",tp10nam[2],"+",tp10nam[3],"+",tp10nam[4],"+",tp10nam[5],"+",
                               tp10nam[6], "+",tp10nam[7],"+",tp10nam[8],"+",tp10nam[9],"+",tp10nam[10]))
  #Number of controls
  nmatch <- sum(dt0$y)*m
  mr <- sample(rownames(dt0[dt0$y==0,]), size=nmatch, replace=FALSE)
  dtm <- dt0[mr,]
  #matched matrix corresponding to cases of jth disease
  dtxx <- as.data.frame(rbind(dt0[dt0$y==1,],dtm))
  n = nrow(dtxx)
  N = nrow(dt0)
  K = 10
  set.seed(1234)
  
  # randomly select one of the K from N, the number of rows of data
  s = sample(1:K, size = n, replace = T) # used for label rows of data titanic for each group       
  pred.outputs.logit <- vector(mode = "numeric", length = N)
  obs.outputs.logit <- vector(mode = "numeric", length = N)
  offset <- 0 # offset is a pointer used for determine where to pending results for the next loop	
  
  #Unmatched data excluded data from sample
  ex.data <- dt0 %>% filter(!(rownames(dt0) %in% (rownames(dtxx))))
  e = sample(1:K, size = N-n, replace = T) # used for label rows of Ex.data titanic for each group       

  i=1  #example for test code
  for(i in 1:K){
    	train <- filter(dtxx, s != i)
    	# test1= the test data from matched data
    	test1 <- filter(dtxx, s == i)
    	# test2=  all unmatched data for taining i
    	test2 <- filter(ex.data,e==i)
    	test <- rbind(test1, test2)
    	obs.outputs.logit[1:(nrow(test) + offset)] <- test$y
    	
    	#tau value are the ratio of cases in the test data, this make the model flexble.
      tau_i=sum(test$y)/nrow(test)
      dt.zlig  <- zelig(y ~ .,
                data =train, model = "relogit", tau = tau_i, cite = FALSE)
      
      # Probability prediction for the test data, a list
      prb <- predict(dt.zlig, test, type = "response")
      pred.outputs.logit[1:(nrow(test) + offset)] <- prb[[1]]  # [[1]] to obtain the vector
     	offset <- offset + nrow(test)
    }
  pred.zeligAll[j] <- list(pred.outputs.logit)
  obs.zeligAll[j] <- list(obs.outputs.logit) 
}

#free some memory
Xmat <- NULL
Xmate10 <- NULL
Y10X <- Ymat <-  NULL
YX <- YXX <- NULL
YY0 <- NULL
mr <- NULL

# ROCs
library(pROC)
i=1 
plot.roc(obs.zeligAll[[i]], as.numeric(pred.zeligAll[[i]]),  lwd = 1, lty=1, ci = TRUE, col = "purple") #CV of svm
legend("bottomright", legend = c("Rare Event Logistic Model"), col = c("purple"), lwd = 2, lty = c(1))
for (i in c(2:12)){
  plot.roc(obs.zeligAll[[i]], pred.zeligAll[[i]], lwd = 1,ci = TRUE, lty=1, col = "purple", add = TRUE) #CV of rf
}

#confusion matrix for 1:11, 13, 16
## confusion table and F score for the above model
f=0.2 # for code test
# matrix for sensitivity, specificity, precision and maximum F-scores
maxl <- matrix(0, nrow = 12, ncol = 7)

#search for the boundary probability that has the maximum F-score
for (f in seq(from =0, to=0.5, by=0.001)){
  # probability x>f = casse, yes disease 
  pred <- lapply(pred.zeligAll, function(x) {x>f})
  
  # numbers for all diseases
  for (i in 1:12){
    t <- table(obs.zeligAll[[i]], pred[[i]])
    if (nrow(t)==2 & ncol(t)==2) {
      sensitivity <- t[2,2]/(t[1,2]+t[2,2])
      specificity <- t[1,1]/(t[1,1]+t[2,1])
      precision <- t[2,2]/(t[2,1]+t[2,2])
      Fscore <- 2*sensitivity*precision/(sensitivity+precision)
      # if an Fscore is larger than old one than change to the new one
      if(Fscore>maxl[i,7] & !is.na(Fscore)) {
        maxF=Fscore
        maxl[i,] <- c(i, f, round(sensitivity,2), round(specificity,2), round(precision,2), round(Fscore,2), round(maxF,4))
      }
    }
  }
}

Fscor <- data.frame(dis_Names[c(1:10,13,16)], maxl[,-1])
names(Fscor) <- c("disease","pred.pr", "sensitivity","specificity","precision","MaxFscore", "maxF")
Fscor
```

When the health of all the surveyed aldults are predicted with the rare event models, the F score and sensitivity is even smaller than the full data logistic model. 

# 10. Conclusion
The project has tabled, plotted, mapped and modeled variations of characters and diseases of interviewed U.S. adults. Logistic model were estimated to capture quantitively these relations. The results are display on a heatmap, and summarized in tables. The model estimation results are similar to the exploratory analyes. The major conclusions include
1. Smoking is a negative factor for health and strongly correlated with most types of diseases;
2. Physical activities are positive to health, and negatively correlated to most diseases;
3. 7 hours of sleep is positively correlated to health, and less sleep is definitely positively correlated with many diseases;
4. Interestingly, drinking has a mixed correlation with health, While former alcohol users may be positively related to diseases, existing alcohol users are relatively healthier than abstainers;
5. Females, compared to males are less likely to have chronic diseases;
6. There are some regional differences of drinking, smoking, and selective diseases;
7. Prediction model have high value of AUC, but the F score for the prediction model is very low for most of the diseases because of the rareness of the diseases, or the imbalanced data. Thus the prediction model may predict too many false cases or not catch true cases. However as a prevention for deadly diseases a positive prediction for a disease even with a low precision can be considered as an alarm . Further medical examination should be performed.
8. Rare event models can't predict disease risks better than logistic models.

# 11. Future Improvement Possible
Although plots show that adults' diseases share some similarities with their parents, but only 754 adults in the sample have parents information, not big enough for modeling the correlation between diseases of adults and their parents.

The models are estimated based on existing cases and predictors without any scientific knowledge of causality. Thus the models do not reflect causality between predictors and diseaese. The significant predictors are only potential candidates of causes. They may also be the results of life style change because of the wareness of risk of disease. Although these models could predict known disease well, they may not perform as well for prediction of risk of unknown cases as they do for known diseases.  

Propensity score matching method could be combined with rare event logistic estimation and may produce consistent estimate with relatively smaller number of matched controls. 

# 12 Aknowledgement 

I would like to give special thanks to Dr. Blanca Himes, at the Institute of Biomedical Informatics of University of Pennsylvania. R methods and code are mostly based on the practicia of her Data Science for Biomedical Informatics class. This project also has the support from Dr. Stephen Kawut at the Center for Clinical Epidermiology and Biostatistics of the Perelman School of Medicine, University of Pennsylvania.


