---
title: "COVID19 and Testing Ratios"
author: "Saba Negash"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

### Overview
My project is hinged on the COVID19 and the unclear testing statistics subject not only to the U.S. but other countries as well. It is has been an uncertainty to others as administrations build their statistics on commendatory perspectives rather than the practical actuality. This project will shed light on this and hopefully provide a rational answer to not only me but whoever is confused and curious on the belief. I will be using primary articles and primary resources online based on the  statistics of the testing ratio in several states as well as a few countries. I have spoken with the following:
-Martine Kersaint, HUP, Biomedical Engineering
-Beryl Greywoode, CHOP, Pediatric Doctor

### Introduction 
The calculation of the prevalence rates and general statistics of the COVID19 Pandemic has been an area of confusion, with administration preferring to relay the statistics that give a favourable outlook, instead of the reality. The project is interested in observing the test ratio for the COVID19 pandemic as a statistic of interest in the pandemic. The project will aim at conducting a comparative study of the testing being done in different countries with special focus in the test ratios among positive and negative diagnoses.
Given that the pandemic itself is an epidemiological event, the field of epidemiology will make significant contribution to the study, in addition to the fields of statistics, demography and spatial studies. 


### Methods
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  

```{r}
## LOading the data set
covid19 <- read.csv2(file = "covid19.csv", header = FALSE, sep = ",")
# viewing the first and the last five rows of the data set
head(covid19)
tail(covid19)
```

```{r}
## checking missing values
sum(is.na(covid19))
```

```{r}
## Performing descriptive statistics
library(psych)
describe(covid19, na.rm = FALSE)
attach(covid19)
```



```{r}
# histogram with added parameters
hist(Recovered...100.Cases,
main="Histogram of recovery rate",
xlab="Recovery rate",
xlim=c(50,100),
col="darkmagenta",
freq=TRUE
)
```

```{r}
# histogram with added parameters
hist(Deaths...100.Cases,
main="Histogram showing death rate",
xlab="Deaths rate",
xlim=c(0,30),
col="red",
freq=TRUE
)
```



```{r}
library(ggplot2)
ggplot(data = covid19) + 
  geom_point(mapping = aes(x = Deaths...100.Cases, y = Recovered...100.Cases))
```

```{r}
boxplot(covid19$Recovered...100.Cases)
```


```{r}
boxplot(covid19$Deaths...100.Cases)
```


```{r}
ggplot(data = covid19) + 
  geom_point(mapping = aes(x = Confirmed, y = Deaths))
```


```{r}
ggplot(data = covid19) + 
  geom_point(mapping = aes(x = Recovered, y = Active))
```


## Results

-The above graph shows that there exist a positive relationship between the deaths/100 cases and recovered/100 cases. Furthermore, we also observe that the recovery rate is higher than the death rate.
-The above boxplot shows that the variable Recovered / 100 Cases is skewed, this is because it has a longer lower whisker than the upper whisker. Therefore, the distribution for this variable is that is skewed to the left or its negatively skewed.
-The above boxplot shows that the variable Deaths / 100 Cases is skewed, this is because it has a longer upper whisker than the lower whisker. Therefore, the distribution for this variable is that is skewed to the right or its positively skewed. Moreover this variable also contain outliers which can be observed from the upper whisker.
-The above graph shows that there exist a positive relationship between confirmed cases and deaths cases. Furthermore, we also observe that confirmed cases are higher than the deaths cases.
-The above graph shows that there exist a positive relationship between active cases and recovered cases. Furthermore, we also observe that recovered cases are higher than active cases.
## Conclusions
From the above results, we conclude that there exist a positive relationship between the rate of death and the rate of recovery. In addition, the rate of recovery is higher than the death rate which simply means that many people contact the disease but they survive. Next, we check the distribution between the recovery rate and death rate and we notice that both of these variables are skewed. Moreover, the variable death rates also have outliers. Next, we check the relationship between the confirmed cases and the number of deaths and we notice that dispite there are several people contacting this disease few of them dies. This is also noticed between the active cases and the number of patients that have recovered that is many patients recover from the disease compared to those who are active.


































