---
title: "BMIN503/EPID600 Project Template"
author: "Jessica Meeker"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but keep the headers.

### Overview
In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

Maternal mortality and morbidity rates have seen a decline globally; however, in the United States these numbers continue to rise, in particular among women of color. I will use Electronic Medical Record (EMR) data from a cohort of women in the University of Pennsylvania Health System (UPHS). The goal of this project is to perform exploratory statistical methods to describe the exposure and outcome variables in the data set and machine learning methods (Lasso) to create a model of neighborhood level stress in Philadelphia that predicts maternal mortality and morbidity. 

I spoke with Drs. Trevor Penning, Ray Bai & Mary Regina Boland about my project. With Dr. Penning I discussed the project from a more broad lens of environmental exposures, both natural and social. Dr. Bai was able to provide much needed insight into the methodology of logistic regression with a lasso penalty on the predictors. Dr. Boland and I discussed overall project context and some of the considerations necessary for which machine learning method to choose. 

[Github Repository](https://github.com/jemeeker10/BMIN503_Final_Project)

### Introduction 
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.

In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

Maternal morbidity and mortality persist as key indicators of women’s health both globally and in the US. Pregnancy related deaths that occur during pregnancy or within a year following pregnancy, have strikingly doubled from 7.2 deaths per 100,000 live births in 1987 to 18.0 deaths in 2014. While mortality is steadily increasing nationally, severe maternal morbidity (SMM) is actually far more common in the US than maternal mortality by nearly 100 times. SMM describes unexpected, poor outcomes of labor or delivery that may result in short or long term consequences that can be significant for the mother, child and their family. However, not all women are at equal risk for SMM or maternal mortality and severe health disparities persist. Notably, African American women are upwards of four times as likely to die of complications from pregnancy as compared to white women. To date, no genetic variant has been found to explain this discrepancy, begging the question of what else is transpiring to propagate such a stark health disparity. In this project I am interested in understanding the role the social, built, and natural environments play on these maternal health disparities. 

This problem is very multidisciplinary in nature as it's success depends upon collaboration of those from a variety of fields including maternal health clinicians, epidemiologists, statisticians, informaticians, environmental toxicologists, and health disparities researchers. I was privileged to be able to meet with three people who were trained in or who have significant experience in all of these roles. This problem is complex as it is a environmental, maternal health, and social disparities problem. In meeting with the three people I consulted regarding this project, I was provided with conceptual and methodological advice that will aid in the success of this project. 


### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

The primary data sets used for this project include: 1) publically available data from the American Community Survey (ACS) from the Census Bureau 2) non-publically available data from the UPHS electronic medical record (EMR) system and 3) shape files from the Census Bureau. Multiple ACS files were used in the lasso prediction model, so as to let the algorithm decide which covariates were of most importance. Please note the socio-demographic variables that were used outlined in the methods code. The maternal health data which was derived from the UPHS EMR, is from 2010-2017 and includes pregnant women who were seen in the Penn System throughout their pregnancy. 

First, all necessary packages are loaded. 

```{r}
#All package libraries
```

Next, we load in all of our predictor variables from the ACS, our outcome variables from the UPHS EMR, and clean the data for analysis. 

```{r}
#ACS DATA

#PENN DATA
```

Histograms were generated to look at overall hemmoraghing rate per deliveries

```{r}
ggplot (data = __, aes (x = hemmoraghingrate)) +
  geom(histogram (bins = 20, aes (y = ..density..), colour = "black", fill = "white") +
  ggtitle("Total hemmoraghing rate per delivery") +
  geom_density(alpha = .2, fill = "lightpurple") +
  faced_grid(~Year)
```

Breakdown of Morbidity by Indicator Category
```{r}
#maybe we'll do this if not too hard 
## e.g hemmoraghing/total morbidities

```


Outcome and Predictor data need to be joined based upon geocoding data.

```{r}
#join geocoded pregnant women with ACS variable data
```

Download shape file for Philadelphia for mapping purposes

```{r}
#import city shapefile as an sf object

Phila.sf <-.......insert code here 

str(Phila.sf)

Phila.sf <- Phila.sf %>%
  select (var x, y, z, etc...) %>%
  rename (GEOID = 'whatever geoid is in sf')

#join shapfile to other data 

alldata <- inner_join(Phila.sf, ACS_UPHS, by = "GEOID")
```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

We will now utilize the machine learning technique of lass to create a predictive neighborhood index for maternal health. 

```{r}
#Insert lasso code here to run logistic regression with lasso penalty on predictors

#normalize to create scale 


```

We will make some interesting maps showing the spatial association between neighborhood stressors and maternal morbidity outcomes. 

```{r}
#insert leaflet code 
```

### Conclusion and Future Directions

