---
title: "An Analysis of the Effects of Operational Factors on the Success of Phase II Industry Sponsored Studies"
author: "Kyle O'Neil"
output:
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

#Overview
###Drug development faces extremely low success rates, with approval dependent on clinical studies that demonstrate the benefits of a particular treatment outweight its risks. With an overall success rate of 14% in clinical development (Estimation of clinical trial success rates and related parameters, Biostatistics - Jan. 2018), evaluating the numerous factors that play a role in clinical studies is crucial. In this study, I evaluated 20 operational parameters that may affect clinical trial success from clinicaltrials.gov data in the first analysis of how operational study design affects trial outcomes. 

#Introduction
###Previous analyses of drug development success have not looked at how trial design might affect study outcomes. Trial design may be particularly relevant in Phase II studies, which have the lowest success rate and are where drugs often need to show efficacy in addition to safety. Therefore, a collection of 50 industry sponsored Phase II studies were evaluated to generate models of succesful clinical trials and determine the relevance of operational factors. Ultimately, this study aims to highlight factors that may improve the chance of trial success in an industry where therapeutic development costs around 2.5 billions dollars (Innovation in the Pharmaceutical Industry: new estimates of R&D costs - Journal of Health Economics). 

###Analyzing clinical studies data incorporates multiple fields of knowledge ranging from biostatics to clinical medicine and physiology. With certain operational parameters defined in clinical trials and others uncertain, feature selection from availible clinicaltrials.gov data was crucial and may affect the final analysis. Variables were chosen that were consistently availible although more relevant variables may exist in alternative data formats. Determining the scope of the analysis will also be critical; even within Phase II studies, there are a wide variety of different study types which can be considered as input or narrowing parameters for the analysis.

#Methods
###50 Phase II studies from 10 large biopharmaceutical companies were selected for analysis on clinicaltrials.gov with results posted in the last 5 years (since January 2014). The 10 biopharmaceutical companies included GSK, Janssen, Gilead, Novartis, Amgen, Abbvie, Genentech, Merck, Pfizer, and Regeneron. Studies were chosen at random across a range of indications and data points were collected manually for each study. The data points recorded included whether there was a comparison group for the main therapeutic, the trial masking, primary endpoints, number of endpoints (primary and secondary), primary endpoint values, number of study sites, number of study participants, the length of study, number of inclusion/exclusion criteria, any p values, and whether the study was advanced into Phase III. Exploratory data analysis and modeling were conducted to see if there were interesting trends in the selected data points. 

#Results - Exploratory Data Analysis
```{r}
library(gridExtra)
library(grid)
library(ggplot2)
library(lattice)
library(ggplot2)
library(tidyverse)
#install.packages("GGally")
library(GGally)
library(dplyr)
dataset <- read.csv("Phase II Study Catalog CSV.csv", header=TRUE)
str(dataset)
summary(dataset)
dataset %>%
    group_by(SOC.Comparison) %>%
    summarise(n())

dataset %>%
    group_by(Masking) %>%
    summarise(n())

dataset %>%
    group_by(Primary.Endpoint.Success) %>%
    summarise(n())

dataset %>%
    group_by(Moved.to.Phase.III) %>%
    summarise(n())

dataset %>%
    group_by(First.time.in.Phase.II) %>%
    summarise(n())

dataset %>%
    group_by(P.Value) %>%
    summarise(n())

#Could call Alcon call center to get NA # of study sites or approximate

#Exploration of the trial length,  subject number, and trial sites numbers
primaryendpointtimeplot <- ggplot(dataset,               
    aes(x=Primary.Endpoint.Success, 
    y=Time.for.Primary.Endpoint..yrs.)) +
    geom_boxplot() + ggtitle("Trial Time vs. Primary 
    Endpoint Sucess") + theme(plot.title = element_text(hjust= 0.5))
studysitesplot <- ggplot(dataset, aes(x=Primary.Endpoint.Success,      y=Number.of.Study.Sites)) +
  geom_boxplot() + ggtitle("Study Site # vs. Primary 
  Endpoint Sucess") + theme(plot.title = element_text(hjust = 0.5))
studysubjectplot <- ggplot(dataset, aes(x=Primary.Endpoint.Success,    y=Number.of.Subjects.Enrolled.in.the.Main.Treatment.Arm)) +
  geom_boxplot() +
  ggtitle("# of Subjects v. Primary 
  Endpoint Sucess") +      
  theme(plot.title = element_text(hjust = 0.5))
grid.arrange(primaryendpointtimeplot, studysitesplot, studysubjectplot, nrow = 1)


#Exploration of Inclusion/Exclusion Criteria Metrics
numberofinclusionplot <- ggplot(dataset,
    aes(x=Primary.Endpoint.Success, y=Number.of.Inclusion.Criteria))     +   geom_boxplot(fill = 'blue', colour = 'red', alpha = 0.7)+ 
    ggtitle("# of IC vs. Prim. Endpoint 
    Sucess") + theme(plot.title = element_text(hjust = 0.5))
numberofexclusionplot <- ggplot(dataset, 
    aes(x=Primary.Endpoint.Success, y=Number.of.Exclusion.Criteria))     + geom_boxplot() + ggtitle("# of EC vs. Prim. Endpoint 
    Sucess") + theme(plot.title = element_text(hjust = 0.5))
ratioinclusionexclusionplot <- ggplot(dataset, 
    aes(x=Primary.Endpoint.Success,      
    y=Ratio.of.Inclusion.to.Exclusion.Criteria)) +
    geom_boxplot() + 
    ggtitle("IC:EC vs. Prim. Endpoint  
    Success") +       
    theme(plot.title = element_text(hjust = 0.5))
grid.arrange(numberofinclusionplot, numberofexclusionplot, ratioinclusionexclusionplot, nrow = 1)

#Exploration of Number of Endpoint Metrics
primendptsplot <- ggplot(dataset, aes(x=Primary.Endpoint.Success,          y=Number.of.Primary.Endpoints)) +
      geom_boxplot() + 
      ggtitle("Pri Endpt # vs. Primary 
   Endpoint Sucess") + theme(plot.title = element_text(hjust =             0.5))
secendpointsplot <- ggplot(dataset, aes(x=Primary.Endpoint.Success, 
      y=Number.of.Secondary.Endpoints)) +
      geom_boxplot() + 
      ggtitle("Sec Endpt # vs. Primary 
  Endpoint Sucess") + theme(plot.title = element_text(hjust =              0.5))
endpointsratioplot <- ggplot(dataset, aes(x=Primary.Endpoint.Success,       y=Ratio.of.Sec..Endpts..To.Prim..Endpts.)) +
      geom_boxplot() + 
      ggtitle("SE:PE Ratio vs. Primary 
  Endpoint Sucess") +        
      theme(plot.title = element_text(hjust = 0.5))
grid.arrange(primendptsplot, secendpointsplot, endpointsratioplot, nrow = 1)

#Histograms of the data
#hist(dataset$Time.for.Primary.Endpoint..yrs.)
#hist(dataset$Number.of.Subjects.Enrolled.in.the.Main.Treatment.Arm)
#hist(dataset$Number.of.Study.Sites) 

#hist(dataset$Number.of.Inclusion.Criteria) 
#hist(dataset$Number.of.Exclusion.Criteria) 
#hist(dataset$Ratio.of.Inclusion.to.Exclusion.Criteria) 

#hist(dataset$Number.of.Primary.Endpoints)
#hist(dataset$Number.of.Secondary.Endpoints)
#hist(dataset$Ratio.of.Sec..Endpts..To.Prim..Endpts.)

smalldataset <- dataset %>%
    select(Primary.Endpoint.Success, Time.for.Primary.Endpoint..yrs., Number.of.Study.Sites, Number.of.Subjects.Enrolled.in.the.Main.Treatment.Arm,  Number.of.Inclusion.Criteria, Number.of.Exclusion.Criteria,  Ratio.of.Inclusion.to.Exclusion.Criteria, Number.of.Primary.Endpoints, Number.of.Secondary.Endpoints, Ratio.of.Sec..Endpts..To.Prim..Endpts.) 
ggpairs(smalldataset)
names = character()
for (i in 2:10){
  names[i] = colnames(smalldataset)[i]
} 
model = numeric()
for (i in 2:10) {
model[i] <- summary(lm(get(names[i])~Primary.Endpoint.Success, data=smalldataset))$coefficients[2,4]
}
model
variableswp <- cbind(names, model)

pvaluedataset <- subset(dataset, P.Value == "Yes")
pvaluedataset2 <- pvaluedataset %>%
    select(Primary.Endpoint.Success, Time.for.Primary.Endpoint..yrs., Number.of.Study.Sites, Number.of.Subjects.Enrolled.in.the.Main.Treatment.Arm,  Number.of.Inclusion.Criteria, Number.of.Exclusion.Criteria,  Ratio.of.Inclusion.to.Exclusion.Criteria, Number.of.Primary.Endpoints, Number.of.Secondary.Endpoints, Ratio.of.Sec..Endpts..To.Prim..Endpts., Number)

pvaluedataset2 %>%
    group_by(Primary.Endpoint.Success) %>%
    summarise(n())

namesp = character()
for (i in 2:11){
  namesp[i] = colnames(pvaluedataset2)[i]
} 
modelp = numeric()
for (i in 2:11) {
modelp[i] <- summary(lm(get(namesp[i])~Primary.Endpoint.Success, data=pvaluedataset2))$coefficients[2,4]
}
modelp
variableswp2 <- cbind(namesp, modelp)
#Secondary endpoint number and ratio of inclusion to exclusion criteria significant - visualize
inclusionratioplot2 <- ggplot(pvaluedataset2, aes(x=Primary.Endpoint.Success, 
      y=Ratio.of.Inclusion.to.Exclusion.Criteria)) +
      geom_boxplot() + 
                          ggtitle("IC:EC Ratio # vs. Primary 
Endpoint Sucess") + theme(plot.title = element_text(hjust =              0.5))
endpointsratioplot2 <- ggplot(pvaluedataset2, aes(x=Primary.Endpoint.Success,       y=Ratio.of.Sec..Endpts..To.Prim..Endpts.)) +
      geom_boxplot() + 
                          ggtitle("SE:PE Ratio vs. Primary
Endpoint Sucess") +        
      theme(plot.title = element_text(hjust = 0.5))
grid.arrange(inclusionratioplot2, endpointsratioplot2, nrow = 1)

pvaluedataset3 <- pvaluedataset2 %>%
    select(Primary.Endpoint.Success, Number.of.Study.Sites, Number.of.Exclusion.Criteria,  Ratio.of.Inclusion.to.Exclusion.Criteria, Ratio.of.Sec..Endpts..To.Prim..Endpts., Number)
ggpairs(pvaluedataset3)

hist(dataset$Number) #to check correlation, likely need to transform this variable 
studyadvancementfullplot <- plot(dataset$Primary.Endpoint.Success, dataset$Moved.to.Phase.III, xlab = "Primary Endpoint Success", ylab = "Moved to Phase III", main = "Association between Primary Endpoint Success and Further Study Advancement") #find four quadrants and see which one is most interesting to further observe
studyadvancementrestrictedplot <- plot(pvaluedataset$Primary.Endpoint.Success, pvaluedataset$Moved.to.Phase.III, xlab = "Primary Endpoint Success", ylab = "Moved to Phase III", main = "Association between Primary Endpoint Success and Further Study Advancement 
in Studies Reporting a P Value") #find four quadrants and see which one is most interesting to further observe - probably no no & yes yes (not influenced by business)
#This did not work - grid.arrange(studyadvancementfullplot, studyadvancementrestrictedplot, nrow = 1)

pvaluedataset %>%
    group_by(Masking) %>%
    summarise(n())

pvaluedataset %>%
    group_by(SOC.Comparison) %>%
    summarise(n())

maskingplot <- ggplot(pvaluedataset, aes(x=Masking, 
      y=Number)) +
      geom_boxplot() + 
      ggtitle("Masking vs. P Val") + theme(plot.title =                    element_text(hjust = 0.5))
socplot <- ggplot(pvaluedataset, aes(x=SOC.Comparison, y=Number)) +
      geom_boxplot() + 
      ggtitle("SOC Test vs. P Val") +        
      theme(plot.title = element_text(hjust = 0.5))
grid.arrange(maskingplot, socplot, nrow = 1)
```

#Results - Modeling
```{r}

```

