---
title: "Title Here"
author: "Alexander Chan"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
# commonly-used packages
library(ggplot2)
library(dplyr)
library(magrittr)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.
```{r, eval=TRUE}

```

### Overview
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.
```{r, eval=TRUE}

```

### Introduction 
Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.
```{r, eval=TRUE}

```

### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

###
Download and install Rtools40 to use the BLS api: https://cran.r-project.org/bin/windows/Rtools/

### Reading and Cleaning Datasets
```{r, eval=TRUE}
# tools for accessing BLS data
install_github("mikeasilva/blsAPI", force = TRUE) # Install the blsAPI by Mike Silva
library(devtools)
library(blsAPI)
library(rjson)
library(blsAPI)
```


```{r, eval=TRUE}
# query data from BLS.gov

# load api key (must have the txt file)
api_key <-  read.csv("bls_api_key.txt", header=FALSE)[[1]]

# signatures to be queried from BLS.gov, 10-year CPI data from major metropolitan areas
signatures <- read.csv("metro_area_cpi_signatures.csv") 

# payload signatures to query BLS 
month_payload <- list('seriesid'= signatures$signature_monthly) #2018-2020 monthly CPI 
annual_payload <- list('seriesid'= signatures$signature_annual) #2018-2020 annual CPI

# query CPI 2018-2020 data from major metropolitan areas from BLS 
months_cpi <- blsAPI(month_payload, return_data_frame = TRUE) # monthly CPI dataframe
annual_cpi <- blsAPI(annual_payload, return_data_frame = TRUE) # annual CPI dataframe
```

```{r, eval=TRUE}

# add columns to join region labels by
months_cpi$signature_monthly <- months_cpi$seriesID
annual_cpi$signature_annual <- annual_cpi$seriesID

# join region and state labels to the cpi data
# complete monthly dataframe for 2018-2020 regional metro cpi data
months_cpi %<>%
  left_join(signatures[, c("signature_monthly","state","area")],by="signature_monthly") %>%
  dplyr::select(-seriesID)

# complete annual dataframe for 2018-2020 regional metro cpi data
annual_cpi %<>%
  left_join(signatures[, c("signature_annual","state","area")],by="signature_annual") %>%
  dplyr::select(-seriesID)

head(months_cpi)
head(annual_cpi)

```
```{r, eval=TRUE}
# Further format dataframes
# pivot data by metro area and years, add column for mean annual CPI

# annual cpi 
a_cpi <- annual_cpi %>% 
  pivot_wider(id_cols = c("area","year"),names_from = "periodName",
              values_from = "value")%>% # pivot on area and year using values from period
  mutate_at(vars(3:4), as.numeric) %>% # change CPI values to numeric
  mutate(annual = rowMeans(select(., c("1st Half", "2nd Half")), na.rm= TRUE))# mean annual cpi

# monthly cpi
m_cpi <- months_cpi %>%
  dplyr::select(-period) %>%
  pivot_wider(id_cols = c("area","year"), names_from = "periodName", 
              values_from = "value") %>% # pivot on area and year using values from period
  mutate_at(vars(2:14), as.numeric) %>% # change CPI values to numeric
  mutate(annual = rowMeans(select(.,-c("area", "year")), na.rm = TRUE)) # mean annual cpi

head(a_cpi)
head(m_cpi)
```


```{r, eval=TRUE}
library(sf)
library(ggsn)
library(stringr)
library(grid)

# function to format the geo id column of FMR dataframes
format_id <- function(df){
  
  df$ID <- str_replace(as.character(df$fips2010),
                       as.character(df$CouSub),"")
  df$ID <- as.numeric(df$ID) + 50000000000
  df$ID <- paste0("0",as.character(df$ID))
  
  return(df)
}

# read in the housing costs: fair market rates (2017-2020)
# format relevent columns into a standard form
fmr_2017 <- read.csv("FY2017_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  mutate_at(c("fmr0", "fmr1", "fmr2", "fmr3", "fmr4"), as.numeric)
fmr_2018 <- read.csv("FY2018_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  mutate_at(c("fmr_0", "fmr_1", "fmr_2", "fmr_3", "fmr_4"), as.numeric) %>%
  rename("CouSub" = "cousub", "fmr0" = "fmr_0","fmr1" = "fmr_1","fmr2" = "fmr_2", 
         "fmr3" = "fmr_3", "fmr4" = "fmr_4", "Metro_code" = "metro_code")
fmr_2019 <- read.csv("FY2019_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  mutate_at(c("fmr_0", "fmr_1", "fmr_2", "fmr_3", "fmr_4"), as.numeric) %>%
  rename("CouSub" = "cousub","fmr0" = "fmr_0","fmr1" = "fmr_1","fmr2" = "fmr_2", 
         "fmr3" = "fmr_3", "fmr4" = "fmr_4", "Metro_code" = "metro_code") 
fmr_2020 <- read.csv("FY2020_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  mutate_at(c("fmr_0", "fmr_1", "fmr_2", "fmr_3", "fmr_4"), as.numeric) %>%
  rename("CouSub" = "cousub", "fmr0" = "fmr_0","fmr1" = "fmr_1","fmr2" = "fmr_2", 
         "fmr3" = "fmr_3", "fmr4" = "fmr_4", "Metro_code" = "metro_code") 
fmr_2021 <- read.csv("FY2021_50_County_FMR.csv", fileEncoding = "UTF-8-BOM") %>%
  mutate_at(c("rent50_0", "rent50_1", "rent50_2", "rent50_3", "rent50_4"),
            as.numeric) %>%
  rename("CouSub" = "cousub", "fmr0" = "rent50_0",
         "fmr1" = "rent50_1","fmr2" = "rent50_2", 
         "fmr3" = "rent50_3", "fmr4" = "rent50_4",
         "Metro_code" = "cbsasub21", "areaname" = "areaname21",
         "countyname" = "cntyname") 

# read in shape file of US counties
counties <- readRDS(gzcon(url('https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/uscounties_2010.rds'))) # decompress  rds file and read in shape file
counties$GEO_ID <- as.character(counties$GEO_ID)#change geo id to characters in counties df
counties$ID <- gsub("US","", counties$GEO_ID)  #remove US from ID

# read in metro area shape file and convert into spatial components
setwd("tl_2017_us_cbsa")
cbsa.sf <- st_read("tl_2017_us_cbsa.shp")
cbsa.sf %<>%
  select(GEOID, geometry,NAME, NAMELSAD) %>%
  rename(ID2 = GEOID, Metro = NAMELSAD)


# function to merge FMR dataframes with county geometries 
geo_format <- function(geo_df){
  fmr_geo <- counties %>%
    dplyr::select(c("NAME", "ID", "geometry")) %>%
    left_join(geo_df[,c("ID","fmr0", "fmr1", "fmr2", "fmr3", "fmr4",
                        "Metro_code", "areaname", "countyname")], by="ID") %>%
    distinct(Metro_code, .keep_all = TRUE)
}

# function to merge FMR dataframes with metro area geometries 
geo_format2 <- function(geo_df){
  
  geo_df %<>%
    dplyr::select(ID,fmr0,fmr1,fmr2, fmr3, fmr4, Metro_code, areaname, countyname)%>%
    distinct(Metro_code, .keep_all = TRUE) %>% # removed repeated counties
    mutate(ID2 = substr(Metro_code, 12,16)) # metro IDs
  
  fmr_geo <- cbsa.sf %>%
    dplyr::select(c("NAME", "ID2", "geometry", "Metro")) %>%
    left_join(geo_df, by="ID2") # join by metro area

}

# format geographical IDs for all FMRs
fmr_2017 <- format_id(fmr_2017)
fmr_2018 <- format_id(fmr_2018)
fmr_2019 <- format_id(fmr_2019)
fmr_2020 <- format_id(fmr_2020)
fmr_2021 <- format_id(fmr_2021)

# add county geographical data to all FMRs 
fmr_2017_geo <- geo_format(fmr_2017)
fmr_2018_geo <- geo_format(fmr_2018)
fmr_2019_geo <- geo_format(fmr_2019)
fmr_2020_geo <- geo_format(fmr_2020)
fmr_2021_geo <- geo_format(fmr_2021)

# add metro area geographical data to all FMRs 
fmr_2017_geo2 <- geo_format2(fmr_2017)
fmr_2018_geo2 <- geo_format2(fmr_2018)
fmr_2019_geo2 <- geo_format2(fmr_2019)
fmr_2020_geo2 <- geo_format2(fmr_2020)
fmr_2021_geo2 <- geo_format2(fmr_2021)


```

```{r, eval=TRUE}
# geographic visualizations of housing costs
cpi_by_year <- read.csv("annual_cpi_us.csv", fileEncoding = "UTF-8-BOM")
cpi_by_year$adjust <- cpi_by_year$cpi/258.2937 # adjustment factor to 2020 dollars

# use the cpi_by_year adjustment to adjust rents to 2010 dollars
# rent in y year = (rent in current year) X (CPI y year/CPI current year)

map_rent <- function(df, year){
    correction = cpi_by_year$adjust[which(cpi_by_year$year == year)] # find correction factor
 
  fmr_map = ggplot() +
    geom_sf(data=df,lwd=0, 
            aes(fill = fmr1/correction)) + # adjust factor included
    ggtitle(paste(year, "Median Rent for 1-bed")) + 
    scale_fill_gradientn(name = "Rent (2020 $)", 
                         colours = myPalette2(100),
                         limits = c(0,2000),
                         breaks = c(500,1000,1500,2000)) + # add limits and breaks to normalize scaling between years  
    north(data = df, symbol = 12, anchor = c(x = -25, y = 39.93)) +
    scalebar(data = combined, dist = 500, dist_unit = 'mi', transform = TRUE,
             location = "bottomleft",
             st.dist = 0.1) +
    my_theme()
}

# call static map function and display
fmr1_2017_map <- map_rent(fmr_2017_geo, 2017) # generate static rent map
fmr1_2017_map # display rent map

# call static map function and display
fmr1_2017_map2 <- map_rent(fmr_2017_geo2, 2017) # generate static rent map
fmr1_2017_map2 # display rent map

```

```{r, eval=TRUE}
# an interactive map of rent prices

library(leaflet)
library(RColorBrewer)
library(htmlwidgets)
library(htmltools)

pal_fun <- colorNumeric("YlOrRd", NULL)  # Yellow-Orange-Red from RColorBrewer


# ------------------------------------------------- #
# HTML code to fix misaligned "na" in map legend
# ------------------------------------------------- #
library(htmlwidgets)
css_fix <- "div.info.legend.leaflet-control br {clear: both;}" # CSS to correct spacing
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML


# Funciton allows to plot interactive leaflet maps of median US rents by year
map_rent_i <- function(df, year){

#---------------------------------------------------#
# HTML code to add title to interactive maps
#---------------------------------------------------#
tag.map.title <- tags$style(HTML("
.leaflet-control.map-title { 
    transform: translate(-50%,20%);
    position: fixed !important;
    left: 50%;
    text-align: center;
    padding-left: 10px; 
    padding-right: 10px; 
    background: rgba(255,255,255,0.75);
    font-weight: bold;
    color: #000000;
    font-size: 20px;
  }
"))

title <- tags$div(
  tag.map.title, HTML(paste(year, "Median Rent:", 1, "Bed"))
)  
    
  correction <- cpi_by_year$adjust[which(cpi_by_year$year == year)] # find correction factor
  
  # Pop-up message
  pu_message <- paste0("County: ", df$countyname,  
                     "<br>Rent: ", "$", floor(df$fmr1/correction)) # in 2010 dollars
  
  # Adding more customization 
  interactive_map <- leaflet(df) %>%
    addPolygons(stroke = FALSE,                        # remove polygon borders
                fillColor = ~pal_fun(fmr1/correction), # in 2010 dollars
                fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
                popup = ~pu_message) %>%
    addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
    #addProviderTiles(providers$Stamen.Toner) %>%
    #addProviderTiles(providers$Esri.NatGeoWorldMap)
    addLegend("bottomright",                           # location of legend
              pal=pal_fun,                             # palette function
              values=~fmr1/correction,                  # variable to be passed to palette function
            title = paste("Rent",1,"bed (2020 $)"),  # legend title
            opacity = 1) %>% # legend opacity (1 = completely opaque)
    addScaleBar() %>%
    addTiles() %>%
    addControl(title, position = "topleft", className="map-title") #add title panel
  
  return(interactive_map %<>% htmlwidgets::prependContent(html_fix))
}

# Call interactive map function and display 
fmr_2020_mapi <- map_rent_i(fmr_2020_geo, 2020) 
fmr_2020_mapi


```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
```{r, eval=TRUE}

```
