---
title: "BMIN503/EPID600 Project Template"
author: "Jessica Meeker"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but keep the headers.

### Overview
In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

Maternal mortality and morbidity rates have seen a decline globally; however, in the United States these numbers continue to rise, in particular among women of color. I will use Electronic Medical Record (EMR) data from a cohort of women in the University of Pennsylvania Health System (UPHS). The goal of this project is to perform exploratory statistical methods to describe the exposure and outcome variables in the data set and machine learning methods (Lasso) to create a model of neighborhood level stress in Philadelphia that predicts maternal mortality and morbidity. 

I spoke with Drs. Trevor Penning, Ray Bai & Mary Regina Boland about my project. With Dr. Penning I discussed the project from a more broad lens of environmental exposures, both natural and social. Dr. Bai was able to provide much needed insight into the methodology of logistic regression with a lasso penalty on the predictors. Dr. Boland and I discussed overall project context and some of the considerations necessary for which machine learning method to choose. 

[Github Repository](https://github.com/jemeeker10/BMIN503_Final_Project)

### Introduction 
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.

In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

Maternal morbidity and mortality persist as key indicators of women’s health both globally and in the US. Pregnancy related deaths that occur during pregnancy or within a year following pregnancy, have strikingly doubled from 7.2 deaths per 100,000 live births in 1987 to 18.0 deaths in 2014. While mortality is steadily increasing nationally, severe maternal morbidity (SMM) is actually far more common in the US than maternal mortality by nearly 100 times. SMM describes unexpected, poor outcomes of labor or delivery that may result in short or long term consequences that can be significant for the mother, child and their family. However, not all women are at equal risk for SMM or maternal mortality and severe health disparities persist. Notably, African American women are upwards of four times as likely to die of complications from pregnancy as compared to white women. To date, no genetic variant has been found to explain this discrepancy, begging the question of what else is transpiring to propagate such a stark health disparity. In this project I am interested in understanding the role the social, built, and natural environments play on these maternal health disparities. 

This problem is very multidisciplinary in nature as it's success depends upon collaboration of those from a variety of fields including maternal health clinicians, epidemiologists, statisticians, informaticians, environmental toxicologists, and health disparities researchers. I was privileged to be able to meet with three people who were trained in or who have significant experience in all of these roles. This problem is complex as it is a environmental, maternal health, and social disparities problem. In meeting with the three people I consulted regarding this project, I was provided with conceptual and methodological advice that will aid in the success of this project. 


### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

The primary data sets used for this project include: 1) publically available data from the American Community Survey (ACS) from the Census Bureau 2) non-publically available data from the UPHS electronic medical record (EMR) system and 3) shape files from the Census Bureau. Multiple ACS files were used in the lasso prediction model, so as to let the algorithm decide which covariates were of most importance. Please note the socio-demographic variables that were used outlined in the methods code. The maternal health data which was derived from the UPHS EMR, is from 2010-2017 and includes pregnant women who were seen in the Penn System throughout their pregnancy. 

First, all necessary packages are loaded. 

```{r}
#All package libraries
library(dplyr)
library(tidycensus)
library(tidyverse)
library(sf)
library(ggplot2)
library(tidycensus)
library(tigris)
```

Next, we load in all of our predictor variables from the ACS and clean the data for analysis. 

```{r}
#ACS DATA
vars <- load_variables(dataset = "acs5",  # specify ACS 5-year estimates 
                       year = 2016)  # specify end year (i.e. this queries ACS 5-year estimates for 2011-2016)

head(vars)


acs.data <- get_acs(geography = "tract", year=2016,
                  state = "PA", county = "Philadelphia", geometry = TRUE,
                  variables = c(female_total = "B01001_026",
                                median.gross.rent = "B25064_001",
                                median.household.income = "B19013_001",
                                rent.burden = "B25071_001",
                                fem_nhwhite = "B01001H_017", 
                                fem_afam = "B01001B_017",
                                fem_hispanic = "B01001I_017",
                                fem_amind_alask = "B01001C_017",
                                fem_asian = "B01001D_017",
                                fem_nh.pi = "B01001E_017",
                                fem_multiple = "B01001G_017",
                                fem_other = "B01001F_017",
                                med.house.income = "B19013_001",
                                med.house.value = "B25077_001", 
                                poverty.estimate = "B17010_002",
                                poverty.total = "B17010_001",
                                fem.educ.total = "B15002_019", #these are for women over 25
                                fem.educ.none = "B15002_020",
                                fem.educ.ninth = "B15002_024",
                                fem.educ.hsgrad = "B15002_028",
                                fem.educ.coll = "B15002_032",
                                fem.educ.doc = "B15002_035",
                                fem.educ.masters = "B15002_033"))

acs.data.2 <- acs.data %>%
  group_by(GEOID) %>%  # return only one row per GEOID
  mutate(pov_fam = estimate[variable == "poverty.estimate"], 
         tot_fam = estimate[variable == "poverty.total"],  # extract and rename total number of families
         percent_poverty = (pov_fam/tot_fam)*100) %>%
  mutate(female_total = estimate[variable == "female_total"],
         fem.educ.total = estimate[variable == "fem.educ.total"],
         fem.educ.none = estimate[variable == "fem.educ.none"],
         fem.educ.ninth = estimate[variable == "fem.educ.ninth"],
         fem.educ.hsgrad = estimate[variable == "fem.educ.hsgrad"],
         fem.educ.coll = estimate[variable == "fem.educ.coll"],
         fem.educ.doc = estimate[variable == "fem.educ.doc"],
         fem.educ.masters = estimate[variable == "fem.educ.masters"],
         percent_hsgrad = (fem.educ.hsgrad/fem.educ.total)*100) %>%
  mutate(median.gross.rent = estimate[variable == "median.gross.rent"],
         median.household.income = estimate[variable == "median.household.income"],
         rent.burden = estimate[variable == "rent.burden"]) %>%
  mutate(fem_nhwhite = estimate[variable == "fem_nhwhite"],
         fem_afam = estimate[variable == "fem_afam"],
         fem_hispanic = estimate[variable == "fem_hispanic"],
         fem_amind_alask = estimate[variable == "fem_amind_alask"],
         fem_asian = estimate[variable == "fem_asian"],
         fem_nh.pi = estimate[variable == "fem_nh.pi"],
         fem_multiple = estimate[variable == "fem_multiple"],
         fem_other = estimate[variable == "fem_other"],
         percent_white = ((fem_nhwhite/(fem_nhwhite + fem_afam + fem_hispanic + fem_amind_alask + fem_asian + fem_nh.pi + fem_multiple + fem_other))*100)) %>%
  mutate(percent_minority = (100-percent_white))%>%
  select(GEOID, NAME, pov_fam, tot_fam, percent_poverty, female_total, fem.educ.total, fem.educ.none, fem.educ.ninth, fem.educ.hsgrad, fem.educ.coll, fem.educ.doc, fem.educ.masters, percent_hsgrad, median.gross.rent, median.household.income, rent.burden, fem_nhwhite, fem_afam, fem_hispanic, fem_amind_alask, fem_asian, fem_nh.pi, fem_multiple, fem_other, percent_white, percent_minority) 

head(acs.data)

```

Next we needed to pull in the patient level maternal health data from Penn. Since this is PHI data, I am ommitting the initial code used to import the data from SQL. 

```{r}
#However, first I imported the data, then I merged a dataset of women with encounters at UPHS with a dataset that had outcome data for their deliveries. These variables include information about stillbirth, preterm birth, preeclampsia, hypertnesion and C-section. 

#these libraries were used in the code I used to tunnel in to the data, which has been ommitted here. 
##library(RMySQL)
##library("DBI")
##library(stringr)

#Merging the data sets together
dim(encounter_at_deliv)
dim(mat_outcomes)
encounter_at_deliv_woutcomes = merge(encounter_at_deliv, mat_outcomes, by="PREG_ID") 
dim(encounter_at_deliv_woutcomes)

#get rid of NAs
encounter_at_deliv_woutcomes_NAlatlonrem=encounter_at_deliv_woutcomes[-which(encounter_at_deliv_woutcomes$LATITUDE=="NA"), ]
encounter_at_deliv_woutcomes_NAlatlonrem=encounter_at_deliv_woutcomes_NAlatlonrem[-which(encounter_at_deliv_woutcomes_NAlatlonrem$LONGITUDE=="NA"), ]
dim(encounter_at_deliv_woutcomes_NAlatlonrem)

#get rid of blanks
encounter_at_deliv_woutcomes_NAlatlonrem_blanks=encounter_at_deliv_woutcomes_NAlatlonrem[-which(encounter_at_deliv_woutcomes_NAlatlonrem$LATITUDE==""), ]
encounter_at_deliv_woutcomes_NAlatlonrem_blanks=encounter_at_deliv_woutcomes_NAlatlonrem[-which(encounter_at_deliv_woutcomes_NAlatlonrem$LONGITUDE==""), ]
dim(encounter_at_deliv_woutcomes_NAlatlonrem_blanks)

outcome.data <- data.frame(encounter_at_deliv_woutcomes_NAlatlonrem_blanks)
class(outcome.data)


#Choose only the variables we want. 
outcome.data_clean <- select(outcome.data,PREG_ID,LATITUDE,LONGITUDE,MRN_HUP,YEAR,NH_AFAM,NH_WHITE,HISPANIC,WEIGHT_LBS,ENC_AGE_DEC.x,STILLBIRTH,PRETERM,CSECTION, HYPERTENSION, PREECLAMPSIA)
dim(outcome.data_clean)

#Take the patient data and append GEOIDs to each patient's address so that we can link it with ACS data
# (!) fill this: 
data_path <- "C:/Users/Jessi/Documents/Blanca Class/finally project/"
load(paste0(data_path,"phillytracts")) 

# (!) let's call the dataset data.pre
data.pre <- outcome.data_clean
data.pre$LATITUDE[which(data.pre$LATITUDE=="")]=NA
data.pre$LONGITUDE[which(data.pre$LONGITUDE=="")]=NA

# GPS location:
temp.pointX <- data.pre$LATITUDE
temp.pointY <- data.pre$LONGITUDE

## actually they have the same NAs, but just in case
rem.ind.points <- is.na(temp.pointX) | is.na(temp.pointY)
## want to make this exact same as old file. Remove extraneous points, keep order:
# (!) add other covariates we care about
data.new <- data.pre[!rem.ind.points,] 
data.new$LATITUDE=as.numeric(as.character(data.new$LATITUDE))
data.new$LONGITUDE=as.numeric(as.character(data.new$LONGITUDE))
pts <- SpatialPoints(data.new[ ,c('LONGITUDE', 'LATITUDE')])
databytracts <- sp::over(tracts, pts, returnList = T)

data.new$tract <- NA
data.new$tractID <- NA

datavector <- unlist(databytracts)
datatractlength <- lapply(databytracts, length)

levels_region <- levels(tracts@data$GEOID10)[tracts@data$GEOID10]

dataind.temp <- c(0, cumsum(datatractlength))
for(i in 1:length(databytracts)){
  temp.index <- (dataind.temp[i] + 1): (dataind.temp[i + 1])
  data.new$tract[datavector[temp.index]] = i
  data.new$tractID[datavector[temp.index]] = levels_region[i]
  if(i %% 100 == 0) cat(i, " ")
}
sum(is.na(data.new$tractID))
dim(data.new)

data.new.philly=data.new[-which(is.na(data.new$tractID)), ]
#View(data.new.philly)
dim(data.new.philly)
table(data.new.philly$tract)
num.tracts = length(unique(data.new.philly$tract))
unique_tracts = unique(data.new.philly$tract)

num_deliv=rep(NA,num.tracts)
num_preterm=rep(NA,num.tracts)
num_stillbirth=rep(NA,num.tracts)
num_Csection=rep(NA,num.tracts)
for(i in 1:length(unique(data.new.philly$tract))){
  data_it = subset(data.new.philly, data.new.philly$tract==unique_tracts[i])
  num_deliv[i] = length(unique(data_it$PREG_ID))
  num_preterm[i] = length(unique(data_it$PREG_ID[which(data_it$PRETERM=="1")]))
  num_stillbirth[i] = length(unique(data_it$PREG_ID[which(data_it$STILLBIRTH=="1")]))
  num_Csection[i] = length(unique(data_it$PREG_ID[which(data_it$CSECTION=="1")]))
}
counts_outcomes_philly_geo = data.frame(tract=unique_tracts,
                                        num_deliv=num_deliv,
                                        num_Csection=num_Csection,
                                        num_preterm=num_preterm,
                                        num_stillbirth=num_stillbirth)


tract_df = data.frame(tract=data.new.philly$tract,
                      tractID=data.new.philly$tractID)

tract_df=tract_df[!(duplicated(tract_df)), ]

counts_outcomes_philly_geo=merge(counts_outcomes_philly_geo,tract_df, by="tract" )

dim(counts_outcomes_philly_geo)

dim(data.new.philly)


```
Now, we need to merge data.new.philly (the geocoded maternal health outcome data) with acs.data.2.
```{r}
class(data.new.philly$tractID)
class(acs.data.2$GEOID)

#first need to make GEOID column in data.new.philly to merge on
data.new.philly$GEOID=data.new.philly$tractID
class(data.new.philly$GEOID)

outcome_acs = merge(acs.data.2, data.new.philly, by="GEOID") 
dim(outcome_acs)
outcome_acs_clean = outcome_acs[!(duplicated(outcome_acs)), ]
dim(outcome_acs_clean)

class(outcome_acs)
colnames(outcome_acs)
```
### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

Bar Plots were generated to look at overall outcome counts

```{r}
#bar plot for c-section
ggplot(data = outcome.data_clean, aes(x = CSECTION, group = YEAR,)) +
  geom_bar(stat = "count", fill = "blue") +
  ggtitle ("Number of c-sections per pregnancies") +
  theme_bw() +
  facet_grid(YEAR ~ .)

table(outcome.data_clean$CSECTION)

#bar plot for stillbirth
ggplot(data = outcome.data_clean, aes(x = STILLBIRTH, group = YEAR)) +
  geom_bar(stat = "count", fill = "black") +
  ggtitle ("Number of stillbirths per pregnancies") +
  theme_bw() +
  facet_grid(YEAR ~ .)

ggplot(data = outcome.data_clean, aes(x = STILLBIRTH, group = YEAR)) +
  geom_bar(stat = "count", fill = "pink") +
  ggtitle ("Number of stillbirths per pregnancies") +
  theme_bw()

table(outcome.data_clean$STILLBIRTH)

#bar plot for pre-term birth by year
ggplot(data = outcome.data_clean, aes(x = PRETERM, group = YEAR)) +
  geom_bar(stat = "count", fill = "purple") +
  ggtitle ("Number of pre-term births per pregnancies") +
  theme_bw() +
  facet_grid(YEAR ~ .)
#overall
ggplot(data = outcome.data_clean, aes(x = PRETERM, group = YEAR)) +
  geom_bar(stat = "count", fill = "purple") +
  ggtitle ("Number of pre-term births per pregnancies") +
  theme_bw() 

#bar plot for pre-eeclampsia
ggplot(data = outcome.data_clean, aes(x = PREECLAMPSIA, group = YEAR)) +
  geom_bar(stat = "count", fill = "dark green") +
  ggtitle ("Number of preeclampsia diagnoses per pregnancies") +
  theme_bw() +
  facet_grid(YEAR ~.)

#bar plot for hypertension
ggplot(data = outcome.data_clean, aes(x = HYPERTENSION, group = YEAR)) +
  geom_bar(stat = "count", fill = "red") +
  ggtitle ("Number of hypertension diagnoses per total pregnancies") +
  theme_bw() +
  facet_grid(YEAR ~.)

#make race category for graph
outcome.data_clean <- outcome.data_clean %>%
  mutate(RACE = ifelse(NH_AFAM == 1, "black",
                  ifelse(NH_WHITE == 1, "white",
                  ifelse(HISPANIC ==1, "hispanic", "no answer"))))

table(outcome.data_clean$RACE)

#bar plot for race 
ggplot (data = outcome.data_clean, aes (x = RACE)) +
  geom_bar(stat = "count") +
  ggtitle("Total race breakdown per pregnancies") 



```
Let's look at a plot that shows all of the outcome variables together as percents.
```{r}
#Looking at A bar plot of all the variables together
#9592/19682 = 49%
outcome.data_clean%>%
  filter(CSECTION==1)%>%
  count()
outcome.data_clean%>%
  filter(CSECTION==0)%>%
  count()
#1760/27515 = 6.4%
outcome.data_clean%>%
  filter(PRETERM==1)%>%
  count()
outcome.data_clean%>%
  filter(PRETERM==0)%>%
  count()
#18%
outcome.data_clean%>%
  filter(HYPERTENSION==1)%>%
  count()
outcome.data_clean%>%
  filter(HYPERTENSION==0)%>%
  count()
#8.1%
outcome.data_clean%>%
  filter(PREECLAMPSIA==1)%>%
  count()
outcome.data_clean%>%
  filter(PREECLAMPSIA==0)%>%
  count()
#1%
outcome.data_clean%>%
  filter(STILLBIRTH==1)%>%
  count()
outcome.data_clean%>%
  filter(STILLBIRTH==0)%>%
  count()

9592/19682
percents <- data.frame("Variable" = c("CSECTION", "PRETERM", "HYPERTENSION", "PREECLAMPSIA", "STILLBIRTH"), "Percent" = c(49,7,18,8,1))

ggplot (data = percents, aes (x = Variable, y=Percent)) +
  geom_bar(stat="identity") +
  ggtitle("Percents of morbidities per total number of pregnancies") 


```

Now let's take a look at the pattern of how certain outcomes map together.
```{r}
# to look at the women who have different combinations of outcomes
outcome.data_clean %>%
  group_by(PRETERM, CSECTION, PREECLAMPSIA, HYPERTENSION, STILLBIRTH) %>%
  summarise(n = n())
```


Download shape file for Philadelphia for mapping purposes

```{r}
#import city shapefile as an sf object

#Pull in Phila cencus tracts shape file 

philly.tracts <- readRDS(gzcon(url("https://raw.githubusercontent.com/HimesGroup/BMIN503/master/DataFiles/philadelphia_tracts_2010.rds")))

# Note philly.tracts is an sf ("simple feature") object of type "MULTIPOLYGON"
head(philly.tracts) 
class(philly.tracts)  

# Once again, we can extract the geometry of philly.tracts with the st_geometry function
pt.geo <- st_geometry(philly.tracts)
pt.geo

# Plot the geometry of philly.tracts with the base plot function
plot(pt.geo)

# The base plot function has some aesthetic options we can use to tweak our plots
plot(pt.geo, col = "lemonchiffon2")
plot(pt.geo, lwd = 2, border = "red")


str(philly.tracts)


#join shapfile to ACS data 
philly.tracts <- mutate(philly.tracts, GEOID = as.character(GEOID10))
class(acs.data)
  
# Now we can perform our join
acs.data.2 <- data.frame(acs.data.2)
acs_philly_sf <- inner_join(philly.tracts, acs.data.2, by = "GEOID")
head(acs_philly_sf)  # success!
class(acs_philly_sf)

#do this when can add GEOID to patient data
alldata <- inner_join(philly.tracts, acs.data.2, by = "GEOID")
```



We will now utilize the machine learning technique of lasso to see which neighborhood variables best predict c-section and preeclampsia. 

```{r}
#Insert lasso code here to run logistic regression with lasso penalty on predictors

##CSECTION OUTCOME 
###make a matrix of all predictors
data<-as.data.frame(outcome_acs_clean)
data<-na.omit(data)
str(data)

#make CSECTION a factor 
data$CSECTION <- factor(data$CSECTION)
class(data$CSECTION)
x.cont<-data %>% select(percent_poverty, percent_hsgrad, median.gross.rent, median.household.income, rent.burden, fem_nhwhite, fem_afam, fem_hispanic, fem_other, fem_amind_alask, fem_asian, fem_nh.pi, fem_multiple, percent_white, percent_minority, ENC_AGE_DEC.x)

x.cat <- data %>% select(CSECTION, NH_AFAM, NH_WHITE, HISPANIC)
x.factors<-model.matrix(CSECTION~NH_AFAM + NH_WHITE + HISPANIC, x.cat)[,-1]

x<-as.matrix(cbind(x.cont,x.factors))

library(glmnet)
set.seed(1234)
csection.lasso <- cv.glmnet(x, y = data$CSECTION, alpha = 1, family = "binomial") #alpha = 1 (coefficient used to penalize variables you're using. binomial (log reg))
csection.lasso$lambda.min
plot(csection.lasso)
lasso.coef <- coef(csection.lasso, s = "lambda.min")
lasso.coef #some are getting higher magnitude weights and some lower (percent poverty is most important)
csection.lasso.pred <- predict(csection.lasso, x, s = "lambda.min", type = "response")
head(csection.lasso.pred) #probabilitiy you had a csection (1) for each person

##

```
We will do some model evaluation.

```{r}
class(data$CSECTION)
class(data)

num_deliv_per_tract=table(data$GEOID)
tracts_to_keep = num_deliv_per_tract[which(num_deliv_per_tract>10)] 
tracts_to_keep=names(tracts_to_keep)
data_all=data
data=data[which(data$GEOID %in% tracts_to_keep), ]

#K-Fold Cross Validation
N = nrow(data)
K = 10
set.seed(1234)
s = sample(1:K, size = N, replace = T)
pred.outputs.lasso <- vector(mode = "numeric", length = N)
obs.outputs <- vector(mode = "numeric", length = N)
offset <- 0


for(i in 1:K){
	train <- filter(data, s != i)
	test <- filter(data, s == i)
	obs.outputs[1:length(s[s == i]) + offset] <- test$CSECTION
	colnames(train)
#lasso train/test
lasso.m <- cv.glmnet(as.matrix(train[, 3:27]), y = train$CSECTION, alpha = 1, family = "binomial")
lasso.pred.curr <- predict(lasso.m, as.matrix(test[, 3:27]), s = "lambda.min", type = "response")
pred.outputs.lasso[1:length(s[s == i]) + offset] <- lasso.pred.curr[ , 1]
	
offset <- offset + length(s[s == i])
}

```
Let's take a look at the ROC Curves

```{r}
library('pROC')

roc(obs.outputs, pred.outputs.lasso, ci = TRUE)
roc(data_all$CSECTION, csection.lasso.pred)
plot.roc(data_all$CSECTION, csection.lasso.pred[ ,1])
plot.roc(obs.outputs, pred.outputs.lasso, ci = TRUE, col = "red", add=TRUE)
legend("bottomright", legend = c("All Data", "Cross-Validation"), col = c("black", "red"), lwd = 1)

```

We will make some interesting maps showing the spatial association between neighborhood stressors and maternal morbidity outcomes. 

```{r}
#ggplot code
library(sf)

ggplot() +
  geom_sf(data = philly.tracts) 

#percent poverty in phila
ggplot() +
  geom_sf(data = acs_philly_sf, aes(fill = percent_poverty ))

#percent hs educated women in phila
ggplot() +
  geom_sf(data = acs_philly_sf, aes(fill = percent_hsgrad ))

#percent minority women in phila
ggplot() +
  geom_sf(data = acs_philly_sf, aes(fill = percent_minority ))

#rent burden in phila
ggplot() +
  geom_sf(data = acs_philly_sf, aes(fill = rent.burden ))


#insert leaflet code 
library(leaflet)

# Select a color palette with which to run the palette function
pal_fun <- colorNumeric("BuPu", NULL)       # Blue-Purple from RColorBrewer
pal_fun2 <- colorNumeric("YlOrRd", NULL)    # Yellow-Orange-Red from RColorBrewer
pal_fun3 <- colorNumeric("viridis", NULL)   # viridis from viridis
pal_fun4 <- colorNumeric("inferno", NULL)   # inferno from viridis
pal_fun5 <- colorNumeric("inferno", NULL, reverse=TRUE)  # reverses the color ramp

# Pop-up message
pu_message <- paste0(philly.tracts2$NAMELSAD10,  # paste0 to append tract name with other relevant text 
                     "<br>Poverty rate: ",       # <br> forces new line
                     # use round function to round continuous poverty rate to one decimal point
                     round(philly.tracts2$percent_poverty, 1), "%")

# Basic leaflet map
leaflet(philly.tracts2) %>%
  addPolygons(fillColor = ~pal_fun(percent_poverty), # addPolygons requires variable and a pass to a palette function  
              popup = pu_message) %>%                # add a popup message
  addTiles()                                         # default basemap

# Adding more customization 
leaflet(philly.tracts2) %>%
  addPolygons(stroke = FALSE,                        # remove polygon borders
              fillColor = ~pal_fun(percent_poverty),
              fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
  #addProviderTiles(providers$Stamen.Toner) %>%
  #addProviderTiles(providers$Esri.NatGeoWorldMap)
  addLegend("bottomright",                           # location of legend
            pal=pal_fun,                             # palette function
            values=~percent_poverty,                 # variable to be passed to palette function
            title = 'Poverty rate',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```

### Conclusion and Future Directions
In conclusion, it looks like percent poverty might be an important predictor for C-section. However, given that our AUC was not very high, in the future we'd like to add additional neighborhood predictors to see if we can create a better model. Furthermore, we'd like to do this for the other outcomes we studied and to create a composite prediction index for maternal morbidity in Philadelphia.  

