---
title: "BMIN503/EPID600 Project: Effect of COVID-19 pandemic on Lung transplantation in US"
author: "Arnab Chowdhury"
output: 
  html_document:
    theme: paper 
    highlight: tango
---

------------------------------------------------------------------------

## 1 Overview

This project aims to study the impact of the COVID-19 pandemic on lung transplantation in the United States utilizing data from OPTN (Oran Procurement and Transplant Network). We aim to answer if lung transplants decreased during the pandemic and if the decrease corelated to publically available data on COVID-19 metrics that include total COVID-19 cases, new COVID-19 cases, number of patients hospitalized for COVID-19 and number of patients with COVID-19 admitted to the ICU. 

## 2 Introduction

The SARS-CoV-2 (COVID-19) pandemic has significantly impacted solid organ transplant programs across the United States. At the height of the pandemic, transplant centers adjusted guidelines for donor selection, management of patients with a positive COVID-19 diagnosis, and prevention of transmission of COVID-19 to transplant recipients and healthcare workers. The COVID-19 pandemic also gave rise to a new subset of patients being referred for lung transplantation as well. The aim of this project is to understand the impact of the COVID-19 pandemic on the lung transplantation program using publically available UNOS OPTN data and local COVID-19 incidence rates.

Lung transplantation requires coordination between multiple disciplines including transplant pulmonology, cardiothoracic surgery, critical care medicine, nursing, respiratory therapy and organ procurement teams, among many others that closely work together to make transplants a success. Data from the effects of COVID-19 pandemic and understanding its impact will help us better prepare for future pandemics.

## 3 Methods

### 3.1 Data

Data was obtained from the OPTN (Organ Procurement and Transplant Network) after completing a data request. The data obtained involved STAR files that contain de-identified individual transplant records. These include all patients in the United States who have ever been placed on a transplant waiting list. The original files were stored in a password protected computer. The original files were in SAS format and the "thoracic_data" file that contained records regarding heart and lung transplant were read using the package "haven".

```{r eval = TRUE}
# Reading SAS files provided by UNOS OPTN. Reading thoracic data. 
library(haven)
thoracic_data <- read_sas("/Users/work/Downloads/SAS Dataset 202209/Thoracic/thoracic_data.sas7bdat")
```

### 3.2 General Methodology

Necessary packages were loaded. Patients waitlisted for lung transplantation after 2015 were selected. The top rows of the dataframe and the dimensions were evaluated. The original table has 206391 records (between 2015 and 2022) and 510 columns. Subsequently, the relevant characteristics (eg age, race, gender, waitlist date, waitlist time, transplant status etc) of this population was also selected. The first few rows of this new table is displayed.

```{r message = FALSE, warning = FALSE}
# Cleaning data to filter transplanted patients and lung transplants and filter by year
library(gtsummary)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(table1)
library(cowplot) 
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(usmap)

dim(thoracic_data)

# Filter patients from 2015 to 2021 who were "activated" for lung transplant
thoracic_data_subset_lung_activation <- thoracic_data %>% 
  dplyr::select(WL_ID_CODE, TXED, WL_ORG, DAYSWAIT_CHRON, ACTIVATE_DATE, TX_DATE, GENDER, PERM_STATE, TXLNG, TX_YEAR, INIT_AGE, TCR_DGN, ETHCAT) %>%
  dplyr::mutate(TXED = (factor(TXED, levels = c(0, 1), labels = c("no", "yes"))), GENDER = factor(GENDER), PERM_STATE = factor(PERM_STATE), TCR_DGN = (factor(TCR_DGN)), ETHCAT = (factor(ETHCAT, levels = c(1, 2, 4, 5, 6, 7, 9, 998), labels =c("White", "Black", "Hispanic", "Asian", "American Indian", "Native Hawaiian", "Multiracial", "Unknown")))) %>% mutate(PERM_STATE=recode(PERM_STATE, "ZZ" = "Unknown")) %>% 
  dplyr::filter(ACTIVATE_DATE >= "2015-01-01" & ACTIVATE_DATE < "2022-01-01") %>%
  dplyr::filter(WL_ORG == "LU")

head(thoracic_data_subset_lung_activation)

```

```{r message = FALSE, warning = FALSE}
# Filter patients from 2015 to 2021 who received a lung transplant
thoracic_data_subset_lung <- thoracic_data %>% 
  dplyr::select(WL_ID_CODE, TXED, WL_ORG, DAYSWAIT_CHRON, ACTIVATE_DATE, TX_DATE, GENDER, PERM_STATE, TXLNG, TX_YEAR, DIAG, INIT_AGE, ETHCAT, TCR_DGN) %>% dplyr::mutate(TXED = (factor(TXED, levels = c(0, 1), labels = c("no", "yes"))), GENDER = factor(GENDER), PERM_STATE = factor(PERM_STATE), TCR_DGN = (factor(TCR_DGN)), ETHCAT = (factor(ETHCAT, levels = c(1, 2, 4, 5, 6, 7, 9, 998), labels =c("White", "Black", "Hispanic", "Asian", "American Indian", "Native Hawaiian", "Multiracial", "Unknown")))) %>% dplyr::filter(TX_DATE >= "2015-01-01" & TX_DATE < "2022-01-01") %>% dplyr::filter(WL_ORG == "LU")
head(thoracic_data_subset_lung)
```

Seperate cohorts of patients who were placed on the waiting list and received lung transplant for COVID-19 related lung disease were also created below. This was done by filtering relevant diagnosis codes (1616 and 1617). 

```{r message = FALSE, warning = FALSE}
covid_tx_waitlist <- thoracic_data_subset_lung_activation %>% dplyr::filter(TCR_DGN == "1616" | TCR_DGN == "1617") 

howmanycovid_tx <- covid_tx_waitlist %>% group_by(TXED) %>% count(TXED)

covid_tx <- covid_tx_waitlist %>% filter(TXED == "yes")
```


Now it was important to select an appropriate interval to further analyze the data. One thing to note is that there is inherent baseline variability in the data. We will visualize the total number of lung transplants and plot the number of lung transplants that occurred daily, weekly and monthly. We cast a wide net and started the visualization from 2015 to see the trend but will narrow down the data further in subsequent codes.

In the graphs below we see that the inherent variability is very pronounced when plotting daily data, decreases slightly with weekly data and is the least with monthly data. Hence, we will utilize monthly data for further analysis of the data.

Visualizing the monthly data qualitatively, there was a steady increase in lung transplants starting in 2016 until 2020 and then transplants may have taken a dip in March 2020, further dips are seen late 2020/early 2021 and possibly another in the last quarter of 2021.

```{r message = FALSE, warning = FALSE}

# Daily lung transplants
lungtx2015_2021 <- thoracic_data_subset_lung %>%
  dplyr::filter(TX_DATE >= "2015-01-01" & TX_DATE <= "2021-12-31") %>% count(TX_DATE) %>% rename("Number.Transplants" = n) 
daily <- ggplot(lungtx2015_2021, aes(x = TX_DATE, y = Number.Transplants)) +
  geom_point(size=0.5) + labs(x = "Date") +
  labs(y = "Lung Transplants") + 
  labs(title = "Daily Lung Transplants Performed (2015-2021)")

# Weekly
weekly <- ggplot(thoracic_data_subset_lung, aes(TX_DATE, fill = TXED)) + 
  geom_histogram(binwidth = 7) + 
  labs(title = "Weekly Lung Transplants Performed (2015-2021)") +
  labs(x = "Year") +
  labs(y = "Lung Transplants")

# Monthly
monthly <- ggplot(thoracic_data_subset_lung, aes(TX_DATE, fill = TXED)) + 
  geom_histogram(binwidth = 30) + 
  labs(title = "Monthly Lung Transplants Performed (2015-2021)") +
  labs(x = "Year") +
  labs(y = "Lung Transplants")

library("patchwork")
daily / weekly / monthly

```

The data was narrowed further to only include transplants from 2018 to 2021. The following data will used to analyze against the COVID-19 metrics.

```{r message = FALSE, warning = FALSE}
# Total number of transplants from 2018 to 2021
lungtx2018_2021 <- thoracic_data_subset_lung %>%
  dplyr::filter(TX_DATE >= "2018-01-01" & TX_DATE <= "2021-12-31") %>% count(TX_DATE) %>% rename("Number.Transplants" = n) 
```

```{r message = FALSE, warning = FALSE}
library(dplyr)
library(lubridate)

# Using monthly data rather than daily data
lungtx2018_2021.ym <- lungtx2018_2021 %>% mutate(
    TX_DATE = ymd(TX_DATE),
    TX_YEAR_MONTH = format_ISO8601(TX_DATE, precision = "ym")
  )
lungtx2018_2021.ym <- lungtx2018_2021.ym %>% group_by(TX_YEAR_MONTH) %>% 
  summarise(total_transplants=sum(Number.Transplants))

lungtx2018_2021.ym$TX_YEAR_MONTH <- as.Date(paste(lungtx2018_2021.ym$TX_YEAR_MONTH, "-01", sep=""))

# Rough plot of lung transplants from 2018-2021. Monthly transplants
library(ggplot2)
ggplot(data = lungtx2018_2021.ym, aes(x = TX_YEAR_MONTH, y = total_transplants)) +
  geom_point() +
  geom_line() +
  geom_smooth() +
  labs(title = "Monthly transplants",
       subtitle = "2018-2021",
       x = "Date", y = "Monthly Transplants")
```

The COVID-19 data was then obtained from the [OWID (Our World in Data) database](https://ourworldindata.org/coronavirus-source-data). This data was then plotted (See figures below). To perform a join with the monthly transplant data, the COVID-19 data was also converted to monthly data rather than daily.

```{r message = FALSE, warning = FALSE}
#   Obtain monthly COVID-19 new cases in US
owid_url <- "https://github.com/owid/covid-19-data/blob/master/public/data/owid-covid-data.csv?raw=true"
country <- "United States"
covid <- read_csv(owid_url)
covid_cases <- covid %>% 
  filter(location == country) %>% 
  select(date, total_cases, new_cases, hosp_patients, icu_patients) %>% 
  arrange(date) %>% mutate(total_cases = as.integer(total_cases), new_cases = as.integer(new_cases), hosp_patients = as.integer(hosp_patients), icu_patients = as.integer(icu_patients))

covid.cases <- ggplot(data=covid_cases, aes(x=date, y=new_cases)) + 
  geom_point(size=0.005) + labs(title = "COVID-19 data", subtitle = "New cases", x = "Date", y = "New cases")
covid.hosp <- ggplot(data=covid_cases, aes(x=date, y=hosp_patients)) + 
  geom_point(size=0.005) + labs(title = "COVID-19 data", subtitle = "Total hospitalized", x = "Date", y = "Hospitalized")
covid.icu <- ggplot(data=covid_cases, aes(x=date, y=icu_patients)) + 
  geom_point(size=0.005) + labs(title = "COVID-19 data", subtitle = "ICU cases", x = "Date", y = "ICU cases")

covid.cases / covid.hosp / covid.icu

covid_cases.ym <- covid_cases %>% mutate(
    date = ymd(date),
    TX_YEAR_MONTH = format_ISO8601(date, precision = "ym")
  )
covid_cases.ym <- covid_cases.ym %>% group_by(TX_YEAR_MONTH) %>% 
  summarise(total_cases2=sum(total_cases), total_new_cases=sum(new_cases), total_hosp=sum(hosp_patients), total_icu=sum(icu_patients)) 

covid_cases.ym$TX_YEAR_MONTH <- as.Date(paste(covid_cases.ym$TX_YEAR_MONTH, "-01", sep=""))

lungtx_covid_join <- left_join(lungtx2018_2021.ym, covid_cases.ym) %>% filter (TX_YEAR_MONTH >= "2020-02-01")

```

## Results

### 4.1 Waitlist characteristics.

Baseline characteristics of patients on the waitlist were then compiled into table 1. Out of 20590 patients placed on the waitlist about 16704 were transplanted. More men are placed on the waiting list compared to women. Among patients not transplanted, there are more women than men. The transplanted group also is slightly older than the group not transplanted.

```{r message = FALSE, warning = FALSE}
library(table1)

thoracic_data_subset_lung_activation$TXED <- 
  factor(thoracic_data_subset_lung_activation$TXED, 
         levels=c("no","yes"),
         labels=c("Not transplanted", "Transplanted"))

label(thoracic_data_subset_lung_activation$TXED) <- "Received Lung Transplant"
label(thoracic_data_subset_lung_activation$DAYSWAIT_CHRON) <- "Days on the wait list"
label(thoracic_data_subset_lung_activation$GENDER) <- "Gender"
label(thoracic_data_subset_lung_activation$ETHCAT) <- "Race"
label(thoracic_data_subset_lung_activation$INIT_AGE) <- "Recipient Age"
label(thoracic_data_subset_lung_activation$PERM_STATE) <- "Recipient State of Residence"
table1(~ DAYSWAIT_CHRON + GENDER + INIT_AGE + ETHCAT + PERM_STATE| TXED, data=thoracic_data_subset_lung_activation)


```

I was also curious how the number of lung transplants and proportion of patients on the waitlist varied by state. Are there states where a higher percentage of patients on the waitlist received a lung transplant? Did it correlate to the number of transplants in a particular state? A geospatial map of the proportion of patients transplanted by state is also depicted below. Observing crudely, there appears to be an inverse relation between number of lung transplants performed in a state and proportion of patients transplanted. In other words, the higher the transplants done in a state, the lower the proportion of patients transplanted. One must be careful interpreting this however as the data is not adjusted for Lung Allocation Score (LAS) and other factors (such as health care resources as a proportion to population) that may affect the data.

```{r message = FALSE, warning = FALSE}

lungtxstate <- thoracic_data %>% 
  dplyr::select(TXED, WL_ORG, DAYSWAIT_CHRON, ACTIVATE_DATE, TX_DATE, PERM_STATE, TX_YEAR, INIT_AGE) %>%
  dplyr::mutate(TXED = (factor(TXED, levels = c(0, 1), labels = c("no", "yes"))), PERM_STATE = factor(PERM_STATE)) %>%
  dplyr::filter(ACTIVATE_DATE >= "2015-01-01" & ACTIVATE_DATE < "2022-01-01") %>%
  dplyr::filter(WL_ORG == "LU") %>% group_by(PERM_STATE, TXED) %>% tally()

lungtxstate.yes <- lungtxstate %>% filter(TXED == "yes") %>% mutate(nyes = n) %>% select(PERM_STATE, nyes)
lungtxstate.no <- lungtxstate %>% filter(TXED == "no") %>% mutate(nno = n) %>% select(PERM_STATE, nno)

lungtxstate_join <- left_join(lungtxstate.yes, lungtxstate.no) %>% select(PERM_STATE, nyes, nno) %>% mutate(proptxed = (nyes/(nno + nyes))*100) %>% mutate(state = PERM_STATE) %>% arrange(desc(proptxed))


percenttx <- plot_usmap(
  data = lungtxstate_join, values = "proptxed", lines = "red"
) + 
  scale_fill_continuous(
    low = "white", high = "red", name = "Proportion Transplanted (%)", label = scales::comma
  ) + 
  labs(title = "US States, Proportion of patients transplanted") +
  theme(legend.position = "right")

numbertx <- plot_usmap(
  data = lungtxstate_join, values = "nyes", lines = "blue"
) + 
  scale_fill_continuous(
    low = "white", high = "purple", name = "Number of Lung Transplants", label = scales::comma
  ) + 
  labs(title = "US States, Number of Lung Transplants") +
  theme(legend.position = "right")

numbertx / percenttx

```

### 4.2 Characteristics of patients who received lung transplants.

Now, we focus our attention to the patients who ended up receiving a lung transplant. A new subset of patients will be created below to select this population of patients who received a lung transplant between 01/01/2015 to 12/31/2021. A separate demographics table was made to assess characteristics of the patients who received lung transplants.

```{r message = FALSE, warning = FALSE}
label(thoracic_data_subset_lung$TXED) <- "Received Lung Transplant"
label(thoracic_data_subset_lung$DAYSWAIT_CHRON) <- "Days on the wait list"
label(thoracic_data_subset_lung$GENDER) <- "Gender"
label(thoracic_data_subset_lung$ETHCAT) <- "Race"
label(thoracic_data_subset_lung$INIT_AGE) <- "Recipient Age"
label(thoracic_data_subset_lung$PERM_STATE) <- "Recipient State of Residence"

table1(~ DAYSWAIT_CHRON + GENDER + INIT_AGE + ETHCAT + PERM_STATE, data=thoracic_data_subset_lung)
```

#### 4.2.1 Transplants due to COVID-19 lung disease

We also looked at patients who were waitlisted for lung transplant due to COVID-19. Their baseline characteristics are in the table below. 
```{r message = FALSE, warning = FALSE}
label(covid_tx_waitlist$TXED) <- "Received Lung Transplant"
label(covid_tx_waitlist$DAYSWAIT_CHRON) <- "Days on the wait list"
label(covid_tx_waitlist$GENDER) <- "Gender"
label(covid_tx_waitlist$ETHCAT) <- "Race"
label(covid_tx_waitlist$INIT_AGE) <- "Recipient Age"
label(covid_tx_waitlist$PERM_STATE) <- "Recipient State of Residence"
table1(~ DAYSWAIT_CHRON + GENDER + INIT_AGE + ETHCAT | TXED, data=covid_tx_waitlist)
```

We see that there were 359 patients placed on the lung transplant list due to COVID-19 related lung disease. The mean age of patients transplanted was 49.1 and more men, more white patients both were placed on the list and received a transplant. We further breakdown the time of transplants by year below. 

```{r message = FALSE, warning = FALSE}
ggplot(data = covid_tx, aes(x = TX_YEAR, fill = TXED)) + geom_histogram(binwidth = 1) + labs(title = "Number of Lung Transplants for COVID-19 related lung disease", x = "Transplant Year", y = "Number of Transplants")
```

In this figure, we see that more patients received a lung transplant for COVID-19 in year 2021 than 2020 or 2022. It will be interesting to see if transplants decreased during 2020 and then increased (or went back to baseline) in 2021. 

### 4.3 Causal Impact analysis of number of transplants

'CausalImpact' package was utilized to perform an analysis if there was a decrease in number of transplants after a particular intervention (start of COVID-19 pandemic) when compared to predicted transplants if the said intervention had not occurred. For this analysis to work, a control time period (where the intervention was not present) is needed. It also assumes the other potential confounding factors were stable before and after the 'intervention'. The pre-period or control period was selected as the period between 01-01-2015 and 03-14-2020 and the post period as 03-15-2020 (day when lockdown measures were first initiated)

```{r message = FALSE, warning = FALSE}
lungtx <- thoracic_data_subset_lung %>%
  filter(TXED == 'yes') %>%
  group_by(TX_DATE) %>%
  count(TX_DATE)
lungtx

library(CausalImpact)

lungtx <- lungtx

time.points <- seq.Date(as.Date("2015-01-01"), by=1, length.out=2803)
number <- ts(lungtx$n)
data <- zoo(number, time.points)
pre.period=as.Date(c("2015-01-01", "2020-02-29"))
post.period=as.Date(c("2020-03-01", "2020-12-31"))
impact <- CausalImpact(data, pre.period, post.period)
impact
plot(impact)
```

A statistically significant negative 'impact' was noted as a result of the pandemic. The number of transplants were 13% lower than predicted with a confidence interval of -21% to -4%. This means the causal effect was statistically significant and there was a decrease in lung transplants after the pandemic began in 2020.

Now let us see if effect from the 'intervention' i.e beginning of the pandemic continued to have a lasting effect on the lung transplant program. Below we performed the causalimpact analysis again extending the post period to Sept 2022 (latest available transplant data).

```{r message = FALSE, warning = FALSE}
time.points <- seq.Date(as.Date("2015-01-01"), by=1, length.out=2803)
number <- ts(lungtx$n)
data <- zoo(number, time.points)
pre.period=as.Date(c("2015-01-01", "2020-02-29"))
post.period=as.Date(c("2020-03-01", "2021-12-31"))
impact2 <- CausalImpact(data, pre.period, post.period)
impact2
plot(impact2)
```

Here the 95% confidence interval of the observed negative effect is [-22% to +2%]. This points that although there appears to be a decrease in the number of lung transplants when compared to predicted, the effect is not statistically significant. In other terms, it appears the effect of the pandemic wore off by the end of 2021, likely corelating to ramp up in transplants in the subsequent years.

### 4.4 Effect of the COVID-19 pandemic on the number of lung transplants done

Now getting to the next part of the analysis, we examine if the number of transplants done were affected by the COVID-19 metrics or predictors (Total number of cases on a given date, new COVID-19 cases, total COVID-19 patients hospitalized and COVID-19 patients admitted to the ICU). We also examine if a model containing all the above factors corelate better to a decrease in transplants.

The data was plotted (Number of transplants against the potential predictors).

```{r message = FALSE, warning = FALSE}
transplant.totalcases <- ggplot(lungtx_covid_join, aes(x = total_cases2, y = total_transplants, )) +
  geom_point(color = "darkseagreen3") +
  geom_smooth(method="lm") + labs(x = "Total COVID-19 cases", y = "Lung Transplants")
transplant.newcases <- ggplot(lungtx_covid_join, aes(x = total_new_cases, y = total_transplants,)) +
  geom_point(color = "darkseagreen3") +
  geom_smooth(method="lm") + labs(x = "New COVID-19 cases", y = "Lung Transplants")
transplant.totalhosp <- ggplot(lungtx_covid_join, aes(x = total_hosp, y = total_transplants)) +
  geom_point(color = "darkseagreen3") +
  geom_smooth(method="lm") + labs(x = "Hospitalized", y = "Lung Transplants")
transplant.icu <- ggplot(lungtx_covid_join, aes(x = total_icu, y = total_transplants)) +
  geom_point(color = "darkseagreen3") +
  geom_smooth(method="lm") + labs(x = "ICU", y = "Lung Transplants")

transplant.totalcases + transplant.newcases + transplant.totalhosp + transplant.icu
```

Examining the plots, it appears that new cases, hospitalized COVID-19 patients and COVID-19 patients in the ICU have a negative effect on the number of transplants. We examine this further with regression models below and tabulate the results.

```{r message = FALSE, warning = FALSE}
# Linear regression analyses
lungtx_covid_join_complete <- tidyr::drop_na(lungtx_covid_join)
lungtx_covid_lm_totalcases <- lm(total_transplants ~ total_cases2, data = lungtx_covid_join_complete)
lungtx_covid_lm_newcases <- lm(total_transplants ~ total_new_cases, data = lungtx_covid_join_complete)
lungtx_covid_lm_totalhosp <- lm(total_transplants ~ total_hosp, data = lungtx_covid_join_complete)
lungtx_covid_lm_totalicu <- lm(total_transplants ~ total_icu, data = lungtx_covid_join_complete)
lungtx_covid_lm_all <- lm(total_transplants ~ total_hosp + total_new_cases + total_icu, data = lungtx_covid_join_complete)

tab_model(lungtx_covid_lm_totalcases, lungtx_covid_lm_newcases, lungtx_covid_lm_totalhosp, lungtx_covid_lm_totalicu, lungtx_covid_lm_all, digits = 8, collapse.ci=TRUE, pred.labels = c("Intercept", "Total cases of COVID-19", "New Cases of COVID-19", "Hospitalized Cases of COVID-19", "ICU Cases of COVID-19"), dv.labels = c("Total Transplants v Total Cases of COVID-19", "Total Transplants v New Cases of COVID-19", "Total Transplants v Hospitalized Cases of COVID-19", "Total Transplants v ICU Cases of COVID-19", "Total Transplants v all"))
```

From the results of the linear regression models, we note that the number of lung transplants per month showed a statistically significant negative correlation to each of the following including: the number of new COVID-19 cases, COVID-19 hospitalized patients and COVID-19 patients in the ICU. The effect of total COVID-19 cases did not have a statistically significant correlation to the number of lung transplants. A model with all the predictors did not perform any better than the individual predictors. This is likely since the predictors themselves correlate highly and a model with all three predictors is not any better when compared to a single significant predictor.

### 4.5 Effect of the COVID-19 pandemic on the duration spent on waitlist

The effect of the pandemic on the waitlist is then examined. We try to answer the question here - Did the waitlist period increase during the pandemic and was this related to the above predictors? This time, we will need to go back to our waitlist data and include all patients (transplanted and not transplanted) who were on the waitlist. Similar plots and regression analysis was performed as above.

```{r message = FALSE, warning = FALSE}

# Filter waitlisted patients, convert to monthly and then join with COVID-19 data
lungtx2018_2021_wait <- thoracic_data_subset_lung_activation %>% dplyr::filter(ACTIVATE_DATE >= "2018-01-01" & ACTIVATE_DATE <= "2021-12-31") %>% mutate(TX_DATE = ymd(TX_DATE), TX_YEAR_MONTH = format_ISO8601(TX_DATE, precision = "ym")) %>% group_by(TX_YEAR_MONTH) %>% summarise(avgwait=mean(DAYSWAIT_CHRON), medianwait=median(DAYSWAIT_CHRON))

lungtx2018_2021_wait$TX_YEAR_MONTH <- as.Date(paste(lungtx2018_2021_wait$TX_YEAR_MONTH, "-01", sep=""))

lungtx_covid_wait_join <- left_join(lungtx2018_2021_wait, covid_cases.ym) %>% filter (TX_YEAR_MONTH >= "2020-02-01")

# Drop NAs
lungtx_covid_wait_complete <- tidyr::drop_na(lungtx_covid_wait_join)
head(lungtx_covid_wait_complete)

# Plots (Waitlist time v predictors)
wait.totalcases <- ggplot(lungtx_covid_wait_complete, aes(x = total_cases2, y = avgwait)) +
  geom_point(color = "darkseagreen3") +
  geom_smooth(method="lm") + labs(x = "Total COVID-19 cases", y = "Waitlist Time")

wait.newcases <- ggplot(lungtx_covid_wait_complete, aes(x = total_new_cases, y = avgwait)) +
  geom_point(color = "darkseagreen3") + 
  geom_smooth(method="lm") + labs(x = "New COVID-19 cases", y = "Lung Transplants")

wait.totalhosp <- ggplot(lungtx_covid_wait_complete, aes(x = total_hosp, y = avgwait)) + 
  geom_point(color = "darkseagreen3") +
  geom_smooth(method="lm") + labs(x = "Hospitalized", y = "Lung Transplants")

wait.icu <- ggplot(lungtx_covid_wait_complete, aes(x = total_icu, y = avgwait)) +
  geom_point(color = "darkseagreen3") +
  geom_smooth(method="lm") + labs(x = "ICU", y = "Lung Transplants")

library("patchwork")
wait.totalcases + wait.newcases + wait.totalhosp + wait.icu

# Linear regression analyses
wait_covid_lm_totcases <- lm(avgwait ~ total_cases2, data = lungtx_covid_wait_complete)

wait_covid_lm_newcases <- lm(avgwait ~ total_new_cases, data = lungtx_covid_wait_complete)

wait_covid_lm_hosp <- lm(avgwait ~ total_hosp, data = lungtx_covid_wait_complete)

wait_covid_lm_icu <- lm(avgwait ~ total_icu, data = lungtx_covid_wait_complete)

tab_model(wait_covid_lm_totcases, wait_covid_lm_newcases, wait_covid_lm_hosp, wait_covid_lm_icu, digits = 8, collapse.ci=TRUE, pred.labels = c("Intercept", "Total cases of COVID-19", "New Cases of COVID-19", "Hospitalized Cases of COVID-19", "ICU Cases of COVID-19"), dv.labels = c("Average Time Spent on Waitlist v Total Cases of COVID-19", "Average Time Spent on Waitlist v New Cases of COVID-19", "Average Time Spent on Waitlist v Hospitalized Cases of COVID-19", "Average Time Spent on Waitlist v ICU Cases of COVID-19"))

```

Interestingly, what we find here is that the waitlist time oositively correlated with the total number of COVID-19 cases. New COVID-19 cases and hospitalized COVID-19 patients did not have a statistically significant correlation with waitlist time. Interestingly, there was a statistically significant negative correlation of waitlist time to the number of patients in the ICU. It is difficult to explain this although one possibility is that those patients in the ICU ended up receiving a lung transplant for COVID-19 related lung disease and had a shorter wait time due to the nature of the illness. 


## Conclusions

We see here that there was a statistically significant decrease in lung transplants performed in the first year of the COVID-19 pandemic (2020). We see this effect dissipate over the next 1.5 years and the decrease was no longer statistically significant. On further analysis, the decrease in lung transplants had a statistically negative correlation to the new cases of COVID-19, COVID-19 hospitalizations and ICU patients with COVID-19. Total number of COVID-19 cases did not significantly correlate to number of transplants.

When we analyze the time spent on waitlist, there was a statistically significant increase in time on wait list with an increase in total COVID-19 cases but did not significantly correlate to hospitalized patients or new COVID-19 cases. Interestingly, there is a statistically significant decrease in time spent on waitlist when the number of COVID-19 ICU patients increased. This is counter intuitive and this phenomenon is not easily explained although one possibility is that those patients in the ICU ended up receiving a lung transplant for COVID-19 related lung disease and had a shorter wait time due to the nature of the illness. 

There are many limitations of the above analysis. The first problem is missing data. There is missing COVID-19 data as it relates to hospitalizations and ICU patients in the beginning of the pandemic. We also do not take into account the lung allocation score of the patients when we analyze the data. Furthermore, it is also possible that with improvement in how we care for patients with chronic lung disease, there may be a decreased need for lung transplantation. Improvement and increased use of extra-corporeal life support (ECLS) may be another potential factor decreasing the need for lung transplants in patients with acute lung injury by allowing time for natural healing.
