---
title: "Title Here"
author: "Alexander Chan"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
# commonly-used packages
library(ggplot2)
library(dplyr)
library(magrittr)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.
```{r, eval=TRUE}

```

### Overview
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.
```{r, eval=TRUE}

```

### Introduction 
Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.
```{r, eval=TRUE}

```

### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

###
Download and install Rtools40 to use the BLS api: https://cran.r-project.org/bin/windows/Rtools/

### Reading and Cleaning Datasets
```{r, eval=TRUE}
# tools for accessing BLS data
library(devtools)
library(blsAPI)
library(rjson)
install_github("mikeasilva/blsAPI", force = TRUE) # Install the blsAPI by Mike Silva
```


```{r, eval=TRUE}
# query data from BLS.gov

# load api key (must have the txt file)
api_key <-  read.csv("bls_api_key.txt", header=FALSE)[[1]]

# signatures to be queried from BLS.gov, 10-year CPI data from major metropolitan areas
signatures <- read.csv("metro_area_cpi_signatures.csv") 

# payload signatures to query BLS 
month_payload <- list('seriesid'= signatures$signature_monthly) #2018-2020 monthly CPI 
annual_payload <- list('seriesid'= signatures$signature_annual) #2018-2020 annual CPI

# query CPI 2018-2020 data from major metropolitan areas from BLS 
months_cpi <- blsAPI(month_payload, return_data_frame = TRUE) # monthly CPI dataframe
annual_cpi <- blsAPI(annual_payload, return_data_frame = TRUE) # annual CPI dataframe
```

```{r, eval=TRUE}
# format the data

# add columns to join region labels by
months_cpi$signature_monthly <- months_cpi$seriesID
annual_cpi$signature_annual <- annual_cpi$seriesID

# join region and state labels to the cpi data
# complete monthly dataframe for 2018-2020 regional metro cpi data
months_cpi %<>%
  left_join(signatures[, c("signature_monthly","state","area")],by="signature_monthly") %>%
  dplyr::select(-seriesID)

# complete annual dataframe for 2018-2020 regional metro cpi data
annual_cpi %<>%
  left_join(signatures[, c("signature_annual","state","area")],by="signature_annual") %>%
  dplyr::select(-seriesID)


head(months_cpi)
head(annual_cpi)

```
```{r, eval=TRUE}
# Further format dataframes
# pivot data by metro area and years, add column for change in cpi over 2 years

# annual cpi 
a_cpi <- annual_cpi %>% 
  pivot_wider(id_cols = c("area","year"),names_from = "periodName",
              values_from = "value")%>% # pivot on area and year using values from period
  mutate_at(vars(3:4), as.numeric) %>% # change CPI values to numeric
  mutate(annual = rowMeans(select(., c("1st Half", "2nd Half")), na.rm= TRUE)) # mean cpi

m_cpi <- months_cpi %>%
  dplyr::select(-period) %>%
  pivot_wider(id_cols = c("area","year"), names_from = "periodName", 
              values_from = "value") %>% # pivot on area and year using values from period
  mutate_at(vars(2:14), as.numeric) %>% # change CPI values to numeric
  mutate(annual = rowMeans(select(.,-c("area", "year")), na.rm = TRUE)) # mean cpi

```


```{r, eval=TRUE}
# read in the housing costs: fair market rates (2017-2020)
fmr_2017 <- read.csv("FY2017_50_County_FMR.csv")
fmr_2018 <- read.csv("FY2018_50_County_FMR.csv")
fmr_2019 <- read.csv("FY2019_50_County_FMR.csv")
fmr_2020 <- read.csv("FY2020_50_County_FMR.csv")
fmr_2021 <- read.csv("FY2021_50_County_FMR.csv")

```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
```{r, eval=TRUE}

```
