---
title: "BMIN503/EPID600 Project"
author: "Sy Hwang"
output: 
  html_document:
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
**Please NOTE that because my project works with PHI and identifiable data from the Penn Medicine EHR(with IRB approval),
much of my code will use the option `results='hide'` in the R code chunks, and therefore results of the code will be hidden.**

### Overview
This project explores the geographical phenomena and processes of SARS-CoV-2(Covid-19) among patients of the University of Pennsylvania
Health System(UPHS), which serves the Greater Philadelphia Metropolitan Area. We dive into a spatiotemporal analysis of Covid-19
and explore how the pandemic has affected certain populations of the area, specifically looking at demographic, socioeconomic, clinical
and environmental correlations with outpatient outcomes like testing(both positive and negative) and inpatient outcomes like 
hospitalization/intubation/mortality.

#### Contributors to the project include:
  * Blanca Himes, PhD
  * Danielle Mowery, PhD
  * Sherry Xie, VMD-PhD candidate
  * Alexandra Rizaldi, BS candidate
  * Emily Schriver, MS


### Introduction
The Covid-19 pandemic has affected millions on a global scale, having a prolonged, devastating impact on public health and economies in practically every part of the world.
It has been no different in Philadelphia, where the infection cases grew rapidly in waves in 2020, but also disproportionately affected communities of color, 
with predominantly African Americans communities experiencing the highest rates of disease severity and mortality. Examining Covid-19 in Philadelphia offers a unique opprotunity and perspective as one of the United States' most racially and ethnically diverse metropolitan cities.
One of the ways in which we can help improve our current understanding of the pandemic is by conducting a multidisciplinary study on how the disease affected the city over time.
Mapping this space-time progression of the disease can highlight the level of spread and factors that facilitated the spread,
as well as bring to light the behavioral, socioeconomic and environmental factors that correlated with outcomes of the disease.
The hope is that this analysis can further be used in the future for determining risk factors, ameliorating resource allocation issues and contributing to improved disease surveillance. 


### Data
This is a restrospective study examining UPHS EHR patients with at least one primary diagnosis ICD-10 code of flu-like symptoms or pneumonia,
or patients who had one procedure code for Covid-19 or influenza testing from January 2020 to January 2021. The cohort consists of 97,710 patients, but due to missing data points the cohort was reduced to 72,908 patients.
The study also incorporates publicly available data such as US Census Bureau data, community health data from the Public Health Management Corporation 
and environmental data from the US Environmental Protection Agency.

### Methods
The primary method of analysis will be divided into three parts:

1. Conduct a chi-squared test to assess difference is sociodemographic characteristics of patietns who tested positive in the early vs latter stages of the pandemic.
2. Visually study the clusters of Covid-19 cases and their distributional shift over the 1-year span.
3. Perform logistic regression to determine the demographic, socioeconomic and environmental factors that are associated with adverse health outcomes of Covid-19.

```{r}
library(dplyr)
labs_filename <- "./labresults_lab4.8.21_7.11.2021.rds"
copd_filename <- "./copd_encounters_2021_01_21.csv"
icd_filename <- "./"
#data <- readRDS(filename)
data <- read.csv(copd_filename)
glimpse(data)
```


### Results
TODO
