---
title: "Predictors of Traumatic Brain Injury Outcome"
author: "Margalit Haber"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
###Overview 
An abnormality on a head computed tomography scan (HCT) is commonly used as an indication that a traumatic brain injury (TBI) event occurred for clinical studies, however, the diagnostic utility of a head CT abnormality for a TBI is not well understood. This project aims to identify whether a HCT abnormality is useful as a prognostic biomarker or whether other factors are more relevant for prognosis.

###Introduction
There are over 2 million occurrences of traumatic brain injury (TBI) in the united states every year. Of those 2 million cases, there is currently no means of predicting who will recover and who will develop chronic long-lasting symptoms. The inability to determine who will recover without intervention makes designing clinical research studies sensitive to patient selection. The inclusion and exclusion criteria selected for a TBI study is crucial to the success or failure of that study. A common inclusion criteria for TBI studies is the presence of a tissue abnormality on a head computed tomography scan (HCT). It is utilized since it ensures that the underlying brain tissue is damaged after the impact, however, the diagnostic utility of HCT abnormalities has not been well characterized. This project aims to determine the diagnostic and prognostic utility of a HCT abnormality for TBI.
	
The analysis from this project will not only focus on HCT abnormalities as a biomarker for TBI, but will also investigate socioeconomic status and age as factors contributing to TBI outcome. The knowledge learned from this project will inform the medical field as to factors that make an individual susceptible to poor outcome after a TBI. This project is also aimed at taking a critical look at clinical research study design, and whether more informed decisions about subject selection is needed for future studies. Initial study design and subject selection criteria is important for the success of clinical research studies as well as epidemiological studies of secondary datasets.


### Methods
The proposed project utilizes a cohort of 99 subjects recruited at Penn Presbyterian Medical Center. The subjects were recruited within 72 hours of their injury, and then followed-up at 2 weeks, 3 months, and 6 months. The dataset includes basic demographic information as well as neuropsychiatric tests for symptomology and functionality. Linear regression will be used to determine whether positive head CT is associated with long lasting symptoms or outcomes. The symptoms and outocmes being tested will be drawn from the standardized tests Rivermead symptoms inventory, BSI, GOSE, and Insomnia severity scale.

```{r eval=TRUE}
Data <- read.csv("/Users/margalithaber/Desktop/Data_12.4.2017.csv", header=TRUE)
#or for dummy data 
#Data <- read.csv(url("https://raw.githubusercontent.com/HaberM/BMIN503_Final_Project/master/DummyData.csv"), header=TRUE)

library(plyr)
library(dplyr)
library(lubridate)
library(ggplot2)

Data.clean <- Data %>%
  mutate(gender=factor(sex_M, levels=c(0, 1, 2), labels=c(NA, "male", "female"))) %>%
  mutate(hct=factor(hct, levels=c(1, 2), labels=c("positive", "negative"))) %>%
  mutate(race=factor(race, levels=c(0, 1, 2, 3, 4, 5, 6), labels=c(NA, "Indian", "AlaskanNative.Inuit", "Asian", "NativeHawaiian.PacificIslander", "Black", "White"))) %>%
  mutate(severity=recode(gcs, '1'=3, '2'=3, '3'=3, '4'=3, '5'=3, '6'=3, '7'=3, '8'=3 , '9'=2, '10'=2, '11'=2, '12'=2 , '13'=1, '14'=1, '15'=1)) %>%
  mutate(severity=factor(severity, levels=c(1, 2, 3), labels=c("Mild", "Moderate","Severe"))) %>%
  mutate(alcohol=factor(alcohol, levels=c(1, 2), labels=c("Sober", "Drunk"))) %>%
  filter(group == 1) %>%
  filter(!is.na(age)) %>%
  filter(!is.na(hct)) %>%
 mutate(education_years=as.integer(as.character(education_years))) %>%
  mutate(Month=month(as.POSIXlt(injury_date, format="%m/%d/%Y")))
```

### Results

```{r eval=TRUE}

#demographics of study
nrow(Data.clean)
summary(Data.clean$age)
summary(Data.clean$education_years)
summary(Data.clean$gcs)

Pos.HCT <- subset(Data.clean,hct=='positive')
Neg.HCT <- subset(Data.clean,hct=='negative')
nrow(Pos.HCT)
summary(Pos.HCT$age)
summary(Pos.HCT$education_years)
summary(Pos.HCT$gcs)
nrow(Neg.HCT)
summary(Neg.HCT$age)
summary(Neg.HCT$education_years)
summary(Neg.HCT$gcs)


#checking for age and education matching
#difference in age between positive and negative hct groups
ggplot(Data.clean,aes(age)) + 
    geom_histogram(data=subset(Data.clean,hct == 'positive'), aes(y=..density..), fill = "red", alpha = 0.4, bins = 20) +
    geom_histogram(data=subset(Data.clean,hct == 'negative'), aes(y=..density..), fill = "blue", alpha = 0.4, bins = 20) +
  labs(title="HCT+/- by Age") +
  theme_bw()

#difference in education level between positive and negative hct groups
ggplot(Data.clean,aes(education_years)) + 
    geom_histogram(data=subset(Data.clean,hct == 'positive'), aes(y=..density..), fill = "red", alpha = 0.4, bins = 10) +
    geom_histogram(data=subset(Data.clean,hct == 'negative'), aes(y=..density..), fill = "blue", alpha = 0.4, bins = 10) +
  labs(title="HCT+/- by Years of Education") +
  theme_bw()

#difference in GCS at ED presentation between positive and negative hct groups
ggplot(Data.clean,aes(gcs)) + 
    geom_histogram(data=subset(Data.clean,hct == 'positive'), aes(y=..density..), fill = "red", alpha = 0.4, bins = 10) +
    geom_histogram(data=subset(Data.clean,hct == 'negative'), aes(y=..density..), fill = "blue", alpha = 0.4, bins = 10) +
  labs(title="HCT+/- by GCS") +
  theme_bw()
```

The dataset consists of 79 TBI subjects with a median age of 54 (18-93) years and median education of 12 (9-20) years. The Glasgow coma scale score (GCS) is used to identify injury severity. The scores range from 3 (comatose) to 15 (awake and responding.) The median GCS was 15 (3-15). The positive HCT group (HCT+) consisted of 63 subjects with a median age of 56 (18-93) and median 12 years of education of (9-20). The median GCS for the HCT+ group was 15 (3-15.) The negative HCT group (HCT-) was slightly younger with a median age of 42 (26-75) years. The median GCS and years of education for the HCT- group was comparable to the the HCT+ group.


```{r eval=TRUE}
ggplot(data=Data.clean, aes(gender)) + 
      geom_bar(fill="black") +
      labs(title="Demographics by gender")+
      labs(x="Gender")+
      theme_bw()
      
ggplot(data=Data.clean, aes(x=race, fill=factor(gender))) +
      geom_bar(position="stack") +
      labs(title="Demographics by race") +
      labs(x="Race")+
      theme_bw()

ggplot(data=Data.clean, aes(x=alcohol, fill=factor(gender))) +
      geom_bar(position="dodge") +
      labs(title="Sobriety at ED presentation") +
      labs(x="Alcohol") +
      theme_bw()

ggplot(data=Data.clean, aes(age, gcs, color=gender)) + 
    geom_point() + 
    labs(title="GCS at ED presentation across age") +
    labs(x="Age", y="GCS") + #Adds a layer with labels
    theme_bw()
```

There are proportionally more males enrolled in the study than females, which is consistant with previous literature regarding incidence of TBIs. The population consists of more white individuals. Incidence of inebriation at emergency department presentation for a TBI was higher in men than in women. The majority of subjects were sober at emergency department presentation. The majority of subjects in the dataset had a GCS of 15.

Severity for TBI is currently classified by glascow coma scale scores (GCS). TBI severity is generally broken down into 3 categories: Mild(gcs=15-13), Moderate(gcs=9-12), and Severe(gcs=1-8). 

```{r eval=TRUE}
ggplot(data=Data.clean, aes(severity, age)) +
    geom_violin(fill="lavender", draw_quantiles = c(0.25, 0.5, 0.75)) +
    geom_jitter(height=0, width=0.1) +
    labs(title="Severity of injury at ED presentation") +
    xlab("Severity (GCS)") +
    ylab("Age") +
    theme_bw() #violin plot indicating age distribution across injury severity 
```

The population in this study showed a high proportion of Mildly injured subjects (primarily gcs=15). The age of subjects was well distributed over the Mild and Severe injury categories. There are few subjects falling within the Moderate injury category.

```{r eval=TRUE}
ggplot(data=Data.clean, aes(x=severity, fill=factor(hct))) +
      geom_bar(position="stack") +
      labs(title="HCT+/- by severity") +
      theme_bw()
```

HCT+ and HCT- subjects were found in both the Mild and Severe injury groups.

```{r eval=TRUE}
#Is a head ct abnormality related to functionality 2wks-1month after injury?
ggplot(data=Data.clean, aes(x=hct, y=gose_score_1mo)) +
    geom_boxplot(fill="lavender") +
    labs(title="HCT and GOSE: 2wk-1mo") +
    theme_bw()
summary((lm(gose_score_1mo~hct, data=Data.clean)))

table(Data.clean$gose_score_3mo, Data.clean$hct)
table(Data.clean$gose_score_6mo, Data.clean$hct)
#not enough subjects in HCT- group for 3 month and 6 month followup visits to analyze. 

#Is a head ct abnormality related to symptoms 2wks-1month after injury?
ggplot(data=Data.clean, aes(x=hct, y=rpq3_1mo)) +
    geom_boxplot(fill="lavender") +
    labs(title="HCT and RPQ3: 2wk-1mo") +
    theme_bw()
summary((lm(rpq3_1mo~hct, data=Data.clean)))
#not significant

ggplot(data=Data.clean, aes(x=hct, y=rpq13_1mo)) +
    geom_boxplot(fill="lavender") +
    labs(title="HCT and RPQ13: 2wk-1mo") +
    theme_bw()
summary((lm(rpq13_1mo~hct, data=Data.clean)))
#not significant

#Is a head ct abnormality related to sleep disturbances after injury?
ggplot(data=Data.clean, aes(x=hct, y=isi_falling_1mo)) +
    geom_boxplot(fill="lavender") +
    labs(title="HCT and Falling Asleep: 2wk-1mo") +
    theme_bw()
summary((glm(isi_falling_1mo~hct, data=Data.clean)))
#not significant

ggplot(data=Data.clean, aes(x=hct, y=isi_staying_1mo)) +
    geom_boxplot(fill="lavender") +
    labs(title="HCT and Staying Asleep: 2wk-1mo") +
    theme_bw()
summary((lm(isi_staying_1mo~hct, data=Data.clean)))
#not significant
```

After reviewing the number of followup visits per HCT group, I discovered that the sample size for the HCT- group was not high enough to compare at 3 months (n=3) and 6 months (n=1) after injury. Analysis was focused on the subacute symptoms and outcome to HCT abnormalities, the 2 weeks to 1 month time point. Linear regression models were generatef using HCT as a predictor of several outcome metrics. GOSE looks at how functional the subject is in their daily life. RPQ is a symptoms inventory looking at self-reported symptoms like depression and headaches. Two insomnia indices were selected from the insomnia severity scale since sleep disturbances is a common symptom after a head injury. There was no significant difference between the HCT+ v HCT– groups for GOSE, RPQ symptoms assessment for 3 and 13, nor for the indicies of difficulty falling asleep or staying asleep. HCT abnormalities do not seem to be associated with subacute outcome measures.

Next I investigated whether there are any predictors for the occurance of HCT abnormalities.

```{r eval=TRUE}
#Does the severity of the injury affect whether the injury results in a head ct abnormality?
ggplot(data=Data.clean, aes(x=hct, fill=factor(severity))) +
      geom_bar(position="dodge") +
      labs(title="Severity vs. presence of HCT abnormalities") +
      labs(x="Severity(GCS)") +
      theme_bw()
summary((glm(hct~severity, data=Data.clean, family=binomial())))
#not significant

#Is education related to whether the injury results in a head ct abnormality?
ggplot(data=Data.clean, aes(x=hct, y=education_years)) +
    labs(title="Years of Education v. HCT abnormalities") +
    labs(y= "Education (yrs)") +
    geom_boxplot(fill="lavender") +
    theme_bw()
summary((glm(hct~education_years, data=Data.clean, family=binomial())))

#Does age affect whether the injury results in a head ct abnormality?
ggplot(data=Data.clean, aes(x=hct, y=age)) +
    geom_boxplot(fill="lavender") +
    labs(title="Age vs. presence of HCT abnormalities") +
    theme_bw()
summary((glm(hct~age, data=Data.clean, family=binomial())))
#slight trend

#Does sobriety affect whether the injury results in a head ct abnormality?
ggplot(data=Data.clean, aes(x=alcohol, fill=factor(hct))) +
      geom_bar(position="fill") +
      labs(title="Sobriety v. HCT abnormalities") +
      labs(x="Alcohol") +
      theme_bw()
summary((glm(hct~alcohol, data=Data.clean, family=binomial())))
#stronger trend

#accounting for age as a covariate for the effect of sobriety on ocurrance of a HCT+
summary((glm(hct~alcohol+age, data=Data.clean, family=binomial())))
#significance only slightly increased
```

Severity and years of education were not significantly associated with HCT+/-. Age showed a slight trend (p=0.18.) Sobriety had a stronger trend at a (p=0.1), suggesting that being drunk on emergency department presentation means you are more likely to have no HCT abnormality than being sober. When age is included in the model for sobriety, the p value increases minimally to 0.14. None of these predictors reached statistical significance. 

```{r eval=TRUE}

#Does age affect injury outcome?
ggplot(data=Data.clean, aes(age, gose_score_1mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Age: 2wk-1mo") +
    labs(y="GOSE", x="Age") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_1mo~age, data=Data.clean)))
#not significant

ggplot(data=Data.clean, aes(age, gose_score_3mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Age: 3mo") +
    labs(y="GOSE", x="Age") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_3mo~age, data=Data.clean)))
#not significant

ggplot(data=Data.clean, aes(age, gose_score_6mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Age: 6mo") +
    labs(y="GOSE", x="Age") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_6mo~age, data=Data.clean)))
#not signficant

ggplot(data=Data.clean, aes(age, rpq13_1mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Age: 2wk-1mo") +
    labs(y="RPQ13", x="Age") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_1mo~age, data=Data.clean)))
#significant with RPQ13

summary((lm(rpq13_1mo~age+gcs, data=Data.clean)))
summary((lm(rpq13_1mo~age+severity, data=Data.clean)))
#age is still significant when adding severity or GCS to the model

ggplot(data=Data.clean, aes(age, rpq13_3mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Age: 3mo") +
    labs(y="RPQ13", x="Age") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_3mo~age, data=Data.clean)))
#not significant

ggplot(data=Data.clean, aes(age, rpq13_6mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Age: 6mo") +
    labs(y="RPQ13", x="Age") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_6mo~age, data=Data.clean)))
#trending but not significant


#Does number of years of eduction affect injury outcome?
ggplot(data=Data.clean, aes(education_years, gose_score_1mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Level of Education: 2wk-1mo") +
    labs(y="GOSE", x="Education (yrs)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_1mo~education_years, data=Data.clean)))
#signficant

summary((lm(gose_score_1mo~education_years+age+gcs, data=Data.clean)))
#still significant when accounting for age and gcs

ggplot(data=Data.clean, aes(education_years, gose_score_3mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Level of Education: 3mo") +
    labs(y="GOSE", x="Education (yrs)") + 
  geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_3mo~education_years, data=Data.clean)))
#significant

summary((lm(gose_score_3mo~education_years+age+gcs, data=Data.clean)))
#still at p=0.5 when accounting for age and gcs

ggplot(data=Data.clean, aes(education_years, gose_score_6mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Level of Education: 6mo") +
    labs(y="GOSE", x="Education (yrs)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_6mo~education_years, data=Data.clean)))
#not signficant for GOSE at 6 months, may be due to small sample size

ggplot(data=Data.clean, aes(education_years, rpq3_1mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Level of Education: 2wk-1mo") +
    labs(y="RPQ13", x="Education (yrs)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_1mo~education_years, data=Data.clean)))
#trending for RPQ13

ggplot(data=Data.clean, aes(education_years, rpq3_3mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Level of Education: 3mo") +
    labs(y="RPQ13", x="Education (yrs)") + 
  geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_3mo~education_years, data=Data.clean)))
#highly significant

summary((lm(rpq13_3mo~education_years+age+gcs, data=Data.clean)))
#still significant when accounting for age and gcs

summary((lm(rpq13_6mo~education_years, data=Data.clean)))
ggplot(data=Data.clean, aes(education_years, rpq3_6mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Level of Education: 6mo") +
    labs(y="RPQ13", x="Education (yrs)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
#almost significant for RPQ13 at p=0.055

#Does race affect injury outcome?
ggplot(data=Data.clean, aes(race, gose_score_1mo)) +
    labs(title="Injury Outcome by Race: 2wk-1mo") +
    geom_boxplot(fill="lavender")+
    theme_bw()
summary((lm(gose_score_1mo~race, data=Data.clean)))

ggplot(data=Data.clean, aes(race, gose_score_3mo)) +
    labs(title="Injury Outcome by Race: 3mo") +
    geom_boxplot(fill="lavender")+
    theme_bw()
summary((lm(gose_score_3mo~race, data=Data.clean)))

ggplot(data=Data.clean, aes(race, gose_score_6mo)) +
    labs(title="Injury Outcome by Race: 6mo") +
    geom_boxplot(fill="lavender") + 
    theme_bw()
summary((lm(gose_score_6mo~race, data=Data.clean)))
#not signifcant at any time point

ggplot(data=Data.clean, aes(race, gose_score_1mo)) +
    labs(title="Injury Outcome by Race: 2wk-1mo") +
    geom_boxplot(fill="lavender")+
    theme_bw()
summary((lm(rpq13_1mo~race, data=Data.clean)))

ggplot(data=Data.clean, aes(race, gose_score_3mo)) +
    labs(title="Injury Outcome by Race: 3mo") +
    geom_boxplot(fill="lavender")+
    theme_bw()
summary((lm(rpq13_3mo~race, data=Data.clean)))

ggplot(data=Data.clean, aes(race, gose_score_6mo)) +
    labs(title="Injury Outcome by Race: 6mo") +
    geom_boxplot(fill="lavender") + 
    theme_bw()
summary((lm(rpq13_6mo~race, data=Data.clean)))
#not significant at any time point

#Does the time of year in which the injury occured affect outcome?
ggplot(data=Data.clean, aes(Month, gose_score_1mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Time of Year: 2wk-1mo") +
    labs(y="GOSE", x="Time of Year (month)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_1mo~Month, data=Data.clean)))
#significant with GOSE at 2wk-1mo

summary((lm(gose_score_1mo~Month+gcs+age, data=Data.clean)))
#still significant when accounting for age and gcs

ggplot(data=Data.clean, aes(Month, gose_score_3mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Time of Year: 3mo") +
    labs(y="GOSE", x="Time of Year (month)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_3mo~Month, data=Data.clean)))
#not significant

ggplot(data=Data.clean, aes(Month, gose_score_6mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Time of Year: 6mo") +
    labs(y="GOSE", x="Time of Year (month)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(gose_score_6mo~Month, data=Data.clean)))
#not signifcant

ggplot(data=Data.clean, aes(Month, rpq13_1mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Time of Year: 2wk-1mo") +
    labs(y="RPQ13", x="Time of Year (month)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_1mo~Month, data=Data.clean)))


ggplot(data=Data.clean, aes(Month, rpq13_3mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Time of Year: 3mo") +
    labs(y="RPQ13", x="Time of Year (month)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_3mo~Month, data=Data.clean)))


ggplot(data=Data.clean, aes(Month, rpq13_6mo, color=severity)) +
    geom_point() + 
    labs(title="Injury Outcome by Time of Year: 6mo") +
    labs(y="RPQ13", x="Time of Year (month)") + 
    geom_smooth(color="black", method="lm") +
    theme_bw()
summary((lm(rpq13_6mo~Month, data=Data.clean)))
#not significant at any time point
```

Age, Race, and time of year did not have strong or consistant relationships with GOSE or RPQ13. The only factor that did show a consistant pattern and significance was years of education. GOSE score improves with years of education. Years of education is a significant predictor of GOSE after injury at 2wk-1mo and 3 months after injury, even when accounting for age and GCS. 6 months is not significant which may be due to the small sample size. RPQ13 which looks at symptoms also shows a significance or a trend toward significance at all time points suggesting that years of education is actually protective against chronic symptoms after TBI.


### Conclusions

Given the heterogeneity of TBI, The sample size in this study is small. An n of 79 subjects with only 16 HCT- subjects makes the results of this study not 100% conclusive. However, the data from this study suggests that a HCT+ finding is an informative inclusion criteria for brain injury studies and should not be utilized going forward. Many of the current studies in our lab have already eliminated that criteria from our inclusion however we have collaborators still requiring it. Years of education appears to be predictive of improved outcome after a brain injury. Several papers have shown a similar finding, suggesting a potential cognitive reserve for TBI similar to what is seen in Alzheimer's disease patients. A large part of this data set is being collected for a multi-center study called TRACK-TBI. There are currently over 15 institutions participating in TRACK-TBI, collecting similar data in the hopes of defining a new taxonomy for TBI. The data is available to any interested party, however, the time to recieve the TRACK-TBI data can take up to a year. The subject pool contains 1000s of subjects and our lab is currently in the process of requesting that data for other projects. These analyses can be extended to more subjects in the future by using the TRACK-TBI dataset.

