---
title: "BMIN503 Final Project: Parsimonious Classifiers for ARDS Subphenotypes"
author: "Heather Giannini"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(width = 400)
```  
***
### Overview
Acute respiratory distress syndrome (ARDS) is a high morbid condition of respiratory failure, resulting from a myriad of insults, including infection (most notably recently, COVID19), trauma, and other underlying etiologies. We are investigating a recently published family of parsimonious models for identifying subphenotypes in this disease. Clinical trajectory of ARDS  can be highly divergent , and yet we do not have a clear understanding of the biological basis for vastly different outcomes and clinical behavior. This project will compare the assignments of  two similar but slightly different biomarker-based algorithms. The parsimonious models use a cytokine, interleukin-6 or interleukin-8, both markers of a hyperinflammatory state, and couple this measurement with the patient's level of soluble tumor necrosis factor receptor 1 (sTNFR1) and bicarbonate level. We will apply both IL6- and IL8- based models, evaluate how the population is categorized into the two subphenotypes, "Hyperinflammatory" and "Hypoinflammatory" and then correlate both clinical features and protein associations.
This work was done in consultation with Dr. Nuala Meyer (Pulmonary and Critical Care) and Dr. Pratik Sinha (UCSF, Clinical Respiratory Physiology and Biostatistics) and Dr.Todd Miano (PharmD, Dept of Epidemiology).
>Nuala Meyer provides primary mentorship for this project, providing expertise in interleukin dynamics of ARDS and associated critical-illness syndromes like sepsis. She has published previously on dynamics of IL6 and IL8 expression in conditions like neutropenia and sepsis.
>Dr. Pratik Sinha derived these parsimonious models after initial derivation of subphenotype classes (Hyper and Hypoinflamm) via Latent Class analysis of ARDS population. He applied these classifiers to our patient data to assign phenotypes.
>Dr. Todd Miano is an expert in epidemiology and study design and provided considerable expertise in coding variables and obtaining data from Penn Data Store to encode bicarbonate for the preliminary variables that were sent to Dr. Sinha for phenotype assignment.

Final Project GitHub Repository: <https://github.com/gianninh/BMIN503_Final_Project>

### Introduction 
Acute respiratory distress syndrome (ARDS) is a devastating pulmonary condition with high patient-level variability that has extremely high mortality and long-term morbidity in survivors. In clinical trials and observational studies for ARDS, two clinical subphenotypes - distinguished by plasma levels of inflammatory proteins and correlating with 30-d mortality - have been consistently identified by latent class analysis (LCA). These two classes are termed “Hyperinflammatory” and “Hypoinflammatory”. These phenotypes demonstrate heterogeneous treatment response to multiple ARDS therapies, including steroids (in COVID19 ARDS). Although iterations of parsimonious models for assigning these LCA classes based on biomarker combinations have similar operating characteristics, their agreement at the patient level has not been directly compared. We hypothesized that the performance of these models may vary in the context of clinical factors that disproportionately affect expression of the plasma proteins used for subphenoytype assignment

This project will evaluate these biomarker-based Parsimonious Models for ARDS subphenotypes. We will apply the models to our ARDS cohort, a subcohort of the Molecular Epidemiology of Sepsis in the ICU (MESSI) sepsis cohort. We will then test associations of these class assignments with various clinical characteristics, including medical comorbidities, biomarker protein levels, laboratory values (ie. white blood cell count). Importantly, we will compare the assignments of two different models, which may vary based on the biomarker used, either IL6 or IL8, two similar cytokines that are typically congruent but may be proportionally different in certain clinical circumstances.

### Methods
471 critically ill patients with ARDS were enrolled in the study within 24 hours of admission to the intensive care unit. Laboratory values and clinical features were extracted from the EHR and input into a redcap data repository. ARDS was defined per Berlin criteria. Plasma proteins (IL6, IL8, sTNFR1) were measured by electrochemiluminescence on blood obtained <24hrs of ICU admission. We applied two parsimonious models for LCA subphenotypes [IL6/bicarbonate/sTNFR1] or [IL8/bicarbonate/sTNFR1] which return probability of assignment to Hyperinflammatory class. Comorbidities were compared between predicted model assignments using the Chi Square or Fisher’s Exact tests. Association of neutropenia (ANC <1000) with class assignments for both models was performed using Fisher’s Exact test.

Derivation of Parsimonious Classifiers for ARDS
<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7543720/>
Development and Validation of Parsimonious Algorithms to Classify ARDS Phenotypes: Secondary Analyses of Randomised Controlled Trials

```{r, echo=F}
# load necessary libraries
library(tidyverse)
library(visdat)
library(naniar)
library(readr)
library(haven)
library(gtsummary)
library(readxl)
library(pheatmap)
library(Hmisc)
library(kableExtra)
library(summarytools)
library(cowplot)
library(dlookr)
library(ggpubr)
library(rstatix)
library(knitr)
```

## Comparing the IL6- and IL8- Model Assignments

First, we will look at descriptive characteristics of the IL6 and
IL8-based models for distribution of probabilities and IL6/IL8 levels.
Here we will also calculate an IL6:IL8 ratio, and test the association
of this ratio (continuous variable) with the categorical assignments.

The two datasets we are reading in contain measures for each MESSI sample ID (each ID is a patient enrolled in the study), with corresponding IL6, IL8, sTNFR1 and bicarbonate level on day 1. The two datasets below also contain the subphenotype class assignments resulting from use of either IL6- or IL8- based model. Finally, the probability of assignment to "Hyperinflammatory" is also included.

```{r reading in datasets}
## Comparing Performance of IL6 vs IL8 models
# read in results from both models
il6mod <- read_excel("~/R_Projects/MESSI_Gene_Exp/ards pars model/ards_model2_calc_done.xlsx")
il8mod <- read_excel("~/R_Projects/MESSI_Gene_Exp/ards pars model/ards_model3_calc_done.xlsx")

head(il6mod)
```

We will need to simplify and merge these in order to compare how the patients are classified by each model.

```{r data cleaning and organizing}

# remove bicarb/sTNFR1 from one dataset so they aren't duplicated in the join
il8mod <- il8mod %>% select(-Bicarbonate, -sTNFR1)

# combine the model 2 and model 3 data for one data table for analysis
# make a new variable called IL6:IL8 ratio so we can use this in our analysis later
mod_join <- left_join(il6mod, il8mod, by="messi_id")
mod_join <- mod_join %>% select(-Model.x, -Model.y) %>%
  rename("IL6_Model"=Class.x, "IL8_Model"=Class.y) %>%
  rename("IL6_Probability"=Probability.x, "IL8_Probability"=Probability.y) %>%
  mutate("IL6:IL8"=IL6/IL8)

# summarize both models assignments
mod_join %>% 
  select(IL6_Model, IL8_Model, IL6_Probability, IL8_Probability, IL6, IL8)%>% 
  tbl_summary(by=IL6_Model)%>%
  modify_caption("by IL6 Model Class Assignments")
```

We see that some of the class assignments are concordant, while some of the class assignments are discordant between IL6 and IL8.

## Evaluating Discordance in Model Probabilities
We will use a heatmap representing the probabilities for each
model to demonstrate that while the models tend to agree most of the time,
there are some samples which clearly diverge depending on IL6 vs IL8
analyte.

```{r heatmap}
mat <- mod_join %>% select(IL6_Probability, IL8_Probability)
mat <- as.matrix(mat)
pheatmap(mat, cluster_cols = F)
```

We can see there is a group of patients at the top which were low
Probability in IL6, but clearly very high in IL8. And a group in the
middle which were low IL8, but high IL6.

## Cohort that demonstrates class switching
There are 98 patients that switch classes between IL6- and IL8-based
models.

```{r dual-class assignments for statistics}
# now make a new variables coding the dual-class assignments to evaluate groups
mod_join <- mod_join %>% 
  mutate(a = case_when(IL6_Probability >= 0.5 ~ "hyperIL6", IL6_Probability <= 0.5 ~ "hypoIL6"))%>% 
  mutate(b = case_when(IL8_Probability >= 0.5 ~ "hyperIL8", IL8_Probability <= 0.5 ~ "hypoIL8"))%>%
  unite("combined_models", a:b, remove = TRUE)

mod_join %>% select(-messi_id, -IL6_Model, -IL8_Model) %>% 
  tbl_summary(by="combined_models") %>%
  modify_caption("Both Model Assignments")

p1 <- ggplot(mod_join, aes(x=combined_models, y=IL6_Probability)) + geom_boxplot() + theme_bw()
p2 <- ggplot(mod_join, aes(x=combined_models, y=IL8_Probability)) + geom_boxplot() + theme_bw()
plot_grid(p1,p2)
```
It does appear that the probabilities - particularly for IL6 Model - are
less extreme, from 0.88 - 0.68 for Hyper and 0.05 - 0.22 for Hypo. IL8
probabilities mellow a bit as well, but maybe not as much.

The more interesting thing here is that the IL8 values are **higher** in
the hypoinflammatory group per IL6 model, which is likely why the IL8
Model tends to put them in hyperinflamm.

**So who are these patients who have relatively higher IL8, despite a
lower IL6 value?**

## Acute Clinical Variables
Next, we will bring in a dataset of clinical variables for each subject. These reflect details about acute illness and demographics including: **apacheIII, los, race, sex, age, hospMort30, mort90, sepsis source** but has a lot of other variables that we will not be working with. 

```{r}
# load in full MESSI clinical dataset, this is over 3000 patients
working_MESSIdata2021_08 <- read_dta("~/R_Projects/MESSI_Gene_Exp/ards pars model/working MESSIdata2021_08.dta")
head(working_MESSIdata2021_08)
```

```{r cleaning clinical data}
pts <- c(mod_join$messi_id)
pt_all <- working_MESSIdata2021_08 %>% 
  filter(pt %in% pts) %>% 
  select (pt, apacheIII, los, race, sex, age, hospMort30, source) %>%
  rename ("messi_id" = pt) %>%
  mutate(hospMort30 = factor(hospMort30, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(sex = factor(sex, levels = c(0, 1), 
                           labels = c("Female", "Male")))

# join the clinical variables with the model variables
pt_all <- left_join(pt_all, mod_join, by="messi_id")

# summarize the data to see what the patients look like
pt_all %>% select(-messi_id, -IL6_Probability, -IL8_Probability, -IL8, -IL6, -IL6_Model, -IL8_Model) %>% 
  tbl_summary(by=combined_models, missing="no") %>% 
  add_overall() %>%
  add_p() %>%
  modify_caption("Split by Dual Class Assignments")
```

**Notable patterns here -- high percentage of race 1 (= white) in the
hypoinflamm-IL6 group. Also we see that the patients
Hypoinflammatory-IL6 have a very high 90-d mortality.**

## Medical History
Next we examine medical history, to see if any clinical factors from a
patient's history correlates with classification and/or class switching. Medical comorbidities that are of interest include coronary disease, heart failure, liver failure/cirrhosis, and particularly, immunosuppression. We tend to have a high level of immune suppression in our ICU ARDS population, and we know from prior work that immunocompromise can affect the level of IL6 and IL8, and interestingly, can lead to relative over-expression of IL8, particularly in neutropenia. We will encode 'immunosuppression' per apacheiii guidelines, which includes hematologic malignancy, metastatic cancer, active chemotherapy, organ transplant (due to administration of immunosuppressant therapy) and use of high dose prednisone (>20mg/day).

```{r}
# first we have to encode immunosuppressed from the data provided, which includes the variables that contribute to immunosuppressed

pt_mh <- working_MESSIdata2021_08 %>% 
  filter(pt %in% pts) %>%
  rename("messi_id"=pt) %>%
  mutate(mhimmuno = ifelse((mhleuk == 1 | 
                            mhmyelo == 1 | 
                            mhlymp == 1 | 
                            mhmeta == 1 |
                            mhchemo == 1 |
                            cmimmuno == 1 |
                            mhorgtrans == 1	| 
                            cmhighpred == 1), 1, 0) , 
                 mhimmuno = ifelse(is.na(mhimmuno),0,mhimmuno) )
```

```{r}
table(pt_mh$mhimmuno)
table(pt_mh$mhimmuno, pt_all$combined_models)
```

There are 223 (47%) nonimmunosuppressed and 248 (53%) immunosuppressed patients in this cohort of ARDS patients.

```{r, echo=FALSE}
# now we will encode the other variables
pt_mh <- pt_mh %>%
  mutate(HTN = factor(mhhtn, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(Diabetes = factor(mhdm, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(CKD = factor(mhcri, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(Immunosuppressed = factor(mhimmuno, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(HF = factor(mhchf, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(CAD = factor(mhcad, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(Cirrhosis = factor(mhcirrhosis, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(LiverFailure = factor(mhence, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(COPD = factor(mhcopd, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  select(messi_id, HTN, Diabetes, CKD, Immunosuppressed, HF, CAD, Cirrhosis, LiverFailure, COPD)

c <- pt_all %>% select(messi_id, combined_models)
pt_mh <- left_join (pt_mh, c, by="messi_id")
# summarize
pt_mh %>% select(-messi_id) %>%
  tbl_summary(by=combined_models, missing = "no") %>%
  add_overall() %>%
  modify_caption("Medical Comorbidities") %>%
  add_p()

#join the clinical variables with the model variables pt_mh \<-
left_join(pt_mh, pt_all, by="messi_id")
#write.csv(pt_mh, file="ards_classifier_medhx.csv", row.names = F)
```

Here, we see the strongest signal in the Immunosuppressed population.

```{r}
table(pt_mh$Immunosuppressed, pt_mh$combined_models)

ic <- chisq.test(table(pt_mh$Immunosuppressed, pt_mh$combined_models))
ic
```

A chi-square test of independence was performed to examine class assignments and immunosuppression. The relation between these variables was significant, X2 (3, N=471) = 14.174, p=0.003).

Class assignment, taking into account both algorithms, is affected by immunosuppression (X2, df=3, p=0.003).

### Neutropenia
Next we evaluate MESSI ARDS patients in this cohort with and without
neutropenia (ANC \< 1), since, per prior publication Reilly et al.,
neutrophil count is inversely correlated with IL6 and IL8 levels. We
pull the neutrophil minumum count on study day 0 (enrollment). Since
there is some missingness for this, we also used day 1 and day -1 to
fill in the value from the 24hr period on either end of Study Day 0.

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
