---
title: "Evaluating Parsimonious Classifiers for ARDS Subphenotypes"
subtitle: "BMIN503 Final Project"
author: "Heather Giannini"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    theme: paper 
    highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r set-options, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(width = 400)
```

------------------------------------------------------------------------

## Overview

Acute respiratory distress syndrome (ARDS) is a high morbid condition of
respiratory failure, resulting from a myriad of insults, including
infection (most notably recently, COVID19), trauma, and other underlying
etiologies. We are investigating a recently published family of
parsimonious models for identifying subphenotypes in this disease.
Clinical trajectory of ARDS can be highly divergent , and yet we do not
have a clear understanding of the biological basis for vastly different
outcomes and clinical behavior. This project will compare the
assignments of two similar but slightly different biomarker-based
algorithms. The parsimonious models use a cytokine, interleukin-6 or
interleukin-8, both markers of a hyperinflammatory state, and couple
this measurement with the patient's level of soluble tumor necrosis
factor receptor 1 (sTNFR1) and bicarbonate level. We will apply both
IL6- and IL8- based models, evaluate how the population is categorized
into the two subphenotypes, "Hyperinflammatory" and "Hypoinflammatory"
and then correlate both clinical features and protein associations. This
work was done in consultation with Dr. Nuala Meyer (Pulmonary and
Critical Care) and Dr. Pratik Sinha (UCSF, Clinical Respiratory
Physiology and Biostatistics) and Dr.Todd Miano (PharmD, Dept of
Epidemiology). \>Nuala Meyer provides primary mentorship for this
project, providing expertise in interleukin dynamics of ARDS and
associated critical-illness syndromes like sepsis. She has published
previously on dynamics of IL6 and IL8 expression in conditions like
neutropenia and sepsis. \>Dr. Pratik Sinha derived these parsimonious
models after initial derivation of subphenotype classes (Hyper and
Hypoinflamm) via Latent Class analysis of ARDS population. He applied
these classifiers to our patient data to assign phenotypes. \>Dr. Todd
Miano is an expert in epidemiology and study design and provided
considerable expertise in coding variables and obtaining data from Penn
Data Store to encode bicarbonate for the preliminary variables that were
sent to Dr. Sinha for phenotype assignment.

Final Project GitHub Repository:
<https://github.com/gianninh/BMIN503_Final_Project>

## Introduction

Acute respiratory distress syndrome (ARDS) is a devastating pulmonary
condition with high patient-level variability that has extremely high
mortality and long-term morbidity in survivors. In clinical trials and
observational studies for ARDS, two clinical subphenotypes -
distinguished by plasma levels of inflammatory proteins and correlating
with 30-d mortality - have been consistently identified by latent class
analysis (LCA). These two classes are termed "Hyperinflammatory" and
"Hypoinflammatory". These phenotypes demonstrate heterogeneous treatment
response to multiple ARDS therapies, including steroids (in COVID19
ARDS). Although iterations of parsimonious models for assigning these
LCA classes based on biomarker combinations have similar operating
characteristics, their agreement at the patient level has not been
directly compared. We hypothesized that the performance of these models
may vary in the context of clinical factors that disproportionately
affect expression of the plasma proteins used for subphenoytype
assignment

This project will evaluate these biomarker-based Parsimonious Models for
ARDS subphenotypes. We will apply the models to our ARDS cohort, a
subcohort of the Molecular Epidemiology of Sepsis in the ICU (MESSI)
sepsis cohort. We will then test associations of these class assignments
with various clinical characteristics, including medical comorbidities,
biomarker protein levels, laboratory values (ie. white blood cell
count). Importantly, we will compare the assignments of two different
models, which may vary based on the biomarker used, either IL6 or IL8,
two similar cytokines that are typically congruent but may be
proportionally different in certain clinical circumstances.

## Methods

471 critically ill patients with ARDS were enrolled in the study within
24 hours of admission to the intensive care unit. Laboratory values and
clinical features were extracted from the EHR and input into a redcap
data repository. ARDS was defined per Berlin criteria. Plasma proteins
(IL6, IL8, sTNFR1) were measured by electrochemiluminescence on blood
obtained \<24hrs of ICU admission. We applied two parsimonious models
for LCA subphenotypes [IL6/bicarbonate/sTNFR1] or
[IL8/bicarbonate/sTNFR1] which return probability of assignment to
Hyperinflammatory class. Comorbidities were compared between predicted
model assignments using the Chi Square or Fisher's Exact tests.
Association of neutropenia (ANC \<1000) with class assignments for both
models was performed using Fisher's Exact test.

Derivation of Parsimonious Classifiers for ARDS
<https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7543720/> Development and
Validation of Parsimonious Algorithms to Classify ARDS Phenotypes:
Secondary Analyses of Randomised Controlled Trials

```{r, echo=F}
# load necessary libraries
library(tidyverse)
library(visdat)
library(naniar)
library(readr)
library(haven)
library(gtsummary)
library(readxl)
library(pheatmap)
library(Hmisc)
library(kableExtra)
library(summarytools)
library(cowplot)
library(dlookr)
library(ggpubr)
library(rstatix)
library(knitr)
```

### Comparing the IL6- and IL8- Model Assignments

First, we will look at descriptive characteristics of the IL6 and
IL8-based models for distribution of probabilities and IL6/IL8 levels.
Here we will also calculate an IL6:IL8 ratio, and test the association
of this ratio (continuous variable) with the categorical assignments.

The two datasets we are reading in contain measures for each MESSI
sample ID (each ID is a patient enrolled in the study), with
corresponding IL6, IL8, sTNFR1 and bicarbonate level on day 1. The two
datasets below also contain the subphenotype class assignments resulting
from use of either IL6- or IL8- based model. Finally, the probability of
assignment to "Hyperinflammatory" is also included.

```{r reading in datasets}
## Comparing Performance of IL6 vs IL8 models
# read in results from both models
il6mod <- read_excel("~/R_Projects/MESSI_Gene_Exp/ards pars model/ards_model2_calc_done.xlsx")
il8mod <- read_excel("~/R_Projects/MESSI_Gene_Exp/ards pars model/ards_model3_calc_done.xlsx")

head(il6mod)
```

We will need to simplify and merge these in order to compare how the
patients are classified by each model.

```{r data cleaning and organizing}

# remove bicarb/sTNFR1 from one dataset so they aren't duplicated in the join
il8mod <- il8mod %>% select(-Bicarbonate, -sTNFR1)

# combine the model 2 and model 3 data for one data table for analysis
# make a new variable called IL6:IL8 ratio so we can use this in our analysis later
mod_join <- left_join(il6mod, il8mod, by="messi_id")
mod_join <- mod_join %>% select(-Model.x, -Model.y) %>%
  rename("IL6_Model"=Class.x, "IL8_Model"=Class.y) %>%
  rename("IL6_Probability"=Probability.x, "IL8_Probability"=Probability.y) %>%
  mutate("IL6:IL8"=IL6/IL8)

# summarize both models assignments
mod_join %>% 
  select(IL6_Model, IL8_Model, IL6_Probability, IL8_Probability, IL6, IL8)%>% 
  tbl_summary(by=IL6_Model)%>%
  modify_caption("by IL6 Model Class Assignments")
```

We see that some of the class assignments are concordant, while some of
the class assignments are discordant between IL6 and IL8.

### Evaluating Discordance in Model Probabilities

We will use a heatmap representing the probabilities for each model to
demonstrate that while the models tend to agree most of the time, there
are some samples which clearly diverge depending on IL6 vs IL8 analyte.

```{r heatmap}
mat <- mod_join %>% select(IL6_Probability, IL8_Probability)
mat <- as.matrix(mat)
pheatmap(mat, cluster_cols = F)
```

We can see there is a group of patients at the top which were low
Probability in IL6, but clearly very high in IL8. And a group in the
middle which were low IL8, but high IL6.

### Cohort that demonstrates class switching

There are 98 patients that switch classes between IL6- and IL8-based
models.

```{r dual-class assignments for statistics}
# now make a new variables coding the dual-class assignments to evaluate groups
mod_join <- mod_join %>% 
  mutate(a = case_when(IL6_Probability >= 0.5 ~ "hyperIL6", IL6_Probability <= 0.5 ~ "hypoIL6"))%>% 
  mutate(b = case_when(IL8_Probability >= 0.5 ~ "hyperIL8", IL8_Probability <= 0.5 ~ "hypoIL8"))%>%
  unite("combined_models", a:b, remove = TRUE)

mod_join %>% select(-messi_id, -IL6_Model, -IL8_Model) %>% 
  tbl_summary(by="combined_models") %>%
  modify_caption("Both Model Assignments")

p1 <- ggplot(mod_join, aes(x=combined_models, y=IL6_Probability)) + geom_boxplot() + theme_bw()
p2 <- ggplot(mod_join, aes(x=combined_models, y=IL8_Probability)) + geom_boxplot() + theme_bw()
plot_grid(p1,p2)
```

It does appear that the probabilities - particularly for IL6 Model - are
less extreme, from 0.88 - 0.68 for Hyper and 0.05 - 0.22 for Hypo. IL8
probabilities mellow a bit as well, but maybe not as much.

The more interesting thing here is that the IL8 values are **higher** in
the hypoinflammatory group per IL6 model, which is likely why the IL8
Model tends to put them in hyperinflamm.

**So who are these patients who have relatively higher IL8, despite a
lower IL6 value?**

### Acute Clinical Variables

Next, we will bring in a dataset of clinical variables for each subject.
These reflect details about acute illness and demographics including:
**apacheIII, los, race, sex, age, hospMort30, mort90, sepsis source**
but has a lot of other variables that we will not be working with.

```{r clinical data fullset}
# load in full MESSI clinical dataset, this is over 3000 patients
working_MESSIdata2021_08 <- read_dta("~/R_Projects/MESSI_Gene_Exp/ards pars model/working MESSIdata2021_08.dta")
head(working_MESSIdata2021_08)
```

```{r cleaning clinical data}
pts <- c(mod_join$messi_id)
pt_all <- working_MESSIdata2021_08 %>% 
  filter(pt %in% pts) %>% 
  select (pt, apacheIII, los, race, sex, age, hospMort30, source) %>%
  rename ("messi_id" = pt) %>%
  mutate(hospMort30 = factor(hospMort30, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(sex = factor(sex, levels = c(0, 1), 
                           labels = c("Female", "Male")))

# join the clinical variables with the model variables
pt_all <- left_join(pt_all, mod_join, by="messi_id")

# summarize the data to see what the patients look like
pt_all %>% select(-messi_id, -IL6_Probability, -IL8_Probability, -IL8, -IL6, -IL6_Model, -IL8_Model) %>% 
  tbl_summary(by=combined_models, missing="no") %>% 
  add_overall() %>%
  add_p() %>%
  modify_caption("Split by Dual Class Assignments")
```

**Notable patterns here -- high percentage of race 1 (= white) in the
hypoinflamm-IL6 group. Also we see that the patients
Hypoinflammatory-IL6 have a very high 90-d mortality.**

### Medical History

Next we examine medical history, to see if any clinical factors from a
patient's history correlates with classification and/or class switching.
Medical comorbidities that are of interest include coronary disease,
heart failure, liver failure/cirrhosis, and particularly,
immunosuppression. We tend to have a high level of immune suppression in
our ICU ARDS population, and we know from prior work that
immunocompromise can affect the level of IL6 and IL8, and interestingly,
can lead to relative over-expression of IL8, particularly in
neutropenia. We will encode 'immunosuppression' per apacheiii
guidelines, which includes hematologic malignancy, metastatic cancer,
active chemotherapy, organ transplant (due to administration of
immunosuppressant therapy) and use of high dose prednisone (\>20mg/day).

```{r}
# first we have to encode immunosuppressed from the data provided, which includes the variables that contribute to immunosuppressed

pt_mh <- working_MESSIdata2021_08 %>% 
  filter(pt %in% pts) %>%
  rename("messi_id"=pt) %>%
  mutate(mhimmuno = ifelse((mhleuk == 1 | 
                            mhmyelo == 1 | 
                            mhlymp == 1 | 
                            mhmeta == 1 |
                            mhchemo == 1 |
                            cmimmuno == 1 |
                            mhorgtrans == 1	| 
                            cmhighpred == 1), 1, 0) , 
                 mhimmuno = ifelse(is.na(mhimmuno),0,mhimmuno) )
```

```{r}
table(pt_mh$mhimmuno)
table(pt_mh$mhimmuno, pt_all$combined_models)
```

There are 223 (47%) nonimmunosuppressed and 248 (53%) immunosuppressed
patients in this cohort of ARDS patients.

```{r, echo=FALSE}
# now we will encode the other variables
pt_mh <- pt_mh %>%
  mutate(HTN = factor(mhhtn, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(Diabetes = factor(mhdm, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(CKD = factor(mhcri, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(Immunosuppressed = factor(mhimmuno, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(HF = factor(mhchf, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(CAD = factor(mhcad, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(Cirrhosis = factor(mhcirrhosis, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(LiverFailure = factor(mhence, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  mutate(COPD = factor(mhcopd, levels = c(0, 1), 
                           labels = c("No", "Yes"))) %>%
  select(messi_id, HTN, Diabetes, CKD, Immunosuppressed, HF, CAD, Cirrhosis, LiverFailure, COPD)

c <- pt_all %>% select(messi_id, combined_models)
pt_mh <- left_join (pt_mh, c, by="messi_id")
# summarize
pt_mh %>% select(-messi_id) %>%
  tbl_summary(by=combined_models, missing = "no") %>%
  add_overall() %>%
  modify_caption("Medical Comorbidities") %>%
  add_p()

#join the clinical variables with the model variables pt_mh \<-
left_join(pt_mh, pt_all, by="messi_id")
#write.csv(pt_mh, file="ards_classifier_medhx.csv", row.names = F)
```

Here, we see the strongest signal in the Immunosuppressed population.

```{r}
table(pt_mh$Immunosuppressed, pt_mh$combined_models)

ic <- chisq.test(table(pt_mh$Immunosuppressed, pt_mh$combined_models))
ic
```

A chi-square test of independence was performed to examine class
assignments and immunosuppression. The relation between these variables
was significant, X2 (3, N=471) = 14.174, p=0.003).

Class assignment, taking into account both algorithms, is affected by
immunosuppression (X2, df=3, p=0.003).

### Neutropenia

Prior research by our group has shown that neutrophil count is inversely
correlated with IL6 and IL8 levels. Therefore, we hypothesize that
neutropenia may play a role in how immunosuppression contributes to
variability in these biomarker-based ARDS classifiers. We evaluate MESSI
ARDS patients in this cohort by stratifying them with and without
neutropenia (defined as ANC \< 1000). I will load in lab data collated
previously which includes the value for Minimum Neutrophil Count for
each study subject on the day of study enrollment.

```{r, echo=FALSE}
# load in cleaned neutrophil data
neutrophilMin <- read_csv("~/R_Projects/MESSI_Gene_Exp/ards pars model/neutrophilMin_clean.csv")
pt_all <- left_join(pt_all, neutrophilMin)

# create a new column called "Neutropenia" for Yes/No dichotomous variable if ANC > 1 (implies 1000)
pt_all <- pt_all %>%
  mutate(neutropenia = case_when(neutrosmin < 1 ~ "Neutropenic",
                        neutrosmin >= 1  ~ "Non-Neutropenic"))
# factor it for analysis
pt_all$neutropenia <- as.factor(pt_all$neutropenia)
# crossTab
table(pt_all$neutropenia)
```

There are 99 neutropenic and 358 non-neutropenic patients in our ARDS
cohort.

```{r, echo=F}
table(pt_all$neutropenia, pt_all$combined_models)
```

```{r, echo=F}
plot_data <- pt_all %>% select(neutropenia, IL6, IL8) %>%
  filter(!is.na(neutropenia))
a <- ggplot(plot_data, aes(x=neutropenia, y=IL6)) + 
  geom_boxplot(outlier.shape = NA) +
  ylim(0,2000)
b <- ggplot(plot_data, aes(x=neutropenia, y=IL8)) + 
  geom_boxplot(outlier.shape = NA) +
  ylim(0,2000)
plot_grid(a,b, labels="auto")
```

```{r}
il8neut <- plot_data %>% filter(neutropenia=="Neutropenic")
il8non_neut <- plot_data %>% filter(neutropenia == "Non-Neutropenic")

res <- wilcox.test(il8neut$IL8, il8non_neut$IL8)
res

res2 <- wilcox.test(il8neut$IL6, il8non_neut$IL6)
res2
```

```{r}
# stats --- categorical outcome (a,b,c,d) vs dichotomous variables (neutropenic or not)
cor.test(pt_all$IL6, pt_all$neutrosmin, method="spearman")
```

There is a significant weak negative correlation between Neutrophil
count and IL6 (-0.3, p-value = 1e-10)

```{r}
cor.test(pt_all$IL8, pt_all$neutrosmin, method="spearman")
```

There is a significant moderate negative association between Neutrophil
count and IL8 level (-0.57, p-value \< 2.2e-16)

```{r}
ggplot(pt_all, aes(x=combined_models, y=neutrosmin)) + geom_boxplot(outlier.shape = NA) + theme_bw() + ylim(0,35)

table(pt_all$neutropenia, pt_all$combined_models)
#can't use ChiSq b/c small group representation present in crossTab
fisher.test(table(pt_all$neutropenia, pt_all$combined_models))
```

## Results

For 471 patients with ARDS, the IL6 model classified 198 (42%) as
Hyperinflammatory whereas the IL8 model classified 258 (55%) as
Hyperinflammatory. Concordant Hyperinflammatory classes had higher
mortality than concordant Hypoinflammatory (p=0.040). There were 98
patients (21%) who switched classes. Among comorbidities,
immunosuppression and neutropenia (ANC\<1000) were statistically
associated with sub-class assignments (p=0.003 and p \<0.001). More
specifically, the switch from Hypoinflammatory IL6 to Hyperinflammatory
IL8 was significantly enriched with patients with immunosuppression (p
\< 0.01) and neutropenia (p \< 0.01). In this ARDS population, IL6 and
IL8 are inversely correlated with absolute neutrophil count (IL6: r =
-0.29, p \< 0.001; IL8: r = -0.57, p \< 0.001), which has been shown
prior in sepsis populations. Both cytokine levels were significantly
higher in the neutropenic compared to non-neutropenic patients (both p
\< 0.001). We could not assess each individual model performance because
they were based on latent class analysis and we do not have the LCA
assignments available in this study.

Immunosuppression, and particularly neutropenia, appear to impact
subphenotyping of ARDS patients when using parsimonious IL6 and
IL8-based classifiers for subphenotype assignment. Plasma protein-based
classifiers may be sensitive to confounders that influence expression of
these plasma biomarkers. Recognizing and describing limitations and
unique populations is a necessary step for future use and application of
cytokine-based subphenotype classifiers in ARDS.
