---
title: "BMIN503/EPID600 Project"
author: "Xingyue Zhu"
output: 
  html_document:
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

required packages
```{r}
library(ggplot2)
library(RColorBrewer)
```

### Overview


### Introduction 
Despite laws banning teenagers from drinking alcohol until they are adults, teenage binge drinking is still a common problem. This study was designed to see if drinking alcohol among teenagers had any adverse effects on them.

The data were obtained in a survey of students math and Portuguese language courses in secondary school. It contains a lot of interesting social, gender and study information about students, especially their alcohol consumption. I would like to use it for finding out the most important factor that related to the students final grade. To see if teen binge drinking has a negative impact on their learning performance. And create a prediction model to predict the student's final score. This problem is related to education, statistic and computer science. Solving it needs a clear and thorough understanding about semantics, regression model as well as solid programming skills. I will apply multiple regression models that we learned in this course and rigorous data pre-processing, multifaceted validation to get a persuasive result.


### Methods
```{r}
#Read in data
Math <- read.table("raw data/student-mat.csv", sep = ",", header = TRUE)
Portuguese <- read.table("raw data/student-por.csv", sep = ",", header = TRUE)


```




### Results

