---
title: "BMIN503/EPID600 Project Template"
author: Kira Nightingale
output: 
  html_document:
    theme: paper 
    highlight: tango
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

### Overview
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

The goal of this project is to assess whether adult dietary habits have an impact on oral health in adults. Data from NHANES years 2015-2016 and 2017-2018 will be analyzed, including the demographic survey, dental examination, and dietary survey. For this project, I was advised by Dr. Blanca Himes (Penn Medicine, informatics), Dr. Kevin Lynch (Penn Medicine, biostatistics), and Jana Yablonski (Penn Dental Medicine, dental hygiene). The project can be found via this GitHub link: https://github.com/kira-nightingale/BMIN503_Final_Project 

### Introduction 
Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

The 2017 Global Burden of Disease Study estimated that approximately 3.5 billion individuals worldwide suffer from oral conditions including carious lesions and periodontal disease, resulting in 18.3 millions years lived with disability globally (1). A study on the economic impact of oral diseases found that the direct and indirect costs associated with poor oral health result in global losses of around $442 billion in 2010. Notably, the estimated indirect costs amounted to $144 billion, a figure within the range of the indirect cost losses of the top ten global causes of death (ranging from $859 billion for cancer to $126 billion for lower respiratory infections) (2). As with all health conditions, prevention is far preferable to treatment. A 10-year follow-up on the 2007 WHO World Health Assembly resolution calling for a focus on oral health promotion and disease prevention (3) found that prevention programs were most frequently targeted at children, with few efforts aimed at improving the oral health of adolescents and adults (4).Therefore, determining how significant the role of diet is on oral health in adults could potentially provide an avenue for future adult-focused public health campaigns. 

### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

First, the necessary libraries were loaded. Data from the 2015-2016 and 2017-2018 years were loaded, and the instruments were combined across years to increase the sample size.
```{r}

#Loading libraries
library(nhanesA) 
library(SASxport)
library(plyr)
library(dplyr)
library(ggplot2)
library(table1)
library(pscl)

#Loading data - 2015-2016 and 2017-2018
demog15 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/15-16.DEMO_I.XPT")
exam15 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/15-16.OHXDEN_I.XPT")
#foodsd115 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/15-16.DR1IFF_I.XPT")
nutritiond115 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/15-16.DR1TOT_I.XPT")
nutritiond215 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/15-16.DR2TOT_I.XPT")

demog17 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/17-18.DEMO_J.XPT")
exam17 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/17-18.OHXDEN_J.XPT")
#foodsd117 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/17-18.DR1IFF_J.XPT")
nutritiond117 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/17-18.DR1TOT_J.XPT")
nutritiond217 <- read.xport("C:/Users/nighki01/Box Sync/Course Materials/2021 Fall_Biomedical Informatics/BMIN503_Final_Project/Data/17-18.DR2TOT_J.XPT")

#Combining instruments across years
demog <- full_join(demog15, demog17)
#foodsd1 <- full_join(foodsd115, foodsd117)
nutritiond1 <- full_join(nutritiond115, nutritiond117)
nutritiond2 <- full_join(nutritiond215, nutritiond217)
nutrition <- full_join(nutritiond1, nutritiond2)


```

Next, the demographic data was cleaned by computing new values, renaming variables, and removing unneeded variables. Individuals were assigned to age groups in 10 year increments.

```{r}

demog <- demog %>%
  select(SEQN, RIAGENDR, RIDAGEYR, RIDRETH3, DMDEDUC2, DMDEDUC3, DMDMARTL, INDHHIN2, INDFMPIR) %>%
  mutate(age.grp = ifelse(RIDAGEYR %in% 10:19, "10-19",
                    ifelse(RIDAGEYR %in% 20:29, "20-29",
                           ifelse(RIDAGEYR %in% 30:39, "30-39",
                                  ifelse(RIDAGEYR %in% 40:49, "40-49",
                                         ifelse(RIDAGEYR %in% 50:59, "50-59",
                                                ifelse(RIDAGEYR %in% 60:69, "60-69",
                                                       ifelse(RIDAGEYR %in% 70:79, "70-79",
                                                              ifelse(RIDAGEYR >79, "80+", "<10"))))))))) %>%
  rename(gender = RIAGENDR, race = RIDRETH3, child.edu = DMDEDUC3, adult.edu = DMDEDUC2, marital = DMDMARTL, income = INDHHIN2, poverty = INDFMPIR)


```  

The Decayed, Missing, Filled Tooth (DMFT) score is a commonly used indicator of tooth health used in dental research. In adults, all teeth except for third molars are assessed for caries, fillings, and missingness due to caries, and assigned a score of either 1 (presence of the aforementioned criteria) or 0 (sound, or missing due to other causes). The DMFT score is a summation of the 1 and 0 values for the full mouth. NHANES examiners assigned each tooth a letter value indicating its status. These letter values were re-coded to either 0 or 1 so that a DMFT score could be calculated for each participant. Root tips were not included as part of this calculation.

```{r}

#Dropping rows with incomplete exams
exam15 <- subset(exam15, OHDEXSTS!=3 & OHDEXSTS!=2)
exam17 <- subset(exam17, OHDEXSTS!=3 & OHDEXSTS!=2)

#Creating new variables with 0 or 1 values
for (i in 37:64) {
  newname <- paste0(names(exam15)[i], "new")
  exam15 <- mutate(exam15, !!newname := ifelse(!!as.name(names(exam15)[i]) %in% c("E", "F", "K", "P", "R", "Z", "A"), 1, 0))
}

for (i in 37:64) {
  newname <- paste0(names(exam17)[i], "new")
  exam17 <- mutate(exam17, !!newname := ifelse(!!as.name(names(exam17)[i]) %in% c("E", "F", "K", "P", "R", "Z", "A"), 1, 0))
}

#Calculating DMFT
exam15$dmft <- exam15 %>%
  select(171:198) %>%
  rowSums()

exam17$dmft <- exam17 %>%
  select(171:198) %>%
  rowSums()

#Selecting necessary variables and joining the datasets for the 2 years
exam15 <- select(exam15, SEQN, dmft)
exam17 <- select(exam17, SEQN, dmft)
exam <- full_join(exam15, exam17)

exam <- exam %>%
  mutate(dmft1 = ifelse(dmft>=1, 1, 0)) %>%
  mutate(dmft_sqrt = sqrt(dmft)) %>%
  mutate(dmft_log = log(dmft))

ggplot(data=exam, aes(dmft_log)) +
  geom_bar()

```

Nutrition data was obtained by NHANES through participant interviews. Participants were asked to report all foods eaten over a period of one or two days. Using an FDA database, NHANES personnel itemized comprehensive nutrition information for all foods eaten. For this project, the total values for relevant nutrients were used. First, participants with incomplete or not done nutrition profiles were removed. Then, the average quantity of nutrients was calculated for each participant. Finally, the amount of each nutrient consumed was compared to the Recommended Dietary Allowance (RDA) based on age and gender, and the applicable percentage was calculated.[5]
```{r}
#Removing participants with incomplete/not done nutrition profiles, and merging nutrition data with demographic information
data <-  full_join(demog, nutrition, by = "SEQN")
data <- subset(data, age.grp!="<10" & age.grp!="10-19")

data <- data %>%
  subset(DR1DRSTZ == 1 | DR2DRSTZ == 1) %>%
  select(SEQN, DRDINT, DR1TNUMF, DR2TNUMF, DR1TKCAL, DR1TPROT, DR1TCARB, DR1TSUGR, DR1TFIBE, DR1TTFAT, DR1TSFAT, DR1TCHOL, DR1TATOC, DR1TVARA, DR1TVB1, DR1TVB2, DR1TVB6, DR1TVB12, DR1TVC, DR1TVD, DR1TVK, DR1TCALC, DR1TIRON, DR1.300, DR1.320Z, DR2TKCAL, DR2TPROT, DR2TCARB, DR2TSUGR, DR2TFIBE, DR2TTFAT, DR2TSFAT, DR2TCHOL, DR2TATOC, DR2TVARA, DR2TVB1, DR2TVB2, DR2TVB6, DR2TVB12, DR2TVC, DR2TVD, DR2TVK, DR2TCALC, DR2TIRON, DR2.300, DR2.320Z, gender, race, child.edu, adult.edu, marital, income, poverty, age.grp)



#Averaging nutrient intake per participant
data$calories <- rowMeans(subset(data, select = c(DR1TKCAL, DR2TKCAL)), na.rm = TRUE)
data$protein <- rowMeans(subset(data, select = c(DR1TPROT, DR2TPROT)), na.rm = TRUE)
data$carbs <- rowMeans(subset(data, select = c(DR1TCARB, DR2TCARB)), na.rm = TRUE)
data$sugar <- rowMeans(subset(data, select = c(DR1TSUGR, DR2TSUGR)), na.rm = TRUE)
data$fiber <- rowMeans(subset(data, select = c(DR1TFIBE, DR2TFIBE)), na.rm = TRUE)
data$fat <- rowMeans(subset(data, select = c(DR1TTFAT, DR2TTFAT)), na.rm = TRUE)
data$satfat <- rowMeans(subset(data, select = c(DR1TSFAT, DR2TSFAT)), na.rm = TRUE)
data$cholesterol <- rowMeans(subset(data, select = c(DR1TCHOL, DR2TCHOL)), na.rm = TRUE)
data$vit.e <- rowMeans(subset(data, select = c(DR1TATOC, DR2TATOC)), na.rm = TRUE)
data$vit.a <- rowMeans(subset(data, select = c(DR1TVARA, DR2TVARA)), na.rm = TRUE)
data$vit.b1 <- rowMeans(subset(data, select = c(DR1TVB1, DR2TVB1)), na.rm = TRUE)
data$vit.b2 <- rowMeans(subset(data, select = c(DR1TVB2, DR2TVB2)), na.rm = TRUE)
data$vit.b6 <- rowMeans(subset(data, select = c(DR1TVB6, DR2TVB6)), na.rm = TRUE)
data$vit.b12 <- rowMeans(subset(data, select = c(DR1TVB12, DR2TVB12)), na.rm = TRUE)
data$vit.c <- rowMeans(subset(data, select = c(DR1TVC, DR2TVC)), na.rm = TRUE)
data$vit.d <- rowMeans(subset(data, select = c(DR1TVD, DR2TVD)), na.rm = TRUE)
data$vit.k <- rowMeans(subset(data, select = c(DR1TVK, DR2TVK)), na.rm = TRUE)
data$calcium <- rowMeans(subset(data, select = c(DR1TCALC, DR2TCALC)), na.rm = TRUE)
data$iron <- rowMeans(subset(data, select = c(DR1TIRON, DR2TIRON)), na.rm = TRUE)
data$water <- rowMeans(subset(data, select = c(DR1.320Z, DR2.320Z)), na.rm = TRUE)

data <- data %>%
  select(SEQN, DRDINT, DR1TNUMF, DR2TNUMF, DR1.300, DR2.300, calories, protein, carbs, sugar, fiber, fat, satfat, cholesterol, vit.e, vit.a, vit.b1, vit.b2, vit.b6, vit.b12, vit.c, vit.d, vit.k, calcium, iron, water, gender, race, child.edu, adult.edu, marital, income, poverty, age.grp)

#Calculating percent daily values based on age and gender
data <- data %>%
  mutate(p.calories = if_else(gender==1 & age.grp=="20-29", (calories/2400)*100, 
                             if_else(gender==2 & age.grp=="20-29", (calories/2000)*100,
                                    if_else(gender==1 & age.grp=="30-39" | age.grp=="40-49", (calories/2200)*100,
                                           if_else(gender==2 & age.grp=="30-39" | age.grp=="40-49", (calories/1800)*100, 
                                                  if_else(gender==1 & age.grp=="50-59" | age.grp=="60-69" | age.grp=="70-79" | age.grp=="80+", (calories/2000)*100, (calories/1600)*100)))))) %>%
  mutate(p.protein = if_else(gender==1, (protein/56)*100, (protein/56)*100)) %>%
  mutate(p.carbs = (carbs/130)*100) %>%
  mutate(p.sugar = ((sugar/calories)/0.1)*100) %>%
  mutate(p.fiber = if_else(gender==1 & age.grp=="20-29", (fiber/34)*100, 
                             if_else(gender==2 & age.grp=="20-29", (fiber/28)*100,
                                    if_else(gender==1 & age.grp=="30-39" | age.grp=="40-49", (fiber/31)*100,
                                           if_else(gender==2 & age.grp=="30-39" | age.grp=="40-49", (fiber/25)*100, 
                                                  if_else(gender==1 & age.grp=="50-59" | age.grp=="60-69" | age.grp=="70-79" | age.grp=="80+", (fiber/28)*100, (fiber/22)*100)))))) %>%
  mutate(p.fat = (fat/35)*100) %>%
  mutate(p.satfat = ((satfat/calories)/0.1)*100) %>%
  mutate(p.vit.e = (vit.e/15)*100) %>%
  mutate(p.vit.a = if_else(gender==1, (vit.a/900)*100, (vit.a/700)*100)) %>%
  mutate(p.vit.b6 = if_else(gender==1 & age.grp=="50-59" | age.grp=="60-69" | age.grp=="70-79" | age.grp=="80+", (vit.b6/1.7)*100,
                            if_else(gender==2 & age.grp=="50-59" | age.grp=="60-69" | age.grp=="70-79" | age.grp=="80+", (vit.b6/1.5)*100, (vit.b6/1.3)*100))) %>%
  mutate(p.vit.b12 = (vit.b12/2.4)*100) %>%
  mutate(p.vit.c = if_else(gender==1, (vit.c/90)*100, (vit.c/75)*100)) %>%
  mutate(p.vit.d = if_else(age.grp=="70-79" | age.grp=="80+", (vit.d/800)*100, (vit.d/600)*100)) %>%
  mutate(p.vit.k = if_else(gender==1, (vit.k/120)*100, (vit.k/90)*100)) %>%
  mutate(p.calcium = if_else(gender==1 & age.grp=="70-79" | age.grp=="80+", (calcium/1200)*100,
                             if_else(gender==2 & age.grp=="50-59" | age.grp=="60-69" | age.grp=="70-79" | age.grp=="80+", (calcium/1200)*100, (calcium/1000)*100))) %>%
  mutate(p.iron = if_else(gender==2 & age.grp=="20-29" | age.grp=="30-39" | age.grp=="40-49", (iron/18)*100, (iron/8)*100))

data$age.grp <- as.factor(data$age.grp)

```


```{r eval=FALSE}
#SAVING THIS STUFF FOR LATER JUST IN CASE
#This one works
exam15$OHX02CTC <- as.character(exam15$OHX02CTC)
exam15$OHX02CTC <- factor(exam15$OHX02CTC, levels = c("A", "D", "M", "Q", "S", "U", "X", "Y", "E", "F", "J", "K", "P", "R", "T", "Z"), labels = c(0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1, 1)) 

#Create new columns for variables %in% c("A", "B", ...) numeric

exam15<- mutate(exam15, ohx02 = function(x) {if (x %in% c("M", "Q", "S")){1} else {0}})
exam15<- mutate(exam15, ohx02 = ifelse(OHX02CTC %in% c("A", "D", "M", "Q", "S", "U", "X", "Y"), 1, 0))

for (i in 37:64) {
  print(i)
  newname <- paste0(names(exam15)[i], "new")
  print(newname)
  print(names(exam15)[i])
  exam15 <- mutate(exam15, !!newname := ifelse(!!as.name(names(exam15)[i]) %in% c("E", "F", "K", "P", "R", "Z", "A"), 1, 0))
}
```

Finally, all remaining data files were combined into a single data frame, and cases that included both nutrition data and DMFT score were included in the final data set. Methods for analysis, the applicable code, and outputs are included in the Results section below.
```{r}
#Merging data into a single data frame
data <-  join_all(list(data, exam), by = "SEQN", type = "full")
data <- data[!is.na(data$calories), ]
data <- data[!is.na(data$dmft), ]

```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

The below table shows the demographic distribution of the population, broken down by age group. The population includes a slightly greater number of females than males. Non-Hispanic whites made up 35% of the population, followed by non-Hispanic blacks (22.4%), Mexican Americans (15.3%), Asians (11.6%), non-Mexican Hispanic (11.2%), and other (4.3%).The mean income for the study population was approximately 2.52 times the national poverty level.

```{r}
#Labeling factor data and variable names
data$gender <- factor(data$gender, levels=c(1, 2), labels=c("Male", "Female"))
data$race <- factor(data$race, levels=c(3, 4, 6, 1, 2, 7), labels=c("White, Non-Hispanic", "Black, Non-Hispanic", "Asian, Non-Hispanic", "Mexican American", "Other Hispanic", "Other/Multi-Race"))
data$adult.edu <- factor(data$adult.edu, levels=c(1, 2, 3, 4, 5, 7, 9), labels=c("Less than 9th grade", "Some highschool", "Highschool/GED or equivalent", "Some college/Associates degree", "College graduate or above", "Refused", "Don't know"))
data$marital <- factor(data$marital, levels=c(1, 2, 3, 4, 5, 6, 77), labels=c("Married", "Widowed", "Divorced", "Separated", "Never married", "Living with partner", "Refused"))

label(data$age.grp) <- "Age Group"
label(data$gender) <- "Sex"
label(data$race) <- "Race"
label(data$adult.edu) <- "Education Level"
label(data$marital) <- "Marital Status"

#Creating a function to show mean (SD) for continuous variables in the table
my.render.cont <- function(x) {
    with(stats.apply.rounding(stats.default(x), digits=3), c("",
        "Mean (SD)"=sprintf("%s (&plusmn; %s)", MEAN, SD)))
}

#Generating Table 1
table1(~gender + race + adult.edu + poverty | age.grp, data=data, overall="Total", render.continuous=my.render.cont)


```

Histogram of DMFT score
```{r}
ggplot(data=exam, aes(dmft)) +
  geom_bar(fill="lightblue") +
  labs(title = "Decayed, Missing, Filled Tooth (DMFT) Score Frequency", x = "DMFT Score", y = "Count") +
  theme_bw()
```

Consumption of sugar and carbohydrates has been associated with increased caries risk due to the resulting change in oral pH that promotes cariogenic bacteria growth.[6] Conversely, inadequate amounts of protein and vitamin C have been linked to periodontal disease, which in turn may lead to increased dental decay due to exposure of the root surfaces. [7] Calcium and vitamin D have also been cited as important nutrients for tooth health, particularly in children. [8] Therefore, to investigate the effects of diet on DMFT score, nutrient information was limited to include total calories, sugar, carbohydrates, protein, vitamin C, calcium, and vitamin D. A summary of the average nutrient consumption by sex and age group can be found below. We can see that most age groups consume close to the RDV of calories and vitamin C on a daily basis. Consumption of carbohydrates and protein is generally higher than recommended, whereas average sugar consumption is surprisingly around half of the RDV. Notably, consumption of vitamin D across all age groups is negligible. This could be due to the fact that the primary foods containing vitamin D are fortified milk, juices, and cereals, and the majority of US adults receive the bulk of their vitamin D requirements from vitamin supplements [9], which were not accounted for in this analysis.

```{r}
label(data$p.calories) <- "% RDV Calories"
label(data$p.sugar) <- "% RDV Sugar"
label(data$p.carbs) <- "% RDV Carbs"
label(data$p.protein) <- "% RDV Protein"
label(data$p.vit.c) <- "% RDV Vitamin C"
label(data$p.vit.d) <- "% RDV Vitamin D"
label(data$p.calcium) <- "% RDV Calcium"
table1(~p.calories + p.sugar + p.carbs + p.protein + p.vit.c + p.calcium + p.vit.d | age.grp + gender, data=data, overall="Total")
```


To assess whether DMFT score is associated with diet, a zero inflated negative binomial regression was performed. This regression model was chosen because the mean DMFT score (11.32) was much lower than the variance of the DMFT score (65.15), and the data included more zeros than expected by the model. 
```{r}

model1 <- zeroinfl(dmft ~ p.calories + p.carbs + p.protein + p.vit.c + p.vit.d + p.sugar + p.calcium + age.grp + adult.edu + gender, data = data, dist = "negbin")
summary(model1)

```
The only nutrient significantly associated with having a DMFT score of 0 versus >0 was sugar, with older adults and males having a higher probability of having a DMFT score of >0. Among those with DMFT scores >0, consumption of sugar was significantly associated with having a greater number of caries, whereas consumption of consumption of vitamin C was significantly associated with having fewer caries. Interestingly, increased carbohydrate consumption was also associated with having fewer caries. There was no significant impact of calcium or vitamin D consumption on caries. This is likely due to the fact that calcium and vitamin D may have a larger impact on tooth health close to the time of eruption, and this study was limited to individuals aged 20 years and older. Older adults were also more likely to have a greater number of caries, which intuitively makes sense as DMFT score can only increase over time, never decrease. Women were significantly more likely than men to have a greater number of caries, a finding that is supported in the literature. [10] Individuals who completed college or higher were significantly more likely to have fewer caries, which is also consistent with existing literature. [11]


```{r}
#Maybe will break up into a logistic regression and linear regression ... start of code is here

glm <- glm(dmft1 ~ p.calories + p.carbs + p.protein + p.vit.c + p.vit.d + p.sugar + p.calcium + age.grp + adult.edu + gender, data = data, family=binomial())

summary(glm)
exp(coef(glm))

```


### References
  
    1. Bernabe E, Marcenes W, Hernandez CR, et al. Global, Regional, and National Levels and Trends in Burden of Oral Conditions from 1990 to 2017: A Systematic Analysis for the Global Burden of Disease 2017 Study. Journal of Dental Research. 2020;99(4):362-373. doi:10.1177/0022034520908533
    2. Listl S, Galloway J, Mossey PA, Marcenes W. Global Economic Impact of Dental Diseases. Journal of Dental Research. 2015;94(10):1355-1361. doi:10.1177/0022034515602879
    3. World Health Organization. World Health Assembly First Special Session: Sixtiety World Health Assembly. 2007. Retrieved from https://apps.who.int/gb/ebwha/pdf_files/WHASSA_WHA60-Rec1/E/WHASS1_WHA60REC1-en.pdf.
    4. Petersen, PE, Baez, RJ, Ogawa, H. Global application of oral disease prevention and health promotion as measured 10 years after the 2007 World Health Assembly statement on oral health. Community Dent Oral Epidemiol. 2020; 48: 338– 348. https://doi-org.proxy.library.upenn.edu/10.1111/cdoe.12538
    5. National Institutes of Health: Office of Dietary Supplements. Nutrient Recommendations: Dietary Reference Intakes. https://ods.od.nih.gov/HealthInformation/nutrientrecommendations.aspx
    6. Shailesh M. Gondivkar, Amol R. Gadbail, Rima S. Gondivkar, Sachin C. Sarode, Gargi S. Sarode, Shankargouda Patil, Kamran H. Awan. Nutrition and oral health. Disease-a-Month. 2019; 65(6): 147-154.
    7. Hujoel, PP, Lingström, P. Nutrition, dental caries and periodontal disease: a narrative review. J Clin Periodontol 2017; 44 (Suppl. 18): S79– S84. doi: 10.1111/jcpe.12672.
    8. Schroth, R.J., Levi, J.A., Sellers, E.A. et al. Vitamin D status of children with severe early childhood caries: a case–control study. BMC Pediatr 13, 174 (2013). https://doi.org/10.1186/1471-2431-13-174
    9. Moore, C., Murphy, M.M., Keast, D.R. and Holick, M.F., 2004. Vitamin D intake in the United States. Journal of the American Dietetic Association, 104(6): 980-983.
    10. Lukacs, J.R. and Largaespada, L.L. (2006), Explaining sex differences in dental caries prevalence: Saliva, hormones, and “life-history” etiologies. Am. J. Hum. Biol., 18: 540-555. https://doi-org.proxy.library.upenn.edu/10.1002/ajhb.20530
    11. Northridge, M.E., Ue, F.V., Borrell, L.N., De La Cruz, L.D., Chakraborty, B., Bodnar, S., Marshall, S. and Lamster, I.B. (2012), Tooth loss and dental caries in community-dwelling older adults in northern Manhattan. Gerodontology, 29: e464-e473. https://doi.org/10.1111/j.1741-2358.2011.00502.x
    
    
    
    11. Blostein, FA, Jansen, EC, Jones, AD, Marshall, TA, Foxman, B. Dietary patterns associated with dental caries in adults in the United States. Community Dent Oral Epidemiol. 2020; 48: 119– 129. https://doi.org/10.1111/cdoe.12509

