---
title: "BMIN503/EPID600 Project Template"
author: "Tuhina Srivastava"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---

```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

## **Overview**
This project aimed to spatially describe and statistically compare social determinants of health, COVID epidemiology, and election results to answer the question: Is there an association between county-level social determinants of health (ex. poverty rates, income), COVID-19 disease epidemiology (case and death rates), and recent election results (outcome for 2020 elections).
    
    I spoke with the following faculty:
    * Julia Lynch, PhD | Professor, Political Science, School of Arts and Sciences
    * John H. Holmes | Professor, Biostatistics and Epidemiology, Perelman School of Medicine
    
 Dr. Holmes advised that my question initially may have had a very obvious answer (conservative voting history may be associated with greater COVID cases and deaths). Dr. Lynch advised me to narrow my question and define it much more clearly, such as which sociodemographic factors I would use. Dr. Lynch also suggested that voter turnout might be an interesting metric that has not yet been looked at in this context. Furthermore, in order to go beyond a purely descriptive project, I could also assess if the socioeconomic status gradient in voting correlates with SES gradient in COVID epidemiology.
    
    [Link to final project GitHub repository](https://github.com/tuhinas/BMIN503_Final_Project)

## **Introduction**
\
    COVID-19 is an urgent and deadly public health crisis that has infected over 15.6 million Americans and claimed over 296,000 lives. Emerging evidence suggests that political partisanship has influenced individuals’ voluntary decisions to engage prevention practices such as distancing and mask-wearing (Grossman G, et.al., Proceedings of the National Academy of Sciences Sep 2020, 117 (39) 24144-24153; DOI: 10.1073/pnas.2007835117). Social determinants of health such as poverty and homelessness also have been linked to increased COVID-19 transmission (Abrams E, Szefler S. COVID-19 and the impact of social determinants of health. The Lancet Respiratory Medicine. 2020;8(7):659-661. doi:10.1016/s2213-2600(20)30234-4). Additionally, the COVID-19-associated death rates for Black and Latino people are 2-3 times higher than for white people, with these estimates rising further when adjusted for age.(Centers for Disease Control and Prevention, 2020) People of color, especially Black Americans, also have an elevated occupational risk of exposure to COVID-19 since they are more likely to be employed as “essential” workers, putting them in higher risk settings. These disparities span across all age groups and across rural to urban counties in numerous states. Underlying health conditions do not solely account for this, as inequities stemming from systemic racism can contribute to disease exposure, morbidity, and mortality at both the individual and neighborhood level. Existing data shows that areas with higher poverty rates, population density, and proportion of minority populations are associated with increased odds for infection. However, the 2020 Presidential election results compared alongside social determinants of health has not yet been evaluated in the context of COVID-19 epidemiology. Given the great partisan divide currently plaguing American politics and the catastrophic toll COVID-19 has had on Americans, such analyses are timely and important in understanding what populations are vulnerable to infection. Therefore, the objective of this analyses is to spatially describe and statistically compare social determinants of health. voting, and COVID-19 epidemiology.
    \
    \
    The problem described above is interdisciplinary because it looks at COVID-19 health outcomes in both a socioeconomic and political context. It is generally accepted that lower socioeconomic status is associated with lower civic engagement, and data has shown that higher levels of income inequality is associated with lower voter turnout (Galbraith, J., & Hale, J. (2008). State Income Inequality and Presidential Election Turnout and Outcomes. Social Science Quarterly, 89(4), 887-901. http://www.jstor.org/stable/42956351). Since COVID-19 transmission is both linked to socioeconomic status and influenced by personal behavior, interdisciplinary analyses using recent election results might help health officials better understand vulnerable populations in their communities. The partisan spin irrationally applied to mask-wearing practices and the lack of federal leadership has only amplified the partisan divide and amplified COVID-19 infections.


## **Methods**

### Data Sources

Publicly available data was used for all measures in this project. The county level Social Vulnerability Index (SVI) data, an index created by the Centers for Disease Control and Prevention (CDC) to evaluate a community’ ability to manage disaster response, was used for all socioeconomic status measures. The SVI uses data from 15 U.S. census variables on social factors in its creation. Although the SVI was designed for environmental disasters, it incorporates information about poverty, access to transportation, and population density, all which are relevant to infectious disease transmission. These SVI data are from 2018, the most recent publicly available SVI data set on the CDC website. Geographic county shapefiles were also obtained from the SVI data.
\
Pennsylvania county-level COVID-19 case and death data was extracted from OpenData Pennsylvania, a portal to which the Pennsylvania Department of Health (DOH) uploads its COVID-19 data. The data includes all reported COVID-19 cases and deaths until December 1,2020. Presidential election data for 2020, which included Donald Trump and Joseph Biden raw votes counts by county, was downloaded from Politico which was sourced from the Associated Press. These data were also as recent as December 1, 2020. I chose to limit my analyses to Pennsylvania since it was a critical swing state in the 2020 Presidential election and has clearly segregated voting outcomes by county.
\
### Analysis

Next, bivariate analyses using logistic regression were performed to determine associations between the dichotomized 2020 presidential election outcome (Trump vs. Biden) and all socioeconomic and COVID-19 (cases, deaths) predictors. Bivariate analyses using linear regression were also performed to determine associations between case and death rates and all socioeconomic and election outcome predictors. Case and death rates were chosen for this instead of cumulative/total counts to account for difference in population size between counties. Using a threshold of p = 0.15, variables were selected for inclusion in the final models. The threshold was selected as p = 0.15, a slightly more stringent than the commonly used p = 0.20 in the literature, to create a parsimonius model. Finally, multivariable logistic and linear regressions were used to look at associations between the 2020 presidential election outcome (Trump vs. Biden), COVID-19 case and death rates, all corresponding significant predictor variables to assess relationships. For the regression models, alpha = 0.05 was used to determine significance. 

###Installing/loading R packages
```{r}
library(dplyr)
library(tidyverse)
library(sf)
library(tidycensus)
library(leaflet)
library(kableExtra)
library(ggplot2)
library(cowplot)
library(gridExtra)
library(ggsn)
library(RColorBrewer)
library(viridis)
```

###Data Retrieval and Cleaning

####Importing CDC Social Vulnerability Index data
```{r}
#CDC Social Vulnerability Index Data
pa.svi.sf <- st_read("SVI2018_PENNSYLVANIA_county.shp")  # read shapefile as an sf object
class(pa.svi.sf)
pa.svi.geo <- st_geometry(pa.svi.sf)
#plot(pa.svi.geo) #checking geometry
head(pa.svi.sf)

#Cleaning SVI shapefile to include variables of interest
pa.svi.sf.clean <- pa.svi.sf %>% dplyr::select(STATE, COUNTY, FIPS, E_POV, E_UNEMP, E_PCI, E_MINRTY, EP_POV, EP_UNEMP, RPL_THEME1, EP_UNINSUR, Shape_STAr, Shape_STLe)
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Total.Poverty = E_POV) #Persons below poverty estimate, 2014-2018 ACS
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Total.Unemployment = E_UNEMP) #Civilian (age 16+) unemployed estimate, 2014-2018 ACS
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Income.Per.Capita = E_PCI) #Per capita income estimate, 2014-2018 ACS
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Total.Minority.Persons = E_MINRTY) #Minority (all persons except white, non-Hispanic) estimate, 2014-2018 ACS
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Percent.Poverty = EP_POV) #Percentage of persons below poverty estimate
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Percent.Unemployment = EP_UNEMP) #Unemployment Rate estimate
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Socioeconomic.Status = RPL_THEME1) #Socioeconomic Status consisting of percentile Below Poverty, Unemployed, Income, No High School Diploma - composite percentile based score
pa.svi.sf.clean <- rename(pa.svi.sf.clean, Percent.Uninsured = EP_UNINSUR) #Percentage Uninsured in the total civilian noninstitutionalized population estimate, 2014-2018 ACS
head(pa.svi.sf.clean)
pa.svi.geo.clean <- st_geometry(pa.svi.sf.clean)
#plot(pa.svi.geo.clean) #checking geometry
```
####Importing COVID-19 case and death data from OpenData Pennsylvania
```{r}
#COVID-19 cumulative case and death data by county by the PA Department of Health
covid19cases_PA <- read.csv("https://raw.githubusercontent.com/tuhinas/BMIN503_Final_Project/master/COVID-19_Aggregate_Cases_Current_Daily_County_Health.csv", header = TRUE, sep = ",")
head(covid19cases_PA)

cases_recent <- filter(covid19cases_PA, Date == "12/01/2020") %>% select(Jurisdiction, Date, Cumulative.cases, Cumulative.Case.Rate)
cases_recent <- rename(cases_recent, County.Name = Jurisdiction)
head(cases_recent)

covid19deaths_PA <- read.csv("https://raw.githubusercontent.com/tuhinas/BMIN503_Final_Project/master/COVID-19_Aggregate_Death_Data_Current_Daily_County_Health.csv", header = TRUE, sep = ",")
head(covid19deaths_PA)
deaths_recent <- filter(covid19deaths_PA, Date.of.Death == "12/01/2020") %>% select(County.Name, X2018.Population, Total.Deaths, Total.Death.Rate)
deaths_recent <- rename(deaths_recent, Population.2018 = X2018.Population)
head(deaths_recent)

covid.pa <- inner_join(cases_recent, deaths_recent, by = "County.Name") #joining cases and deaths based on county
head(covid.pa)

covid.pa <- rename(covid.pa, COUNTY = County.Name) #Renaming county variables for future inner join with other data
covid.pa$COUNTY[covid.pa$COUNTY == "Mckean"] <- "McKean" #Fixing misspelling in county name
```

####2020 Presidential Election Data - Pennsylvania
```{r}
#PA Election Data Sourced from Politico on 12/01/2020
pa.presidential.2020 <- read.csv("https://raw.githubusercontent.com/tuhinas/BMIN503_Final_Project/master/20201201_pa_results.csv", header = TRUE, sep = ",")
head(pa.presidential.2020)
pa.presidential.2020$Winner #Creating Binary winner outcome
pa.presidential.2020$Winner <- ifelse(pa.presidential.2020$biden.votes > pa.presidential.2020$trump.votes, "Biden", "Trump") #Generating winner per county
head(pa.presidential.2020)
```

####Creating Aggregate Data Set with Social Vulnerability Index, COVID-19 Cases and Deaths, and 2020 Presidential Election Results
```{r}
#Checking consistency of number of rows
nrow(pa.svi.sf.clean)
nrow(covid.pa) #this dataset has an extra row for all of Pennsylvania
nrow(pa.presidential.2020)
covid.pa.clean <- covid.pa[-c(4),] 
nrow(covid.pa.clean)
head(covid.pa.clean)
head(pa.presidential.2020)
pa.presidential.2020$COUNTY <- trimws(pa.presidential.2020$COUNTY, which = c("both"), whitespace = "[ \t\r\n]") #removing extra psace after some county names

#Creating aggregate data set with SVI, COVID-19, and 2020 Presidential election results
svi.covid <- inner_join(pa.svi.sf.clean, covid.pa.clean, by = "COUNTY")
svi.covid.geo <- st_geometry(svi.covid)
plot(svi.covid.geo)
head(svi.covid)

pa.full <- inner_join(svi.covid, pa.presidential.2020, by = "COUNTY")
pa.full.geo <- st_geometry(pa.full)
plot(pa.full.geo)
head(pa.full)

#Renaming Variables for clarity
pa.full <- rename(pa.full, Cumulative.Cases = Cumulative.cases)
pa.full <- rename(pa.full, Total.Biden.Votes = biden.votes)
pa.full <- rename(pa.full, Total.Biden.Percentage = biden.pct)
pa.full <- rename(pa.full, Total.Trump.Votes = trump.votes)
pa.full <- rename(pa.full, Total.Trump.Percentage = trump.pct)

#Factoring outcome for logistic regression
pa.full$pres.outcome #Creating Presidential outcome variable
pa.full$pres.outcome <- as.factor(ifelse(pa.full$Winner == "Biden", 1, 0)) #Biden = 1
head(pa.full)
```
### Logistic Regression
#### Bivariate Analyses for Variable Selection
```{r}
#We selected all variables not associated with geography and created a new data frame - logistic regression
pa.bivariate <- pa.full %>% dplyr::select(2, 4:11, 15:24, 26)
pa.bivariate <- as.data.frame(pa.bivariate)
pa.bivariate$pres.outcome <- as.factor(pa.bivariate$pres.outcome) #Making sure the outcome variable is a factor - Biden = 1
head(pa.bivariate)
biv.pa <- cbind(Variable=rep(NA,20),P.Value=rep(NA,20)) #Creating a new data frame for bivariate associations output (variable name and p-value)
#Run bivariate logistic regression with outcome and 1 predictor each time in the loop
for (i in 2:20) {
 glm.pa <- glm(pres.outcome ~ pa.bivariate[[i]], data = pa.bivariate, family = binomial(logit))
 biv.pa[i, "Variable"] <- names(pa.bivariate[i])
 biv.pa[i, "P.Value"] <- summary(glm.pa)$coefficients[2,4]
 }
biv.pa <- as.data.frame(biv.pa)
biv.pa$P.Value <- as.numeric(biv.pa$P.Value)
#Variable Selection using common cutoff of p = 0.15
biv.var.sel <- biv.pa %>% filter(P.Value<0.15)
biv.var.sel[,"P.Value"] <- round(as.numeric(biv.var.sel[,"P.Value"]),3)
biv.var.sel %>% kbl() %>% kable_classic_2(full_width = F) %>% add_header_above(c(" ", "Table 1. Logistic Bivariate Analyses Results (Outcome: Presidential Election Winner)" = 1))

#Linear Regression Variable Selection - Case Rate Outcome
biv.pa.linear <- cbind(Variable=rep(NA,20),P.Value=rep(NA,20)) #Creating a new data frame for bivariate associations output (variable name and p-value)
#Run bivariate logistic regression with outcome and 1 predictor each time in the loop
for (i in 2:20) {
 lm.pa <- lm(Cumulative.Case.Rate ~ pa.bivariate[[i]], data = pa.bivariate)
 biv.pa.linear[i, "Variable"] <- names(pa.bivariate[i])
 biv.pa.linear[i, "P.Value"] <- summary(lm.pa)$coefficients[2,4]
 }
biv.pa.linear <- as.data.frame(biv.pa.linear)
biv.pa.linear$P.Value <- as.numeric(biv.pa.linear$P.Value)
#Variable Selection using common cutoff of p = 0.15
biv.linear.varsel <- biv.pa.linear %>% filter(P.Value<0.15)
biv.linear.varsel[,"P.Value"] <- round(as.numeric(biv.linear.varsel[,"P.Value"]),3)
biv.linear.varsel %>% kbl() %>% kable_classic_2(full_width = F) %>% add_header_above(c(" ", "Table 2. Linear Bivariate Analyses Results (Outcome: Case Rate)" = 1))

#Linear Regression Variable Selection - Death Rate Outcome
biv.pa.lindeath <- cbind(Variable=rep(NA,20),P.Value=rep(NA,20)) #Creating a new data frame for bivariate associations output (variable name and p-value)
#Run bivariate logistic regression with outcome and 1 predictor each time in the loop
for (i in 2:20) {
 lmdeath.pa <- lm(Total.Death.Rate ~ pa.bivariate[[i]], data = pa.bivariate)
 biv.pa.lindeath[i, "Variable"] <- names(pa.bivariate[i])
 biv.pa.lindeath[i, "P.Value"] <- summary(lmdeath.pa)$coefficients[2,4]
 }
biv.pa.lindeath <- as.data.frame(biv.pa.lindeath)
biv.pa.lindeath$P.Value <- as.numeric(biv.pa.lindeath$P.Value)
#Variable Selection using common cutoff of p = 0.15
biv.lindeath.varsel <- biv.pa.lindeath %>% filter(P.Value<0.15)
biv.lindeath.varsel[,"P.Value"] <- round(as.numeric(biv.lindeath.varsel[,"P.Value"]),3)
biv.lindeath.varsel %>% kbl() %>% kable_classic_2(full_width = F) %>% add_header_above(c(" ", "Table 3. Linear Bivariate Analyses Results (Outcome: Death Rate)" = 1))
```


#### Regression Models
```{r}
#Creating a new data frame with the outcome and variables chosen by variable selection
logistic.pa <- as.data.frame(pa.full) %>% dplyr::select(Total.Poverty, Total.Unemployment, Income.Per.Capita, Total.Minority.Persons, Socioeconomic.Status, Cumulative.Cases, Cumulative.Case.Rate, Population.2018, Total.Deaths, Total.Death.Rate, Total.Biden.Votes, Total.Trump.Votes, pres.outcome)
logistic.pa <- as.data.frame(logistic.pa)
logistic.pa$pres.outcome <- as.factor(logistic.pa$pres.outcome) #Making sure the outcome variable is a factor - Biden = 1

#Creating a full multivariable logistic model
glm.mvlogistic <- glm(pres.outcome ~ Total.Poverty + Total.Unemployment + Income.Per.Capita + Total.Minority.Persons + Socioeconomic.Status +  Cumulative.Cases + Cumulative.Case.Rate + Population.2018 + Total.Deaths + Total.Death.Rate, data = logistic.pa, family = binomial(logit))
summary(glm.mvlogistic)
exp(coef(glm.mvlogistic))
confint(glm.mvlogistic) #Confidence intervals for fit
pval.results <- as.data.frame(coef(summary(glm.mvlogistic))[,4])
colnames(pval.results) <- "P.value"
multi.results <- as.data.frame(cbind(exp(cbind(OR = coef(glm.mvlogistic), CI = confint(glm.mvlogistic))), pval.results$P.value))  #OR, 95%CI, pvalue
multi.results <- rename(multi.results, P.value = V4)

#Logistic Regression with SES and Election
glm.sociologistic <- glm(pres.outcome ~ Total.Poverty + Total.Unemployment + Income.Per.Capita + Total.Minority.Persons + Socioeconomic.Status + Population.2018, data = logistic.pa, family = binomial(logit))
summary(glm.sociologistic)
exp(coef(glm.sociologistic))
confint(glm.sociologistic) #Confidence intervals for fit
pval.socioresults <- as.data.frame(coef(summary(glm.sociologistic))[,4])
colnames(pval.socioresults) <- "P.value"
socio.results <- as.data.frame(cbind(exp(cbind(OR = coef(glm.sociologistic), CI = confint(glm.sociologistic))), pval.results$P.value))  #OR, 95%CI, pvalue
socio.results <- rename(socio.results, P.value = V4)

#Linear Regression for case rate based on variables selected above
linear.pa <- as.data.frame(pa.full) %>% dplyr::select(Total.Poverty, Total.Unemployment, Income.Per.Capita, Total.Minority.Persons, Percent.Uninsured, Cumulative.Cases, Cumulative.Case.Rate, Population.2018, Total.Deaths, Total.Death.Rate, Total.Biden.Votes, Total.Trump.Votes, Winner)
lm.cases <- lm(Cumulative.Case.Rate ~ Total.Poverty + Total.Unemployment + Income.Per.Capita + Total.Minority.Persons + Percent.Uninsured + Winner, data = linear.pa)
summary.lm(lm.cases)
summary.lm.cases <- summary.lm(lm.cases)
names(lm.cases) #We can retrieve output from summary statistics for model fit
summary.lm.cases$adj.r.squared
confint(lm.cases) #Confidence intervals for fit
pval.lm.cases <- as.data.frame(coef(summary.lm.cases)[,4])
colnames(pval.lm.cases) <- "P.value"
lm.cases.results <- as.data.frame(cbind(Coefficients = coef(lm.cases), pval.lm.cases$P.value))  #OR, 95%CI, pvalue
lm.cases.results <- rename(lm.cases.results, P.value = V2)

#Linear Regression for death rate outcome based on variables selected above
lm.deaths <- lm(Cumulative.Case.Rate ~ Total.Poverty + Total.Unemployment + Income.Per.Capita + Total.Minority.Persons + Percent.Uninsured + Winner, data = linear.pa)
summary.lm(lm.deaths)
summary.lm.deaths <- summary.lm(lm.deaths)
names(lm.deaths) #We can retrieve output from summary statistics for model fit
summary.lm.deaths$adj.r.squared
confint(lm.deaths) #Confidence intervals for fit
pval.lm.deaths <- as.data.frame(coef(summary.lm.deaths)[,4])
colnames(pval.lm.deaths) <- "P.value"
lm.deaths.results <- as.data.frame(cbind(Coefficients = coef(lm.deaths), pval.lm.deaths$P.value))  #OR, 95%CI, pvalue
lm.deaths.results <- rename(lm.deaths.results, P.value = V2)
```
## **Results**
\
I mapped social determinants of health, COVID epidemiology, and election results and then created regression models to assess associations.

### Mapping Socioeconomic Factors, COVID-19 Epidemiology, and 2020 Presidential Election Results

I spatially mapped numerous socioeconomic factors from the Social Vulnerability Index, COVID-19 cases and deaths, and election results by county in one of the critical swing states in the 2020 Presidential election, Pennsylvania.

```{r, echo=FALSE}
#Plots
my_theme <- function() {
      theme_minimal() +                                  
      theme(axis.line = element_blank(),                 
            axis.text = element_blank(),                 
            axis.title = element_blank(),
            panel.grid = element_line(color = "white"),  
            legend.key.size = unit(0.8, "cm"),          
            legend.text = element_text(size = 16),       
            legend.title = element_text(size = 16),
            plot.title = element_text(size = 22))      
      }
    
    myPalette <- colorRampPalette(brewer.pal(9, "YlOrRd")) #Worse outcomes depicted as darker red
    myPalette2 <- colorRampPalette(brewer.pal(9, "RdYlBu")) #Higher values depicted as blue
    myPalette3 <- colorRampPalette(brewer.pal(9, "PuOr")) #Higher values depicted as purple
    myPalette4 <- colorRampPalette(brewer.pal(9, "RdPu")) #Higher values depicted as purple
    
#Total Poverty
total.pov.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Total.Poverty), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Total # of People Living Below Poverty Level") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "# of People", colours = myPalette(100))
total.pov.plot
```
\
The above map shows the total number of people living in poverty at the county level. The county with the highest number of people living in poverty is Philadelphia county (380,873 people) followed by Allegheny county (144,842 people). Those are also the two most populous counties in Pennsylvania (PA) with 1,584,138 and 1,218,452 people, respectively. Philadelphia county contains the city of Philadelphia, and Allegheny county contains the city of Pittsburgh, two large metropolitan centers in PA.
```{r, echo=FALSE}
#Percent Poverty
percent.pov.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Percent.Poverty), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Percent of People Living Below Poverty Level") + #theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "% of People in County", colours = myPalette(100))
percent.pov.plot
```
\
The above map shows the percent of people living in poverty at the county level. Philadelphia county is still the highest, at 10.2% unemployed, but many counties in Central and Eastern Pennsylvania such as Centre (18.4%), Fayette (17.7%), and Clinton (17.4%) county are high as well.
```{r, echo=FALSE}
#Total Unemployment
total.unemploy.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Total.Unemployment), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Total # of People (age 16+) Unemployed") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "# of People", colours = myPalette(100))
total.unemploy.plot
```
\
The above map shows the total number of people (age 16+) unemployed at the county level. The county with the highest number of people (age 16+) is Philadelphia county (78,170 people) followed by Allegheny county (34,903 people).
```{r, echo=FALSE}
#Percent Unemployment
percent.unemploy.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Percent.Unemployment), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Unemployment Rate") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "Rate (%)", colours = myPalette(100))
percent.unemploy.plot
```
\
The above map shows the percent of people (age 16+) unemployed at the county level. Philadelphia county is still the highest, at 24.9% living in poverty, but many counties in Northeast, Central, and Eastern Pennsylvania such as Sullivan (8.0%), Monroe (7.9%), and Pike (7.4%) counties are high as well.
```{r, echo=FALSE}
#Per Capita Income
income.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Income.Per.Capita), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Per Capita Income") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "Income ($)", colours = myPalette2(100))
income.plot
```
\
The above map shows the per capita income by county in PA. Chester, Montgomery, and Bucks counties, all which suburbs of Philadelphia, have the highest per capita income values. The lowest per capita income values are in Forest, Clearfield, and Clinton counties.
```{r, echo=FALSE}
#Minorities
minority.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Total.Minority.Persons), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Total # of Racial/Ethnic Minority Individuals", subtitle = "All persons excluding white, non-Hispanic people") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "# of People", colours = myPalette3(100))
minority.plot
```
\
The above map shows the estimate of total number of people who are minorities, or all persons except white, non-Hispanic, at the county level. The county with the highest estimated number of minorities is Philadelphia county (1,029,738 people).

```{r, echo=FALSE}
#Percent Uninsured
percent.uninsured.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Percent.Uninsured), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Percent Uninsured") + theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "% of People in County", colours = myPalette(100))
percent.uninsured.plot
```
\
The above map shows the estimate of the percent of people uninsured in the civilian non-institutionalized population at the county level. Mifflin (12.5%), Lancaster(12.0%), and Snyder (11.9%) counties have the highest percent uninsured.
```{r, echo=FALSE}
#Socioeconomic Status
ses.plot <- ggplot() + geom_sf(data = pa.svi.sf.clean, aes(fill = Socioeconomic.Status), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Composite Socioeconomic (SES) Vulnerability Score", subtitle = "Composite score based on percentile Below Poverty, Unemployed, Income, No High School Diploma ") + 
      #theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "SES Score", colours = myPalette2(100))
ses.plot
```
\
The above map shows the composite Socioeconomic Vulnerability Score from the SVI based on percentile of the following social factors: Below Poverty, Unemployed, Income, and No High School Diploma. Blue-shaded areas indicate greater relative socioeconomic vulnerability. Clearfield County in Central PA had the highest score, followed by Philadelphia, Fayette, and Forest counties.
```{r, echo=FALSE}
#COVID Cases
cases.plot <- ggplot() + geom_sf(data = pa.full, aes(fill = Cumulative.Cases), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Cumulative COVID-19 Cases by County") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "# of Cases", colours = myPalette4(100))
cases.plot
```
\
The above map shows the total number cases of COVID-19 at the county level. The county with the highest number of COVID-19 cases is Philadelphia county (63,719 cases) followed by Allegheny county (29,515 cases).
```{r, echo=FALSE}
#COVID Case Rate
caserate.plot <- ggplot() + geom_sf(data = pa.full, aes(fill = Cumulative.Case.Rate), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("COVID-19 Case Rate by County") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "Case Rate (per 100,000)", colours = myPalette4(100))
caserate.plot
```
\
The above map shows the cumulative case rate, or number of cumulative confirmed and probable cases reported through December 1,2020 per 100,000 population, of COVID-19 at the county level. The counties with the highest cumulative case rates are Mifflin, Philadelphia, and Centre counties with case rates 4026.2, 4022.3, and 3975.9 per 100,000, respectively.
```{r, echo=FALSE}
#COVID Deaths
deaths.plot <- ggplot() + geom_sf(data = pa.full, aes(fill = Total.Deaths), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("Cumulative COVID-19 Deaths by County") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "# of Deaths", colours = myPalette4(100))
deaths.plot
```
\
The above map shows the total number deaths of COVID-19 at the county level. The county with the highest number of COVID-19 deaths is Philadelphia county (2,065 cases) followed by Montgomery county (947 cases).
```{r, echo=FALSE}
#COVID Death Rate
deathrate.plot <- ggplot() + geom_sf(data = pa.full, aes(fill = Total.Death.Rate), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("COVID-19 Death Rate by County") + 
      theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_gradientn(name = "Death Rate (per 100,000)", colours = myPalette4(100))
deathrate.plot
```
\
The above map shows the cumulative case rate, or total COVID-19 deaths reported through December 1,2020 per 100,000 population, of COVID-19 at the county level. The county with the highest death rate is Northumberland county (149.31 per 100,000) followed by Delaware county (142.72 per 100,000).
```{r pressure, echo=FALSE}
#Presidential Election Results
ptx.colors <- c("Trump" = "red1", "Biden" = "blue1")
presidential.pa.plot <- ggplot() + geom_sf(data = pa.full, aes(fill = Winner), lwd = 0, color = "black", size = 3) +
      my_theme() + ggtitle("2020 Presidential Election Results") +
      theme(plot.title = element_text(hjust = 0.5)) +
      scale_fill_manual(values = ptx.colors)
presidential.pa.plot
```
\
Finally, the above map shows the 2020 Presidential election results by winner (as determined by raw total votes) of that county. A blue county denotes a Joseph Biden win while a red county denotes a Donald Trump win.\

####Logistic Regression Models
```{r}
#Printing results
multi.results %>% kbl() %>% kable_classic_2(full_width = F) %>% add_header_above(c(" ", "Table 4. Full Multivariable Logistic Regression Model Results" = 4))
```
The overall full multivariable logistic regression yielded no significant association between predictors and the outcome of the election (Biden), with wide confidence intervals all crossing 1 and most odds ratios close to 1 (Table 4, above).
```{r}
socio.results %>% kbl() %>% kable_classic_2(full_width = F) %>% add_header_above(c(" ", "Table 5. Multivariable Logistic Model (Outcome: Election | Predictors: Socioeconomic) Results" = 4))
```
I created a slightly more parsimonious model by removing COVID-19-related variables. The multivariable logistic regression assessing if any socioeconomic predictors were associated with election outcome yielded no significant association between predictors and the outcome of the election (Biden), with wide confidence intervals all crossing 1 and most odds ratios close to 1 (Table 5, above).

```{r}
lm.cases.results %>% kbl() %>% kable_classic_2(full_width = F) %>% add_header_above(c(" ", "Table 6. Linear Regression Model (Outcome: Cumulative Case Rate | Predictors: Socioecnomic and Election) Results" = 2))
```
```{r}
lm.deaths.results %>% kbl() %>% kable_classic_2(full_width = F) %>% add_header_above(c(" ", "Table 7. Linear Regression Model (Outcome: Total Death Rate | Predictors: Socioecnomic and Election) Results" = 2))
```

Based on multivariable linear regression models (Table 6 and 7, above), the percent of population uninsured seems to be the only significant predictor of both cumulative case rates (p = 0.005) and total death rates (p = 0.003). While both linear models are significant overall (p = 0.021 and p = 0.013, respectively), the models fits are poor, as is evidences by R-squared values below 0.5, and none of the other selected socioeconomic factors are significant predictors in the model.


## **Conclusions**

Philadelphia county seems to be one of the msot diverse counties in Pennsylvania. Philadelphia county also also has the highest amount and proportion of people unemployed and living in poverty. Many rural counties in PA have high poverty, unemployment, percent uninsured, and COVID case rates. Rural counties generally voted for Trump, while urban and suburban counties generally voted for Biden.

While many socioeconomic factors were significantly associated with election results, COVID-19 case rates, and COVID-19 death rates individually, when added to a multivariable model, only percent uninsured remaining a significant predictor for COVID-19 epidemiology at the county level.

It would be interesting to further refine variable selection and perform such analyses at a more granular geographic unit of analyses. One limitation of this study is the small sample size in the total number of counties (n = 67) in Pennsylvania, especially with a majority of those going to Trump. To increase sample size and power this study further, it would be interesting to replicate and expand these analyses to include other swing states (ex. Georgia, Michigan, Arizona, Nevada, and Wisconsin) and the whole country.