---
title: "BMIN503/EPID600 Project Template"
author: "Katie Sullivan"
output: 
  html_document:
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

### Overview
<ul style="list-style-type:circle">
# Overview
<li> This project aims to understand whether epigenetic changes contribute to a higher risk of type 2 diabetes and DKD in Pima Indians.</li>
<li> The objective of this project is analyze cytosine methylation changes (5mC) which exist in the Pima Indian population with diabetes, and compare these to individuals without diabetes. We also would like to investigate how these methylation changes relate to kidney outcomes. </li>
</ul> 

### Introduction
+ ***first paragraph***:  Pima Indians of Arizona have an extremely high prevalence of type 2 diabetes and Diabetic Kidney Diseasae (DKD) . Genetically related Pima Indians living in Mexico, whose lifestyle remains traditional, have a much lower prevalence of these morbidities. The differences in lifestyle indicate that environmental factors play an important role in disease origins and suggest involvement of epigenetic programming.

###Methods
<ul style="list-style-type:circle">
<li> 327 Pima Indians (205 women, 122 men) were selected from a longitudinal cohort, all had an eGFR greater than 60ml/min and ACR <300mg/g at baselinem ie absence of diabetic kidney disease. DNA methylation from peripheral blood leukocytes was analyzed on an Illumina Infinium HumanMethylation 450 Beadchip. Preprocessing and Quality Control will be performed using the SeSame package in R, Methylation changes will be transformed to M values. I plan to use the following covariates in my regression models: age, sex, duration of diabetes, mean blood pressure, HbA1c, genotype, batch, cell count and conversion efficiency. P-values will be corrected for multiple comparisons.




### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 
                                                                                           
```{r, eval=TRUE}

local({r <- getOption("repos")
       r["CRAN"] <- "http://cran.r-project.org"
       options(repos=r)})

```
                                                                                                                                            
```{r, eval=TRUE}
                                                                                                                                            
# install sesame package
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("sesame")

install.packages("dplyr")

```                                                                                                                                           

```{r,eval=TRUE}
install.packages("ggpubr")
install.packages("lmer")
```

```{r}
library(sesame)
library(ggplot2)
library(dplyr)
library(ggpubr)
library(FactoMineR)
library(factoextra)
#install packages needed. sesame will load and pre-process methylation files, ggplot for visualization
```

```{r, eval=FALSE}
install.packages("lme4",
   repos=c("http://lme4.r-forge.r-project.org/repos",
      getOption("repos")[["CRAN"]]))

```

```{r,eval=FALSE}
library(lme4)

```

```{r}
#CRITICAL: After a new installation, one needs to cache the associated annotation data using the following command. This needs to be done only once per SeSAMe installation.
#sesameDataCacheAll()
#This function caches the needed SeSAMe annotation for all the supported platforms. For our analysis we will use the platform HM450
sesameDataCache("HM450")
```

```{r}
tools::R_user_dir("ExperimentHub", which="cache")

```
```{r, eval=FALSE}
HM450.hg38.manifest <- sesameDataGet('HM450.hg38.manifest')
head(HM450.hg38.manifest)
```

```{r,eval=FALSE}
mft <- sesameDataGetAnno("HM450.hg38.manifest.tsv.gz")



```





```{r, eval=FALSE}
                                                                                            ## Import data
idat.dir <- "/home/sullka/idat"   #pmacs own directory
#Here idat.dir is the directory containing all the IDAT files. 
betas <- openSesame(idat.dir, quality.mask = FALSE)  #cannot run on own system due to computational space

```

Open Sesame is a pipeline available from the Sesame Package for loading beta values, see https://www.bioconductor.org/packages/release/bioc/vignettes/sesame/inst/doc/sesame.html#The_openSesame_Pipeline for more info

Open Sesame contains many in-built pre-processing steps. More can be found from this paper https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6237738/. It includes the pooBAH algorithm, which removes artefactual readings. It also ioncludes background subtraction with noob and non linear dye bias correction.

After obtaining the beta values we can carry out some quality control using the QC package within Sesame. This is again run on Pmacs due to the size of the file,

```{r, eval=FALSE}
idat_dir<-"/home/sullka/idat"
ssets<-lapply(searchIDATprefixes(idat_dir),readIDATpair)
sesameQC(ssets[[1]])
#Quality Control

=======================
=      Intensities    =
=======================
No. probes                      485577 
mean (M/U) (in-band InfI):      4985.92 
mean (M+U) (in-band InfI):      9971.84 

-- Infinium II --
No. probes:                    350076 (72.095%)
Mean Intensity:                 4645.133 

-- Infinium I (Red) -- 
No. probes:                     89203 (18.371%)
No. Probes Consistent Channel:  88957 
No. Porbes Swapped Channel:     67 
No. Probes Low Intensity:       179 
Mean Intensity (in-band):       6767.643 
Mean Intensity (out-of-band):   687.4307 

-- Infinium I (Grn) -- 
No. probes:                     46298 (9.535%)
No. Probes Consistent Channel:  45662 
No. Probes Swapped Channel:     551 
No. Probes Low Intensity:       85 
Mean Intensity (in-band):       4129.861 
Mean Intensity (out-of-band):   430.347 

=======================
=      Beta Values    =
=======================
No. probes:                     485577 
No. probes with NA:             67684 (13.939%)
Mean Betas:                     0.4860155 
Median Betas:                   0.5611176 

-- cg probes --
No. Probes:                     482421 
No. Probes with NA:             67063 (13.901%)
Mean Betas:                     0.4885846 
Median Betas:                   0.5704924 
% Unmethylated (Beta < 0.3):    42.167%
% Methylated (Beta > 0.7):      44.517%

-- ch probes --
No. Probes:                     3091 
No. Probes with NA:             621 (20.091%)
Mean Betas:                     0.05389193 
Median Betas:                   0.04912089 
% Unmethylated (Beta < 0.3):    100.000%
% Methylated (Beta > 0.7):      0.000%

-- rs probes --
No. Probes:                     65 
No. Probes with NA:             0 (0.000%)
Mean Betas:                     0.4901543 
Median Betas:                   0.4403832 
% Unmethylated (Beta < 0.3):    32.308%
% Methylated (Beta > 0.7):      35.385%

=======================
=      Inferences     =
=======================
Sex:                            FEMALE 
Ethnicity:                      WHITE 
Age:                            43.94463 
Bisulfite Conversion (GCT):     1.199467 


```

```{r, eval=FALSE}
   #perform inverse normalization of beta values 

for (i in 1:485577 {
betas2[i,]<-qnorm((rank(betas,na.last="keep")-0.5)/sum(!is.na(betas)))  #again ran in pmacs 
}
#reference https://www.biostars.org/p/80597/

#save object to a RDS file

saveRDS(betas2, file = "inversenormalizedbetamatrix.rds" 


```

```{r, eval=FALSE}
#read in manifest file
hm450 <- sesameDataGet("HM450.hg38.manifest")
#view manifest file
head(hm450)
str(hm450)
#convert GR ranges object into dataframe
hm450df<-as.data.frame(hm450)
str(hm450df)
#filter out sex_chromosomes as I will not be including these
hm450df_filtered = filter(hm450df, seqnames!= "chrX" & seqnames!="chrY")
# write hm450df_filtered to file
saveRDS(hm450df_filtered, file = "hm450df_filtered.rds")


```

I now have matrix of inverse normalized beta values produced by Sesame. In this matrix the rows are the CpG sites from the array and the columns are the names of the samples. In the manifest file ("hm450") the row names are also the CpG sites. We want to see which rows are overlapping from our inverse beta values rds file, with the manifest file, as this will allow us to annotate the CpG sites 
```{r, eval=FALSE}
inverse_beta_matrix = readRDS("inversenormalizedbetamatrix.rds")
# large file so need to load on PMACS
head(inverse_beta_matrix)
colnames(inverse_beta_matrix)
rownames(inverse_beta_matrix)
# First load manifest file and then convert row names into a vector
hm450df_filtered = readRDS("hm450df_filtered.rds")
cpg_sites_manifest<-as.vector(row.names(hm450df_filtered))
print(cpg_sites_manifest)

# with the variable inverse_beta_matrix_filtered, we now have the overlapping CpG sites
inverse_beta_matrix_filtered<-subset(inverse_beta_matrix, rownames(inverse_beta_matrix) %in% cpg_sites_manifest)
dim(inverse_beta_matrix_filtered)
# a total of 473914 CpGs are included for the final analysis
# Now save inverse beta matrix filtered to a rds file
saveRDS(inverse_beta_matrix_filtered, file = "inverse_beta_matrix_filtered.rds")

```
### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

```{r,eval=FALSE}
# Make Principal Component Analysis(PCA) plot
# I will use the following libraries : FactoMineR and factoextra
# To perform a PCA plot I first need to transform matrix (inverse_beta_matrix_filtered)


dat = readRDS("inverse_beta_matrix_filtered.rds")
dat = t(dat)
 
ncol(dat)
dat[,ncol(dat)]

dat.pca <- PCA(dat[,-ncol(dat)], graph = FALSE)

plot(dat.pca,choix="ind")

fviz_pca_ind(dat.pca,

             geom.ind = "point", # show points only (nbut not "text")

             palette = c("#ff0000", "#0000ff", '#CC00FF', '#FF0099', '#0000FF', '#008B8B'),

             addEllipses = TRUE, # Concentration ellipses # circle

)

ggsave('7.PCA_all_samples.png')


```


```{r,eval=TRUE}

# Phenotype data is stored in a file called blood327.csv. I will load this file and then view the phenotype charachteristics

pheno_data<-read.csv("/Users/katiesullivan/Documents/blood327_katie.csv")
print(head(pheno_data))
str(pheno_data)

```

```{r,eval=TRUE}
# Start formatting data set. First step: change 1 to male and 2 to female 
pheno_data<- pheno_data %>%
  mutate(sex = recode(sex, "1" = "male",
                             "2" = "female",
                            )
  )

```

```{r,eval=TRUE}
# Each column of the data set contains data. We wanted to make a summary of each column
summary_data <- sapply(pheno_data, summary)
```

```{r,eval=TRUE}
#Here I wanted to plot out some variables against each other. egfr stands for estimated glomrrular filtration rate, which is a marker of kidney health. hbA1c stands for glycosylated haemoglobin which is a measure of diabetes severity. 
plot_hba1c_egfr<-ggplot(pheno_data, aes(x = hba1c, y =egfr)) + 
  geom_point()

plot_hba1c_egfr

plot_age_egfr<-ggplot(pheno_data, aes(x = age, y =egfr)) + 
  geom_point()

plot_age_egfr

```

```{r,eval=TRUE}
# Understand if there is a signficant correlation between hba1c and egfr and egfr and age. 
#For this I will use the ggpubr package.
# hba1c, age and egfr are normally distributed so I will use pearson. 

cor.test(pheno_data$hba1c, pheno_data$egfr, method="pearson")
cor.test(pheno_data$age, pheno_data$egfr, method="pearson")
         
      
```

```{r, eval=TRUE}
# linear regression analysis preparation of clinical analysis file, labelled pheno_data
# read in csv file. Sample_ID is the ID of the different subject samples
pheno_data<-read.csv("/Users/katiesullivan/Documents/blood327_katie.csv")
head(pheno_data)
Sample_ID<- pheno_data[, c("Sample_ID")] 
Sample_ID
```



```{r, eval=TRUE}
# linear regression analysis preparation of the beta matrix file
getwd()
inverse_beta_matrix_filtered<-readRDS("~/Documents/R_Final_Project_Methylation/BMIN503_Final_Project/inverse_beta_matrix_filtered.rds")
SampleNameBeta<-as.vector(colnames(inverse_beta_matrix_filtered))
print(SampleNameBeta)
rownames(inverse_beta_matrix_filtered)
# Convert rds file into a data frame 
df_methylation=as.data.frame(inverse_beta_matrix_filtered)
#write.csv(df_methylation,"~/Documents/R_Final_Project_Methylation/BMIN503_Final_Project/inverse_beta_matrix_filtered.csv", row.names = FALSE)

```

```{r, eval=TRUE}
# Transpose metylation matrix data frame so columns are rows
df_methylation_new<-as.data.frame(t(df_methylation))

```

```{r, eval=TRUE}
# Filter out to include the samples you have clinical info for 
df_methylation_clinical_info<-df_methylation_new[Sample_ID,]
# save this to file
saveRDS(df_methylation_clinical_info, file ="df_methylation_clinical_info.rds")


```

```{r,eval=TRUE}
# reload df_methylation_clinical_info
df_methylation_clinical_info<-readRDS("df_methylation_clinical_info.rds")

```

```{r, eval=TRUE}

# transpose df methylation clinical info so that row names are probe names
df_methylation_clinical_info_transpose = t(df_methylation_clinical_info)
ProbeName <- row.names(df_methylation_clinical_info_transpose)
```

```{r, eval=TRUE}
#convert df-methylation_clinical_info_transpose into a matrix
beta <- as.matrix(df_methylation_clinical_info_transpose)
```

```{r, eval=TRUE}
# load text file which contains info on genotypes. This data was prepared by a previous lab member
### genotype PCs
geno <- read.table("genet_pcs.txt", header=T, row.names=1, sep=",")
row.names(geno) <- paste("x", row.names(geno), sep="")

```

```{r, eval=TRUE}
### phenotypes with slope data added to orginal file
        pd <- read.csv("clinical_phenotype3.csv", header=TRUE, row.names=1)
SampleName <- row.names(pd)
     pd$id <- paste("x", pd$id, sep="")
  pd$batch <- factor(pd$batch)
    pd$sex <- factor(pd$sex)
      geno <- geno[pd$id,]
   PheName <- names(pd)
### 316 of 318 have genotype data

```

```{r, eval=FALSE}
n=length(ProbeName)
res <- matrix(,n,3)
for (i in c(1:n)){
     fit1 <- lmer(beta[i,]~pd$batch+pd$CD8T+pd$CD4T+pd$NK+pd$Bcell+pd$Mono+pd$Gran+pd$converse)
       fit2 <- lm(pd$slope~residuals(fit1)+pd$age+pd$sex+geno$pc1+geno$pc2+geno$pc3+geno$pc4+geno$pc5+geno$pc6+geno$pc7+geno$pc8+geno$pc9+geno$pc10, weights=(1/pd$var))
    res[i,1] <- summary(fit2)$coefficients[2,1]
    res[i,2] <- summary(fit2)$coefficients[2,4]
}
```

I keep getting an error saying vari