---
title: "BMIN503/EPID600 Project"
author: "Yifei Duan"
output: 
  html_document:
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.
>This is the new template

### Overview

Goal: Use statistical and mathematical methods to analyze the COVID-19 spreading pattern in the past one year. 

Data used in this Project: Package("COVID19") / Package("covid19.analytics) / data[us_states]

Three faculties/Staff/Students: Mingjia Li: Candidate Ph.D in Perelman School of Medicine
                               Shufan Lin: Candidate Ph.D in Bioengineering Department
                               Bomyi Lim: Assistant Professor of Chemical and Biomolecular Engineering 
      They helped me found the data resources used in this project. And they gave me some ideas how to use the data to do the analysis.
      
Link to the repo: https://github.com/ddyifei/BMIN503_Final_Project
                


### Introduction 

Coronavirus disease (COVID-19) is an infectious disease caused by the SARS-CoV-2 virus. Most people infected with the virus will experience mild to moderate respiratory illness and recover without requiring special treatment. However, some will become seriously ill and require medical attention. Older people and those with underlying medical conditions like cardiovascular disease, diabetes, chronic respiratory disease, or cancer are more likely to develop serious illness. Anyone can get sick with COVID-19 and become seriously ill or die at any age. The COVID-19 pandemic is a formidable global public health challenge. Since the initial emergence of a novel coronavirus in late 2019, the spread of
SARS-CoV-2 has been unrelenting, impacting nearly every aspect of society worldwide. The pandemic has required a substantial response by public health authorities at all levels.

As the pandemic continues to unfold across the United States and world, the consequences of inadequate preparation and implementation are abundantly clear. Despite decades of various commissions highlighting the threat of global pandemics and international planning for their inevitability, neither the United States nor the broader international system were ready to execute those plans and respond to a severe pandemic. 

The COVID-19 pandemic is far from over and could yet evolve in unanticipated ways, but one of its most important lessons is already clear: preparation and early execution are essential in detecting, containing, and rapidly responding to and mitigating the spread of potentially dangerous emerging infectious diseases.
Analyzing the trend of COVID-19 is a possible method to prevent the new spreading of COVID-19 pandemic. Analysis the pattern of COVID-19 in the past one year in the U.S. could help make the prediction of COVID-19 in the future.Different statistical methods will be used in this project to analyze the spreading situation of COVID-19 in the U.S. Mathematical methods and coding will be used for this project.


### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 
```{r, eval = TRUE}
# install.packages("covid19.analytics")
```

```{r, eval = TRUE}
library("covid19.analytics")
library("dplyr")
library("spData")
library(sf)
```

In this project, I used COVID19 data provided by package("COVID19"). This package provides a daily summary of COVID-19 cases, deaths, recovered, tests, vaccinations, and hospitalizations for 230+ countries, 760+ regions, and 12000+ administrative divisions of lower level. Includes policy measures, mobility data, and geospatial identifiers. This package can be found on the website: COVID-19 Data Hub https://covid19datahub.io	
```{r, eval = TRUE}
# get the library of covid19 data
library("COVID19")
library(leaflet)
library(RColorBrewer)
library(ggplot2)
library(sf)
```

The only function in the package is covid19(). By default, the function downloads worldwide data by country. But, you can choose the data from specific countries. What I did was that I put "US" argument in the covid19() function to extract all the data from US. Another argument I used in this function is "level", which specifies the granularity of the data. In this case, level2 means the function can retrun state-level data. So, the code below will return the COVID19 data from US in a state-level.
```{r, eval = TRUE}
# get the covid19 data by US and for each state
covid19.us.state <- covid19(c("US"),level=2)
head(covid19.us.state)

```

"Confirmed" column and "administrative_area_level_2" columns were extracted from the COVID19 dataset. Column "confirmed" has the the total number of confirmed cases. "administrative_area_level_2" is the column which has the state name. And then the "administrative_area_level_2" renamed as "state" which is convenient to track the data.
```{r, eval = TRUE}

# get the date; confirmed cases and state column from the dataframe
covid19.confirmed.death <- covid19.us.state[, c("confirmed","administrative_area_level_2")]

# sort data
covid19.confirmed.death <- covid19.confirmed.death %>%
  arrange(administrative_area_level_2) %>%
  rename(state=administrative_area_level_2)
head(covid19.confirmed.death)
```

Created another vector called "state.name" which contains all the names of the US states. Used a for loop to iterate this vector -- state.name, and then filtered the whole dataset in which the state has the same name as in the for loop condition. Then used max to get the largest number of confirmed cases in each state. This number was stored in a vector called confirmed.number.
```{r, eval = TRUE}
# list of all State names
state.list <- state.name
# create a list which contain the data which show the highest number of confirmed cases in a day 
confirmed.number <- c()

# get the highest confirmed cases in each state
for(name in state.list)
{
  # get the confirmed cases data by each state
  confirmed.cases.state <- covid19.confirmed.death %>% filter(state == name)
  # get highest number of the cases
  highest.number.case <- max(na.omit(confirmed.cases.state$confirmed))
  # add the number in to the list
  confirmed.number <- c(confirmed.number, highest.number.case)
}

```

Next step was to create another data frame called "confirmed.cases.by.state" which contains the name of each state in one column and the largest number of confirmed cases in each state. Then extracted the geometry data of each state from "data(us_states)[NAME}" and did the inner join with data frame "confirmed.cases.by.state" by state name. 
```{r, eval = TRUE}
data(us_states)
# create a new dataset which only contains the confirmed case and state
confirmed.cases.by.state <- data.frame(state.list, confirmed.number)
head(confirmed.cases.by.state)

nrow(confirmed.cases.by.state)

confirmed.cases.by.state <- confirmed.cases.by.state %>%
  rename(NAME=state.list)
nrow(confirmed.cases.by.state)

head(confirmed.cases.by.state)

# get the geometry data in US
data(us_states)

# get the geometry data for each state
state.geo <- us_states["NAME"]
head(state.geo)

# add this geometry data to the covid data set
confirmed.cases.by.state <- inner_join(x=state.geo, y=confirmed.cases.by.state, by="NAME")
head(confirmed.cases.by.state)
```

Used the data frame which was inner joined with "data(us_states)["NAME"]", an interactive map was created to show the confirmed cases so far in each state in the US. A scale bar was shown below as well.
```{r, eval = TRUE}

pal_fun <- colorNumeric("RdBu", NULL)  

pu_message <- paste0("State: ", confirmed.cases.by.state$NAME, # paste0 to append tract name with other relevant text
"<br> Total Number of Confirmed Cases: ",confirmed.cases.by.state$confirmed.number)  # <br> forces new line

leaflet(confirmed.cases.by.state) %>%
  addPolygons(stroke = FALSE,                        # remove polygon borders
              fillColor = ~pal_fun(confirmed.number),
              fillOpacity = 1, smoothFactor = 1, # increase opacity and resolution
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
  addLegend("bottomright",                           # location of legend
            pal=pal_fun,                             # palette function
            values= ~confirmed.number,                 # variable to be passed to palette function
            title = 'Total number of confirmed cases',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()

```



"deaths" column and "administrative_area_level_2" columns were extracted from the COVID19 dataset. Column "deaths" has the the total number of deaths. "administrative_area_level_2" is the column which has the state name. And then the "administrative_area_level_2" renamed as "state" which is convenient to track the data.

```{r, eval = TRUE}

head(covid19.us.state)
covid19.deaths.by.state <- covid19.us.state[, c("deaths","administrative_area_level_2")]

# sort data
covid19.deaths.by.state <- covid19.deaths.by.state %>%
  arrange(administrative_area_level_2) %>%
  rename(state=administrative_area_level_2)
head(covid19.deaths.by.state)
```


Used another vector called "state.name" which was created above and this vector contains all the names of the US states. Used a for loop to iterate this vector -- state.name, and then filtered the whole dataset in which the state has the same name as in the for loop condition. Then used max to get the total number of deaths in each state. This number was stored in a vector called deaths.state.
```{r, eval = TRUE}
deaths.state <- c()

for(name in state.list)
{
  # get the confirmed cases data by each state
  data.by.state <- covid19.deaths.by.state %>% filter(state == name)
  deaths.number <- max(na.omit(data.by.state$deaths))

  deaths.state <- c(deaths.state, deaths.number)
}

total.deaths.by.state <- data.frame(state.list, deaths.state)
total.deaths.by.state <- total.deaths.by.state %>% rename(NAME = state.list)
total.deaths.by.state
```

Next step was to create another data frame called "total.deaths.by.state" which contains the name of each state in one column and the total number of deaths in each state. Then extracted the geometry data of each state from "data(us_states)[NAME}" and did the inner join with data frame "total.deaths.by.state" by state name. Used the data frame which was inner joined with "data(us_states)["NAME"]", an interactive map was created to show the total number of deaths so far in each state in the US. A scale bar was shown below as well.
```{r, eval = TRUE}
head(state.geo)
# inner join the state geometry data and deaths data
total.deaths.by.state <- inner_join(x=state.geo, y=total.deaths.by.state, by="NAME")
head(total.deaths.by.state)

pal_fun <- colorNumeric("RdPu", NULL)  

pu_message <- paste0("State: ", total.deaths.by.state$NAME, # paste0 to append tract name with other relevant text
"<br> Total Number of Deaths: ",total.deaths.by.state$deaths.state)  # <br> forces new line

leaflet(total.deaths.by.state) %>%
  addPolygons(stroke = FALSE,                        # remove polygon borders
              fillColor = ~pal_fun(deaths.state),
              fillOpacity = 1, smoothFactor = 1, # increase opacity and resolution
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
  addLegend("bottomright",                           # location of legend
            pal=pal_fun,                             # palette function
            values= ~deaths.state,                 # variable to be passed to palette function
            title = 'Total number of confirmed cases',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```

"tests" column and "administrative_area_level_2" columns were extracted from the COVID19 dataset. Column "tests" has the the total number of tests on each state. "administrative_area_level_2" is the column which has the state name. And then the "administrative_area_level_2" renamed as "state" which is convenient to track the data.
```{r, eval = TRUE}

covid19.tests.by.state <- covid19.us.state[, c("tests","administrative_area_level_2")]
head(covid19.tests.by.state)

covid19.tests.by.state <- covid19.tests.by.state %>%
  arrange(administrative_area_level_2) %>%
  rename(state=administrative_area_level_2)
head(covid19.tests.by.state)
```

Used another vector called "state.name" which was created above and this vector contains all the names of the US states. Used a for loop to iterate this vector -- state.name, and then filtered the whole dataset in which the state has the same name as in the for loop condition. Then used max to get the total number of tests in each state. This number was stored in a vector called tests.state.
```{r, eval = TRUE}
tests.state <- c()

for(name in state.list)
{
  # get the total number of tests data by each state
  tests.number.by.state <- covid19.tests.by.state %>% filter(state == name)
  total.tests.number <- max(na.omit(tests.number.by.state$tests))

  tests.state <- c(tests.state, total.tests.number)
}

total.tests.by.state <- data.frame(state.list, tests.state)
total.tests.by.state <- total.tests.by.state %>% rename(NAME = state.list, total.tests.number = tests.state)
head(total.tests.by.state)
```
Next step was to create another data frame called "total.tests.by.state" which contains the name of each state in one column and the total number of tests in each state. Then extracted the geometry data of each state from "data(us_states)[NAME}" and did the inner join with data frame "confirmed.cases.by.state" by state name. 

Used the data frame which was inner joined with "data(us_states)["NAME"]", an interactive map was created to show the total number of tests so far in each state in the US. A scale bar was shown below as well.
```{r, eval = TRUE}
total.tests.by.state <- inner_join(x=state.geo, y=total.tests.by.state, by="NAME")
head(total.deaths.by.state)

pal_fun <- colorNumeric("OrRd", NULL)  

pu_message <- paste0("State: ", total.tests.by.state$NAME, # paste0 to append tract name with other relevant text
"<br> Total Number of tests: ",total.tests.by.state$total.tests.number)  # <br> forces new line

leaflet(total.tests.by.state) %>%
  addPolygons(stroke = FALSE,                        # remove polygon borders
              fillColor = ~pal_fun(total.tests.number),
              fillOpacity = 1, smoothFactor = 1, # increase opacity and resolution
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
  addLegend("bottomright",                           # location of legend
            pal=pal_fun,                             # palette function
            values= ~total.tests.number,                 # variable to be passed to palette function
            title = 'Total number of confirmed cases',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```


"vaccines" column and "administrative_area_level_2" columns were extracted from the COVID19 dataset. Column "vaccines" has the the total number of people who are fully vaccinated in each state. "administrative_area_level_2" is the column which has the state name. And then the "administrative_area_level_2" renamed as "state" which is convenient to track the data.
```{r, eval = TRUE}

covid19.tests.by.state <- covid19.us.state[, c("vaccines","administrative_area_level_2")]
head(covid19.tests.by.state)

covid19.tests.by.state <- covid19.tests.by.state %>%
  arrange(administrative_area_level_2) %>%
  rename(state=administrative_area_level_2)
head(covid19.tests.by.state)
```

Used another vector called "state.name" which was created above and this vector contains all the names of the US states. Used a for loop to iterate this vector -- state.name, and then filtered the whole dataset in which the state has the same name as in the for loop condition. Then used max to get the total number of people who are fully vaccinated in each state. This number was stored in a vector called vaccines.state.
```{r, eval = TRUE}
vaccines.state <- c()

for(name in state.list)
{
  # get the total number of tests data by each state
  vaccines.number.by.state <- covid19.tests.by.state %>% filter(state == name)
  total.vaccines.number <- max(na.omit(vaccines.number.by.state$vaccines))

  vaccines.state <- c(vaccines.state, total.vaccines.number)
}

total.vaccines.by.state <- data.frame(state.list, vaccines.state)
total.vaccines.by.state <- total.vaccines.by.state %>% rename(NAME = state.list)
head(total.vaccines.by.state)
```

Next step was to create another data frame called "total.vaccines.by.state" which contains the name of each state in one column and the total number of people who are fully vaccinated in each state. Then extracted the geometry data of each state from "data(us_states)[NAME}" and did the inner join with data frame "total.vaccines.by.state" by state name. 

Used the data frame which was inner joined with "data(us_states)["NAME"]", an interactive map was created to show the total number of people who are fully vaccinated so far in each state in the US. A scale bar was shown below as well.
```{r, eval = TRUE}
total.vaccines.by.state <- inner_join(x=state.geo, y=total.vaccines.by.state, by="NAME")

pal_fun <- colorNumeric("Greens", NULL)  

pu_message <- paste0("State: ", total.vaccines.by.state$NAME, # paste0 to append tract name with other relevant text
"<br> Total Number of vaccinations: ",total.vaccines.by.state$vaccines.state)  # <br> forces new line

leaflet(total.vaccines.by.state) %>%
  addPolygons(stroke = FALSE,                        # remove polygon borders
              fillColor = ~pal_fun(vaccines.state),
              fillOpacity = 1, smoothFactor = 1, # increase opacity and resolution
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%   # add third party provider tile
  addLegend("bottomright",                           # location of legend
            pal=pal_fun,                             # palette function
            values= ~vaccines.state,                 # variable to be passed to palette function
            title = 'Total number of confirmed cases',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```

Another package I uased in this project is library("covid19.analytics").This package also provides worldwide COVID19 data. But this package includes lots of number of functions.
```{r, eval = TRUE}
library("covid19.analytics")
```

The function covid19.data() can return worldwide data. By using argument 'ts-confirmed', this covid19.data function will return the total number of confirmed cases.
```{r, eval = TRUE}
# get the data of confirmed cases globally from dataset
covid19.confirmed.cases <- covid19.data('ts-confirmed')
covid19.confirmed.cases_us <- covid19.confirmed.cases %>% 
  filter(Country.Region == "US")

# show the trend of increasing number of confirmed cases
tots.per.location(covid19.confirmed.cases_us)
```

The growth rate function will return the growth rate of the number of confirmed cases. And this function will return the growth pattern of confirmed cases in the future.
```{r, eval = TRUE}
# get the growth rate of cases in the US.(Total growth )
TS.data <- covid19.data("ts-confirmed")
growth.rate(TS.data, geo.loc="US")
```

### Results and Conclusion

Based on the output of interactive maps, it is obvious that New York, California, Texas and Florida are the four states which have the higher number of confirmed cases compared to other states in the US. The numbers of confirmed cases in these four states are around 300,0000. As for the deaths data, there are five states: New York, Florida, Texas, California and Pennsylvania which have the higher number of deaths compared to other states. The deaths numbers in these states are around 50000 - 70000. For the tests and vaccines data, the states which have the higher number of confirmed cases also have higher number of tests and people who are fully vaccinated. Those states are New York, Florida, Texas and California. Besides, Illinois also has the higher testing number. The result suggests that people in New York, California, Texas and Florida these states have higher risks to be infected.

The growth rate of confirmed cases in US graph shows that the growth rate kept increasing from the end of May to the end of September.And the peak occurred in the September. The number of cases changed graph shows that the curve is kind of flat during the first half year 2021, however, the curve shows the sharp increase during the second half of the year 2021.And the curve shows that it has the trend which will continue to increase next year.
