---
title: "Multiple micronutrient supplementation pilot intervention for pregnant women in Botswana"
subtitle: "BMIN503/EPID600 Final Project"
author: "Jingni Pan"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

## Overview {#sec-overview}

This project was a hybrid implementation-effectiveness pilot intervention of Multiple Micronutrient Supplementation (MMS) for pregnant women in Botswana. The primary goal was to assess the program's reach (penetration) and its effects on maternal hemoglobin levels and adverse birth outcomes using a Difference-in-Difference (DiD) analysis of data from the Tsepamo Study.Dr. Caniglia shared this idea and access to the dataset with me, and help me set up a rigorous analysis plan. From her, I learned the rigorous process of defining the intervention periods and using the non-intervention clinics as a comparison group to ensure the validity of the causal estimate when studying the effects of MMS strategies. Dr. Ruggieri helped narrow down to my current topic focusing on Botswana.

## Introduction {#sec-introduction}

Micronutrient deficiencies during pregnancy are a major public health issue, particularly in low- and middle-income countries (LMICs), where they contribute significantly to maternal anemia, poor fetal growth, and increased risk of adverse birth outcomes.1,2 According to the World Health Organization (WHO), an estimated 32 million pregnant women globally are affected by anemia, most of whom live in LMICs.3 Iron and folic acid supplementation (IFA) has long been the standard of care, yet it often fails to address the multiple concurrent micronutrient deficiencies that women face during pregnancy.

Multiple micronutrient supplementation (MMS), particularly the United Nations International Multiple Micronutrient Antenatal Preparation (UNIMMAP), has been recommended as a more effective alternative to IFA. MMS provides 15 essential vitamins and minerals, including iron, folic acid, zinc, vitamin A, and iodine.4 A systematic review of randomized controlled trials found that compared to IFA, MMS reduced the risk of low birthweight by 12%, preterm birth by 8%, and small for gestational age (SGA) births by 9%.2 MMS has also been shown to improve maternal hemoglobin levels, reduce maternal mortality in some contexts, and promote better long-term health outcomes in children.5 Moreover, recent cost-effectiveness analyses suggest that MMS offers greater health and economic benefits than IFA, with studies reporting positive benefit–cost ratios ranging from 2:1 to 23:1 depending on country context.6,7

Despite these promising findings, MMS is not yet widely adopted in sub-Saharan Africa. In Botswana, maternal and child health continue to face significant challenges. The country has a maternal mortality ratio estimated at 155 deaths per 100,000 live births, and nearly 50% of pregnant women register for antenatal care (ANC) after the first trimester, missing the opportunity for early intervention.8,9 An HIV prevalence of 24% among pregnant women further compounds the risks of anemia and poor pregnancy outcomes.10 A recent large-scale cohort study in Botswana involving over 96,000 women found that MMS users had lower risks of preterm birth, very preterm birth, and low birth weight compared to those on IFA, with the strongest link seen among women with HIV.11 Further, this study found that fewer than 60% of pregnant women in Botswana received guideline-concordant supplementation. Qualitative interviews conducted with pregnant women identified important barriers to IFA, such as stockouts, insufficient education, side effects, poverty, pill burden, and poor adherence, which may similarly affect MMS uptake.12 Taken together, this evidence suggests that MMS may outperform both IFA and iron-only supplementation in Botswana’s clinic-based ANC system, while also underscoring the need to address structural barriers like supplement stockouts, education, and delayed ANC registration.

In light of the efficacy of MMS, well documented in clinical trials, the WHO has updated its guidelines to now recommend MMS “in the context of rigorous research.” WHO and UNICEF specifically call for countries considering MMS programs to generate rigorous implementation research to understand the acceptability, feasibility, equity, and sustainability of switching from IFA to MMS.13 Despite studies in Bangladesh, Indonesia, and Mali, no implementation study has yet been conducted in Botswana. 14–16 This highlights the importance of generating evidence specific to Botswana’s health system and maternal population.

To address these gaps, we conducted a hybrid implementation-effectiveness pilot intervention of providing MMS to ANCs in Botswana. Specifically, we assessed the reach, feasibility, acceptability, and appropriateness of providing MMS to pregnant women through routine ANC services at clinics across three urban, rural, and semi-urban regions in Botswana and measured its effect on maternal hemoglobin levels and adverse birth outcomes through linkage with a nationwide birth outcomes surveillance study (The Tsepamo Study). In doing this, we hope to prove the necessity of transitioning the policy recommendation from IFA to MMS.

## Methods {#sec-methods}

Describe the data used and general methodological approach used to address the problem described in the @sec-introduction. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

###Linkage to Tsepamo Study 

To assess the primary implementation outcome and all effectiveness outcomes of interest, we leveraged individual-level clinical data collected in the Tsepamo study, which has been conducting birth outcomes surveillance in Botswana since 2014. The Tsepamo study captures delivery data from government hospitals throughout Botswana, including hospitals in Gaborone, Molepolole, and Palapye, and covers up to 72% of births nationwide. Data are abstracted from maternity obstetric records at the time of delivery, including ANC clinic and visits, diagnoses, medications (including MMS), HIV status, and birth outcomes. We used the ANC clinic and date of the first ANC clinic visit (the booking date) recorded in Tsepamo to link our MMS intervention data with the Tsepamo database. To minimize the potential for oversampling shorter pregnancies and to ensure all individuals included in the analysis had the potential to be pregnant for 40 weeks, we restricted our analysis to individuals with an estimated start of pregnancy at least 40 weeks before the day we abstracted the data from Tsepamo (Sep 3, 2025).

```{r}
# Load all neceaary packages
install.packages("haven")
library(haven)
install.packages("dplyr")
library(dplyr)
install.packages("lubridate")
library(lubridate)
install.packages("tidyverse")
library(tidyverse)

#Read a SAS dataset (.sas7bdat)
tsepamo <- read_sas("~/Desktop/tsepamo_updated_5sep25.sas7bdat")
# View the data
print(tsepamo)

# Only keep data from the 13 intervention clinics
mms <- tsepamo %>%         #=intervclinics
  filter(
    cliniccode %in% c(5302, 5303, 5323, 15312, 3319, 15325, 5306, 
                      15331, 15341, 15329, 3201, 5318, 15335)
  )

# Separately, only keep data from the non-intervention clinics (all other clinics in gabs, moleps, palapye
# Creating non-intervention clinics dataset
nonintervclinics <- tsepamo %>%
  mutate(flagtokeep = if_else((cliniccode >= 5000 & cliniccode < 6000) |
                                (cliniccode >= 3000 & cliniccode < 4000) |
                                (cliniccode >= 15000 & cliniccode < 16000), 1, 0))
# Filter the non-intervention clinics based on the flag
nonintervclinics <- nonintervclinics %>%
  filter(flagtokeep == 1) %>%
  filter(!(cliniccode %in% c(5302, 5303, 5323, 15312, 3319, 15325, 5306, 
                             15331, 15341, 15329, 3201, 5318, 15335)))

# Creating the MMS start date and end date based on clinic code T1
mms_timeline <- mms %>%                                 # Has 158+2 variables here
  mutate(
    mmsstartdate = case_when(
      cliniccode == 5302 ~ ymd("2024-08-23"),
      cliniccode == 5303 ~ ymd("2024-08-23"),
      cliniccode == 5323 ~ ymd("2024-08-23"),
      cliniccode == 15312 ~ ymd("2024-08-13"),
      cliniccode == 3319 ~ ymd("2024-09-06"),
      cliniccode == 15325 ~ ymd("2024-08-16"),
      cliniccode == 5306 ~ ymd("2024-08-23"),
      cliniccode == 15331 ~ ymd("2024-08-21"),
      cliniccode == 15341 ~ ymd("2024-08-16"),
      cliniccode == 15329 ~ ymd("2024-08-19"),
      cliniccode == 3201 ~ ymd("2024-09-06"),
      cliniccode == 5318 ~ ymd("2024-08-23"),
      cliniccode == 15335 ~ ymd("2024-08-16")
    ),
    mmsenddate = case_when(
      cliniccode == 5302 ~ ymd("2025-02-13"),
      cliniccode == 5303 ~ ymd("2025-03-17"),
      cliniccode == 5323 ~ ymd("2025-02-13"),
      cliniccode == 15312 ~ ymd("2025-02-20"),
      cliniccode == 3319 ~ ymd("2025-04-15"),
      cliniccode == 15325 ~ ymd("2025-02-20"),
      cliniccode == 5306 ~ ymd("2025-02-13"),
      cliniccode == 15331 ~ ymd("2025-02-20"),
      cliniccode == 15341 ~ ymd("2025-02-20"),
      cliniccode == 15329 ~ ymd("2025-02-20"),
      cliniccode == 3201 ~ ymd("2025-04-15"),
      cliniccode == 5318 ~ ymd("2025-02-13"),
      cliniccode == 15335 ~ ymd("2025-02-20")
    )
  ) %>%
  # Restriction step, 40 weeks before cut-off date Sep 03, 2025
  filter(lmp <= ymd("2024-11-27"))

# Filter mms data for women whose booking date is on or after the start of the pilot
Interv_T1 <- mms_timeline %>%    #Rename from mms_T1 to Interv_T1, for intervention clincis during intervention period  
  filter(bookdate >= mmsstartdate)       

# Creating T0 for intervention clinics 
Interv_timeline_T0 <- mms_timeline %>%             #List past three timelines 
  mutate(
    t0_start_1 = mmsstartdate - years(1),
    t0_end_1   = mmsenddate - years(1),
    t0_start_2 = mmsstartdate - years(2),
    t0_end_2   = mmsenddate - years(2),
    t0_start_3 = mmsstartdate - years(3),
    t0_end_3   = mmsenddate - years(3)
  ) %>%
  distinct(cliniccode, .keep_all = TRUE) %>%
  select(cliniccode, starts_with("t0_"))

Interv_T0 <- mms %>%        #For intervention clincis during non-intervention period  
  left_join(Interv_timeline_T0, by = "cliniccode") %>%
  filter(
    (bookdate >= t0_start_1 & bookdate <= t0_end_1) |
      (bookdate >= t0_start_2 & bookdate <= t0_end_2) |
      (bookdate >= t0_start_3 & bookdate <= t0_end_3)
  )

# Create start and end dates for non-intervention clinics
nonintervclinics_timeline <- nonintervclinics %>%                          # Has 159+2 variables
  mutate(
    mmsstartdate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2024-08-23"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2024-08-16"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2024-09-06")
    ),
    mmsenddate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2025-02-13"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2025-02-20"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2025-04-15")
    )
  ) %>%

  # Restriction step, 40 weeks before cut-off date Sep 03, 2025
  filter(lmp <= ymd("2024-11-27"))

# Filter non-intervention clinics for women whose booking date is on or after the start of the pilot
noninterv_T1 <- nonintervclinics_timeline %>%
  filter(bookdate >= mmsstartdate)

# Creating T0 for nonintervention clinics 
Noninterv_timeline_T0_ <- nonintervclinics_timeline %>%             #List past three timelines 
  mutate(
    t0_start_1 = mmsstartdate - years(1),
    t0_end_1   = mmsenddate - years(1),
    t0_start_2 = mmsstartdate - years(2),
    t0_end_2   = mmsenddate - years(2),
    t0_start_3 = mmsstartdate - years(3),
    t0_end_3   = mmsenddate - years(3)
  ) %>%
  distinct(cliniccode, .keep_all = TRUE) %>%
  select(cliniccode, starts_with("t0_"))

noninterv_T0 <- nonintervclinics %>%   #use original nonintervention clinics dataset without T1 timeline
  left_join(Noninterv_timeline_T0_, by = "cliniccode") %>%
  filter(
    (bookdate >= t0_start_1 & bookdate <= t0_end_1) |
      (bookdate >= t0_start_2 & bookdate <= t0_end_2) |
      (bookdate >= t0_start_3 & bookdate <= t0_end_3)
  )# Create start and end dates for non-intervention clinics
nonintervclinics_timeline <- nonintervclinics %>%                          # Has 159+2 variables
  mutate(
    mmsstartdate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2024-08-23"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2024-08-16"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2024-09-06")
    ),
    mmsenddate = case_when(
      cliniccode >= 5000 & cliniccode < 6000 ~ ymd("2025-02-13"),
      cliniccode >= 15000 & cliniccode < 16000 ~ ymd("2025-02-20"),
      cliniccode >= 3000 & cliniccode < 4000 ~ ymd("2025-04-15")
    )
  ) %>%
  # Restriction step, 40 weeks before cut-off date Sep 03, 2025
  filter(lmp <= ymd("2024-11-27"))

# Filter non-intervention clinics for women whose booking date is on or after the start of the pilot
noninterv_T1 <- nonintervclinics_timeline %>%
  filter(bookdate >= mmsstartdate)

# Creating T0 for nonintervention clinics 
Noninterv_timeline_T0_ <- nonintervclinics_timeline %>%             #List past three timelines 
  mutate(
    t0_start_1 = mmsstartdate - years(1),
    t0_end_1   = mmsenddate - years(1),
    t0_start_2 = mmsstartdate - years(2),
    t0_end_2   = mmsenddate - years(2),
    t0_start_3 = mmsstartdate - years(3),
    t0_end_3   = mmsenddate - years(3)
  ) %>%
  distinct(cliniccode, .keep_all = TRUE) %>%
  select(cliniccode, starts_with("t0_"))

noninterv_T0 <- nonintervclinics %>%   #use original nonintervention clinics dataset without T1 timeline
  left_join(Noninterv_timeline_T0_, by = "cliniccode") %>%
  filter(
    (bookdate >= t0_start_1 & bookdate <= t0_end_1) |
      (bookdate >= t0_start_2 & bookdate <= t0_end_2) |
      (bookdate >= t0_start_3 & bookdate <= t0_end_3)
  )
```
### Implementation Outcomes

The implementation outcome was MMS penetration (often called “reach”),18 which is defined as the proportion of eligible individuals who participate in the intervention. We estimated penetration by calculating the proportion of deliveries for which there was documentation in the obstetric card that MMS was the first supplement prescribed during pregnancy. We used a difference-in-difference (DiD) approach to compare this proportion across time and clinic. We fit a linear probability model with indicators for the intervention group, the post-intervention period, and an interaction between the intervention group and post-intervention period. The coefficient for the interaction term quantifies the change in the proportion who received MMS from the pre-intervention to the post-intervention period, comparing the intervention clinics with the non-intervention clinics (a difference-in-difference). Each clinic had a unique intervention period, defined as the time between the first MMS distribution and 30 days after the sixth month of distribution. The pre-intervention period refers to the same calendar period in the three consecutive prior years. We included all clinics in Gaborone, Molepolole, and Palapye as non-intervention clinics in the primary analysis. Individuals were included in the analysis if they initiated ANC at one of the 13 intervention clinics or at a non-intervention clinic during either the intervention period or the corresponding pre-intervention period and delivered at a site included in Tsepamo. For non-intervention clinics, the intervention and pre-intervention periods were defined using the median intervention dates from clinics within the same geographic region.

```{r}
#Coding variable for MMNS, IFAS only, Iron only, No iron
tsepamo <- tsepamo %>%
  mutate(
    mmns = case_when(
      multivit == 1 ~ 3,                                      # MMNS
      (iron == 1 & folate == 1) | (fefol == 1) ~ 2,            # Iron + folatic acid but no other micronutrients
      (iron == 1 & folate == 0 & fefol == 0 & multivit == 0) ~ 1, # Iron only
      (iron == 0 & fefol == 0 & multivit == 0) ~ 0             # Not on iron
    )
  )

# Number MMS filled in Tsepamo among Women Presenting to ANC
# During T1
##Intervention
Interv_MMS_T1 <- Interv_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))
##nonintervention
noninterv_MMS_T1 <- noninterv_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))

# During T0
##Intervention
Interv_MMS_T0 <- Interv_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))
##nonintervention
noninterv_MMS_T0 <- noninterv_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    multivit_count = sum(multivit == 1, na.rm = TRUE))
```

### Effectiveness Outcomes

The primary effectiveness outcome was the mean change in hemoglobin during pregnancy, calculated as the difference between the second and first hemoglobin measurements recorded in the Tsepamo database. Specifically, the first hemoglobin must be measured before 24 weeks of gestation, and the second hemoglobin must be measured on or after 24 weeks of gestation. Secondary outcomes included the second hemoglobin value measured in pregnancy and a composite indicator of adverse birth outcomes, including stillbirth, preterm delivery (\<37 weeks of gestation), in-hospital neonatal death (\<28 days), and small-for-gestational-age (SGA). Preterm delivery, SGA, and low birth weight (\<2500 grams) were also examined individually. As above, analyses of the effectiveness outcomes were restricted to individuals who delivered at a site included in Tsepamo.

Adverse Birth Outcome
```{r}
# Number any adverse birth outcome in Tsepamo among Women Presenting to ANC

## SGA and VSGA calculation functions

calculate_sga <- function(nga, bwt, gender, bwtgm) {
  # Vectorized version using case_when
  dplyr::case_when(
    is.na(nga) | is.na(bwt) | is.na(gender) | is.na(bwtgm) ~ NA_real_,
    bwt < bwtgm ~ 1L,  # Meets SGA criteria
    TRUE ~ 0L          # Does not meet criteria
  )
}

calculate_vsga <- function(nga, bwt, gender, bwtgm) {
  # Identical structure since logic is the same, just different thresholds
  dplyr::case_when(
    is.na(nga) | is.na(bwt) | is.na(gender) | is.na(bwtgm) ~ NA_real_,
    bwt < bwtgm ~ 1L,  # Meets VSGA criteria
    TRUE ~ 0L          # Does not meet criteria
  )
}

# Correct SGA and VSGA weight thresholds
sgawham <- setNames(c(530,600,690,780,880,1000,1130,1270,1430,1550,1820,2050,2270,2470,2640,2800,2940,3060,3170),
                    24:42)
sgawhaf <- setNames(c(500,570,650,740,830,940,1070,1200,1350,1530,1790,2020,2230,2410,2570,2710,2830,2930,3010),
                    24:42)

vsgawham <- setNames(c(470,540,610,690,790,890,1000,1130,1270,1300,1560,1800,2020,2220,2390,2550,2690,2820,2920),
                     24:42)
vsgawhaf <- setNames(c(440,510,580,660,740,840,950,1070,1200,1320,1570,1800,2000,2180,2340,2480,2600,2690,2770),
                     24:42)

calculate_growth_status <- function(bwt, bwtgm) {
  dplyr::case_when(
    is.na(bwt) | is.na(bwtgm) ~ NA_real_,  # Handle missing data
    bwt < bwtgm ~ 1L,                     # Meets criteria
    TRUE ~ 0L                              # Does not meet criteria
  )
}


## For Intervention period

# Intervention clinics 
Interv_BD_T1 <- Interv_T1 %>%
  # First create prerequisite variables
  mutate(
    
    # Set implausible values to NA
    nga_clean = if_else(nga %in% 24:42, nga, NA_real_),
    
    # Flag outliers for review
    nga_flag = case_when(
      nga < 24 ~ "Below viability threshold",
      nga > 42 ~ "Post-term error",
      is.na(nga) ~ "Missing GA",
      TRUE ~ NA_character_
    ),
    
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga_clean %in% 24:36 ~ 1L,  # WHO preterm definition
      nga_clean >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga_clean %in% 24:31 ~ 1L,  # WHO very preterm
      nga_clean >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birth weight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(round(nga_clean)),  # Handle decimal weeks if needed
    bwtgm_sga = case_when(
      gender == 0 ~ sgawham[nga_char],
      gender == 1 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 ~ vsgawham[nga_char],
      gender == 1 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt <= bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt <= bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

Interv_BD_clinics_T1 <- Interv_BD_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )


# Non-intervention clincis
noninterv_BD_T1 <- noninterv_T1 %>%
  # First create prerequisite variables
  mutate(
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga >= 24 & nga < 37 ~ 1L,
      nga >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga >= 24 & nga < 32 ~ 1L,
      nga >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birthweight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(nga),
    bwtgm_sga = case_when(
      gender == 0 & nga %in% 24:42 ~ sgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 & nga %in% 24:42 ~ vsgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt < bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt < bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

noninterv_BD_clinics_T1 <- noninterv_BD_T1 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )

## For Non-interventnion period
#Intervenetion clincis
Interv_BD_T0 <- Interv_T0 %>%
  # First create prerequisite variables
  mutate(
    
    # Set implausible values to NA
    nga_clean = if_else(nga %in% 24:42, nga, NA_real_),
    
    # Flag outliers for review
    nga_flag = case_when(
      nga < 24 ~ "Below viability threshold",
      nga > 42 ~ "Post-term error",
      is.na(nga) ~ "Missing GA",
      TRUE ~ NA_character_
    ),
    
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga_clean %in% 24:36 ~ 1L,  # WHO preterm definition
      nga_clean >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga_clean %in% 24:31 ~ 1L,  # WHO very preterm
      nga_clean >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birthweight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(round(nga_clean)),  # Handle decimal weeks if needed
    bwtgm_sga = case_when(
      gender == 0 ~ sgawham[nga_char],
      gender == 1 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 ~ vsgawham[nga_char],
      gender == 1 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt <= bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt <= bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

Interv_BD_clinics_T0 <- Interv_BD_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )

# Non-interverntion clinics
noninterv_BD_T0 <- noninterv_T0 %>%
  # First create prerequisite variables
  mutate(
    # Stillbirth calculation
    stillbirth = case_when(
      infantstat %in% c(2, 3) ~ 1L,
      infantstat == 1 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Preterm calculations
    preterm = case_when(
      nga >= 24 & nga < 37 ~ 1L,
      nga >= 37 ~ 0L,
      TRUE ~ NA_integer_
    ),
    verypreterm = case_when(
      nga >= 24 & nga < 32 ~ 1L,
      nga >= 32 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Low birthweight calculations
    lowbirthweight = case_when(
      bwt < 2500 ~ 1L,
      bwt >= 2500 ~ 0L,
      TRUE ~ NA_integer_
    ),
    
    # Neonatal death recoding
    nndeath_old = case_when(
      nndeath == 1 ~ 1L,
      nndeath == 0 ~ 0L,
      TRUE ~ NA_integer_
    ),
    nndeath_new = case_when(
      nndeath_old == 1 & alivebaby == 1 ~ NA_integer_,
      nndeath_old == 0 & alivebaby == 0 ~ NA_integer_,
      TRUE ~ nndeath_old
    ),
    
    # SGA/VSGA thresholds
    nga_char = as.character(nga),
    bwtgm_sga = case_when(
      gender == 0 & nga %in% 24:42 ~ sgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ sgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    bwtgm_vsga = case_when(
      gender == 0 & nga %in% 24:42 ~ vsgawham[nga_char],
      gender == 1 & nga %in% 24:42 ~ vsgawhaf[nga_char],
      TRUE ~ NA_real_
    ),
    
    # SGA/VSGA statuses (assuming these are sgawho_all/vsgawho_all)
    sgawho_all = case_when(
      bwt < bwtgm_sga ~ 1L,
      bwt >= bwtgm_sga ~ 0L,
      TRUE ~ NA_integer_
    ),
    vsgawho_all = case_when(
      bwt < bwtgm_vsga ~ 1L,
      bwt >= bwtgm_vsga ~ 0L,
      TRUE ~ NA_integer_
    )
  ) %>%
  
  # Now create composite outcomes
  mutate(
    composite = case_when(
      sgawho_all == 1 | preterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(sgawho_all, preterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    ),
    severe_composite = case_when(
      vsgawho_all == 1 | verypreterm == 1 | stillbirth == 1 | nndeath_new == 1 ~ 1L,
      if_any(c(vsgawho_all, verypreterm, stillbirth, nndeath_new), is.na) ~ NA_integer_,
      TRUE ~ 0L
    )
  ) %>%
  select(-nga_char)

noninterv_BD_clinics_T0 <- noninterv_BD_T0 %>%
  group_by(cliniccode) %>%
  summarise(
    n_total = sum(!is.na(composite)),                #n_total is the number of non-missing composite observations per clinic
    n_events = sum(composite == 1, na.rm = TRUE),    #n_events is the number of cases where composite == 1
    percent = mean(composite, na.rm = TRUE) * 100
  )
```

Hemoglobins
```{r}
# Number deliveries in Tsepamo among Women Presenting to ANC with 2 hemoglobins
#During T1

##Interventnion
Interv_2hb_T1 <- Interv_T1 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())
##Non~
noninterv_2hb_T1 <- noninterv_T1 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())

# During T0

##Interventnion
Interv_2hb_T0 <- Interv_T0 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())
##Non~
noninterv_2hb_T0 <- noninterv_T0 %>%
  filter(!is.na(firsthb) & !is.na(secondhb)) %>%
  group_by(cliniccode) %>%
  summarise(n_with_2_hb = n())

```

## Results {#sec-results}

Describe your results and include relevant tables, plots, and code/comments used to obtain them. You may refer to the @sec-methods as needed. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

## Conclusion

This the conclusion. The @sec-results can be invoked here.
