---
title: "Health Disparities Between Rural and Urban Populations in Philadelphia Region"
author: "Joyce Wang"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
### Overview
Rural population are generally less well off than urban population with respect to health due to lack of access and increased risky health behaviors among others. I'm interested to see if these health disparities exist in the greater Philadelphia region and what factors are the greatest predictors of adverse health events. I will be applying statistical and machine learning analysis to the Southeastern Pennsylvania Household Survey results from 2015 to evaluate the difference between rural and urban populations. You can get access to my Github repository [here](https://github.com/joycehelenwang/BMIN503_Final_Project).

I conceptualized my project after speaking with three faculty members: Andria Johnson, Rebecca Hubbard, and Ariana Chao. Andria Johnson is a Professor in the Department of History and Sociology of Science and recommended that I look at health disparities in the local Philadelphia region. Rebecca Hubbard is an Associate Professor of Biostatistics in the Department of Biostatistics, Epidemiology and Informatics and suggested using machine learning techniques to create a model to predict health outcomes and determine source variability. Ariana Chao is an Assistant Professor of Nursing and reccomended some datasets and discussed the impact of mental health on BMI and other health outcomes as a possibile route.

### Introduction 
In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.

In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.


### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

``` {r, eval = TRUE}
library(haven)

data <- read_sav(url("https://raw.githubusercontent.com/joycehelenwang/BMIN503_Final_Project/master/HS15COM1b.sav"))


```


### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
