---
title: "BMIN503/EPID600 Project Template"
author: "Your Name"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

### Overview
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.


### Introduction 
Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.


### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 
```{r}
library(tidyverse)
#load raw dataset
seer <- read.csv("C:/Users/kmroh/Desktop/BMIN503/dataset1.csv", header = T)
#view column names and rename variables of interest
colnames(seer)
seer <- seer %>%
  rename(patid = Patient.ID, 
         sex = Sex, 
         year.dx = Year.of.diagnosis,
         mstatus = Marital.status.at.diagnosis,
         race = Race.recode..White..Black..Other.,
         surv.months = Survival.months,
         cod.site = COD.to.site.recode,
         cod = SEER.cause.specific.death.classification,
         vital.status = Vital.status.recode..study.cutoff.used.,
         yob = Year.of.birth,
         prim.site = Primary.Site...labeled) 
seer$surv.months <- as.integer(seer$surv.months)
#see possible marital statuses
table(seer$mstatus)
#remove entries which are not married, partnered, or single at time of diagnosis
seer <- seer %>%
  filter(mstatus == "Married (including common law)" |
         mstatus == "Unmarried or Domestic Partner" |
         mstatus == "Single (never married)")
#recode to those who are partnered or not at the time of diagnosis
seer$part <- ifelse(seer$mstatus != "Single (never married)",
                    1,
                    0)
seer$part <- factor(seer$part, labels = c("Single","Partnered"))
#recode race variable
table(seer$race)
seer$race <- ifelse(seer$race == "Other (American Indian/AK Native, Asian/Pacific Islander)",
                    "Other", ifelse(seer$race == "White",
                                    "White", ifelse(seer$race == "Black",
                                                    "Black", ifelse(seer$race == "Unknown",
                                                                    "Unknown", 0))))
#remove those of unknown race 
seer <- seer %>%
  filter(race != "Unknown")
#remove those whose cause of death is unknown, or whose tumor is not the first
table(seer$cod)
seer <- seer %>%
  filter(cod == "Alive or dead of other cause" |
         cod == "Dead (attributable to this cancer dx)")
#remove observations with unknown primary site
seer <- seer %>%
  filter(prim.site != "C80.9-Unknown primary site")

#create group average of survival month by mar variable
seer <- seer %>%
  group_by(part) %>%
  mutate(avg.pstatus = mean(surv.months, na.rm = T))
#plot survival months by recoded marital status
ggplot(seer, aes(x=as.factor(part), y=surv.months)) +
  geom_boxplot() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(title = "Survival Months by Marital Status at Diagnosis \nAll Sexes", 
       x = "Relationship status", 
       y = "Survival Months")

seer <- seer %>%
  group_by(sex) %>%
  mutate(avg.part.sex = mean(surv.months, na.rm = T))

ggplot(seer, aes(x=as.factor(sex), y=surv.months)) +
  geom_boxplot() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(title = "Survival Months by Sex", 
       x = "Sex", 
       y = "Survival Months")

ggplot(seer, aes(x=as.factor(part), y=surv.months)) +
  geom_boxplot(aes(color = sex)) +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black")) +
  labs(title = "Survival Months by Marital Status at Diagnosis \nand Sex", 
       x = "Relationship status", 
       y = "Survival Months")
#
```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
