---
title: "Predicting the occurrence of steatosis and cirrhosis based on transient elastography in U.S. adults using supervised and unsupervised machine learning models: An analysis of the National Health and Nutrition Examination Survey, 2017-2020"
subtitle: "BMIN503/EPID600 Final Project"
author: "Samiran Mukherjee"
format: html
editor: visual
number-sections: true
embed-resources: true
---

------------------------------------------------------------------------

## Overview {#sec-overview}

Using NHANES data, we aim to predict the determinants of liver steatosis and fibrosis (using the publically available fibroscan data). In addition, we will also learn to train a machine learning model to identify variables that can best predict the occurence of fibrosis. .

## Introduction {#sec-introduction}

Metabolic dysfunction-associated steatotic liver disease (MASLD) is prevalent in 30% of the US population and is strongly associated with the presence of cardiometabolic risk factors like obesity, type 2 diabetes, hypertension and dyslipidemia. Given that 1% of patients will develop cirrhosis, it is imperative to identify preventable independent variables that may influence its occurrence. We aimed to predict the determinants of liver steatosis and fibrosis and train a machine learning models to identify variables that can best predict the occurrence of fibrosis. 

## Methods {#sec-methods}

We will be using NHANES data to assess the determinants of advanced fibrosis from the link https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?Cycle=2017-2020

From the multiple data sets availabe from 2017-2020, we will be using the following files for the following: 1) demo_data; Demographic data 2) hepq_data; Hepatitis questionnaire data 3) lab_data; Lab data 4) dmq_data; dmq (diabetes questionnaire) data 5) a1c_data; a1c (diabetes lab) data 6) hepb_data; hepatitis b data 7) hepc_data; Hepatitis c data 8) fibroscan_data; Fibroscan (firbous elastography) data 9) insurance_data; insurance data 10)alcohol_data; Alcohol data

The study plan will comprise of

1\) Identifying cases and controls using the fibroscan data

2\) Use logistic regression models to identify variables that are associated with cases

3)create a supervised machine learning model to predict the occurence of steatosis

4\) Plotting AUC curves for the same

6\) create an unsupervised machine model (Kmeans clustering) and plot cases vs controls

Install packages

```{r}
#install.packages("modelsummary")
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("haven")
#install.packages("tidyverse")
#install.packages("tidyr")
#install.packages("kernlab")
#install.packages("randomForest")
#install.packages("glmnet")
#install.packages("tidymodels")
#install.packages("dotwhisker")
#install.packages("vip")
#install.packages("Rtsne")
#install.packages("cowplot")
#install.packages("ggdendro")
```

Load packages

```{r}
library(tidyverse)
library(modelsummary)
library(dplyr)
library(broom)
library(ggplot2)
library(dplyr)
library(haven) #required to read XPT files
library(tidyr)
library(kernlab)
library(randomForest)
library(glmnet)
library(tidymodels)
tidymodels_prefer()
library(dotwhisker)
library(vip)
library(Rtsne)
library(cowplot)
library("ggdendro")
library(statsExpressions)
library(ggstatsplot)
```

```{r}
##read XPT files
demo_data <- read_xpt("P_DEMO.XPT") |> data.frame("P_DEMO.XPT")
hepq_data <- read_xpt("P_HEQ.XPT") |> data.frame("P_HEQ.XPT")
lab_data <- read_xpt("P_BIOPRO.XPT") |> data.frame("P_BIOPRO.XPT")
dmq_data <- read_xpt("P_DIQ.XPT") |> data.frame("P_DIQ.XPT") 
a1c_data <- read_xpt("P_GHB.XPT") |> data.frame("P_GHB.XPT")
hepb_data <- read_xpt("P_HEPBD.XPT") |> data.frame("P_HEPBD.XPT")
hepc_data <- read_xpt("P_HEPC.XPT") |> data.frame("P_HEPC.XPT")
fibroscan_data <- read_xpt("P_LUX.XPT") |> data.frame("P_LUX.XPT")
insurance_data <- read_xpt("P_HIQ.XPT") |> data.frame("P_HIQ.XPT")
alcohol_data <-  read_xpt("P_ALQ.XPT") |> data.frame("P_ALQ.XPT")
bodymeasurements_data <-  read_xpt("P_BMX.XPT") |> data.frame("P_BMX.XPT")

#Merge each of the tables to create one master data sheet. This has to happen step-wise
merged_df <- merge(demo_data, hepq_data, by = "SEQN")
merged_df <- merge(merged_df, hepb_data, by = "SEQN")
merged_df <- merge(merged_df, hepc_data, by = "SEQN")
merged_df <- merge(merged_df, lab_data, by = "SEQN")
merged_df <- merge(merged_df, dmq_data, by = "SEQN")
merged_df <- merge(merged_df, a1c_data, by = "SEQN")
merged_df <- merge(merged_df, fibroscan_data, by = "SEQN")
merged_df <- merge(merged_df, insurance_data, by = "SEQN")
merged_df <- merge(merged_df, alcohol_data, by = "SEQN")
merged_df <- merge(merged_df, bodymeasurements_data, by = "SEQN")

```

Now to modify each of the variables names of interest accordingly

```{r}
#Alcohol 
merged_df$etoh <- merged_df$ALQ151 #1->Y, 2-> No, 7-> refused, 9-> ?, .-> Missing

#biomarkers (continuous variables)
merged_df$alt <- merged_df$LBXSATSI #Values .-> Missing
merged_df$alb <- merged_df$LBXSAL 
merged_df$alp <- merged_df$LBXSAPSI
merged_df$ast <- merged_df$LBXSASSI 
merged_df$urea <- merged_df$LBXSBU
merged_df$cr <- merged_df$LBXSCR
merged_df$ggt <- merged_df$LBXSGTSI
merged_df$ldh <- merged_df$LBXSLDSI
merged_df$na <- merged_df$LBXSNASI
merged_df$tbil <- merged_df$LBXSTB
merged_df$chol <- merged_df$LBXSCH
merged_df$tg <- merged_df$LBXSTR

#HepQ data
merged_df$hb_self_reported <- merged_df$HEQ010 #1 Yes, 2 No, 7 Refused, 9 ?, . missing)
merged_df$hb_self_reported_treated <- merged_df$HEQ020 #Same as above
merged_df$hc_self_reported <- merged_df$HEQ030
merged_df$hc_self_reported_treated <- merged_df$HEQ040

#HepB data
merged_df$hbcore_positive <- merged_df$LBXHBC #1 yes, 2 neg, 3 +/-, . missing
merged_df$hbsag_positive <- merged_df$LBDHBG #1 yes, 2 neg, 3 +/-, . missing

#HepC data
merged_df$hcrna <- merged_df$LBXHCR #1 yes, 2 neg, 3 neg HCAb
merged_df$hcab <- merged_df$LBDHCI # 1 yes, 2 neg, 3 neg screening , 4 pos HCVRNA, . missing
merged_df$hcgenotype <- merged_df$LBXHCG # 1 1a, 2 1b, 3 1a/b?, 4 2, 5 3, 6 4, 7 5, 8, 6, 9 1

#A1c data
merged_df$ha1c <- merged_df$LBXGH #2.8-16.2 

#fibroscan data
merged_df$fibroscan_meanstiffness_kpa <- merged_df$LUXSMED
merged_df$fibroscan_stiffnessiqr <- merged_df$LUXSIQR
merged_df$fibroscan_cap <- merged_df$LUXCAPM
merged_df$hc_self_cpiqr <- merged_df$LUXCPIQR

#Insurance data 
merged_df$insurance_covered <- merged_df$HIQ011 # 1 yes, 2 no, 7 refused, 9 ?, . missing

#Demographic data 
merged_df$demo_gender <- merged_df$RIAGENDR #1 male, 2 female
merged_df$demo_age <- merged_df$RIDAGEYR # 0-79 and >80
merged_df$demo_race <- merged_df$RIDRETH3 # 1 Mexican american, 2 other hisp, 3 NH White, 4 NH black, 6 NH Asian, 7 other race
merged_df$demo_education <- merged_df$DMDEDUC2 #1 <9thgr, 2 9-11th gr, 3 HSGrad, 4 college, 5 college grad

#Body measures
merged_df$body_bmi <- merged_df$BMXBMI

#convert data type to  factors / numerics
merged_df$hb_self_reported <- as.factor(merged_df$hb_self_reported)
merged_df$hb_self_reported_treated <- as.factor(merged_df$hb_self_reported_treated)
merged_df$hc_self_reported <- as.factor(merged_df$hc_self_reported)
merged_df$hc_self_reported_treated <- as.factor(merged_df$hc_self_reported_treated)
merged_df$hbcore_positive <- as.factor(merged_df$hbcore_positive)
merged_df$hbsag_positive <- as.factor(merged_df$hbsag_positive)
merged_df$hcrna <- as.factor(merged_df$hcrna)
merged_df$hcab <- as.factor(merged_df$hcab)
merged_df$hcgenotype <- as.factor(merged_df$hcgenotype )
merged_df$insurance_covered <- as.factor(merged_df$insurance_covered)
merged_df$demo_gender <- as.factor(merged_df$demo_gender)
merged_df$demo_race <- as.factor(merged_df$demo_race)
merged_df$demo_education <- as.factor(merged_df$demo_education)


```

## Definition of Phenotype

***Cases:***

[1) Steatosis]{.underline}

Steatosis cut offs using fibroscan (https://pubmed.ncbi.nlm.nih.gov/30689971/) CAP value \> 288

[2) Cirrhosis]{.underline}

We will use a fibroscan mean liver stiffness measurement of \>20 kPA for the definition of cirrhosis https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1856085/ https://journals.lww.com/hep/Fulltext/2023/05000/AASLD_Practice_Guidance_on_the_clinical_assessment.31.aspx\>

***Controls:***

1\) [Steatosis controls]{.underline}: CAP \< 288

2\) [Cirrhosis controls]{.underline}: Fibroscan \<8kPA

Create columns that represent the phenotype

```{r}

#Cirrhosis to 0 and 1 using fibroscan data 

merged_df <- merged_df |> mutate(fibroscan_cirrhosis = ifelse(fibroscan_meanstiffness_kpa > 20, 1, 0))

#Steatosis to 0 an 1 using fibroscan data 
merged_df <- merged_df |> mutate(fibroscan_steatosis = ifelse(merged_df$fibroscan_cap > 288, 1, 0))


merged_df$fibroscan_cirrhosis <- as.factor(merged_df$fibroscan_cirrhosis)
merged_df$fibroscan_steatosis <- as.factor(merged_df$fibroscan_steatosis)

#check data type

summary(merged_df$fibroscan_steatosis) #2888 patients with steatosis
summary(merged_df$fibroscan_cirrhosis) #142 patients with cirrhosis 

```

We will also truncate the table to only the columns of interest

```{r}

df <- merged_df[, c("alt", "alb","alp", "ast", "urea", "cr", "ggt", "ldh", "na", "tbil", "chol", "tg", "etoh", "fibroscan_meanstiffness_kpa", "ha1c", "fibroscan_stiffnessiqr", "fibroscan_cap", "hc_self_cpiqr", "insurance_covered","demo_gender", "demo_age" , "demo_race", "demo_education", "body_bmi", "hb_self_reported", "hc_self_reported", "hc_self_reported", "hbcore_positive", "hcrna", "hcab",  "fibroscan_steatosis", "fibroscan_cirrhosis")]

# we will have to remove columns hc_self_reported_treated, hb_self_reported_treated, hcgenotype, hbsag as these are the columns with the maximum number of missings

#We now remove all the missings
df <- na.omit(df)


```

## Results

## Identifying variables associated with Steatosis

```{r}
model_steatosis_binary <- glm(fibroscan_steatosis ~ alt + alb + alp + ast + urea + cr + ggt + ldh + na + tbil + chol + tg + etoh + ha1c +  insurance_covered + demo_gender + demo_age + demo_race + demo_education + body_bmi, data = df, family = binomial)

summary(model_steatosis_binary)

significant_results_steatosis_binary <- tidy(model_steatosis_binary)|> filter(p.value < 0.05) |> arrange(p.value) |> mutate(Odds_Ratio = exp(estimate))


table(df$fibroscan_steatosis)

#How does the strongest variable associate with the outcome? # BMI
#PLOT 
plot1 <- ggplot(df, aes(x = fibroscan_steatosis, y = body_bmi)) +
  geom_violin(aes(fill = fibroscan_steatosis)) +
  geom_boxplot(width = 0.1, alpha = 0.2)



plot1


summary(df$body_bmi)




```

```{r}


ggbetweenstats(
  data  = df,
  x     = fibroscan_steatosis,
  y     = body_bmi,
  title = "Steatosis vs BMI ")


```

As a continuous variable

```{r}

model_steatosis_continuous <- glm(fibroscan_cap ~ alt + alb + alp + ast + urea + cr + ggt + ldh + na + tbil + chol + tg + etoh + ha1c +  insurance_covered + demo_gender + demo_age + demo_race + demo_education + body_bmi, data = df)

significant_results_steatosis_continuous <- tidy(model_steatosis_continuous)|> filter(p.value < 0.05) |> arrange(p.value) |> mutate(Odds_Ratio = exp(estimate))

significant_results_steatosis_continuous




#How does the strongest variable associate with the outcome? # BMI
#PLOT 
plot2 <- ggplot(df, aes(x = body_bmi, y = fibroscan_cap)) + geom_point()+
  geom_smooth(method = "lm") 


plot2

```

## Identifying variables associated with Cirrhosis

```{r}

table(df$hbcore_positive)

model_cirrhosis_binary <- glm(fibroscan_cirrhosis ~ alt + alb + alp + ast + urea + cr + ggt + ldh + na + tbil + chol + tg + etoh + ha1c + insurance_covered + demo_gender + demo_age + demo_race + demo_education + body_bmi + hcab + hbcore_positive + hcrna, data = df, family = binomial)

summary(model_cirrhosis_binary)

significant_results_cirrhosis_binary <- tidy(model_cirrhosis_binary)|> filter(p.value < 0.05) |> arrange(p.value) |> mutate(Odds_Ratio = exp(estimate))

significant_results_cirrhosis_binary

#How does the strongest variable associate with the outcome? \# BMI

#PLOT

plot10 <- ggplot(data=df, aes(x = fibroscan_cirrhosis, y = body_bmi)) +

geom_violin(aes(fill = fibroscan_cirrhosis)) +

geom_boxplot(width = 0.1, alpha = 0.2)


plot10


```

```{r}


ggbetweenstats(
  data  = df,
  x     = fibroscan_steatosis,
  y     = body_bmi,
  title = "Steatosis vs BMI ")


```

As a continuous variable

```{r}

model_cirrhosis_continuous <- glm(fibroscan_meanstiffness_kpa ~ alt + alb + alp + ast + urea + cr + ggt + ldh + na + tbil + chol + tg + etoh + ha1c + insurance_covered + demo_gender + demo_age + demo_race + demo_education + body_bmi + hcab + hbcore_positive + hcrna, data = df)

significant_results_cirrhosis_continuous <- tidy(model_cirrhosis_continuous)|> filter(p.value < 0.05) |> arrange(p.value) |> mutate(Odds_Ratio = exp(estimate))

significant_results_cirrhosis_continuous


#How does the strongest variable associate with the outcome? # BMI
#PLOT to see the number of hcrna
plot11 <- ggplot(df, aes(x = body_bmi, y = fibroscan_meanstiffness_kpa, color=hcrna)) + geom_point()+
  geom_smooth(method = "lm") 

plot11

#PLOT to see the numebr of hb
plot12 <- ggplot(df, aes(x = body_bmi, y = fibroscan_meanstiffness_kpa, color=hbcore_positive)) + geom_point()+
  geom_smooth(method = "lm") 

plot12

```

## Supervised machine learning using Random Forest Model to assess most important factors/GINI in Steatosis

For this we will have to convert the cases to 1 and controls to 0 as the "truth" argument of the predictor variable regonizes "Cases" and "Controls" best

```{r}
#We will convert 1-> Cases and 0-> Controls
#Steatosis
df<- df |>
  mutate(fibroscan_steatosis = ifelse(fibroscan_steatosis == 1, "Cases", "Controls"))

table(df$fibroscan_steatosis)

df$fibroscan_steatosis <-  as.factor(df$fibroscan_steatosis)

```

create a new data frame without any other fibroscan data except for fibroscan_steatosis and fibroscan_cirrhosis as they will automatically serve as a predictor

```{r}

df <- subset(df, select = -c(fibroscan_meanstiffness_kpa, fibroscan_cap, fibroscan_stiffnessiqr))


```

```{r}
#Random forest model
rf_cls_spec <- 
  rand_forest(trees = 1000, min_n = 5) |> 
  set_engine("randomForest", importance = TRUE) |>
  set_mode("classification")

rf_cls_spec

rf_cls_fit_steatosis <- rf_cls_spec |>
  fit(fibroscan_steatosis ~ alt + alb + alp + ast + urea + cr + ggt + ldh + na + tbil + chol + tg + etoh + ha1c +  insurance_covered + demo_gender + demo_age + demo_race + demo_education + body_bmi, data = df)


rf_cls_fit_steatosis

#for importance scores 
rf_cls_fit_steatosis |>
  extract_fit_engine() |>
  importance()

#to plot MeanDecreaseGini
plot3 <- rf_cls_fit_steatosis |>
  extract_fit_engine() |>
  vip()

plot3 

#save
rf.predicted.steatosis <- bind_cols(
  truth = df$fibroscan_steatosis,
  predict(rf_cls_fit_steatosis, df),
  predict(rf_cls_fit_steatosis, df, type = "prob")
)


rf.predicted.steatosis

#RF plots

plot4 <- autoplot(roc_curve(rf.predicted.steatosis, 
                   truth, .pred_Cases ))

plot4 

plot5 <- roc_auc(rf.predicted.steatosis, 
                   truth,  .pred_Cases )


plot5


```

For obtaining 10-fold cross validation classification vectors for the model

```{r}
set.seed(1234)

data.split <- initial_split(df, 
                            strata = fibroscan_steatosis, 
                            prop = 0.80)

data.split

data.train <- training(data.split)
data.test <- testing(data.split)
```

Training based on RF model

```{r}

rf_cls_fit_train<- rf_cls_spec |>
  fit(fibroscan_steatosis ~ ., data = data.train)


rf_cls_fit_train
```

Testing based on RF model

```{r}
data.rf.pred.values.test <-  bind_cols(
  truth = data.test$fibroscan_steatosis,
  predict(rf_cls_fit_train, data.test),
  predict(rf_cls_fit_train, data.test, type = "prob")
)


data.rf.pred.values.test
```

ROC and AUC

```{r}
plot6 <- autoplot(roc_curve(data.rf.pred.values.test, 
                   truth, .pred_Cases))

plot6

plot7 <- roc_auc(data.rf.pred.values.test, 
                   truth, .pred_Cases)

plot7
```

## Unsupervised machine learning using Kmeans clustering in Steatosis

```{r}
set.seed(1234)

df.kmeans <- kmeans(df[ , 1:24], 2)
df.kmeans
table(df$fibroscan_steatosis, df.kmeans$cluster)
centers <- as.data.frame(df.kmeans$centers)


plot8 <- ggplot(data = df) +
    geom_point(aes(body_bmi, ha1c, 
                        color = factor(df.kmeans$cluster))) +
    geom_point(data = centers, aes(body_bmi, ha1c), 
               color = "purple", size = 2)

plot8
```

## Supervised machine learning using Random Forest Model to assess most important factors/GINI in Cirrhosis

For this we will have to convert the cases to 1 and controls to 0 as the "truth" argument of the predictor variable regonizes "Cases" and "Controls" best

```{r}
#We will convert 1-> Cases and 0-> Controls
#Steatosis
df<- df |>
  mutate(fibroscan_cirrhosis = ifelse(fibroscan_cirrhosis == 1, "Cases", "Controls"))

table(df$fibroscan_cirrhosis)

df$fibroscan_cirrhosis <-  as.factor(df$fibroscan_cirrhosis)


```

```{r}


ggbetweenstats(
  data  = df,
  x     = fibroscan_cirrhosis,
  y     = body_bmi,
  title = "Cirrhosis vs BMI ")


```

```{r}
#Random forest model
rf_cls_spec <- 
  rand_forest(trees = 1000, min_n = 5) |> 
  set_engine("randomForest", importance = TRUE) |>
  set_mode("classification")

rf_cls_spec

rf_cls_fit_cirrhosis <- rf_cls_spec |>
  fit(fibroscan_cirrhosis ~ alt + alb + alp + ast + urea + cr + ggt + ldh + na + tbil + chol + tg + etoh + ha1c + insurance_covered + demo_gender + demo_age + demo_race + demo_education + body_bmi + hcab + hbcore_positive + hcrna, data = df)


rf_cls_fit_cirrhosis

#for importance scores 
rf_cls_fit_cirrhosis |>
  extract_fit_engine() |>
  importance()

#to plot MeanDecreaseGini
rf_cls_fit_cirrhosis |>
  extract_fit_engine() |>
  vip()


#save
rf.predicted.cirrhosis <- bind_cols(
  truth = df$fibroscan_cirrhosis,
  predict(rf_cls_fit_cirrhosis, df),
  predict(rf_cls_fit_cirrhosis, df, type = "prob")
)


rf.predicted.cirrhosis

#RF plots

plot13 <- autoplot(roc_curve(rf.predicted.cirrhosis, 
                   truth, .pred_Cases ))

plot13 

plot14 <- roc_auc(rf.predicted.cirrhosis, 
                   truth,  .pred_Cases )


plot14



```

For obtaining 10-fold cross validation classification vectors for the model

```{r}
set.seed(1234)

data.split <- initial_split(df, 
                            strata = fibroscan_cirrhosis, 
                            prop = 0.80)

data.split

data.train <- training(data.split)
data.test <- testing(data.split)




```

Training based on RF model

```{r}
rf_cls_fit_train<- rf_cls_spec |>
  fit(fibroscan_cirrhosis ~ ., data = data.train)


rf_cls_fit_train
```

Testing based on RF model

```{r}
data.rf.pred.values.test <-  bind_cols(
  truth = data.test$fibroscan_cirrhosis,
  predict(rf_cls_fit_train, data.test),
  predict(rf_cls_fit_train, data.test, type = "prob")
)


data.rf.pred.values.test
```

ROC and AUC Curves

```{r}
autoplot(roc_curve(data.rf.pred.values.test, 
                   truth, .pred_Cases))


roc_auc(data.rf.pred.values.test, 
                   truth, .pred_Cases)
```

## Unsupervised machine learning using Kmeans clustering in cirrhosis

```{r}
set.seed(1234)

df.kmeans <- kmeans(df[ , 1:24], 2)
df.kmeans
table(df$fibroscan_cirrhosis, df.kmeans$cluster)
centers <- as.data.frame(df.kmeans$centers)


ggplot(data = df) +
    geom_point(aes(body_bmi, ha1c, 
                        color = factor(df.kmeans$cluster))) +
    geom_point(data = centers, aes(body_bmi, ha1c), 
               color = "purple", size = 2)




```

Unsupervised Machine Learning using hierarchical clustering (removing the last 2 columns #31 and #32 as they are the phenotype columns)

```{r}
df.hclust <- hclust(dist(df), method = "complete")

df.dend <- dendro_data(as.dendrogram(df.hclust))

labels <- label(df.dend)

labels$fibroscan_steatosis <- df$fibroscan_steatosis[as.numeric(labels$label)]

plot15 <- ggplot(segment(df.dend)) +
    geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_text(data = labels, aes(label = label, x = x, y = 0, 
                                 color = fibroscan_steatosis), size = 4)


plot15
```

```{r}
df.cut <- cutree(df.hclust, 11) #Cut where there are 11 clusters
df.comparison <- table(df$fibroscan_steatosis, df.cut)

df.comparison

```

all plots (using cowplot)

```{r}
combined_plot <- plot_grid(plot1, plot2, plot3, plot4, plot5, plot6, plot7, plot8, plot10, plot11, plot12, plot13, plot14,plot15)


combined_plot

```

## Interpretation of results

Using transient elastography, phenotypes were created for Steatosis (n=2209) and Cirrhosis (n =109). Multivariate logistic regressions revealed that body mass index (BMI) (OR=1.17, *p* \<0.001), HbA1c (OR=1.37, *p* \<0.001), triglyceride levels (OR =1.004, *p* \<0.001) were associated with steatosis, whereas BMI (OR = 1.12, *p*\<0.001), albumin (OR = 0.2299, P\<0.0001), ALT (OR = 0.268, P= 0.029) were associated with cirrhosis. An 8:2 split of the data to training and testing sets yielded an AUC of 0.836 and 0.848 for steatosis and cirrhosis, respectively. Unsupervised machine learning using K-means clustering yielded 2 clusters with 60% (598/995) Steatosis + 2.9%(29/995) Cirrhosis in [Cluster 1]{.underline} and 31%(1611/5124) Steatosis +  1.5%(80/5124) in [Cluster 2.]{.underline}

## Conclusion

In summary, we realize that BMI is a strong predictor for the occurence of hepatic steatosis and cirrhosis. Other variables which are associated with each include alt, alb, gender (cirrhosis) and a1c, tg and alt (steatosis).

We have also successfully created supervised (using random forest) and unsupervised (k means clustering and hierarchihcal clustering) models to predict occurence of steatosis and cirrhosis.
