---
title: "Sarai_Garcia_TMS_Analysis"
subtitle: "BMIN503/EPID600 Final Project"
author: "Sarai Garcia"
format: html
editor: visual
embed-resources: true
---

------------------------------------------------------------------------

Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers. Feel free to change the theme and other display settings, although this is not required.

### Overview

Give a brief a description of your project and its goal(s), what data you are using to complete it, and what two faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

```{r}
# This study investigates the efficacy of neuroimaging-guided repetitive transcranial magnetic stimulation (rTMS) in individuals with trauma-related depression. I am analyzing data from a study that was completed in 2018. However, so many clinical/behavioral measures were taken that there is a wealth of data to be analyzed. I spoke with my PI, Dr. Desmond Oathes, who directs the Center for Brain Imaging and Stimulation. It was a study of his design and his familiarity with the procedures and trends that had already been observed in the data made him extremely helpful to guide me through this analysis as a beginner. A data analyst in my lab, Julie Grier, as well as a staff scientist, Dr. Romain Duprat, helped with any issues I ran into with data cleaning and constructing my model, respectively. As her job is largely data analysis and organization, I learned about the complex process of data cleaning. There were particular issues with this data set due to the way it was collected in REDCap as separate items rather than repeared measures of the same item. I ran into a lot of issues making sure I only had unique values and that the clinical measures were recorded in the correct order. With Dr. Duprat I learned a lot about the importance of data structure/context and how these influence the choice of what model suits my data the best. I ended up using a linear mixed effects model which allowed me to compare differences in clinical outcomes between different predictor variables, but became increasingly more difficult to interpret as the number of predictors I was testing increased. 
```

### Introduction

```{r}
# Major Depressive Disorder (MDD) is a common sequelae of trauma and is highly comorbid with Post Traumatic Stress Disorder (PTSD). About 50% of those with PTSD also suffer from MDD. Repetitive transcranial magnetic stimulation (rTMS) is an established non-invasive tool for treating MDD (FDA cleared) and PTSD. This treatment modulates neural activity by delivering electrical stimulation to targeted brain regions. Neural circuits associated with affective disorders have characteristic abnormalities that might be further exacerbated by trauma experience. Recent data from our group show that using functional Magnetic Resonance Imaging (fMRI) to guide rTMS produces significantly stronger clinical effects than using traditional scalp-based targets. 

# This research aims to deepen our understanding of the use of imaging tools in non-invasive brain stimulation treatment of psychiatric disorders by 1) examining trauma preponderance and 2) manipulating brain activation with neuromodulation treatment. We hypothesized that fMRI-guided TMS would produce a greater benefit to patients than the standard targeting method. Targeting the delivery of TMS based on individualized readings of sgACC resting functional connectivity may be especially effective in reducing its hyperactivity. We also hypothesized that psychological tasks would differentially facilitate treatment outcomes. While preliminary data reveals no significant influence by task condition, participants treated with neuromodulation in the fMRI-guided target reported a significantly greater reduction in depression and PTSD symptoms compared to participants treated with stimulation in the standard target. Further analyses are being conducted on self-report ratings of positive and negative affectivity. Specifically, we are interested in how these aspects are differentially influenced by trauma preponderance. These findings emphasize the significance of imaging-guided neuromodulation in more effective treatments for MDD and PTSD diagnosed patients. 

# This research is interdisciplinary across neuroscience, psychiatry, and psychobiology. We utilize neuroscience methods to identify relevant neurocircuitry and administer TMS. We measure clinical outcomes and individual factors that influence TMS receptivity by introducing psychological tasks and measuring heart rate during TMS sessions. Though this was not a clinical trial, this study has implications in psychiatry and in practical applications of these individualized TMS methods to patient populations with treatment-resistant psychopathologies. Promising results of this study could demonstrate that individualized TMS is an effetive tool at improving affective ratings (acute mood) associated with depression and PTSD. 
```

Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

### Methods

Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why.

```{r}
# Data is still messy so I do not want to put it in yet. But, data cleaning/organization was complicated due to the way data was collected. Affective ratings were assessed through the Positive and Negative Affect Schedule Survey (PANAS), a clinical measure that allowed us to observe acute mood changes during a single visit as well as over the duration of the study. 
# There were 45 participants in total who participated in two phases of treatment. Each phase lasted 10 days with 2 TMS sessions per day. Within each phase participants were randomly assigned to the fMRI-guided TMS target or the Standard TMS target. They were additionally randomly assigned to receive a psychological task (n-back task) that would induce a higher state of arousal or simply a video that would induce a lower state of arousal.
# The PANAS was taken prior to TMS, during the 20 minute break between TMS sessions, following the psychological task introduced during the break, and again post TMS. We focused our analysis on the first phase of TMS and the pre TMS ratings of the PANAS, so as to avoid acute mood changes falsely attributed to the unpleasant physical experience of receiving TMS so as to more accurately collect mood inherent to their emotional state.
# It was a lot of data and most of my time was spent cleaning and restructuring it into a manner that was useful for me. 45 participants x 11 days of TMS (10 + baseline) x 4 times a day the PANAS was taken 
# My main question was to determine whether the fMRI-guided site reported more significant clinical improvement in affective ratings than the standard site. I also had exploratory analyses related to the influence of the psychological tasks as well as trauma experience on affective ratings improvement. My PI advised me to account for the influence of baseline score. Recent data shows that participants who have higher baseline scores typically have greater improvement.
# I conducted a linear mixed effects model that looked like this:
# lmer(PANAS SCORE ~ TIME * TMS TARGET SITE * TRAUMA EXPERIENCE * TASK CONDITION + BASELINE SCORE)
```

### Results

Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
