---
title: "BMIN503/EPID600 Project Template"
author: "Amelia Van Pelt"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

### Overview
Include a link to your final project GitHub repository

This research aims to determine reasons for poor vaccine uptake in Uganda. To accomplish this goal, this research proposes to visualize the geographic distribution of vaccine receipt across the regions and to assess the relationship between various demographic factors and rates of vaccination coverage in Uganda. This research will utilize the 2016 Demographic and Health Survey dataset, specifically the women's recoded dataset. Scholars in three disciplines provided guidance on the project: Irma Elo, PhD, MPA, Professor in the Department of Sociology, Luca Pesando, PhD, MSc, Post-doctoral Fellow in the Population Studies Center, and Vicky Tam, MA, resident expert in Geographic Information Systems at the University of Pennsylvania and the Children's Hostpial of Philadelphia. Pesando works extensively with the Demographic and Health Survey datasets, so he provided the data, data dictionary, and guidance on the quirks of Demographic and Health Survey data. For example, he explained the importance of the survey weights. Elo discussed suggestions on demographic variables to include in the analyses and potential hypotheses for the results including fertility. Tam introduced geospatial analyses and suggested implications.  

### Introduction 

According to the 2016 Demographic Health Survey (DHS), Uganda experiences low vaccine coverage. Approximately half of all children aged 12-23 months included in the survey received all of the basic vaccinations. Poor vaccine coverage causes significant consequences. A lack of immunological protection increases the susceptibility to disease. As a result, a greater potential for an outbreak of a vaccine-preventable disease (VPD) exists. The increase in the incidence of VPDs results in increased morbidity and mortality. Further, the effect of an outbreak of a VPD extends beyond the physical health of a population, for treatment of disease, compared to receipt of a vaccine, contributes more economic burden. 

Many factors influence vaccine uptake. The final report of the DHS provides a table that illustrates the relationship between some common variables, such as mother's education level and birth order, and the percentage of children who received various vaccines. However, vaccination uptake is a complex problem that extends beyond the simple analyses provided. For example, religion and access to healthcare influence vaccine uptake. Thus, to fully understand the complexity of vaccination coverage, research must propose an interdisciplinary approach. Experts in public health, sociology, population science, medicine, and many other disciplines can contribute perspectives to develop a more robust understanding of the issue. This research will draw on conversations with scholars in such disciplines to incorporate relevant variables for statistical analyses and data visualization. 


### Methods

One of the 2016 Demographic and Health Survey datasets was used for the analyses. The DHS program is funded by the United States Agency for International Development (USAID) and facilitates "nationally-representative household surveys" in over 90 countries. Officials utilize the data for monitoring and evaluating population, health, and nutrition indicators. Many recoded versions of the data exist, such as household recode, children's recode, and women's recode. For this research, the 2016 Uganda women's recoded dataset was analyzed, for it contains observations on children's vaccination status and mother's demographic variables for the same respondents. The dataset contains 18,506 observations and 4,966 variables. The large size of the dataset continuously crashed the program, so the unrelated variables were dropped outside of RStudio. In addition, the observations that corresponded to women with zero children were dropped as well, because the primary outcome of interest was child's vaccine receipt. The final dataset used for this project contained 13,740 observations. Data visualization and statistical analyses were performed. 


#### Loading the Dataset
```{r}
library(readxl)
womendataset <- read_excel("C:/Users/vanpel/Downloads/datasciencefinaldata.xls")
View(womendataset)
```

#### Additional Data Cleaning
```{r}
head(womendataset)

womendataset$'ever had vaccination'[womendataset$'ever had vaccination'=='no'] <- 0
womendataset$'ever had vaccination'[womendataset$'ever had vaccination'=='yes'] <- 1
womendataset$`ever had vaccination` <- as.numeric(womendataset$`ever had vaccination`)
class(womendataset$`ever had vaccination`)

womendataset$region[womendataset$region==0] <- "Kampala"
womendataset$region[womendataset$region==1] <- "Central 1"
womendataset$region[womendataset$region==2] <- "Central 2"
womendataset$region[womendataset$region==3] <- "Busoga"
womendataset$region[womendataset$region==4] <- "Bukedi"
womendataset$region[womendataset$region==5] <- "Bugishu"
womendataset$region[womendataset$region==6] <- "Teso"
womendataset$region[womendataset$region==7] <- "Karamoja"
womendataset$region[womendataset$region==8] <- "Lango"
womendataset$region[womendataset$region==9] <- "Acholi"
womendataset$region[womendataset$region==10] <- "West Nile"
womendataset$region[womendataset$region==11] <- "Bunyoro"
womendataset$region[womendataset$region==12] <- "Tooro"
womendataset$region[womendataset$region==13] <- "Ankole"
womendataset$region[womendataset$region==14] <- "Kigezi"

womendataset$'type of place of residence'[womendataset$'type of place of residence'==1] <- "urban"
womendataset$'type of place of residence'[womendataset$'type of place of residence'==2] <- "rural"

womendataset$`birth order number` <- as.factor(womendataset$`birth order number`)

```


#### Descriptive Statistics

To determine the distribution of responses for the demographic variables, simple descriptive statistics were performed. The frequency for each item for the categorical variables was calculated. 
```{r}
table(womendataset$region)
table(womendataset$`type of place of residence`)
table(womendataset$`highest educational level`)
table(womendataset$religion)
table(womendataset$literacy)
table(womendataset$`wealth index combined`)
table(womendataset$`birth order number`)
table(womendataset$`total children ever born`)
table(womendataset$`covered by health insurance`)
table(womendataset$`ever had vaccination`)
```


#### Visualization of the Distribution of Demographic Variables

To better visualize the distribution of respondents for the demomographic varaibles, bar charts were created. 
```{r}
library(dplyr)
library(ggplot2)

ggplot(data=womendataset, aes(x=womendataset$region)) +
  geom_bar (fill="blue") +
  ggtitle("Distribution of Respondents by Region")+
  labs(x="DHS Region", y="Number of Respondents", title="Distribution of Respondents by Region")

ggplot(data=womendataset, aes(x=womendataset$'type of place of residence')) +
  geom_bar (fill="purple") +
  ggtitle("Distribution of Respondents by Environment")+
  labs(x="Environment", y="Number of Respondents", title="Distribution of Respondents by Environment")

ggplot(data=womendataset, aes(x=womendataset$'highest educational level')) +
  geom_bar (fill="green") +
  ggtitle("Distribution of Respondents by Education Level")+
  labs(x="Highest Educational Level", y="Number of Respondents", title="Distribution of Respondents by Education Level")

ggplot(data=womendataset, aes(x=womendataset$religion)) +
  geom_bar (fill="orange") +
  ggtitle("Distribution of Respondents by Religion")+
  labs(x="Religion", y="Number of Respondents", title="Distribution of Respondents by Religion")

ggplot(data=womendataset, aes(x=womendataset$literacy)) +
  geom_bar (fill="red") +
  ggtitle("Distribution of Respondents by Literacy Level")+
  labs(x="Literacy Level", y="Number of Respondents", title="Distribution of Respondents by Literacy Level")

ggplot(data=womendataset, aes(x=womendataset$'wealth index combined')) +
  geom_bar (fill="yellow") +
  ggtitle("Distribution of Respondents by Wealth Index")+
  labs(x="Wealth Index", y="Number of Respondents", title="Distribution of Respondents by Wealth Index")

ggplot(data=womendataset, aes(x=womendataset$'birth order number')) +
  geom_bar (fill="pink") +
  ggtitle("Distribution of Respondents by Birth Order")+
  labs(x="Birth Order Number", y="Number of Respondents", title="Distribution of Respondents by Birth Order")

ggplot(data=womendataset, aes(x=womendataset$'total children ever born')) +
  geom_bar (fill="black") +
  ggtitle("Distribution of Respondents by Total Number of Children")+
  labs(x="Total Number of Children", y="Number of Respondents", title="Distribution of Respondents by Total Number of Children")

ggplot(data=womendataset, aes(x=womendataset$'covered by health insurance')) +
  geom_bar (fill="lightblue") +
  ggtitle("Distribution of Respondents by Health Insurance Status")+
  labs(x="Health Insurance Status", y="Number of Respondents", title="Distribution of Respondents by Health Insurance Status")
```

Respondents were sampled from each of the regions, and the distribution of the number of participants per region is not greatly skewed. Due to the landscape of Uganda, the majority of respondents lived in a rural environment compared to an urban environment. The distribution of the respondents across the religion categories was not equal. The majority of respondents identified with the Catholic religion and Anglican religion; six of the religions had less than ten respondents.Across the four educational categories, most participants indicated primary education as their highest level of completion. The literacy level varied between the extremes, for a large proportion of the sample could read complete sentences or could not read at all. However, the distribution of respondents across wealth index categories was fairly equal. Yet, the overwhelming majority of respondents were not covered by health insurance. Most respondents had five or fewer children, but the total number of children appears well distributed across the number categories. Due to the methodology of the data collection, the birth order variable reflects the total number of children.  


### Results

Regression Models

To test the association between the demographic variables and the receipt of a vaccine, bivariate and multivariate logistic regression models were created. Vaccination status was dichotomized into yes or no. All of the included predictors are statistically-significant at the 0.05 threshold level. 
```{r}
summary((glm(womendataset$`ever had vaccination`~womendataset$region, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$`type of place of residence`, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$`highest educational level`, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$religion, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$literacy, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$`wealth index combined`, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$`birth order number`, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$`total children ever born`, family=binomial())))
summary((glm(womendataset$`ever had vaccination`~womendataset$`covered by health insurance`, family=binomial())))

summary((glm(womendataset$`ever had vaccination`~womendataset$`highest educational level` + womendataset$region + womendataset$`type of place of residence`+ womendataset$religion + womendataset$literacy + womendataset$`wealth index combined` + womendataset$`birth order number` + womendataset$`total children ever born` + womendataset$`covered by health insurance`, family=binomial())))
```
In the bivariate models, living in an urban environment and being in the "richest" wealth category increased the odds of the child getting vaccinated. Conversely, the total number of children/birth order number, not having an education, and belonging to the "other" religion category decreased the odds of the child getting vaccinated. Literacy level and health insurance status was not a statistically-significant predictor of vaccine receipt. All of these relationships remain in the mutilivariate model except for the influence of wealth, for the wealth index variables are no longer a statistically-significant predictor.   


#### Geographical Visualization of the Vaccination Coverage

To visualize the spatial distribution of the vaccination coverage, static and interactive maps were created. The vaccination coverage variable represent the number respondents whose child ever received a vaccine divided by the total number of respondents in that region. The Uganda shape file was obtained directly from the DHS, because the dataset aggregated 112 districts into 15 regions. As a result, the normal geographic boundary map of Uganda did not match up with the data. Percentages of vaccine receipt are plotted on the maps.
```{r}
library(readxl)
vaccinerates <- read_excel("C:/Users/vanpel/Downloads/datasciencefinalspatial(1).xlsx")

library(dplyr)
library(sf)
ugandashape <- st_read('C:/Users/vanpel/Desktop/shps/sdr_subnational_boundaries.shp')


visualization <- inner_join(ugandashape, vaccinerates, by = "REGNAME")
head(visualization)


library(RColorBrewer)
library(viridis)

myPalette <- colorRampPalette(brewer.pal(9, "BuPu"))

my_theme <- function() {
  theme_minimal() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),
        legend.key.size = unit(0.8, "cm"),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16))
}

library(ggplot2)

ggplot() +
  geom_sf(data = visualization, aes(fill = percentvaccinatedmissing)) +
  my_theme() +                   
  ggtitle("Rates of Vaccination Coverage in Uganda") +
  scale_fill_gradientn(name = "Ever Receiving a Vaccine (%)",
                    colours = myPalette(100))
  

library(leaflet)

pal_fun <- colorNumeric("BuPu", NULL)

pu_message <- paste0(visualization$REGNAME, 
                     "<br>Vaccination Rate: ",
                     round(visualization$percentvaccinatedmissing, 1), "%")


leaflet(visualization) %>%
  addPolygons(stroke = FALSE,
              fillColor = ~pal_fun(percentvaccinatedmissing),
              fillOpacity = 0.5, smoothFactor = 0.5,
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  #addProviderTiles(providers$Stamen.Toner) %>%
  #addProviderTiles(providers$Esri.NatGeoWorldMap)
  addLegend("bottomright",
            pal=pal_fun,
            values=~percentvaccinatedmissing,
            title = 'Vaccination Coverage(%)',
            opacity = 1) %>%
  addScaleBar()
```
Qualitatively, the vaccination rates differ by region and ranged from 71.7% to 98.2%. The regions in the northeastern part of Uganda (Bugishu at 98.2% and Karamoja at 97.4%) have the higheset rates of vaccination coverage. The capital, Kampala, indicated by the small purple region in the center of the map and and Kigezi in the southwestern part of Uganda experience high coverage as well at 94% and 94.7%, respectively. The lowest vaccination coverage exists in Bunyoro (71.7%) in the western part of the country. The variation in the location of the highest and lowest percentage of covers illustrates the random distribution. 


#### Spatial Statistics

To assess the spatial auto-correlation quantitatively, Moran's Index was calculated. ArcGIS was utilized to calculate the value, so it is not possible to include the map outputs directly in this document. However, the input for the code was the following:  
```{r}
#input class= Uganda regions
#input field= vaccination coverage
#conceptualization of spatial relationships= inverse distance
#distance method= euclidean distance
#unit of measurement= meters
#standardization= none
#coincidence points= mean

```
The spatial auto-correlation was not statistically-significant (Moran's I=-0.06, z-score=0.05, p-value=0.96). Thus, the overall distribution of vaccine coverage is not clustered but instead random. However, there were two regions in which the pattern of distribution is significant (Teso as a high-high cluster and Tooro as a high-low outlier). 

Since the logistic regression models do not account for the spatial component of the data, geographically-weighted regression models were run for statistically-significant predictors of diarrhea. ArcGIS was used to perform the geographically-weighted regression models and produce the resulting maps of the beta coeficients. A sample of the code includes the following: 

```{r}
#input feature= Uganda regions
#dependent variable= vaccination coverage
#explanatory variable= percent urban
#kernel type= fixed
#bandwidth method= AICc
#number of neighbors= 30

#input feature= Uganda regions
#dependent variable= vaccination coverage
#explanatory variable= percent with no education
#kernel type= fixed
#bandwidth method= AICc
#number of neighbors= 30

#input feature= Uganda regions
#dependent variable= vaccination coverage
#explanatory variable= percent within richest category
#kernel type= fixed
#bandwidth method= AICc
#number of neighbors= 30

##multivariate model
#input feature= Uganda regions
#dependent variable= vaccination coverage
#explanatory variable= percent within richest category + percent with no education + percent within richest category
#kernel type= fixed
#bandwidth method= AICc
#number of neighbors= 30
```

Again, since these statistics were performed in ArcGIS, the map outputs are not included. However, they are available upon request, and samples were included in the presentation slides. If referencing the slides, the darker blue color indicates the regions in which the variable was a strong predictor for the child not getting vaccinated. The lighter green colors indicate the regions in which the variable was not as strong of a predictor of the child getting vaccinated. The adjusted R-squared value for the geographically-weighted regression between no education and vaccine coverage was -0.039. Not having an education was a strong predictor of the child not getting vaccinated in regions in the southwestern part of the country. The adjusted R-squared value for the geographically-weighted regression between urban place of residence and vaccine coverage was -0.074, and the adjusted R-squared value for the geographically-weighted regression between the richest wealth index and vaccine coverage was -0.070. Qualitatively, living in an urban evironment and being in the highest wealth index strongly predicted the child receiving a vaccine in the southwestern part of Uganda. In the multivariate model, the distribution of the strength of the coefficients differed. Not having an education and being the richest wealth index category was the strongest predictor in the western part of the country.   


#### Post-hoc Analyses for Discussion
```{r}

library(RColorBrewer)
library(viridis)

myPalette <- colorRampPalette(brewer.pal(9, "BuPu"))

my_theme <- function() {
  theme_minimal() +
  theme(axis.line = element_blank(),
        axis.text = element_blank(),
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),
        legend.key.size = unit(0.8, "cm"),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16))
}

library(ggplot2)

ggplot() +
  geom_sf(data = visualization, aes(fill = percentnoeducation)) +
  my_theme() +                   
  ggtitle("Rates of No Education") +
  scale_fill_gradientn(name = "No Education (%)",
                    colours = myPalette(100))

ggplot() +
  geom_sf(data = visualization, aes(fill = percentirchest)) +
  my_theme() +                   
  ggtitle("Rates of Richest Wealth Index") +
  scale_fill_gradientn(name = "Richest (%)",
                    colours = myPalette(100))

library(leaflet)

pal_fun <- colorNumeric("BuPu", NULL)

pu_message <- paste0(visualization$REGNAME, 
                     "<br>Richest Rate: ",
                     round(visualization$percentirchest, 1), "%")


leaflet(visualization) %>%
  addPolygons(stroke = FALSE,
              fillColor = ~pal_fun(percentirchest),
              fillOpacity = 0.5, smoothFactor = 0.5,
              popup = pu_message) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  #addProviderTiles(providers$Stamen.Toner) %>%
  #addProviderTiles(providers$Esri.NatGeoWorldMap)
  addLegend("bottomright",
            pal=pal_fun,
            values=~percentirchest,
            title = 'Richest Wealth Index (%)',
            opacity = 1) %>%
  addScaleBar()
  
```


### Conclusion
The high vaccination coverage seen in the population does not reflect the low coverage articulated by the literature in the introduction. This discrepancy results from the difference in the measurement of vaccination coverage. The literature details the coverage for all basic vaccines, which include four at various doses. This analysis examines the rate of a child ever receiving a vaccine, so the proportion of individuals in the population will likely be greater than the other measure. 

The random distribution of vaccination coverage in Uganda may result from the underlying distribution of the demographic variables in each region. For example, the regions in the northeastern part of the country house the lowest percentage of respondents in the richest wealth index, and Kampala has the highest percentage of participants in the richest wealth category. The relationship between wealth and vaccination coverage in the capital is not novel, as wealth increases access to medical care, such as immunizations. The inverse relationship in Karamoja may allude to public health interventions. Often, mass vaccination campaigns are completed in areas with the worst access to care. Thus, the wealth index may serve not only as a proxy to income level but also as a proxy to access to care, which may explain the higher vaccination rates in the region. This theory also applies to the spatial relationship seen between not having an education and high vaccination coverage.     

The "other" religion category statistically-significantly predicted vaccine status. This category reduced the odds of a child receiving a vaccine. Literature articulates the relationship between religion and vaccine refusal. Thus, qualiative analyses should explore the vaccination beliefs within this group. 

Limitations with these analyses exist. First, geographically-weighted regression is usually completed on a larger sample size. The 15 regions included in the models in this research may have influenced the results. Second, there was missing data on the child's vaccine status, which may have biased the analyses. However, the most inclusive variable was selected. Third, the shapefile provided by the DHS program distorts the geographical boundaries of Uganda by excluding the indentations of the major lakes (Lake Victoria, Lake Albert, and Lake Edward), which may influence the visual observations and conclusions drawn about the spatial relationships.  

The results highlight the areas for public health interventions aimed at increasing vaccination uptake. In the regions with the lowest coverage, efforts should be emphasized. Geographically-weighted regression results should inform the implementation of the intervention, for the interventions should be tailored based on the strongest predictor in the region. However, some of the predictors, such as education level, may be easier to address than others, such as type of residence. Further investigations through qualitative methods should explore whether or not each of the predictors is actually significant in the region.
