---
title: "Predicting Multi-Drug Resistance Status in Hospitalized Dogs"
author: "Sondra Calhoun"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation in class will be based on the contents of this document. Replace the title/name and text below with your own, but leave the headers.

### Overview
  The goal of this project is to develop and compare Random Forest and logistic regression models designed to predict whether or not a hospitalized dog is infected with multi-drug resistant (MDR) bacteria. MDR-bacteria pose a threat to other hospital patients and staff, so identifying high risk patients before the results of culture and sensitivity testing are available may reduce the risk to patients and staff.  This project will leverage two years of demographic, procedure, diagnosis, and perscription data for canines at a large tertiary veterinary clinic to predict canine MDR status. 
  
Faculty consulted for this project included 
  
In this section, give a brief a description of your project and its goal, what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.


### Introduction 
  Antimicrobial resistance in bacteria, particularly multi-drug resistance (MDR) is a growing concern in not only human but veterinary medicine. Animals with MDR infections have poor treatment outcomes, the antimicrobials needed to treat MDR infections are significantly more expensive for pet owners, and many MDR organisms are zoonotic, putiing not only other animals but humans at risk. When a patient's bacterial culture and sensitivity tests reveals MDR, various additional infectious disease control procedures are initiated to protect other animal patients and human staff. The ability to identify patients at high risk of MDR would allow for these procedures to be initiated earlier, reducing the spread of MDR bacteria and the risk to animal and human health.
  
  Dr. Shelley Rankin's experience and knowledge of culture and sensitivity testing as Chief of Clinical Microbiology at PennVet's Ryan Veterinary Hospital helped me develop the initial question for this project. My studies in veterinary medicine, epidemiology and the sociology of antimicrobial prescribing influenced the choice of the variables to include in the dataset.  Many veterinary studies are limited to small case-control or cohort studies and prospective clinical trials, largely due to a lack of electronic records at many clinics. This study will utilize a data set of approximately 40,000 canine admissions to a tertiary veterinary hospital.   The scale of the data for this project allows us to branch into "big data" approaches that are rare in veterinary medicine.  As such my methods are developed with the assistance of Dr. La Cava, a Data Sscientist post-doctroal fellow, and John Holmes, an faculty member in the Department of Epidemiology, Biostatistics, and Informatics. 

In the first paragraph, describe the problem addressed, its significance, and some background to motivate the problem.

In the second paragraph, explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.


### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

Retrieving the vet data (given as a .csv), and saving as a dataframe in R. 
```{r eval=FALSE}
vet<-read.csv("FullData.csv", header=TRUE, na.strings= "", stringsAsFactors = FALSE)
save(vet,file="vet.Rda")

```

Loading the dataframe and cleaing the data. 
```{r eval=TRUE}
library(dplyr)
library(knitr)
load("vet.Rda")

#Stripping the time and date from the MDR-Alert and fee code so that only procedures and visits prior to the MDR alert are included in the analysis. 
vet<-mutate(vet,MDRtime=MDRAert)
vet<-mutate(vet,MDRtime=(gsub('MDR-Alert! ','',MDRtime)))
vet$MDRtime<-strptime(vet$MDRtime, "%H:%M:%S %d %b %Y")
vet$MDRtime<-as.POSIXct(vet$MDRtime)
class(vet$MDRtime)
vet$MDRtime<-as.POSIXct(vet$MDRtime)

vet <- mutate(vet, feedate=paste(vet$PostDate, vet$PostTime, sep=" ")) 
vet$feedate<-strptime(vet$feedate, "%m/%d/%Y %H:%M:%S")
vet$feedate<-as.POSIXct(vet$feedate)
class(vet$feedate)



#Identifying the total number of patients with an MDR alert by year. Removing patients with an MDR alert prior to 2013 from the analsis.  
visit<-vet[!duplicated(vet$VisitID),]
visit$year = factor(format(visit$MDRtime,'%Y'))
vet$year = factor(format(vet$MDRtime,'%Y'))
summary(visit$year)

#Remove patients with MDR statuses prior to 2013
vet<-subset(vet, year=="2013"|year=="2014"|year=="2015" |year=="2016"| is.na(year))
visit<-subset(visit, year=="2013"|year=="2014"|year=="2015" |year=="2016"| is.na(year))

#Remove MDRtime for mutation--returning error otherwise"Error: Column `MDRtime` is a date/time and must be stored as POSIXct, not POSIXlt"
#var<-names(vet1) %in% c("MDRtime")
#vet2<-vet1[!var]
#visit2<-visit1[!var]

#Generate Case status
vet <-mutate(vet, case = ifelse(is.na(year)|year == "2015" | year== "2016" |year=="2008"|year=="2009"|year=="2010"|year=="2011"|year=="2012", 0,1))
visit <-mutate(visit, case = ifelse(is.na(year)|year == "2015" | year== "2016" |year=="2008"|year=="2009"|year=="2010"|year=="2011"|year=="2012", 0,1))
vet <- mutate(vet, case = factor(case, levels=c(0, 1), labels=c("No", "Yes"))) 
visit <- mutate(visit, case = factor(case, levels=c(0, 1), labels=c("No", "Yes"))) 
table(vet$case)
table(visit$case)

```
Of the 45,312 canine patients seen in 2013-2014, 423 had MDR alerts in their medical records.  Of these, 320 had MDR Alerts placed in the study period. The 68 patients with MDR alerts prior to 2013 will be removed from analysis. The 35 patients with MDR alerts after 2014 will be included in the analysis, but will be treated as non-MDR patients since during their hospital stays during the study period they did not have an MDR. The final number of MDR patient admissions during the study period will be 320, out of 45,244 admissions. 

 Remove all the procedure codes after the MDR alert time from the patients that had an MDR (since procedures after the MDR alert are not relevant to predicting MDR).  
```{r eval=TRUE}

vet<-subset(vet,vet$case=="No"|(vet$case=="Yes"&!(vet$feedate < vet$MDRtime)))

```

Create new variables to be used in the analysis: antibiotic use (y/n), antibiotic prescriptions per visit, length of stay, month, admit service, sex, referred (y/n), surgery patient(y/n), Philadelphia residence (y/n), admit time (from opening of hospital), age at admission.  Each will be calculated at the visit level. 

```{r eval=TRUE}
#Age
vet<-rename(vet, dob=MDR)
vet$dob<-strptime(vet$dob, "%m/%d/%Y")
vet$dob<-as.POSIXct(vet$dob)
vet$AdmitDate<-strptime(vet$AdmitDate, "%d%b%Y")
vet$AdmitDate<-as.POSIXct(vet$AdmitDate)
vet<-mutate(vet, age=as.numeric(difftime(AdmitDate, dob, 
                                        unit="weeks"))/52.25)

```


### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
