---
title: "BMIN503/EPID600 Project Template"
author: "Dominique Medaglio"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

### Overview
Give a brief a description of your project and its goal(s), what data you are using to complete it, and what three faculty/staff in different fields you have spoken to about your project with a brief summary of what you learned from each person. Include a link to your final project GitHub repository.

> Overview: Medication nonadherence amongst diabetics is a widespread issue in the United States, and differences in insurance coverage (which differs by residential location) may be a contributing factor. The purpose of this project is to determine if there are differences in cost-related nonadherence (CRN) prevalence based on urban/rural residence. I will acheive this aim by creating static maps and logistic regression models based on adult respondents of the 2019 National Health Interview Survey (NHIS) who are diabetic.

>I have spoken to Mary Regina Boland, Robert Gross, Sean Hennessy, and Karen Glanz. With Mary Regina, I learned about special considerations for using zip code data, which is what I'll be using for my maps. Robert and Sean provided guidance on how to interpret CRN measures appropriately. Karen provided overall advice on adherence behaviors as a part of self-care behaviors, and variables that I should consider for my model.

>The link to the final GitHub repository can be found here: https://github.com/dmedaglio/BMIN503_Final_Project


### Introduction 
Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.

>Introduction: Medication adherence is a mainstay to the management of chronic conditions. Medication error, including the non-initiation of life-saving treatment, is one of the top reasons why patients are readmitted to the hospital. Unfortunately, medication nonadherence is very common; it is estimated that 30-50% of medications are not taken as prescribed.

>Cost is a common reason why patients to not their medications as they should. This can especially be compounded if patients take multiple medications to manage a condition, such as common with diabetes. Insurance coverage can be important to help many afford their medications, yet sufficient coverage may vary by where patients live. In urban areas, one is more likely to benefit from more provider competition in the Marketplace and have lower premiums. In rural areas, such competition may not exist, resulting in higher premiums on average. It is unknown if urban/rural residence also has an impact on the affordability of prescription medications.

>This research question requires an interdisciplinary approach, given the need for geospacial approaches, pharmacoepi expertise, and an understanding of health policy. It will be important to accurately interpreret NHIS survey responses, as CRN is only one form of medication nonadherence. Additionally, the NHIS is administered using geographically clustered sampling techniques, so static maps will likely zoom in on these clusters. Finally, an understanding of policies around the health insurance marketplace will be needed to interpret access to pharmacy benefits. 


### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 


### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
```{r}
library(readxl)
adult19 <- read_excel("C:/Users/dmcom/Documents/NHISadult2019.xlsx")

```


#consider using gtsummary to create a Table 1 and univariate analyses
