---
title: "Final_Project_Extras"
author: "Nawar Naseer"
date: "11/16/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description
I am using this document to store extra code chunks that yielded in results/maps/graphs that I didn't think were useful for the final report, but wanted to maintain a copy of in case I changed my mind. Where possible, I indicate why I didn't think it was helpful to include. 

Create a map of change in percent population with low access to stores from 2010 to 2015. 
```{r, eval=TRUE}
#Create new dataframe with variables of interest
LA2015change <- Access %>%
  select(FIPS, PCH_LACCESS_POP_10_15)
LA2015change$GEO_ID <- paste0("0500000US", LA2015change$FIPS)
LA2015changemap <- inner_join(counties, LA2015change, by="GEO_ID")

myPalette <- colorRampPalette(brewer.pal(9, "YlOrRd"))

my_theme <- function() {
  theme_minimal() +                                  
  theme(axis.line = element_blank(),                 
        axis.text = element_blank(),                 
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),  
        legend.key.size = unit(0.8, "cm"),          
        legend.text = element_text(size = 16),       
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 22))      
}

ggplot() +
  geom_sf(data = LA2015changemap) + 
  geom_sf(data = LA2015changemap, aes(fill = PCH_LACCESS_POP_10_15), lwd = 0) +
  my_theme() +
  ggtitle("County-Level Percent Change in Population with Low Access to Stores in the US 2010-2015") +
  scale_fill_gradientn(name = "Percent Change in Population", colours = myPalette(100))
```

Create a map of percent change in population that had low access to stores and low income between 2010 and 2015.
```{r, eval=TRUE}
#Create new dataframe with variables of interest
LAIchange <- Access %>%
  select(FIPS, PCH_LACCESS_LOWI_10_15)
LAIchange$GEO_ID <- paste0("0500000US", LAIchange$FIPS)
LAIchangemap <- inner_join(counties, LAIchange, by="GEO_ID")

myPalette <- colorRampPalette(brewer.pal(9, "YlOrRd"))

my_theme <- function() {
  theme_minimal() +                                  
  theme(axis.line = element_blank(),                 
        axis.text = element_blank(),                 
        axis.title = element_blank(),
        panel.grid = element_line(color = "white"),  
        legend.key.size = unit(0.8, "cm"),          
        legend.text = element_text(size = 16),       
        legend.title = element_text(size = 16),
        plot.title = element_text(size = 22))      
}

ggplot() +
  geom_sf(data = LAIchangemap) + 
  geom_sf(data = LAIchangemap, aes(fill = PCH_LACCESS_LOWI_10_15), lwd = 0) +
  my_theme() +
  ggtitle("County-Level Percent Population with Low Income and Low Access to Stores in the US in 2015") +
  scale_fill_gradientn(name = "Percent Population", colours = myPalette(100))
```
Visually not as helpful since there were a few outlier counties with large changes, whereas most places had relatively little change.

```{r, eval=TRUE}
#Transposing the dataframe to make the bar graphing easier
Access_age_PHLBAL_t <- data.frame(t(Access_age_PHLBAL))
Access_age_PHLBAL_t %>%
  rename(Baltimore = X1, Philadelphia = X2)

AAAt <- Access_age_PHLBAL_t[-c(1),] 
```
Ended up doing this manually.

```{r, eval=TRUE}
#Create new dataframe with variables of interest
SNAP2016 <- Stores %>%
  select(FIPS, SNAPSPTH16)
SNAP2016$GEO_ID <- paste0("0500000US", SNAP2016$FIPS)
SNAP2016map <- inner_join(counties, SNAP2016, by="GEO_ID")

ggplot() +
  geom_sf(data = SNAP2016map) + 
  geom_sf(data = SNAP2016map, aes(fill = SNAPSPTH16), lwd = 0) +
  my_theme() +
  ggtitle("County-Level SNAP-authorized stores in the US, 2016") +
  scale_fill_gradientn(name = "SNAP-authorized stores/1000 population", colours = myPalette(100))
```

```{r, eval=TRUE}
#Create new dataframe with variables of interest
SNAP2016PA <- Stores %>%
  select(FIPS, State, SNAPSPTH16) %>%
  filter(State=="PA")
SNAP2016PA$GEO_ID <- paste0("0500000US", SNAP2016PA$FIPS)
SNAP2016PAmap <- inner_join(counties, SNAP2016PA, by="GEO_ID")

ggplot() +
  geom_sf(data = SNAP2016PAmap) + 
  geom_sf(data = SNAP2016PAmap, aes(fill = SNAPSPTH16), lwd = 0) +
  my_theme() +
  ggtitle("County-Level SNAP-authorized stores in Pennsylvania, 2016") +
  scale_fill_gradientn(name = "SNAP-authorized stores/1000 population", colours = myPalette(100))
```
Kept the dataframe but changed to interactive map. 

Realized that if I have an interactive map at the nation level, then I don't need a separate PA map. Keeping it here incase I decide to change the nation-wide maps to static and need the PA info back.

PA-specific - SNAP stores
```{r, eval=TRUE}
#Create new dataframe with variables of interest
SNAP2016PA <- Stores %>%
  select(FIPS, State, County, SNAPSPTH16) %>%
  filter(State=="PA")
SNAP2016PA$GEO_ID <- paste0("0500000US", SNAP2016PA$FIPS)
SNAP2016PAmap <- inner_join(counties, SNAP2016PA, by="GEO_ID")

pu_SNAP2016PA <- paste0(SNAP2016PAmap$County, ",", " ", SNAP2016PAmap$State,  # paste0 to append tract name with other relevant text
                     "<br>SNAP-authorized stores/1,000 population, 2016: ", round(SNAP2016PAmap$SNAPSPTH16, 1))

# Basic leaflet map
leaflet(SNAP2016PAmap) %>%
  addPolygons(stroke = FALSE, # remove polygon borders
              fillColor = ~pal_fun2(SNAPSPTH16), # addPolygons requires variable and a pass to a palette function  
              fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
              popup = pu_SNAP2016PA) %>%                # add a popup message
  addTiles() %>%                                         # default basemap
  addLegend("bottomright",                           # location of legend
            pal=pal_fun2,                             # palette function
            values=~SNAPSPTH16,                 # variable to be passed to palette function
            title = 'SNAP-authorized stores/1,000 population, 2016',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```

PA-specific WIC stores
```{r, eval=TRUE}
#Create new dataframe with variables of interest
WIC2012PA <- Stores %>%
  select(FIPS, State, County, WICSPTH12) %>%
  filter(State=="PA")
WIC2012PA$GEO_ID <- paste0("0500000US", WIC2012PA$FIPS)
WIC2012PAmap <- inner_join(counties, WIC2012PA, by="GEO_ID")

pu_WIC2012PA <- paste0(WIC2012PAmap$County, ",", " ", WIC2012PAmap$State,  # paste0 to append tract name with other relevant text
                     "<br>WIC-authorized stores/1,000 population, 2016: ", round(WIC2012PAmap$WICSPTH12, 1))

# Basic leaflet map
leaflet(WIC2012PAmap) %>%
  addPolygons(stroke = FALSE, # remove polygon borders
              fillColor = ~pal_fun2(WICSPTH12), # addPolygons requires variable and a pass to a palette function  
              fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
              popup = pu_WIC2012PA) %>%                # add a popup message
  addTiles() %>%                                         # default basemap
  addLegend("bottomright",                           # location of legend
            pal=pal_fun2,                             # palette function
            values=~WICSPTH12,                 # variable to be passed to palette function
            title = 'WIC-authorized stores/1,000 population, 2012',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```

PA-specific Grocery stores
```{r, eval=TRUE}
#Create new dataframe with variables of interest
GroceryPA2014 <- Stores %>%
  select(FIPS, State, County, GROCPTH14) %>%
  filter(State=="PA")
GroceryPA2014$GEO_ID <- paste0("0500000US", GroceryPA2014$FIPS)
GroceryPA2014map <- inner_join(counties, GroceryPA2014, by="GEO_ID")

# Pop-up message
pu_GroceryPA2014 <- paste0(GroceryPA2014map$County, ",", " ", GroceryPA2014map$State,  # paste0 to append tract name with other relevant text 
                     "<br>Grocery stores/1,000 population, 2014: ", round(GroceryPA2014map$GROCPTH14, 1))

# Basic leaflet map
leaflet(GroceryPA2014map) %>%
  addPolygons(stroke = FALSE, # remove polygon borders
              fillColor = ~pal_fun2(GROCPTH14), # addPolygons requires variable and a pass to a palette function  
              fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
              popup = pu_GroceryPA2014) %>%                # add a popup message
  addTiles() %>%                                         # default basemap
  addLegend("bottomright",                           # location of legend
            pal=pal_fun2,                             # palette function
            values=~GROCPTH14,                 # variable to be passed to palette function
            title = 'Grocery stores/1,000 population, 2014',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```

PA-specific - SNAP redemption
```{r, eval=TRUE}
#Create new dataframe with variables of interest
SNAPRed2016PA <- SNAPRed2016 %>%
  filter(State=="PA")
SNAPRed2016PAmap <- inner_join(counties, SNAPRed2016PA, by="GEO_ID")

# Pop-up message
pu_SNAPRed2016PA <- paste0(SNAPRed2016PAmap$County, ",", " ", SNAPRed2016PAmap$State,  # paste0 to append tract name with other relevant text 
                     "<br>SNAP Redemption/SNAP-authorized stores, 2016: ", round(SNAPRed2016PAmap$REDEMP_SNAPS16, 1))

# Basic leaflet map
leaflet(SNAPRed2016PAmap) %>%
  addPolygons(stroke = FALSE, # remove polygon borders
              fillColor = ~pal_fun2(REDEMP_SNAPS16), # addPolygons requires variable and a pass to a palette function  
              fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
              popup = pu_SNAPRed2016PA) %>%                # add a popup message
  addTiles() %>%                                         # default basemap
  addLegend("bottomright",                           # location of legend
            pal=pal_fun2,                             # palette function
            values=~REDEMP_SNAPS16,                 # variable to be passed to palette function
            title = 'SNAP Redemption/SNAP-authorized stores, 2016',                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```

##### Farmers' markets
Most of the numbers were 0, although it's illuminating that that's the case. but doesn't sound correct. Philadelphia definitely has lots of farmers' markets. 
```{r, eval=TRUE}
#Create new dataframe with variables of interest
Farmers2016 <- Local %>%
  select(FIPS, State, County, FMRKTPTH16)
Farmers2016$GEO_ID <- paste0("0500000US", Farmers2016$FIPS)
Farmers2016map <- inner_join(counties, Farmers2016, by="GEO_ID")

# Pop-up message
pu_Farmers2016 <- paste0(Farmers2016map$County, ",", " ", Farmers2016map$State,  # paste0 to append tract name with other relevant text 
                     "<br>Farmers' market/1,000 population, 2016: ", round(Farmers2016map$FMRKTPTH16, 1))

# Basic leaflet map
leaflet(Farmers2016map) %>%
  addPolygons(stroke = FALSE, # remove polygon borders
              fillColor = ~pal_fun2(FMRKTPTH16), # addPolygons requires variable and a pass to a palette function  
              fillOpacity = 0.5, smoothFactor = 0.5, # increase opacity and resolution
              popup = pu_Farmers2016) %>%                # add a popup message
  addTiles() %>%                                         # default basemap
  addLegend("bottomright",                           # location of legend
            pal=pal_fun2,                             # palette function
            values=~FMRKTPTH16,                 # variable to be passed to palette function
            title = "Farmers' market/1,000 population, 2016",                  # legend title
            opacity = 1) %>%                         # legend opacity (1 = completely opaque)
  addScaleBar()
```


