---
title: "BMIN503/EPID600 Project Template"
author: "Your Name"
output: 
  html_document:
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

```{r eval = FALSE}
This is a template.

```

### Overview
The goal of this project is to use a randomized phase III trial study CALGB 50303 , "Rituximab and Combination Chemotherapy in Treating Patients With Diffuse Large B-Cell Non-Hodgkin's Lymphoma",  to compare the efficacy of rituximab when given together with the two conventional chemotherapy regimens in treating stage II to IV DLBCL patients. 


### Introduction 
Diffuse large B-cell lymphomas (DLBCLs), accounting for 30% to 40% of adult non-Hodgkin lymphomas (NHLs), are the most common histologic subtype of lymphoid neoplasms in the United States (1). Conventional chemotherapy regimens for DLBCL include CHOP (cyclophosphamide, doxorubicine, vincristine, prednisone) and EPOCH (etoposide, prednisone, vincristine, cyclophosphamide, doxorubicin), but the complete response rates are not satisfactory (1). To further improve the efficacy of these combination chemotherapies, rituximab has been introduced additionally in frontline therapies, because its monoclonal antibody properties allows it to prevent cancer cell proliferation in many different ways (2). The modified treatments are called EPOCH-R and R-CHOP respectively. 

Describe the problem addressed, its significance, and some background to motivate the problem.

Explain why your problem is interdisciplinary, what fields can contribute to its understanding, and incorporate background related to what you learned from meeting with faculty/staff.


### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 


### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
