---
title: "BMIN503/EPID600 Project Template"
author: Tara Friebel Klingner
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
    
---

### Overview: 
For my final project I will be using looking at biological and demographic factors associated with mortality from breast cancer among patients in a cohort from Botswana.  Women enrolled into the study between 2015 & 2019.  Variables of interest are HIV status, age at diagnosis, stage of tumor at diagnosis, BMI, hormone receptor status, distance from treatment center, marital status and smoking.

### Introduction: 
Cancer is a burden around the globe.   Looking at the epidemiology of cancer worldwide, there are 18 million incident cancer cases per year and 9.6 million deaths annually.  85% of these deaths occur in low to middle income countries and Africa is disproportionately affected, as is  most of sub-Saharan Africa.  As motivation for this project, in recognizing the high burden of morbidity and mortality of this disease in Africa and that there is a paucity of research being done focusing on breast cancer incidence and mortality  Africa, particularly in Botswana, I will be studying factors associated with mortality from breast cancer.  Identifying and reporting on these characteristics will allow  a better understanding of cancer in Botswana and will help aid in the direction of resources and help direct future educational and screening interventions. I will use logsitic regression, chi2 tests and other statistical techniques to examine at these factors.

This is a multidisciplinary because it pulls in expertise from chronic diseases, medical oncology, infectious diseases, global oncology, and epidemiology. For a clinical understanding of breast cancer in Botswana I spoke with Surbhi Grover, MD, a radition oncologist in Botswana.  To have an understanding of the infectous disease burden, particulrly HIV in Botswana, I spoke with Erle Roberston, PhD, and for global oncology and epidemiology expertise I spoke with Timothy Rebbeck, PhD. 

### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

The data I will be using is from the Botswana cancer cohort.  The infrastructure for this cohort formed from a joint collaboration between UPenn, Botswana's Ministry of Health, and the University of Botswana.  The health care system in Botswana is government funded and provides free or low cost healthcare to all citizens.  The system is decentralized and is comprised of a network of facilities from health posts to referral hospitals.  Referral hospitals are the only facilities equipped to care for specialized diseases like cancer.  The three referral hospitals are Princess Marina Hospital (PMH), Gaborone Private Hospital (GPH), and Nyangabgwe Referral Hospital (NRH). PMH and GPH are located in the capital of Gaborone and NRH is in Francistown.  GPH is the only hospital in the country equipped with a
radiation oncology facility and capable to providing chemo-radiation treatment for cancer  Patients for this study will be recruited from PMH and GPH in Gaborone.  After consent, patients complete an initial consult questionnaire and if undergoing treatment will also complete an On-Site Treatment Visit (OTV) questionnaire at each visit.  Patients are then followed prospectively every three months until death.  Eligible subjects for this proposal are females, 18 years or older, with a pathologically confirmed diagnosis of cervical cancer. Pregnant patients are ineligible.  Study data are collected and managed using REDCap, a secure web-based software platform designed to support data capture, hosted at U Penn.  There are > 2,000 Botswana residents with 17 different cancer types in this cohort, comprising the population for the proposed study.    

Statistical Analyses will examine differences in mortality and characteristics of breast cancer patients diagnosed from January 2015 through December 2019.  Univariate analysis will be used to examine the differences between variables of interest, such as HIV, age, BMI, stage at diagnosis, hormone receptor status, distance from cancer care center, marital status and smoking history.  Descriptive statistics will be presented to quantify central tendencies of the data as well as the variability, including the mean of covariates mentioned for breast malignancies presented in the follow-up period.  Additional descriptive will include standard deviation, variance, quartiles and minimum and maximum variables for each variable to evaluate the dispersion of data within the BCC.  Results will be presented in graphs, plots, and tables to visually demonstrate the distribution and emphasize key features. In addition to chi-square tests for binary and categorical variables, univariate logistic regression, multivariate logistic regression will be used to estimate risk ratios for stage III or IV versus stage I or II stage disease.   Covariates found to be significant in the univariate analysis (p<0.20) will be retained in the model for further analysis and possible confounding and interaction between covariates will be explored.  

First, I imported the Botswana Cancer Cohort relevant data, using a de-identified and HIPPA compliant .csv file.  Next, I will select the outcome and variables of interest, as well as perform and relevant coding and labeling of the variables.  For each binary or categorical variable I created a table to see the quantiative statistcs for each variable.  For continous variables I obtained the mean and quartile ranges for each. Also, using HIV as a main exposure, I explored relationsips between HIV affected breast cancer patients and HIV unaffacted breast cancer patients with the covariates mentioend above.

When studying health in Botswana, it is important to recognize the high prevalence of HIV in this population.  In addition, there are known HIV associated cancers.  In this cohort, 28% of the population is HIV positive.  Breast cancer is not known to be associated with HIV.  However, it remains important to account for this co-morbidity in patients as it may affect their diagnosis, treatment and survival from cancer.  

---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
#from Botswana Cancer Cohort
getwd()
library(readxl)
library(tidyverse)
library(knitr)
library(ggplot2)
library(readr)
library(rmarkdown)


#Import the Botswana Cancer Cohort csv file
final_project_BCC <- read_csv("final_project_BCC.csv")
#head(final_project_BCC)


library(xtable)
#data(final_BCC)
#print(xtable(head(final_project_BCC, 10)))
kable(head(final_project_BCC), format="pandoc", caption="Variables in the BCC dataset")
colSums(!is.na(final_project_BCC))

```

####Variables & Outcome
From the main dataset, I selected the outcome of interest, mortality, and variables of interest to keep in this analysis.  I also assigned labels to some numeric variables.


```{r}


outcome_exposures<- final_project_BCC %>%
           select("record_id", "marital_status", "smoker", "bmi", "age", "age_group", "hiv", "yearofvisit", "stage_new", "stage_high", "dead", "distance", "er", "pr", "her2", "tnb_1", "hormone") %>%
  mutate(hiv = factor(hiv, levels = c(0, 1), labels = c("HIV unaffected", "HIV affected"))) %>%
  mutate(age_group = factor(age_group, levels = c(1, 2, 3), labels = c("<50", "50-70", ">70"))) %>%
  mutate(stage_high = factor(stage_high, levels = c(0, 1), labels = c("early stage", "late stage"))) %>%
  mutate(dead = factor(dead, levels = c(0, 1), labels = c("Alive", "Deceased"))) %>%
  mutate(smoker = factor(er, levels = c(0, 1), labels = c("Never Smoker", "Ever Smoker"))) %>%
  mutate(er = factor(er, levels = c(0, 1), labels = c("negative", "positive"))) %>%
  mutate(pr = factor(pr, levels = c(0, 1), labels = c("negative", "positive"))) %>%
  mutate(her2 = factor(her2, levels = c(0, 1), labels = c("negative", "positive"))) %>%
  mutate(tnb_1 = factor(tnb_1, levels = c(0, 1), labels = c("negative", "positive"))) %>%
  mutate(hormone = factor(hormone, levels = c(1, 2, 3, 4, 5, 6 ,7), labels = c("ER+", "ER/PR+", "ER/PR/Her2+", "PR+", "PR/Her2+", "Her2+", "tnb"))) 

kable(head(outcome_exposures), format="pandoc", caption="Variables in the BCC dataset")
```

```{r}
#filter out missing data
no_missing<-outcome_exposures %>%
  filter(complete.cases(outcome_exposures))

kable(head(no_missing), format="pandoc", caption="Variables in the BCC dataset")

```

####Exploratory analysis
First, I did an exploratory analysis. I took a look at individual variables, then a look at bivariate relationships.  Visual plots and quantitative statistics for all variables of interest.  For bivariate/categorical variables I used bar plots for visual display.
```{r}
library(RColorBrewer)
library(viridis)
library(ggplot2)
library(devtools)
library(cowplot)

#For dichotomous variables, tables of outcome and variable of interest:
table(outcome_exposures$hiv)
ggplot(data = outcome_exposures, aes(x = hiv)) + 
    geom_bar(fill=" grey") +
  ggtitle(paste("HIV status")) 

table(outcome_exposures$age_group)
ggplot(data = outcome_exposures, aes(x = age_group)) +
    geom_bar(fill="dark grey") +
  ggtitle(paste("Age Groups")) 

table(outcome_exposures$stage_high)
ggplot(data = outcome_exposures, aes(x = stage_high)) + 
    geom_bar(fill="blue") +
  ggtitle(paste("Early Stage (I/II) vs. Late Stage (III/IV)")) 

table(outcome_exposures$er)
ggplot(data = outcome_exposures, aes(x = er)) +
    geom_bar(fill="light blue") +
  ggtitle(paste("Estrogen Receptor")) 

table(outcome_exposures$pr)
ggplot(data = outcome_exposures, aes(x = pr)) +
    geom_bar(fill="grey") +
  ggtitle(paste("Progesterone Receptor"))

table(outcome_exposures$her2)
ggplot(data = outcome_exposures, aes(x = her2)) +
    geom_bar(fill="dark grey") +
  ggtitle(paste("Her2 Receptor"))

table(outcome_exposures$hormone)
ggplot(data = outcome_exposures, aes(x = hormone)) +
    geom_bar(fill="blue")+
  ggtitle(paste("Hormone status")) 

table(outcome_exposures$marital_status)
ggplot(data = outcome_exposures, aes(x = marital_status)) +
    geom_bar(fill="light blue") +
  ggtitle(paste("Marital status")) 

table(outcome_exposures$smoker)
ggplot(data = outcome_exposures, aes(x = smoker)) +
    geom_bar(fill="grey") +
  ggtitle(paste("Smoking History ")) 

table(outcome_exposures$dead)
ggplot(data = outcome_exposures, aes(x = dead)) + 
    geom_bar(fill="dark grey") +
  ggtitle(paste("Breast Cancer Mortality"))
    
```

####Visual plots and quantitative statistics of variables of interest:
To get a sense of the distribution of a continuous variable, I used histograms and density plotsand obtained the mean value of each continuous variable.




```{r, eval = TRUE}


summarise(outcome_exposures, age = mean(age, na.rm=TRUE))
ggplot(data = outcome_exposures, aes(x = age)) +
    geom_histogram(binwidth=5, fill="grey") +
  ggtitle(paste("Age distribution"))


#Density plot
ggplot(data = outcome_exposures, aes(age)) + 
    geom_histogram(aes(y = ..density..), breaks = seq(0, 80, 1), color = "blue", fill = "blue", alpha = 0.7) +
    geom_density(color = "red")

summarise(outcome_exposures, bmi = mean(bmi, na.rm=TRUE))
ggplot(data = outcome_exposures, aes(x = bmi)) +
    geom_histogram(binwidth=5, fill="light blue")+
  ggtitle(paste("BMI distribution"))


#Density plot
ggplot(data = outcome_exposures, aes(bmi)) + 
    geom_histogram(aes(y = ..density..), breaks = seq(0, 80, 1), color = "blue", fill = "blue", alpha = 0.7) +
    geom_density(color = "red")


summarise(outcome_exposures, distance = mean(distance, na.rm=TRUE))
ggplot(data = outcome_exposures, aes(x = distance)) +
  geom_histogram(binwidth=5, color="blue") +
  ggtitle(paste("Euclidian Distance from Gaborone"))


#Density plot
ggplot(data = outcome_exposures, aes(distance)) + 
    geom_histogram(aes(y = ..density..), breaks = seq(0, 80, 1), color = "blue", fill = "blue", alpha = 0.7) +
    geom_density(color = "red")
 
  
```
####Exploring  differences between HIV affacted and HIV unaffected individuals.
I further explored visual and quantitaive relationships for binary/categorical variable using barplots that were split by  HIV status. I used the "fill" option for proportions to  visually compare the two groups.

```{r}

#exploring visual relation ship between main variable (HIV) and covariates of interest
ggplot(data = outcome_exposures, aes(x = hiv, fill=age_group)) + 
    geom_bar(position="fill") +
  ggtitle(paste("Age group and HIV"))

summary((glm(hiv ~ age_group, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ age_group, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(x = hiv, fill=stage_high)) + 
    geom_bar(position="fill") +
  ggtitle(paste("Stage and HIV"))
summary((glm(hiv ~ stage_high, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ stage_high, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(x = hiv, fill=hormone)) + 
    geom_bar(position="fill") +
  ggtitle(paste("Hormone Receptor Status & HIV"))
summary((glm(hiv ~ hormone, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ hormone, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(x = hiv, fill=marital_status)) + 
    geom_bar(position="fill") +
  ggtitle(paste("Marital Status"))
summary((glm(hiv ~ marital_status, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ marital_status, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(x = hiv, fill=smoker)) + 
    geom_bar(position="fill") +
  ggtitle(paste("Smoking and HIV"))

summary((glm(hiv ~ smoker, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ smoker, data = outcome_exposures, family = binomial()))) #Odds ratio
```



####Boxplots for continuous predictors

```{r}

ggplot(data = outcome_exposures, aes(hiv, age)) + 
    geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("HIV Status of all Patients by Age"))
summary((glm(hiv ~ age, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ age, data = outcome_exposures, family = binomial()))) #Odds ratio


ggplot(data = outcome_exposures, aes(hiv, bmi)) + 
    geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("BMI and HIV"))

summary((glm(hiv ~ bmi, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ bmi, data = outcome_exposures, family = binomial()))) 

ggplot(data = outcome_exposures, aes(hiv, distance)) + 
     geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("Distance and HIV"))
summary((glm(hiv ~ distance, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ distance, data = outcome_exposures, family = binomial()))) 


##Exploring Hormone Status across the age distribution of the cohort
ggplot(data = outcome_exposures, aes(er, age)) + 
     geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("Estrogen and Age"))
summary((glm(er ~ age, data = outcome_exposures, family = binomial())))
exp(coef(glm(er ~ age, data = outcome_exposures, family = binomial()))) 

ggplot(data = outcome_exposures, aes(pr, age)) + 
     geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("Progesterone and Age"))
summary((glm(pr ~ age, data = outcome_exposures, family = binomial())))
exp(coef(glm(pr ~ age, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(her2, age)) + 
     geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("Her2 and Age"))

summary((glm(her2 ~ age, data = outcome_exposures, family = binomial())))
exp(coef(glm(her2 ~ age, data = outcome_exposures, family = binomial()))) 

ggplot(data = outcome_exposures, aes(hormone, age)) + 
     geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("Hormone receptor Status and Age"))
summary((glm(hormone ~ age, data = outcome_exposures, family = binomial())))
exp(coef(glm(hormone ~ age, data = outcome_exposures, family = binomial()))) 

```
####Histogram
To get a sense of the distribution of a continuous variable, I used histograms.
```{r}

ggplot(data = outcome_exposures, aes(x = age, fill=hiv)) +
    geom_histogram(binwidth=5)+ 
  ggtitle(paste("Age and HIV"))
summary((glm(hiv ~ age, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ age, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(age, fill=hiv)) + 
    geom_histogram(aes(y = ..density..), breaks = seq(0, 80, 1), alpha = 0.7) +
    geom_density(color = "red") +
  ggtitle(paste("Age and HIV"))

ggplot(data = outcome_exposures, aes(x = bmi, fill=hiv)) +
    geom_histogram(binwidth=5) +
  ggtitle(paste("BMI and HIV"))
summary((glm(hiv ~ bmi, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ bmi, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(bmi, fill=hiv)) + 
    geom_histogram(aes(y = ..density..), breaks = seq(0, 80, 1),  alpha = 0.7) +
    geom_density(color = "red") +
  ggtitle(paste("BMI and HIV"))


ggplot(data = outcome_exposures, aes(x = distance, fill=hiv)) +
    geom_histogram(binwidth=5) +
  ggtitle(paste("Distance and HIV")) 
summary((glm(hiv ~ distance, data = outcome_exposures, family = binomial())))
exp(coef(glm(hiv ~ distance, data = outcome_exposures, family = binomial()))) #Odds ratio

ggplot(data = outcome_exposures, aes(distance, fill=hiv)) + 
    geom_histogram(aes(y = ..density..), breaks = seq(0, 80, 1),  alpha = 0.7) +
    geom_density(color = "red") +
  ggtitle(paste("Distance and HIV"))



ggplot(data=outcome_exposures, aes(x=hiv, fill=dead)) + 
  geom_bar() +
  ggtitle(paste("HIV and Mortality")) + 
  theme (
    plot.title=element_text(hjust=0, face="bold", size=20))
```


####scatterplots
To see how two continuous variables are related to each other, age and hormone use, I used a scatterplot.

Raster of 2D density estimates
I also used density plotsto help me interpret data relationships.


```{r eval = TRUE}

#Scatterplot
ggplot(final_project_BCC, aes(x = age, y = hormone)) +
    geom_point()

# Raster of 2D density estimates
ggplot(final_project_BCC, aes(x = age, y = hormone)) +
    geom_point() +
    stat_density2d(aes(fill = ..level.., alpha = ..level..), geom = "polygon", color = "black") +
    scale_fill_continuous() +
    guides(alpha = "none")

#viridis` color scheme
ggplot(final_project_BCC, aes(x = age , y = hormone )) +
    stat_density2d(aes(fill = ..density..), geom = "raster", contour = FALSE) +
    scale_fill_viridis_c() 
``` 


###Results


####Chi-square
Here I will test univariate relationships with dichotomous and categorical variables with a Pearson's Chi-square test and display using bar graphs.

Chi-square tests and bar plots
```{r}

chisq.test(table(outcome_exposures$dead, outcome_exposures$hiv))
ggplot(data = outcome_exposures, aes(x = dead, fill= hiv, na.rm=TRUE)) +
    geom_bar(na.rm=TRUE) +
  ggtitle(paste("Mortality and HIV"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$age_group))
ggplot(data = outcome_exposures, aes(x = dead, fill = age_group, na.rm=TRUE)) +
    geom_bar() +
  ggtitle(paste("Mortality and Age group"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$er))
ggplot(data = outcome_exposures, aes(x = dead, fill = er, na.rm=TRUE)) +
    geom_bar() +
  ggtitle(paste("Mortality and Estrogen Status"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$pr))
ggplot(data = outcome_exposures, aes(x = dead, fill = pr, na.rm=TRUE)) +
    geom_bar() +
  ggtitle(paste("Mortality and Progesterone Status"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$her2))
ggplot(data = outcome_exposures, aes(x = dead, fill = her2, na.rm=TRUE)) +
    geom_bar() +
   ggtitle(paste("Mortality and Her2 Status"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$hormone))
ggplot(data = outcome_exposures, aes(x = dead, fill = hormone, na.rm=TRUE)) +
    geom_bar() +
   ggtitle(paste("Mortality and Hormone Receptor Status"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$marital_status))
ggplot(data = outcome_exposures, aes(x = dead, fill = marital_status, na.rm=TRUE)) +
    geom_bar() +
   ggtitle(paste("Mortality and Marital Status"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$smoker))
ggplot(data = outcome_exposures, aes(x = dead, fill = smoker, na.rm=TRUE)) +
    geom_bar() +
   ggtitle(paste("Mortality and Smoking History"))

chisq.test(table(outcome_exposures$dead, outcome_exposures$stage_high))
ggplot(data = outcome_exposures, aes(x = dead, fill = stage_high, na.rm=TRUE)) +
    geom_bar() +
   ggtitle(paste("Mortality and Stage of Tumor"))


```

#### Univariate Logistic Regression

To fit a model to a binary outcome, I will use logistic regression with the binomial family, due to the binary outcome, as the type of function to be used for the fit. I will retrieve the summary output and also report the odds rations and 95% conifence intervals.  

For the binary and categorial variables, results from Pearson's Chi-square test and logistic regression models should be the same.  I will plot the results of the continuous indepedent variables with histograms.


```{r eval = TRUE}

outcome_hiv<-glm(dead ~ hiv, data=outcome_exposures, family=binomial())
summary(outcome_hiv)
exp(cbind(OR = coef(outcome_hiv), CI = confint(outcome_hiv))) #OR and 95%CI

outcome_age<-glm(dead ~ age, data=outcome_exposures, family=binomial())
summary(outcome_age)
exp(cbind(OR = coef(outcome_age), CI = confint(outcome_age))) 

ggplot(data = outcome_exposures, aes(x = age, fill=dead)) +
    geom_histogram(binwidth=5) +
  ggtitle(paste("Age and Mortality")) 

ggplot(data = outcome_exposures, aes(dead, age)) + 
    geom_boxplot(outlier.colour = "red", outlier.size = 1) +
  geom_jitter(width=0.1, pch=21, fill="grey", na.rm=TRUE) +
  ggtitle(paste("Mortality Status by Age"))

outcome_age_group<-glm(dead ~ age_group, data=outcome_exposures, family=binomial())
summary(outcome_age_group)
exp(cbind(OR = coef(outcome_age_group), CI = confint(outcome_age_group))) 

outcome_stage<-glm(dead ~ stage_high, data=outcome_exposures, family=binomial())
summary(outcome_stage)
exp(cbind(OR = coef(outcome_stage), CI = confint(outcome_stage))) 

outcome_bmi<-glm(dead ~ bmi, data=outcome_exposures, family=binomial())
summary(outcome_bmi)
exp(cbind(OR = coef(outcome_bmi), CI = confint(outcome_bmi))) 

ggplot(data = outcome_exposures, aes(x = bmi, fill=dead)) +
    geom_histogram(binwidth=5) +
  ggtitle(paste("BMI and Mortality")) 


outcome_hormone<-glm(dead ~ hormone, data=outcome_exposures, family=binomial())
summary(outcome_hormone)
exp(cbind(OR = coef(outcome_hormone), CI = confint(outcome_hormone))) 

outcome_er<-glm(dead ~ er, data=outcome_exposures, family=binomial())
summary(outcome_er)
exp(cbind(OR = coef(outcome_er), CI = confint(outcome_er))) 

outcome_pr<-glm(dead ~ pr, data=outcome_exposures, family=binomial())
summary(outcome_pr)
exp(cbind(OR = coef(outcome_pr), CI = confint(outcome_pr))) 

outcome_her2<-glm(dead ~ her2, data=outcome_exposures, family=binomial())
summary(outcome_her2)
exp(cbind(OR = coef(outcome_her2), CI = confint(outcome_her2))) 

outcome_marital<-glm(dead ~ marital_status, data=outcome_exposures, family=binomial())
summary(outcome_marital)
exp(cbind(OR = coef(outcome_marital), CI = confint(outcome_marital))) 

outcome_distance<-glm(dead ~ distance, data=outcome_exposures, family=binomial())
summary(outcome_distance)
exp(cbind(OR = coef(outcome_distance), CI = confint(outcome_distance))) 

outcome_smoker<-glm(dead ~ smoker, data=outcome_exposures, family=binomial())
summary(outcome_smoker)
exp(cbind(OR = coef(outcome_smoker), CI = confint(outcome_smoker))) 

```

####Multivariate Logistic Regression
I will use Multivariate Logostoc Regression with the retained variables that had a p-value <0.20 in the univariate analysis.
```{r}

all_hiv<- outcome_exposures %>%
  select("hiv", "age", "stage_high", "pr", "dead")
outcome_all_hiv<-glm(dead ~ ., data=all_hiv, family=binomial())
summary(outcome_all_hiv)
exp(cbind(OR = coef(outcome_all_hiv), CI = confint(outcome_all_hiv)))



all<- outcome_exposures %>%
  select("age", "stage_high", "pr", "dead")
outcome_all<-glm(dead ~ ., data=all, family=binomial())
summary(outcome_all)
exp(cbind(OR = coef(outcome_all), CI = confint(outcome_all))) 



outcome_all<-glm(dead ~ ., data=all, family=binomial())
summary(outcome_all)
exp(cbind(OR = coef(outcome_all), CI = confint(outcome_all)))

outcome<-glm(dead ~ hiv+ age + pr, data=outcome_exposures, family=binomial())
summary(outcome)


```

###Botswana Shape File
####Spatial Data and Mapping
Mapping a Botswana Shape file- but no data at this time to converge.  This turned out to be more difficult than I anticipated because there are not a lot of avaiabe files for geogrpahy and Botswana.  Also, there are no FIPs codes or census blocks to merge data sets.  This is something I would like to explore further as I appreciated the geographic capabilties of R with spatial data.
```{r}

library(sp)
library(rgdal)
library(sf)
dir.create("Bostwana")
setwd("Bostwana")
shape <- readOGR(dsn = ".", layer = "BWA_adm2")
plot(shape)
```


###Conclusion
This is the first look at demographic and biological characteristics associated with breast cancer mortality in Botswana. The pathology data for breast cancer is extremely limited in this database.  Abstraction from medical records is needed to complete missing data.  The variables that had a statistically significant higher odds of dying from Breast cancer were: progesterone neagtive tumors, late stage (stage III/IV) of tumor at diagnosis, and older age.  For the multivariate analysis, none of these factors remained signficanct.

For exploratory purposes I also looked at variables of interest with HIV affacted and HIV unaffected women.  Botswana has the third highest prevalence of HIv in the world.  Many oncology studies in Botswana account for this co-mordity.  For this sub-analysis, younger age was associated with a greater odds of having HIV (p<0.000).

Also, future directions for this area of research includes learning information that can be applicable to other cancers, informing policy about the need for breast cancer screening programs, and lessons learned from this can be applied to other LMICs.



