---
title: "Heart Rate Monitors"
author: "Avantika"
date: "October 20, 2016"
output: html_document
---

Wrist Worn Heart Rate Monitors
by Avantika Diwadkar 
Masters in Biotechnology, University of Pennsylvania
EPID 600 Final Project

#Overview
The main focus of this project is to investigate the use of commercial wearables as possible patient monitoring devices by testing the precision of heart rate measurement by a wrist worn heart rate monitor (FitBit) against a portable ECG-based heart rate computer (Bitalino). The project will involve data collection, extraction, cleaning and conversion followed by a statistical model to test the accuracy of the FitBit band and correlation statistics with ECG.

#Introduction 
Wearable devices are being studied and explored for use as diagnostic and patient monitoring devices replacing the current complicated and expensive medical devices. Now-a-days, wrist monitors like FitBit, Apple and Samsung are popular in the market. These devices use photoplethysmography (PPG) which detects blood racing through the veins by optical sensors. Majority of the limited literature review proves that the bands are mostly accurate in heart rate measurement with few abnormalities at high bpm. 
Heart rate variation is the physiological phenomenon of variation in the time interval between heartbeats. It is measured by the variation in the beat-to-beat interval. The SA node receives several different inputs from the Sympathetic and Parasympathetic Nervous system. Factors that affect the input are the baroreflex, thermoregulation, hormones, sleep-wake cycle, meals, physical activity, and stress. HRV can be used for stress response detection due to behavioral, bacterial or other abnormalities. The HRV can be also used to determine outcome of psychological conditions like Autism and other behavioral conditions. It can also be used to detect the outcome of a bacterial or behavioral abnormality before the patient actually shows indications of it. Thus, wearables have application in diagnostics and patient monitoring.

The problem addressed in this project covers the field of diagnostics, biomedical devices, signal processing and data analysis. Through this project, I am trying to address two questions - 
1)Does the heart rate measurement accuracy of FitBit vary with range? 
2)Is the heart rate measurement by Fitbit accurate?

The project layout consists of device set up and measurement of heart rate with 2 samples. The first sample is heart rate measured at different physical conditions (4 minutes walking - 3 minutes standing - 4 minutes sitting) with both the devices -Fitbit and Bitalino taking measurement at the same time. The second sample is heart rate measured over the course of an hour for use in correlation testing. The data extracted was filtered and converted to appropriate units for comparison. The first and second samples are being used to answer the first and second question respectively. Statistical tests like t-test, Pearson correlation are calculated and histograms and boxplots are used for visualization.

My supervisor Mr. Jeff Pennington from Children's Hospital of Philadelphia is guiding me in this project. The devices have been provided by him. CHOP is interested in exploring the use of wearable devices and this project is a preliminary device study that may aid them in determining the accuracy of these devices for their future study. More than accuracy, they are mainly interested in feature extraction thus, the correlation statistics is of great interest to them.

#Methods
The data analysis in this project involves the following steps - 

##Device Setup
The devices that I am using are Bitalino for the ECG measurement and the Fitbit watch as a wearable device. For the ECG, I used hand electrodes for first sample and chest electrodes for the second sample. The Bitalino is connected to the laptop by Bluetooth and the software OpenSignals is used to take measurements.

##Data Acquisition
I measured the heart rate in different situations with varying time periods. The total data was acquired from two samples - 
1) Heart rate measured at different physical conditions (4 minutes walking - 3 minutes standing - 4 minutes sitting)
2)Heart rate measured over the course of 60 minutes. 

##Data Extraction
This is the important step after acquisition of data. I am using R package fitbitScraper to extract data from the Fitbit cloud. The data extracted consists of multiple variables like confidence level, date, time, heart rate zones, calories burned, etc. but the one I am interested in is the bpm (beats per minute data). The FitBit band measures data every one minute but allows us to extract data at the interval of 5 minutes only. So the data I got was the beats per minute value for that specific minute. Fitbit is currently working on a plug-in to give access to the entire heart rate data. This has proved to be a major setback of any wearable device. 
For Bitalino, the Opensignals software produces the data in two formats - h5 and txt. I am showing the method to extract data from both types of files but I will be using the data from txt file for my further analysis. nSeq is the sequence, while I1 -I4 are the inputs. The interesting and useful variable is the raw one.


##Data Cleaning/Conversion
The raw output data received from the Bitalino device needs to be converted into the standard units of measurement, which in this case is Volts. I converted the raw values obtained from bitalino into appropriate values by the transfer function mentioned below.

Transfer function:
ECGV = (ECGB * Vcc / 2^n - Vcc / 2) / GECG                                        
Where: 
ECGV - ECG value in Volts (V)
ECGmV - ECG value in miliVolts (mV)
ECGB - ECG value obtained from BITalino
Vcc - Operating Voltage (V) 
n - number of bits (bit) 
GECG - ECG Sensor Gain 

Values: 
Vcc = 3.3 (V)
GECG = 1100 
n= 10 (for Channel 3)

The next step involves converting the ECG output to RR intervals. Kubios is a software that is used to calculate heart rate variation. I used the ecgdata.txt file as input in the software to convert the ECG data into RR intervals. The RR interval data has to be converted into beats per minute. The formula for conversion is RR interval for 1 minute = 1/bpm. As the RR is in seconds I am converting it to minutes and then inversing it to get the beats per minute value.
60/average of RR interval in 1 minute = beats per minute 

##Statistical Analysis
The statistical model involves quantification of the second sample by Bland-Altman plots, t-test calculation, accuracy percentage and correlation coefficient. In the first sample, the variation in heart rate over the different episodes is visualized by plots and histograms to get an idea of variation in accuracy of measurement. The second sample is being analyzed statistically by correlation coefficient, t-tests and accuracy percentage. The data points are being quantified by Bland Altman plot analysis.

#Results
The heart rate measurement by Fitbit shows variation for high beats per minute measurement. The accuracy of measurement by the wearable device shows low accuracy with a weak correlation with the positive control

##Sample 1
The final bpm measurement for Sample 1 have been shown in dataset "total". By viewing the histogram for total data, you can see rise in the heart rate during walking. The high bpm measurement during walking and standing has been incorrectly calculated by the Fitbit band while the heart rate measurement while sitting is almost accurate.  This indicates that accuracy in heart beat measurement by Fitbit is influenced by high range of bpm measurement.

##Sample 2
The boxplot figure for dataset "total2" shows considerable variation in the data collected from both the devices with mean values 71.50 and 75.70 for Fitbit and Bitalino respectively. The t-test gives a p-value of 0.002331. Thus, as the p-value is significant, we reject the null hypothesis. The correlation coefficient gives us a value of 0.3725712 which is pretty weak. The correlation plot gives us no information about the two device measurements. The Bland Altman plot shows a proportional error trend with more variability for mid-range heart rate values and less for tail ends. All of the data points except one falls within the limits of agreement. 

#Future Goals
The negative results may be due to error in ECG measurement using Bitalino with the possibility of the device not being as accurate as claimed. The RR interval to beats per minute calculation could have affected the results too, as the conversion method is a crude one and may produce an error. There could also be human error involved with time. Keeping in mind these possibilities, my next step is to change the positive control to a more accurate device namely, Biopac. The advantage with Biopac is it gives an output in beats per minute so it will help minimize the process by eliminating the entire data conversion and cleaning step. 
 I am also working on getting more data from Fitbit and re-checking if there is a method to get data within a shorted time interval. The increase in data extracted would also help me to calculate RR intervals i.e. HRV which is the main feature we are interested in. We will also be experimenting with other wearables like Apple watch and Polar as they are known to be more accurate than the Fitbit watch

#References
1)Wallen MP, Gomersall SR, Keating SE, Wisløff U, Coombes JS (2016) Accuracy of Heart Rate Watches: Implications for Weight Management. PLoS ONE 11(5): e0154420
2)El-Amrawy F, Nounou MI.   Are Currently Available Wearable Devices for Activity Tracking and Heart Rate Monitoring Accurate, Precise, and Medically Beneficial? Healthc Inform Res. 2015 Oct;21(4):315-320.
3)http://biostatmatt.com/R/ecg.R


##Code below
```{r setup, eval=FALSE,include=FALSE}
install.packages("fitbitScraper")
install.packages("lubridate")
install.packages("BlandAltmanLeh")
install.packages("ggplot2")
install.packages("dplyr")
install.packages("plyr")
```

##Libaries used
```{r message=FALSE, warning=FALSE}
library(fitbitScraper)
library(plyr)
library(dplyr)
library(lubridate)
library(BlandAltmanLeh)
library(ggplot2)
```

#Data extraction
##FitBit
##Sample1
I constructed the login cookie for accessing fibit data from cloud and extracted heart rate data (bpm within 5 minute intervals)
```{r}
cookie = login("penningtonjeff@yahoo.com","wearable10", rememberMe = TRUE)
startdate = as.Date("2016-11-23", format="%Y-%m-%d")
my.hrdata <- get_intraday_data(cookie,what="heart-rate",date=sprintf("%s",startdate))

#Printing out bpm for the measured time period. Rest all are zeros.
newdata <- my.hrdata[c(144:152),]
newdata

#Plotting the data
plot(newdata$time,newdata$bpm, type="l",xlab="time",ylab="bpm",main="Heart rate (bpm) after 5 min intervals for 9th October 2016")

```

##Sample2
```{r}
cookie = login("anupamalur@gmail.com","anupam123", rememberMe = TRUE)
startdate = as.Date("2016-12-02", format="%Y-%m-%d")

my.hrdata2 <- get_intraday_data(cookie,what="heart-rate",date=sprintf("%s",startdate))
fitdata <- my.hrdata2[219:230,]
```

##Bitalino
The Opensignals software produces the data in two formats - h5 and txt. I am showing the method to extract the data from txt file for my further analysis. nSeq is the sequence, while I1 -I4 are the inputs. The interesting and useful variable is the raw one.

##Sample 1
```{r}
#txt file 
my.data <- read.table("sample1.txt")
names(my.data) <- c("nSeq","I1","I2","I3","I4","raw")
head(my.data)
str(my.data)

```

#Sample 2
```{r}
bitdata <- read.table("sample2.txt")
names(bitdata) <- c("nSeq","I1","I2","I3","I4","raw")
head(bitdata)
str(bitdata)

```
 
#Data conversion/cleaning
The data received from the Bitalino device needs to be converted into the standard units of measurement, which in this case is Volts.I am converting the raw values obtained from bitalino into appropriate values by the formula: ECGV = (ECGB * Vcc / 2^n - Vcc / 2) / GECG 

Where: 
ECGV - ECG value in Volts (V)
ECGmV - ECG value in miliVolts (mV)
ECGB - ECG value obtained from BITalino
Vcc - Operating Voltage (V) 
n - number of bits (bit) 
GECG - ECG Sensor Gain 

Values: 
Vcc = 3.3 (V)
GECG = 1100 
n= 10 (for Channel 3)
##Sample 1
```{r}
#Converting to ECG data in V
my.data$ECG <- sapply(my.data$raw,function(x)(x*3.3/2^10 - 3.3/2)/1100)
head(my.data$ECG)

#Write in file for Kubios software
ecgdatalist1 <- c(my.data$ECG)
write.table(ecgdatalist1,file="ecgdata1.txt")
#I manually removed the first column from this file
```

##Sample 2
```{r}
#Converting to ECG data in V
bitdata$ECG <- sapply(bitdata$raw,function(x)(x*3.3/2^10 - 3.3/2)/1100)
head(bitdata$ECG)

#Write in file for Kubios software
ecgdatalist2 <- c(bitdata$ECG)
write.table(ecgdatalist2,file="ecgdata2.txt")
#I manually removed the first column from this file
```

##Convert Bitalino data to RR interval using Kubios- opening the converted files
#Sample1
```{r}
hrv1.kubios <- read.csv("ecghrv1.csv")
names(hrv1.kubios) <- c("Time","RR Interval","FFT Frequency","PSD","AR Frequency","PSD")
head(hrv1.kubios)

```

#Sample 2
```{r}
hrv2.kubios <- read.csv("ecghrv2.csv")
names(hrv2.kubios) <- c("Time","RR Interval","FFT Frequency","PSD","AR Frequency","PSD")
head(hrv2.kubios)

```

##Data cleaning and conversion
The RR interval data has to be converted into beats per minute. The formula for conversion is RR interval for 1 minute = 1/bpm. As the RR is in seconds I am converting it to minutes and then inversing it to get the beats per minute value. 


#Sample 1
```{r}
#Episode one - sitting - calculate RR for 12:25
B1 <- 60/(mean(hrv1.kubios[1:85,2])) #Bitalino
F1 <- newdata$bpm[6] #time=12:25

#Episode two - walking - calculate RR for 12:30
B2 <- 60/(mean(hrv1.kubios[337:422,2])) 
F2 <- newdata$bpm[7] #time=12:30

#Episode three - sitting again - calculate RR for 12:35
B3 <- 60/(mean(hrv1.kubios[653:726,2]))
F3 <- newdata$bpm[8] #time=12:35


bitalino <- c(B1, B2, B3)
fitbit <- c(F1, F2, F3)
```

#Sample 2
```{r}
#Calculating for every 5 minutes - thus, calculating 12 data points
a <- 60/mean((hrv2.kubios[314:394,2])) #18:15 - 240 to 300 seconds
b <- 60/mean((hrv2.kubios[716:795,2]))  #18:20 - 540 to 600 seconds
c <- 60/mean((hrv2.kubios[1091:1164,2]))  #18:25 - 840 to 900 seconds
d <- 60/mean((hrv2.kubios[1443:1521,2]))  #18:30 - 1140 to 1200 seconds
e <- 60/mean((hrv2.kubios[1813:1893,2]))  #18:35 - 1440 to 1500 seconds
f <- 60/mean((hrv2.kubios[2181:2256,2]))  #18:40 - 1740 to 1800 seconds
g <- 60/mean((hrv2.kubios[2548:2620,2]))  #18:45 - 2040 to 2100 seconds
h <- 60/mean((hrv2.kubios[2917:2989,2]))  #18:50 - 2340 to 2400 seconds
i <- 60/mean((hrv2.kubios[3280:3354,2]))  #18:55 - 2640 to 2700 seconds
j <- 60/mean((hrv2.kubios[3654:3727,2]))  #19:00 - 2940 to 3000 seconds
k <- 60/mean((hrv2.kubios[4022:4097,2]))  #19:05 - 3240 to 3300 seconds
l <- 60/mean((hrv2.kubios[4401:4476,2]))  #19:10 - 3540 to 3600 seconds

bitalino2 <- c(a,b,c,d,e,f,g,h,i,j,k,l)
fitbit2 <- c(fitdata$bpm)

```


#visualization
The data can be visualized to check overall trend - ecg RR interval data 
```{r}
plot(hrv1.kubios$`RR Interval`,type="l")
plot(hrv2.kubios$`RR Interval`,type="l")
```

##Signal processing
I have used a code [3] to process the ECG data obtained from Bitalino to pass low and high pass filters so as to visualize the data. The sampling rate used was 100Hz. The purpose of this is to only visualization of the data. Sample 2 has been visualized here. 
```{r}
#Transform the real and imaginary portions of the 
#FFT into magnitude and phase. 
amplitude <- function( x ) { sqrt(Re(x)^2+Im(x)^2) }
phase     <- function( x ) { atan(Im(x)/Re(x)) }

#sinc function of frequency f
sinc      <- function( x, f ) { ifelse(x==0, 2*pi*f, sin(2*pi*f*x)/x) }

#Blackman window from 0..m
Blackman  <- function( m ) { 0.42-0.5*cos(2*pi*(0:m)/m)+0.08*cos(4*pi*(0:m)/m) }

#Hamming window from 0..m
Hamming   <- function( m ) { 0.54-0.46*cos(2*pi*(0:m)/m) }

#simple low pass filter
#y - vector to filter
#t - time interval between measurements (s)
#f - low pass frequency (Hz)
lpf <- function( y, t, f ) {
  rc <- 1 / ( 2 * pi * f )
  a  <- t / ( t + rc )
  n  <- length( y )
  yf <- y
  for( i in 2:length(y) ) {
    yf[i] <- a * y[i] + (1-a) * yf[i-1]
  }
  return( yf )
}  

#windowed sinc low pass filter
#y - vector to filter
#t - time interval between measurements (s)
#f - low pass frequency (Hz)
wlpf <- function( y, t, f ) {
  m  <- min(floor(length(y)/2), 500)
  #generate the sinc kernel
  rk <- sinc(-m:m, f*t)  
  #apply the Blackman window
  bk <- Blackman(2*m) * rk
  #pad the filter with zeros
  k  <- c(bk, rep(0,length(y)-length(bk)))
  #convolve y with the filter kernel
  fy  <- fft(fft(k)*fft(y), inverse=TRUE)
  return(Re(fy))
}

dat  <- scan(file="ecgdata2.csv")
dat  <- (dat - mean(dat)) / sd(dat)

#filter high frequency noise 
fdat <- wlpf(dat, 1/100, 30)
fdat <- (fdat-mean(fdat))/sd(fdat)

#filter low frequency noise
rdat <- wlpf(dat, 1/100, 1)
rdat <- (rdat-mean(rdat))/sd(rdat)

#use low pass filter in ecg signal
edat <- fdat - rdat

require(lattice)
xplot <- rep((0:(length(dat)-1))/1000,4)
yplot <- c(dat,fdat,rdat,edat)
gplot <- c(rep("Raw",length(dat)),
           rep("Low Frequency Filter",length(dat)),
           rep("High Frequency Filter",length(dat)),
           rep("ECG",length(dat)))

tp <- xyplot(yplot~xplot|gplot,type="l",layout=c(1,4), xlab="Time", ylab="V")

#uncomment the following to save an image
#trellis.device(png, file="ecgfilter.png", height=750, width=750)
print(tp)
#dev.off()

```
#Results
The results of the data can be first visualized by histograms:

##Histograms
##Sample 1
```{r}
total <- data.frame(c(fitbit,bitalino))
names(total)<- c("bpm")
total$episode <- c("walking","standing","sitting")
total$device <- c("fitbit","fitbit","fitbit","bitalino","bitalino","bitalino")
total

ggplot(total, aes(factor(episode),bpm, fill = device)) + 
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_brewer(palette = "Set1")
```

##Sample 2
```{r}
total2 <- data.frame(c(fitbit2,bitalino2))
names(total2)<- c("bpm")
total2$time <- c("05","10","15","20","25","30","35","40","45","50","55","60")
total2$device <- c("fitbit","fitbit","fitbit","fitbit","fitbit","fitbit","fitbit","fitbit","fitbit","fitbit","fitbit","fitbit","bitalino","bitalino","bitalino","bitalino","bitalino","bitalino","bitalino","bitalino","bitalino","bitalino","bitalino","bitalino")
total2

ggplot(total2, aes(factor(time),bpm, fill = device)) + 
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_brewer(palette = "Set2")

#boxplots
boxplot(bpm~device, data=total2,col="blue", main="Heart rate measurement in bpm")

#Histograms
ggplot(total2[1:12,], aes(factor(time),bpm)) + 
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_brewer(palette = "Set1")

ggplot(total2[12:24,], aes(factor(time),bpm)) + 
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_brewer(palette = "Set2")
```


##BlandAltman Plots
The converted results can be analysed by plotting BlandAltman plots to check the accuracy of the measurements.The majority of the points are within the limits of agreement (the 2 dotted lines at the end representing +1.96 SD and -1.96 SD from top to bottom. The middle dotted line is the mean difference between the two methods.)

```{r}

bland.altman.plot(fitbit2, bitalino2, xlab = "Mean measurement", ylab= "Differences")
bland.altman.stats(fitbit2, bitalino2)

```

##T-test
The data can be analysed by performing t-test. The p-value is 0.002331 thus, significant. 
```{r}
t.test(fitbit2,mu=75.70883)

```

##correlation 
For sample 2 gives a R value of 0.3725712
```{r eval=TRUE}
cor(fitbit2, bitalino2, use="complete.obs", method="pearson")
plot(fitbit2 ~ bitalino2)
```





