---
title: "Changing Nature of Death in Infancy in the United States"
author: "Kevin Dysart"
date: "12/1/2019"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading packages, echo=FALSE, warning=FALSE, message=FALSE}
library("tidyverse")
library("urbnmapr")
library("lubridate")
library("survival")
library("survminer")
library("RColorBrewer")
library("knitr")
library("plotrix")
library("ggfortify")
```

```{r reading in dataframes for slides, echo=FALSE, warning=FALSE, message=FALSE}
#Reading CDC Wonder Data Files
Wonder_File_Names <- list.files(path = "Linked_Birth_Death_Data/CDC_Wonder_All_Death_By_Year/", pattern = "*.txt")
Wonder_Files <- list.files(path = "Linked_Birth_Death_Data/CDC_Wonder_All_Death_By_Year/", pattern = "*.txt", full.names = T)
Wonder_Files_df <- lapply(Wonder_Files, read.delim)
names(Wonder_Files_df) <- Wonder_File_Names
#Reading Linked Birth and Death Data
Linked_Birth_Death_Names <- list.files(path = "Linked_Birth_Death_Data/Linked_Year_CSV_Zip/", pattern = "*.csv.zip")
Linked_Birth_Death_Files <- list.files(path = "Linked_Birth_Death_Data/Linked_Year_CSV_Zip/", pattern = "*.csv.zip", full.names = T)
Linked_Birth_Death_Files_df <- lapply(Linked_Birth_Death_Files, read_csv)
names(Linked_Birth_Death_Files_df) <- Linked_Birth_Death_Names
```

```{r removing Notes column, echo=FALSE, warning=FALSE, message=FALSE}
for (i in 1:length(Wonder_Files_df)) {
  Wonder_Files_df[[i]] <-  Wonder_Files_df[[i]] %>%
                    select(-Notes) %>%
                    slice(1:51)
}
```

```{r Reading in IMR files for VLBW and ELBW, echo=FALSE, warning=FALSE, message=FALSE}
## Reading in IMR files from CDC 1995 through 2017 ELBW VLBW.
IMR_VLBW_ELBW_Names <- list.files(path = "Linked_Birth_Death_Data/IMR_VLBW_ELBW/", pattern = "*.txt")
IMR_VLBW_ELBW_Files <- list.files(path = "Linked_Birth_Death_Data/IMR_VLBW_ELBW/", pattern = "*.txt", full.names = T)
IMR_VLBW_ELBW_df <- lapply(IMR_VLBW_ELBW_Files, read.delim)
names(IMR_VLBW_ELBW_df) <- IMR_VLBW_ELBW_Names
## Cleaning each dataframe in the list. ELBW and VLBW Dataframes are paired down the list.
ELBW_1995 <- IMR_VLBW_ELBW_df[[1]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_1995 <- IMR_VLBW_ELBW_df[[2]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_1999 <- IMR_VLBW_ELBW_df[[3]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_1999 <- IMR_VLBW_ELBW_df[[4]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_2003 <- IMR_VLBW_ELBW_df[[5]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_2003 <- IMR_VLBW_ELBW_df[[6]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_2007 <- IMR_VLBW_ELBW_df[[7]] %>%
                select(-Notes) %>%
                    slice(1:11)
VLBW_2007 <- IMR_VLBW_ELBW_df[[8]] %>%
                select(-Notes) %>%
                    slice(1:11)
All_ELBW_IMR <- do.call("rbind", list(ELBW_1995, ELBW_1999, ELBW_2003, ELBW_2007))
All_VLBW_IMR <- do.call("rbind", list(VLBW_1995, VLBW_1999, VLBW_2003, VLBW_2007))
##Creating a grouping variable for data cohorts
All_VLBW_IMR$Year.Group <- c(rep("1995-1998", 4), rep("1999-2002", 4),
                             rep("2003-2006", 4), rep("2007-2017", 11))
All_ELBW_IMR$Year.Group <- c(rep("1995-1998", 4), rep("1999-2002", 4),
                             rep("2003-2006", 4), rep("2007-2017", 11))
All_ELBW_IMR$Year.Group <- as.factor(All_ELBW_IMR$Year.Group)
All_VLBW_IMR$Year.Group <- as.factor(All_VLBW_IMR$Year.Group)
```

```{r Cleaning Data for IMR Description, echo=FALSE, warning=FALSE, message=FALSE}
## Demographic data cleaning
linked_1995 <- Linked_Birth_Death_Files_df[[1]]
linked_1995_VLBW <- filter(linked_1995, dbirwt<1500)
Percent_1995_IMR_VLBW <- length(linked_1995_VLBW$matchs)/length(linked_1995$matchs)
linked_2017 <- Linked_Birth_Death_Files_df[[19]]
linked_2017_VLBW <- filter(linked_2017, dbwt<1500)
Percent_2017_IMR_VLBW <- length(linked_2017_VLBW$idnumber)/length(linked_2017$idnumber)
linked_1995_ELBW <- filter(linked_1995, dbirwt<1000)
linked_2017_ELBW <- filter(linked_2017, dbwt<1000)
Percent_1995_IMR_ELBW <- length(linked_1995_ELBW$matchs)/length(linked_1995$matchs)
Percent_2017_IMR_ELBW <- length(linked_2017_ELBW$idnumber)/length(linked_2017$idnumber)
linked_1995_lt_23 <- filter(linked_1995, gestat<23)
linked_2017_lt_23 <- filter(linked_2017, combgest<23)
Percent_1995_IMR_lt_23 <- length(linked_1995_lt_23$matchs)/length(linked_1995$matchs)
Percent_2017_IMR_lt_23 <- length(linked_2017_lt_23$idnumber)/length(linked_2017$idnumber)
```


```{r Creating Data Frames for boxplot of age of death, echo=FALSE, warning=FALSE, message=FALSE}

##This chunk creates a large data frames for the year of birth for each infant in the linked death and birth certificate data. Because infants less than 22 weeks have not routinely been offered care in the United States unitl recent years. VLBW and ELBW data frames. Unfortunately each data frame needs to be dealt with from the list of data frames separately. Over the approximately 20 year time period variables have been added to the birth or death certificate 

##ELBW Infants
ELBW_1995_linked_ad <- Linked_Birth_Death_Files_df[[1]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1996_linked_ad <- Linked_Birth_Death_Files_df[[2]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1997_linked_ad <- Linked_Birth_Death_Files_df[[3]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1998_linked_ad <- Linked_Birth_Death_Files_df[[4]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1999_linked_ad <- Linked_Birth_Death_Files_df[[5]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2000_linked_ad <- Linked_Birth_Death_Files_df[[6]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2001_linked_ad <- Linked_Birth_Death_Files_df[[7]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2002_linked_ad <- Linked_Birth_Death_Files_df[[8]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2003_linked_ad <- Linked_Birth_Death_Files_df[[9]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2004_linked_ad <- Linked_Birth_Death_Files_df[[10]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2005_linked_ad <- Linked_Birth_Death_Files_df[[11]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2006_linked_ad <- Linked_Birth_Death_Files_df[[12]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2007_linked_ad <- Linked_Birth_Death_Files_df[[13]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2008_linked_ad <- Linked_Birth_Death_Files_df[[14]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2009_linked_ad <- Linked_Birth_Death_Files_df[[15]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2010_linked_ad <- Linked_Birth_Death_Files_df[[16]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2011_linked_ad <- Linked_Birth_Death_Files_df[[17]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2012_linked_ad <- Linked_Birth_Death_Files_df[[18]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
ELBW_2013_linked_ad <- Linked_Birth_Death_Files_df[[19]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 

ELBW_rbind_ad <- do.call("rbind", list(ELBW_1995_linked_ad,
                                       ELBW_1996_linked_ad,
                                       ELBW_1997_linked_ad,
                                       ELBW_1998_linked_ad,
                                       ELBW_1999_linked_ad,
                                       ELBW_2000_linked_ad,
                                       ELBW_2001_linked_ad,
                                       ELBW_2002_linked_ad,
                                       ELBW_2003_linked_ad,
                                       ELBW_2004_linked_ad,
                                       ELBW_2005_linked_ad,
                                       ELBW_2006_linked_ad,
                                       ELBW_2007_linked_ad,
                                       ELBW_2008_linked_ad,
                                       ELBW_2009_linked_ad,
                                       ELBW_2010_linked_ad,
                                       ELBW_2011_linked_ad,
                                       ELBW_2012_linked_ad,
                                       ELBW_2013_linked_ad))
ELBW_rbind_ad$Year <- as.factor(ELBW_rbind_ad$Year)
ELBW_rbind_ad_group <- group_by(ELBW_rbind_ad, Year)
##User function to compute standard error.
se <- function(x) {
  sqrt(var(x)/length(x))
}
##Summary table for ELBW Infants
ELBW_rbind_ad_summ <- summarise(ELBW_rbind_ad_group, mean_aged = mean(aged),
                                median_aged = median(aged), 
                                sd_aged = sd(aged),
                                se_aged = se(aged),
                                se_2_aged = std.error(aged),
                                n=n())
##VLBW Infants
VLBW_1995_linked_ad <- Linked_Birth_Death_Files_df[[1]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_1996_linked_ad <- Linked_Birth_Death_Files_df[[2]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_1997_linked_ad <- Linked_Birth_Death_Files_df[[3]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_1998_linked_ad <- Linked_Birth_Death_Files_df[[4]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_1999_linked_ad <- Linked_Birth_Death_Files_df[[5]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_2000_linked_ad <- Linked_Birth_Death_Files_df[[6]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_2001_linked_ad <- Linked_Birth_Death_Files_df[[7]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_2002_linked_ad <- Linked_Birth_Death_Files_df[[8]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1500) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
VLBW_2003_linked_ad <- Linked_Birth_Death_Files_df[[9]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2004_linked_ad <- Linked_Birth_Death_Files_df[[10]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2005_linked_ad <- Linked_Birth_Death_Files_df[[11]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2006_linked_ad <- Linked_Birth_Death_Files_df[[12]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2007_linked_ad <- Linked_Birth_Death_Files_df[[13]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2008_linked_ad <- Linked_Birth_Death_Files_df[[14]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2009_linked_ad <- Linked_Birth_Death_Files_df[[15]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2010_linked_ad <- Linked_Birth_Death_Files_df[[16]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2011_linked_ad <- Linked_Birth_Death_Files_df[[17]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2012_linked_ad <- Linked_Birth_Death_Files_df[[18]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 
VLBW_2013_linked_ad <- Linked_Birth_Death_Files_df[[19]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1500) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 

VLBW_rbind_ad <- do.call("rbind", list(VLBW_1995_linked_ad,
                                       VLBW_1996_linked_ad,
                                       VLBW_1997_linked_ad,
                                       VLBW_1998_linked_ad,
                                       VLBW_1999_linked_ad,
                                       VLBW_2000_linked_ad,
                                       VLBW_2001_linked_ad,
                                       VLBW_2002_linked_ad,
                                       VLBW_2003_linked_ad,
                                       VLBW_2004_linked_ad,
                                       VLBW_2005_linked_ad,
                                       VLBW_2006_linked_ad,
                                       VLBW_2007_linked_ad,
                                       VLBW_2008_linked_ad,
                                       VLBW_2009_linked_ad,
                                       VLBW_2010_linked_ad,
                                       VLBW_2011_linked_ad,
                                       VLBW_2012_linked_ad,
                                       VLBW_2013_linked_ad))
VLBW_rbind_ad$Year <- as.factor(VLBW_rbind_ad$Year)
VLBW_rbind_ad_group <- group_by(VLBW_rbind_ad, Year)
VLBW_rbind_ad_summ <- summarise(VLBW_rbind_ad_group, mean_aged = mean(aged),
                                median_aged = median(aged), 
                                sd_aged = sd(aged),
                                se_aged = se(aged),
                                n=n())
## Filter the dataframes based on surviving more than 7 days.
ELBW_surv_8d <- filter(ELBW_rbind_ad_group, aged>7)
VLBW_surv_8d <- filter(VLBW_rbind_ad_group, aged>7)
## Create Summary tables for ELBW and VLBW that survive more than 7 days.
ELBW_surv_8d_group <- group_by(ELBW_surv_8d, Year)
VLBW_surv_8d_group <- group_by(VLBW_surv_8d, Year)
ELBW_surv_8d_summ <- summarise(ELBW_surv_8d_group, mean_aged = mean(aged),
                                median_aged = median(aged), 
                                sd_aged = sd(aged),
                                se_aged = se(aged),
                                n=n())
VLBW_surv_8d_summ <- summarise(VLBW_surv_8d_group, mean_aged = mean(aged),
                                median_aged = median(aged), 
                                sd_aged = sd(aged),
                                se_aged = se(aged),
                                n=n())
```

```{r Creating Data frames for Kaplan Meier Analysis, echo=FALSE, warning=FALSE, message=FALSE}
km_infant_deaths_1995_VLBW <- Linked_Birth_Death_Files_df[[1]] %>%
                      dplyr::filter(dbirwt < 1500) %>%
                      dplyr::filter(gestat>22) %>%
                      dplyr::select(aged, biryr, csex, mrace3) %>%
                      dplyr::rename(dob_yy = biryr, sex = csex, mracerec = mrace3) %>%
                      dplyr::mutate(race = ifelse((mracerec==1), "white", 
                                          ifelse((mracerec==3), "black", "other"))) %>%
                      dplyr::mutate(sex=ifelse(sex==1,"M","F")) %>%
                      dplyr::select(-mracerec) %>%
                      dplyr::mutate(status=1)
km_infant_deaths_1995_ELBW <- Linked_Birth_Death_Files_df[[1]] %>%
                      dplyr::filter(dbirwt < 1000) %>%
                      dplyr::filter(gestat>22) %>%
                      dplyr::select(aged, biryr, csex, mrace3) %>%
                      dplyr::rename(dob_yy = biryr, sex = csex, mracerec = mrace3) %>%
                      dplyr::mutate(race = ifelse((mracerec==1), "white", 
                                          ifelse((mracerec==3), "black", "other"))) %>%
                      dplyr::mutate(sex=ifelse(sex==1,"M","F")) %>%
                      dplyr::select(-mracerec) %>%
                      dplyr::mutate(status=1)
km_infant_deaths_2007_VLBW <- Linked_Birth_Death_Files_df[[13]] %>%
                      dplyr::filter(dbwt < 1500) %>%
                      dplyr::filter(combgest > 22) %>%
                      dplyr::select(aged, dob_yy, sex, mracerec) %>%
                      dplyr::mutate(race = ifelse((mracerec==1), "white", 
                                          ifelse((mracerec==2), "black", "other"))) %>%
                      dplyr::select(-mracerec) %>%
                      dplyr::mutate(status=1)
km_infant_deaths_2007_ELBW <- Linked_Birth_Death_Files_df[[13]] %>%
                      dplyr::filter(dbwt < 1000) %>%
                      dplyr::filter(combgest > 22) %>%
                      dplyr::select(aged, dob_yy, sex, mracerec) %>%
                      dplyr::mutate(race = ifelse((mracerec==1), "white", 
                                          ifelse((mracerec==2), "black", "other"))) %>%
                      dplyr::select(-mracerec) %>%
                      dplyr::mutate(status=1)
km_infant_deaths_2013_VLBW <- Linked_Birth_Death_Files_df[[19]] %>%
                      dplyr::filter(dbwt < 1500) %>%
                      dplyr::filter(combgest > 22) %>%
                      dplyr::select(aged, dob_yy, sex, mracerec) %>%
                      dplyr::mutate(race = ifelse((mracerec==1), "white", 
                                          ifelse((mracerec==2), "black", "other"))) %>%
                      dplyr::select(-mracerec) %>%
                      dplyr::mutate(status=1)
km_infant_deaths_2013_ELBW <- Linked_Birth_Death_Files_df[[19]] %>%
                      dplyr::filter(dbwt < 1000) %>%
                      dplyr::filter(combgest > 22) %>%
                      dplyr::select(aged, dob_yy, sex, mracerec) %>%
                      dplyr::mutate(race = ifelse((mracerec==1), "white", 
                                          ifelse((mracerec==2), "black", "other"))) %>%
                      dplyr::select(-mracerec) %>%
                      dplyr::mutate(status=1)

km_infant_deaths_VLBW <- do.call("rbind", list(km_infant_deaths_1995_VLBW, 
                                               km_infant_deaths_2007_VLBW,
                                               km_infant_deaths_2013_VLBW))
km_infant_deaths_ELBW <- do.call("rbind", list(km_infant_deaths_1995_ELBW, 
                                               km_infant_deaths_2007_ELBW,
                                               km_infant_deaths_2013_ELBW))
km_infant_deaths_VLBW$race <- as.factor(km_infant_deaths_VLBW$race)
km_infant_deaths_ELBW$race <- as.factor(km_infant_deaths_ELBW$race)
km_infant_deaths_VLBW$sex <- as.factor(km_infant_deaths_VLBW$sex)
km_infant_deaths_ELBW$sex <- as.factor(km_infant_deaths_ELBW$sex)
```


## Introduction{.bigger}

- Infant mortality in the United States is a much discussed topic among population health policy makers and researchers.
  - Infant mortality rates are often associated with the health of nations across the world.
    - In the US much of infant mortality is amongst the most immature infants. 
- Much of the recent conversation in neonatal medicine is on the increasing survival amongst infants born less than 1,500g (VLBW) or less than 32 weeks.
  - Recent studies have shown increases in survival amongst the smallest and most premature infants. 
    - This is largely the result of therapeutic interventions introduced over the last 20 years and, potentially, more frequent applications to younger infants.
- Little information is available on how trends in care practices have impacted death in infancy.
    - Specifically among infants born VLBW and ELBW.

## Hypothesis {.bigger}

- Among infants born VLBW or ELBW there will be a temporal trend towrds increasing time before death among infants born less than 1,000g. 

## Data Sources {.bigger}

- Linked infant birth and death certificate data available from the National Bureau of Economic Reserach from **1995 through 2013**.
- CDC Wonder queries made throught the CDC Wonder website. **1995 through 2017**.


## Infant Mortality Rate 1995

```{r IMR 1995, echo = FALSE, warning=FALSE, fig.align = "center", fig.width=10, fig.height=6}
All_Infant_Death_1995_By_State <- Wonder_Files_df[[1]]
All_Infant_Death_1995_By_State <- rename(All_Infant_Death_1995_By_State, state_name = State)
All_Infant_Death_1995_By_State <- rename(All_Infant_Death_1995_By_State, death_rate = "Death.Rate")

state_data_infant_death_1995 <- left_join(All_Infant_Death_1995_By_State, states, 
                                          by = "state_name")
state_data_infant_death_1995 %>%
  ggplot(aes(long, lat, group = group, fill = death_rate)) +
  ggtitle("Infant Death Rate in the US 1995") +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Infant Death Rate")
```

## Infant Mortality Rate 2007

```{r IMR 2007, echo = FALSE, warning=FALSE, fig.align = "center", fig.width=10, fig.height=6}
All_Infant_Death_2007_By_State <- Wonder_Files_df[[13]]
All_Infant_Death_2007_By_State <- rename(All_Infant_Death_2007_By_State, state_name = State)
All_Infant_Death_2007_By_State <- rename(All_Infant_Death_2007_By_State, death_rate = "Death.Rate")

state_data_infant_death_2007 <- left_join(All_Infant_Death_2007_By_State, states, 
                                          by = "state_name")
state_data_infant_death_2007 %>%
  ggplot(aes(long, lat, group = group, fill = death_rate)) +
  ggtitle("Infant Death Rate in the US 2007") +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Infant Death Rate")
```

## Infant Mortality Rate 2017

```{r IMR 2017, echo = FALSE, warning=FALSE, fig.align = "center", fig.width=10, fig.height=6}
All_Infant_Death_2017_By_State <- Wonder_Files_df[[23]]
All_Infant_Death_2017_By_State <- rename(All_Infant_Death_2017_By_State, state_name = State)
All_Infant_Death_2017_By_State <- rename(All_Infant_Death_2017_By_State, death_rate = "Death.Rate")

state_data_infant_death_2017 <- left_join(All_Infant_Death_2017_By_State, states, 
                                          by = "state_name")
state_data_infant_death_2017 %>%
  ggplot(aes(long, lat, group = group, fill = death_rate)) +
  ggtitle("Infant Death Rate in the US 2017") +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Infant Death Rate")
```

## What Portion of the IMR is Accounted for by VLBW Infants?{.bigger}
- In 1995 **`r round(Percent_1995_IMR_VLBW*100, digits=1)`** percent of the IMR in the United States was accounted for by those infants born VLBW and **`r round(Percent_1995_IMR_ELBW*100, digits=1)`** percent from infants born ELBW.
- By 2017 the percent of US infant mortality accounted for by VLBW infants rose to **`r round(Percent_2017_IMR_VLBW*100, digits=1)`**. **`r round(Percent_2017_IMR_ELBW*100, digits=1)`** percent of IMR is now from infants born ELBW.
- Interestingly in 1995 **`r round(Percent_1995_IMR_lt_23*100, digits = 1)`** percent of IMR was account for by infants 22 weeks or less and that percent also rose to **`r round(Percent_2017_IMR_lt_23*100, digits = 1)`** percent.
  - These infants, until recently, were considered non-viable and not offered support.

## Improvements in IMR in Infants 1,000 to 1,499g.
```{r IMR in VLBW, echo = FALSE, warning=FALSE, message=FALSE}
aa <- ggplot(All_VLBW_IMR, aes(x=Year.of.Death, y=Death.Rate))
bb <- aa +geom_point(aes(colour = Year.Group, size=Deaths)) +
      scale_color_brewer(palette="Blues") +
      ggtitle("IMR in Infants 1,000 to 1,499g") + 
      xlab("Year") +
      ylab("IMR per 1,000 live births") +
      labs(Size="Number of Deaths", colour="Year Group") +
      theme_dark()
bb
```

## Improvements in IMR in Infants 500-999g.
```{r IMR in ELBW, echo = FALSE, warning=FALSE, message=FALSE}
cc <- ggplot(All_ELBW_IMR, aes(x=Year.of.Death, y=Death.Rate))
dd <- cc +geom_point(aes(colour = Year.Group, size=Deaths)) +
      scale_color_brewer(palette="Blues") +
      ggtitle("IMR in Infants 500 to 999g") + 
      xlab("Year") +
      ylab("IMR per 1,000 live births") +
      labs(Size="Number of Deaths", colour="Year Group") + 
      theme_dark()
dd
```

## Timing Prior to Death in Infacy Over Time for ELBW.

```{r ELBW Day Before Death, echo = FALSE, warning=FALSE, message=FALSE}
ee <- ggplot(data=ELBW_rbind_ad_summ, aes(x=Year, y=mean_aged))
ff <- ee + geom_point(color="red")+
        geom_errorbar(aes(ymin=mean_aged-se_aged, ymax=mean_aged+se_aged), width=.1) +
        ggtitle("Mean Age of Death Among ELBW Infants") +
        ylab("Age of Death(Days)") + 
        xlab("Year")
ff
```

## Timing Prior to Death in Infacy Over Time for VLBW.

```{r VLBW Day Before Death, echo = FALSE, warning=FALSE, message=FALSE}
gg <- ggplot(data=VLBW_rbind_ad_summ, aes(x=Year, y=mean_aged))
hh <- gg + geom_point(color="red") +
        geom_errorbar(aes(ymin=mean_aged-se_aged, ymax=mean_aged+se_aged), width=.1) +
        ggtitle("Mean Age of Death Among VLBW Infants") +
        ylab("Age of Death(Days)") +
        xlab("Year")
hh
```

## Distribution of Time to Death for the Entire ELBW Cohort.

```{r Distribution of Age of Death ELBW, message=FALSE}
ggplot(data=ELBW_rbind_ad, aes(x=aged)) + geom_histogram(color="darkblue", fill="lightblue") +xlab("Aged of Death") + ylab("Number of Deaths") +
  ggtitle("Distribution of Time of Death, Entire ELBW Cohort") + annotate("text", x = 200, y = 62500, label = "- Clearly non-normally distributed.") + annotate("text", x = 200, y = 56250, label = "- Median for most individual years between 0-2.")
```

## Distribution of Time to Death for the Entire VLBW Cohort.

```{r Distribution of Age of Death VLBW, message=FALSE}
ggplot(data=VLBW_rbind_ad, aes(x=aged)) + 
  geom_histogram(color="darkblue", fill="lightblue") +
  xlab("Aged of Death") + 
  ylab("Number of Deaths") +
  ggtitle("Distribution of Time of Death, Entire ELBW Cohort") + 
  annotate("text", x = 200, y = 62500, label = "- Clearly non-normally distributed again.") + 
  annotate("text", x = 200, y = 56250, label = "- Median values similar.") +
  annotate("text", x = 200, y = 50000, label = "- ELBW cohort is contained withing the VLBW cohort.")
```

## Timing Prior to Death for ELBW Infants Survivng More Than 7d.

```{r ELBW  GT 7d Day Before Death, echo = FALSE, warning=FALSE, message=FALSE}
ii <- ggplot(data=ELBW_surv_8d_summ, aes(x=Year, y=mean_aged))
jj <- ii + geom_point(color="red")+
        geom_errorbar(aes(ymin=mean_aged-se_aged, ymax=mean_aged+se_aged), width=.1) +
        ggtitle("Mean Age of Death Among ELBW Infants Surviving > 7d") +
        ylab("Age of Death(Days)") + 
        xlab("Year")
jj
```

## Timing Prior to Death for VLBW Infants Survivng More Than 7d.

```{r VLBW  GT 7d Day Before Death, echo = FALSE, warning=FALSE, message=FALSE}
kk <- ggplot(data=VLBW_surv_8d_summ, aes(x=Year, y=mean_aged))
ll <- kk + geom_point(color="red")+
        geom_errorbar(aes(ymin=mean_aged-se_aged, ymax=mean_aged+se_aged), width=.1) +
        ggtitle("Mean Age of Death Among VLBW Infants Surviving > 7d") +
        ylab("Age of Death(Days)") + 
        xlab("Year")
ll
```

## Kaplan Meier Analysis 1995, 2007 and 2013 - Basic Survival Plot

```{r KM analysis 1995 through 2013}
km_infant_death <- with(km_infant_deaths_VLBW, Surv(aged, status))
km_infant_death_fit <- survfit(Surv(aged, status) ~ 1, data=km_infant_deaths_VLBW)
#summary(km_infant_death_fit, times = c(0,7,14,28,42,56,70,84,98,112,140,168,196,224,365))
#plot(km_infant_death_fit, xlab="Days", main = 'Kaplan Meier Curve for Entire Cohort')
autoplot(km_infant_death_fit)
```

## Kaplan Meier Analysis 1995, 2007 and 2013 - Impact of Year

```{r KM Year}
km_infant_death_year_fit <- survfit(Surv(aged, status) ~ dob_yy, data=km_infant_deaths_VLBW)
#autoplot(km_infant_death_year_fit)
ggsurvplot(km_infant_death_year_fit, data = km_infant_deaths_VLBW)
```

## Kaplan Meier Analysis 1995, 2007 and 2013 - Impact of Race

```{r}
km_infant_death_race_fit <- survfit(Surv(aged, status) ~ race, data=km_infant_deaths_VLBW)
#autoplot(km_infant_death_race_fit)
ggsurvplot(km_infant_death_race_fit, data = km_infant_deaths_VLBW)
```

## Kaplan Meier Analysis 1995, 2007 and 2013 - Impact of Sex

```{r}
km_infant_death_sex_fit <- survfit(Surv(aged, status) ~ sex, data=km_infant_deaths_VLBW)
#autoplot(km_infant_death_sex_fit)
ggsurvplot(km_infant_death_sex_fit, data = km_infant_deaths_VLBW)
```

## Cox Proportional Hazard Model

```{r cph Forest Plot, message=FALSE, warning=FALSE}
infant_death_VLBW_cox <- coxph(Surv(aged, status) ~ dob_yy + sex + race , data = km_infant_deaths_VLBW)
#summary(infant_death_VLBW_cox)
ggforest(infant_death_VLBW_cox, data = km_infant_deaths_VLBW)
```

## Summary {.bigger}

- It appears that VLBW infants that died in infancy did see an increase in the time before death from **1995 through 2007**. 
_ However, timing more recently appears to be back to previous values.
  - This likely represents more infants living overall and not dying in infancy while those infants unlikely to respond to current therapies are dying early again.
    - Caffeine, CPAP, iNO may have have roles to play.
    - So may parental and health care giver attitudes towards resuscitation.
- Amongst the smallest birth weight infants there does seem to be an increasing time beofre death in the most recent era.
- African American infants appear to live longer then caucasian or other racial groups before death in infancy.

