---
title: "Changing Nature of Death in Infancy in the United States"
author: "Kevin Dysart"
date: "12/1/2019"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r loading packages, echo=FALSE, warning=FALSE, message=FALSE}
library("tidyverse")
library("urbnmapr")
library("lubridate")
library("survival")
library("survminer")
library("RColorBrewer")
library("knitr")
```

```{r reading in dataframes for slides, echo=FALSE, warning=FALSE, message=FALSE}
#Reading CDC Wonder Data Files
Wonder_File_Names <- list.files(path = "Linked_Birth_Death_Data/CDC_Wonder_All_Death_By_Year/", pattern = "*.txt")
Wonder_Files <- list.files(path = "Linked_Birth_Death_Data/CDC_Wonder_All_Death_By_Year/", pattern = "*.txt", full.names = T)
Wonder_Files_df <- lapply(Wonder_Files, read.delim)
names(Wonder_Files_df) <- Wonder_File_Names
#Reading Linked Birth and Death Data
Linked_Birth_Death_Names <- list.files(path = "Linked_Birth_Death_Data/Linked_Year_CSV_Zip/", pattern = "*.csv.zip")
Linked_Birth_Death_Files <- list.files(path = "Linked_Birth_Death_Data/Linked_Year_CSV_Zip/", pattern = "*.csv.zip", full.names = T)
Linked_Birth_Death_Files_df <- lapply(Linked_Birth_Death_Files, read_csv)
names(Linked_Birth_Death_Files_df) <- Linked_Birth_Death_Names
```

```{r removing Notes column, echo=FALSE, warning=FALSE, message=FALSE}
for (i in 1:length(Wonder_Files_df)) {
  Wonder_Files_df[[i]] <-  Wonder_Files_df[[i]] %>%
                    select(-Notes) %>%
                    slice(1:51)
}
```

```{r Reading in IMR files for VLBW and ELBW, echo=FALSE, warning=FALSE, message=FALSE}
## Reading in IMR files from CDC 1995 through 2017 ELBW VLBW.
IMR_VLBW_ELBW_Names <- list.files(path = "Linked_Birth_Death_Data/IMR_VLBW_ELBW/", pattern = "*.txt")
IMR_VLBW_ELBW_Files <- list.files(path = "Linked_Birth_Death_Data/IMR_VLBW_ELBW/", pattern = "*.txt", full.names = T)
IMR_VLBW_ELBW_df <- lapply(IMR_VLBW_ELBW_Files, read.delim)
names(IMR_VLBW_ELBW_df) <- IMR_VLBW_ELBW_Names
## Cleaning each dataframe in the list. ELBW and VLBW Dataframes are paired down the list.
ELBW_1995 <- IMR_VLBW_ELBW_df[[1]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_1995 <- IMR_VLBW_ELBW_df[[2]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_1999 <- IMR_VLBW_ELBW_df[[3]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_1999 <- IMR_VLBW_ELBW_df[[4]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_2003 <- IMR_VLBW_ELBW_df[[5]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_2003 <- IMR_VLBW_ELBW_df[[6]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_2007 <- IMR_VLBW_ELBW_df[[7]] %>%
                select(-Notes) %>%
                    slice(1:11)
VLBW_2007 <- IMR_VLBW_ELBW_df[[8]] %>%
                select(-Notes) %>%
                    slice(1:11)
All_ELBW_IMR <- do.call("rbind", list(ELBW_1995, ELBW_1999, ELBW_2003, ELBW_2007))
All_VLBW_IMR <- do.call("rbind", list(VLBW_1995, VLBW_1999, VLBW_2003, VLBW_2007))
All_VLBW_IMR$Year.Group <- c(rep("1995-1998", 4), rep("1999-2002", 4),
                             rep("2003-2006", 4), rep("2007-2017", 11))
All_ELBW_IMR$Year.Group <- c(rep("1995-1998", 4), rep("1999-2002", 4),
                             rep("2003-2006", 4), rep("2007-2017", 11))
All_ELBW_IMR$Year.Group <- as.factor(All_ELBW_IMR$Year.Group)
All_VLBW_IMR$Year.Group <- as.factor(All_VLBW_IMR$Year.Group)
```

```{r Cleaning Data for IMR Description, echo=FALSE, warning=FALSE, message=FALSE}
## Demographic data cleaning
linked_1995 <- Linked_Birth_Death_Files_df[[1]]
linked_1995_VLBW <- filter(linked_1995, dbirwt<1500)
Percent_1995_IMR_VLBW <- length(linked_1995_VLBW$matchs)/length(linked_1995$matchs)
linked_2017 <- Linked_Birth_Death_Files_df[[19]]
linked_2017_VLBW <- filter(linked_2017, dbwt<1500)
Percent_2017_IMR_VLBW <- length(linked_2017_VLBW$idnumber)/length(linked_2017$idnumber)
linked_1995_ELBW <- filter(linked_1995, dbirwt<1000)
linked_2017_ELBW <- filter(linked_2017, dbwt<1000)
Percent_1995_IMR_ELBW <- length(linked_1995_ELBW$matchs)/length(linked_1995$matchs)
Percent_2017_IMR_ELBW <- length(linked_2017_ELBW$idnumber)/length(linked_2017$idnumber)
linked_1995_lt_23 <- filter(linked_1995, gestat<23)
linked_2017_lt_23 <- filter(linked_2017, combgest<23)
Percent_1995_IMR_lt_23 <- length(linked_1995_lt_23$matchs)/length(linked_1995$matchs)
Percent_2017_IMR_lt_23 <- length(linked_2017_lt_23$idnumber)/length(linked_2017$idnumber)
```


```{r Creating Data Frames for boxplot of age of death, echo=FALSE, warning=FALSE, message=FALSE}

##This chunk creates a large data frame for the year of birth for each infant in the linked death and birth certificate data. Because infants less than 22 weeks have not routinely been offered care in the United States unitl recent years.

ELBW_1995_linked_ad <- Linked_Birth_Death_Files_df[[1]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1996_linked_ad <- Linked_Birth_Death_Files_df[[2]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1997_linked_ad <- Linked_Birth_Death_Files_df[[3]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1998_linked_ad <- Linked_Birth_Death_Files_df[[4]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_1999_linked_ad <- Linked_Birth_Death_Files_df[[5]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2000_linked_ad <- Linked_Birth_Death_Files_df[[6]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2001_linked_ad <- Linked_Birth_Death_Files_df[[7]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2002_linked_ad <- Linked_Birth_Death_Files_df[[8]] %>%
                select(biryr, aged, dbirwt, gestat) %>%
                filter(dbirwt<1000) %>%
                filter(gestat > 22) %>%
                rename(Year=biryr, GA=gestat, BW=dbirwt) 
ELBW_2003_linked_ad <- Linked_Birth_Death_Files_df[[9]] %>%
                select(dob_yy, aged, dbwt, combgest) %>%
                filter(dbwt<1000) %>%
                filter(combgest > 22) %>%
                rename(Year=dob_yy, GA=combgest, BW=dbwt) 

ELBW_rbind_ad <- do.call("rbind", list(ELBW_1995_linked_ad,
                                       ELBW_1996_linked_ad,
                                       ELBW_1997_linked_ad,
                                       ELBW_1998_linked_ad,
                                       ELBW_1999_linked_ad,
                                       ELBW_2000_linked_ad))
ELBW_rbind_ad$Year <- as.factor(ELBW_rbind_ad$Year)
ELBW_rbind_ad_group <- group_by(ELBW_rbind_ad, Year)
ELBW_rbind_ad_summ <- summarise(ELBW_rbind_ad_group, mean_aged = mean(aged),
                                median_aged = median(aged), 
                                sd_aged = sd(aged),
                                n=n())

VLBW_1995_linked_ad <- Linked_Birth_Death_Files_df[[1]] %>%
                select(biryr, aged, dbirwt) %>%
                filter(dbirwt<1500) 
VLBW_1995 <- IMR_VLBW_ELBW_df[[2]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_1999 <- IMR_VLBW_ELBW_df[[3]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_1999 <- IMR_VLBW_ELBW_df[[4]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_2003 <- IMR_VLBW_ELBW_df[[5]] %>%
                select(-Notes) %>%
                    slice(1:4)
VLBW_2003 <- IMR_VLBW_ELBW_df[[6]] %>%
                select(-Notes) %>%
                    slice(1:4)
ELBW_2007 <- IMR_VLBW_ELBW_df[[7]] %>%
                select(-Notes) %>%
                    slice(1:11)
VLBW_2007 <- IMR_VLBW_ELBW_df[[8]] %>%
                select(-Notes) %>%
                    slice(1:11)
All_ELBW_IMR <- do.call("rbind", list(ELBW_1995, ELBW_1999, ELBW_2003, ELBW_2007))
All_VLBW_IMR <- do.call("rbind", list(VLBW_1995, VLBW_1999, VLBW_2003, VLBW_2007))

```

## Introduction{.bigger}

- Infant mortality in the United States is a much discussed topic among population health policy makers and researchers.
  - Infant mortality rates are often associated with the health of nations across the world.
    - In the US much of infant mortality is amongst the most immature infants. 
- Much of the recent conversation in neonatal medicine is on the increasing survival amongst infants born less than 1,500g (VLBW) or less than 32 weeks.
  - Recent studies have shown increases in survival amongst the smallest and most premature infants. 
    - This is largely the result of therapeutic interventions introduced over the last 20 years and, potentially, more frequent applications to younger infants.
- Little information is available on how trends in care practices have impacted death in infancy.
    - Specifically among infants born VLBW and ELBW.

## Hypothesis {.bigger}

- Among infants born VLBW or ELBW there will be a temporal trend towrds increasing time before death among infants born less than 1,000g. 

## Infant Mortality Rate 1995

```{r IMR 1995, echo = FALSE, warning=FALSE}
All_Infant_Death_1995_By_State <- Wonder_Files_df[[1]]
All_Infant_Death_1995_By_State <- rename(All_Infant_Death_1995_By_State, state_name = State)
All_Infant_Death_1995_By_State <- rename(All_Infant_Death_1995_By_State, death_rate = "Death.Rate")

state_data_infant_death_1995 <- left_join(All_Infant_Death_1995_By_State, states, 
                                          by = "state_name")
state_data_infant_death_1995 %>%
  ggplot(aes(long, lat, group = group, fill = death_rate)) +
  ggtitle("Infant Death Rate in the US 1995") +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Infant Death Rate")
```

## Infant Mortality Rate 2007

```{r IMR 2007, echo = FALSE, warning=FALSE}
All_Infant_Death_2007_By_State <- Wonder_Files_df[[13]]
All_Infant_Death_2007_By_State <- rename(All_Infant_Death_2007_By_State, state_name = State)
All_Infant_Death_2007_By_State <- rename(All_Infant_Death_2007_By_State, death_rate = "Death.Rate")

state_data_infant_death_2007 <- left_join(All_Infant_Death_2007_By_State, states, 
                                          by = "state_name")
state_data_infant_death_2007 %>%
  ggplot(aes(long, lat, group = group, fill = death_rate)) +
  ggtitle("Infant Death Rate in the US 2007") +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Infant Death Rate")
```

## Infant Mortality Rate 2017

```{r IMR 2017, echo = FALSE, warning=FALSE}
All_Infant_Death_2017_By_State <- Wonder_Files_df[[23]]
All_Infant_Death_2017_By_State <- rename(All_Infant_Death_2017_By_State, state_name = State)
All_Infant_Death_2017_By_State <- rename(All_Infant_Death_2017_By_State, death_rate = "Death.Rate")

state_data_infant_death_2017 <- left_join(All_Infant_Death_2017_By_State, states, 
                                          by = "state_name")
state_data_infant_death_2017 %>%
  ggplot(aes(long, lat, group = group, fill = death_rate)) +
  ggtitle("Infant Death Rate in the US 2017") +
  geom_polygon(color = NA) +
  coord_map(projection = "albers", lat0 = 39, lat1 = 45) +
  labs(fill = "Infant Death Rate")
```

## What Portion of the IMR is Accounted for by VLBW Infants?{.bigger}
- In 1995 `r round(Percent_1995_IMR_VLBW*100, digits=1)` percent of the IMR in the United States was accounted for by those infants born VLBW and `r round(Percent_1995_IMR_ELBW*100, digits=1)` percent from infants born ELBW.
- By 2017 the percent of US infant mortality accounted for by VLBW infants rose to `r round(Percent_2017_IMR_VLBW*100, digits=1)`. `r round(Percent_2017_IMR_ELBW*100, digits=1)` percent of IMR is now from infants born ELBW.
- Interestingly in 1995 `r round(Percent_1995_IMR_lt_23*100, digits = 1)` percent of IMR was account for by infants 22 weeks or less and that percent also rose to `r round(Percent_1995_IMR_lt_23*100, digits = 1)` percent.
  - These infants, until recently, were considered non-viable and not offered support.

## Improvements in IMR in Infants 1,000 to 1,499g.
```{r IMR in VLBW}
aa <- ggplot(All_VLBW_IMR, aes(x=Year.of.Death, y=Death.Rate))
bb <- aa +geom_point(aes(colour = Year.Group, size=Deaths)) +
      scale_color_brewer(palette="Blues") +
      ggtitle("IMR in Infants 1,000 to 1,499g") + 
      xlab("Year") +
      ylab("IMR") +
      labs(Size="Number of Deaths", colour="Year Group") +
      theme_dark()
bb
```

## Improvements in IMR in Infants 500-999g.
```{r IMR in ELBW}
cc <- ggplot(All_ELBW_IMR, aes(x=Year.of.Death, y=Death.Rate))
dd <- cc +geom_point(aes(colour = Year.Group, size=Deaths)) +
      scale_color_brewer(palette="Blues") +
      ggtitle("IMR in Infants 500 to 999g") + 
      xlab("Year") +
      ylab("IMR") +
      labs(Size="Number of Deaths", colour="Year Group") + 
      theme_dark()
dd
```

## Timing Prior to Death in Infacy Over Time.

```{r}

```

