---
title: "Preliminary Statistics"
author: "Erin Kennedy"
date: "11/24/2020"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Purpose
This document aligns with the preliminary analysis section in the main analysis rmd file. Although I did not eliminate any values based on univariable analysis, I did run them for all variables of interest using Chi-squared tests and univariable logistic regression models comparing categorized to dichotomized results where applicable. Those results are included below, starting with chi-squared tests. 

Loading data 
```{r}
# Dataset
mimic <- read.csv(file = 'mimic.csv')
```


# Univariable Analysis Pre-Dichotomization

## Chi Squared Tests
Chi-Squared Tests of Variables (Pre-Dichotomization) to look at univariable relationship between clinically important predictors and 30-day readmissions. Significant variables included  

**Categorical/continuous variables**  

* Admission type (p=0.019)
* Discharge categories (p<0.001)
* Insurance (p=0.02)
* Ethnicity (p<0.001)
* Elixhauser comorbidities (p<0.001)
* LOS (p<0.001)

  
**Dichotomized variables**  

* Admission type (p=0.006)
* Discharge Disposition (p<0.001)
* Insurance (p=0.01)
* Ethnicity (p=0.004)
* Surgical vs non.surgical (p=0.03)
* Elixhauser comorbidities (p<0.001)
* LOS (p<0.001)
```{r}
# Making the readmission variable a factor
mimic$readmitted_30 <- as.factor(mimic$readmitted_30)

# Gender
chisq.test(table(mimic$readmitted_30, mimic$GENDER)) 

# Admission Type
chisq.test(table(mimic$readmitted_30, mimic$ADMISSION_TYPE)) 
chisq.test(table(mimic$readmitted_30, mimic$admission_dichotomous))

# Discharge Location (not going to be in the model but interesting to look at)
chisq.test(table(mimic$readmitted_30, mimic$discharge_categories)) 
chisq.test(table(mimic$readmitted_30, mimic$discharge_dichotomous))

# Insurance
chisq.test(table(mimic$readmitted_30, mimic$INSURANCE)) 
chisq.test(table(mimic$readmitted_30, mimic$insurance_categories)) 

# Religion
chisq.test(table(mimic$readmitted_30, mimic$religion_categories)) 

# Marital Status
chisq.test(table(mimic$readmitted_30, mimic$marital_dichotomous)) 

# Ethnicity
chisq.test(table(mimic$readmitted_30, mimic$ethnicity_categories)) 
chisq.test(table(mimic$readmitted_30, mimic$ethnicity_dichotomous)) 

# Service
chisq.test(table(mimic$readmitted_30, mimic$service)) 

# Elixhauser
chisq.test(table(mimic$readmitted_30, mimic$elixhauser_categories)) 
chisq.test(table(mimic$readmitted_30, mimic$elixhauser_dichotomous)) 

# Age
chisq.test(table(mimic$readmitted_30, mimic$AGE)) 
chisq.test(table(mimic$readmitted_30, mimic$age_dichotomous)) 

# LOS - 
chisq.test(table(mimic$readmitted_30, mimic$LOS)) 
chisq.test(table(mimic$readmitted_30, mimic$los_dichotomous)) 
```

## Univariable Logistic Regression
Univariable Logistic Regression between clinically important predictors and readmissions. 

**Categorical/continuous variables**  

* Admission type: Emergency (p=0.005)
* LOS (p<0.001)


**Dichotomized variables**  

* Admission type: Emergency/Urgent (p<0.001)
* Discharge Disposition: Non-Home (p<0.001)
* Insurance: Public (p=0.004)
* Ethnicity: Non-white (p=0.004)
* Surgical vs non.surgical (p=0.02)
* Elixhauser comorbidities: 1 or more (p<0.001)
* LOS (p<0.001)
```{r}
# Gender
summary((glm(readmitted_30 ~ GENDER, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ GENDER, data = mimic, family = binomial()))) #gives us the coefficients

# Admission Type
summary((glm(readmitted_30 ~ ADMISSION_TYPE, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ ADMISSION_TYPE, data = mimic, family = binomial())))

summary((glm(readmitted_30 ~ admission_dichotomous, data = mimic, family = binomial()))) #dichotomous
exp(coef(glm(readmitted_30 ~ admission_dichotomous, data = mimic, family = binomial())))

# Discharge Location (not going to be in the model but interesting to look at)
summary((glm(readmitted_30 ~ discharge_categories, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ discharge_categories, data = mimic, family = binomial())))

summary((glm(readmitted_30 ~ discharge_dichotomous, data = mimic, family = binomial()))) #dichotomous
exp(coef(glm(readmitted_30 ~ discharge_dichotomous, data = mimic, family = binomial())))

# Insurance
summary((glm(readmitted_30 ~ INSURANCE, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ INSURANCE, data = mimic, family = binomial())))

summary((glm(readmitted_30 ~ insurance_categories, data = mimic, family = binomial()))) #categorized
exp(coef(glm(readmitted_30 ~ insurance_categories, data = mimic, family = binomial())))

# Marital Status
summary((glm(readmitted_30 ~ marital_dichotomous, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ marital_dichotomous, data = mimic, family = binomial())))

# Ethnicity
mimic$ethnicity_categories <- as.factor(mimic$ethnicity_categories)
mimic$ethnicity_categories = relevel(mimic$ethnicity_categories, ref = 5) #making white the reference category
summary((glm(readmitted_30 ~ ethnicity_categories, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ ethnicity_categories, data = mimic, family = binomial()))) 

mimic$ethnicity_dichotomous <- as.factor(mimic$ethnicity_dichotomous)
mimic$ethnicity_dichotomous = relevel(mimic$ethnicity_dichotomous, ref = 2) #making white the reference category
summary((glm(readmitted_30 ~ ethnicity_dichotomous, data = mimic, family = binomial()))) #dichotomous
exp(coef(glm(readmitted_30 ~ ethnicity_dichotomous, data = mimic, family = binomial())))

# Service
summary((glm(readmitted_30 ~ service, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ service, data = mimic, family = binomial())))

# Insurance
summary((glm(readmitted_30 ~ elixhauser_categories, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ elixhauser_categories, data = mimic, family = binomial())))

summary((glm(readmitted_30 ~ elixhauser_dichotomous, data = mimic, family = binomial()))) #categorized
exp(coef(glm(readmitted_30 ~ elixhauser_dichotomous, data = mimic, family = binomial())))

# Age
summary((glm(readmitted_30 ~ AGE, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ AGE, data = mimic, family = binomial())))

summary((glm(readmitted_30 ~ age_dichotomous, data = mimic, family = binomial()))) #dichotomous
exp(coef(glm(readmitted_30 ~ age_dichotomous, data = mimic, family = binomial())))

# LOS - 
summary((glm(readmitted_30 ~ LOS, data = mimic, family = binomial())))
exp(coef(glm(readmitted_30 ~ LOS, data = mimic, family = binomial())))

summary((glm(readmitted_30 ~ los_dichotomous, data = mimic, family = binomial()))) #dichotomous
exp(coef(glm(readmitted_30 ~ los_dichotomous, data = mimic, family = binomial())))
```

