---
title: "BMIN503 Project: Protein Crystal Structure Disulfide Bond Analysis"
author: "Shruthi Murali"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: cerulean 
    highlight: tango
    editor_options: 
    chunk_output_type: inline
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
Use this template to complete your project throughout the course. Your Final Project presentation will be based on the contents of this document. Replace the title/name above and text below with your own, but keep the headers.

### Overview
In proteins, disulfide bonds form when two cysteine residues are in close contact. These bonds increase the shelf stability of the protein and have been shown to increase melting temperature in mutation studies. The goal of my project is to take an input protein crystal structure text file of type .pdb from the Protein Data Bank (PDB) to clean and format the data into a usable dataframe with minimal manual changes and determine if I can find the existing disulfide bonds and propose
new mutations to form new disulfide bonds. 

The goal of my project has changed since I spoke to Dr. Tsourkas and a scientist in the lab, but much of what would be useful to use in PDB data is similar. I have learned from both of them that it is possible to find angles and positions of residues from the PDB data file, which I previously have not done, as I have mainly only used the common visualization tools in the field. I also have realized that excluding certain residues from the mutation analysis is important to retain function of the protein. Since every protein is different, these residues have to be hard-coded. 

Link to my final project GitHub Repository: https://github.com/coffeebean21/BMIN503_Final_Project

### Introduction 
The problem being addressed with this code is how to process and make PDB data usable in R. More specifically, this project helps to identify disulfide bonds that are already formed in the protein based on cysteine pairs, or to identify residues to be mutated to a cysteine pair such that they form a disulfide bond. This information would be useful to engineer new disulfide bonds
into a protein to increase stability, but would be preliminary data and likely need further validation with visualization tools to determine if mutations are in optimal positions. 

This problem is interdisciplinary and contributes to protein engineering because it provides a high-throughput computational approach to determine information about a protein without using limited wet-lab resources.  

### Methods
Describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

```{r eval = TRUE, include = FALSE, message = FALSE, warning = FALSE}
#Load packages
library(dplyr)
library(ggplot2)
```


>Import .pdb as .txt file. This is the protein data file for IgG2a monoclonal antibody.
The PDB website reference is: https://www.rcsb.org/structure/1igt
The website for the primary manuscript for this PDB file is: https://pubs.acs.org/doi/10.1021/bi962514%2B. 
Load in text file with only "ATOM" entries. This is a limitation because I am not sure of a way to do this automatically in R, as every .pdb file may have a different header/number of rows at the beginning. 

### STEP 1: FINDING THE DATA IN THE TEXT FILE AND ARRANGING INTO A DATAFRAME

```{r eval = TRUE, message = FALSE, warning = FALSE}
lines <- readLines("./1igt_clean.txt")

```

> This is what the raw data looks like:

```{r eval = TRUE, warning = FALSE}
head(lines)
```

```{r eval = TRUE, message = FALSE, warning = FALSE}
#Initialize vectors to contain the split entries and one to merge all of them together
temp.clean.split <- c()
temp.clean.merge <- c()

#Loop through all of the lines in the .txt 
for (i in 1:length(lines))
{
  temp.line <- lines[i] #Store the working line
  temp.split <- as.data.frame(strsplit(temp.line, " ", fixed = FALSE)) #Split the working line
  temp.index <- temp.split != ""
  temp.no.space <- as.data.frame(temp.split[temp.index]) #Extracting only the values and not the spaces in between the values that are present in the original data format
  temp.clean.merge <- c(temp.clean.merge, temp.no.space) #
}
```

> This list now contains only the values and no spaces from each of the rows in the .txt file:

```{r eval = TRUE, message = FALSE, warning = FALSE}
head(temp.clean.merge)
```


### STEP 2: ORGANIZING CLEANED TEXT FILE DATA INTO LABELED DATAFRAME COLUMNS

```{r eval = TRUE, warning = FALSE, message = FALSE}
#Initialize the new data frame with all columns labeled with information from .pdb
pdb.df <- data.frame(matrix(ncol = 12))
columns <- c("Type", "Atom Number", "Element", "Residue AA", "Chain", 
             "Residue Number", "X-Position", "Y-Position", "Z-Position", 
             "Occupancy Factor", "Temperature Factor", "Element2")
colnames(pdb.df) <- columns
  
#Outer: loop through all the list rows and extract each row
for (j in 1:length(temp.clean.merge))
{
  temp.atom <- temp.clean.merge[[j]]
  
  #Inner: loop through the columns of the stored row and store in labeled df column
  for (k in 1:length(temp.atom))
  {
    pdb.df[j,k] <- temp.atom[k]
  }
}
```

>This is what the organized data in a dataframe looks like:

```{r eval = TRUE, warning = FALSE}
head(pdb.df)
```

### STEP 3: GENERATION OF A NEW DATAFRAME FOR EACH OF THE CHAINS PRESENT

```{r eval = TRUE}
#.pdb files often contain multiple chains representing different parts of the protein structure. The structure being used in this case is the Human IgG structure (antibody), which consists of 4 chains: A, B, C, and D, where each chain consists of parts of the whole antibody, for example, the constant regions and variable regions. 

#Remove "ter" type and retain "atom" type. These are some values that are not relevant to my question that I am removing.
ter.type.index <- pdb.df$Type == 'TER'
pdb.df <- pdb.df[!ter.type.index,]

#Find the unique chains
chains <- pdb.df %>% distinct(pdb.df$Chain)

#Create a new dataframe for each chain
#Loop: through the number of chains and find matching entries in the pdb.df large dataframe to assign to a new variable named as the chain. 
for (l in 1:nrow(chains))
{
  df.name <- paste0("Chain", chains[[1]][l])
  temp.df.index <- pdb.df$Chain == chains[[1]][l]
  assign(df.name, pdb.df[temp.df.index,])
}
```

> This is what one of the new chain dataframes looks like. All the chain values should be the same and it should have split the entire pdb.df dataframe by chain label. 

```{r eval = TRUE, warning = FALSE}
head(ChainA)
```


### STEP 4: GENERATION OF A NEW DATAFRAME WITH AVERAGE XYZ POSITION PER RESIDUE PER CHAIN

> Calculation of avg positions of each residue = avg(all atoms for that residue). Needed to hard code for the chains I know are in this dataset. By this, I mean that there is no general loop to generate new dataframes for the average XYZ position. I have created and filled in using loops for each individual chain dataframe. Not sure how I can make this automatic for any pdb file.

```{r eval = TRUE, message = FALSE, warning = FALSE}
#Find number of residues total
num.resi.A <- ChainA[nrow(ChainA), 6]
num.resi.B <- ChainB[nrow(ChainB), 6]
num.resi.C <- ChainC[nrow(ChainC), 6]
num.resi.D <- ChainD[nrow(ChainD), 6]

#Initialize new dataframes for the average values per residue. These avg dataframes will only have the relevant columns. I have removed "type", "atom number", "chain", "element", and "element2" since those are all atom-specific and chain is already specified by the DF name. 

ChainA.avg <- data.frame(matrix(ncol = 7))
ChainB.avg <- data.frame(matrix(ncol = 7))
ChainC.avg <- data.frame(matrix(ncol = 7))
ChainD.avg <- data.frame(matrix(ncol = 7))
cols <- c("Residue Number", "Residue AA", "Avg X-Position", 
          "Avg Y-Position", "Avg Z-Position", "Avg Occupancy Factor", 
          "Avg Temperature Factor")
colnames(ChainA.avg) <- cols
colnames(ChainB.avg) <- cols
colnames(ChainC.avg) <- cols
colnames(ChainD.avg) <- cols

##CREATING CHAIN A DF

#Loop: Go through the number of residues per chain (not number of atoms)
for (m in 1:num.resi.A)
{
  #Index out and access the residue matching the loop counter, m. This will contain all the atoms that correspond to that residue
  temp.resi.index <- ChainA[,6] == m
  temp.resi <- ChainA[temp.resi.index,]
  
  #Index out all the x-values from the temp.resi dataframe and take average
  x.pos.vals <- as.numeric(temp.resi[,7])
  x.avg <- mean(x.pos.vals)
  
  #Repeat on y values
  y.pos.vals <- as.numeric(temp.resi[,8])
  y.avg <- mean(y.pos.vals)
  
  #Repeat on z values
  z.pos.vals <- as.numeric(temp.resi[,9])
  z.avg <- mean(z.pos.vals)
  
  #Repeat on occupancy values
  occ.vals <- as.numeric(temp.resi[,10])
  occ.avg <- mean(occ.vals)
  
  #Repeat on temperature factors
  temperature.vals <- as.numeric(temp.resi[,11])
  temperature.avg <- mean(temperature.vals)
  
  #Assign all the values to each of the predefined columns (from the new dataframes initialized outside of the loop)
  ChainA.avg[m,1] <- m               #Residue number
  ChainA.avg[m,2] <- temp.resi[1,4]  #Residue amino acid
  ChainA.avg[m,3] <- x.avg           #X-avg position for all atoms in this residue
  ChainA.avg[m,4] <- y.avg           #Y-avg ''
  ChainA.avg[m,5] <- z.avg           #Z-avg ''
  ChainA.avg[m,6] <- occ.avg         #Occupancy ''
  ChainA.avg[m,7] <- temperature.avg #Temperature ''
}
```

> Here's what the temp.resi looks like that is giving information to extract in the loop

```{r eval = TRUE, warning = FALSE}
print(temp.resi)
```

> Here's what the ChainA.avg dataframe now looks like

```{r eval = TRUE, warning = FALSE}
head(ChainA.avg)
```


> Created ChainB.avg, ChainC.avg, and ChainD.avg dataframes with the same code as that for ChainA. For those code chunks, I have set echo = FALSE. 

```{r eval = TRUE, warning = FALSE, message = FALSE, echo = FALSE}
for (m in 1:num.resi.B)
{
  temp.resi.index <- ChainB[,6] == m
  temp.resi <- ChainB[temp.resi.index,]
  
  x.pos.vals <- as.numeric(temp.resi[,7])
  x.avg <- mean(x.pos.vals)
  
  y.pos.vals <- as.numeric(temp.resi[,8])
  y.avg <- mean(y.pos.vals)
  
  z.pos.vals <- as.numeric(temp.resi[,9])
  z.avg <- mean(z.pos.vals)
  
  occ.vals <- as.numeric(temp.resi[,10])
  occ.avg <- mean(occ.vals)
  
  temperature.vals <- as.numeric(temp.resi[,11])
  temperature.avg <- mean(temperature.vals)
  
  ChainB.avg[m,1] <- m
  ChainB.avg[m,2] <- temp.resi[1,4]
  ChainB.avg[m,3] <- x.avg
  ChainB.avg[m,4] <- y.avg
  ChainB.avg[m,5] <- z.avg
  ChainB.avg[m,6] <- occ.avg
  ChainB.avg[m,7] <- temperature.avg
}

##CREATING CHAIN C DF - same as above
for (m in 1:num.resi.C)
{
  temp.resi.index <- ChainC[,6] == m
  temp.resi <- ChainC[temp.resi.index,]
  
  x.pos.vals <- as.numeric(temp.resi[,7])
  x.avg <- mean(x.pos.vals)
  
  y.pos.vals <- as.numeric(temp.resi[,8])
  y.avg <- mean(y.pos.vals)
  
  z.pos.vals <- as.numeric(temp.resi[,9])
  z.avg <- mean(z.pos.vals)
  
  occ.vals <- as.numeric(temp.resi[,10])
  occ.avg <- mean(occ.vals)
  
  temperature.vals <- as.numeric(temp.resi[,11])
  temperature.avg <- mean(temperature.vals)
  
  ChainC.avg[m,1] <- m
  ChainC.avg[m,2] <- temp.resi[1,4]
  ChainC.avg[m,3] <- x.avg
  ChainC.avg[m,4] <- y.avg
  ChainC.avg[m,5] <- z.avg
  ChainC.avg[m,6] <- occ.avg
  ChainC.avg[m,7] <- temperature.avg
}


##CREATING CHAIN D DF - same as above
for (m in 1:num.resi.D)
{
  temp.resi.index <- ChainD[,6] == m
  temp.resi <- ChainD[temp.resi.index,]
  
  x.pos.vals <- as.numeric(temp.resi[,7])
  x.avg <- mean(x.pos.vals)
  
  y.pos.vals <- as.numeric(temp.resi[,8])
  y.avg <- mean(y.pos.vals)
  
  z.pos.vals <- as.numeric(temp.resi[,9])
  z.avg <- mean(z.pos.vals)
  
  occ.vals <- as.numeric(temp.resi[,10])
  occ.avg <- mean(occ.vals)
  
  temperature.vals <- as.numeric(temp.resi[,11])
  temperature.avg <- mean(temperature.vals)
  
  ChainD.avg[m,1] <- m
  ChainD.avg[m,2] <- temp.resi[1,4]
  ChainD.avg[m,3] <- x.avg
  ChainD.avg[m,4] <- y.avg
  ChainD.avg[m,5] <- z.avg
  ChainD.avg[m,6] <- occ.avg
  ChainD.avg[m,7] <- temperature.avg
}
```

### STEP 5: Remove CDRs and VH regions. 

> These are the regions where the antibody binds its target, so for the mutation studies, avoiding mutations in these regions will ensure that the antibody's function is retained. From the PDB Entry, I know that Chain A and Chain C resi 1-107 is Variable Kappa (VL) and Chain B and Chain D resi 1-118 is Variable Heavy (VL). These new DFs will be the actual regions that we can mutate without altering Ab binding.

```{r eval = TRUE, warning = FALSE, message = FALSE}
ChainA.no.V <- ChainA.avg[108:nrow(ChainA.avg),]
ChainB.no.V <- ChainB.avg[119:nrow(ChainB.avg),]
ChainC.no.V <- ChainC.avg[108:nrow(ChainC.avg),]
ChainD.no.V <- ChainD.avg[119:nrow(ChainD.avg),]
```

### STEP 6: Generate dataframe with the distances between all residues from each other.  

```{r eval = TRUE, warning = FALSE, message = FALSE}

##STEP 6A: Loop through the ChainX.no.V dataframes and acquire distances between residues
#Generate a new df with distances between all of the residues - x,y, and z

###FOR CHAIN A###
temp_ChainA_distDF <- data.frame(matrix(ncol = 3))
ChainA_allDist <- data.frame(matrix(ncol = 3))
cols_distDF <- c("Residue1", "Residue2", "Distance")
colnames(temp_ChainA_distDF) <- cols_distDF
colnames(ChainA_allDist) <- cols_distDF

xvals <- ChainA.no.V$`Avg X-Position`
yvals <- ChainA.no.V$`Avg Y-Position`
zvals <- ChainA.no.V$`Avg Z-Position`

#Formula for distance between 2 points in 3D: sqrt((x2-x1)^2 + (y2-y1)^2 + (z2-z1)^2)

for (i in 1:nrow(ChainA.no.V))
{
  #Store the first residue and its coordinates
  resi1 <- ChainA.no.V[i, 1]
  x1 <- ChainA.no.V[i,3]
  y1 <- ChainA.no.V[i,4]
  z1 <- ChainA.no.V[i,5]
  
  #Generate the distances between the working row and all other rows and store in a newly
  #formatted DF
 for (j in 1:nrow(ChainA.no.V))
  {
    #Store the second residue and its coordinates
    resi2 <- ChainA.no.V[j, 1]
    x2 <- ChainA.no.V[j,3]
    y2 <- ChainA.no.V[j,4]
    z2 <- ChainA.no.V[j,5]
    
    #Calculation
    xdiff <- x2-x1
    ydiff <- y2-y1
    zdiff <- z2-z1
    distance <- sqrt(xdiff^2 + ydiff^2 + zdiff^2)
    
    #Store resi1, resi2, and the distance between the 2 residues in the new DF
    temp_ChainA_distDF[j,1] <- resi1
    temp_ChainA_distDF[j,2] <- resi2
    temp_ChainA_distDF[j,3] <- distance
 }
  ChainA_allDist <- rbind(ChainA_allDist, temp_ChainA_distDF)
}

```

> This is what the ChainA_allDist dataframe looks like. It contains distances between each of the points to all the other points. 

```{ r eval = TRUE, warning = FALSE}
head(ChainA_allDist)
```

> Created ChainB_allDist, ChainC_allDist, and ChainD_allDist dataframes with the same code as that for ChainA. For those code chunks, I have set echo = FALSE. 

```{r eval = TRUE, warning = FALSE, message = FALSE, echo = FALSE}
###FOR CHAIN B###
temp_ChainB_distDF <- data.frame(matrix(ncol = 3))
ChainB_allDist <- data.frame(matrix(ncol = 3))
colnames(temp_ChainB_distDF) <- cols_distDF
colnames(ChainB_allDist) <- cols_distDF

xvals <- ChainB.no.V$`Avg X-Position`
yvals <- ChainB.no.V$`Avg Y-Position`
zvals <- ChainB.no.V$`Avg Z-Position`

#Formula for distance between 2 points in 3D: sqrt((x2-x1)^2 + (y2-y1)^2 + (z2-z1)^2)

for (i in 1:nrow(ChainB.no.V))
{
  #Store the first residue and its coordinates
  resi1 <- ChainB.no.V[i, 1]
  x1 <- ChainB.no.V[i,3]
  y1 <- ChainB.no.V[i,4]
  z1 <- ChainB.no.V[i,5]
  
  #Generate the distances between the working row and all other rows and store in a newly
  #formatted DF
 for (j in 1:nrow(ChainB.no.V))
  {
    #Store the second residue and its coordinates
    resi2 <- ChainB.no.V[j, 1]
    x2 <- ChainB.no.V[j,3]
    y2 <- ChainB.no.V[j,4]
    z2 <- ChainB.no.V[j,5]
    
    #Calculation
    xdiff <- x2-x1
    ydiff <- y2-y1
    zdiff <- z2-z1
    distance <- sqrt(xdiff^2 + ydiff^2 + zdiff^2)
    
    #Store resi1, resi2, and the distance between the 2 residues in the new DF
    temp_ChainB_distDF[j,1] <- resi1
    temp_ChainB_distDF[j,2] <- resi2
    temp_ChainB_distDF[j,3] <- distance
 }
  ChainB_allDist <- rbind(ChainB_allDist, temp_ChainB_distDF)
}

###FOR CHAIN C###
temp_ChainC_distDF <- data.frame(matrix(ncol = 3))
ChainC_allDist <- data.frame(matrix(ncol = 3))
colnames(temp_ChainC_distDF) <- cols_distDF
colnames(ChainC_allDist) <- cols_distDF

xvals <- ChainC.no.V$`Avg X-Position`
yvals <- ChainC.no.V$`Avg Y-Position`
zvals <- ChainC.no.V$`Avg Z-Position`

#Formula for distance between 2 points in 3D: sqrt((x2-x1)^2 + (y2-y1)^2 + (z2-z1)^2)

for (i in 1:nrow(ChainC.no.V))
{
  #Store the first residue and its coordinates
  resi1 <- ChainC.no.V[i, 1]
  x1 <- ChainC.no.V[i,3]
  y1 <- ChainC.no.V[i,4]
  z1 <- ChainC.no.V[i,5]
  
  #Generate the distances between the working row and all other rows and store in a newly
  #formatted DF
 for (j in 1:nrow(ChainC.no.V))
  {
    #Store the second residue and its coordinates
    resi2 <- ChainC.no.V[j, 1]
    x2 <- ChainC.no.V[j,3]
    y2 <- ChainC.no.V[j,4]
    z2 <- ChainC.no.V[j,5]
    
    #Calculation
    xdiff <- x2-x1
    ydiff <- y2-y1
    zdiff <- z2-z1
    distance <- sqrt(xdiff^2 + ydiff^2 + zdiff^2)
    
    #Store resi1, resi2, and the distance between the 2 residues in the new DF
    temp_ChainC_distDF[j,1] <- resi1
    temp_ChainC_distDF[j,2] <- resi2
    temp_ChainC_distDF[j,3] <- distance
 }
  ChainC_allDist <- rbind(ChainC_allDist, temp_ChainC_distDF)
}


###FOR CHAIN D###
temp_ChainD_distDF <- data.frame(matrix(ncol = 3))
ChainD_allDist <- data.frame(matrix(ncol = 3))
colnames(temp_ChainD_distDF) <- cols_distDF
colnames(ChainD_allDist) <- cols_distDF

xvals <- ChainD.no.V$`Avg X-Position`
yvals <- ChainD.no.V$`Avg Y-Position`
zvals <- ChainD.no.V$`Avg Z-Position`

#Formula for distance between 2 points in 3D: sqrt((x2-x1)^2 + (y2-y1)^2 + (z2-z1)^2)

for (i in 1:nrow(ChainD.no.V))
{
  #Store the first residue and its coordinates
  resi1 <- ChainD.no.V[i, 1]
  x1 <- ChainD.no.V[i,3]
  y1 <- ChainD.no.V[i,4]
  z1 <- ChainD.no.V[i,5]
  
  #Generate the distances between the working row and all other rows and store in a newly
  #formatted DF
 for (j in 1:nrow(ChainD.no.V))
  {
    #Store the second residue and its coordinates
    resi2 <- ChainD.no.V[j, 1]
    x2 <- ChainD.no.V[j,3]
    y2 <- ChainD.no.V[j,4]
    z2 <- ChainD.no.V[j,5]
    
    #Calculation
    xdiff <- x2-x1
    ydiff <- y2-y1
    zdiff <- z2-z1
    distance <- sqrt(xdiff^2 + ydiff^2 + zdiff^2)
    
    #Store resi1, resi2, and the distance between the 2 residues in the new DF
    temp_ChainD_distDF[j,1] <- resi1
    temp_ChainD_distDF[j,2] <- resi2
    temp_ChainD_distDF[j,3] <- distance
 }
  ChainD_allDist <- rbind(ChainD_allDist, temp_ChainD_distDF)
}
```

### STEP 7: Access the new residues that we could mutate to form disulfide bonds 

> Find residue pairs that are 3.5-7 Angstroms apart (between the atoms of the peptide bonds) (which may form a disulfide bond if mutated to cysteines). Because I have chosen to look at residues instead of atoms, I am constraining this distance to be 5 to 5.5 Angstroms apart (arbitrary guess). Reference: https://www.nature.com/articles/s41598-020-67230-z#:~:text=3%2C%20the%20distances%20between%20the,acid%20pairs%20is%20significantly%20reduced

```{r eval = TRUE}
#Store the indices where the distance fits the criteria of being 3.5-7 A apart
indexA <- ChainA_allDist[,3] > 5 & ChainA_allDist[,3] < 5.5
indexB <- ChainB_allDist[,3] > 5 & ChainB_allDist[,3] < 5.5
indexC <- ChainC_allDist[,3] > 5 & ChainC_allDist[,3] < 5.5
indexD <- ChainD_allDist[,3] > 5 & ChainD_allDist[,3] < 5.5

A_true <- ChainA_allDist[indexA,]
B_true <- ChainB_allDist[indexB,]
C_true <- ChainC_allDist[indexC,]
D_true <- ChainD_allDist[indexD,]

#In the original ChainA, ChainB, ChainC, ChainD datasets, some residues were missing. In this final distance dataframe, we now have a bunch of NA that arose from the indexing to residues that weren't present. I am removing all NA element rows here:

A_true <- na.omit(A_true)
B_true <- na.omit(B_true)
C_true <- na.omit(C_true)
D_true <- na.omit(D_true)
```

### STEP 8: See if I can access existing disulfide bonds with the same constraints I used to propose mutation sites. I will use the disulfide bonds characterized from the PDB entry and check if they match the constraint in this dataset (using the ChainX.avg dataframes). Since Chain A and C are the same entity and B and D are the same entity, I will just check A and B for the bonds. 
> From the PDB entry for 1IGT, there are the following disulfide bonds:
> A23-A88; A134-A194; A214-B133; B22-96; B145-200; B262-322; B368-426; B228-D228.

```{r eval = TRUE}
#A23-A88
A23 <- ChainA.avg[23,]
A88 <- ChainA.avg[88,]
dis1 <- sqrt((A23[3]-A88[3])^2 + (A23[4]-A88[4])^2 + (A23[5]-A88[5])^2)

#A134-A194
A134 <- ChainA.avg[134,]
A194 <- ChainA.avg[194,]
dis2 <- sqrt((A134[3]-A194[3])^2 + (A134[4]-A194[4])^2 + (A134[5]-A194[5])^2)

#A214-B133
A214 <- ChainA.avg[214,]
B133 <- ChainB.avg[133,]
dis3 <- sqrt((A214[3]-B133[3])^2 + (A214[4]-B133[4])^2 + (A214[5]-B133[5])^2)

#B22-B96
B22 <- ChainB.avg[22,]
B96 <- ChainB.avg[96,]
dis4 <- sqrt((B22[3]-B96[3])^2 + (B22[3]-B96[3])^2 + (B22[3]-B96[3])^2)

#B145-B200
B145 <- ChainB.avg[145,]
B200 <- ChainB.avg[200,]
dis5 <- sqrt((B145[3]-B200[3])^2 + (B145[3]-B200[3])^2 + (B145[3]-B200[3])^2)

#B262-B322
B262 <- ChainB.avg[262,]
B322 <- ChainB.avg[322,]
dis6 <- sqrt((B262[3]-B322[3])^2 + (B262[4]-B322[4])^2 + (B262[5]-B322[5])^2)

#B368-B426
B368 <- ChainB.avg[368,]
B426 <- ChainB.avg[426,]
dis7 <- sqrt((B368[3]-B426[3])^2 + (B368[4]-B426[4])^2 + (B368[5]-B426[5])^2)

#B228-D228
B228 <- ChainB.avg[228,]
D228 <- ChainD.avg[228,]
dis8 <- sqrt((B228[3]-D228[3])^2 + (B228[4]-D228[4])^2 + (B228[5]-D228[5])^2)

dis.validation <- c(dis1, dis2, dis3, dis4, dis5, dis6, dis7, dis8)
```

> Here's the contents of dis.validation, which contains the distances between the residues' average positions that are involved in true disulfide bonds present in the protein:

```{r eval = TRUE}
print(dis.validation)
```

> The vector dis.validation contains some of the true disulfide bond distances calculated with the same method as I calculated for the mutation study. Some of these values are within a close range, but some are much larger. Notably, between chains the distance is greater than it is within a chain. I think I need to look into additional constraints to make my predictors more accurate, and I have learned that distance is not enough of a parameter to consider. 

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.

> The residue pairs that could be useful to form a disulfide bond are now listed in the X_true dataframe!! Here are the headers of those dataframes

```{r eval = TRUE}
head(A_true)
head(B_true)
head(C_true)
head(D_true)
```

> I know it might seem like very little results, but my overall question was to try to find residues that I could mutate in a protein to generate disulfide bonds in the structure. The headers of the dataframes shown in the previous R chunk show the finalized residue pairs that could be used to introduce the mutations. I am very happy with these results and how the code turned out because this cuts out lots of time and guesswork in determining suitable residues for mutation. I don't have any plots to show since the objective of the project doesn't have visual deliverables.
> For future studies, I would like to do the mutations on an atom basis instead of a residue basis as I think that will help to screen out mutations that may not work that might be in the current final dataset. 
