---
title: "Health Disparities Between Rural and Urban Populations in Philadelphia Region"
author: "Joyce Wang"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***
### Overview
Rural population are generally less well off than urban population with respect to health due to lack of access and increased risky health behaviors among others. I'm interested to see if these health disparities exist in the greater Philadelphia region and what factors are the greatest predictors of adverse health events. I will be applying statistical and machine learning analysis to the Southeastern Pennsylvania Household Survey results from 2015 to evaluate the difference between rural and urban populations. You can get access to my Github repository [here](https://github.com/joycehelenwang/BMIN503_Final_Project).

I conceptualized my project after speaking with three faculty members: Andria Johnson, Rebecca Hubbard, and Ariana Chao. Andria Johnson is a Professor in the Department of History and Sociology of Science and recommended that I look at health disparities in the local Philadelphia region. Rebecca Hubbard is an Associate Professor of Biostatistics in the Department of Biostatistics, Epidemiology and Informatics and suggested using machine learning techniques to create a model to predict health outcomes and determine source variability. Ariana Chao is an Assistant Professor of Nursing and reccomended some datasets and discussed the impact of mental health on BMI and other health outcomes as a possibile route.

### Introduction 
While there exists many difference between rural and urban populations, there's been a growing interest in health disparities between the two groups. The difference in demographics of the groups along with less access to healthcare providers in rural regions leads to health disparities. In an era where healthcare issues are at the forefront of political agendas, it's important to create policies that benefit all Americans regardless of their location. For this project in particular, looking at the greater Philadelphia region will provide insights locally. Through this analysis I hope to elucidate differences in health outcomes between rural and suburban groups and also what factors are the greatest predictors of adverse health outcomes to inform healthcare policies in the region.

This problem interdiciplinary as it spans several fields. There is a large data science aspect since the analysis is performed through coding in R. The problem itself draws insights from epidemiology, public health and policy, and sociology. Since healthcare policies are so influential, it's vital that we have adequate data and effective analysis in order to inform these decisions. Datasets will need to be properly cleaned and transformed, making note of missing values. Additionally, being able to apply context to the obtained results and drawing the correct conclusions is crucial in best representing the population.

### Methods
In the first paragraph, describe the data used and general methodological approach. Subsequently, incorporate full R code necessary to retrieve and clean data, and perform analysis. Be sure to include a description of code so that others (including your future self) can understand what you are doing and why. 

The data I am analyzing is from the 2014-2015 Southeastern Pennsylvania Household Health Survey with results from adults and children combined. After loading in the data, it'll be cleaned accordingly. First and foremost, the entires will be classified based on rural or urban according to the zip code. Afterwards, relevant variables will be cleaned so that the responses are renamed to useful information.

``` {r, eval = TRUE}
library(haven)

# read in raw data
raw.data <- read_sav(url("https://raw.githubusercontent.com/joycehelenwang/BMIN503_Final_Project/master/HS15COM1b.sav"))

# define urban and rural



```


### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
