---
title: "Drug Overdose Trends and Forecasts in the United States"
author: "Katherine Slovik"
output: 
  html_document:
    toc: false 
    depth: 3 
    theme: paper 
    highlight: tango
---
```{r set-options, echo=FALSE, cache=FALSE}
options(width = 400)
```  
***

### Overview
The opioid epidemic is a serious public health problem in the United States. The National Vital Statistics System provides data made available by the CDC and U.S. Department of Health regarding national mortality for drug overdoses over the past several years. The goal of this project is to perform statistical analyses to learn about the relationships between variables in the data set, geospatial analyses to understand the critical centers of the epidemic, as well as forcasting to predict future overdose trajectories.

[Github Repository](https://github.com/kslovik/BMIN503_Final_Project)

### Introduction 
Over the past 20 years, the United States has experienced an increase in the amount of deaths that can be attributed to drug overdoses.  In particular, opioids have been at the heart of this problem, including both prescription medications and illegal opioids. Opioids are used to treat pain by interacting with opioid receptors in the body and brain. In addition to pain relief, opioids also elicit a feeling of pleasure.  This combination of pain relief and pleasure provides some explanation for the addictive quality of opioid drugs. Over-prescribing by clinicians and access to illicit drugs have allowed this to become a severe epidemic that has shown little indication of slowing down, that both local and national agencies are attempting to address. 

To understand the nature of the problem of drug overdose mortality, an interdisciplinary approach can provide the most insight.  Firstly this is a public health issue, requiring the knowledge of clinicians, psychiatrists, and public health officials. This is also an economic and geographic issue, as socioeconomic status and location may have a large affect on drug overdose mortality. Using additional data sets from SAMHSA, NIDA and the DEA to integrate with the drug overdose mortality data will provide more insight for correlative analysis and trends regarding the opioid epidemic.

### Methods
The primary data set used for this project was obtained from the publicly available data from the National Vital Statistics System.  The link to the data download is [here](https://catalog.data.gov/dataset/vsrr-provisional-drug-overdose-death-counts-54e35). The data analyzed is titled "VSRR Provisional Drug Overdose Death Counts" and contains provisional monthly counts for drug overdose deaths and total number of deaths (from mortality data) for all 50 states in the United States over the years 2015-2017.  The data also includes information on the specific drug categories determined to be the cause of overdose for 18 states.

The main goal is to analyze the data with some exploratory analysis as well as graphical analysis to visualize data in a meaningful way to help understand the scale and severity of the opioid epidemic in the United States.

First, necessary packages are loaded.
```{r, message=FALSE}
# load necessary packages
library(tidyverse)
library(sf)
library(RColorBrewer)
library(tigris)
library(tmap)
library(tmaptools)
library(grid)
library(leaflet)
```

### Results
Describe your results and include relevant tables, plots, and code/comments used to obtain them. End with a brief conclusion of your findings related to the question you set out to address. You can include references if you'd like, but this is not required.
