---
title: "Drug Treatment Centers and Neighborhood Crime: Evidence from Nine U.S. Cities Over Two Decades"
subtitle: "BMIN5030/EPID6000 Final Project"
format:
  clean-revealjs:
    self-contained: true
author:
  - name: Jamie S. Benson
    degrees:
      - B.A.
    orcid: 0000-0002-0709-4711
    corresponding: true
   # email: jamie.benson@pennmedicine.upenn.edu
    roles:
      - investigation
      - data curation
    affiliation:
      - ref: ggeb
  - name: Nicholas J. Seewald
    degrees:
      - PhD
    orcid: 0000-0002-8367-0522
    corresponding: false
   # email: nicholas.seewald@pennmedicine.upenn.edu
    roles: []
    affiliation:
      - ref: ggeb
  - name: Elizabeth K. Nesoff
    degrees:
      - PhD
      - MPH
    orcid: 0000-0001-8562-8646
    corresponding: false
   # email: elizabeth.nesoff@pennmedicine.upenn.edu
    roles: []
    affiliation:
      - ref: ggeb
affiliations:
  - id: ggeb
    name: Perelman School of Medicine, Department of
      Biostatistics, Epidemiology, and Informatics
    isni: 123012088
date: last-modified
csl: jama.csl
bibliography: references.bib
---

```{r}
#| label: load-env
#| echo: false
#| 
library(quartabs)
library(targets)
library(cols4all)
library(tidyverse)
library(tmap)
library(flextable)
library(kableExtra)
library(dplyr)
library(broom)
library(panelView)
library(ggplot2)
library(purrr)
library(visNetwork)
library(ggeffects)
library(stringr)
library(widgetframe)
library(ggthemes)
library(sf)
library(units)
library(fixest)
library(glmmTMB)
library(modelsummary)

units::install_unit("persons")
units::install_unit("crimes")

tar_load(c(
  buffers,
  cities,
  clinics_sample,
  clinics_open,
  clinics_close,
  bg_poly,
  acs_sample,
  #crime_geo,
  cities_sample,
  included_cities,
  att_dist,
  att_buffer,
  att_dist_city,
  att_buffer_city,
  dist_forest,
  dist_forest_city,
  buffer_forest,
  buffer_forest_city,
  clinic_year_summary,
  all_layer_map
  #plot_clinic_year,
  #plot_clinic_year_matrix
))

cb_palette <- c(
  "#E69F00", # orange
  "#56B4E9", # sky blue
  "#009E73", # bluish green
  "#F0E442", # yellow
  "#0072B2", # blue
  "#D55E00", # vermillion
  "#CC79A7", # reddish purple
  "#999999", # grey
  
  # extended set (similar luminance/saturation)
  "#117733", # dark green
  "#882255", # wine
  "#44AA99", # teal
  "#332288", # indigo
  "#DDCC77", # sand
  "#AA4499", # magenta
  "#88CCEE", # light blue
  "#661100"  # dark brown
)

tar_source(list.files("code", full.names = TRUE)) # Loads helper functions

```

# Project Motivation {background-color="#40666e"}

## Motivation {#sec-background}
### Background
The opioid epidemic in the U.S. is an unceasing public health [crisis]{.alert}

:::{.fragment}
- Despite encouraging post-pandemic trends (13% annual decline in fatalities), it remains worse than pre-COVID[@alexanderEpidemicMidstPandemic2020; @florenceEconomicBurdenPrescription2016; @gomesTrendsOpioidToxicity2023]
:::
:::{.fragment}
- Improvements in overdose fatalities have mirrored decreases in violent crime rates across the country, dropping 13% over-all in 2024[@CrimeTrendsUS2024]
:::
:::{.fragment}
- Access to opioid use disorder (OUD) treatment services is essential, and has been shown to decrease deaths from overdose and crime recidivism, and improve quality of life[@cantorPatternsGeographicDistribution2022; @amiriDisparitiesAccessOpioid2021]
:::
:::{.fragment}
- Drug treatment centers (DTCs) take many forms, and provide much of this care in American communities
:::

## Motivation
### Background
Social, structural, and political barriers remain opposing expansion of these essential services

:::{.fragment}
- Often, "Not in My Backyard" (NIMBY) arguments are an impediment[@rouhaniNIMBYismHarmReduction2022]
:::
:::{.fragment}
- County-level analyses have shown an [inverse relationship]{.alert} to crime and all-cause mortality with facility openings[@bondurantSubstanceAbuseTreatment]
:::
:::{.fragment}
- Tract and bock-group level models in New Jersey have found a slight cross-sectional increase in crime in proximity to facilities[@furr-holdenNotMyBack2016]
    - This is hypothesized to be a function of population and activity density, and was a uniform effect across DTCs, grocery stores, and liquor outlets
:::

## Motivation
:::{.fragment}
- Evidence for, or indeed against, "NIMBY" arguments against DTC proliferation at the local level is lacking
:::
:::{.fragment .center}
### Project Aims
- This project aims to explore the [spatial and temporal]{.alert} relationship between Drug Treatment Center operation and neighborhood crime density over a [large timespan]{.alert} in major U.S. metropolitan areas
:::

# Methods {background-color="#40666e"}

## Methods
### Analytic Pipeline
```{r}
#| label: pipeline-viz
#| echo: false
#| warnings: false

# Define a representative subset of targets
# subset_targets <- c(
#   "cities",
#   "counties_geo",
#   "acs_sample",
#   "clinics_keyed",
#   "clinics_geo",
#   "clinics_sample",
#   "crime_geo",
#   "clinic_year_summary"
# )

# # Build minimal dependency graph
# g <- tar_network(targets_only = FALSE, exclude = NULL)

# visNetwork(g$nodes, g$edges) %>%
#   visNodes(shape = "box") %>%
#   visLayout(randomSeed = 12)

# visNetwork(g$nodes, g$edges, width = "100%", height = 700) %>%
#   visNodes() %>%
#   visOptions(highlightNearest = TRUE) %>%
#   visInteraction(
#     navigationButtons = TRUE,
#     dragNodes = FALSE,
#     dragView = TRUE,
#     zoomView = TRUE
#   ) %>%
#   visEdges(arrows = 'from')

tar_visnetwork(physics = TRUE, zoom_speed = .75, reporter = "silent",  targets_only = TRUE, exclude = "keys_crosswalk")

```

<!-- ## Data Sources ```{r, results = "asis", echo = FALSE}
  cat(c("```{mermaid}", targets::tar_mermaid(reporter = "silent"),  targets_only = TRUE, exclude = "keys_crosswalk", "```"), sep = "\n")
  ``` -->

## Methods
### Data Sources
#### Drug Treatment Centers (DTC) 

:::{.fragment}
- National Survey of Substance Abuse Treatment Services (N-SSATS) is conducted by the Substance Abuse and Mental Health Services Administration (SAMSHA) annually
:::
:::{.fragment}
- Facilities provide information on services, licensure, funding structure, address, and contact information
:::
:::{.fragment}
- We obtained N-SSATS directory listings in PDF format for the years 2005-2021, which were ingested using optical character recognition, and merged with CSV data covering 2022-2024
:::
:::{.fragment}
- Clinics were geocoded and linked across years using nearest neighbor and Jaro-Winkler/SOUNDEX similarity
:::

## Methods
### Data Sources
#### Violent Crime Data

:::{.fragment}
- Accessed open data portals with geolocated crime incident records for large US cities (populations >300,000)
:::
:::{.fragment}
- Included records with date of incident, latitude and longitude, and crime description data 
:::
:::{.fragment}
- As drug-related and petty crimes are subject to reporting biases, we focused on "Part I" violent crimes: murder, non-negligent manslaughter (homicide), rape, robbery, or aggravated assault
:::
:::{.fragment}
- Subset by Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) standards: against persons, and against property
:::

## Methods
### Data Sources
#### Neighborhood Data

:::{.fragment}
- Community demographics for each block group in each city were taken from five-year American Community Survey (ACS) estimates using TidyCensus
:::
:::{.fragment}
- Shapefiles and source tables were retrieved from the IPUMS National Historical Geographic Information System (NHGIS).
:::
:::{.fragment}
- Population-weighted geographic centroids representing each block group similarly obtained from NHGIS
:::
:::{.fragment}
- Variables including population totals and median household income in $10,000s, educational attainment, property values, race / ethnicity / sex distribution, and household composition
:::

## Methods
### Sample Definition

:::{.fragment}
- Clinics were included for analysis if they had at least 5 years of operation observed, provided addiction treatment services, and were geocoded with at least 95% street address match confidence
:::
:::{.fragment}
- In each year, crimes within a maximum of 0.5mi Euclidean distance were considered for buffered density and distance calculations, as this is widely considered the upper bound of *"walking distance"* in an urban environment
:::
:::{.fragment}
- For clinics which opened in a given year, the point-locations of future clinics (which had not yet opened) were used as control units
:::

## Methods
### Exposures

:::{.fragment}
- The unit of analysis is the point location of a DTC, which is tracked over time before it opens, during operation, and (if relevant) after it closes
:::
:::{.fragment}
- Clinics which responded to NSSATS for listing in a given year were considered open
:::
:::{.fragment}
- Clinics were considered closed after the final year in which they responded to NSSATS
:::
:::{.fragment}
- As DTCs may have multiple listed offices within the same block address, DTCs which geocoded to within 100ft of one another were considered as one "treatment" unit
:::

## Methods
### Outcome Measures

:::{.fragment}
- Primary outcome was the density of violent crime in the immediate area surrounding a DTC before, during, and after its operation in a neighborhood
:::
:::{.fragment}
- Crime density was measured as: 
  1) Count of violent crimes within concentric 200ft buffers, up to 0.5mi
  2) Median Euclidean distance from crime location to each DTC
:::
:::{.fragment}
- Separate estimands were included for each crime subtype (over-all, against persons, against property)
:::
:::{.fragment}
- SAMSHA updates the directory annually, temporal resolution is 1yr
:::
:::{.fragment}
- We allowed a 1-year wash-in period until effects were estimated
:::

## Methods
### Social Measures

:::{.fragment}
Social Deprivation
:::
:::{.fragment}
- Neighborhood deprivation operationalized using the *Neighborhood Deprivation Score*, calculated as: 
$$\text{NDS} = \frac{\left(\frac{c}{10} + \frac{d}{10}\right) - \left(\frac{a}{10} + \frac{b}{10}\right)}{4}$$
:::
:::{.fragment}
- Range -5:5, where: 
  (a) adults ≥25 years with a college degree
  (b) owner-occupied housing
  (c) households with incomes below the federal poverty threshold
  (d) female-headed households with children
:::

## Methods
### Social Measures

:::{.fragment}
Residential Segregation
:::
:::{.fragment}
- Assessed using *Index of Concentration at the Extremes* (ICE), calculated as: 

$$\text{ICE} = \frac{\text{Non-Hispanic White}-\text{Non-Hispanic Black}}{\text{Block Group Population}} $$
:::
:::{.fragment}
- Range: -1 is 100% Black | 0 is 50% Black, 50% White | 1 is 100% White
:::

## Methods
### Modeling Approach

:::{.fragment}
- Within each year, treated units (newly opened or closed DTCs) were compared to not-yet-treated units using an event-study model.
:::
:::{.fragment}
- Paired t-tests assessed crime density before and after openings and closures.
:::
:::{.fragment}
- Initial GLMMs predicted crime density and distance following opening and closure events.
:::
:::{.fragment}
- A difference-in-differences model with staggered treatment timing and robust standard errors clustered at the clinic and city levels was estimated using the **did** package.
:::
:::{.fragment}
- Group-time-averaged treatment effects are reported for clarity.
:::

# Results {background-color="#40666e"}

## Results
### Clinics Identified

:::{.fragment}
In total, we identified ```r length(unique(clinics_sample$clinic_id))``` clinics in ```r length(unique(included_cities))``` cities, across ```r max(clinics_sample$year)-min(clinics_sample$year)``` years:
:::
:::: {.columns}
::: {.column width="40%"}
:::{.fragment}
| City           | State                  |
|----------------|------------------------|
| Atlanta        | Georgia                |
| Baltimore      | Maryland               |
| Chicago        | Illinois               |
| Denver         | Colorado               |
| Los Angeles    | California             |
| New York       | New York               |
| Philadelphia   | Pennsylvania           |
| San Francisco  | California             |
| Seattle        | Washington             |
:::
:::
::: {.column width="60%"}
:::{.fragment}
```{r}
clinic_year_summary[1] %>%
  as.data.frame() %>%
  pivot_longer(cols=c(counts_compare.n,counts_compare.official_count), names_to = "Count",values_to = "value") %>%
  mutate(Count = factor(Count, levels = c("counts_compare.n","counts_compare.official_count"), labels = c("Respondent Facilities","Official Count"))) %>%
  rename(Year = counts_compare.year) %>%
  ggplot(aes(x=Year, y = value, group = Count, color = Count)) +
    geom_point() +
    geom_line() +
    annotate("rect", xmin = 2021, xmax = 2025, ymin = 0, ymax = 19000, 
           alpha = .2) +
    ylim(0,19000) +
    theme_minimal() +
    scale_color_brewer(palette = "Dark2") +
    labs(
      x = "Year",
      y = "Clinics (n)",
      title = "Count of Clinics by Year\nSAMSHA Certified vs NSSATS Respondents"
    )
```
:::
:::
::::

## Results
### DTC & Crime Distribution: Philadelphia & Seattle {.center}

:::: {.columns}

::: {.column width="60%"}
:::{.fragment}
```{r}
#| label: philly-map
#| fig-height: 10
#| 
all_layer_map[[12]]
```
:::
:::
::: {.column width="40%"}
:::{.fragment}
```{r}
#| label: seattle-map
#| fig-height: 15
#| 
all_layer_map[[15]]
```
:::
:::

::::

## Results {visibility="hidden"}
### Clinic Treatment Status

:::{.fragment}
```{r}
#| label: panel-view
#| fig-height: 5
#| 

panelview(violent_count ~ open, # Using the treatment var which allows treatment switching after initial assignment
          data = clinics_open %>% 
            filter(buffer==min(set_units(buffers,"m"))), 
          index = c("group","year"), 
          by.timing = TRUE,
          main = "Treatment Status - Clinic Opening\nAll Clinics",
          xlab = "Year", ylab = "Clinic ID")
```
:::

## Results
### Unadjusted Differences: Clinic Opening

::: {.panel-tabset}

#### Crimes Against Persons
```{r}
clinics_open %>%
  st_drop_geometry() %>%
  filter(city == "Philadelphia") %>%
  mutate(period = factor(period, levels = c(0,1), labels = c("Pre-Opening", "Post-Opening"))) %>%
  ggplot(aes(x=person_median_distance, group = period, fill = period)) +
  geom_density(alpha=.8) +
  scale_fill_brewer(palette = 10) +
  theme_minimal() +
  labs(
    x = "Median Distance to Crime",
    y = "Density",
    title = "Median Distance: DTC to Crimes Against Persons"
  )

```

#### Crimes Against Property
```{r}

clinics_open %>%
  st_drop_geometry() %>%
  filter(city == "Philadelphia") %>%
  mutate(period = factor(period, levels = c(0,1), labels = c("Pre-Opening", "Post-Opening"))) %>%
  ggplot(aes(x=property_median_distance, group = period, fill = period)) +
  geom_density(alpha=.8) +
  scale_fill_brewer(palette = 10) +
  theme_minimal() +
  labs(
    x = "Median Distance to Crime",
    y = "Density",
    title = "Median Distance: DTC to Crime Against Property"
  )

```

:::

## Results
### Unadjusted Differences: Clinic Closure
::: {.panel-tabset}

#### Crimes Against Persons
```{r}
clinics_close %>%
  st_drop_geometry() %>%
  filter(city == "Philadelphia") %>%
  mutate(period = factor(period_closure, levels = c(0,1), labels = c("Pre-Closure", "Post-Closure"))) %>%
  ggplot(aes(x=person_median_distance, group = period, fill = period)) +
  geom_density(alpha=.8) +
  scale_fill_brewer(palette = 10) +
  theme_minimal() +
  labs(
    x = "Median Distance to Crime",
    y = "Density",
    title = "Median Distance: DTC to Crimes Against Persons"
  )

```

#### Crimes Against Property
```{r}

clinics_close %>%
  st_drop_geometry() %>%
  filter(city == "Philadelphia") %>%
  mutate(period = factor(period_closure, levels = c(0,1), labels = c("Pre-Closure", "Post-Closure"))) %>%
  ggplot(aes(x=property_median_distance, group = period, fill = period)) +
  geom_density(alpha=.8) +
  scale_fill_brewer(palette = 10) +
  theme_minimal() +
  labs(
    x = "Median Distance to Crime",
    y = "Density",
    title = "Median Distance: DTC to Crime Against Property"
  )

```
:::

## Results
### Unadjusted Differences: T-Tests & Parallel Trends

::: {.panel-tabset}

#### Clinic Opening
```{r}

clinics_open %>%
  filter(time_point %in% seq(-3,3)) %>%
  st_drop_geometry() %>%
  group_by(buffer) %>%
  do(tidy(t.test(units::drop_units(violent_count/area) ~ period, data = .))) %>%
  ggplot() +
  geom_line(aes(x=buffer,y=estimate*-1), color=cb_palette[1]) +
  geom_ribbon(aes(x=buffer,ymin=conf.low*-1,ymax=conf.high*-1), alpha = 0.3, fill=cb_palette[1]) +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  ylab("Violent Crime Count Change (t-test)") +
  xlab("Buffer Size") +
  labs(
    title = "T-test: Unadjusted Violent Crime Counts - Opening",
    subtitle = "2004-2024, Offers Treatment, Geolocation Score >95%, 5+ Years of Clinic Operation",
    caption = "Data: U.S. Census Bureau, NHGIS, OpenStreetMap, SAMSHA"
  ) +
  theme_clean()
```

#### Clinic Closure
```{r}
clinics_close %>%
  filter(time_point_closure %in% seq(-3,3)) %>%
  st_drop_geometry() %>%
  group_by(buffer) %>%
  do(tidy(t.test(units::drop_units(violent_count/area) ~ period_closure, data = .))) %>%
  ggplot() +
  geom_line(aes(x=buffer,y=estimate*-1), color=cb_palette[1]) +
  geom_ribbon(aes(x=buffer,ymin=conf.low*-1,ymax=conf.high*-1), alpha = 0.3, fill=cb_palette[1]) +
  geom_hline(aes(yintercept = 0), linetype = "dashed") +
  ylab("Violent Crime Count Change (t-test)") +
  xlab("Buffer Size") +
  labs(
    title = "T-test: Unadjusted Violent Crime Counts - Closure",
    subtitle = "2004-2024, Offers Treatment, Geolocation Score >95%, 5+ Years of Clinic Operation",
    caption = "Data: U.S. Census Bureau, NHGIS, OpenStreetMap, SAMSHA"
  ) +
  theme_clean()

```
:::

## Results {visibility="hidden" .scrollable}
### OLS Model Results - **Opening**
```{r}
#| echo: false

df <- clinics_open %>%
  mutate(
    crime_density = drop_units(violent_count / area),
    buffer = drop_units(buffer)
  ) %>%
  select(
    group,
    city,
    buffer,
    violent_count,
    property_count,
    crime_density,
    person_count,
    area,
    violent_median_distance,
    person_median_distance,
    property_median_distance,
    tot_pop_estimate,
    ice,
    nds,
    year,
    time_point,
    period
  ) %>%
  na.omit() %>%
  drop_units()

# Dictionary of regression coefficients for plotting / tables
dict = c("ice"   = "ICE",
"nds" = "NDS",
"ice:nds" = "ICExNDS",
"bufer" = "Buffer Radius",
"i(time_point, ref = 0, keep = -3:3)-3" = "Open -3",
"i(time_point, ref = 0, keep = -3:3)-2" = "Open -2",
"i(time_point, ref = 0, keep = -3:3)-1" = "Open -1",
"i(time_point, ref = 0, keep = -3:3)1" = "Open +1",
"i(time_point, ref = 0, keep = -3:3)2" = "Open +2",
"i(time_point, ref = 0, keep = -3:3)3" = "Open +3")

m_nb_person_open <- glmmTMB(
  person_count ~ 
     i(time_point, ref=0, keep = -3:3) + ice*nds +
    (1 | city),
  offset = log(area),
  family = gaussian(),
  data = df %>% filter(buffer == 548.64)
)

m_nb_property_open <- glmmTMB(
  property_count ~ 
     i(time_point, ref=0, keep = -3:3) + ice*nds +
    (1 | city),
  offset = log(area),
  family = gaussian(),
  data = df %>% filter(buffer == 548.64)
)

models <- list(
  Property = m_nb_property_open,
  Person   = m_nb_person_open
)
```

::: {.panel-tabset}
#### Table: Density
```{r}
modelsummary(
    setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict, stars = TRUE, statistic = 'conf.int',
  gof_map = NA
  )
```

#### Fig: Density
```{r}
modelplot(setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict) +
    labs(x = 'Coef: Crime Density', 
         y = 'Term',
         title = 'Negative Binomial GLMM\nPre/Post Opening vs Crime Density') + 
          geom_vline(xintercept = 0, color = 'orange', alpha = 0.6) +
    scale_color_brewer(palette = "Dark2")

```
#### Table: Distance
```{r}
#| echo: false
dist_nb_person_open <-  glmmTMB(
  person_median_distance ~ 
    i(time_point, ref = 0, keep =-3:3) + ice*nds +
  (1 | city),
  data = df %>% filter(buffer == min(buffers)),
  family = gaussian()
)

dist_nb_property_open <-  glmmTMB(
  property_median_distance ~ 
    i(time_point, ref = 0, keep =-3:3) + ice*nds +
  (1 | city),
  data = df %>% filter(buffer == min(buffers)),
  family = gaussian()
)

models <- list(
  Property = dist_nb_property_open,
  Person   = dist_nb_person_open
)
```

```{r}
modelsummary(
    setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict, stars = TRUE, statistic = 'conf.int',
  gof_map = NA
  )
```
#### Fig: Distance
```{r}
modelplot(setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict) +
    labs(x = 'Coef: Crime Distance', 
         y = 'Term',
         title = 'Negative Binomial GLMM\nPre/Post Opening vs Crime Distance') + 
          geom_vline(xintercept = 0, color = 'orange', alpha = 0.6) +
    scale_color_brewer(palette = "Dark2")

```
:::

## Results {visibility="hidden" .scrollable}
### OLS Model Results - **Closure**
```{r}
#| echo: false

df <- clinics_close %>%
  mutate(
    time_point = time_point_closure,
    crime_density = drop_units(violent_count / area),
    buffer = drop_units(buffer)
  ) %>%
  select(
    group,
    city,
    buffer,
    violent_count,
    property_count,
    crime_density,
    person_count,
    area,
    violent_median_distance,
    person_median_distance,
    property_median_distance,
    tot_pop_estimate,
    ice,
    nds,
    year,
    time_point,
    period
  ) %>%
  na.omit() %>%
  drop_units()

# Dictionary of regression coefficients for plotting / tables
dict = c("ice"   = "ICE",
"nds" = "NDS",
"ice:nds" = "ICExNDS",
"bufer" = "Buffer Radius",
"i(time_point, ref = 0, keep = -3:3)-3" = "Close -3",
"i(time_point, ref = 0, keep = -3:3)-2" = "Close -2",
"i(time_point, ref = 0, keep = -3:3)-1" = "Close -1",
"i(time_point, ref = 0, keep = -3:3)1" = "Close +1",
"i(time_point, ref = 0, keep = -3:3)2" = "Close +2",
"i(time_point, ref = 0, keep = -3:3)3" = "Close +3")

m_nb_person_close <- glmmTMB(
  person_count ~ 
     i(time_point, ref=0, keep = -3:3) + ice*nds +
    (1 | city),
  offset = log(area),
  family = gaussian(),
  data = df %>% filter(buffer == 548.64)
)

m_nb_property_close <- glmmTMB(
  property_count ~ 
     i(time_point, ref=0, keep = -3:3) + ice*nds +
    (1 | city),
  offset = log(area),
  family = gaussian(),
  data = df %>% filter(buffer == 548.64)
)

models <- list(
  Property = m_nb_property_close,
  Person   = m_nb_person_close
)
```

::: {.panel-tabset}
#### Table: Density
```{r}
modelsummary(
    setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict, stars = TRUE, statistic = 'conf.int',
  gof_map = NA
  )
```

#### Fig: Density
```{r}
modelplot(setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict) +
    labs(x = 'Coef: Crime Density', 
         y = 'Term',
         title = 'Negative Binomial GLMM\nPre/Post Closure vs Crime Density') + 
          geom_vline(xintercept = 0, color = 'orange', alpha = 0.6) +
    scale_color_brewer(palette = "Dark2")

```
#### Table: Distance
```{r}
#| echo: false
dist_nb_person_close <-  glmmTMB(
  person_median_distance ~ 
    i(time_point, ref = 0, keep =-3:3) + ice*nds +
  (1 | city),
  data = df %>% filter(buffer == min(buffers)),
  family = gaussian()
)

dist_nb_property_close <-  glmmTMB(
  property_median_distance ~ 
    i(time_point, ref = 0, keep =-3:3) + ice*nds +
  (1 | city),
  data = df %>% filter(buffer == min(buffers)),
  family = gaussian()
)

models <- list(
  Property = dist_nb_property_close,
  Person   = dist_nb_person_close
)
```

```{r}
modelsummary(
    setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict, stars = TRUE, statistic = 'conf.int',
  gof_map = NA
  )
```
#### Fig: Distance
```{r}
modelplot(setNames(models, c("(1: Property)", "(2: Persons)")),
  coef_map = dict) +
    labs(x = 'Coef: Crime Distance', 
         y = 'Term',
         title = 'Negative Binomial GLMM\nPre/Post Closure vs Crime Distance') + 
          geom_vline(xintercept = 0, color = 'orange', alpha = 0.6) +
    scale_color_brewer(palette = "Dark2")

```
:::

## Results
### Difference-in-Difference
::: {.panel-tabset}
#### Distance
```{r}
dist_forest
```

#### Distance - By City
```{r}
dist_forest_city
```

#### Density
```{r}
buffer_forest
```

#### Density - By City
```{r}
buffer_forest_city
```
:::

# Conclusions {background-color="#40666e"}

## Conclusions
### Primary Findings

:::{.fragment}
- Substantial growth in DTC availability in U.S. cities from 2004 to 2024
:::
:::{.fragment}
- Neighborhoods with greater social disadvantage had higher levels of violent crime
:::
:::{.fragment}
- In aggregate, across all cities and years, we found **no significant relationship** between DTC operation and violent crime, crimes against persons, or crimes against property
:::

## Conclusions
### Primary Findings

:::{.fragment}
City-level estimates are more varied: 
:::
:::{.fragment}
- Most cities have no significant relationship by either measure
:::
:::{.fragment}
- NYC and Atlanta showed protective associations with DTC opening
:::
:::{.fragment}
- Denver and Philadelphia showed potential positive associations when measured via Euclidean distance, and negative to null associations when measured with buffered-density
:::

## Conclusions
### Limitations, Future Directions

:::{.fragment}
- Causal inference is dependent upon assumed parallel trends
:::
:::{.fragment}
- Correction for multiple comparisons in confidence interval selection is needed
:::
:::{.fragment}
- Social covariates treated as uniform across time (2005 starting estimates)
:::
:::{.fragment}
- Euclidean distance measurements do not reflect street-network based travel times
:::
:::{.fragment}
- Spatial autocorrelation should be addressed using k-means clustered nearest neighbor crime values
:::

## References
::: {#refs}
:::

<!-- [[Back to top]{.button}](#sec-background) -->
