---
title: "Admin Intensity"
author: "David Roberts"
format: html
editor: visual

project:
  execute-dir: project
---

## Purpose

-   Compare daily pbj journal derived staffing ratios to quarterly reports.
-   Construct a measure which attempts to measure diversion of resources to admin duties.
    -   How can I measure this?
-   Test correlation between case-mix and diversion of resources to admin duties.
-   105319 - Sun Terrace, admin spike in 2019

## Imports

```{r}
library(dplyr)
library(ggplot2)
library(here)
```

## Load Data

```{r}
load(here("data", "interim", "snf_provider_info.Rda"))
pbj <- read.csv(here("data", "interim", "pbj_facility_level.csv"))
```

## Calculate YOY increases in staffing rating

Landed in 4-5 range is a case

```{r}
staffing_next_yr <- select(snf_level_data, c("federal_provider_number", "staffing_rating", "collection_yr", "adjusted_rn_staffing_hours_per_resident_per_day", "adjusted_total_nurse_staffing_hours_per_resident_per_day", "ownership_type", "average_number_of_residents_per_day", "provider_changed_ownership_in_last_12_months", "continuing_care_retirement_community","average_number_of_residents_per_day",
"overall_rating", "health_inspection_rating", "qm_rating"))
staffing_next_yr$collection_yr = staffing_next_yr$collection_yr - 1
snf_level_data <- snf_level_data %>%
  left_join(staffing_next_yr, by=c("federal_provider_number", "collection_yr"))
staffing_yoy <- select(snf_level_data,
    c("federal_provider_number", "staffing_rating.x", "collection_yr",
       "staffing_rating.y","adjusted_rn_staffing_hours_per_resident_per_day.x", "adjusted_total_nurse_staffing_hours_per_resident_per_day.x",
      "adjusted_rn_staffing_hours_per_resident_per_day.y", "adjusted_total_nurse_staffing_hours_per_resident_per_day.y",
      "ownership_type.x", "ownership_type.y", "average_number_of_residents_per_day.x",
      "average_number_of_residents_per_day.y",
      "provider_changed_ownership_in_last_12_months.y", "continuing_care_retirement_community.y",
      "average_number_of_residents_per_day.x", "average_number_of_residents_per_day.y",
      "overall_rating.x", "overall_rating.y", 
      "health_inspection_rating.x", "health_inspection_rating.y", 
      "qm_rating.x", "qm_rating.y")) %>%
  filter(is.na(staffing_rating.y) == FALSE & is.na(staffing_rating.x) == FALSE) %>%
  mutate(overall_star_change=overall_rating.y - overall_rating.x) %>%
  mutate(inspect_star_change=health_inspection_rating.y - health_inspection_rating.x) %>%
  mutate(qm_star_change=qm_rating.y - qm_rating.x) %>%
  mutate(staff_star_change=staffing_rating.y - staffing_rating.x) %>%
  mutate(rn_hr_prpd_change=adjusted_rn_staffing_hours_per_resident_per_day.y - adjusted_rn_staffing_hours_per_resident_per_day.x) %>%
  mutate(total_hr_prpd_change=adjusted_total_nurse_staffing_hours_per_resident_per_day.y - adjusted_total_nurse_staffing_hours_per_resident_per_day.x) %>%
  mutate(status=as.factor(
           ifelse(staff_star_change > 0 & staffing_rating.x < 4 & staffing_rating.y >=4,
          "Case", "Control"))) %>%
  mutate(status_v2=as.factor(
           ifelse(staff_star_change > 1 & staffing_rating.x < 4 & staffing_rating.y >=4,
          "Case", "Control"))) %>%
  mutate(status_v3=as.factor(
           ifelse((staff_star_change > 0 & staffing_rating.x < 4 & staffing_rating.y >=4)
                  & (qm_star_change < 0 | inspect_star_change < 0),
          "Case", "Control")))
```

Assess missingness... To create a clean analysis, probably worth removing observations where an staffing observation is not available. This drops the staffing rating to 1...

Among the cases, it only occurs with +3 and +4 stars...

```{r}
staffing_yoy %>% count(status, staff_star_change)
staffing_yoy %>% count(status_v2, staff_star_change)
staffing_yoy %>% count(status_v3, staff_star_change)
filter(staffing_yoy, is.na(rn_hr_prpd_change)) %>% count(status, staff_star_change)
staffing_yoy.complete <- filter(staffing_yoy, !(is.na(rn_hr_prpd_change) | is.na(total_hr_prpd_change)))
```

Dropped \~3k observations

Pull corresponding PBJ (Payroll Based Journal), also Peanut Butter and Jelly

```{r}
pbj <- pbj %>%
  mutate(year = as.integer(substr(CY_Qtr, 1, 4))) %>%
  mutate(quarter = as.integer(substr(CY_Qtr, 6, 6)))

pbj.q4 <- pbj %>%
  filter(quarter==4)

pbj.q4.merged <- pbj.q4 %>%
  mutate(year = year + 1) %>%
  left_join(pbj.q4, by=c("PROVNUM", "year")) %>%
  filter(!is.na(quarter.y)) %>%
  mutate(admin_change_pct_pt = (AdminIntensity.y - AdminIntensity.x) * 100)
```

55939 obs in pbj.q4.merged 55473 obs in staffing_yoy 54807 obs in final dataframe

Assess reliability of data calculated from pbj detail

```{r}
library(stringr)

staffing_yoy.complete$year = staffing_yoy.complete$collection_yr + 1
pbj.q4.merged <- rename(pbj.q4.merged, federal_provider_number = PROVNUM )
final.df <- staffing_yoy.complete %>%
  inner_join(pbj.q4.merged, by=c("year", "federal_provider_number"))
final.df <- subset(final.df, select=-c(
  collection_yr,
  CY_Qtr.x,
  CY_Qtr.y,
  quarter.x,
  quarter.y
))
final.df$year_factor <- as.factor(final.df$year)
final.df$status <- factor(final.df$status, levels=c("Control", "Case")) # easier to interpet
final.df <- final.df %>%
  mutate(for_profit = ifelse(str_detect(ownership_type.y, 'For profit'), TRUE, FALSE))
```

Visualize

```{r}
plt <- ggplot(final.df, aes(x=year_factor, y=admin_change_pct_pt)) + geom_violin()
plt
```

Analysis: Large staffing star changes, +3 and +4 do appear more likely to be cases. There appears to be a greater portion of cases with positive YOY admin intensity changes.

```{r}
plt <- ggplot(final.df, aes(x=admin_change_pct_pt, y=staff_star_change, color=status)) + geom_point()
plt
```

```{r}
plt <- ggplot(final.df, aes(x=as.factor(staffing_rating.x), y=admin_change_pct_pt)) + geom_violin()
plt
```

```{r}
library(gtsummary)
select(final.df, 
      c(staffing_rating.x, staff_star_change, status, year_factor),
      c(admin_change_pct_pt)
      ) %>% tbl_summary(by=year_factor)
```

Admin intensity reducing the probability of case group One thought... Incorporate drop in quality scores. We're worried about increasing admin intensity in ways that doesn't benefit residents.

```{r}
model <- glm(status ~ admin_change_pct_pt, family=binomial, final.df)
summary(model)
```

```{r}
model <- glm(status ~ admin_change_pct_pt + for_profit:admin_change_pct_pt +  provider_changed_ownership_in_last_12_months.y + continuing_care_retirement_community.y + average_number_of_residents_per_day.x + average_number_of_residents_per_day.y + year_factor, family=binomial, final.df)
summary(model)
```

```{r}
model <- glm(status ~ admin_change_pct_pt + for_profit:admin_change_pct_pt +  provider_changed_ownership_in_last_12_months.y + continuing_care_retirement_community.y, family=binomial, filter(final.df, year %in% c(2018,2019)))
summary(model)
```

```{r}
model <- lm(admin_change_pct_pt ~ 0 + staff_star_change + for_profit + provider_changed_ownership_in_last_12_months.y + continuing_care_retirement_community.y, family=binomial, filter(final.df, year %in% c(2018,2019)))
summary(model)
```

```{r}
model <- lm(admin_change_pct_pt ~ staff_star_change + qm_star_change + inspect_star_change + for_profit + provider_changed_ownership_in_last_12_months.y + continuing_care_retirement_community.y, family=binomial, filter(final.df, year %in% c(2018,2019)))
summary(model)
```

```{r}
linear.model <- lm(staff_star_change ~ admin_change_pct_pt + admin_change_pct_pt * for_profit, final.df)
summary(linear.model)
```

```{r}
model <- glm(status ~ admin_change_pct_pt + for_profit:admin_change_pct_pt +  provider_changed_ownership_in_last_12_months.y, family=binomial, filter(final.df, year_factor==2018))
summary(model)
```

```{r}
model <- glm(top_5_percentile_admin_change ~ status, family=binomial, final.df)
summary(model)
```

```{r}
model <- glm(status_v2 ~ admin_change_pct_pt + for_profit:admin_change_pct_pt +  provider_changed_ownership_in_last_12_months.y + continuing_care_retirement_community.y, family=binomial, final.df)
summary(model)
```

Model star change as a function of admin intensity change

```{r}
linear.model <- lm(staff_star_change ~ 0 + admin_change_pct_pt, filter(final.df, staff_star_change > 0))
summary(linear.model)
```

```{r}
linear.model <- lm(staff_star_change ~ 0 + admin_change_pct_pt + for_profit:admin_change_pct_pt, final.df)
summary(linear.model)
```

What is the impact of an admin intensity spike, on average?

```{r}
summary(final.df$admin_change_pct_pt)
quantile(final.df$admin_change_pct_pt, .95)
quantile(final.df$admin_change_pct_pt, .99)

spike.facilities <- filter(final.df, admin_change_pct_pt > quantile(admin_change_pct_pt, .99)) %>%
    select(federal_provider_number)

final.df.spikes <- inner_join(final.df, spike.facilities, by="federal_provider_number")
```

```{r}
linear.model <- lm(qm_star_change ~ 0 + admin_change_pct_pt, final.df)
summary(linear.model)
```

Model by sequence of staffing rating

```{r}
final.df <- final.df %>%
  mutate(top_5_percentile_admin_change = ifelse(admin_change_pct_pt >= quantile(admin_change_pct_pt, .95), TRUE, FALSE)) %>%
  mutate(top_1_percentile_admin_change = ifelse(admin_change_pct_pt >= quantile(admin_change_pct_pt, .99), TRUE, FALSE)) %>%
  mutate(staffing_rating_seq=paste(staffing_rating.x, staffing_rating.y))
```

```{r}
select(final.df, 
      staffing_rating_seq, year_factor) %>% tbl_summary(by=year_factor)
```

2-5 seq has very few observations...

```{r}
model <- glm(top_5_percentile_admin_change ~ staffing_rating_seq, family=binomial, filter(final.df, year %in% c(2018, 2019)))
summary(model)

model <- glm(top_5_percentile_admin_change ~ staffing_rating_seq, family=binomial, final.df)
summary(model)

model <- glm(top_1_percentile_admin_change ~ staffing_rating_seq + inspect_star_change, family=binomial, final.df)
summary(model)
```

# Analysis

Top 5 more likely to be in cases... Switch levels, case and control.

```{r}
model <- glm(status ~ top_5_percentile_admin_change, family=binomial, final.df)
summary(model)
```
