---
title: "County Level Regressions"
author: "David Roberts"
format: html
editor: visual

project:
  execute-dir: project
---

## Purpose

-   Merge NHC and VBC penetration data
-   Visualize and explore patterns across time
-   Linear modeling at the county level

## Imports

```{r}
library(dplyr)
library(ggplot2)
library(here)
```

## Load Data

```{r}
load(here("data", "interim", "county_vbc_penetration.Rda"))
load(here("data", "interim", "county_level_staffing.Rda"))
county_pene <- merged.subset
rm(merged.subset)
```

```{r}
county_staffing <- rename(
  county_staffing,
  Year = collection_yr
)

merged <- merge(
  county_pene,
  county_staffing,
  by=c("FIPS_Code", "Year"),
  all.x = TRUE
)
merged$VBCPenetration = merged$VBCPenetration * 100

merged$MAPenetration = merged$MAPenetration * 100
merged$ACOPenetration = merged$ACOPenetration * 100

merged$YearFactor = as.factor(merged$Year)
```

# Visualize across time

```{r}
plt <- ggplot(filter(merged, Year == 2017), mapping=aes(x = VBCPenetration, y = mean_cna_reported_hprd)) + geom_point() + xlim(0,100) + ylim(0,6) + ggtitle(label="Reported Nursing Aid Hours Per Resident Per Day By VBC Penetration (2017)")
plt
plt <- ggplot(filter(merged, Year == 2018), mapping=aes(x = VBCPenetration, y = mean_cna_reported_hprd)) + geom_point()  + xlim(0,100) + ylim(0,6) + ggtitle(label="Reported Nursing Aid Hours Per Resident Per Day By VBC Penetration (2018)")
plt
plt <- ggplot(filter(merged, Year == 2019), mapping=aes(x = VBCPenetration, y = mean_cna_reported_hprd)) + geom_point()  + xlim(0,100) + ylim(0,6) + ggtitle(label="Reported Nursing Aid Hours Per Resident Per Day By VBC Penetration (2019)")
plt
plt <- ggplot(filter(merged, Year == 2020), mapping=aes(x = VBCPenetration, y = mean_cna_reported_hprd)) + geom_point()  + xlim(0,100) + ylim(0,6)+ ggtitle(label="Reported Nursing Aid Hours Per Resident Per Day By VBC Penetration (2020)")
plt
plt <- ggplot(filter(merged, Year == 2021), mapping=aes(x = VBCPenetration, y = mean_cna_reported_hprd)) + geom_point() + xlim(0,100) + ylim(0,6) + ggtitle(label="Reported Nursing Aid Hours Per Resident Per Day By VBC Penetration (2021)")
plt
```

How to explain the spike in VBC penetration in 2019? Something appears to be wrong with the data... Seeing rates \>1

https://www.naacos.com/press-release--medicare-aco-participation-flat-in-2022 Drop in ACO's in 2019. So why did VBC pene increase in this year? So why did VBC pene increase in this year? Is COVID impactful?

# Regresion Models

### VBC Penetration by year

Negligible association, not clinically meaningful .002 \~= 7 seconds per resident per year In 2020-2021, increases to .005 \~= 17 seconds

Interesting that without adding state fixed effects, CNA casemix variable is not statistically significant.

```{r}
for (yr in 2017:2021) {
  filtered <- filter(merged, Year==yr)
  model <- lm(mean_cna_reported_hprd ~ VBCPenetration + mean_cna_cm_hprd, filtered)
  print(summary(model))
}
```

### VBC Penetration, Decomposed by Year

Still negligible. ACO pene loses stat signficance in 2019. This is the strange year where VBC penetration \> 1 in \~60 cases.

```{r}
for (yr in 2017:2021) {
  filtered <- filter(merged, Year==yr)
  model <- lm(mean_cna_reported_hprd ~ MAPenetration + ACOPenetration + mean_cna_cm_hprd, filtered)
  print(summary(model))
}
```

### VBC Decomposed with State / Year fixed effects

State / year fixed effects should prove influential for two reasons: - Regulatory environment varies by state and year. - COVID had a differential impact depending on statewide policies and cultural norms regarding vaccination.

\~10 second drop per pct point increase of MA penetration. ACO penetration is stat insignificant and tiny magnitude. Results are robust to incorporating case mix variable. In fact, the impact increases 3 seconds when adjusting for case-mix.

This is interesting. Reflects the differing levers in the MA program vs. the ACO program. MA payors contract directly with SNF's. A hospital based ACO's levers are indirect. May influence referral patterns, but only to the extent that they can influence hospital-based case managers. A physician based ACO probably has very limited influence in the SNF world.

```{r}
model <- lm(mean_cna_reported_hprd ~ MAPenetration + ACOPenetration + StateAbbrev.x:YearFactor + mean_cna_cm_hprd, merged)
print(summary(model))
```

### VBC Decomposed, total staffing

Total staffing initially dropping \~7 seconds with each point of MA penetration. Adjusting for case-mix makes magnitude even more negligible.

```{r}
model <- lm(mean_total_reported_hprd ~ MAPenetration + ACOPenetration + StateAbbrev.x:YearFactor + mean_total_cm_hprd, merged)
print(summary(model))
```

### Conclusions

Domain experts told me it was a nothing burger. My results support their claim that it's a nothing burger.

What counter-balancing forces may contribute to the outcome?

-   SNF's receiving pressure in Part A space, SNF admin responds by increasing services in Part B. AKA, the water balloon effect.
-   ACO's exert very little influence. Do not have many levers to pull in order to affect behavior. This outcome contradicts portrayals of SNF's as the "ACO piggybank".
-   Some SNF's may receive financial supports via integration in larger systems. For instance, a SNF onsite at a CCRC or Hospital may be subsidized against financial shocks.
-   MA contracting may be supportive in some respects. Depending on contract parameters, it may bring predictability that a FFS environment does not.
-   VBC penetration may actually increase concentration and adversely affect specific SNF's, but not decrease staffing levels on average.
